### 2019-09-02 README.Rmd
first commit

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
# Please put your title here to include it in the file below.
Title <- "Title of your paper goes here"
```

# kwl.pottery

[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh///master?urlpath=rstudio)

This repository contains the data and code for our paper:

> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>

Our pre-print is online here:

> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to cite

Please cite this compendium as:

> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to download or install

You can download the compendium as a zip from from this URL: </archive/master.zip>

Or you can install this compendium as an R package, kwl.pottery, from GitHub with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
remotes::install_github("/")
```

### Licenses

**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code :** See the [DESCRIPTION](DESCRIPTION) file

**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse

### Contributions

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

### 2019-09-02 analysis/paper/paper.Rmd
added sentences to introduction

---
title: "Exploring pottery specialization during the colonial contact in the 17-19th century in northeast Taiwan"
author:
  - Li-Ying Wang
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/"
)
```

# Introduction

Pottery production can reflect prehistoric socioeconomic patterns since it relates to not only economic base for households but also societies as whole when specialized production was organized that hints changes in production systems. Craft specialization can provide further evidence about the type of production organization, and the presence of elite control (Arnold 2000; Blackman, et al. 1993; Costin 1991; Hirshman, et al. 2010). The underlying assumption is that the specialized mass production of ceramics will lead to uniformity or homogeneity of the product due to increased skills, routinization, and lower diversity of producers (Arnold 2000: 334).

We observed pottery production in northeast Taiwan in the late Iron Age period and historical period (AD 600-100) presenting a high consistency in form throughout the region compared to other places in the island.

Here we examine pottery measurements and shapes from Kiwulan, a large multi-component archaeological site in NE Taiwan, to investigate if there are any changes resulting from different periods of foreign contact, European and then Chinese, that might indicate social changes in the indigenous society. We asked the questions: Did foreign contact have an impact on indigenous pottery production in Northeast Taiwan that can be detected in the shape of the vessels? Did the pottery production become more standardized and homogeneous in shape after foreign contact? Did the contact with European cause more homogeneous shape than Chinese contact period because of large scale exchange network? This is important because the

# Background and Research area

# Hypotheses and expectations

The goal for locally made pottery data collection is to explore whether pottery production could reflect the social change, from more corporate modes to more network modes after European contact. The level of craft specialization usually relates to the presence of elite control. In this case, if competitions for prestige and European resources among individuals gradually lead to change in social organization, then the production of pottery will show more homogeneous features after European contact due to the craft specialization. The standardization of metric attributes for pottery will be used to measure the craft specialization.

# Methods

In this study, I examine 291 pots recovered from Kiwulan site.  The layer from 1 to 6 could be divided into 3 time periods. Layer 5 and 6 represent pre-contact period, layer 4 represents contact period, and layer1, 2, and 3 represent post-contact period.  The amount of pots for each layer shows below. Although most pots are not complete, the thickness from rim, neck, to body can be measured. I have also measured the diameter of rim, neck, and body. For those pots that are incomplete, the diameter is measured by its curvature. Since the height is incomplete for most pots, this preliminary analysis focuses more on the possible change in thickness and diameter of pot for different parts, and their ratio over time.

# Results

```{r get-data, eval = FALSE}
# Note the path that we need to use to access our data files when rendering this document
my_data <- read.csv(here::here('analysis/data/raw_data/my_csv_file.csv'))
```

```{r demo-plot, fig.cap="A plot of random numbers"}
plot(rnorm(10))
```

Figure \@ref(fig:demo-plot) shows how we can have a caption and cross-reference for a plot

```{r demo-inline-code}
x <- round(pi, 2)
```

Here is an example of inline code `r x` in the middle of a sentence.

# Discussion

# Conclusion

# Acknowledgments

<!-- The following line inserts a page break when the output is MS Word. For page breaks in PDF, use \newpage on its own line.  -->
##### pagebreak

# References 
<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
<div id="refs"></div>

##### pagebreak

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```

### 2019-10-01 analysis/paper/paper.Rmd
test PAT
revise introdution and change the title

@@
 ---
-title: "Exploring pottery specialization during the colonial contact in the 17-19th century in northeast Taiwan"
+title: "Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeast Taiwan"
 author:
@@
 output: 
-    bookdown::html_document2:
+    bookdown::word_document2:
       fig_caption: yes
@@
 
+Culture contact between indigenous people and colonizer with imperial power usually leads to profound changes in local indigenous societies.
+We explore the culture changes in indigenous societies in northeast Taiwan that experienced foreign contacts with the Europeans and Chinese since 17th century by examining the local pottery production.
 Pottery production can reflect prehistoric socioeconomic patterns since it relates to not only economic base for households but also societies as whole when specialized production was organized that hints changes in production systems.
-Craft specialization can provide further evidence about the type of production organization, and the presence of elite control (Arnold 2000; Blackman, et al.
+Based on the assumption that the specialized mass production will lead to uniformity of the product due to increased skills, routinization, and lower diversity of producers, we may be able to determine the type of production, and the presence of elite control, from artefact analysis (Arnold 2000: 334; Blackman, et al.
 1993; Costin 1991; Hirshman, et al.
 2010).
-The underlying assumption is that the specialized mass production of ceramics will lead to uniformity or homogeneity of the product due to increased skills, routinization, and lower diversity of producers (Arnold 2000: 334).
 
-We observed pottery production in northeast Taiwan in the late Iron Age period and historical period (AD 600-100) presenting a high consistency in form throughout the region compared to other places in the island.
-
-Here we examine pottery measurements and shapes from Kiwulan, a large multi-component archaeological site in NE Taiwan, to investigate if there are any changes resulting from different periods of foreign contact, European and then Chinese, that might indicate social changes in the indigenous society.
+We observed pottery production in northeast Taiwan in the late Iron Age period and historical period (1400-1900 AD) presenting a high consistency in form and shape compared to other pottery found from other regions on this island that hints the emergence of pottery specialization.
 We asked the questions: Did foreign contact have an impact on indigenous pottery production in Northeast Taiwan that can be detected in the shape of the vessels?
@@
 Did the contact with European cause more homogeneous shape than Chinese contact period because of large scale exchange network?
-This is important because the
 
-# Background and Research area
+Pottery specialization can be identified by examining the standardization of pottery measurements (citation).
+However, traditional typological and linear measurements are limited because they can be insensitive to subtle variations resulting from changes in craft specialization.
+Taking a new approach to the measurement of craft specialization, we studied shape and standardization of locally made pottery to identify changes in pottery production at Kiwulan (1400-1900 AD), a large multi-component archaeological site in northeast Taiwan.
+We apply geometric morphometric methods to study artifact shapes to investigate if there are any changes resulting from foreign contact of European and Chinese that might indicate social changes in the indigenous society.
+
+We find differences in shape and shape standardization of pottery that indicate changes in pottery production resulting from foreign contact, suggesting increasing craft specialization and changes in local social organization at Kiwulan.
+These results are important to understand the influence of culture contact on local indigenous societies and answer the anthropological question that concerns the the mechanisms for social changes.
+In addition, our case study, which includes an openly available research compendium of R code suitable for use with any other assemblage, will help to expand the use of shape-based quantitative methods to questions about craft specialization and standardization in prehistoric technologies.
+
+# pottery shape analysis 
+- review about pottery shape analysis recently
+
+# Research area
+Kiwulan is located in the north part of Yilan plain in northeast Taiwan. Yilan is
 
@@
 
-In this study, I examine 291 pots recovered from Kiwulan site.
+In this study, we examine 291 pots recovered from Kiwulan site.
 The layer from 1 to 6 could be divided into 3 time periods.
@@
 
-# Acknowledgments
+# Acknowledgements
 
@@
 ```
+
+Word count: `r wordcountaddin::word_count()`

### 2019-10-03 analysis/paper/paper.Rmd
added paragraphs to method nd result section

@@
 # Research area
-Kiwulan is located in the north part of Yilan plain in northeast Taiwan. Yilan is
+Kiwulan is located at the north part of Yilan plain in northeast Taiwan.
 
 # Hypotheses and expectations
-
 The goal for locally made pottery data collection is to explore whether pottery production could reflect the social change, from more corporate modes to more network modes after European contact.
@@
 # Methods
+
+## metric measurements
 
@@
 
+## geometric morphometric approach
+
+To prepare the data for shape analysis, we first traced the outlines of pottery from previously outline drawings done by excavation team using Inkscape software.
+The digitised outlines were converted into a list of x-y pixel coordinates for elliptic Fourier analysis (EFA), which can assess pottery morphological difference among pottery outlines from three occupation contexts.
+The EFA generates harmonic coefficients were analysed by principal component analysis (PCA) to identity major patterns in data through dimensionality reduction.
+The principal components (PCs) were analysed with a multivariate analysis of variance (MANOVA) to test for significant effects of the occupation context on shape variances.
+Finally, we computed a significance test for the equality of coefficients of variation (CVs) among multiple groups of PCs to understand the differences in shape standardisation.
+Outline analyses were carried out using the package Momocs in the software R.
+
 # Results
+
+Elliptic Fourier analysis (EFA): Thin-plate splines compares the average shapes of vessels from each period to visualize the outline deformations required to pass from an extreme of one morphospace to the other.
+The results of PCA show that the first three PCs explain 86% of shape variation.
+PC1 relates to height of vessel, PC2 relates to width of neck and mouth, and PC3 relates to width and flare of neck.
+The PCA biplot of vessel shape shows grouping by each occupation phase.
+MANOVA shows significant differences in shape between Pre-European and Post-European (p = 0.02) and Pre-European and Chinese contact shapes (p < 0.01).
+The distributions of the first three PCs for each vessel suggest variations in shape standardization by period.
+PC1 shows more variation in the pre-Euro period compared to the Chinese period.
+Standardization appears strong in the pre-Euro period but weak in the Chinese period.
+P-values for a modified signed-likelihood ratio test of equality of CVs of PC1 & PC2 show significant differences in standardization of vessel shape across some periods, especially between Chinese contact with either pre-European or post-European.
 

### 2019-10-04 analysis/paper/paper.Rmd
add a paragraph about Kiwulan in the Materials section

@@
 
+- start with researches about shape analysis for artefacts and what is the implication.
+
 Culture contact between indigenous people and colonizer with imperial power usually leads to profound changes in local indigenous societies.
@@
 
-# Research area
-Kiwulan is located at the north part of Yilan plain in northeast Taiwan.
-
 # Hypotheses and expectations
@@
 
-# Methods
+# Materials and methods
+
+## Archaeological pottery from Kiwulan
+
+Kiwulan is an late Iron Age site situated on a hill near a riverside at the northern margin of Yilan, northeast Taiwan.
+Yilan is characterized by a triangular alluvial plain facing eastwards the Pacific and mountains on three other sides.
+Kiwulan contains two archaeological layers, Lower Culture Layer (1300 BP to 800 BP) and Upper Culture Layer (600 BP to 100 BP, i.e.
+AD 1400-1900), with a sterile deposit in between.
+The Upper Culture Layer is the main focus in this paper which covers 500 years from the late Iron Age to the historical period.
+This time period experienced frequent foreign contacts including the European colonial presence in the 17th century and great waves of Chinese immigrant since 19th century.
+The excavation revealed abundant artifacts in which potsherds are the dominate materials throughout the site.
+Imported ceramics from mainland China, stonewares, and ornament elements such as beads were usually found in the Upper Culture Layer that indicates the prosperous international trading activities in the 17th century during the European presence in Taiwan.
+In addition to artifacts, features were also found, such as burials, middens, and postholes with in-situ posts that explains Kiwulan was a settlement site.
 

### 2019-10-08 analysis/paper/paper.Rmd
added bibtex, and revised introduction and method section

--- 2019-10-03 analysis/paper/paper.Rmd
@@
 
+Ceramic production can reflect prehistoric socioeconomic patterns since it relates to not only economic base for households but also societies as a whole when specialized production was organized.
+Identifying the specialization in ceramic production will help to understand how complex societies form and further explore the underlying mechanism.
+One way to examine ceramic specialization is to identify whether ceramics present standardization in production, which based on the assumption that the specialized mass production will lead to uniformity of the product due to increased skills, routinization, and lower diversity of producers that enable us to interpret social organization or social relations such as type of production, and the presence of elite control [@Arnold2000, pp.
+334; @Stark1995; @Costin1991].
+
+Several variables have been proposed to measure ceramic standardization, including metric, compositional, and technological attributes [@Rice1991; @Blackman1993; @Costin1991; @Hirshman2010; @Tite1999].
+However, traditional typological and linear measurements are limited because they can be insensitive to subtle variations resulting from changes in craft specialization.
+Taking a new approach to the measurement of craft specialization, we studied shape and standardization of locally made pottery to identify changes in pottery production at Kiwulan (1400-1900 AD), a large multi-component archaeological site in northeast Taiwan.
+We apply geometric morphometric methods to study artifact shapes to investigate if there are any changes resulting from foreign contact of European and Chinese that might indicate social changes in the indigenous society.
+
 Culture contact between indigenous people and colonizer with imperial power usually leads to profound changes in local indigenous societies.
 We explore the culture changes in indigenous societies in northeast Taiwan that experienced foreign contacts with the Europeans and Chinese since 17th century by examining the local pottery production.
-Pottery production can reflect prehistoric socioeconomic patterns since it relates to not only economic base for households but also societies as whole when specialized production was organized that hints changes in production systems.
-Based on the assumption that the specialized mass production will lead to uniformity of the product due to increased skills, routinization, and lower diversity of producers, we may be able to determine the type of production, and the presence of elite control, from artefact analysis (Arnold 2000: 334; Blackman, et al.
-1993; Costin 1991; Hirshman, et al.
-2010).
-
 We observed pottery production in northeast Taiwan in the late Iron Age period and historical period (1400-1900 AD) presenting a high consistency in form and shape compared to other pottery found from other regions on this island that hints the emergence of pottery specialization.
@@
 Did the contact with European cause more homogeneous shape than Chinese contact period because of large scale exchange network?
-
-Pottery specialization can be identified by examining the standardization of pottery measurements (citation).
-However, traditional typological and linear measurements are limited because they can be insensitive to subtle variations resulting from changes in craft specialization.
-Taking a new approach to the measurement of craft specialization, we studied shape and standardization of locally made pottery to identify changes in pottery production at Kiwulan (1400-1900 AD), a large multi-component archaeological site in northeast Taiwan.
-We apply geometric morphometric methods to study artifact shapes to investigate if there are any changes resulting from foreign contact of European and Chinese that might indicate social changes in the indigenous society.
 
@@
 
-# Methods
+# Materials and methods
+
+## Archaeological pottery from Kiwulan
+
+Ceramics data analysed in this paper come from the Upper Culture Layer of Kiwulan, northeast Taiwan.
+Those ceramics dated from AD 1400 to AD 1900, which covers 500 years from the late Iron Age to the historical period with the onset of the European presence in the 17th century.
+Kiwulan is situated on a hill near a riverside at the northern margin of Yilan, which is characterized by a triangular alluvial plain facing eastwards the Pacific and mountains on three other sides.
+The Upper Culture Layer experienced frequent foreign contacts including the European colonial presence in the 17th century and great waves of Chinese immigrants since 19th century.
+The excavation revealed abundant artifacts in which potsherds are the dominate materials throughout the site.
+Imported ceramics from mainland China, stonewares, and ornament elements such as beads were usually found in the Upper Culture Layer that indicates the prosperous international trading activities in the 17th century during the European presence in Taiwan.
+In addition to artifacts, features were also found, such as burials, middens, and postholes with in-situ posts that explains Kiwulan was a large settlement site.
 
@@
 
-To prepare the data for shape analysis, we first traced the outlines of pottery from previously outline drawings done by excavation team using Inkscape software.
+The drawings of pottery with half cross-section from rim to body were imported into Inkscape software for outline digitalization.
+All half cross-section images were copied and flipped to create enclosed images in order to be ran in the R environment.
 The digitised outlines were converted into a list of x-y pixel coordinates for elliptic Fourier analysis (EFA), which can assess pottery morphological difference among pottery outlines from three occupation contexts.

### 2019-10-09 analysis/paper/paper.Rmd
added sentences to method section

@@
 
-# pottery shape analysis 
+# Shape analysis for artefacts
 - review about pottery shape analysis recently
+
+There are two different approaches for shape analysis, one is landmark analysis and another one is outline analysis.
+In this paper, we conducted outline analysis because it is useful to compare geometric structures where individual homologous landmarks are difficult to locate.
+
+GMM analysis: measuring standardisation through shape
 
@@
 
-## geometric morphometric approach
+## outline analysis approach
 
-The drawings of pottery with half cross-section from rim to body were imported into Inkscape software for outline digitalization.
-All half cross-section images were copied and flipped to create enclosed images in order to be ran in the R environment.
-The digitised outlines were converted into a list of x-y pixel coordinates for elliptic Fourier analysis (EFA), which can assess pottery morphological difference among pottery outlines from three occupation contexts.
-The EFA generates harmonic coefficients were analysed by principal component analysis (PCA) to identity major patterns in data through dimensionality reduction.
+The scanned cross-section drawings for pots covering from rim to bottom were acquired from Bureau of Cultural Affairs in Yilan.
+Scanned drawings were imported into Inkscape software for outline tracing to remove some additional information on the orginal drawings.
+Each finished half cross-section image was duplicated, flipped, and then joined with original one to create a complete cross-section image in order to be performed in the R software (www.rproject.org, Core-Team, 2015).
+Geometric morphometric analyses were conducted in R using the functions included in the Momocs, a package intended to quantify the shape and compare its variation, especially for outline analysis [@Bonhomme2014].
+The digitised outlines were converted into a list of x-y pixel coordinates for elliptic Fourier analysis (EFA), which assesses morphological differences among pottery shapes from three occupation contexts.
+EFA is powerful method to extract geometric information and decompose the shape into a series of trigonometric functions referred as harmonic coefficients or harmonics [@Claude2008; @Bonhomme2014].
+Harmonic..
+The EFA generates harmonic coefficients were analysed by principal component analysis (PCA) to identity major patterns of variation among the shape data through dimensionality reduction.
 The principal components (PCs) were analysed with a multivariate analysis of variance (MANOVA) to test for significant effects of the occupation context on shape variances.
-Finally, we computed a significance test for the equality of coefficients of variation (CVs) among multiple groups of PCs to understand the differences in shape standardisation.
-Outline analyses were carried out using the package Momocs in the software R.
+Finally, we computed a significance test to assess the equality of coefficients of variation (CVs) among multiple groups of PCs to understand the differences in shape standardisation.
+The R package cvequality (Version 0.2.0; Marwick and Krishnamoorthy 2019) in the CRAN repository was used to conduct the significance test.
 

### 2019-10-11 analysis/paper/paper.Rmd
added sentences to method section

--- 2019-10-08 analysis/paper/paper.Rmd
@@
 
-# pottery shape analysis 
-- review about pottery shape analysis recently
+# Geometric Morphometric approach: shape analysis for pottery specialization
+
+Geometric morphometry methods (GMM) has been widely applied recently in archaeology to study the morphological variability and similarity of archaeological materials to address the questions of anthropological interests.
+Different from taking linear measurements of objects as variables, geometric morphometrics methods use geometric coordinates of morphological structures as variables to specify the shapes [@Adams2004; @Lawing2010].
+According to different focuses, there are two common methods used to capture and analyze the shapes of objects, landmark analysis and outline analysis [@Adams2004].
+Landmark analysis
+
+- related to the issue of specialization
 
@@
 Ceramics data analysed in this paper come from the Upper Culture Layer of Kiwulan, northeast Taiwan.
-Those ceramics dated from AD 1400 to AD 1900, which covers 500 years from the late Iron Age to the historical period with the onset of the European presence in the 17th century.
+Those ceramics dated from AD 1400 to AD 1900, which covers 500 years from the late Iron Age to the historical period  by the European presence in the 17th century.
 Kiwulan is situated on a hill near a riverside at the northern margin of Yilan, which is characterized by a triangular alluvial plain facing eastwards the Pacific and mountains on three other sides.
@@
 
-## geometric morphometric approach
+## Outline analysis approach
 
-The drawings of pottery with half cross-section from rim to body were imported into Inkscape software for outline digitalization.
-All half cross-section images were copied and flipped to create enclosed images in order to be ran in the R environment.
-The digitised outlines were converted into a list of x-y pixel coordinates for elliptic Fourier analysis (EFA), which can assess pottery morphological difference among pottery outlines from three occupation contexts.
-The EFA generates harmonic coefficients were analysed by principal component analysis (PCA) to identity major patterns in data through dimensionality reduction.
-The principal components (PCs) were analysed with a multivariate analysis of variance (MANOVA) to test for significant effects of the occupation context on shape variances.
-Finally, we computed a significance test for the equality of coefficients of variation (CVs) among multiple groups of PCs to understand the differences in shape standardisation.
-Outline analyses were carried out using the package Momocs in the software R.
+The scanned pottery drawings for reconstructed pots that mostly covers from rim to bottom were acquired from Bureau of Cultural Affairs in Yilan.
+All drawing presents two-dimensional view of the section of a vessel with indications of metric measurements.
+The scanned drawings were imported into the Inkscape software for outlines tracing to remove some information such as marks, lines, and numbers on the original file.
+Each traced half cross-section image was duplicated, flipped, and then joined with another one to create a 2D closed outline that can be processed later in the R software (www.rproject.org, Core-Team, 2015).
+Geometric morphometric analyses were conducted using the functions included in the Momocs, a R package intended to quantify the shape and compare its variation, especially for outline analysis [@Bonhomme2014].
+The digitised outlines were converted into a list of successsive x-y pixel coordinates for elliptic Fourier analysis (EFA), which assesses morphological differences among pottery shapes from three occupation contexts.
+EFA is a powerful method to extract geometric information where closed outlines are considered as periodic functions and decomposed into a series of trigonometric functions referred as harmonic coefficients [@Claude2008; @Bonhomme2014].
+The number of harmonics determines the quality of the geometry that more harmonics can capture the shapes more precisely [@Bonhomme2013].
+Here we used 13 harmonics to describe pottery shapes.
+The harmonic coefficients generated by EFA were analysed by principal component analysis (PCA) to illustrate the diversity of the shape data and identify the major patterns of variation through dimensionality reduction.
+The principal components (PCs) scores were analysed with a multivariate analysis of variance (MANOVA) to test for significant effects of the groups of occupation context on shape variances.
+Finally, we computed a significance test to assess the equality of coefficients of variation (CVs) among multiple groups of PCs to understand the differences in shape standardisation among context groups.
+The R package cvequality (Version 0.2.0; Marwick and Krishnamoorthy 2019) in the CRAN repository was used to conduct the significance test for CVs.
 

### 2019-10-23 analysis/paper/paper.Rmd
updates for methods

@@
 
-Geometric morphometry methods (GMM) has been widely applied recently in archaeology to study the morphological variability and similarity of archaeological materials to address the questions of anthropological interests.
-Different from taking linear measurements of objects as variables, geometric morphometrics methods use geometric coordinates of morphological structures as variables to specify the shapes [@Adams2004; @Lawing2010].
-According to different focuses, there are two common methods used to capture and analyze the shapes of objects, landmark analysis and outline analysis [@Adams2004].
-Landmark analysis
+Geometric morphometry methods (GMM) has been widely applied recently in archaeology for shape analysis, which explores morphological variability and similarity of archaeological materials to address the questions of anthropological interests.
+Different from linear measurement approach that captures shapes by measuring distances, angles or ratios of objects, geometric morphometrics methods use Cartesian coordinates of morphological structures to define the shapes.
+Landmarks, curves or outlines of objects can be transformed into coordinates by their unique location and orientation with respect to numerical values of coordinate axes [@Adams2004; @Lawing2010; @Slice2007].
+According to different focuses and forms of original shape data, there are two common morphometric methods used to capture and analyze the shapes, landmark analysis and outline analysis [@Adams2004].
+Landmark approaches apply landmark points to specimen as variables that can be specified on coordinate system.
+Based on the points, objects are translated and scaled into the same orientation and size using Procrustes superimposition [@Lawing2010].
+This method commonly applied to zooarchaeology and
 
 - related to the issue of specialization
+
+- CV is hard to test, we apply some statistics testing for it
 
@@
 
-## Outline analysis approach
+## Outline analysis
 
-The scanned pottery drawings for reconstructed pots that mostly covers from rim to bottom were acquired from Bureau of Cultural Affairs in Yilan.
+```{r}
+# read in data
+# EFA
+```
+
+The scanned pottery drawings of reconstructed pots that mostly covers from rim to bottom were acquired from Bureau of Cultural Affairs in Yilan.
 All drawing presents two-dimensional view of the section of a vessel with indications of metric measurements.
-The scanned drawings were imported into the Inkscape software for outlines tracing to remove some information such as marks, lines, and numbers on the original file.
+The scanned drawings were imported into the Inkscape software for outlines tracing to remove some information such as marks, lines, and numbers on the original drawing.
 Each traced half cross-section image was duplicated, flipped, and then joined with another one to create a 2D closed outline that can be processed later in the R software (www.rproject.org, Core-Team, 2015).
@@
 The number of harmonics determines the quality of the geometry that more harmonics can capture the shapes more precisely [@Bonhomme2013].
-Here we used 13 harmonics to describe pottery shapes.
+Here we used \13\ harmonics to describe pottery shapes.
 The harmonic coefficients generated by EFA were analysed by principal component analysis (PCA) to illustrate the diversity of the shape data and identify the major patterns of variation through dimensionality reduction.

### 2019-10-29 analysis/paper/paper.Rmd
added code chucks for reading data and preparing data for EFA

@@
 
-```{r}
-# read in data
-# EFA
-```
-
 The scanned pottery drawings of reconstructed pots that mostly covers from rim to bottom were acquired from Bureau of Cultural Affairs in Yilan.
@@
 # Results
+
+```{r read-data}
+library(here)
+lf <- list.files(here("analysis", "data", "raw_data"), full.names = TRUE)
+
+library(xml2)
+library(stringr)
+remove_scan_and_fill <- function(svg_file) {
+  # remove the scan image from the SVG
+  t1 <- read_xml(svg_file)
+  t2 <- xml_children(t1)
+  scan_layer <- t2[[4]]
+  xml_remove(scan_layer)
+  # now we have a SVG without the scan layer
+
+  # change from outline to fill
+  stroke_and_fill_node <- xml_children(t1)[4]
+  stroke_and_fill_node_inner <- xml_children(stroke_and_fill_node)
+  xml_attr(stroke_and_fill_node_inner, "style") <- 
+    "fill:#000000;stroke:none;stroke-width:5;stroke-linecap:butt;stroke-linejoin:miter;stroke-miterlimit:4;stroke-dasharray:none;stroke-opacity:1"
+
+  write_xml(t1,
+            str_glue("analysis/data/raw_data/fill_{basename(svg_file)}"))
+
+  # fit image object to canvas in svg so it's all captured during the conversion to JPG
+  system(str_glue('inkscape --verb=EditSelectAll --verb=AlignHorizontalCenter --verb=AlignVerticalCenter  --verb=FitCanvasToSelectionOrDrawing --verb=FileSave --verb=FileQuit "$(pwd)/analysis/data/raw_data/fill_{basename(svg_file)}"'))
+}
+
+# run the function on all SVGs
+library(purrr)
+walk(lf, remove_scan_and_fill)
+
+# convert SVG to PNG
+filled_svgs <- list.files("analysis/data/raw_data", pattern = "^fill_", full.names = TRUE)
+
+# convert svgs to to jpg
+library(magick)
+library(rsvg)
+map(filled_svgs,
+  ~image_read_svg(.x) %>% 
+  image_background("white") %>%  
+  image_convert("jpg") %>% 
+  image_write(str_glue("{.x}_output.jpg")))
+```
+
+```{r tidy-data}
+# Once we have the jpgs, start from here
+library(tidyverse)
+
+# get list of jpgs only
+jpgs <- list.files("analysis/data/raw_data", pattern = ".jpg", full.names = TRUE)
+
+# join file names to phase
+pottery_data <- readr::read_csv("Pottery_outlines/KWL_Ceramic_final2018_with_phases.csv")
+pottery_data <- pottery_data %>% 
+  mutate(filename = str_glue('fill_{filename}.svg_output.jpg'))
+
+# some outlines that we have no phases for... 
+sdiff <- setdiff(pottery_data$filename, basename(jpgs))
+
+# the ones we do have both outlines and phases
+isect <- intersect(pottery_data$filename, basename(jpgs))
+
+jpgs <- jpgs[basename(jpgs) %in% isect]
+pottery_data <- 
+  pottery_data %>% 
+  filter(filename %in% isect) %>% 
+  distinct(filename, phase, .keep_all = TRUE)
+
+phases_isect <- 
+  pottery_data %>%  
+  select(filename, phase) %>% 
+  filter(filename %in% isect) %>% 
+  mutate_all(as.factor) %>% 
+  distinct(filename, phase) %>% 
+  mutate(phase = as.factor(case_when(
+    phase == 'post-e' ~ "Post-European",
+    phase == 'pre-e' ~ "Pre-European",
+    phase == 'ch-con' ~ "Post-European" # we combine Chinese and post-Euro
+  )))
+
+# how many pots in each phase?
+phases_isect %>% 
+  group_by(phase) %>% 
+  tally(sort = T)
+
+# start shape exploration
+library(Momocs)
+#  it freezes on the empty ones...
+jpgs_imported <- import_jpg(jpgs, threshold = 0.5)
+
+# Normalise the shapes, from 
+# https://www.repository.cam.ac.uk/bitstream/handle/1810/266423/Supplementary%20Code%20S1.pdf?sequence=20&isAllowed=y
+jpgs_imported_coo <- 
+  jpgs_imported %>% 
+   Out(., fac = phases_isect)  %>% 
+  # Centring, scaling and sampling pseudo-landmarks
+   coo_center %>% 
+   coo_scale %>% 
+   coo_sample(., 1000) %>% 
+  # Procrustes superimposition
+   fgProcrustes %>% 
+  # Normalisation of the starting points
+   coo_slidedirection
+
+jpgs_imported_coo %>% 
+  panel(., fac = phases_isect$phase)
+
+pile(jpgs_imported_coo)
+
+```
 

### 2019-10-29 analysis/paper/paper.Rmd
update code

@@
 library(here)
-lf <- list.files(here("analysis", "data", "raw_data"), full.names = TRUE)
+lf <- list.files(here("analysis", "data", "raw_data", "pottery_svg"), full.names = TRUE)
 
@@
   write_xml(t1,
-            str_glue("analysis/data/raw_data/fill_{basename(svg_file)}"))
+            str_glue("analysis/data/raw_data/fill_pottery_svg/fill_{basename(svg_file)}"))
 
   # fit image object to canvas in svg so it's all captured during the conversion to JPG
-  system(str_glue('inkscape --verb=EditSelectAll --verb=AlignHorizontalCenter --verb=AlignVerticalCenter  --verb=FitCanvasToSelectionOrDrawing --verb=FileSave --verb=FileQuit "$(pwd)/analysis/data/raw_data/fill_{basename(svg_file)}"'))
+  system(str_glue('inkscape --verb=EditSelectAll --verb=AlignHorizontalCenter --verb=AlignVerticalCenter  --verb=FitCanvasToSelectionOrDrawing --verb=FileSave --verb=FileQuit "$(pwd)/analysis/data/raw_data/fill_pottery_svg/fill_{basename(svg_file)}"'))
 }
@@
 # convert SVG to PNG
-filled_svgs <- list.files("analysis/data/raw_data", pattern = "^fill_", full.names = TRUE)
+filled_svgs <- list.files("analysis/data/raw_data/fill_pottery_svg", pattern = "^fill_", full.names = TRUE)
 
@@
 # get list of jpgs only
-jpgs <- list.files("analysis/data/raw_data", pattern = ".jpg", full.names = TRUE)
+jpgs <- list.files("analysis/data/raw_data/pottery_jpg", pattern = ".jpg", full.names = TRUE)
 
 # join file names to phase
-pottery_data <- readr::read_csv("Pottery_outlines/KWL_Ceramic_final2018_with_phases.csv")
+pottery_data <- readr::read_csv("analysis/data/raw_data/KWL_Ceramic_final2018_with_phases.csv")
 pottery_data <- pottery_data %>% 
@@
 pile(jpgs_imported_coo)
+```
+
+```{r EFA-analysis}
+op <- efourier(jpgs_imported_coo, norm = FALSE)
+
+# The progressive capture of shape geometry along the number of harmonics can be visualized with:
+
+calibrate_reconstructions_efourier(jpgs_imported_coo,  
+                          id = 1, 
+                          range = c(1, 2, 4, 8, 14, 20))
+
+# mean shape, per group
+bot.ms <- mshapes(op, 2)
+bot.ms$shp$`Post-European`   <- bot.ms$shp$`Post-European` %T>% coo_plot(border="blue")
+`Post-European`  <- bot.ms$shp$`Post-European`  %T>% coo_draw(border="blue")
+`Pre-European` <-   bot.ms$shp$`Pre-European`  %T>% coo_draw(border="red")
+legend("topright", 
+       lwd=1,
+       col=c("blue", "red"), 
+       legend=c( "Post-European", "Pre-European"))
+
+# Thin plate spline analysis (TPS)
+# Thin plate splines (TPS) was used to compare average shapes and visualise the outline deformations required to pass from an extreme of the morphospace to the other.
+# Deformation grids
+png("Pottery_outlines/tps-pre-Euro-to-post-Euro.png", res = 300, w = 900, h = 900)
+tps_grid(`Pre-European`, `Post-European` , shp.lwd = c(2, 2), shp.border = c("red", "blue"),
+         amp = 1, grid.size = 18, legend = T)
+dev.off()
+
+png("Pottery_outlines/tps-pre-Euro-to-Chinese.png", res = 300, w = 900, h = 900)
+tps_grid(`Pre-European`, `Chinese Contact`, shp.lwd = c(2, 2), shp.border = c("red", "blue"),
+         amp = 1, grid.size = 18, legend = T)
+dev.off()
+
+# Isodeformation lines
+tps_iso(`Pre-European`, `Post-European`,  shp.lwd = c(2, 2), shp.border = c("red", "blue"),
+        amp = 0.5, iso.nb = 10000, iso.levels = 12, legend.text = c("Pre-contact", "Post-contact"),
+        palette = col_summer, grid = F)
 

### 2019-11-01 analysis/paper/paper.Rmd
update of code

@@
 According to different focuses and forms of original shape data, there are two common morphometric methods used to capture and analyze the shapes, landmark analysis and outline analysis [@Adams2004].
-Landmark approaches apply landmark points to specimen as variables that can be specified on coordinate system.
-Based on the points, objects are translated and scaled into the same orientation and size using Procrustes superimposition [@Lawing2010].
-This method commonly applied to zooarchaeology and
+Landmark approaches apply landmark points to specimen as variables that can be specified on a coordinate system.
+According to the specific location of each point, objects are translated and scaled into the same orientation and size using Procrustes superimposition [@Lawing2010].
+This method can be applied to 3D objects not just limited to 2D that commonly applied to zooarchaeology and
 
@@
 Here we used \13\ harmonics to describe pottery shapes.
+
 The harmonic coefficients generated by EFA were analysed by principal component analysis (PCA) to illustrate the diversity of the shape data and identify the major patterns of variation through dimensionality reduction.
@@
   image_convert("jpg") %>% 
-  image_write(str_glue("{.x}_output.jpg")))
+  image_write(str_glue("analysis/data/raw_data/pottery_jpg/{basename(.x)}_output.jpg")))
 ```

### 2019-11-01 analysis/paper/paper.Rmd
Momocs code worked!

@@
 
-```{r read-data}
-library(here)
-lf <- list.files(here("analysis", "data", "raw_data", "pottery_svg"), full.names = TRUE)
-
-library(xml2)
-library(stringr)
-remove_scan_and_fill <- function(svg_file) {
-  # remove the scan image from the SVG
-  t1 <- read_xml(svg_file)
-  t2 <- xml_children(t1)
-  scan_layer <- t2[[4]]
-  xml_remove(scan_layer)
-  # now we have a SVG without the scan layer
-
-  # change from outline to fill
-  stroke_and_fill_node <- xml_children(t1)[4]
-  stroke_and_fill_node_inner <- xml_children(stroke_and_fill_node)
-  xml_attr(stroke_and_fill_node_inner, "style") <- 
-    "fill:#000000;stroke:none;stroke-width:5;stroke-linecap:butt;stroke-linejoin:miter;stroke-miterlimit:4;stroke-dasharray:none;stroke-opacity:1"
-
-write_xml(t1,
-            str_glue("analysis/data/raw_data/fill_pottery_svg/fill_{basename(svg_file)}"))
-
-  # fit image object to canvas in svg so it's all captured during the conversion to JPG
-  system(str_glue('inkscape --verb=EditSelectAll --verb=AlignHorizontalCenter --verb=AlignVerticalCenter  --verb=FitCanvasToSelectionOrDrawing --verb=FileSave --verb=FileQuit "$(pwd)/analysis/data/raw_data/fill_pottery_svg/fill_{basename(svg_file)}"'))
-}
-
-# run the function on all SVGs
-library(purrr)
-walk(lf, remove_scan_and_fill)
-
-# convert SVG to PNG
-filled_svgs <- list.files("analysis/data/raw_data/fill_pottery_svg", pattern = "^fill_", full.names = TRUE)
-
-# convert svgs to to jpg
-library(magick)
-library(rsvg)
-map(filled_svgs,
-  ~image_read_svg(.x) %>% 
-  image_background("white") %>%  
-  image_convert("jpg") %>% 
-  image_write(str_glue("analysis/data/raw_data/pottery_jpg/{basename(.x)}_output.jpg")))
-```
-
 ```{r tidy-data}
@@
 # get list of jpgs only
-jpgs <- list.files("analysis/data/raw_data/pottery_jpg", pattern = ".jpg", full.names = TRUE)
+jpgs <- list.files("analysis/data/raw_data/outline_jpg", pattern = ".jpg", full.names = TRUE)
 

### 2019-11-05 analysis/paper/paper.Rmd
update the method section

@@
 According to different focuses and forms of original shape data, there are two common morphometric methods used to capture and analyze the shapes, landmark analysis and outline analysis [@Adams2004].
-Landmark approaches apply landmark points to specimen as variables that can be specified on a coordinate system.
-According to the specific location of each point, objects are translated and scaled into the same orientation and size using Procrustes superimposition [@Lawing2010].
-This method can be applied to 3D objects not just limited to 2D that commonly applied to zooarchaeology and
+Landmark approaches assign a set of landmark points to specimen that can be specified on a coordinate system.
+The landmark coordinates are represented by numerical values with respect to defined coordinate axes.
+Based on the specific position of each point, objects are translated, rotated, and scaled into the same orientation and size using Procrustes superimposition technique to find the best fit between configuration shapes [@Lawing2010].
+Landmarks
+
+In archaeology, the landmarks are according to “type two” landmarks of Bookstein (1991), which define them as points located in the maximum of curvature or extreme points in morphology (Bookstein 1991; Zelditch et al.
+2004).
+
+were chosed  be applied to 3D objects not just limited to 2D that commonly applied to zooarchaeology and
+
+the analysis of shape coordinates generated by the least-squares superimposition of configurations of landmarks
 
@@
 The principal components (PCs) scores were analysed with a multivariate analysis of variance (MANOVA) to test for significant effects of the groups of occupation context on shape variances.
-Finally, we computed a significance test to assess the equality of coefficients of variation (CVs) among multiple groups of PCs to understand the differences in shape standardisation among context groups.
-The R package cvequality (Version 0.2.0; Marwick and Krishnamoorthy 2019) in the CRAN repository was used to conduct the significance test for CVs.
+Finally, we computed coefficients of variation statistic (CVs) among multiple groups of PCs with a significance test for the equality of CVs, which enable us to compare CVs in a statistical way (Karl Pearson 1896).
+The coefficient of variation is a common and widely-used statistical measure of the spread of a set of measurements of a sample.
+It is defined as the standard deviation divided by the mean in a ratio scale format.
+
+$$c_v = \frac{\sigma}{\mu}$$
+
+By standardizing standard deviation in each data set, the coefficients of variation statistics allows us to directly compare variation in samples measured with different units or means.
+Although it is a useful method, there is a question that how should we make a decision that the differences among the multiple CVs are significant or not?
+To answer this question, we need convenient methods for testing the equality of $c_v$ values from different samples, which has been described in many publications and here we proposed a significant test for CVs using the R package cvequality [@Marwick2019] in the CRAN repository.
+The package cvequality includes two tests, the "Asymptotic test for the equality of coefficients of variation from k populations" (Feltz and Miller 1996) and the "Modified signed-likelihood ratio test (SLRT) for equality of CVs" (Krishnamoorthy and Lee 2014).
+Feltz and Miller (1996) test is widely cited as an authoritative test for the equality of $c_v$ values that performs better than many approximative tests.
+It is a 'gold standard' test for comparing multiple $c_v$.
+Our implementation is based on the following from Feltz and Miller (1996):
+
+Let $s_i$ be the observed standard deviation of the $i$th sample (or group of measurements), $x_i$ be the observed mean of the $i$th sample, and let
+
+$$m_i=n_i-1$$
+
+and we do not know the population $c_v$, so let $\widehat{\tau }_c$ be an estimate as follows:
+
+$$\widehat{\tau }_c = \frac{\left( \sum _{j=1}^km_j\frac{S_j}{\bar{x}_j} \right)}{M}$$ 
+where 
+$$M=\sum _{j=1}^m m_j$$ 
+Then the test statistic can be computed with: 
+$$\begin{aligned} D'A D= \widehat{\tau }_c^{-2}\left( 0.5+\widehat{\tau }_c^2\right) ^{-1}\left[ \sum _{i=1}^km_i\left(\frac{s_i}{\bar{x}_i} \right)^2-\frac{\widehat{\tau }_c^2}{M} \right] \end{aligned}$$
+The $D'AD$ value measures how far each sample $c_v$ is from our estimate of the population $c_v$.
+Feltz and Miller (1996) note that the $D'AD$ value distributes as a central $\chi^2$ random variable with $k-1$ degrees of freedom, from which a p-value can be computed.
+
+We also include the Krishnamoorthy and Lee (2014) test as a more recent development with lower rates of type I error, better performance with uneven sample numbers, and more power across a range of conditions.
 

### 2019-11-05 analysis/paper/paper.Rmd
updates about code

@@
 
-In archaeology, the landmarks are according to “type two” landmarks of Bookstein (1991), which define them as points located in the maximum of curvature or extreme points in morphology (Bookstein 1991; Zelditch et al.
-2004).
-
-were chosed  be applied to 3D objects not just limited to 2D that commonly applied to zooarchaeology and
-
-the analysis of shape coordinates generated by the least-squares superimposition of configurations of landmarks
-
 - related to the issue of specialization  
-
 - CV is hard to test, we apply some statistics testing for it
@@
 
-```{r tidy-data}
+```{r tidy-data, eval=FALSE}
 # Once we have the jpgs, start from here
@@
 
-jpgs_imported_coo %>% 
-  panel(., fac = phases_isect$phase)
-
-pile(jpgs_imported_coo)
+saveRDS(jpgs_imported_coo,
+        here::here("analysis/data/derived_data/jpgs_imported_coo.rds"))
 ```
@@
 ```{r EFA-analysis}
+jpgs_imported_coo <- read_rds(here::here("analysis/data/derived_data/jpgs_imported_coo.rds"))
+
 op <- efourier(jpgs_imported_coo, norm = FALSE)
@@
 # Deformation grids
-png("Pottery_outlines/tps-pre-Euro-to-post-Euro.png", res = 300, w = 900, h = 900)
+png(here::here("analysis/figures/tps-pre-Euro-to-post-Euro.png"), res = 300, w = 900, h = 900)
 tps_grid(`Pre-European`, `Post-European` , shp.lwd = c(2, 2), shp.border = c("red", "blue"),
@@
 
-png("Pottery_outlines/tps-pre-Euro-to-Chinese.png", res = 300, w = 900, h = 900)
-tps_grid(`Pre-European`, `Chinese Contact`, shp.lwd = c(2, 2), shp.border = c("red", "blue"),
+png(here::here("analysis/figures/tps-pre-Euro-to-Chinese.png"), res = 300, w = 900, h = 900)
+tps_grid(`Pre-European`, `Post-European`, shp.lwd = c(2, 2), shp.border = c("red", "blue"),
          amp = 1, grid.size = 18, legend = T)
@@
 P-values for a modified signed-likelihood ratio test of equality of CVs of PC1 & PC2 show significant differences in standardization of vessel shape across some periods, especially between Chinese contact with either pre-European or post-European.
-
-```{r get-data, eval = FALSE}
-# Note the path that we need to use to access our data files when rendering this document
-my_data <- read.csv(here::here('analysis/data/raw_data/my_csv_file.csv'))
-```
-
-```{r demo-plot, fig.cap="A plot of random numbers"}
-plot(rnorm(10))
-```
-
-Figure \@ref(fig:demo-plot) shows how we can have a caption and cross-reference for a plot
-
-```{r demo-inline-code}
-x <- round(pi, 2)
-```
-
-Here is an example of inline code `r x` in the middle of a sentence.
 

### 2019-11-09 analysis/paper/paper.Rmd
update for the background section

@@
 Geometric morphometry methods (GMM) has been widely applied recently in archaeology for shape analysis, which explores morphological variability and similarity of archaeological materials to address the questions of anthropological interests.
-Different from linear measurement approach that captures shapes by measuring distances, angles or ratios of objects, geometric morphometrics methods use Cartesian coordinates of morphological structures to define the shapes.
-Landmarks, curves or outlines of objects can be transformed into coordinates by their unique location and orientation with respect to numerical values of coordinate axes [@Adams2004; @Lawing2010; @Slice2007].
+Different from linear measurement approach that captures shapes by measuring distances, angles or ratios of objects, geometric morphometrics methods use Cartesian coordinates of morphological structures to define the shapes [@Adams2004; @Lawing2010; @Slice2007].
+Landmarks, curves or outlines of objects can be transformed into coordinates by their unique location and orientation with respect to numerical values of coordinate axes .
 According to different focuses and forms of original shape data, there are two common morphometric methods used to capture and analyze the shapes, landmark analysis and outline analysis [@Adams2004].
-Landmark approaches assign a set of landmark points to specimen that can be specified on a coordinate system.
-The landmark coordinates are represented by numerical values with respect to defined coordinate axes.
-Based on the specific position of each point, objects are translated, rotated, and scaled into the same orientation and size using Procrustes superimposition technique to find the best fit between configuration shapes [@Lawing2010].
-Landmarks
+Landmark approaches assign a set of landmark points onto specimens that can be specified on a coordinate system.
+Each point has a coordinate represented by numerical values with respect to defined coordinate axes.
+Based on the distances and angles of interlandmarks, the objects are translated, rotated, and scaled into the same orientation and size using generalized Procrustes analysis (GPA).
+GPA method enables the comparison between shapes because it brings the landmark configurations of all objects into a common coordinate system, where differences are reflected based on landmark coordinate values [@Slice2007].
 

### 2019-11-09 analysis/paper/paper.Rmd
update for 11/08

@@
 # Hypotheses and expectations
-The goal for locally made pottery data collection is to explore whether pottery production could reflect the social change, from more corporate modes to more network modes after European contact.
-The level of craft specialization usually relates to the presence of elite control.
-In this case, if competitions for prestige and European resources among individuals gradually lead to change in social organization, then the production of pottery will show more homogeneous features after European contact due to the craft specialization.
-The standardization of metric attributes for pottery will be used to measure the craft specialization.
+The goal of my data collection for locally pottery is to explore whether pottery production could reflect the social change, from a more corporate mode to a more network mode after European contact.
+Craft specialization usually appears when the presence of elite control.
+In this case, if such feel of competitions for prestige and wonderful European resources among proud individuals gradually lead to change in social organization, then I expect the production of pottery will show more homogeneous features after European contact due to the craft specialization and the inspiration of the Europeans advisor.
+The standardization of metric attributes for pottery has being used to measure the craft specialization.
 

### 2019-11-11 analysis/paper/paper.Rmd
updates for backgrounds of GM

--- 2019-11-05 analysis/paper/paper.Rmd
@@
 Geometric morphometry methods (GMM) has been widely applied recently in archaeology for shape analysis, which explores morphological variability and similarity of archaeological materials to address the questions of anthropological interests.
-Different from linear measurement approach that captures shapes by measuring distances, angles or ratios of objects, geometric morphometrics methods use Cartesian coordinates of morphological structures to define the shapes.
-Landmarks, curves or outlines of objects can be transformed into coordinates by their unique location and orientation with respect to numerical values of coordinate axes [@Adams2004; @Lawing2010; @Slice2007].
-According to different focuses and forms of original shape data, there are two common morphometric methods used to capture and analyze the shapes, landmark analysis and outline analysis [@Adams2004].
-Landmark approaches assign a set of landmark points to specimen that can be specified on a coordinate system.
-The landmark coordinates are represented by numerical values with respect to defined coordinate axes.
-Based on the specific position of each point, objects are translated, rotated, and scaled into the same orientation and size using Procrustes superimposition technique to find the best fit between configuration shapes [@Lawing2010].
-Landmarks
+Different from linear measurement approaches that capture shapes by measuring distances, angles or ratios of objects, geometric morphometrics methods use Cartesian coordinates of morphological structures to define the shapes [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007].
+Landmarks, curves or outlines of objects can be represented by coordinates with their unique location and orientation with respect to numerical values of coordinate axes.
+According to different focuses and forms of original shape data, there are two common morphometric methods used to capture and analyse the shapes, landmark approaches and outline approaches [@Adams2004].
+Landmark approaches assign a set of landmarks onto specimens as reference points that can be specified on a coordinate system as x, y coordinates for two dimensions or x, y, z coordinates for three dimensions.
+Based on the distances and angles of landmark points, the objects are translated, rotated, and scaled into the same orientation and size using generalized Procrustes analysis (GPA) [@Bookstein1991].
+GPA method extracts shape information and compares shapes by taking sets of landmark configurations into a common coordinate system, where coordinate values represent shape variables that can be further analysed by further multivariate statistical analyses [@Slice2007].
+Landmark approach is a useful method for archaeological objects with extreme morphological signature or distinctive curvature that provide suitable locations for landmark placement, such as projectile point tips or biologically meaningful points indicating osteological differences [@Birch2019; @Buchanan2019; @Lycett2013; @Cardillo2010; @Haruda2019; @Meloro2015].
+
+However, landmark approaches have a limitation that a set of landmarks may not be able to capture the shape differences of a morphological structure when the curve between landmarks conveys more important information than landmarks themselves.
+In addition, suitable locations for landmark placement may not be available for some complex or rounded shaped structures, such as pottery.
 
@@
 # Hypotheses and expectations
-The goal for locally made pottery data collection is to explore whether pottery production could reflect the social change, from more corporate modes to more network modes after European contact.
-The level of craft specialization usually relates to the presence of elite control.
-In this case, if competitions for prestige and European resources among individuals gradually lead to change in social organization, then the production of pottery will show more homogeneous features after European contact due to the craft specialization.
-The standardization of metric attributes for pottery will be used to measure the craft specialization.
+
+My hypothesis is that if foreign contacts including European and Chinese had impact on the emergence of social inequality in local indigenous society due to the monopolies for trade between a small number of indigenous people, then I expect to see the changes in social organization from a more corporate to a more network organization that can be reflected by pottery production at Kiwulan.
+The emergence of craft specialization, here pottery specialization, are usually related to the presence of elite control.
+In this case, if the competitions for foreign resources and being trade partners of European or Chinese colonizer among individuals gradually lead to the emergence of social inequality, then I expect the local pottery will show more homogeneous features after contact due to the pottery specialization caused by control of small group of individuals.
 

### 2019-11-19 analysis/paper/paper.Rmd
add semilandmark and outline approaches, specialization of pottery

@@
 
-Geometric morphometry methods (GMM) has been widely applied recently in archaeology for shape analysis, which explores morphological variability and similarity of archaeological materials to address the questions of anthropological interests.
-Different from linear measurement approaches that capture shapes by measuring distances, angles or ratios of objects, geometric morphometrics methods use Cartesian coordinates of morphological structures to define the shapes [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007].
+Geometric morphometric methods (GMM) has been widely applied recently in archaeology for shape analysis, which explores morphological variability and similarity of archaeological materials to address the questions of anthropological interests.
+Different from linear measurement approaches that capture shapes by measuring length, width or ratios of objects, geometric morphometrics methods use Cartesian coordinates of morphological structures to define the shapes [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007].
 Landmarks, curves or outlines of objects can be represented by coordinates with their unique location and orientation with respect to numerical values of coordinate axes.
-According to different focuses and forms of original shape data, there are two common morphometric methods used to capture and analyse the shapes, landmark approaches and outline approaches [@Adams2004].
-Landmark approaches assign a set of landmarks onto specimens as reference points that can be specified on a coordinate system as x, y coordinates for two dimensions or x, y, z coordinates for three dimensions.
-Based on the distances and angles of landmark points, the objects are translated, rotated, and scaled into the same orientation and size using generalized Procrustes analysis (GPA) [@Bookstein1991].
-GPA method extracts shape information and compares shapes by taking sets of landmark configurations into a common coordinate system, where coordinate values represent shape variables that can be further analysed by further multivariate statistical analyses [@Slice2007].
-Landmark approach is a useful method for archaeological objects with extreme morphological signature or distinctive curvature that provide suitable locations for landmark placement, such as projectile point tips or biologically meaningful points indicating osteological differences [@Birch2019; @Buchanan2019; @Lycett2013; @Cardillo2010; @Haruda2019; @Meloro2015].
+According to different focuses and nature of original shape data, there are two common morphometric methods used to capture and analyse the shapes, landmark approaches and outline approaches [@Adams2004].
+Landmark approaches assign a set of landmarks onto objects as reference points that can be specified on a coordinate system as x, y coordinates for two dimensions or x, y, z coordinates for three dimensions.
+Based on raw landmarks, objects are translated to a common centroid, rescaled into the same size, and rotated until the summed squared distances between corresponding landmarks is minimized using generalized Procrustes analysis (GPA) [@Bookstein1991].
+GPA superimposes sets of landmark configurations into a common coordinate system, where superimposed landmark coordinates are used as shape variables that can be further analysed by multivariate statistics [@Slice2007].
+Landmark-based morphometrics are widely used for archaeological objects with extreme morphological signature or distinctive points that provide suitable locations for landmark placement, such as projectile point tips or biologically meaningful points indicating osteological differences [@Birch2019; @Buchanan2019; @Lycett2013; @Cardillo2010; @Haruda2019; @Meloro2015].
 
-However, landmark approaches have a limitation that a set of landmarks may not be able to capture the shape differences of a morphological structure when the curve between landmarks conveys more important information than landmarks themselves.
-In addition, suitable locations for landmark placement may not be available for some complex or rounded shaped structures, such as pottery.
+However, one limitation of landmark-based morphometrics is that landmark data may not be able to capture the shape differences of a morphological structure of which the curving outlines between landmarks convey important information.
+In addition, obvious and reliable landmark points may be unavailable for some complex, rounded or globular shaped structures.
+In those cases, outline approaches are more suitable to study shape by focusing more on on the overall shape of a morphological structure.
+One of the outline approaches is semi-landmarks method, which assign points along a outline curve at defined intervals between two landmarks to capture the variations of the outlines [@Bookstein1997; @Lawing2010].
+Another approach, also the method used in this paper, is elliptic Fourier Analysis (EFA) commonly used for two-dimensional closed shape [@Kuhl1982].
+EFA is a powerful method to extract geometric information where outlines are considered as periodic functions and decomposed into a series of ellipses described by trigonometric functions, called harmonic coefficients [@Adam2004; @Claude2008; @Bonhomme2014].
+The number of harmonics determines the quality of the geometry that more harmonics can capture the shapes more precisely.
+The use of EFA in archaeology are mostly for bone morphology to understand the biological variation and recently more for stone artefacts to explore stone tool technology [@Fox2015; @Hoggard2019; @Iovictua2010].
+Outline approach based on EFA is more suitable for shapes lacking of obvious landmarks or curves convey more important information indicating the meaningful variation in shapes, for example, ceramics.
+@Wilczek2014 evaluate the concordance between outline-based approaches, including the EFA and the Discrete Cosine Transform (DCT), and traditional typology by studying 154 complete ceramic vessels in varied forms from the Bibracte oppidum in France.
+They found that the variation in ceramic shapes analysed by EFA and DCT matches traditional ceramic typology, which supports that outline-based approaches can be efficiently used for studying variations in ceramic forms.
+EFA enables the study of the level of production standardisation to answer questions about the evolution of shape over space and time.
 
-- related to the issue of specialization  
+In this paper, we use EFA for pottery shape analysis to evaluate the level of standardization of pottery from an Iron Age site in northeast Taiwan in relation to the European presence in the 17th century that might hint the emergence of craft specialization for pottery production.
+Craft specialization can provide further evidence about the type of production organization which is closely related to economic or political aspects of a society [@Arnold2000; @Blackman1993; @Costin1991; @Hirshman2010; @Stark1995].
+The emergence of craft specialization can also be an indicator for social inequality that hints the possibility of the presence of production groups worked for specific individuals.
+The presence of craft specialization is commonly studied by measuring the standardization of ceramics based on the assumption that the specialized mass production will lead to uniformity or homogeneity of the product due to increased skills, routinization, and lower diversity of producers [@Arnold2000].
+To test the prediction of a shift in the level of social inequality in this case study in northeast Taiwan influnced by the colonial European presence, we use standardization of pottery in shapes as an index for craft specialization to identify the presence of a few individuals controlling large scale pottery production [@Costin1995; @Hirshman2010] using outline-based geometric morphometrics couples with traditional metric measurements to compare and evaluate the use of outline approaches for pottery with high consistent in forms.
+
 - CV is hard to test, we apply some statistics testing for it
@@
 The digitised outlines were converted into a list of successsive x-y pixel coordinates for elliptic Fourier analysis (EFA), which assesses morphological differences among pottery shapes from three occupation contexts.
-EFA is a powerful method to extract geometric information where closed outlines are considered as periodic functions and decomposed into a series of trigonometric functions referred as harmonic coefficients [@Claude2008; @Bonhomme2014].
-The number of harmonics determines the quality of the geometry that more harmonics can capture the shapes more precisely [@Bonhomme2013].
 Here we used \13\ harmonics to describe pottery shapes.
@@
    coo_sample(., 1000) %>% 
-  # Procrustes superimposition
+  # Procrustes superimposition for pseudo-landmarks
    fgProcrustes %>% 
@@
 
+# take a look when they stack together
+stack(jpgs_imported_coo)
+
+# fit the x and y coordinates separately
+coo_oscillo(jpgs_imported_coo[1], "efourier")
+
+# default is 13 harmanics that gather 99% of the harmonic power 
 op <- efourier(jpgs_imported_coo, norm = FALSE)
@@
 # The progressive capture of shape geometry along the number of harmonics can be visualized with:
-
 calibrate_reconstructions_efourier(jpgs_imported_coo,  
@@
 bot.ms <- mshapes(op, 2)
-bot.ms$shp$`Post-European`   <- bot.ms$shp$`Post-European` %T>% coo_plot(border="blue")
-`Post-European`  <- bot.ms$shp$`Post-European`  %T>% coo_draw(border="blue")
-`Pre-European` <-   bot.ms$shp$`Pre-European`  %T>% coo_draw(border="red")
+bot.ms$shp$`Post-European` <- bot.ms$shp$`Post-European`  %>% coo_plot(border="blue")
+`Post-European`  <- bot.ms$shp$`Post-European`  %>% coo_draw(border="blue")
+`Pre-European` <-   bot.ms$shp$`Pre-European`  %>% coo_draw(border="red")
 legend("topright", 
@@
        legend=c( "Post-European", "Pre-European"))
+
+# or from mshapes directly
+jpgs_imported_coo %>% efourier(6) %>% mshapes(~phase) %>% plot_mshapes()
 

### 2019-11-21 analysis/paper/paper.Rmd
added code chunk for PCA, MANOVA, and CV tests

@@
 The presence of craft specialization is commonly studied by measuring the standardization of ceramics based on the assumption that the specialized mass production will lead to uniformity or homogeneity of the product due to increased skills, routinization, and lower diversity of producers [@Arnold2000].
-To test the prediction of a shift in the level of social inequality in this case study in northeast Taiwan influnced by the colonial European presence, we use standardization of pottery in shapes as an index for craft specialization to identify the presence of a few individuals controlling large scale pottery production [@Costin1995; @Hirshman2010] using outline-based geometric morphometrics couples with traditional metric measurements to compare and evaluate the use of outline approaches for pottery with high consistent in forms.
+To test the prediction of a shift in the level of social inequality in this case study in northeast Taiwan influenced by the colonial European presence, we use standardization of pottery in shapes as an index for craft specialization to identify the presence of a few individuals controlling large scale pottery production [@Costin1995; @Hirshman2010] using outline-based geometric morphometrics couples with traditional metric measurements to compare and evaluate the use of outline approaches for pottery with high consistent in forms.
 
@@
         palette = col_summer, grid = F)
+```
+
+```{r PCA-analysis}
+# pca
+# these are fast
+op.p <- PCA(op)
+
+# We can see the pots groups by phase, and their shape distribution
+# according to the first two principle components of the EFA 
+# PC1 and PC2 explain 79% of variation
+png("index/data/pottery-shape/pca-biplot.png", res = 300, w = 1500, h = 1500)
+plot(op.p, 2, 
+     morpho = TRUE, 
+     ellipses = TRUE,
+     cex = 1, 
+     cex.labelsgroups = 1.5)
+
+dev.off()
+
+# different visualization 
+plot(PCA(op), 2, pos.shp="full")
+plot(PCA(op), 2, pos.shp="range")
+plot(PCA(op), 2, pos.shp="xy")
+plot(PCA(op), 2, pos.shp="range_axes")
+plot(PCA(op), 2, pos.shp="full_axes")
+
+# the contributions of each PC to the shapes
+pc_contrib <- PCcontrib(op.p, nax = 1:3) 
+pc_contrib$gg +  
+  geom_polygon(fill="slategrey", col="black") + 
+  ggtitle("Shape variation along PC axes") +
+  theme_minimal(base_size = 6)
+
+ggsave("index/data/pottery-shape/pca-contrib.png")
+scree_plot(op.p)
+
+# PC1 is height of vessel
+# PC2 is neck constriction
+# PC3 is height of neck and roundness of body
+```
+
+```{r LDA-analysis}
+# lda
+
+# we performed a linear discriminant analysis (LDA) based on the new shape variables (PCs), with a leave-one-out cross-validation procedure, to identify the linear combination of shape features that was able to discriminate between phases
+
+# we work on PCA scores
+bot.l <- LDA(op.p, 2)
+# print a summary, along with the leave-one-out cross-validation table. 
+bot.l 
+# plot.LDA works pretty much with the same grammar as plot.PCA
+# here we only have one LD
+plot(bot.l)
+# plot the cross-validation table
+plot_CV(bot.l)
+
+# Does not produce an efficient separation between groups and a cross-validation (leave-one-out) at phase level showing a low confidence in the reclassification.
+But best for Pre-e, indicating these have the most distinctive shapes
+
+#--------------------------------------------------
+# Unsupervised model-based clustering
+
+# what are the best grouping of the pots by shape?
+library(mclust)
+
+BIC <- mclustBIC(op.p$x[ , 1:5])
+mod1 <- Mclust(op.p$x[ , 1:5], x = BIC)
+summary(mod1, parameters = TRUE)
+plot(mod1, what = "classification")
+
+# no clusters apparent 
+```
+
+```{r MANOVA}
+# MANOVA: Multivariate Analysis of (co)variace
+# We can test for a difference in the distribution of PC scores by calculating a pairwise combination between every levels of a fac.
+
+manv <-   MANOVA_PW(op.p, "phase")
+manv$summary
+
+#----------------------------------------------------
+# GAMMS from https://www-nature-com.offcampus.lib.washington.edu/articles/s41598-018-20122-9
+
+# Generalised additive mixed models (GAMMs) were used to explain shape variance with respect to mean environmental parameters and shell size, and to compare between individual shape features (PCs).
+A GAMM is a generalised linear mixed model with a linear predictor involving a sum of smooth functions of covariates.
+The model allowed for flexible specification of the dependence of the response on the covariates by defining regression splines (smoothing functions) and estimating their optimal degree of smoothness, rather than calculating parametric relationships
+
+ggplot(pottery_data,
+       aes(phase,
+           as.numeric(Thick1_Neck))) +
+  geom_boxplot()
+
+pottery_data$thick_neck1 <- as.numeric(pottery_data$Thick1_Neck)
+pottery_data$thick_body1 <- as.numeric(pottery_data$Thick1_Body)
+pottery_data$height <- as.numeric(pottery_data$Hight)
+
+plot(op.p$x[ , 2], pottery_data$thick_body1)
+
+gams_input <- data.frame(thick_body1 = pottery_data$thick_body1 , 
+                         height = pottery_data$height, 
+                         shape_pc1 = op.p$x[ , 1])
+
+library(brms)
+m2 <- brm(bf(shape_pc1 ~ s(thick_body1)),
+          data = gams_input, family = gaussian(), cores = 4, seed = 17,
+          iter = 4000, warmup = 1000, thin = 10, refresh = 0,
+          control = list(adapt_delta = 0.99))
+summary(m2) # so many zeros... not very promising 
+# plot marginal effects for each predictor
+plot(marginal_effects(m2), ask = FALSE)
+loo(m2)
+pp_check(m2)
+```
+
+```{r CV-test}
+# CVs on shape PCs for standardisation
+
+len <- length(op.p$x[,1])
+# plot PC1-3 distribution by phase
+pc_plot <- data_frame(pc = c(op.p$x[,1],
+                             op.p$x[,2],
+                             op.p$x[,3]),
+                      phase = rep(op.p$fac$phase, 3),
+                        pcn = str_glue('PC{sort(rep(1:3, len))}'))
+
+pc_plot$phase <- fct_relevel(pc_plot$phase, 
+                            c("Pre-European", 
+                              "Post-European"))
+
+ggplot(pc_plot,
+       aes(phase,
+           pc)) +
+  geom_violin() +
+  ggforce::geom_sina() +
+  xlab("") +
+  facet_wrap(~ pcn, ncol = 1) +
+  theme_minimal()
+ggsave("index/data/pottery-shape/pcs-boxplot-for-shape-variation.png",
+       h = 10, w = 5)
+
+# what are the variences for each PC in each period?
+# we don't use CV because that is not good for measurements with no
+# absolute zero 
+pc_plot %>% 
+  group_by(pcn, phase) %>% 
+  summarise(variance = var(pc))
+
+# statistical test
+pc_plot_tests <- 
+pc_plot %>% 
+  nest_legacy(pc, phase) %>% 
+  mutate(var_test = purrr::map(data, ~broom::tidy(var.test(.x$pc, .x$phase)))) %>% 
+  unnest_legacy(var_test)
+
+pc_plot_tests <- 
+  pc_plot %>% 
+  group_by_at(vars(-pc)) %>% 
+  mutate(i = row_number()) %>% 
+  ungroup() %>% 
+  mutate(phase = as.character(phase),
+         pcn = as.character(pcn)) %>% 
+  spread(phase, pc) %>% 
+  select(-i) %>% 
+  nest_legacy(-pcn) %>% 
+  mutate(var_test = purrr::map(data, ~broom::tidy(var.test(.x$`Post-European` ,
+                                            .x$`Pre-European`)))) %>% 
+  unnest(var_test)
+
+library(cvequality)
+pc1 <- data.frame(
+x  = op.p$x[,1],
+y = as.character(op.p$fac$phase),
+stringsAsFactors = F)
+
+cv_test_pc1 <- mslr_test(nr = 1000, pc1$x, pc1$y)
+
+pc2 <- data.frame(
+  x  = op.p$x[,2],
+  y = as.character(op.p$fac$phase),
+  stringsAsFactors = F)
+
+cv_test_pc2 <- mslr_test(nr = 1000, pc2$x, pc2$y)
+
+pc3 <- data.frame(
+  x  = op.p$x[,3],
+  y = as.character(op.p$fac$phase),
+  stringsAsFactors = F)
+
+cv_test_pc3 <- mslr_test(nr = 1000, pc3$x, pc3$y)
+
+# do this pairwise
+cv_test_pc1.1 <- 
+  mslr_test(nr = 1000, 
+            pc1$x[pc1$y %in% c("Post-European", 
+                               "Pre-European")], 
+            pc1$y[pc1$y %in% c("Post-European", 
+                               "Pre-European")])
+
+cv_test_pc1.2 <- 
+  mslr_test(nr = 1000, 
+            pc1$x[pc1$y %in% c("Chinese Contact", 
+                               "Pre-European")], 
+            pc1$y[pc1$y %in% c("Chinese Contact", 
+                               "Pre-European")])
+
+cv_test_pc1.3 <- 
+  mslr_test(nr = 1000, 
+            pc1$x[pc1$y %in% c("Chinese Contact", 
+                               "Post-European")], 
+            pc1$y[pc1$y %in% c("Chinese Contact", 
+                               "Post-European")])
+
+PC1_cv_tests <- 
+bind_rows(cv_test_pc1.1, cv_test_pc1.2, cv_test_pc1.3) %>% 
+  bind_cols(phases = c("Post-European vs Pre-European",
+              "Chinese Contact vs Pre-European",
+              "Chinese Contact vs Post-European"))
+
+cv_test_pc2.1 <- 
+  mslr_test(nr = 1000, 
+            pc2$x[pc2$y %in% c("Post-European", 
+                               "Pre-European")], 
+            pc2$y[pc2$y %in% c("Post-European", 
+                               "Pre-European")])
+
+cv_test_pc2.2 <- 
+  mslr_test(nr = 1000, 
+            pc2$x[pc2$y %in% c("Chinese Contact", 
+                               "Pre-European")], 
+            pc2$y[pc2$y %in% c("Chinese Contact", 
+                               "Pre-European")])
+
+cv_test_pc2.3 <- 
+  mslr_test(nr = 1000, 
+            pc2$x[pc2$y %in% c("Chinese Contact", 
+                               "Post-European")], 
+            pc2$y[pc2$y %in% c("Chinese Contact", 
+                               "Post-European")])
+
+PC2_cv_tests <- 
+  bind_rows(cv_test_pc2.1, 
+            cv_test_pc2.2, 
+            cv_test_pc2.3) %>% 
+  bind_cols(phases = c("Post-European vs Pre-European",
+                       "Chinese Contact vs Pre-European",
+                       "Chinese Contact vs Post-European"))
+
+bind_rows(PC1_cv_tests, 
+          PC2_cv_tests, .id = "PC") %>% 
+  mutate(PC = str_glue('PC{PC}')) %>% 
+  mutate(colour = if_else(p_value < 0.05, "red", "black")) %>% 
+  ggplot(aes(str_wrap(phases, 40),
+             p_value,
+             colour = I(colour),
+             shape = PC)
+         ) +
+  geom_point(size = 8,  position = position_dodge(width = 0.5)) +
+  theme_minimal(base_size = 12) +
+  coord_cartesian(clip = 'off') +
+  xlab("") +
+  ylab("CV equality test p-values")
+
+ggsave("index/data/pottery-shape/cv-test-on-pcs.png", h = 3, w = 12)
 

### 2019-12-08 analysis/paper/paper.Rmd
some editting

@@
 
+A total of 73 vessels from pre-contact contexts (n = 32), post-European contexts (n = 27), and Chinese contact contexts (n = 14) were selected for this study because of their completeness mostly covering parts from rim to bottom.
+The scanned pottery drawings of those pots were acquired from Bureau of Cultural Affairs in Yilan.
+All drawing presents two-dimensional view of the section of a vessel with indications of metric measurements.
+The scanned drawings were imported into the Inkscape software for outlines tracing to remove additional information such as marks, lines, and measurements on the original drawings.
+Each traced half cross-section image was duplicated, flipped, and then joined with another one to create a 2D closed outline.
+
 ## metric measurements
@@
 
-The scanned pottery drawings of reconstructed pots that mostly covers from rim to bottom were acquired from Bureau of Cultural Affairs in Yilan.
-All drawing presents two-dimensional view of the section of a vessel with indications of metric measurements.
-The scanned drawings were imported into the Inkscape software for outlines tracing to remove some information such as marks, lines, and numbers on the original drawing.
-Each traced half cross-section image was duplicated, flipped, and then joined with another one to create a 2D closed outline that can be processed later in the R software (www.rproject.org, Core-Team, 2015).
-Geometric morphometric analyses were conducted using the functions included in the Momocs, a R package intended to quantify the shape and compare its variation, especially for outline analysis [@Bonhomme2014].
+Geometric morphometric analyses were conducted in the R software (www.rproject.org, Core-Team, 2015) using the functions included in the Momocs, a R package intended to quantify the shape and compare its variation, especially for outline analysis [@Bonhomme2014].
 The digitised outlines were converted into a list of successsive x-y pixel coordinates for elliptic Fourier analysis (EFA), which assesses morphological differences among pottery shapes from three occupation contexts.
-Here we used \13\ harmonics to describe pottery shapes.
 
@@
 ```
+
+The elliptic Fourier coefficients of 73 pottery from three different phases were calculated.
+Reliable pottery outline was captured by 13 harmonics that gather 99 % of the total harmonic power.
+Thin-plate splines compares the average shapes of vessels from each period to visualize the outline deformations required to pass from an extreme of one morphospace to the other.
 
@@
 
-Elliptic Fourier analysis (EFA): Thin-plate splines compares the average shapes of vessels from each period to visualize the outline deformations required to pass from an extreme of one morphospace to the other.
 The results of PCA show that the first three PCs explain 86% of shape variation.

### 2019-12-08 analysis/paper/paper.Rmd
typos fixed

@@
 Landmark approaches assign a set of landmarks onto objects as reference points that can be specified on a coordinate system as x, y coordinates for two dimensions or x, y, z coordinates for three dimensions.
-Based on raw landmarks, objects are translated to a common centroid, rescaled into the same size, and rotated until the summed squared distances between corresponding landmarks is minimized using generalized Procrustes analysis (GPA) [@Bookstein1991].
+Based on raw landmarks, objects are translated to a common centroid, rescaled into the same size, and rotated until the summed squared distances between corresponding landmarks are minimized using generalized Procrustes analysis (GPA) [@Bookstein1991].
 GPA superimposes sets of landmark configurations into a common coordinate system, where superimposed landmark coordinates are used as shape variables that can be further analysed by multivariate statistics [@Slice2007].
@@
 In addition, obvious and reliable landmark points may be unavailable for some complex, rounded or globular shaped structures.
-In those cases, outline approaches are more suitable to study shape by focusing more on on the overall shape of a morphological structure.
-One of the outline approaches is semi-landmarks method, which assign points along a outline curve at defined intervals between two landmarks to capture the variations of the outlines [@Bookstein1997; @Lawing2010].
+In those cases, outline approaches are more suitable to study shape by focusing more on the overall shape of a morphological structure.
+One of the outline approaches is semi-landmarks method, which assigns points along an outline curve at defined intervals between two landmarks to capture the variations of the outlines [@Bookstein1997; @Lawing2010].
 Another approach, also the method used in this paper, is elliptic Fourier Analysis (EFA) commonly used for two-dimensional closed shape [@Kuhl1982].
@@
 The number of harmonics determines the quality of the geometry that more harmonics can capture the shapes more precisely.
-The use of EFA in archaeology are mostly for bone morphology to understand the biological variation and recently more for stone artefacts to explore stone tool technology [@Fox2015; @Hoggard2019; @Iovictua2010].
-Outline approach based on EFA is more suitable for shapes lacking of obvious landmarks or curves convey more important information indicating the meaningful variation in shapes, for example, ceramics.
+The use of EFA in archaeology is mostly for bone morphology to understand the biological variation and recently more for stone artefacts to explore stone tool technology [@Fox2015; @Hoggard2019; @Iovictua2010].
+Outline approach based on EFA is more suitable for shapes lacking obvious landmarks or curves convey more important information indicating the meaningful variation in shapes, for example, ceramics.
 @Wilczek2014 evaluate the concordance between outline-based approaches, including the EFA and the Discrete Cosine Transform (DCT), and traditional typology by studying 154 complete ceramic vessels in varied forms from the Bibracte oppidum in France.
@@
 In this paper, we use EFA for pottery shape analysis to evaluate the level of standardization of pottery from an Iron Age site in northeast Taiwan in relation to the European presence in the 17th century that might hint the emergence of craft specialization for pottery production.
-Craft specialization can provide further evidence about the type of production organization which is closely related to economic or political aspects of a society [@Arnold2000; @Blackman1993; @Costin1991; @Hirshman2010; @Stark1995].
-The emergence of craft specialization can also be an indicator for social inequality that hints the possibility of the presence of production groups worked for specific individuals.
+Craft specialization can provide further evidence about the type of production organization which is closely related to economic or political aspects of society [@Arnold2000; @Blackman1993; @Costin1991; @Hirshman2010; @Stark1995].
+The emergence of craft specialization can also be an indicator of social inequality that hints the possibility of the presence of production groups worked for specific individuals.
 The presence of craft specialization is commonly studied by measuring the standardization of ceramics based on the assumption that the specialized mass production will lead to uniformity or homogeneity of the product due to increased skills, routinization, and lower diversity of producers [@Arnold2000].
-To test the prediction of a shift in the level of social inequality in this case study in northeast Taiwan influenced by the colonial European presence, we use standardization of pottery in shapes as an index for craft specialization to identify the presence of a few individuals controlling large scale pottery production [@Costin1995; @Hirshman2010] using outline-based geometric morphometrics couples with traditional metric measurements to compare and evaluate the use of outline approaches for pottery with high consistent in forms.
+To test the prediction of a shift in the level of social inequality in this case study in northeast Taiwan influenced by the colonial European presence, we use standardization of pottery in shapes as an index for craft specialization to identify the presence of a few individuals controlling large scale pottery production [@Costin1995; @Hirshman2010] using outline-based geometric morphometrics couples with traditional metric measurements to compare and evaluate the use of outline approaches for pottery with high consistency in forms.
 

### 2019-12-10 analysis/paper/paper.Rmd
changed the file name, removed "."

@@
 ---
-title: "Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeast Taiwan"
+title: "Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeastern Taiwan"
 author:
@@
 However, traditional typological and linear measurements are limited because they can be insensitive to subtle variations resulting from changes in craft specialization.
-Taking a new approach to the measurement of craft specialization, we studied shape and standardization of locally made pottery to identify changes in pottery production at Kiwulan (1400-1900 AD), a large multi-component archaeological site in northeast Taiwan.
+Taking a new approach to the measurement of craft specialization, we studied shape and standardization of locally made pottery to identify changes in pottery production at Kiwulan (1400-1900 AD), a large multi-component archaeological site in northeastern Taiwan.
 We apply geometric morphometric methods to study artifact shapes to investigate if there are any changes resulting from foreign contact of European and Chinese that might indicate social changes in the indigenous society.
@@
 Culture contact between indigenous people and colonizer with imperial power usually leads to profound changes in local indigenous societies.
-We explore the culture changes in indigenous societies in northeast Taiwan that experienced foreign contacts with the Europeans and Chinese since 17th century by examining the local pottery production.
-We observed pottery production in northeast Taiwan in the late Iron Age period and historical period (1400-1900 AD) presenting a high consistency in form and shape compared to other pottery found from other regions on this island that hints the emergence of pottery specialization.
-We asked the questions: Did foreign contact have an impact on indigenous pottery production in Northeast Taiwan that can be detected in the shape of the vessels?
+We explore the culture changes in indigenous societies in northeastern Taiwan that experienced foreign contacts with the Europeans and Chinese since 17th century by examining the local pottery production.
+We observed pottery production in northeastern Taiwan in the late Iron Age period and historical period (1400-1900 AD) presenting a high consistency in form and shape compared to other pottery found from other regions on this island that hints the emergence of pottery specialization.
+We asked the questions: Did foreign contact have an impact on indigenous pottery production in northeastern Taiwan that can be detected in the shape of the vessels?
 Did the pottery production become more standardized and homogeneous in shape after foreign contact?
@@
 
-In this paper, we use EFA for pottery shape analysis to evaluate the level of standardization of pottery from an Iron Age site in northeast Taiwan in relation to the European presence in the 17th century that might hint the emergence of craft specialization for pottery production.
+In this paper, we use EFA for pottery shape analysis to evaluate the level of standardization of pottery from an Iron Age site in northeastern Taiwan in relation to the European presence in the 17th century that might hint the emergence of craft specialization for pottery production.
 Craft specialization can provide further evidence about the type of production organization which is closely related to economic or political aspects of society [@Arnold2000; @Blackman1993; @Costin1991; @Hirshman2010; @Stark1995].
@@
 The presence of craft specialization is commonly studied by measuring the standardization of ceramics based on the assumption that the specialized mass production will lead to uniformity or homogeneity of the product due to increased skills, routinization, and lower diversity of producers [@Arnold2000].
-To test the prediction of a shift in the level of social inequality in this case study in northeast Taiwan influenced by the colonial European presence, we use standardization of pottery in shapes as an index for craft specialization to identify the presence of a few individuals controlling large scale pottery production [@Costin1995; @Hirshman2010] using outline-based geometric morphometrics couples with traditional metric measurements to compare and evaluate the use of outline approaches for pottery with high consistency in forms.
+To test the prediction of a shift in the level of social inequality in this case study in northeastern Taiwan influenced by the colonial European presence, we use standardization of pottery in shapes as an index for craft specialization to identify the presence of a few individuals controlling large scale pottery production [@Costin1995; @Hirshman2010] using outline-based geometric morphometrics couples with traditional metric measurements to compare and evaluate the use of outline approaches for pottery with high consistency in forms.
 
@@
 
-Ceramics data analysed in this paper come from the Upper Culture Layer of Kiwulan, northeast Taiwan.
+Ceramics data analysed in this paper come from the Upper Culture Layer of Kiwulan, northeastern Taiwan.
 Those ceramics dated from AD 1400 to AD 1900, which covers 500 years from the late Iron Age to the historical period  by the European presence in the 17th century.

### 2019-12-19 analysis/paper/paper.Rmd
some editing

@@
 Different from linear measurement approaches that capture shapes by measuring length, width or ratios of objects, geometric morphometrics methods use Cartesian coordinates of morphological structures to define the shapes [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007].
-Landmarks, curves or outlines of objects can be represented by coordinates with their unique location and orientation with respect to numerical values of coordinate axes.
+Landmarks, curves or outlines of objects can be represented by coordinates by their unique point location with respect to numerical values of coordinate axes.
 According to different focuses and nature of original shape data, there are two common morphometric methods used to capture and analyse the shapes, landmark approaches and outline approaches [@Adams2004].
-Landmark approaches assign a set of landmarks onto objects as reference points that can be specified on a coordinate system as x, y coordinates for two dimensions or x, y, z coordinates for three dimensions.
-Based on raw landmarks, objects are translated to a common centroid, rescaled into the same size, and rotated until the summed squared distances between corresponding landmarks are minimized using generalized Procrustes analysis (GPA) [@Bookstein1991].
-GPA superimposes sets of landmark configurations into a common coordinate system, where superimposed landmark coordinates are used as shape variables that can be further analysed by multivariate statistics [@Slice2007].
-Landmark-based morphometrics are widely used for archaeological objects with extreme morphological signature or distinctive points that provide suitable locations for landmark placement, such as projectile point tips or biologically meaningful points indicating osteological differences [@Birch2019; @Buchanan2019; @Lycett2013; @Cardillo2010; @Haruda2019; @Meloro2015].
+Landmark approaches assign a set of landmarks or semilandmarks onto objects as reference points that can be specified on a coordinate system as x, y coordinates for two dimensions, or x, y, z coordinates for three dimensions.
+Based on distances between raw landmarks, objects are translated to a common centroid, rescaled into the same size, and rotated until the summed squared distances between corresponding landmarks are minimized using generalized Procrustes analysis (GPA) [@Bookstein1991].
+GPA is a method that superimposes sets of landmark configurations into a common coordinate system, where superimposed landmark coordinates are used as shape variables for further multivariate statistics analyses[@Slice2007].
+Landmark-based morphometrics are widely used for archaeological objects where morphological signature is obvious to identity which provides ideal locations for landmark placement, such as projectile point tips or biologically meaningful points indicating osteological features [@Birch2019; @Buchanan2019; @Lycett2013; @Cardillo2010; @Haruda2019; @Meloro2015].
 
-However, one limitation of landmark-based morphometrics is that landmark data may not be able to capture the shape differences of a morphological structure of which the curving outlines between landmarks convey important information.
+However, one limitation of landmark-based morphometrics is that landmark data may not be able to capture the shape differences of a morphological structure where the curving outlines between landmarks convey important information.
 In addition, obvious and reliable landmark points may be unavailable for some complex, rounded or globular shaped structures.
-In those cases, outline approaches are more suitable to study shape by focusing more on the overall shape of a morphological structure.
+In those cases, outline approaches are more suitable to study shape by focusing on the overall shape of a morphological structure.
 One of the outline approaches is semi-landmarks method, which assigns points along an outline curve at defined intervals between two landmarks to capture the variations of the outlines [@Bookstein1997; @Lawing2010].
 Another approach, also the method used in this paper, is elliptic Fourier Analysis (EFA) commonly used for two-dimensional closed shape [@Kuhl1982].
-EFA is a powerful method to extract geometric information where outlines are considered as periodic functions and decomposed into a series of ellipses described by trigonometric functions, called harmonic coefficients [@Adam2004; @Claude2008; @Bonhomme2014].
-The number of harmonics determines the quality of the geometry that more harmonics can capture the shapes more precisely.
-The use of EFA in archaeology is mostly for bone morphology to understand the biological variation and recently more for stone artefacts to explore stone tool technology [@Fox2015; @Hoggard2019; @Iovictua2010].
-Outline approach based on EFA is more suitable for shapes lacking obvious landmarks or curves convey more important information indicating the meaningful variation in shapes, for example, ceramics.
-@Wilczek2014 evaluate the concordance between outline-based approaches, including the EFA and the Discrete Cosine Transform (DCT), and traditional typology by studying 154 complete ceramic vessels in varied forms from the Bibracte oppidum in France.
-They found that the variation in ceramic shapes analysed by EFA and DCT matches traditional ceramic typology, which supports that outline-based approaches can be efficiently used for studying variations in ceramic forms.
-EFA enables the study of the level of production standardisation to answer questions about the evolution of shape over space and time.
+EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions, called harmonic coefficients [@Adam2004; @Claude2008; @Bonhomme2014].
+The number of harmonics determines the quality and precision of the geometry of an object.
+The harmonic power expressed by the cumulated sum of squared harmonic coefficient usually provides a robust rule for choosing desired number of harmonics [@Bonhomme2014].
+The application of EFA in archaeology is mostly used for bone morphology to understand the biological variation and recently more for stone artefacts to explore stone tool technology [@Fox2015; @Hoggard2019; @Iovictua2010].
+Outline approach based on EFA is suitable for shapes lacking obvious landmarks or curves convey more important information indicating the meaningful variations, for example, ceramics.
+@Wilczek2014 evaluate the concordance between outline-based approaches, the EFA and the Discrete Cosine Transform (DCT), and traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France.
+They found that the variation in ceramic shapes analysed by EFA and DCT matches traditional ceramic typology, which supports that outline-based approaches can be efficiently used for studying variations in ceramic shapes.
+In addition, the results of shape variation by EFA helps to understand the level of production standardisation over time across that region.
 
@@
 The presence of craft specialization is commonly studied by measuring the standardization of ceramics based on the assumption that the specialized mass production will lead to uniformity or homogeneity of the product due to increased skills, routinization, and lower diversity of producers [@Arnold2000].
-To test the prediction of a shift in the level of social inequality in this case study in northeastern Taiwan influenced by the colonial European presence, we use standardization of pottery in shapes as an index for craft specialization to identify the presence of a few individuals controlling large scale pottery production [@Costin1995; @Hirshman2010] using outline-based geometric morphometrics couples with traditional metric measurements to compare and evaluate the use of outline approaches for pottery with high consistency in forms.
+To test the prediction of a shift in the level of social inequality in this case study in northeastern Taiwan influenced by the colonial European presence, we use standardization of pottery in shapes as an index for craft specialization to identify the presence of a few individuals controlling large scale pottery production [@Costin1995; @Hirshman2010] using outline-based geometric morphometrics couples with traditional metric measurements to compare and evaluate the use of outline approaches for pottery with high consistency in shapes.
 
@@
 
-Ceramics data analysed in this paper come from the Upper Culture Layer of Kiwulan, northeastern Taiwan.
-Those ceramics dated from AD 1400 to AD 1900, which covers 500 years from the late Iron Age to the historical period  by the European presence in the 17th century.
+Ceramics data analysed in this paper comes from the upper component of Kiwulan, northeastern Taiwan.
+Those ceramics dated from AD 1400 to AD 1900, which covers 500 years from the late Iron Age to the historical period started with the European presence in the 17th century.
 Kiwulan is situated on a hill near a riverside at the northern margin of Yilan, which is characterized by a triangular alluvial plain facing eastwards the Pacific and mountains on three other sides.
-The Upper Culture Layer experienced frequent foreign contacts including the European colonial presence in the 17th century and great waves of Chinese immigrants since 19th century.
+The upper component experienced frequent foreign contacts including the European colonial presence in the 17th century and great waves of Chinese immigrants since 19th century.
 The excavation revealed abundant artifacts in which potsherds are the dominate materials throughout the site.
@@
 
-The results of PCA show that the first three PCs explain 86% of shape variation.
-PC1 relates to height of vessel, PC2 relates to width of neck and mouth, and PC3 relates to width and flare of neck.
-The PCA biplot of vessel shape shows grouping by each occupation phase.
-MANOVA shows significant differences in shape between Pre-European and Post-European (p = 0.02) and Pre-European and Chinese contact shapes (p < 0.01).
+Outlines generated by EFA are examined by PCA, which shows the first three principal components (PCs) can explain 86% of shape variation.
+The first PC is related to the height of the vessels from short to high, the second PC reflects the width of their necks and mouths from narrow to wide, and the third PC relates to the degree of flare of neck from curve to straight.
+The biplot of PCs presents the grouping by each occupation phase.
+Multivariate analysis of variance (MANOVA) is used to test the difference in three shape varaibles from three phases.
+The result shows significant differences in shape between Pre-European and Post-European (p = 0.02) and Pre-European and Chinese contact shapes (p < 0.01).
 The distributions of the first three PCs for each vessel suggest variations in shape standardization by period.
-PC1 shows more variation in the pre-Euro period compared to the Chinese period.
+The first PC, the hight of the vessels, shows more variations in the pre-Euro period compared to the Chinese period.
 Standardization appears strong in the pre-Euro period but weak in the Chinese period.

### 2019-12-19 analysis/paper/paper.Rmd
add figures, update code, work on PCA

@@
 One of the outline approaches is semi-landmarks method, which assigns points along an outline curve at defined intervals between two landmarks to capture the variations of the outlines [@Bookstein1997; @Lawing2010].
-Another approach, also the method used in this paper, is elliptic Fourier Analysis (EFA) commonly used for two-dimensional closed shape [@Kuhl1982].
+Another approach is elliptic Fourier Analysis (EFA) commonly used for two-dimensional closed shape which turns coordinates into coefficients [@Kuhl1982].
 EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions, called harmonic coefficients [@Adam2004; @Claude2008; @Bonhomme2014].
 The number of harmonics determines the quality and precision of the geometry of an object.
-The harmonic power expressed by the cumulated sum of squared harmonic coefficient usually provides a robust rule for choosing desired number of harmonics [@Bonhomme2014].
+The harmonic power, a cumulated sum of squared harmonic coefficient, provides a robust rule for desired number of harmonics [@Bonhomme2014].
 The application of EFA in archaeology is mostly used for bone morphology to understand the biological variation and recently more for stone artefacts to explore stone tool technology [@Fox2015; @Hoggard2019; @Iovictua2010].
@@
     phase == 'pre-e' ~ "Pre-European",
-    phase == 'ch-con' ~ "Post-European" # we combine Chinese and post-Euro
+    phase == 'ch-con' ~ "Chinese" # we combine Chinese and post-Euro
   )))
@@
   tally(sort = T)
+```
 
+```{r jpg-coordinates, eval = FALSE}
 # start shape exploration
@@
 # mean shape, per group
-bot.ms <- mshapes(op, 2)
-bot.ms$shp$`Post-European` <- bot.ms$shp$`Post-European`  %>% coo_plot(border="blue")
-`Post-European`  <- bot.ms$shp$`Post-European`  %>% coo_draw(border="blue")
-`Pre-European` <-   bot.ms$shp$`Pre-European`  %>% coo_draw(border="red")
+pot.ms <- mshapes(op, 2)
+pot.ms$shp$`Post-European` <- pot.ms$shp$`Post-European` %>% coo_plot(border="blue")
+`Post-European` <- pot.ms$shp$`Post-European` %>% coo_draw(border="blue")
+`Pre-European` <- pot.ms$shp$`Pre-European`  %>% coo_draw(border="red")
+`Chinese` <- pot.ms$shp$`Chinese`  %>% coo_draw(border="green")
 legend("topright", 
        lwd=1,
-       col=c("blue", "red"), 
-       legend=c( "Post-European", "Pre-European"))
+       col=c("red", "blue", "green"), 
+       legend=c( "Pre-European", "Post-European", "Chinese"))
 
-# or from mshapes directly
-jpgs_imported_coo %>% efourier(6) %>% mshapes(~phase) %>% plot_mshapes()
+# from mshapes directly
+jpgs_imported_coo %>% efourier() %>% mshapes(~phase) %>% plot_mshapes()
 
@@
 png(here::here("analysis/figures/tps-pre-Euro-to-post-Euro.png"), res = 300, w = 900, h = 900)
-tps_grid(`Pre-European`, `Post-European` , shp.lwd = c(2, 2), shp.border = c("red", "blue"),
+
+tps_grid(`Pre-European`, `Post-European`, shp.lwd = c(2, 2), shp.border = c("red", "blue"),
          amp = 1, grid.size = 18, legend = T)
@@
 png(here::here("analysis/figures/tps-pre-Euro-to-Chinese.png"), res = 300, w = 900, h = 900)
-tps_grid(`Pre-European`, `Post-European`, shp.lwd = c(2, 2), shp.border = c("red", "blue"),
+tps_grid(`Pre-European`, `Chinese`, shp.lwd = c(2, 2), shp.border = c("red", "blue"),
          amp = 1, grid.size = 18, legend = T)
@@
 
+The elliptic Fourier coefficients of 73 pottery from three different phases were calculated.
+Reliable pottery outline was captured by 13 harmonics that gather 99 % of the total harmonic power.
+Thin-plate splines compares the average shapes of vessels from each period to visualize the outline deformations required to pass from an extreme of one morphospace to the other.
+
 ```{r PCA-analysis}
-# pca
 # these are fast
 op.p <- PCA(op)
+# proportion of variances explained, PC1 and PC2 explain 79% of variation
+pro_vari <-scree_plot(op.p)$data
+vari_two <- round(pro_vari$cumsum[2],4)*100
+vari_three <- round(pro_vari$cumsum[3],4)*100
 
-# We can see the pots groups by phase, and their shape distribution
-# according to the first two principle components of the EFA 
-# PC1 and PC2 explain 79% of variation
-png("index/data/pottery-shape/pca-biplot.png", res = 300, w = 1500, h = 1500)
+# the pots groups by phase, and their shape distribution according to 
+# the first two principle components of the EFA 
+png(here::here("analysis/figures/pca-biplot.png"), res = 300, w = 1500, h = 1500)
 plot(op.p, 2, 
@@
 plot(PCA(op), 2, pos.shp="range_axes")
+
 plot(PCA(op), 2, pos.shp="full_axes")
@@
 
-ggsave("index/data/pottery-shape/pca-contrib.png")
-scree_plot(op.p)
+ggsave(here::here("analysis/figures/pca-contrib.png"))
 
@@
 
-The elliptic Fourier coefficients of 73 pottery from three different phases were calculated.
-Reliable pottery outline was captured by 13 harmonics that gather 99 % of the total harmonic power.
-Thin-plate splines compares the average shapes of vessels from each period to visualize the outline deformations required to pass from an extreme of one morphospace to the other.
+Outlines generated by EFA are examined by principal components analysis to explore the shape variations across three phases.
+The first two principal components (PCs) can explain `r vari_two`% of variances, and the first three principal components (PCs) can explain `r vari_three`% of variances.
+The first PC is related to the height of the vessels from short to high, the second PC reflects the width of their necks and mouths from narrow to wide, and the third PC relates to the degree of flare of neck from curve to straight.
+The biplot of PCs presents the grouping by each occupation phase.
 
@@
 
-Outlines generated by EFA are examined by PCA, which shows the first three principal components (PCs) can explain 86% of shape variation.
-The first PC is related to the height of the vessels from short to high, the second PC reflects the width of their necks and mouths from narrow to wide, and the third PC relates to the degree of flare of neck from curve to straight.
-The biplot of PCs presents the grouping by each occupation phase.
 Multivariate analysis of variance (MANOVA) is used to test the difference in three shape varaibles from three phases.

### 2019-12-19 analysis/paper/paper.Rmd
move LDA code to script file

@@
 
-```{r LDA-analysis}
-# lda
-
-# we performed a linear discriminant analysis (LDA) based on the new shape variables (PCs), with a leave-one-out cross-validation procedure, to identify the linear combination of shape features that was able to discriminate between phases
-
-# we work on PCA scores
-bot.l <- LDA(op.p, 2)
-# print a summary, along with the leave-one-out cross-validation table. 
-bot.l 
-# plot.LDA works pretty much with the same grammar as plot.PCA
-# here we only have one LD
-plot(bot.l)
-# plot the cross-validation table
-plot_CV(bot.l)
-
-# Does not produce an efficient separation between groups and a cross-validation (leave-one-out) at phase level showing a low confidence in the reclassification.
-But best for Pre-e, indicating these have the most distinctive shapes
-
-#--------------------------------------------------
-# Unsupervised model-based clustering
-
-# what are the best grouping of the pots by shape?
-library(mclust)
-
-BIC <- mclustBIC(op.p$x[ , 1:5])
-mod1 <- Mclust(op.p$x[ , 1:5], x = BIC)
-summary(mod1, parameters = TRUE)
-plot(mod1, what = "classification")
-
-# no clusters apparent 
-```
-
 ```{r MANOVA}

### 2019-12-19 analysis/paper/LDA_analysis.R

@@ -1,0 +1,27 @@
+# LDA-analysis
+# we performed a linear discriminant analysis (LDA) based on the new shape variables (PCs), with a leave-one-out cross-validation procedure, to identify the linear combination of shape features that was able to discriminate between phases
+
+# we work on PCA scores
+bot.l <- LDA(op.p, 2)
+# print a summary, along with the leave-one-out cross-validation table.
+bot.l
+# plot.LDA works pretty much with the same grammar as plot.PCA
+# here we only have one LD
+plot(bot.l)
+# plot the cross-validation table
+plot_CV(bot.l)
+
+# Does not produce an efficient separation between groups and a cross-validation (leave-one-out) at phase level showing a low confidence in the reclassification. But best for Pre-e, indicating these have the most distinctive shapes
+
+#--------------------------------------------------
+# Unsupervised model-based clustering
+
+# what are the best grouping of the pots by shape?
+library(mclust)
+BIC <- mclustBIC(op.p$x[ , 1:5])
+mod1 <- Mclust(op.p$x[ , 1:5], x = BIC)
+summary(mod1, parameters = TRUE)
+plot(mod1, what = "classification")
+
+# no clusters apparent
+

### 2019-12-19 analysis/paper/shapes.R
move some code for shape to script file

@@ -1,0 +1,13 @@
+# some plots generated from momocs
+jpgs_imported_coo <- read_rds(here::here("analysis/data/derived_data/jpgs_imported_coo.rds"))
+
+# take a look when they stack together
+stack(jpgs_imported_coo)
+
+# fit the x and y coordinates separately
+coo_oscillo(jpgs_imported_coo[1], "efourier")
+
+# The progressive capture of shape geometry along the number of harmonics can be visualized with:
+calibrate_reconstructions_efourier(jpgs_imported_coo,
+                                   id = 1,
+                                   range = c(1, 2, 4, 8, 14, 20))

### 2019-12-19 analysis/paper/paper.Rmd
move some code for shape to script file
all code runs

@@
 
-```{r tidy-data, eval=FALSE}
+```{r tidy-data}
 # Once we have the jpgs, start from here
 library(tidyverse)
+library(here)
 
 # get list of jpgs only
-jpgs <- list.files("analysis/data/raw_data/outline_jpg", pattern = ".jpg", full.names = TRUE)
+jpgs <- list.files(here("analysis/data/raw_data/outline_jpg"), pattern = ".jpg", full.names = TRUE)
 
 # join file names to phase
-pottery_data <- readr::read_csv("analysis/data/raw_data/KWL_Ceramic_final2018_with_phases.csv")
+pottery_data <- readr::read_csv(here("analysis/data/raw_data/KWL_Ceramic_final2018_with_phases.csv"))
 pottery_data <- pottery_data %>% 
@@
 
-```{r jpg-coordinates, eval = FALSE}
+```{r jpg-coordinates, eval=FALSE}
 # start shape exploration
@@
 ```{r EFA-analysis}
-jpgs_imported_coo <- read_rds(here::here("analysis/data/derived_data/jpgs_imported_coo.rds"))
-
-# take a look when they stack together
-stack(jpgs_imported_coo)
-
-# fit the x and y coordinates separately
-coo_oscillo(jpgs_imported_coo[1], "efourier")
+library(Momocs)
+jpgs_imported_coo <- readr::read_rds(here::here("analysis/data/derived_data/jpgs_imported_coo.rds"))
 
 # default is 13 harmanics that gather 99% of the harmonic power 
-op <- efourier(jpgs_imported_coo, norm = FALSE)
+op <- efourier(jpgs_imported_coo, norm = FALSE, 13)
 
@@
 # mean shape, per group
-pot.ms <- mshapes(op, 2)
-pot.ms$shp$`Post-European` <- pot.ms$shp$`Post-European` %>% coo_plot(border="blue")
-`Post-European` <- pot.ms$shp$`Post-European` %>% coo_draw(border="blue")
-`Pre-European` <- pot.ms$shp$`Pre-European`  %>% coo_draw(border="red")
-`Chinese` <- pot.ms$shp$`Chinese`  %>% coo_draw(border="green")
+pot.ms <- MSHAPES(op, 'phase')
+mshp <- pot.ms$shp
+names(mshp) <- c("chi", "poe", "pre")
+
+`Post-European` <- mshp$poe %T>%  coo_plot( border="blue")
+`Pre-European` <-  mshp$pre  %T>% coo_plot(border="red", plot.new=FALSE)
+`Chinese` <-       mshp$chi  %T>% coo_plot(border="green", plot.new=FALSE)
 legend("topright", 
@@
 
-# from mshapes directly
-jpgs_imported_coo %>% efourier() %>% mshapes(~phase) %>% plot_mshapes()
+# from mshapes directly, nice but the labels spill outside of the plot area, not clear how to fix
+op %>% MSHAPES(~phase) %>% plot_MSHAPES()
 
@@
 # Isodeformation lines
+png(here::here("analysis/figures/tps-iso.png"), res = 600, w = 1200, h = 1800)
 tps_iso(`Pre-European`, `Post-European`,  shp.lwd = c(2, 2), shp.border = c("red", "blue"),
@@
         palette = col_summer, grid = F)
+dev.off()
 ```
@@
 # plot PC1-3 distribution by phase
-pc_plot <- data_frame(pc = c(op.p$x[,1],
+pc_plot <- tibble(pc = c(op.p$x[,1],
                              op.p$x[,2],
@@
   theme_minimal()
-ggsave("index/data/pottery-shape/pcs-boxplot-for-shape-variation.png",
+ggsave(here("analysis/figures/pcs-boxplot-for-shape-variation.png"),
        h = 10, w = 5)
@@
 # statistical test
-pc_plot_tests <- 
-pc_plot %>% 
-  nest_legacy(pc, phase) %>% 
-  mutate(var_test = purrr::map(data, ~broom::tidy(var.test(.x$pc, .x$phase)))) %>% 
-  unnest_legacy(var_test)
 
@@
 
-ggsave("index/data/pottery-shape/cv-test-on-pcs.png", h = 3, w = 12)
+ggsave(here("analysis/figures/cv-test-on-pcs.png"), h = 3, w = 12)
 

### 2019-12-19 README.Rmd
update readme

@@
 
-[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh///master?urlpath=rstudio)
+[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/LiYingWang/kwl.pottery/master?urlpath=rstudio)
 
@@
 # install.packages("devtools")
-remotes::install_github("/")
+remotes::install_github("LiYingWang/kwl.pottery")
 ```

### 2019-12-20 analysis/paper/paper.Rmd
fix knitting problem, add paragraph of reproducibility

@@
 We also include the Krishnamoorthy and Lee (2014) test as a more recent development with lower rates of type I error, better performance with uneven sample numbers, and more power across a range of conditions.
+
+# Reproducibility and open source materials
+
+To enable re-use of materials and improve reproducibility and transparency [@Marwick2017], the entire R code [@Rlanguage2019] used for all the analysis and visualizations contained in this paper is included in http:XXX.
+Also in this version-controlled compendium [@Marwick2018] are the raw data for all the visualizations and tests reported here.
+All of the figures, tables, and statistical test results presented here can be independently reproduced with the code and data in this repository.
+The code is released under the MIT license, the data as CC-0, and figures as CC-BY, to enable maximum re-use.
 

### 2019-12-20 analysis/paper/paper.Rmd
remove some paragraphs to script file

@@
 - CV is hard to test, we apply some statistics testing for it
-
-# Hypotheses and expectations
-
-My hypothesis is that if foreign contacts including European and Chinese had impact on the emergence of social inequality in local indigenous society due to the monopolies for trade between a small number of indigenous people, then I expect to see the changes in social organization from a more corporate to a more network organization that can be reflected by pottery production at Kiwulan.
-The emergence of craft specialization, here pottery specialization, are usually related to the presence of elite control.
-In this case, if the competitions for foreign resources and being trade partners of European or Chinese colonizer among individuals gradually lead to the emergence of social inequality, then I expect the local pottery will show more homogeneous features after contact due to the pottery specialization caused by control of small group of individuals.
 
@@
        legend=c( "Pre-European", "Post-European", "Chinese"))
-
-# from mshapes directly, nice but the labels spill outside of the plot area, not clear how to fix
-op %>% MSHAPES(~phase) %>% plot_MSHAPES()
 

### 2019-12-20 analysis/paper/scrap.R

@@ -1,0 +1,3 @@
+# Hypotheses and expectations
+
+My hypothesis is that if foreign contacts including European and Chinese had impact on the emergence of social inequality in local indigenous society due to the monopolies for trade between a small number of indigenous people, then I expect to see the changes in social organization from a more corporate to a more network organization that can be reflected by pottery production at Kiwulan. The emergence of craft specialization, here pottery specialization, are usually related to the presence of elite control. In this case, if the competitions for foreign resources and being trade partners of European or Chinese colonizer among individuals gradually lead to the emergence of social inequality, then I expect the local pottery will show more homogeneous features after contact due to the pottery specialization caused by control of small group of individuals.

### 2019-12-20 analysis/paper/shapes.R

@@ -11,3 +11,7 @@ coo_oscillo(jpgs_imported_coo[1], "efourier")
 calibrate_reconstructions_efourier(jpgs_imported_coo,
                                    id = 1,
                                    range = c(1, 2, 4, 8, 14, 20))
+
+
+# from mshapes directly, nice but the labels spill outside of the plot area, not clear how to fix
+op %>% MSHAPES(~phase) %>% plot_MSHAPES()

### 2019-12-21 analysis/paper/paper.Rmd
add in-text code

@@
 GPA is a method that superimposes sets of landmark configurations into a common coordinate system, where superimposed landmark coordinates are used as shape variables for further multivariate statistics analyses[@Slice2007].
-Landmark-based morphometrics are widely used for archaeological objects where morphological signature is obvious to identity which provides ideal locations for landmark placement, such as projectile point tips or biologically meaningful points indicating osteological features [@Birch2019; @Buchanan2019; @Lycett2013; @Cardillo2010; @Haruda2019; @Meloro2015].
+Landmark-based morphometrics are widely used for archaeological objects where morphological signature is obvious to identity which provides ideal reference points for landmark placement, such as projectile point tips or biologically measurements points indicating osteological features [@Birch2019; @Buchanan2019; @Lycett2013; @Cardillo2010; @Haruda2019; @Meloro2015].
 
 However, one limitation of landmark-based morphometrics is that landmark data may not be able to capture the shape differences of a morphological structure where the curving outlines between landmarks convey important information.
-In addition, obvious and reliable landmark points may be unavailable for some complex, rounded or globular shaped structures.
-In those cases, outline approaches are more suitable to study shape by focusing on the overall shape of a morphological structure.
-One of the outline approaches is semi-landmarks method, which assigns points along an outline curve at defined intervals between two landmarks to capture the variations of the outlines [@Bookstein1997; @Lawing2010].
+In addition, obvious and reliable landmark points may be unavailable for complex or rounded shaped structures.
+In those cases, outline approaches serve as a solution by analyzing the overall shape of a morphological structure.
+One of the outline approaches is semi-landmarks method, also called sliding landmarks, which assigns points along the curve between two landmarks at defined intervals [@Bookstein1997; @Lawing2010].
+Those semi-landmarks are allowed to slide along the curve to remove the effect of the arbitrary landmark spacing by minimizing either Procrustes distance or bending energy [@Bookstein1997; @Slice2007; @Gunz2013].
 Another approach is elliptic Fourier Analysis (EFA) commonly used for two-dimensional closed shape which turns coordinates into coefficients [@Kuhl1982].
-EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions, called harmonic coefficients [@Adam2004; @Claude2008; @Bonhomme2014].
+EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions, called harmonic coefficients [@Adams2004; @Claude2008; @Bonhomme2014].
 The number of harmonics determines the quality and precision of the geometry of an object.
@@
 
-A total of 73 vessels from pre-contact contexts (n = 32), post-European contexts (n = 27), and Chinese contact contexts (n = 14) were selected for this study because of their completeness mostly covering parts from rim to bottom.
+```{r tidy-data}
+# Once we have the jpgs, start from here
+library(tidyverse)
+library(here)
+
+# get list of jpgs only
+jpgs <- list.files(here("analysis/data/raw_data/outline_jpg"), pattern = ".jpg", full.names = TRUE)
+
+# join file names to phase
+pottery_data <- readr::read_csv(here("analysis/data/raw_data/KWL_Ceramic_final2018_with_phases.csv"))
+pottery_data <- pottery_data %>% 
+  mutate(filename = str_glue('fill_{filename}.svg_output.jpg'))
+
+# some outlines that we have no phases for... 
+sdiff <- setdiff(pottery_data$filename, basename(jpgs))
+
+# the ones we do have both outlines and phases
+isect <- intersect(pottery_data$filename, basename(jpgs))
+
+jpgs <- jpgs[basename(jpgs) %in% isect]
+pottery_data <- 
+  pottery_data %>% 
+  filter(filename %in% isect) %>% 
+  distinct(filename, phase, .keep_all = TRUE)
+
+phases_isect <- 
+  pottery_data %>%  
+  select(filename, phase) %>% 
+  filter(filename %in% isect) %>% 
+  mutate_all(as.factor) %>% 
+  distinct(filename, phase) %>% 
+  mutate(phase = as.factor(case_when(
+    phase == 'post-e' ~ "Post-European", # include European period and after it before Chinese contact
+    phase == 'pre-e' ~ "Pre-European",
+    phase == 'ch-con' ~ "Chinese" # we can combine Chinese and post-Euro
+  )))
+
+# how many pots in each phase?
+phase_fre <- phases_isect %>% 
+  group_by(phase) %>% 
+  tally(sort = T)
+
+total_no <- sum(phase_fre$n)
+```
+
+A total of `r total_no` vessels from pre-contact contexts (n = 32), post-European contexts (n = 27), and Chinese contact contexts (n = 14) were selected for this study because of their completeness mostly covering parts from rim to bottom.
 The scanned pottery drawings of those pots were acquired from Bureau of Cultural Affairs in Yilan.
@@
 
-## metric measurements
+## Metric measurements
 
@@
 # Results
-
-```{r tidy-data}
-# Once we have the jpgs, start from here
-library(tidyverse)
-library(here)
-
-# get list of jpgs only
-jpgs <- list.files(here("analysis/data/raw_data/outline_jpg"), pattern = ".jpg", full.names = TRUE)
-
-# join file names to phase
-pottery_data <- readr::read_csv(here("analysis/data/raw_data/KWL_Ceramic_final2018_with_phases.csv"))
-pottery_data <- pottery_data %>% 
-  mutate(filename = str_glue('fill_{filename}.svg_output.jpg'))
-
-# some outlines that we have no phases for... 
-sdiff <- setdiff(pottery_data$filename, basename(jpgs))
-
-# the ones we do have both outlines and phases
-isect <- intersect(pottery_data$filename, basename(jpgs))
-
-jpgs <- jpgs[basename(jpgs) %in% isect]
-pottery_data <- 
-  pottery_data %>% 
-  filter(filename %in% isect) %>% 
-  distinct(filename, phase, .keep_all = TRUE)
-
-phases_isect <- 
-  pottery_data %>%  
-  select(filename, phase) %>% 
-  filter(filename %in% isect) %>% 
-  mutate_all(as.factor) %>% 
-  distinct(filename, phase) %>% 
-  mutate(phase = as.factor(case_when(
-    phase == 'post-e' ~ "Post-European",
-    phase == 'pre-e' ~ "Pre-European",
-    phase == 'ch-con' ~ "Chinese" # we combine Chinese and post-Euro
-  )))
-
-# how many pots in each phase?
-phases_isect %>% 
-  group_by(phase) %>% 
-  tally(sort = T)
-```
 
@@
 
-The elliptic Fourier coefficients of 73 pottery from three different phases were calculated.
+The elliptic Fourier coefficients of `r total_no` pottery from three different phases were calculated.
 Reliable pottery outline was captured by 13 harmonics that gather 99 % of the total harmonic power.
@@
 Outlines generated by EFA are examined by principal components analysis to explore the shape variations across three phases.
-The first two principal components (PCs) can explain `r vari_two`% of variances, and the first three principal components (PCs) can explain `r vari_three`% of variances.
-The first PC is related to the height of the vessels from short to high, the second PC reflects the width of their necks and mouths from narrow to wide, and the third PC relates to the degree of flare of neck from curve to straight.
+The first two principal components (PCs) explain `r vari_two`% of variances, and the first three principal components (PCs) explain `r vari_three`% of variances.
+The PC1 relate to the height of the vessels from short to high, the PC2 reflects the width of their necks and mouths from narrow to wide, and the PC3 relates to the degree of flare of neck from curve to straight.
 The biplot of PCs presents the grouping by each occupation phase.

### 2019-12-22 analysis/paper/paper.Rmd
minor updates

@@
 
-# The progressive capture of shape geometry along the number of harmonics can be visualized with:
-calibrate_reconstructions_efourier(jpgs_imported_coo,  
-                          id = 1, 
-                          range = c(1, 2, 4, 8, 14, 20))
-
 # mean shape, per group
@@
 Thin-plate splines compares the average shapes of vessels from each period to visualize the outline deformations required to pass from an extreme of one morphospace to the other.
+The
 
@@
 The first two principal components (PCs) explain `r vari_two`% of variances, and the first three principal components (PCs) explain `r vari_three`% of variances.
-The PC1 relate to the height of the vessels from short to high, the PC2 reflects the width of their necks and mouths from narrow to wide, and the PC3 relates to the degree of flare of neck from curve to straight.
+The PC1 relate to the height of the vessels from short to high, the PC2 reflects the neck and mouth constriction from wide to narrow, and the PC3 relates to the degree of flare of neck from curved to straight shape.
 The biplot of PCs presents the grouping by each occupation phase.

### 2019-12-22 analysis/paper/shapes.R

@@ -1,10 +1,10 @@
 # some plots generated from momocs
 jpgs_imported_coo <- read_rds(here::here("analysis/data/derived_data/jpgs_imported_coo.rds"))
 
-# take a look when they stack together
+# take a look when all the images stack together
 stack(jpgs_imported_coo)
 
-# fit the x and y coordinates separately
+# fit the x and y coordinates separately, that is the blue and red curves in the right hand side plot
 coo_oscillo(jpgs_imported_coo[1], "efourier")
 
 # The progressive capture of shape geometry along the number of harmonics can be visualized with:
@@ -12,6 +12,6 @@ calibrate_reconstructions_efourier(jpgs_imported_coo,
                                    id = 1,
                                    range = c(1, 2, 4, 8, 14, 20))
 
-
 # from mshapes directly, nice but the labels spill outside of the plot area, not clear how to fix
+# paired comparison for every two periods shown in one plot with 3 * 3 variables
 op %>% MSHAPES(~phase) %>% plot_MSHAPES()

### 2019-12-22 analysis/paper/paper.Rmd
add sentence to pottery background

@@
 We observed pottery production in northeastern Taiwan in the late Iron Age period and historical period (1400-1900 AD) presenting a high consistency in form and shape compared to other pottery found from other regions on this island that hints the emergence of pottery specialization.
-We asked the questions: Did foreign contact have an impact on indigenous pottery production in northeastern Taiwan that can be detected in the shape of the vessels?
-Did the pottery production become more standardized and homogeneous in shape after foreign contact?
-Did the contact with European cause more homogeneous shape than Chinese contact period because of large scale exchange network?
+Our aim is to explore the impact of foreign contacts on local indigenous societies in northeastern Taiwan by examining pottery production as an indicator for social changes.
+Using pottery shape as a proxy to study craft specialization, we answer these questions: Did foreign contacts have impacts on indigenous pottery production that can be detected in the shape of the vessels?
+How does pottery change in shape throughout 600 years at this site?
+Did pottery become more homogeneous and standardized in shape after foreign contacts with the European colonizer or the Chinese immigrant?
 
-We find differences in shape and shape standardization of pottery that indicate changes in pottery production resulting from foreign contact, suggesting increasing craft specialization and changes in local social organization at Kiwulan.
-These results are important to understand the influence of culture contact on local indigenous societies and answer the anthropological question that concerns the the mechanisms for social changes.
-In addition, our case study, which includes an openly available research compendium of R code suitable for use with any other assemblage, will help to expand the use of shape-based quantitative methods to questions about craft specialization and standardization in prehistoric technologies.
+Our hypothesis is that if foreign contacts including European and Chinese had impact on the emergence of social inequality in local indigenous society due to the monopolies for trade between a small number of indigenous people, then we expect to see the changes in social organization from a more corporate to a more network organization that would be reflected by pottery production at Kiwulan.
+The emergence of craft specialization, here pottery specialization, are usually related to the presence of elite control.
+In this case, if the competitions for foreign resources and being trade partners of European or Chinese colonizer among individuals gradually lead to the emergence of social inequality, then I expect the local pottery will show more homogeneous features after contact due to the craft specialization caused by control of small group of individuals.
+The first step is to detecting pottery standardization when discussing the emergence of craft specialization and the possible changes in social organization.
+This paper uses Elliptic Fourier Analysis, one outline approach in Geometric morphometric methods, to test whether there was increasing pottery standardization in relation to foreign contacts.
 
@@
 
-Ceramics data analysed in this paper comes from the upper component of Kiwulan, northeastern Taiwan.
-Those ceramics dated from AD 1400 to AD 1900, which covers 500 years from the late Iron Age to the historical period started with the European presence in the 17th century.
+Pottery analysed in this paper comes from Kiwulan, northeastern Taiwan.
 Kiwulan is situated on a hill near a riverside at the northern margin of Yilan, which is characterized by a triangular alluvial plain facing eastwards the Pacific and mountains on three other sides.
+Kiwulan dates to AD 1350 to AD 1950, which covers around 600 years from the late Iron Age to the historical period started with the European presence in Taiwan in the early 17th century.
+The Dutch first came and occupied southern Taiwan in 1624 and then the Spanish occupied northern Taiwan in 1626.
+In 1642, the Spanish was expelled by the Dutch, who took over their forts in northern Taiwan.
+Since then, western Taiwan was mostly under the colonial rule of the Dutch until 1662 when the Kingdom of Tungning was founded by Koxinga, a loyalist of Ming dynasty of China.
+
 The upper component experienced frequent foreign contacts including the European colonial presence in the 17th century and great waves of Chinese immigrants since 19th century.
 The excavation revealed abundant artifacts in which potsherds are the dominate materials throughout the site.
-Imported ceramics from mainland China, stonewares, and ornament elements such as beads were usually found in the Upper Culture Layer that indicates the prosperous international trading activities in the 17th century during the European presence in Taiwan.
-In addition to artifacts, features were also found, such as burials, middens, and postholes with in-situ posts that explains Kiwulan was a large settlement site.
+Imported ceramics from mainland China, stonewares, and ornament elements such as beads were also commonly found in the upper component that indicates the frequent long-distance trade activities in the 17th century during the European presence.
+In addition to artifacts, features were also found, such as burials, middens, and post-holes with in-situ posts that explains Kiwulan was a large settlement site.
+
+Pottery found at Kiwulan are all locally made. 
+Most sherds are fired to an orange to brownish color with a fully oxidized core or a reduced core with oxidized fringes.
+It seems that there is no significant relationship between color and phases.
+Petrographic analysis for thin sections was conducted at the Department of Geosciences, National Taiwan University.
+The results shows high percentage of inclusions (15-50%), including argillite, quartz, feldspar, metasandstone, sandstone fragment, and slate.
+Those minerals are believed to be local materials of the Yilan Plain.
+The most common type of sherds, also the dominant one, has a high percentage of inclusion (25-40%), which is composed mainly by argillite (15-40%), followed by metasandstone (1-10%), sandstone (1-6%), and quartz (1-5%).
+The size of the particles ranges from 500 micron to 1300 micron (=1 mm).
+Overall, this type presents a mixture of fine, rounded argillite with a small part of rounded metasandstone and rounded to sub-angular monocrystalline quartz.
 
@@
 
+The elliptic Fourier coefficients of `r total_no` pottery from three different phases were calculated.
+Reliable pottery outline was captured by 13 harmonics that gather 99 % of the total harmonic power.
+Thin-plate splines compares the average shapes of vessels from each period to visualize the outline deformations required to pass from an extreme of one morphospace to the other.
+
+The mean petal shape of each variety was then drawn using the mean values of the standardized Fourier coefficients within each variety.
+Figure 1 shows the wide variation of petal shape that was observed among the varieties examined in this study.
+
 ```{r jpg-coordinates, eval=FALSE}
@@
 ```
-
-The elliptic Fourier coefficients of `r total_no` pottery from three different phases were calculated.
-Reliable pottery outline was captured by 13 harmonics that gather 99 % of the total harmonic power.
-Thin-plate splines compares the average shapes of vessels from each period to visualize the outline deformations required to pass from an extreme of one morphospace to the other.
-The
 
@@
 
+We find differences in shape and shape standardization of pottery that indicate changes in pottery production resulting from foreign contact, suggesting increasing craft specialization and changes in local social organization at Kiwulan.
+These results are important to understand the influence of culture contact on local indigenous societies and answer the anthropological question that concerns the the mechanisms for social changes.
+In addition, our case study, which includes an openly available research compendium of R code suitable for use with any other assemblage, will help to expand the use of shape-based quantitative methods to questions about craft specialization and standardization in prehistoric technologies.
+
 # Acknowledgements

### 2019-12-22 analysis/paper/paper.Rmd
add references

@@
   comment = "#>",
+  dpi = 300,
   fig.path = "../figures/"
@@
 
-The elliptic Fourier coefficients of `r total_no` pottery from three different phases were calculated.
-Reliable pottery outline was captured by 13 harmonics that gather 99 % of the total harmonic power.
-Thin-plate splines compares the average shapes of vessels from each period to visualize the outline deformations required to pass from an extreme of one morphospace to the other.
-
-The mean petal shape of each variety was then drawn using the mean values of the standardized Fourier coefficients within each variety.
-Figure 1 shows the wide variation of petal shape that was observed among the varieties examined in this study.
-
 ```{r jpg-coordinates, eval=FALSE}
@@
 
+The elliptic Fourier coefficients of `r total_no` pottery from three phases were calculated.
+Reliable pottery outline was captured by 13 harmonics that gather 99 % of the total harmonic power.
+The mean pottery shape of each phase was visualized by calculating the mean of the standardized Fourier coefficients within each group.
+Thin-plate splines compares the average shapes of vessels from each period to visualize the outline deformations required to pass from an extreme of one morphospace to the other.
+
 ```{r EFA-analysis}
 library(Momocs)
+library(cowplot)
 jpgs_imported_coo <- readr::read_rds(here::here("analysis/data/derived_data/jpgs_imported_coo.rds"))
@@
 # Deformation grids
-png(here::here("analysis/figures/tps-pre-Euro-to-post-Euro.png"), res = 300, w = 900, h = 900)
 
@@
 
-png(here::here("analysis/figures/tps-pre-Euro-to-Chinese.png"), res = 300, w = 900, h = 900)
 tps_grid(`Pre-European`, `Chinese`, shp.lwd = c(2, 2), shp.border = c("red", "blue"),
@@
 png(here::here("analysis/figures/tps-iso.png"), res = 600, w = 1200, h = 1800)
+de_pre_post_plot <- 
   tps_iso(`Pre-European`, `Post-European`,  shp.lwd = c(2, 2), shp.border = c("red", "blue"),
@@
 dev.off()
+
+de_pre_chi_plot <- 
+  tps_iso(`Pre-European`, `Chinese`,  shp.lwd = c(2, 2), shp.border = c("red", "blue"),
+          amp = 0.5, iso.nb = 10000, iso.levels = 12, legend.text = c("Pre-contact", "Chinese"),
+          palette = col_summer, grid = F)
+dev.off()
+
+# Doesn't work if the plot is not produced by ggplot?
+# plot_two_de_by_period <- 
+  #plot_grid(de_pre_post_plot, de_pre_chi_plot, labels = c('A', 'B'))
+
+#knitr::include_graphics(here::here("analysis", "figures", "tps-pre-Euro-to-post-Euro.png"))
+#knitr::include_graphics(here::here("analysis", "figures", "tps-pre-Euro-to-Chinese.png"))
+
 ```

### 2019-12-22 analysis/paper/paper.Rmd
seperate code blocks and working on PCA result

@@
 )
+
+image_width_1_col_inch = 90/25.4 # image Single column	w = 90 mm 
+image_width_2_col_inch = 190/25.4 # image Double column w = 190 mm 
 ```
@@
 Reliable pottery outline was captured by 13 harmonics that gather 99 % of the total harmonic power.
-The mean pottery shape of each phase was visualized by calculating the mean of the standardized Fourier coefficients within each group.
-Thin-plate splines compares the average shapes of vessels from each period to visualize the outline deformations required to pass from an extreme of one morphospace to the other.
+The average shape of vessels from each phase was visualized with the mean of the standardized Fourier coefficients within each phase group (Figure \@ref(fig:EFA-analysis)).
 
-```{r EFA-analysis}
+```{r EFA-analysis, fig.cap= "Mean pottery shapes of three phases"}
 library(Momocs)
@@
        legend=c( "Pre-European", "Post-European", "Chinese"))
+```
 
+```{r thin-plate-splines}
 # Thin plate spline analysis (TPS)
@@
            amp = 1, grid.size = 18, legend = T)
-dev.off()
 
@@
 png(here::here("analysis/figures/tps-iso.png"), res = 600, w = 1200, h = 1800)
-de_pre_post_plot <- 
-  tps_iso(`Pre-European`, `Post-European`,  shp.lwd = c(2, 2), shp.border = c("red", "blue"),
-          amp = 0.5, iso.nb = 10000, iso.levels = 12, legend.text = c("Pre-contact", "Post-contact"),
-          palette = col_summer, grid = F)
-dev.off()
-
-de_pre_chi_plot <- 
-  tps_iso(`Pre-European`, `Chinese`,  shp.lwd = c(2, 2), shp.border = c("red", "blue"),
-          amp = 0.5, iso.nb = 10000, iso.levels = 12, legend.text = c("Pre-contact", "Chinese"),
-          palette = col_summer, grid = F)
-dev.off()
 
@@
 # plot_two_de_by_period <- 
-  #plot_grid(de_pre_post_plot, de_pre_chi_plot, labels = c('A', 'B'))
+  #plot_grid(, labels = c('A', 'B'))
 
 #knitr::include_graphics(here::here("analysis", "figures", "tps-pre-Euro-to-post-Euro.png"))
-#knitr::include_graphics(here::here("analysis", "figures", "tps-pre-Euro-to-Chinese.png"))
-
 ```
 
-```{r PCA-analysis}
+```{r PCA-table}
 # these are fast
@@
 vari_three <- round(pro_vari$cumsum[3],4)*100
+```
 
-# the pots groups by phase, and their shape distribution according to 
-# the first two principle components of the EFA 
-png(here::here("analysis/figures/pca-biplot.png"), res = 300, w = 1500, h = 1500)
-plot(op.p, 2, 
-     morpho = TRUE, 
-     ellipses = TRUE,
-     cex = 1, 
-     cex.labelsgroups = 1.5)
+Outlines generated by EFA are examined by principal components analysis to explore the shape variations across three phases.
+The first two principal components (PCs) explain `r vari_two`% of variances, and the first three principal components (PCs) explain `r vari_three`% of variances.
+The PC1 relate to the height of the vessels from short to high, the PC2 reflects the neck and mouth constriction from wide to narrow, and the PC3 relates to the degree of flare of neck from curved to straight shape.
+The contribution of each principal component on the shape of vessels was listed in table ().
+The PC1 relative to the hight of the vessels accounts for nearly half of variances.
+Figure\@ref(fig:first-two-PCs-byplot) shows that a wider variation in pottery shape from the pre-European and post-European period compared to those from Chinese period.
 
-dev.off()
-
-# different visualization 
-plot(PCA(op), 2, pos.shp="full")
-plot(PCA(op), 2, pos.shp="range")
-plot(PCA(op), 2, pos.shp="xy")
-plot(PCA(op), 2, pos.shp="range_axes")
-
-plot(PCA(op), 2, pos.shp="full_axes")
-
+```{r contribution-PCs}
 # the contributions of each PC to the shapes
@@
 
-Outlines generated by EFA are examined by principal components analysis to explore the shape variations across three phases.
-The first two principal components (PCs) explain `r vari_two`% of variances, and the first three principal components (PCs) explain `r vari_three`% of variances.
-The PC1 relate to the height of the vessels from short to high, the PC2 reflects the neck and mouth constriction from wide to narrow, and the PC3 relates to the degree of flare of neck from curved to straight shape.
-The biplot of PCs presents the grouping by each occupation phase.
+```{r first-two-PCs-byplot, fig.cap= "Pottery shape distribution by each occupation phase according to the first two PCs"}
+# the pots groups by phase, and their shape distribution according to 
+# the first two principle components of the EFA 
+png(here::here("analysis/figures/pca-biplot.png"), res = 300, w = 1500, h = 1500)
+plot(op.p, 2, 
+     morpho = TRUE, 
+     ellipses = TRUE,
+     cex = 1, 
+     cex.labelsgroups = 1.5)
+dev.off()
+
+# PCA byplot
+plot(PCA(op), 2, pos.shp="range")
+```
 
@@
 manv <-   MANOVA_PW(op.p, "phase")
-manv$summary
+# manv$summary
 

### 2019-12-22 analysis/paper/shapes.R

@@ -15,3 +15,21 @@ calibrate_reconstructions_efourier(jpgs_imported_coo,
 # from mshapes directly, nice but the labels spill outside of the plot area, not clear how to fix
 # paired comparison for every two periods shown in one plot with 3 * 3 variables
 op %>% MSHAPES(~phase) %>% plot_MSHAPES()
+
+# calculates deformations between two configurations and map them with or without isolines.
+tps_iso(`Pre-European`, `Post-European`,  shp.lwd = c(2, 2), shp.border = c("red", "blue"),
+        amp = 0.5, iso.nb = 10000, iso.levels = 12, legend.text = c("Pre-contact", "Post-contact"),
+        palette = col_summer, grid = F)
+dev.off()
+
+tps_iso(`Pre-European`, `Chinese`,  shp.lwd = c(2, 2), shp.border = c("red", "blue"),
+        amp = 0.5, iso.nb = 10000, iso.levels = 12, legend.text = c("Pre-contact", "Chinese"),
+        palette = col_summer, grid = F)
+dev.off()
+
+# PCA byplot, different visualization
+plot(PCA(op), 2, pos.shp="full")
+plot(PCA(op), 2, pos.shp="range")
+plot(PCA(op), 2, pos.shp="xy")
+plot(PCA(op), 2, pos.shp="range_axes")
+plot(PCA(op), 2, pos.shp="full_axes")

### 2019-12-23 analysis/paper/paper.Rmd
work on PCA results

@@
 
-`Post-European` <- mshp$poe %T>% coo_plot( border="blue")
+`Post-European` <- mshp$poe %T>% coo_plot(border="blue")
 `Pre-European` <-  mshp$pre %T>% coo_plot(border="red", plot.new=FALSE)
@@
 # Doesn't work if the plot is not produced by ggplot?
-# plot_two_de_by_period <- 
-  #plot_grid(, labels = c('A', 'B'))
+# plot_grid(a, b, labels = c('A', 'B'))
 
@@
 pro_vari <-scree_plot(op.p)$data
+pc1 <- round(pro_vari$proportion[1],4)*100
+pc3 <- round(pro_vari$proportion[3],4)*100
 vari_two <- round(pro_vari$cumsum[2],4)*100
@@
 
-Outlines generated by EFA are examined by principal components analysis to explore the shape variations across three phases.
-The first two principal components (PCs) explain `r vari_two`% of variances, and the first three principal components (PCs) explain `r vari_three`% of variances.
-The PC1 relate to the height of the vessels from short to high, the PC2 reflects the neck and mouth constriction from wide to narrow, and the PC3 relates to the degree of flare of neck from curved to straight shape.
-The contribution of each principal component on the shape of vessels was listed in table ().
-The PC1 relative to the hight of the vessels accounts for nearly half of variances.
-Figure\@ref(fig:first-two-PCs-byplot) shows that a wider variation in pottery shape from the pre-European and post-European period compared to those from Chinese period.
+The elliptic Fourier coefficients are examined by principal components analysis to explore the shape variations across three phases.
+The first two principal components (PCs) explain `r vari_two`% of the total variances, of which `r pc1`% is explained by the first principle component.
+With the third component, the first three principal components explain `r vari_three`% of the total variances.
+Figure \@ref(fig:contribution-PCs) shows the shape variation associated to the first three principal components.
+From the negative to the positive scores, the PC1 represents the height of the vessels from tall to short and the roundness of the body from round to oval shaped.
+The PC2 relates to the neck and mouth constriction from narrow to wide.
+The PC3 explains a smaller portion of the variance (`r pc3`%), which relates to the degree of flare of neck from curved to straight shape.
+The first two components account for most of the variance in relation to three phases were represented in Figure \@ref(fig:first-two-PCs-byplot).
+The result reflects a large overlap of three groupings especially for shapes in the pre-European and post-European period that shows some extent of similarity and consistency.
+However, the spread of shape distribution indications a wider variation in shapes in the pre-European and post-European period compared to those in Chinese period along both PC1 and PC2 axes.
+In other words, there is a decreasing in shape variance in the Chinese period that shares the features of shorter height and narrower mouth.
 
-```{r contribution-PCs}
+```{r contribution-PCs, fig.cap= "Pottery shape variation based on scores along principal component axes.
+The rows from the top to the bottom represent the first to third principal component respectively.
+Each column shows the reconstructed shape variable according to the score mean ±2 standard deviation (SD) with a 0.5 interval"}
 # the contributions of each PC to the shapes
 pc_contrib <- PCcontrib(op.p, nax = 1:3) 
+dev.off()
 pc_contrib$gg +  
   geom_polygon(fill="slategrey", col="black") + 
-  ggtitle("Shape variation along PC axes") +
   theme_minimal(base_size = 6)
@@
 # PC2 is neck constriction
-# PC3 is height of neck and roundness of body
+# PC3 is the degree of flare of neck
 ```

### 2019-12-23 analysis/paper/scrap.R

@@ -1,3 +1,3 @@
 # Hypotheses and expectations
 
-My hypothesis is that if foreign contacts including European and Chinese had impact on the emergence of social inequality in local indigenous society due to the monopolies for trade between a small number of indigenous people, then I expect to see the changes in social organization from a more corporate to a more network organization that can be reflected by pottery production at Kiwulan. The emergence of craft specialization, here pottery specialization, are usually related to the presence of elite control. In this case, if the competitions for foreign resources and being trade partners of European or Chinese colonizer among individuals gradually lead to the emergence of social inequality, then I expect the local pottery will show more homogeneous features after contact due to the pottery specialization caused by control of small group of individuals.
+# My hypothesis is that if foreign contacts including European and Chinese had impact on the emergence of social inequality in local indigenous society due to the monopolies for trade between a small number of indigenous people, then I expect to see the changes in social organization from a more corporate to a more network organization that can be reflected by pottery production at Kiwulan. The emergence of craft specialization, here pottery specialization, are usually related to the presence of elite control. In this case, if the competitions for foreign resources and being trade partners of European or Chinese colonizer among individuals gradually lead to the emergence of social inequality, then I expect the local pottery will show more homogeneous features after contact due to the pottery specialization caused by control of small group of individuals.

### 2019-12-23 analysis/paper/paper.Rmd
work on MANOVA table

@@
 The first two components account for most of the variance in relation to three phases were represented in Figure \@ref(fig:first-two-PCs-byplot).
-The result reflects a large overlap of three groupings especially for shapes in the pre-European and post-European period that shows some extent of similarity and consistency.
+The result reflects a large overlap of three groupings especially for shapes in the pre-European and post-European period that might show some extent of similarity.
 However, the spread of shape distribution indications a wider variation in shapes in the pre-European and post-European period compared to those in Chinese period along both PC1 and PC2 axes.
@@
 pc_contrib <- PCcontrib(op.p, nax = 1:3) 
-dev.off()
 pc_contrib$gg +  
@@
 
-```{r MANOVA}
+To test differences in the distribution of shape variables indicated by the PC scores, we use multivariate analysis of variance (MANOVA) to compare a pairwise combination between every sample means across three phases.
+The results in Table \@ref(tab:MANOVA-table) show significant differences in shape between Pre-European and Post-European (p = 0.02) and Pre-European and Chinese contact shapes (p < 0.01), which corresponds to the previous observation of the different distribution in shapes in Chinese period according to PCA byplot.
+In addition, although there is a considerable overlap of shape variables between Pre-European and Post-European, their PC scores are significantly different to each other.
+There is not much differences between the shapes in Post-European and Chinese contact period.
+
+```{r MANOVA-table}
 # MANOVA: Multivariate Analysis of (co)variace
@@
 manv <- MANOVA_PW(op.p, "phase")
-# manv$summary
+su_manv <- manv$summary
 
+knitr::kable(su_manv,
+             caption = "The statistics summary of MANOVA test for the PC scores")
+```
+
+```{r}
 #----------------------------------------------------
@@
 
+# metric data
 ggplot(pottery_data,
@@
 
+# Bayesian Regression Models using 'Stan'
 library(brms)
@@
 
-Multivariate analysis of variance (MANOVA) is used to test the difference in three shape varaibles from three phases.
-The result shows significant differences in shape between Pre-European and Post-European (p = 0.02) and Pre-European and Chinese contact shapes (p < 0.01).
 The distributions of the first three PCs for each vessel suggest variations in shape standardization by period.

### 2019-12-23 analysis/paper/paper.Rmd
add in-text code for p-value, edit background

@@
 
-Pottery analysed in this paper comes from Kiwulan, northeastern Taiwan.
+Pottery analysed in this paper comes from of Kiwulan, northeastern Taiwan.
 Kiwulan is situated on a hill near a riverside at the northern margin of Yilan, which is characterized by a triangular alluvial plain facing eastwards the Pacific and mountains on three other sides.
-Kiwulan dates to AD 1350 to AD 1950, which covers around 600 years from the late Iron Age to the historical period started with the European presence in Taiwan in the early 17th century.
+Kiwulan can be divided into two cultural components, the upper and the lower component, with a sterile layer in between.
+This paper focus on the upper component dated to AD 1350 to AD 1950, which covers around 600 years from the late Iron Age to the historical period started with the European presence in Taiwan in the early 17th century.
 The Dutch first came and occupied southern Taiwan in 1624 and then the Spanish occupied northern Taiwan in 1626.
@@
 
-Pottery found at Kiwulan are all locally made. 
+There is only one type of locally made pottery found at the upper component of Kiwulan.
+There is around 1,200 pots were reconstructed showing the common round body form with a wide variety of impressed geometric motifs.
 Most sherds are fired to an orange to brownish color with a fully oxidized core or a reduced core with oxidized fringes.
@@
 
-To test differences in the distribution of shape variables indicated by the PC scores, we use multivariate analysis of variance (MANOVA) to compare a pairwise combination between every sample means across three phases.
-The results in Table \@ref(tab:MANOVA-table) show significant differences in shape between Pre-European and Post-European (p = 0.02) and Pre-European and Chinese contact shapes (p < 0.01), which corresponds to the previous observation of the different distribution in shapes in Chinese period according to PCA byplot.
-In addition, although there is a considerable overlap of shape variables between Pre-European and Post-European, their PC scores are significantly different to each other.
-There is not much differences between the shapes in Post-European and Chinese contact period.
-
 ```{r MANOVA-table}
@@
 manv <- MANOVA_PW(op.p, "phase")
-su_manv <- manv$summary
+su_manv <- as.data.frame(manv$summary)
+
+pval_chi_pre <- signif(unname(su_manv[,'Pr(>F)'][2]), 4)
+pval_post_pre <- signif(unname(su_manv[,'Pr(>F)'][3]), 4)
 
@@
 
-```{r}
+To test differences in the distribution of shape variables indicated by the PC scores, we use multivariate analysis of variance (MANOVA) to compare a pairwise combination between every sample means across three phases.
+The results in Table \@ref(tab:MANOVA-table) show significant differences in shape between Pre-European and Post-European (p = $`r pval_post_pre`$) and Pre-European and Chinese contact shapes (p = $`r pval_chi_pre`$), which corresponds to the previous observation of the different distribution in shapes in Chinese period according to PCA byplot.
+In addition, although there is a considerable overlap of shape variables between Pre-European and Post-European, their PC scores are significantly different to each other.
+There is not much differences between the shapes in Post-European and Chinese contact period.
+
+```{r GAMM, eval = FALSE}
 #----------------------------------------------------

### 2019-12-23 analysis/paper/paper.Rmd
some questions about code

@@
        legend=c( "Pre-European", "Post-European", "Chinese")) 
+#? How to put the legend out of the plot, tried change the margin using par(), but doesn't work
 ```
@@
 
-# Doesn't work if the plot is not produced by ggplot?
+#? How to plot them next to each other, cowplot doesn't work when the plot not ggplot objects?
 # plot_grid(a, b, labels = c('A', 'B'))
@@
 pc_contrib <- PCcontrib(op.p, nax = 1:3) 
+#?How to hide above plot, tried dev.off but seems it only stop to compute the following one
 pc_contrib$gg +  
@@
 
-manv <- MANOVA_PW(op.p, "phase")
+manv <- MANOVA_PW(op.p, "phase") #? how to hide the process, echo = FALSE?
 su_manv <- as.data.frame(manv$summary)
+# how to reduce the number in the table to only 4 digits, tried mutate with signif but doesn't work?
+#su_manv %>% 
+  #mutate('Pr(>F)' = signif('Pr(>F)', 4))
 

### 2019-12-24 analysis/paper/paper.Rmd
updates for CV test

@@
 
+- craft or pottery specialization
 - CV is hard to test, we apply some statistics testing for it
 
-# Materials and methods
+# Materials and historical background
 
@@
 
-Pottery analysed in this paper comes from of Kiwulan, northeastern Taiwan.
+Pottery analysed in this paper comes from Kiwulan, northeastern Taiwan.
 Kiwulan is situated on a hill near a riverside at the northern margin of Yilan, which is characterized by a triangular alluvial plain facing eastwards the Pacific and mountains on three other sides.
@@
 
-## Outline analysis
+# Methods
 
@@
 # statistical test
-
 pc_plot_tests <- 
@@
 
-The distributions of the first three PCs for each vessel suggest variations in shape standardization by period.
+The standardization of pottery shape for each phase is explored by the distributions of the first three PC scores for each vessel that suggests variations in shape standardization by phase.
 The first PC, the hight of the vessels, shows more variations in the pre-Euro period compared to the Chinese period.
-Standardization appears strong in the pre-Euro period but weak in the Chinese period.
-P-values for a modified signed-likelihood ratio test of equality of CVs of PC1 & PC2 show significant differences in standardization of vessel shape across some periods, especially between Chinese contact with either pre-European or post-European.
+That shows standardization appears strong in the Chinese period but weak in the pre-European period.
+To see whether the differences of the distribution of PCs between each phase is statistically significant, coefficients of variation statistic (CVs) among phase groups of PCs is examined with a modified signed-likelihood ratio test for the equality of CVs.
+P-values for the significant test of equality of CVs of PC1 and PC2 show significant differences in standardization of vessel shape across some periods, especially between Chinese contact with either pre-European or post-European.
 

### 2019-12-25 analysis/paper/paper.Rmd
work on CV test result

@@
 
-```{r CV-test}
+```{r PCs-distribution, fig.cap= "The distribution of PC scores by phases "}
+# what are the variences for each PC in each period?
 # CVs on shape PCs for standardisation
-
 len <- length(op.p$x[,1])
@@
        h = 10, w = 5)
+```
 
-# what are the variences for each PC in each period?
+```{r CV-test-PCs}
 # we don't use CV because that is not good for measurements with no
@@
 
-The standardization of pottery shape for each phase is explored by the distributions of the first three PC scores for each vessel that suggests variations in shape standardization by phase.
-The first PC, the hight of the vessels, shows more variations in the pre-Euro period compared to the Chinese period.
-That shows standardization appears strong in the Chinese period but weak in the pre-European period.
-To see whether the differences of the distribution of PCs between each phase is statistically significant, coefficients of variation statistic (CVs) among phase groups of PCs is examined with a modified signed-likelihood ratio test for the equality of CVs.
-P-values for the significant test of equality of CVs of PC1 and PC2 show significant differences in standardization of vessel shape across some periods, especially between Chinese contact with either pre-European or post-European.
+The variation of pottery shape for each phase is explored by the distributions of the first three PC scores to compare the extent of standardization across phases.
+Figure \@ref(fig:PCs-distribution) suggests variations of PC scores for each vessel by phase.
+The first PC, the hight of the vessels, shows more variations in the pre-European period compared to the Chinese period.
+That indicates standardization appears strong in the Chinese period but weak in the pre-European period.
+The second PC also presents the similar pattern that stronger standardization of shape in the Chinese period compared to other two phases.
+To see whether the differences of the distribution of PCs between any two phases is statistically significant, coefficients of variation statistic (CVs) among phase groups of PCs is examined with a modified signed-likelihood ratio test for the equality of CVs.
+P-values for the significant test of CVs of PC1 and PC2 show significant differences in standardization of vessel shape across some periods, especially between Chinese contact with either pre-European or post-European.
+<--seems not the same plot used before-->
 

### 2019-12-25 analysis/paper/paper.Rmd
update the CV test plot

@@
 
-```{r PCs-distribution, fig.cap= "The distribution of PC scores by phases "}
+```{r PCs-distribution, fig.cap= "The distribution of PC scores by phases"}
 # what are the variences for each PC in each period?
@@
 
-```{r CV-test-PCs}
-# we don't use CV because that is not good for measurements with no
-# absolute zero 
+```{r variance-test-PCs}
+# we don't use CV because that is not good for measurements with no absolute zero
 pc_plot %>% 
@@
 
-# statistical test
+# statistical test to compare the variances of two samples
 pc_plot_tests <- 
@@
   unnest(var_test)
+```
 
+```{r CV-test-PCs, fig.cap= "P-values of the CV equality test of PC1 and PC2 between phases, red color indicates p-value is less than 0.05"}
 library(cvequality)
@@
   mslr_test(nr = 1000, 
-            pc1$x[pc1$y %in% c("Chinese Contact", 
+            pc1$x[pc1$y %in% c("Chinese", 
                                "Pre-European")], 
-            pc1$y[pc1$y %in% c("Chinese Contact", 
+            pc1$y[pc1$y %in% c("Chinese", 
                                "Pre-European")])
@@
   mslr_test(nr = 1000, 
-            pc1$x[pc1$y %in% c("Chinese Contact", 
+            pc1$x[pc1$y %in% c("Chinese", 
                                "Post-European")], 
-            pc1$y[pc1$y %in% c("Chinese Contact", 
+            pc1$y[pc1$y %in% c("Chinese", 
                                "Post-European")])
@@
   mslr_test(nr = 1000, 
-            pc2$x[pc2$y %in% c("Chinese Contact", 
+            pc2$x[pc2$y %in% c("Chinese", 
                                "Pre-European")], 
-            pc2$y[pc2$y %in% c("Chinese Contact", 
+            pc2$y[pc2$y %in% c("Chinese", 
                                "Pre-European")])
@@
   mslr_test(nr = 1000, 
-            pc2$x[pc2$y %in% c("Chinese Contact", 
+            pc2$x[pc2$y %in% c("Chinese", 
                                "Post-European")], 
-            pc2$y[pc2$y %in% c("Chinese Contact", 
+            pc2$y[pc2$y %in% c("Chinese", 
                                "Post-European")])
@@
              colour = I(colour),
-             shape = PC)
-         ) +
+             shape = PC)) +
   geom_point(size = 8,  position = position_dodge(width = 0.5)) +
@@
 The second PC also presents the similar pattern that stronger standardization of shape in the Chinese period compared to other two phases.
-To see whether the differences of the distribution of PCs between any two phases is statistically significant, coefficients of variation statistic (CVs) among phase groups of PCs is examined with a modified signed-likelihood ratio test for the equality of CVs.
-P-values for the significant test of CVs of PC1 and PC2 show significant differences in standardization of vessel shape across some periods, especially between Chinese contact with either pre-European or post-European.
-<--seems not the same plot used before-->
+To see whether the differences in the distribution of PCs between any two phases is statistically significant, coefficients of variation statistic (CVs) among phase groups of PCs is examined with a modified signed-likelihood ratio test for the equality of CVs.
+P-values for the significant test of CVs of PC1 and PC2 show significant differences (P-value < 0.01) in standardization of vessel shape across some periods, especially between Chinese contact with either pre-European or post-European Figure \@ref(fig:CV-test-PCs).
 

### 2019-12-27 analysis/paper/paper.Rmd
some updates about background

@@
 
-There is only one type of locally made pottery found at the upper component of Kiwulan.
-There is around 1,200 pots were reconstructed showing the common round body form with a wide variety of impressed geometric motifs.
-Most sherds are fired to an orange to brownish color with a fully oxidized core or a reduced core with oxidized fringes.
-It seems that there is no significant relationship between color and phases.
+More than 550 thousands of potsherds were excavated from the upper component at Kiwulan, and around 1,200 vessels were reconstructed.
+Those vessels show high consistency in shape, which is globular body with short neck and wide mouth.
+A wide variety of impressed geometric motifs applied all over the exterior body surface are observed.
+They were believed as cooking pots according to the evidence of chars commonly observed on both interior and exterior surface.
+Most pots are fired to an orange to brownish color with a fully oxidized core or a reduced core with oxidized fringes.
+
+## Analyses for pottery: Metric measurements and petrography
+
+We examine 291 pots recovered from Kiwulan site.
+The layer from 1 to 6 could be divided into 3 time periods.
+Layer 5 and 6 represent pre-contact period, layer 4 represents contact period, and layer1, 2, and 3 represent post-contact period.
+The amount of pots for each layer shows below.
+Although most pots are not complete, the thickness from rim, neck, to body can be measured.
+I have also measured the diameter of rim, neck, and body.
+For those pots that are incomplete, the diameter is measured by its curvature.
+Since the height is incomplete for most pots, this preliminary analysis focuses more on the possible change in thickness and diameter of pot for different parts, and their ratio over time.
+
 Petrographic analysis for thin sections was conducted at the Department of Geosciences, National Taiwan University.
-The results shows high percentage of inclusions (15-50%), including argillite, quartz, feldspar, metasandstone, sandstone fragment, and slate.
-Those minerals are believed to be local materials of the Yilan Plain.
+The results show high percentage of inclusions (15-50%), including argillite, quartz, feldspar, metasandstone, sandstone fragment, and slate.
 The most common type of sherds, also the dominant one, has a high percentage of inclusion (25-40%), which is composed mainly by argillite (15-40%), followed by metasandstone (1-10%), sandstone (1-6%), and quartz (1-5%).
@@
 Overall, this type presents a mixture of fine, rounded argillite with a small part of rounded metasandstone and rounded to sub-angular monocrystalline quartz.
+Those minerals are believed to be local materials of the Yilan Plain.
 
@@
 
+## Samples
+
 A total of `r total_no` vessels from pre-contact contexts (n = 32), post-European contexts (n = 27), and Chinese contact contexts (n = 14) were selected for this study because of their completeness mostly covering parts from rim to bottom.
@@
 Each traced half cross-section image was duplicated, flipped, and then joined with another one to create a 2D closed outline.
-
-## Metric measurements
-
-In this study, we examine 291 pots recovered from Kiwulan site.
-The layer from 1 to 6 could be divided into 3 time periods.
-Layer 5 and 6 represent pre-contact period, layer 4 represents contact period, and layer1, 2, and 3 represent post-contact period.
-The amount of pots for each layer shows below.
-Although most pots are not complete, the thickness from rim, neck, to body can be measured.
-I have also measured the diameter of rim, neck, and body.
-For those pots that are incomplete, the diameter is measured by its curvature.
-Since the height is incomplete for most pots, this preliminary analysis focuses more on the possible change in thickness and diameter of pot for different parts, and their ratio over time.
 
@@
 
+We would like to thank the Yilan County Cultural Affairs Bureau in Taiwan for permitting access to the pottery used in this study.
+We thank Dr.
+Wen-Shan Chen in the Department of Geosciences, National Taiwan University for his invaluable guidance for petrographic analysis in his lab.
+
 <!-- The following line inserts a page break when the output is MS Word. For page breaks in PDF, use \newpage on its own line.  -->

### 2019-12-27 analysis/paper/paper.Rmd
fix typo and grammer check

@@
 Landmarks, curves or outlines of objects can be represented by coordinates by their unique point location with respect to numerical values of coordinate axes.
-According to different focuses and nature of original shape data, there are two common morphometric methods used to capture and analyse the shapes, landmark approaches and outline approaches [@Adams2004].
+According to different focuses and nature of original shape data, there are two common morphometric methods used to capture and analyze the shapes, landmark approaches and outline approaches [@Adams2004].
 Landmark approaches assign a set of landmarks or semilandmarks onto objects as reference points that can be specified on a coordinate system as x, y coordinates for two dimensions, or x, y, z coordinates for three dimensions.
@@
 GPA is a method that superimposes sets of landmark configurations into a common coordinate system, where superimposed landmark coordinates are used as shape variables for further multivariate statistics analyses[@Slice2007].
-Landmark-based morphometrics are widely used for archaeological objects where morphological signature is obvious to identity which provides ideal reference points for landmark placement, such as projectile point tips or biologically measurements points indicating osteological features [@Birch2019; @Buchanan2019; @Lycett2013; @Cardillo2010; @Haruda2019; @Meloro2015].
+Landmark-based morphometrics is widely used for archaeological objects where a morphological signature is obvious to identity which provides ideal reference points for landmark placement, such as projectile point tips or biologically measurements points indicating osteological features [@Birch2019; @Buchanan2019; @Lycett2013; @Cardillo2010; @Haruda2019; @Meloro2015].
 
@@
 In those cases, outline approaches serve as a solution by analyzing the overall shape of a morphological structure.
-One of the outline approaches is semi-landmarks method, also called sliding landmarks, which assigns points along the curve between two landmarks at defined intervals [@Bookstein1997; @Lawing2010].
+One of the outline approaches is the semi-landmarks method, also called sliding landmarks, which assigns points along the curve between two landmarks at defined intervals [@Bookstein1997; @Lawing2010].
 Those semi-landmarks are allowed to slide along the curve to remove the effect of the arbitrary landmark spacing by minimizing either Procrustes distance or bending energy [@Bookstein1997; @Slice2007; @Gunz2013].
@@
 The number of harmonics determines the quality and precision of the geometry of an object.
-The harmonic power, a cumulated sum of squared harmonic coefficient, provides a robust rule for desired number of harmonics [@Bonhomme2014].
-The application of EFA in archaeology is mostly used for bone morphology to understand the biological variation and recently more for stone artefacts to explore stone tool technology [@Fox2015; @Hoggard2019; @Iovictua2010].
+The harmonic power, a cumulated sum of squared harmonic coefficient, provides a robust rule for the desired number of harmonics [@Bonhomme2014].
+The application of EFA in archaeology is mostly used for bone morphology to understand the biological variation and recently more for stone artifacts to explore stone tool technology [@Fox2015; @Hoggard2019; @Iovictua2010].
 Outline approach based on EFA is suitable for shapes lacking obvious landmarks or curves convey more important information indicating the meaningful variations, for example, ceramics.
 @Wilczek2014 evaluate the concordance between outline-based approaches, the EFA and the Discrete Cosine Transform (DCT), and traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France.
-They found that the variation in ceramic shapes analysed by EFA and DCT matches traditional ceramic typology, which supports that outline-based approaches can be efficiently used for studying variations in ceramic shapes.
-In addition, the results of shape variation by EFA helps to understand the level of production standardisation over time across that region.
+They found that the variation in ceramic shapes analyzed by EFA and DCT matches traditional ceramic typology, which supports that outline-based approaches can be efficiently used for studying variations in ceramic shapes.
+In addition, the results of shape variation by EFA helps to understand the level of production standardization over time across that region.
 
@@
 
-Pottery analysed in this paper comes from Kiwulan, northeastern Taiwan.
+Pottery analyzed in this paper comes from Kiwulan, northeastern Taiwan.
 Kiwulan is situated on a hill near a riverside at the northern margin of Yilan, which is characterized by a triangular alluvial plain facing eastwards the Pacific and mountains on three other sides.
@@
 In 1642, the Spanish was expelled by the Dutch, who took over their forts in northern Taiwan.
-Since then, western Taiwan was mostly under the colonial rule of the Dutch until 1662 when the Kingdom of Tungning was founded by Koxinga, a loyalist of Ming dynasty of China.
+Since then, western Taiwan was mostly under the colonial rule of the Dutch until 1662 when the Kingdom of Tungning was founded by Koxinga, a loyalist of the Ming dynasty of China.
 
-The upper component experienced frequent foreign contacts including the European colonial presence in the 17th century and great waves of Chinese immigrants since 19th century.
+The upper component experienced frequent foreign contacts including the European colonial presence in the 17th century and great waves of Chinese immigrants since the 19th century.
 The excavation revealed abundant artifacts in which potsherds are the dominate materials throughout the site.
 Imported ceramics from mainland China, stonewares, and ornament elements such as beads were also commonly found in the upper component that indicates the frequent long-distance trade activities in the 17th century during the European presence.
-In addition to artifacts, features were also found, such as burials, middens, and post-holes with in-situ posts that explains Kiwulan was a large settlement site.
+In addition to artifacts, features were also found, such as burials, middens, and post-holes with in-situ posts that explain Kiwulan was a large settlement site.
 
 More than 550 thousands of potsherds were excavated from the upper component at Kiwulan, and around 1,200 vessels were reconstructed.
-Those vessels show high consistency in shape, which is globular body with short neck and wide mouth.
+Those vessels show high consistency in shape - globular body with short neck and wide mouth.
 A wide variety of impressed geometric motifs applied all over the exterior body surface are observed.
-They were believed as cooking pots according to the evidence of chars commonly observed on both interior and exterior surface.
-Most pots are fired to an orange to brownish color with a fully oxidized core or a reduced core with oxidized fringes.
+They were believed as cooking pots according to the evidence of chars commonly observed on both interior and exterior surfaces.
+Most pots are fired to orange to brownish color with a fully oxidized core or a reduced core with oxidized fringes.
 
@@
 Although most pots are not complete, the thickness from rim, neck, to body can be measured.
-I have also measured the diameter of rim, neck, and body.
+I have also measured the diameter of the rim, neck, and body.
 For those pots that are incomplete, the diameter is measured by its curvature.
@@
 Geometric morphometric analyses were conducted in the R software (www.rproject.org, Core-Team, 2015) using the functions included in the Momocs, a R package intended to quantify the shape and compare its variation, especially for outline analysis [@Bonhomme2014].
-The digitised outlines were converted into a list of successsive x-y pixel coordinates for elliptic Fourier analysis (EFA), which assesses morphological differences among pottery shapes from three occupation contexts.
+The digitised outlines were converted into a list of successive x-y pixel coordinates for elliptic Fourier analysis (EFA), which assesses morphological differences among pottery shapes from three occupation contexts.
 
@@
 With the third component, the first three principal components explain `r vari_three`% of the total variances.
-Figure \@ref(fig:contribution-PCs) shows the shape variation associated to the first three principal components.
-From the negative to the positive scores, the PC1 represents the height of the vessels from tall to short and the roundness of the body from round to oval shaped.
+Figure \@ref(fig:contribution-PCs) shows the shape variation associated with the first three principal components.
+From the negative to the positive scores, the PC1 represents the height of the vessels from tall to short and the roundness of the body from round to oval-shaped.
 The PC2 relates to the neck and mouth constriction from narrow to wide.
-The PC3 explains a smaller portion of the variance (`r pc3`%), which relates to the degree of flare of neck from curved to straight shape.
+The PC3 explains a smaller portion of the variance (`r pc3`%), which relates to the degree of the flare of the neck from curved to straight shape.
 The first two components account for most of the variance in relation to three phases were represented in Figure \@ref(fig:first-two-PCs-byplot).
-The result reflects a large overlap of three groupings especially for shapes in the pre-European and post-European period that might show some extent of similarity.
-However, the spread of shape distribution indications a wider variation in shapes in the pre-European and post-European period compared to those in Chinese period along both PC1 and PC2 axes.
+The result reflects a large overlap of three groupings especially for shapes in the pre-European and post-European periods that might show some extent of the similarity.
+However, the spread of shape distribution indications a wider variation in shapes in the pre-European and post-European periods compared to those in the Chinese period along both PC1 and PC2 axes.
 In other words, there is a decreasing in shape variance in the Chinese period that shares the features of shorter height and narrower mouth.
@@
 The results in Table \@ref(tab:MANOVA-table) show significant differences in shape between Pre-European and Post-European (p = $`r pval_post_pre`$) and Pre-European and Chinese contact shapes (p = $`r pval_chi_pre`$), which corresponds to the previous observation of the different distribution in shapes in Chinese period according to PCA byplot.
-In addition, although there is a considerable overlap of shape variables between Pre-European and Post-European, their PC scores are significantly different to each other.
-There is not much differences between the shapes in Post-European and Chinese contact period.
+In addition, although there is a considerable overlap of shape variables between Pre-European and Post-European, their PC scores are significantly different from each other.
+There are not much differences between the shapes in the Post-European and Chinese contact periods.
 
@@
 That indicates standardization appears strong in the Chinese period but weak in the pre-European period.
-The second PC also presents the similar pattern that stronger standardization of shape in the Chinese period compared to other two phases.
+The second PC also presents a similar pattern that stronger standardization of shape in the Chinese period compared to the other two phases.
 To see whether the differences in the distribution of PCs between any two phases is statistically significant, coefficients of variation statistic (CVs) among phase groups of PCs is examined with a modified signed-likelihood ratio test for the equality of CVs.
-P-values for the significant test of CVs of PC1 and PC2 show significant differences (P-value < 0.01) in standardization of vessel shape across some periods, especially between Chinese contact with either pre-European or post-European Figure \@ref(fig:CV-test-PCs).
+P-values for the significant test of CVs of PC1 and PC2 show significant differences (P-value < 0.01) in the standardization of vessel shape across some periods, especially between Chinese contact with either pre-European or post-European Figure \@ref(fig:CV-test-PCs).
 

### 2019-12-28 analysis/paper/paper.Rmd
add sentence to discussion section and reorganize some sections

@@
 
-Ceramic production can reflect prehistoric socioeconomic patterns since it relates to not only economic base for households but also societies as a whole when specialized production was organized.
+Ceramic specialization can reflect prehistoric socioeconomic patterns since it relates to not only economic base for households but also societies as a whole when specialized production was organized.
 Identifying the specialization in ceramic production will help to understand how complex societies form and further explore the underlying mechanism.
@@
 The emergence of craft specialization, here pottery specialization, are usually related to the presence of elite control.
-In this case, if the competitions for foreign resources and being trade partners of European or Chinese colonizer among individuals gradually lead to the emergence of social inequality, then I expect the local pottery will show more homogeneous features after contact due to the craft specialization caused by control of small group of individuals.
+In this case, if the competitions for foreign resources and being trade partners of European or Chinese colonizer among individuals gradually lead to the emergence of social inequality, then we expect the local pottery will show more homogeneous features after contact due to the craft specialization caused by control of small group of individuals.
 The first step is to detecting pottery standardization when discussing the emergence of craft specialization and the possible changes in social organization.
@@
 
-# Geometric Morphometric approach: shape analysis for pottery specialization
+## Geometric Morphometric approach: shape analysis for pottery specialization
 
@@
 
-# Materials and historical background
+# Materials
 
-## Archaeological pottery from Kiwulan
+## Archaeological background
 
@@
 Most pots are fired to orange to brownish color with a fully oxidized core or a reduced core with oxidized fringes.
-
-## Analyses for pottery: Metric measurements and petrography
 
@@
 
-## Samples
+A total of `r total_no` vessels from pre-contact contexts (n = 32), post-European contexts (n = 27), and Chinese contact contexts (n = 14) were selected for this study because of their completeness mostly covering parts from rim to bottom.
 
-A total of `r total_no` vessels from pre-contact contexts (n = 32), post-European contexts (n = 27), and Chinese contact contexts (n = 14) were selected for this study because of their completeness mostly covering parts from rim to bottom.
+# Methods
+
+## Digitising and analysing by EFA
+
 The scanned pottery drawings of those pots were acquired from Bureau of Cultural Affairs in Yilan.
@@
 
-# Methods
-
 Geometric morphometric analyses were conducted in the R software (www.rproject.org, Core-Team, 2015) using the functions included in the Momocs, a R package intended to quantify the shape and compare its variation, especially for outline analysis [@Bonhomme2014].
@@
 The harmonic coefficients generated by EFA were analysed by principal component analysis (PCA) to illustrate the diversity of the shape data and identify the major patterns of variation through dimensionality reduction.
+
+## Statistical testing
+
 The principal components (PCs) scores were analysed with a multivariate analysis of variance (MANOVA) to test for significant effects of the groups of occupation context on shape variances.
@@
 
+The results confirm that the differences in pottery shapes can be detected using outline analysis approaches.
+It shows there is differences in pottery shapes between pre-European and post-European, and Pre-European and Chinese contact.
+A increasing standardization of shape of pottery during the Chinese contact period compared to the other two periods is observed.
+This study provide an comparable basis for our understanding of the standardized pottery as an indicator of social changes in a cultural contact conditions.
+Northeastern Taiwan had been encountered two major foreign contacts in the 17th century with the Europeans and 19th century with the Chines immigrants.
+It shows that the standardization of pottery shape appears during the European presence and increased in the Chinese period.
+This may suggest that an emergence of specialized groups or a consensus of the adoption of similar shape of pottery in the society.
+The changes in the standardization of pottery may indicate a major social changes in the local indigenous organization.
+When it comes to ceramic specialization, metric measurements and clay paste are two major indicators.
+Shape analysis could be viewed as a replacement of metric measurement that enables subtle changes to be detected.
+We observe the increasing standardization of the pottery shape, however, it is interesting to note that there is no obvious changes in the clay paste across different phases.
+
 # Conclusion

### 2019-12-28 analysis/paper/paper.Rmd
add references

@@
 Kiwulan is situated on a hill near a riverside at the northern margin of Yilan, which is characterized by a triangular alluvial plain facing eastwards the Pacific and mountains on three other sides.
-Kiwulan can be divided into two cultural components, the upper and the lower component, with a sterile layer in between.
+Kiwulan can be divided into two cultural components, the upper and the lower component, with a sterile layer in between [@Chen2007].
 This paper focus on the upper component dated to AD 1350 to AD 1950, which covers around 600 years from the late Iron Age to the historical period started with the European presence in Taiwan in the early 17th century.
@@
 In 1642, the Spanish was expelled by the Dutch, who took over their forts in northern Taiwan.
-Since then, western Taiwan was mostly under the colonial rule of the Dutch until 1662 when the Kingdom of Tungning was founded by Koxinga, a loyalist of the Ming dynasty of China.
+Since then, western Taiwan was mostly under the colonial rule of the Dutch until 1662 when the Kingdom of Tungning was founded by Koxinga, a loyalist of the Ming dynasty of China [@Andrade2007].
 
@@
 Imported ceramics from mainland China, stonewares, and ornament elements such as beads were also commonly found in the upper component that indicates the frequent long-distance trade activities in the 17th century during the European presence.
-In addition to artifacts, features were also found, such as burials, middens, and post-holes with in-situ posts that explain Kiwulan was a large settlement site.
+In addition to artifacts, features were also found, such as burials, middens, and post-holes with in-situ posts that explain Kiwulan was a large settlement site [@Chen2007].
 

### 2019-12-31 analysis/paper/paper.Rmd
some updates, work on introduction a little

@@
 abstract: |
-  Text of abstract
+  The emergence of ceramic specialization in prehistoric societies is often linked to shifts in the complexity of social structures, because specialized pottery production can reflect craft specialization and the presence of elite control.
+Previous work on identifying specialization relies on typological or linear metric analysis.
+Here we demonstrate how to investigate ceramic standardization by analyzing outlines of ceramic vessels.
+Outline analysis is an important method because, unlike more commonly-used landmark analysis methods, it can effectively quantify shape differences for objects that lack distinctive measurement points needed for landmark analysis.
+We demonstrate this method using pottery from Kiwulan, a large multi-component Iron Age site (AD 1350-1850) in northeast Taiwan.
+To measure ceramic specialization, we quantified pottery standardization by analyzing shape variables with reproducible geometric morphometric methods.
+We computed coefficients of variation (CVs) for shape coefficients obtained by elliptical Fourier analysis to test for shape standardization.
+We found significant differences in pottery shape and shape standardization that indicate changes in pottery production resulting from contact with mainland Han Chinese groups in northeast Taiwan.
+We infer increasing craft specialization and changes in social organization.
+Our case study, which includes an openly available research compendium of R code, represents an innovative application of outline-based methods in geometric morphometry to answer the anthropological questions of craft specialization.
+This study implies that craft specialization can be a deliberate act of resistance to show indigenous identity and ethnicity in a direct culture contact situation.
 keywords: |
@@
 
-Ceramic specialization can reflect prehistoric socioeconomic patterns since it relates to not only economic base for households but also societies as a whole when specialized production was organized.
+- the definition of craft specialization
+
+The emergence of craft specialization is usually associated with the changes in a socio-economic system of a society since it links to the organization of production
+
+context, concentration, scale, intensity
+
+Ceramic specialization can reflect prehistoric socio-economic patterns since it relates to not only economic base for households but also societies as a whole when specialized production was organized.
+
+Identifying the specialization in ceramic production will help to understand how complex societies form and further explore the underlying mechanism.
+One way to examine ceramic specialization is to identify whether ceramics present standardization in production, which based on the assumption that the specialized mass production will lead to uniformity of the product due to increased skills, routinization, and lower diversity of producers that enable us to interpret social organization or social relations such as type of production, and the presence of elite control [@Arnold2000, pp.
+334; @Stark1995; @Costin1991].
+
+social changes -  culture contact framework 
+- the cause of emergence of craft specialization, elite control, rice intensification, political, economic 
+- here I have the third cultural resistance, identity take a look at social boundary
+- standardization as indicator, metric and technological
+- GM is more sensitive for subtle changes, and the outline
+- innovative: CV test 
+- hypothesis, goals
+
+Ceramic specialization can reflect prehistoric socio-economic patterns since it relates to not only economic base for households but also societies as a whole when specialized production was organized.
 Identifying the specialization in ceramic production will help to understand how complex societies form and further explore the underlying mechanism.
@@
 
-Pottery analyzed in this paper comes from Kiwulan, northeastern Taiwan.
+Ceramics analyzed in this paper come from 40 units (4m by 4m) sampled from 4 connected trenches at Kiwulan, northeastern Taiwan.
 Kiwulan is situated on a hill near a riverside at the northern margin of Yilan, which is characterized by a triangular alluvial plain facing eastwards the Pacific and mountains on three other sides.
-Kiwulan can be divided into two cultural components, the upper and the lower component, with a sterile layer in between [@Chen2007].
-This paper focus on the upper component dated to AD 1350 to AD 1950, which covers around 600 years from the late Iron Age to the historical period started with the European presence in Taiwan in the early 17th century.
-The Dutch first came and occupied southern Taiwan in 1624 and then the Spanish occupied northern Taiwan in 1626.
+The chronology of Kiwulan can be divided into two cultural components, the upper and the lower component, with a sterile layer in between [@Chen2007].
+This paper focuses on the upper component dating from AD 1350 to AD 1950, which covers the late Iron Age and the historical period defined by the European presence in Taiwan in the early 17th century.
+The Dutch firstly came and occupied southern Taiwan in 1624 and then the Spanish occupied northern Taiwan in 1626.
 In 1642, the Spanish was expelled by the Dutch, who took over their forts in northern Taiwan.
@@
 
-More than 550 thousands of potsherds were excavated from the upper component at Kiwulan, and around 1,200 vessels were reconstructed.
-Those vessels show high consistency in shape - globular body with short neck and wide mouth.
-A wide variety of impressed geometric motifs applied all over the exterior body surface are observed.
-They were believed as cooking pots according to the evidence of chars commonly observed on both interior and exterior surfaces.
+More than 550 thousands of sherds were excavated from the upper component at Kiwulan, and around 1,200 vessels were reconstructed.
+There are only two forms of vessels, one is cooking pot and another one is steamer made of two pots stacking together.
+The pots show high consistency in shape - globular body with short neck and wide mouth.
+A wide variety of impressed geometric motifs are applied all over the exterior body surface.
+They were believed used for cooking according to the evidence of charred residue commonly observed on interior surface and soot on exterior surface.
+Besides, there is no other utilitarian vessels found at this site.
 Most pots are fired to orange to brownish color with a fully oxidized core or a reduced core with oxidized fringes.
+Those pots might be pinched and shaped by hand according to the pressed finger marks usually observed inside of pots.
 

### 2019-12-31 analysis/paper/paper.Rmd
work on introduction

@@
 
-The emergence of craft specialization is usually associated with the changes in a socio-economic system of a society since it links to the organization of production
-
-context, concentration, scale, intensity
+The development of craft specialization is usually associated with the social changes since it reflects production systems that inform social, economic, or political conditions of a society.
+Specialization is defined as "differential participation in specific economic activities", in which a particular good is produced by fewer producers compared to the total population who consume it [@Costin1991; @Costin2001].
+The implication of specialized production in relation to social structure or social process became a interest in archaeological investigations.
+From a political perspective, craft specialization is viewed as a strategy for elites to control over production systems in order to maintain their power and authority [@Costin2001].
+The relationship between craft specialization and social differentiation is broadly discussed and many cases use specialized production as an indicator of increasing socio-political complexity [@Stark1995; @Hirshman2010; @Junker1999].
 
 Ceramic specialization can reflect prehistoric socio-economic patterns since it relates to not only economic base for households but also societies as a whole when specialized production was organized.
-
 Identifying the specialization in ceramic production will help to understand how complex societies form and further explore the underlying mechanism.

### 2019-12-31 analysis/paper/paper.Rmd
add some references

@@
 
-- the definition of craft specialization
-
 The development of craft specialization is usually associated with the social changes since it reflects production systems that inform social, economic, or political conditions of a society.
@@
 
+- the cause of emergence of craft specialization, elite control, rice intensification, political, economic
+
 Ceramic specialization can reflect prehistoric socio-economic patterns since it relates to not only economic base for households but also societies as a whole when specialized production was organized.
@@
 
-social changes -  culture contact framework 
-- the cause of emergence of craft specialization, elite control, rice intensification, political, economic 
+- social changes -  culture contact framework 
 - here I have the third cultural resistance, identity take a look at social boundary
@@
 
-Several variables have been proposed to measure ceramic standardization, including metric, compositional, and technological attributes [@Rice1991; @Blackman1993; @Costin1991; @Hirshman2010; @Tite1999].
+Several variables have been proposed to measure ceramic standardization, including metric, compositional, and technological attributes [@Rice1991; @Blackman1993; @Costin1991; @Hirshman2010; @Tite1999; @Eerkens2001].
 However, traditional typological and linear measurements are limited because they can be insensitive to subtle variations resulting from changes in craft specialization.

### 2020-01-01 analysis/paper/paper.Rmd
work on introduction

--- 2019-12-31 analysis/paper/paper.Rmd
@@
 
-- the definition of craft specialization
+The development of craft specialization is usually associated with the social changes since it reflects production systems that inform social, economic, or political conditions of a society.
+Specialization is defined as "differential participation in specific economic activities", in which a particular good is produced by fewer producers compared to the total population who consume it [@Costin1991; @Costin2001, pp.
+43].
+The implication of the emergence of specialization for social structure or social process has been a theme in archaeological investigations.
+According to the political approaches, craft specialization, attached production specifically, is viewed as a strategy used by elites to control over production systems for the sake of power and authority [@Costin2001].
+The relationship between craft specialization and social differentiation is broadly discussed, and many cases show that specialized production can be a useful indicator of increasing socio-political complexity [@Hirshman2010; @Junker1999].
+Economic models consider production, exchange, and consumption as a whole and explain that specialization is a subsistence strategy to accommodate uneven distribute of resources [@Costin2001; @Arnold1994], or a reaction to the changes in subsistence practices [@Stark1995economi].
+However, the explanation for the emergence of specialized production emerges might not be exclusive or straightforward.
+Also, specialization is not necessarily accompanied with social differentiation that is observed in lowland northern Philippines with low degree of ceramic specialization during the Spanish colonial period [@Acabado2018].
+This implies that the condition and contexts under which specialists organized are important for interpretation of changes in social organization [@Roux2018].
 
-The emergence of craft specialization is usually associated with the changes in a socio-economic system of a society since it links to the organization of production
-
-context, concentration, scale, intensity
-
-Ceramic specialization can reflect prehistoric socio-economic patterns since it relates to not only economic base for households but also societies as a whole when specialized production was organized.
-
-Identifying the specialization in ceramic production will help to understand how complex societies form and further explore the underlying mechanism.
-One way to examine ceramic specialization is to identify whether ceramics present standardization in production, which based on the assumption that the specialized mass production will lead to uniformity of the product due to increased skills, routinization, and lower diversity of producers that enable us to interpret social organization or social relations such as type of production, and the presence of elite control [@Arnold2000, pp.
-334; @Stark1995; @Costin1991].
-
-social changes -  culture contact framework 
-- the cause of emergence of craft specialization, elite control, rice intensification, political, economic 
-- here I have the third cultural resistance, identity take a look at social boundary
 - standardization as indicator, metric and technological
@@
 - innovative: CV test
+
+craft specialization is commonly identified through statistical standardization of specific dimensions of a particular item – standardization as evidence of specialization.
+
+The variation of standardization of specialized product is commonly used as indicator to identify craft specialization based on the assumption that specialized mass production will lead to uniformity of the product due to routinization, increased skills, and fewer number of producers involved in production [@Arnold2000, pp.
+334; @Stark1995; @Costin1991; @Costin2001].
+For ceramic standardization, several measurements have been proposed, including metric, compositional, and technological approaches [@Rice1991; @Blackman1993; @Costin1991; @Tite1999; @Roux2015].
+Among those variables, linear measurements is widely used by calculation coefficient of variation (CV) that lower CV value indicates higher degree of standardization [@Hirshman2010; @Eerkens2001].
+However, traditional typological and linear measurements are limited because they can be insensitive to subtle variations resulting from changes in craft specialization.
+
+Taking a new approach to the measurement of craft specialization, we studied shape and standardization of locally made pottery to identify changes in pottery production at Kiwulan (1400-1900 AD), a large multi-component archaeological site in northeastern Taiwan.
+We apply geometric morphometric methods to study artifact shapes to investigate if there are any changes resulting from foreign contact of European and Chinese that might indicate social changes in the indigenous society.
+
+- social changes -  culture contact framework 
+- here I have the third cultural resistance, identity take a look at social boundary
 - hypothesis, goals
 
-Ceramic specialization can reflect prehistoric socio-economic patterns since it relates to not only economic base for households but also societies as a whole when specialized production was organized.
-Identifying the specialization in ceramic production will help to understand how complex societies form and further explore the underlying mechanism.
-One way to examine ceramic specialization is to identify whether ceramics present standardization in production, which based on the assumption that the specialized mass production will lead to uniformity of the product due to increased skills, routinization, and lower diversity of producers that enable us to interpret social organization or social relations such as type of production, and the presence of elite control [@Arnold2000, pp.
-334; @Stark1995; @Costin1991].
-
-Several variables have been proposed to measure ceramic standardization, including metric, compositional, and technological attributes [@Rice1991; @Blackman1993; @Costin1991; @Hirshman2010; @Tite1999].
-However, traditional typological and linear measurements are limited because they can be insensitive to subtle variations resulting from changes in craft specialization.
-Taking a new approach to the measurement of craft specialization, we studied shape and standardization of locally made pottery to identify changes in pottery production at Kiwulan (1400-1900 AD), a large multi-component archaeological site in northeastern Taiwan.
-We apply geometric morphometric methods to study artifact shapes to investigate if there are any changes resulting from foreign contact of European and Chinese that might indicate social changes in the indigenous society.
+This paper
 
@@
 The excavation revealed abundant artifacts in which potsherds are the dominate materials throughout the site.
-Imported ceramics from mainland China, stonewares, and ornament elements such as beads were also commonly found in the upper component that indicates the frequent long-distance trade activities in the 17th century during the European presence.
+Imported ceramics from mainland China, stonewares, and ornaments elements such as beads were also commonly found in the upper component that indicates the frequent long-distance trade activities in the 17th century during the European presence.
 In addition to artifacts, features were also found, such as burials, middens, and post-holes with in-situ posts that explain Kiwulan was a large settlement site [@Chen2007].
+The predominant artefact found at Kiwulan is locally made ceramics, which distributed throughout the sequence across the site.
 
 More than 550 thousands of sherds were excavated from the upper component at Kiwulan, and around 1,200 vessels were reconstructed.
-There are only two forms of vessels, one is cooking pot and another one is steamer made of two pots stacking together.
-The pots show high consistency in shape - globular body with short neck and wide mouth.
-A wide variety of impressed geometric motifs are applied all over the exterior body surface.
-They were believed used for cooking according to the evidence of charred residue commonly observed on interior surface and soot on exterior surface.
-Besides, there is no other utilitarian vessels found at this site.
+It is interesting to note that there are only two forms of vessels recovered, one is cooking pot and another one is steamer made of two cooking pots stacked together with a filter in between.
+Those pots show high consistency in shape - globular body with short neck and wide mouth.
+The exterior surface below the neck is treated with a wide variety of impressed geometric motifs.
+They were believed used for cooking because of the evidence of charred residues or carbon deposits commonly observed on the interior and soot on the exterior.
+Besides, there is no other utilitarian earthenware ceramics found at this site suggests this type of pot was mainly used for daily cooking.
 Most pots are fired to orange to brownish color with a fully oxidized core or a reduced core with oxidized fringes.
-Those pots might be pinched and shaped by hand according to the pressed finger marks usually observed inside of pots.
+Finger impressions and seams usually on the interior indicate they were pinched using slabs of clay and shaped by hand.
 

### 2020-01-02 analysis/paper/paper.Rmd
updates for introduction

@@
 The relationship between craft specialization and social differentiation is broadly discussed, and many cases show that specialized production can be a useful indicator of increasing socio-political complexity [@Hirshman2010; @Junker1999].
-Economic models consider production, exchange, and consumption as a whole and explain that specialization is a subsistence strategy to accommodate uneven distribute of resources [@Costin2001; @Arnold1994], or a reaction to the changes in subsistence practices [@Stark1995economi].
+Economic models consider production, exchange, and consumption as a whole and explain that specialization is a subsistence strategy to accommodate uneven distribute of resources [@Costin2001; @Arnold1994], or a reaction to the changes in subsistence practices [@Stark1995economic].
 However, the explanation for the emergence of specialized production emerges might not be exclusive or straightforward.
@@
 
-- standardization as indicator, metric and technological
-- GM is more sensitive for subtle changes, and the outline
-- innovative: CV test
+The variation of standardization of specialized product is commonly used as indicator to identify craft specialization based on the assumption that specialized mass production will lead to uniformity of the product due to routinization, increased skills, and fewer number of producers involved in production [@Stark1995; @Costin1991; @Costin2001].
+For ceramic standardization, several measurements have been proposed, such as metric, compositional, and technological approaches [@Arnold2000; @Rice1991; @Blackman1993; @Costin1991; @Tite1999; @Roux2015].
+Among those variables, metric measurement is widely applied in archaeological assemblages by calculating coefficient of variation between sets of data coupled with statistical comparison that lower value reflects higher degree of standardization [@Eerkens2001; @Roux2003].
+However, this traditional typological and linear measurements might be limited because they can be insensitive to subtle variations resulting from changes in assembles.
+Shape analysis based on geometric morphometrics method are new approaches for measuring the degree of standardization since it takes the overall shape as variable for comparison [@Slice2007].
+Taking geometric morphomerics approaches to the measurement of craft specialization, we study shape and standardization of locally made ceramic to identify changes in ceramic production in cross-cultural interaction contexts at Kiwulan (1350-1950 AD), a large multi-component archaeological site in northeastern Taiwan.
+The aim of this paper is to investigate if there are increasing ceracmic specialization resulting from interaction with the Europeans and the Chinese, two major foreign influences in early historical Taiwan that might indicate social changes in the indigenous society.
 
-craft specialization is commonly identified through statistical standardization of specific dimensions of a particular item – standardization as evidence of specialization.
+## Geometric Morphometric approach: shape analysis for ceramic specialization
 
-The variation of standardization of specialized product is commonly used as indicator to identify craft specialization based on the assumption that specialized mass production will lead to uniformity of the product due to routinization, increased skills, and fewer number of producers involved in production [@Arnold2000, pp.
-334; @Stark1995; @Costin1991; @Costin2001].
-For ceramic standardization, several measurements have been proposed, including metric, compositional, and technological approaches [@Rice1991; @Blackman1993; @Costin1991; @Tite1999; @Roux2015].
-Among those variables, linear measurements is widely used by calculation coefficient of variation (CV) that lower CV value indicates higher degree of standardization [@Hirshman2010; @Eerkens2001].
-However, traditional typological and linear measurements are limited because they can be insensitive to subtle variations resulting from changes in craft specialization.
-
-Taking a new approach to the measurement of craft specialization, we studied shape and standardization of locally made pottery to identify changes in pottery production at Kiwulan (1400-1900 AD), a large multi-component archaeological site in northeastern Taiwan.
-We apply geometric morphometric methods to study artifact shapes to investigate if there are any changes resulting from foreign contact of European and Chinese that might indicate social changes in the indigenous society.
-
-- social changes -  culture contact framework 
-- here I have the third cultural resistance, identity take a look at social boundary
-- hypothesis, goals
-
-This paper
-
-Culture contact between indigenous people and colonizer with imperial power usually leads to profound changes in local indigenous societies.
-We explore the culture changes in indigenous societies in northeastern Taiwan that experienced foreign contacts with the Europeans and Chinese since 17th century by examining the local pottery production.
-We observed pottery production in northeastern Taiwan in the late Iron Age period and historical period (1400-1900 AD) presenting a high consistency in form and shape compared to other pottery found from other regions on this island that hints the emergence of pottery specialization.
-Our aim is to explore the impact of foreign contacts on local indigenous societies in northeastern Taiwan by examining pottery production as an indicator for social changes.
-Using pottery shape as a proxy to study craft specialization, we answer these questions: Did foreign contacts have impacts on indigenous pottery production that can be detected in the shape of the vessels?
-How does pottery change in shape throughout 600 years at this site?
-Did pottery become more homogeneous and standardized in shape after foreign contacts with the European colonizer or the Chinese immigrant?
-
-Our hypothesis is that if foreign contacts including European and Chinese had impact on the emergence of social inequality in local indigenous society due to the monopolies for trade between a small number of indigenous people, then we expect to see the changes in social organization from a more corporate to a more network organization that would be reflected by pottery production at Kiwulan.
-The emergence of craft specialization, here pottery specialization, are usually related to the presence of elite control.
-In this case, if the competitions for foreign resources and being trade partners of European or Chinese colonizer among individuals gradually lead to the emergence of social inequality, then we expect the local pottery will show more homogeneous features after contact due to the craft specialization caused by control of small group of individuals.
-The first step is to detecting pottery standardization when discussing the emergence of craft specialization and the possible changes in social organization.
-This paper uses Elliptic Fourier Analysis, one outline approach in Geometric morphometric methods, to test whether there was increasing pottery standardization in relation to foreign contacts.
-
-## Geometric Morphometric approach: shape analysis for pottery specialization
-
-Geometric morphometric methods (GMM) has been widely applied recently in archaeology for shape analysis, which explores morphological variability and similarity of archaeological materials to address the questions of anthropological interests.
+Geometric morphometric methods (GMM) has been increasingly applied in archaeology for shape analysis, which explores morphological variability and similarity of archaeological materials to address the questions of anthropological interests.
 Different from linear measurement approaches that capture shapes by measuring length, width or ratios of objects, geometric morphometrics methods use Cartesian coordinates of morphological structures to define the shapes [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007].
@@
 
+## social changes in cross-cultural interaction contexts
+- here I have the third cultural resistance, identity take a look at social boundary
+- hypothesis, goals
+
+Culture contact between indigenous people and colonizer with imperial power usually leads to profound changes in local indigenous societies.
+We explore the culture changes in indigenous societies in northeastern Taiwan that experienced foreign interactions with the Europeans and Chinese since 17th century by examining the local pottery production.
+We observed ceramic production in northeastern Taiwan in the late Iron Age period and historical period (1400-1900 AD) presenting a high consistency in form and shape compared to other pottery found from other regions on this island that hints the emergence of pottery specialization.
+Our aim is to explore the impact of foreign contacts on local indigenous societies in northeastern Taiwan by examining pottery production as an indicator for social changes.
+Using pottery shape as a proxy to study craft specialization, we answer these questions: Did foreign contacts have impacts on indigenous pottery production that can be detected in the shape of the vessels?
+How does pottery change in shape throughout 600 years at this site?
+Did pottery become more homogeneous and standardized in shape after foreign contacts with the European colonizer or the Chinese immigrant?
+
+Our hypothesis is that if foreign contacts including European and Chinese had impact on the emergence of social inequality in local indigenous society due to the monopolies for trade between a small number of indigenous people, then we expect to see the changes in social organization from a more corporate to a more network organization that would be reflected by pottery production at Kiwulan.
+The emergence of craft specialization, here pottery specialization, are usually related to the presence of elite control.
+In this case, if the competitions for foreign resources and being trade partners of European or Chinese colonizer among individuals gradually lead to the emergence of social inequality, then we expect the local pottery will show more homogeneous features after contact due to the craft specialization caused by control of small group of individuals.
+The first step is to detecting pottery standardization when discussing the emergence of craft specialization and the possible changes in social organization.
+This paper uses Elliptic Fourier Analysis, one outline approach in Geometric morphometric methods, to test whether there was increasing pottery standardization in relation to foreign contacts.
+
 In this paper, we use EFA for pottery shape analysis to evaluate the level of standardization of pottery from an Iron Age site in northeastern Taiwan in relation to the European presence in the 17th century that might hint the emergence of craft specialization for pottery production.
@@
 - craft or pottery specialization
-- CV is hard to test, we apply some statistics testing for it
+- CV is hard to test, we apply some statistics testing for it, - innovative: CV test
 

### 2020-01-03 analysis/paper/paper.Rmd
reorganize and add references

@@
 Economic models consider production, exchange, and consumption as a whole and explain that specialization is a subsistence strategy to accommodate uneven distribute of resources [@Costin2001; @Arnold1994], or a reaction to the changes in subsistence practices [@Stark1995economic].
-However, the explanation for the emergence of specialized production emerges might not be exclusive or straightforward.
-Also, specialization is not necessarily accompanied with social differentiation that is observed in lowland northern Philippines with low degree of ceramic specialization during the Spanish colonial period [@Acabado2018].
-This implies that the condition and contexts under which specialists organized are important for interpretation of changes in social organization [@Roux2018].
+However, the explanation for the emergence of specialized production might not be exclusive or straightforward.
+Furthermore, specialization is not necessarily accompanied with social differentiation observed in lowland northern Philippines with a low degree of ceramic specialization in Ifugao societies with social ranking during the Spanish colonization [@Acabado2018].
+This implies that the condition and contexts under which specialists organized are important for interpretation of changes in social organization [@Roux2015].
+
+## social changes in cross-cultural interaction contexts
+- here I have the third cultural resistance, identity take a look at social boundary
+- hypothesis, goals
+- trade [@Alizadeh2018]
+
+Culture contact between indigenous people and colonizer with imperial power usually leads to profound changes in local indigenous societies.
+We explore the culture changes in indigenous societies in northeastern Taiwan that experienced foreign interactions with the Europeans and Chinese since 17th century by examining the local pottery production.
+We observed ceramic production in northeastern Taiwan in the late Iron Age period and historical period (1400-1900 AD) presenting a high consistency in form and shape compared to other pottery found from other regions on this island that hints the emergence of pottery specialization.
+Our aim is to explore the impact of foreign contacts on local indigenous societies in northeastern Taiwan by examining pottery production as an indicator for social changes.
+Using pottery shape as a proxy to study craft specialization, we answer these questions: Did foreign contacts have impacts on indigenous pottery production that can be detected in the shape of the vessels?
+How does pottery change in shape throughout 600 years at this site?
+Did pottery become more homogeneous and standardized in shape after foreign contacts with the European colonizer or the Chinese immigrant?
+
+Our hypothesis is that if foreign contacts including European and Chinese had impact on the emergence of social inequality in local indigenous society due to the monopolies for trade between a small number of indigenous people, then we expect to see the changes in social organization from a more corporate to a more network organization that would be reflected by pottery production at Kiwulan.
+The emergence of craft specialization, here pottery specialization, are usually related to the presence of elite control.
+In this case, if the competitions for foreign resources and being trade partners of European or Chinese colonizer among individuals gradually lead to the emergence of social inequality, then we expect the local pottery will show more homogeneous features after contact due to the craft specialization caused by control of small group of individuals.
+The first step is to detecting pottery standardization when discussing the emergence of craft specialization and the possible changes in social organization.
+This paper uses Elliptic Fourier Analysis, one outline approach in Geometric morphometric methods, to test whether there was increasing pottery standardization in relation to foreign contacts.
+
+## Measuring ceramic standardization
 
@@
 For ceramic standardization, several measurements have been proposed, such as metric, compositional, and technological approaches [@Arnold2000; @Rice1991; @Blackman1993; @Costin1991; @Tite1999; @Roux2015].
-Among those variables, metric measurement is widely applied in archaeological assemblages by calculating coefficient of variation between sets of data coupled with statistical comparison that lower value reflects higher degree of standardization [@Eerkens2001; @Roux2003].
+Among those variables, metric measurement is widely applied in archaeological assemblages by calculating coefficient of variation between sets of data coupled with statistical comparison that lower value reflects higher degree of standardization [@Eerkens2001; @Roux2003; @Roux2018; @Stark1995].
 However, this traditional typological and linear measurements might be limited because they can be insensitive to subtle variations resulting from changes in assembles.
 Shape analysis based on geometric morphometrics method are new approaches for measuring the degree of standardization since it takes the overall shape as variable for comparison [@Slice2007].
-Taking geometric morphomerics approaches to the measurement of craft specialization, we study shape and standardization of locally made ceramic to identify changes in ceramic production in cross-cultural interaction contexts at Kiwulan (1350-1950 AD), a large multi-component archaeological site in northeastern Taiwan.
+Taking geometric morphomerics approaches to the measurement of craft specialization, we study shape and standardization of locally made ceramic at Kiwulan (1350-1950 AD), a large multi-component archaeological site in northeastern Taiwan, to identify changes in ceramic production in cross-cultural interaction contexts.
 The aim of this paper is to investigate if there are increasing ceracmic specialization resulting from interaction with the Europeans and the Chinese, two major foreign influences in early historical Taiwan that might indicate social changes in the indigenous society.
@@
 In addition, the results of shape variation by EFA helps to understand the level of production standardization over time across that region.
-
-## social changes in cross-cultural interaction contexts
-- here I have the third cultural resistance, identity take a look at social boundary
-- hypothesis, goals
-
-Culture contact between indigenous people and colonizer with imperial power usually leads to profound changes in local indigenous societies.
-We explore the culture changes in indigenous societies in northeastern Taiwan that experienced foreign interactions with the Europeans and Chinese since 17th century by examining the local pottery production.
-We observed ceramic production in northeastern Taiwan in the late Iron Age period and historical period (1400-1900 AD) presenting a high consistency in form and shape compared to other pottery found from other regions on this island that hints the emergence of pottery specialization.
-Our aim is to explore the impact of foreign contacts on local indigenous societies in northeastern Taiwan by examining pottery production as an indicator for social changes.
-Using pottery shape as a proxy to study craft specialization, we answer these questions: Did foreign contacts have impacts on indigenous pottery production that can be detected in the shape of the vessels?
-How does pottery change in shape throughout 600 years at this site?
-Did pottery become more homogeneous and standardized in shape after foreign contacts with the European colonizer or the Chinese immigrant?
-
-Our hypothesis is that if foreign contacts including European and Chinese had impact on the emergence of social inequality in local indigenous society due to the monopolies for trade between a small number of indigenous people, then we expect to see the changes in social organization from a more corporate to a more network organization that would be reflected by pottery production at Kiwulan.
-The emergence of craft specialization, here pottery specialization, are usually related to the presence of elite control.
-In this case, if the competitions for foreign resources and being trade partners of European or Chinese colonizer among individuals gradually lead to the emergence of social inequality, then we expect the local pottery will show more homogeneous features after contact due to the craft specialization caused by control of small group of individuals.
-The first step is to detecting pottery standardization when discussing the emergence of craft specialization and the possible changes in social organization.
-This paper uses Elliptic Fourier Analysis, one outline approach in Geometric morphometric methods, to test whether there was increasing pottery standardization in relation to foreign contacts.
 
@@
 Each traced half cross-section image was duplicated, flipped, and then joined with another one to create a 2D closed outline.
-
 Geometric morphometric analyses were conducted in the R software (www.rproject.org, Core-Team, 2015) using the functions included in the Momocs, a R package intended to quantify the shape and compare its variation, especially for outline analysis [@Bonhomme2014].
@@
 
+- identity, social boundary
+
 The results confirm that the differences in pottery shapes can be detected using outline analysis approaches.

### 2020-01-03 analysis/paper/paper.Rmd
add sentences to introduction

@@
 
-## social changes in cross-cultural interaction contexts
-- here I have the third cultural resistance, identity take a look at social boundary
-- hypothesis, goals
-- trade [@Alizadeh2018]
+The emergence of social inequality in the context of cross-cultural interaction especially in a colonial situation was observed in many parts of world when the introduction of foreign trade goods to local indigenous societies.
+The monopolization for long-distance trade goods might cause substantial transformations of indigenous economic, cultural, and socio-political systems [@Dietler2005; @Dietler1997; @Junker1993; @Silliman2005].
+The archaeological evidence in Northeastern Taiwan show some extent of social differentiation when encountered by the Spanish and the Dutch in the 17th century based on the distribution pattern of the trade ornaments that were viewed as prestige goods in local historical context.
+In addition to the differential distribution of trade ornaments in domestic area and burial contexts, the locally made ceramic present a high consistency in form and shape compared to other pottery found from other regions on this island that hints the possibility of pottery specialization [@Chen2007].
+To further understand the relationship among the use of prestige goods, the degree of ceramic specialization and the arrival of foreign colonizers, this paper uses pottery production as an example for identifying social changes in a cross-cultural contact context.
 
-Culture contact between indigenous people and colonizer with imperial power usually leads to profound changes in local indigenous societies.
-We explore the culture changes in indigenous societies in northeastern Taiwan that experienced foreign interactions with the Europeans and Chinese since 17th century by examining the local pottery production.
-We observed ceramic production in northeastern Taiwan in the late Iron Age period and historical period (1400-1900 AD) presenting a high consistency in form and shape compared to other pottery found from other regions on this island that hints the emergence of pottery specialization.
-Our aim is to explore the impact of foreign contacts on local indigenous societies in northeastern Taiwan by examining pottery production as an indicator for social changes.
-Using pottery shape as a proxy to study craft specialization, we answer these questions: Did foreign contacts have impacts on indigenous pottery production that can be detected in the shape of the vessels?
-How does pottery change in shape throughout 600 years at this site?
-Did pottery become more homogeneous and standardized in shape after foreign contacts with the European colonizer or the Chinese immigrant?
+<---[@Acabado2018] In order to maintain status, it was important for elites to monopolize foreign trade and control the manufacture of earthenware pots connected to it through centralized and specialized production.
+--->
 
@@
 
+<--Culture contact between indigenous people and colonizer with imperial power usually leads to profound changes in local indigenous societies.
+We explore the culture changes in indigenous societies in northeastern Taiwan that experienced foreign interactions with the Europeans and Chinese since 17th century by examining the local pottery production.
+We observed ceramic production in northeastern Taiwan in the late Iron Age period and historical period (1400-1900 AD) presenting a high consistency in form and shape compared to other pottery found from other regions on this island that hints the emergence of pottery specialization.
+Our aim is to explore the impact of foreign contacts on local indigenous societies in northeastern Taiwan by examining pottery production as an indicator for social changes.
+Using pottery shape as a proxy to study craft specialization, we answer these questions: Did foreign contacts have impacts on indigenous pottery production that can be detected in the shape of the vessels?
+How does pottery change in shape throughout 600 years at this site?
+Did pottery become more homogeneous and standardized in shape after foreign contacts with the European colonizer or the Chinese immigrant?
+-->
+
 ## Measuring ceramic standardization
+- intentional and mechanical
 
@@
 
-- identity, social boundary
+## social changes in cross-cultural interaction contexts
+- cultural resistance
+- trade [@Alizadeh2018]
+- identity, social boundary, Raux2015, Boness2017
 

### 2020-01-04 analysis/paper/paper.Rmd
updates for introduction and add references

--- 2020-01-03 analysis/paper/paper.Rmd
@@
 Furthermore, specialization is not necessarily accompanied with social differentiation observed in lowland northern Philippines with a low degree of ceramic specialization in Ifugao societies with social ranking during the Spanish colonization [@Acabado2018].
-This implies that the condition and contexts under which specialists organized are important for interpretation of changes in social organization [@Roux2015].
+This implies that the condition and contexts under which specialists organized are important for interpretation of changes in social organization [@Roux2015; @Costin2001].
 
-## social changes in cross-cultural interaction contexts
-- here I have the third cultural resistance, identity take a look at social boundary
-- hypothesis, goals
-- trade [@Alizadeh2018]
+The emergence of social inequality in the context of cross-cultural interaction especially in a colonial situation was observed in many parts of world when the introduction of foreign trade goods to local indigenous societies.
+The monopolization of long-distance trade goods might cause substantial transformations of indigenous economic, cultural, and socio-political systems [@Dietler2005; @Dietler1997; @Junker1993; @Silliman2005].
+Recently more studies discuss the indirect effects of colonialism or pericolonial archaeology that investigates areas where European colonial rule was limited or their conquests were unsuccessful, but their colonial activities had economically and politically impacts on indigenous peoples in the periphery of colonial control [@Trabert2017; @Acabado2017].
+For example, @Acabado2017's study on Ifugao society in the highland Philippines suggests economic and political intensification during the Spanish presence that indicates the responses of indigenous peoples to Spanish cooptation.
+The archaeological evidence in northeastern Taiwan show evidence of use of prestige goods such as trade ornaments ornaments and their uneven distribution pattern in domestic area and burial contexts indicates some extent of social inequality when encountered by the Spanish and the Dutch in the 17th century.
+In addition to the differential distribution of trade ornaments, the locally made ceramic present a high consistency in form and shape compared to other pottery found from other regions on this island that hints the possibility of pottery specialization.
+To further understand the relationship among the use of prestige goods, the degree of ceramic specialization, and the influence of foreign colonizers, this paper uses pottery production as a proxy for identifying social changes in a cross-cultural contact context.
 
-Culture contact between indigenous people and colonizer with imperial power usually leads to profound changes in local indigenous societies.
-We explore the culture changes in indigenous societies in northeastern Taiwan that experienced foreign interactions with the Europeans and Chinese since 17th century by examining the local pottery production.
-We observed ceramic production in northeastern Taiwan in the late Iron Age period and historical period (1400-1900 AD) presenting a high consistency in form and shape compared to other pottery found from other regions on this island that hints the emergence of pottery specialization.
-Our aim is to explore the impact of foreign contacts on local indigenous societies in northeastern Taiwan by examining pottery production as an indicator for social changes.
-Using pottery shape as a proxy to study craft specialization, we answer these questions: Did foreign contacts have impacts on indigenous pottery production that can be detected in the shape of the vessels?
-How does pottery change in shape throughout 600 years at this site?
-Did pottery become more homogeneous and standardized in shape after foreign contacts with the European colonizer or the Chinese immigrant?
-
-Our hypothesis is that if foreign contacts including European and Chinese had impact on the emergence of social inequality in local indigenous society due to the monopolies for trade between a small number of indigenous people, then we expect to see the changes in social organization from a more corporate to a more network organization that would be reflected by pottery production at Kiwulan.
-The emergence of craft specialization, here pottery specialization, are usually related to the presence of elite control.
-In this case, if the competitions for foreign resources and being trade partners of European or Chinese colonizer among individuals gradually lead to the emergence of social inequality, then we expect the local pottery will show more homogeneous features after contact due to the craft specialization caused by control of small group of individuals.
-The first step is to detecting pottery standardization when discussing the emergence of craft specialization and the possible changes in social organization.
-This paper uses Elliptic Fourier Analysis, one outline approach in Geometric morphometric methods, to test whether there was increasing pottery standardization in relation to foreign contacts.
-
-## Measuring ceramic standardization
-
-The variation of standardization of specialized product is commonly used as indicator to identify craft specialization based on the assumption that specialized mass production will lead to uniformity of the product due to routinization, increased skills, and fewer number of producers involved in production [@Stark1995; @Costin1991; @Costin2001].
+To identify craft specialization, the variation of standardization of specialized product is commonly used as indicator based on the assumption that specialized mass production will lead to uniformity of the product due to routinization, increased skills, and fewer number of producers involved in production [@Stark1995; @Costin1991; @Costin2001].
 For ceramic standardization, several measurements have been proposed, such as metric, compositional, and technological approaches [@Arnold2000; @Rice1991; @Blackman1993; @Costin1991; @Tite1999; @Roux2015].
@@
 Shape analysis based on geometric morphometrics method are new approaches for measuring the degree of standardization since it takes the overall shape as variable for comparison [@Slice2007].
-Taking geometric morphomerics approaches to the measurement of craft specialization, we study shape and standardization of locally made ceramic at Kiwulan (1350-1950 AD), a large multi-component archaeological site in northeastern Taiwan, to identify changes in ceramic production in cross-cultural interaction contexts.
-The aim of this paper is to investigate if there are increasing ceracmic specialization resulting from interaction with the Europeans and the Chinese, two major foreign influences in early historical Taiwan that might indicate social changes in the indigenous society.
+Taking geometric morphomerics approaches to the measurement of craft specialization, we study shape and standardization of locally made ceramic at Kiwulan (1350-1950 AD) [@Chen2007], a large multi-component archaeological site in northeastern Taiwan, to identify changes in ceramic production in cross-cultural interaction contexts.
+The aim of this paper is to investigate if there are increasing ceramic specialization resulting from interaction with the Europeans and the Chinese, two major foreign influences in early historical Taiwan that might indicate social changes in the indigenous society.
+
+Our hypothesis is that if foreign contacts including European in the 17th century and Chinese in the 19th century had impacts on the emergence of social inequality in local indigenous society due to the monopolies for trade between a small number of indigenous people, then we expect to see the changes in social organization from a more corporate to a more network organization that would be reflected by pottery production at Kiwulan.
+The emergence of craft specialization, here pottery specialization, are usually related to the presence of elite control.
+In this case, if the competitions for foreign resources and being trade partners of European or Chinese colonizer among individuals gradually lead to the emergence of social inequality, then we expect the local pottery will show more homogeneous features after contact due to the craft specialization caused by control of small group of individuals.
+The aim of this paper is to explore the impact of foreign contacts on local indigenous societies in northeastern Taiwan by examining pottery production as an indicator for social changes.
+Using pottery shape as a proxy to study craft specialization, we answer these questions: Did foreign contacts have impacts on indigenous pottery production that can be detected in the shape of the vessels?
+How does pottery change in shape throughout 500 years at this site?
+Did pottery become more homogeneous and standardized in shape after foreign contacts with the European colonizer or the Chinese immigrant?
 
@@
 The application of EFA in archaeology is mostly used for bone morphology to understand the biological variation and recently more for stone artifacts to explore stone tool technology [@Fox2015; @Hoggard2019; @Iovictua2010].
+
 Outline approach based on EFA is suitable for shapes lacking obvious landmarks or curves convey more important information indicating the meaningful variations, for example, ceramics.
@@
 In addition, the results of shape variation by EFA helps to understand the level of production standardization over time across that region.
+<--Topi2007-->
 
@@
 
-# Materials
-
-## Archaeological background
+# Materials and archaeological background
 
@@
 It shows there is differences in pottery shapes between pre-European and post-European, and Pre-European and Chinese contact.
-A increasing standardization of shape of pottery during the Chinese contact period compared to the other two periods is observed.
+An increasing standardization of shape of pottery during the Chinese contact period compared to the other two periods is observed.
 This study provide an comparable basis for our understanding of the standardized pottery as an indicator of social changes in a cultural contact conditions.
@@
 This may suggest that an emergence of specialized groups or a consensus of the adoption of similar shape of pottery in the society.
-The changes in the standardization of pottery may indicate a major social changes in the local indigenous organization.
 When it comes to ceramic specialization, metric measurements and clay paste are two major indicators.
@@
 We observe the increasing standardization of the pottery shape, however, it is interesting to note that there is no obvious changes in the clay paste across different phases.
+In the vitiation that the shape become more standardized without changes in raw materials,
 

### 2020-01-04 analysis/paper/paper.Rmd
move a paragraph to scrape file

@@
 Taking geometric morphomerics approaches to the measurement of craft specialization, we study shape and standardization of locally made ceramic at Kiwulan (1350-1950 AD) [@Chen2007], a large multi-component archaeological site in northeastern Taiwan, to identify changes in ceramic production in cross-cultural interaction contexts.
-The aim of this paper is to investigate if there are increasing ceramic specialization resulting from interaction with the Europeans and the Chinese, two major foreign influences in early historical Taiwan that might indicate social changes in the indigenous society.
+The aim of this paper is to investigate if there are increasing ceramic specialization resulting from interaction with the Europeans and the Chinese, two major foreign influences in early historical Taiwan, that might indicate social changes in the indigenous society.
 
-Our hypothesis is that if foreign contacts including European in the 17th century and Chinese in the 19th century had impacts on the emergence of social inequality in local indigenous society due to the monopolies for trade between a small number of indigenous people, then we expect to see the changes in social organization from a more corporate to a more network organization that would be reflected by pottery production at Kiwulan.
-The emergence of craft specialization, here pottery specialization, are usually related to the presence of elite control.
-In this case, if the competitions for foreign resources and being trade partners of European or Chinese colonizer among individuals gradually lead to the emergence of social inequality, then we expect the local pottery will show more homogeneous features after contact due to the craft specialization caused by control of small group of individuals.
-The aim of this paper is to explore the impact of foreign contacts on local indigenous societies in northeastern Taiwan by examining pottery production as an indicator for social changes.
+Our hypothesis is that if foreign contacts including European in the 17th century and Chinese in the 19th century had impacts on the emergence of social inequality in local indigenous society due to the monopolies for trade between a small number of indigenous people, then we expect to see the changes in social organization from a more corporate to a more network organization that would be reflected by pottery production at Kiwulan [@Feinman2000].
+The emergence of craft specialization, here pottery specialization, are usually related to changes in social organization towards a society with increasing status inequality.
+In this case, if the competitions for foreign resources and being trade partners of European or Chinese colonizer among individuals gradually lead to the emergence of social inequality, then we expect the local ceramic will show more homogeneous features after contact due to the craft specialization caused by control of small group of individuals.
 Using pottery shape as a proxy to study craft specialization, we answer these questions: Did foreign contacts have impacts on indigenous pottery production that can be detected in the shape of the vessels?
@@
 Did pottery become more homogeneous and standardized in shape after foreign contacts with the European colonizer or the Chinese immigrant?
+Those question are important to understand the indirect influences of the European colonists on the local indigenous societies.
+Also, it would help to understand the relationship between craft specialization and increasing social inequality in a pericolonial context [@Acabado2017].
 
-## Geometric Morphometric approach: shape analysis for ceramic specialization
+## Geometric Morphometric approaches
 
@@
 In this paper, we use EFA for pottery shape analysis to evaluate the level of standardization of pottery from an Iron Age site in northeastern Taiwan in relation to the European presence in the 17th century that might hint the emergence of craft specialization for pottery production.
-Craft specialization can provide further evidence about the type of production organization which is closely related to economic or political aspects of society [@Arnold2000; @Blackman1993; @Costin1991; @Hirshman2010; @Stark1995].
-The emergence of craft specialization can also be an indicator of social inequality that hints the possibility of the presence of production groups worked for specific individuals.
-The presence of craft specialization is commonly studied by measuring the standardization of ceramics based on the assumption that the specialized mass production will lead to uniformity or homogeneity of the product due to increased skills, routinization, and lower diversity of producers [@Arnold2000].
-To test the prediction of a shift in the level of social inequality in this case study in northeastern Taiwan influenced by the colonial European presence, we use standardization of pottery in shapes as an index for craft specialization to identify the presence of a few individuals controlling large scale pottery production [@Costin1995; @Hirshman2010] using outline-based geometric morphometrics couples with traditional metric measurements to compare and evaluate the use of outline approaches for pottery with high consistency in shapes.
 
-- craft or pottery specialization
 - CV is hard to test, we apply some statistics testing for it, - innovative: CV test

### 2020-01-04 analysis/paper/scraps.Rmd
move a paragraph to scrape file

# Hypotheses and expectations

My hypothesis is that if foreign contacts including European and Chinese had impact on the emergence of social inequality in local indigenous society due to the monopolies for trade between a small number of indigenous people, then I expect to see the changes in social organization from a more corporate to a more network organization that can be reflected by pottery production at Kiwulan. The emergence of craft specialization, here pottery specialization, are usually related to the presence of elite control. In this case, if the competitions for foreign resources and being trade partners of European or Chinese colonizer among individuals gradually lead to the emergence of social inequality, then I expect the local pottery will show more homogeneous features after contact due to the pottery specialization caused by control of small group of individuals.

# Part of Introdution

In this paper, we use EFA for pottery shape analysis to evaluate the level of standardization of pottery from an Iron Age site in northeastern Taiwan in relation to the European presence in the 17th century that might hint the emergence of craft specialization for pottery production. Craft specialization can provide further evidence about the type of production organization which is closely related to economic or political aspects of society [@Arnold2000; @Blackman1993; @Costin1991; @Hirshman2010; @Stark1995]. The emergence of craft specialization can also be an indicator of social inequality that hints the possibility of the presence of production groups worked for specific individuals. The presence of craft specialization is commonly studied by measuring the standardization of ceramics based on the assumption that the specialized mass production will lead to uniformity or homogeneity of the product due to increased skills, routinization, and lower diversity of producers [@Arnold2000]. To test the prediction of a shift in the level of social inequality in this case study in northeastern Taiwan influenced by the colonial European presence, we use standardization of pottery in shapes as an index for craft specialization to identify the presence of a few individuals controlling large scale pottery production [@Costin1995; @Hirshman2010] using outline-based geometric morphometrics couples with traditional metric measurements to compare and evaluate the use of outline approaches for pottery with high consistency in shapes.

### 2020-01-04 analysis/paper/scrap.R

@@ -1,3 +1,3 @@
 # Hypotheses and expectations
 
-# My hypothesis is that if foreign contacts including European and Chinese had impact on the emergence of social inequality in local indigenous society due to the monopolies for trade between a small number of indigenous people, then I expect to see the changes in social organization from a more corporate to a more network organization that can be reflected by pottery production at Kiwulan. The emergence of craft specialization, here pottery specialization, are usually related to the presence of elite control. In this case, if the competitions for foreign resources and being trade partners of European or Chinese colonizer among individuals gradually lead to the emergence of social inequality, then I expect the local pottery will show more homogeneous features after contact due to the pottery specialization caused by control of small group of individuals.
+My hypothesis is that if foreign contacts including European and Chinese had impact on the emergence of social inequality in local indigenous society due to the monopolies for trade between a small number of indigenous people, then I expect to see the changes in social organization from a more corporate to a more network organization that can be reflected by pottery production at Kiwulan. The emergence of craft specialization, here pottery specialization, are usually related to the presence of elite control. In this case, if the competitions for foreign resources and being trade partners of European or Chinese colonizer among individuals gradually lead to the emergence of social inequality, then I expect the local pottery will show more homogeneous features after contact due to the pottery specialization caused by control of small group of individuals.

### 2020-01-05 analysis/paper/paper.Rmd
work on GMM and finish the first part of introduction

@@
 The emergence of social inequality in the context of cross-cultural interaction especially in a colonial situation was observed in many parts of world when the introduction of foreign trade goods to local indigenous societies.
-The monopolization of long-distance trade goods might cause substantial transformations of indigenous economic, cultural, and socio-political systems [@Dietler2005; @Dietler1997; @Junker1993; @Silliman2005].
+The monopolization of long-distance trade goods usaully cause substantial transformations of indigenous economic, cultural, and socio-political systems [@Dietler2005; @Dietler1997; @Junker1993; @Silliman2005].
 Recently more studies discuss the indirect effects of colonialism or pericolonial archaeology that investigates areas where European colonial rule was limited or their conquests were unsuccessful, but their colonial activities had economically and politically impacts on indigenous peoples in the periphery of colonial control [@Trabert2017; @Acabado2017].
-For example, @Acabado2017's study on Ifugao society in the highland Philippines suggests economic and political intensification during the Spanish presence that indicates the responses of indigenous peoples to Spanish cooptation.
-The archaeological evidence in northeastern Taiwan show evidence of use of prestige goods such as trade ornaments ornaments and their uneven distribution pattern in domestic area and burial contexts indicates some extent of social inequality when encountered by the Spanish and the Dutch in the 17th century.
-In addition to the differential distribution of trade ornaments, the locally made ceramic present a high consistency in form and shape compared to other pottery found from other regions on this island that hints the possibility of pottery specialization.
-To further understand the relationship among the use of prestige goods, the degree of ceramic specialization, and the influence of foreign colonizers, this paper uses pottery production as a proxy for identifying social changes in a cross-cultural contact context.
+For example, @Acabado2017's study on Ifugao society in the highland Philippines suggests economic and political intensification during the Spanish presence as the response of indigenous peoples to the Spanish cooptation.
+The archaeological evidence in northeastern Taiwan show evidence of increasing use of prestige goods such as trade ornaments and uneven distribution pattern in domestic area and burial contexts when the society encountered by the Spanish and the Dutch in the 17th century.
+In addition to the differential distribution of trade ornaments, the locally made ceramic present a high consistency in form and shape compared to other pottery throughout northeastern Taiwan that hints the possibility of pottery specialization.
 
@@
 For ceramic standardization, several measurements have been proposed, such as metric, compositional, and technological approaches [@Arnold2000; @Rice1991; @Blackman1993; @Costin1991; @Tite1999; @Roux2015].
-Among those variables, metric measurement is widely applied in archaeological assemblages by calculating coefficient of variation between sets of data coupled with statistical comparison that lower value reflects higher degree of standardization [@Eerkens2001; @Roux2003; @Roux2018; @Stark1995].
-However, this traditional typological and linear measurements might be limited because they can be insensitive to subtle variations resulting from changes in assembles.
+Among those variables, metric measurement is widely applied in archaeological assemblages by calculating coefficient of variation for sets of data coupled with statistical comparison that lower value reflects higher degree of standardization [@Eerkens2001; @Roux2003; @Roux2018; @Stark1995].
+However, this traditional typological and linear measurements might be limited because they can be insensitive to subtle variations resulting from changes in assemblages.
 Shape analysis based on geometric morphometrics method are new approaches for measuring the degree of standardization since it takes the overall shape as variable for comparison [@Slice2007].
-Taking geometric morphomerics approaches to the measurement of craft specialization, we study shape and standardization of locally made ceramic at Kiwulan (1350-1950 AD) [@Chen2007], a large multi-component archaeological site in northeastern Taiwan, to identify changes in ceramic production in cross-cultural interaction contexts.
-The aim of this paper is to investigate if there are increasing ceramic specialization resulting from interaction with the Europeans and the Chinese, two major foreign influences in early historical Taiwan, that might indicate social changes in the indigenous society.
+Using geometric morphomerics approaches, we study shape and standardization of locally made ceramics at Kiwulan (1350-1950 AD) [@Chen2007], a large multi-component archaeological site in northeastern Taiwan, to identify changes in ceramic production in cross-cultural interaction contexts.
+The aim of this paper is to investigate if there are increasing ceramic specialization resulting from interaction with the Europeans in the 17th century or the Chinese in the 19th century, two major foreign influences in early historical Taiwan, that might indicate social changes in the indigenous society.
 
-Our hypothesis is that if foreign contacts including European in the 17th century and Chinese in the 19th century had impacts on the emergence of social inequality in local indigenous society due to the monopolies for trade between a small number of indigenous people, then we expect to see the changes in social organization from a more corporate to a more network organization that would be reflected by pottery production at Kiwulan [@Feinman2000].
+Our hypothesis is that if foreign influences had impacts on the emergence of social inequality in local indigenous society due to the monopolies for trade between a small number of indigenous people, then we expect to see the changes in social organization from a more corporate to a more network organization that would be reflected by pottery production at Kiwulan [@Feinman2000].
 The emergence of craft specialization, here pottery specialization, are usually related to changes in social organization towards a society with increasing status inequality.
-In this case, if the competitions for foreign resources and being trade partners of European or Chinese colonizer among individuals gradually lead to the emergence of social inequality, then we expect the local ceramic will show more homogeneous features after contact due to the craft specialization caused by control of small group of individuals.
+In this case, if the competitions for foreign resources and being trade partners of European or Chinese colonizer among individuals gradually lead to increasing social inequality, then we expect the local ceramic will show more homogeneous features after contact due to the craft specialization caused by control of small group of individuals.
 Using pottery shape as a proxy to study craft specialization, we answer these questions: Did foreign contacts have impacts on indigenous pottery production that can be detected in the shape of the vessels?
@@
 Did pottery become more homogeneous and standardized in shape after foreign contacts with the European colonizer or the Chinese immigrant?
-Those question are important to understand the indirect influences of the European colonists on the local indigenous societies.
-Also, it would help to understand the relationship between craft specialization and increasing social inequality in a pericolonial context [@Acabado2017].
+Those question are important to understand the indirect influences of the European colonists on the local indigenous societies, which remains unknown in East Asia where the colonial power was not successful compared to other places in Southeast Asia.
+Also, it would help to understand the relationship among the use of prestige goods, the degree of ceramic specialization, and the influence of foreign colonizers in a pericolonial context [@Acabado2017].
 
@@
 Geometric morphometric methods (GMM) has been increasingly applied in archaeology for shape analysis, which explores morphological variability and similarity of archaeological materials to address the questions of anthropological interests.
-Different from linear measurement approaches that capture shapes by measuring length, width or ratios of objects, geometric morphometrics methods use Cartesian coordinates of morphological structures to define the shapes [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007].
-Landmarks, curves or outlines of objects can be represented by coordinates by their unique point location with respect to numerical values of coordinate axes.
-According to different focuses and nature of original shape data, there are two common morphometric methods used to capture and analyze the shapes, landmark approaches and outline approaches [@Adams2004].
+Different from linear measurement approaches that capture shapes by measuring length, width or ratios of objects, geometric morphometrics methods use Cartesian coordinates of morphological structures to define shapes [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007].
+Landmarks, curves or outlines of objects can be represented by coordinates in terms of their unique point locations with respect to numerical values of coordinate axes.
+According to different focuses and nature of original shape data, there are two common morphometric methods, landmark approaches and outline approaches [@Adams2004].
 Landmark approaches assign a set of landmarks or semilandmarks onto objects as reference points that can be specified on a coordinate system as x, y coordinates for two dimensions, or x, y, z coordinates for three dimensions.
 Based on distances between raw landmarks, objects are translated to a common centroid, rescaled into the same size, and rotated until the summed squared distances between corresponding landmarks are minimized using generalized Procrustes analysis (GPA) [@Bookstein1991].
-GPA is a method that superimposes sets of landmark configurations into a common coordinate system, where superimposed landmark coordinates are used as shape variables for further multivariate statistics analyses[@Slice2007].
-Landmark-based morphometrics is widely used for archaeological objects where a morphological signature is obvious to identity which provides ideal reference points for landmark placement, such as projectile point tips or biologically measurements points indicating osteological features [@Birch2019; @Buchanan2019; @Lycett2013; @Cardillo2010; @Haruda2019; @Meloro2015].
+GPA is a method that superimposes sets of landmark configurations into a common coordinate system, where superimposed landmark coordinates are used as shape variables for further multivariate statistics analyses [@Slice2007].
+Landmark-based morphometrics are widely applied to archaeological objects with a obvious morphological signature that can provide ideal reference points for landmark placement, such as projectile point tips or biologically measurements points indicating osteological features [@Birch2019; @Buchanan2019; @Lycett2013; @Cardillo2010; @Haruda2019; @Meloro2015].
 
-However, one limitation of landmark-based morphometrics is that landmark data may not be able to capture the shape differences of a morphological structure where the curving outlines between landmarks convey important information.
-In addition, obvious and reliable landmark points may be unavailable for complex or rounded shaped structures.
-In those cases, outline approaches serve as a solution by analyzing the overall shape of a morphological structure.
+However, one limitation of landmark approaches is that landmark data may not be able to capture the shape differences of a morphological structure where the curving outlines between landmarks are crucial for variation.
+Furthermore, representative and reliable landmark points may be unavailable for complex or rounded shaped structures.
+In those cases, outline approaches are better way for analyzing the overall shape of an object.
 One of the outline approaches is the semi-landmarks method, also called sliding landmarks, which assigns points along the curve between two landmarks at defined intervals [@Bookstein1997; @Lawing2010].
 Those semi-landmarks are allowed to slide along the curve to remove the effect of the arbitrary landmark spacing by minimizing either Procrustes distance or bending energy [@Bookstein1997; @Slice2007; @Gunz2013].
-Another approach is elliptic Fourier Analysis (EFA) commonly used for two-dimensional closed shape which turns coordinates into coefficients [@Kuhl1982].
+Another approach is elliptic Fourier Analysis (EFA) that turns coordinates into coefficients using mathematical function, commonly applied to two-dimensional closed shape [@Kuhl1982].
 EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions, called harmonic coefficients [@Adams2004; @Claude2008; @Bonhomme2014].
@@
 The harmonic power, a cumulated sum of squared harmonic coefficient, provides a robust rule for the desired number of harmonics [@Bonhomme2014].
-The application of EFA in archaeology is mostly used for bone morphology to understand the biological variation and recently more for stone artifacts to explore stone tool technology [@Fox2015; @Hoggard2019; @Iovictua2010].
 
-Outline approach based on EFA is suitable for shapes lacking obvious landmarks or curves convey more important information indicating the meaningful variations, for example, ceramics.
-@Wilczek2014 evaluate the concordance between outline-based approaches, the EFA and the Discrete Cosine Transform (DCT), and traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France.
-They found that the variation in ceramic shapes analyzed by EFA and DCT matches traditional ceramic typology, which supports that outline-based approaches can be efficiently used for studying variations in ceramic shapes.
-In addition, the results of shape variation by EFA helps to understand the level of production standardization over time across that region.
-<--Topi2007-->
+Outline approaches especially EFA is suitable for shapes lacking representative landmarks or curves convey more meaningful variations.
+The applications in archaeology are seen on human remains or zooarchaeology to understand the biological variation and recently more on stone artifacts to explore stone tool technology [@Fox2015; @Hoggard2019; @Iovictua2010].
+Ceramics is another archaeological materials that outline approaches can provide new insights into shape variation to answer anthropological questions related to production.
+@Wilczek2014 evaluate the concordance between outline-based approaches, EFA and Discrete Cosine Transform (DCT), and traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France.
+The results show the variation indicated by EFA and DCT matches traditional ceramic typology, which supports that outline-based approaches can be efficiently used for studying variations in ceramic shapes.
+Furthermore, the EFA results help to understand the level of production standardization over time across the region.
+Similarly, @Topi2018 examine 89 photographs of globular jars from the Casas Grandes of northwest Mexico using semi-landmarks method to explore the relationship between social complexity and the emergence of craft specialization.
+Their results suggest that some ceramic types from Medio period, AD 1200 to 1450, were made by specialists.
+Among those high standardized jars, some were made by attached specialists, while others were made by independent specialists according to distribution patterns of jars.
+It shows that outline approaches can distinguish variation in ceramic shapes in a finer resolution to understand deeply the issue of craft specialisation.
 
-In this paper, we use EFA for pottery shape analysis to evaluate the level of standardization of pottery from an Iron Age site in northeastern Taiwan in relation to the European presence in the 17th century that might hint the emergence of craft specialization for pottery production.
+In this paper, we use EFA for ceramic shape analysis to evaluate the level of standardization of pottery from an Iron Age site in northeastern Taiwan in relation to the European presence in the 17th century that might hint the emergence of craft specialization for pottery production.
 

### 2020-01-05 analysis/paper/paper.Rmd
updates for intro and complete the GMM section

@@
 In this case, if the competitions for foreign resources and being trade partners of European or Chinese colonizer among individuals gradually lead to increasing social inequality, then we expect the local ceramic will show more homogeneous features after contact due to the craft specialization caused by control of small group of individuals.
-Using pottery shape as a proxy to study craft specialization, we answer these questions: Did foreign contacts have impacts on indigenous pottery production that can be detected in the shape of the vessels?
-How does pottery change in shape throughout 500 years at this site?
-Did pottery become more homogeneous and standardized in shape after foreign contacts with the European colonizer or the Chinese immigrant?
-Those question are important to understand the indirect influences of the European colonists on the local indigenous societies, which remains unknown in East Asia where the colonial power was not successful compared to other places in Southeast Asia.
+This study is important to understand the indirect influences of the European colonists on the local indigenous societies, which remains unclear in East Asia where the colonial power was not successful compared to other places in Southeast Asia.
 Also, it would help to understand the relationship among the use of prestige goods, the degree of ceramic specialization, and the influence of foreign colonizers in a pericolonial context [@Acabado2017].
@@
 
-In this paper, we use EFA for ceramic shape analysis to evaluate the level of standardization of pottery from an Iron Age site in northeastern Taiwan in relation to the European presence in the 17th century that might hint the emergence of craft specialization for pottery production.
-
-- CV is hard to test, we apply some statistics testing for it, - innovative: CV test
+Taking the ceramics data from Kiwulan, northeaster Taiwan, we use EFA to evaluate the level of standardization of pottery in relation to the European presence in the 17th century that might hint the emergence of craft specialization for pottery production.
+In addition, we use significance test for the equality of coefficient of variation of shape variables to compare the vessel shapes from different periods in a statistical way.
+Using pottery shape as a proxy to study craft specialization, we answer these questions: Did foreign influence have impacts on indigenous pottery production that can be detected in the shape of the vessels?
+How does pottery change in shape throughout 500 years from a single site that can represent the development of social organization in northeastern Taiwan?
+Did pottery become more homogeneous and standardized in shape after foreign contacts with the European colonizer or the Chinese immigrants?
+Starting from the historical background of colonial influences and its reflection in the material record, this study explore the relationship between the emergence of social inequality and craft specialization that will extend our understanding of the reactions of indigenous societies in the periphery of colonial control, providing a perspective that might  be applied in other parts of the world in a similar context.
 

### 2020-01-05 analysis/paper/paper.Rmd
updates for intro

@@
 The relationship between craft specialization and social differentiation is broadly discussed, and many cases show that specialized production can be a useful indicator of increasing socio-political complexity [@Hirshman2010; @Junker1999].
-Economic models consider production, exchange, and consumption as a whole and explain that specialization is a subsistence strategy to accommodate uneven distribute of resources [@Costin2001; @Arnold1994], or a reaction to the changes in subsistence practices [@Stark1995economic].
-However, the explanation for the emergence of specialized production might not be exclusive or straightforward.
-Furthermore, specialization is not necessarily accompanied with social differentiation observed in lowland northern Philippines with a low degree of ceramic specialization in Ifugao societies with social ranking during the Spanish colonization [@Acabado2018].
-This implies that the condition and contexts under which specialists organized are important for interpretation of changes in social organization [@Roux2015; @Costin2001].
+Economic models consider production, exchange, and consumption as a whole to explain specialization as a subsistence strategy to accommodate uneven distribute of resources [@Costin2001; @Arnold1994], a reaction to the changes in subsistence practices [@Stark1995economic], or an outcome stimulated by long-distance trade and inter-regional exchange [@Alizadeh2018].
+An present-day investigation of ceramic specialization in the Jodhpur region of India suggests that the adoption of standardized products across cultural boundaries is a reaction to a collapse of previous economic system, indicating the correlation with major socio-economic changes [@Roux2015].
+However, it is important to note that the explanation for the emergence of specialized production might not be exclusive or straightforward.
+Moreover, @Acabado2018's case study in lowland northern Philippines shows that specialization is not necessarily accompanied with social differentiation based on the result of a low degree of ceramic specialization in Ifugao ranked societies during the Spanish colonization.
+The development of specialization could be history dependent, where condition and contexts under which specialists organized are important for interpretation of changes in social organization [@Roux2015; @Costin2001].
 
@@
 For ceramic standardization, several measurements have been proposed, such as metric, compositional, and technological approaches [@Arnold2000; @Rice1991; @Blackman1993; @Costin1991; @Tite1999; @Roux2015].
-Among those variables, metric measurement is widely applied in archaeological assemblages by calculating coefficient of variation for sets of data coupled with statistical comparison that lower value reflects higher degree of standardization [@Eerkens2001; @Roux2003; @Roux2018; @Stark1995].
-However, this traditional typological and linear measurements might be limited because they can be insensitive to subtle variations resulting from changes in assemblages.
+Among those variables, metric measurement is widely applied in archaeological assemblages by calculating coefficient of variation for sets of data coupled with statistical comparison to identify lower value as an indicator of higher degree of standardization [@Eerkens2001; @Roux2003; @Stark1995].
+@Roux2018's ethnographic study in the Jodhpur region of India suggests the number of artisans can be assessed by comparing coefficient of variation and the differences between artisans can be detected at both intra-individual level and inter-individual level in a region.
+Although this traditional typological and linear measurements are useful for identifying differences in archaeological assembles, it might be limited because dimensional measurements can be insensitive to subtle variations resulting from changes in assemblages.
 Shape analysis based on geometric morphometrics method are new approaches for measuring the degree of standardization since it takes the overall shape as variable for comparison [@Slice2007].
@@
 
-## social changes in cross-cultural interaction contexts
+Did foreign influence have impacts on indigenous pottery production that can be detected in the shape of the vessels?
+How does pottery change in shape throughout 500 years from a single site that can represent the development of social organization in northeastern Taiwan?
+Did pottery become more homogeneous and standardized in shape after foreign contacts with the European colonizer or the Chinese immigrants?
+
 - cultural resistance

### 2020-01-06 analysis/paper/paper.Rmd
updates intro, add references, work on discussion

@@
 For example, @Acabado2017's study on Ifugao society in the highland Philippines suggests economic and political intensification during the Spanish presence as the response of indigenous peoples to the Spanish cooptation.
+Indigenous societies might not just passively accept the colonial rules, but instead actively negotiated with the colonist, and accommodate or resist the foreign intrusion through their daily cultural practices, such as consumption patterns of foreign goods [@Dietler2015; @Given2004; @Mullins2011; @Scaramelli2005; @Silliman2001; @TorrenceandClarke2000].
 The archaeological evidence in northeastern Taiwan show evidence of increasing use of prestige goods such as trade ornaments and uneven distribution pattern in domestic area and burial contexts when the society encountered by the Spanish and the Dutch in the 17th century.
@@
 Similarly, @Topi2018 examine 89 photographs of globular jars from the Casas Grandes of northwest Mexico using semi-landmarks method to explore the relationship between social complexity and the emergence of craft specialization.
-Their results suggest that some ceramic types from Medio period, AD 1200 to 1450, were made by specialists.
+Their results suggest that some ceramic types from Medio period (AD 1200-1450), were made by specialists.
 Among those high standardized jars, some were made by attached specialists, while others were made by independent specialists according to distribution patterns of jars.
@@
 
-Taking the ceramics data from Kiwulan, northeaster Taiwan, we use EFA to evaluate the level of standardization of pottery in relation to the European presence in the 17th century that might hint the emergence of craft specialization for pottery production.
+Taking the ceramics data from Kiwulan, northeaster Taiwan, we use EFA to evaluate the level of standardization of ceramics in relation to the European presence in the 17th century that might hint the emergence of ceramic specialization.
 In addition, we use significance test for the equality of coefficient of variation of shape variables to compare the vessel shapes from different periods in a statistical way.
-Using pottery shape as a proxy to study craft specialization, we answer these questions: Did foreign influence have impacts on indigenous pottery production that can be detected in the shape of the vessels?
+Using pottery shape as a proxy to study craft specialization, we answer the questions: Did foreign influence have impacts on indigenous pottery production that can be detected in the shape of the vessels?
 How does pottery change in shape throughout 500 years from a single site that can represent the development of social organization in northeastern Taiwan?
 Did pottery become more homogeneous and standardized in shape after foreign contacts with the European colonizer or the Chinese immigrants?
-Starting from the historical background of colonial influences and its reflection in the material record, this study explore the relationship between the emergence of social inequality and craft specialization that will extend our understanding of the reactions of indigenous societies in the periphery of colonial control, providing a perspective that might  be applied in other parts of the world in a similar context.
+Starting from the historical background of colonial influences and its reflection in the material record, this study explore the relationship between the emergence of social inequality and craft specialization that can extend our understanding of the reactions of indigenous societies in the periphery of colonial control, providing a perspective that could be applied in other parts of the world in a similar context.
 
@@
 
-Ceramics analyzed in this paper come from 40 units (4m by 4m) sampled from 4 connected trenches at Kiwulan, northeastern Taiwan.
+Ceramics analyzed in this paper come from 40 units (4m by 4m) sampled from four connected trenches at Kiwulan, northeastern Taiwan.
 Kiwulan is situated on a hill near a riverside at the northern margin of Yilan, which is characterized by a triangular alluvial plain facing eastwards the Pacific and mountains on three other sides.
-The chronology of Kiwulan can be divided into two cultural components, the upper and the lower component, with a sterile layer in between [@Chen2007].
+It is an ideal context to study peripheral colonial influence because it was isolated from intensive direct contact with the Spanish and Dutch colonial presence in other parts of Taiwan.
+The chronology of Kiwulan is divided into two cultural components, the upper and the lower component, with a sterile layer in between [@Chen2007].
 This paper focuses on the upper component dating from AD 1350 to AD 1950, which covers the late Iron Age and the historical period defined by the European presence in Taiwan in the early 17th century.
@@
 In 1642, the Spanish was expelled by the Dutch, who took over their forts in northern Taiwan.
-Since then, western Taiwan was mostly under the colonial rule of the Dutch until 1662 when the Kingdom of Tungning was founded by Koxinga, a loyalist of the Ming dynasty of China [@Andrade2007].
+Since then, western Taiwan was mostly under the colonial rule of the Dutch until 1662 when the Kingdom of Tungning in Taiwan was founded by Koxinga, a loyalist of the Ming dynasty of China [@Andrade2007].
 
-The upper component experienced frequent foreign contacts including the European colonial presence in the 17th century and great waves of Chinese immigrants since the 19th century.
-The excavation revealed abundant artifacts in which potsherds are the dominate materials throughout the site.
-Imported ceramics from mainland China, stonewares, and ornaments elements such as beads were also commonly found in the upper component that indicates the frequent long-distance trade activities in the 17th century during the European presence.
-In addition to artifacts, features were also found, such as burials, middens, and post-holes with in-situ posts that explain Kiwulan was a large settlement site [@Chen2007].
-The predominant artefact found at Kiwulan is locally made ceramics, which distributed throughout the sequence across the site.
-
+The upper component experienced foreign contacts including the European colonial presence in the 17th century and great waves of Chinese immigrants since the 19th century.
+Imported ceramics from mainland China, stonewares, and ornaments elements such as beads were trade goods commonly found in the upper component that indicates the frequent long-distance trade activities in the 17th century during the European presence.
+In addition to artifacts, features such as burials, middens, and post-holes with in-situ posts that explain Kiwulan was a large settlement site [@Chen2007].
+The predominant artefact found is locally made ceramics, which distributed throughout the sequence across the site.
 More than 550 thousands of sherds were excavated from the upper component at Kiwulan, and around 1,200 vessels were reconstructed.
-It is interesting to note that there are only two forms of vessels recovered, one is cooking pot and another one is steamer made of two cooking pots stacked together with a filter in between.
-Those pots show high consistency in shape - globular body with short neck and wide mouth.
+It is interesting to note that there are only two forms of vessels recovered, one is cooking pot and another one is steamer made of two cooking pots stacked together with a filter clay layer in between.
+Those pots show high consistency in shape.
+They are globular body with short neck and wide mouth.
 The exterior surface below the neck is treated with a wide variety of impressed geometric motifs.
-They were believed used for cooking because of the evidence of charred residues or carbon deposits commonly observed on the interior and soot on the exterior.
-Besides, there is no other utilitarian earthenware ceramics found at this site suggests this type of pot was mainly used for daily cooking.
+They were believed used for cooking because of the evidence of charred residues and carbon deposits commonly observed on the interior and soot on the exterior.
+Moveover, there is no other utilitarian earthenware ceramics found at this site suggests this type of pot was mainly used for daily cooking.
 Most pots are fired to orange to brownish color with a fully oxidized core or a reduced core with oxidized fringes.
@@
 The digitised outlines were converted into a list of successive x-y pixel coordinates for elliptic Fourier analysis (EFA), which assesses morphological differences among pottery shapes from three occupation contexts.
-
 The harmonic coefficients generated by EFA were analysed by principal component analysis (PCA) to illustrate the diversity of the shape data and identify the major patterns of variation through dimensionality reduction.
@@
 P-values for the significant test of CVs of PC1 and PC2 show significant differences (P-value < 0.01) in the standardization of vessel shape across some periods, especially between Chinese contact with either pre-European or post-European Figure \@ref(fig:CV-test-PCs).
+Significant difference is also detected between pre-Eruopean and post-European (P-value < 0.05).
 
 # Discussion
+
+- cultural resistance
+- trade [@Alizadeh2018]
+- identity, social boundary, Raux2015, Boness2017
+
+The results confirm that the differences in pottery shapes can be detected using oEFA-based outline analysis approach.
+There are differences in pottery shapes detected when indigenous society encountered the foreign presence, including the European in the 17th century and the Chinese in the 19th century.
+The significant difference especially presented between Chinese contact and either pre-European or post-European (P-value < 0.01).
+Previous investigations at Kiwulan suggest an unequal distribution pattern of prestige goods, trade ornaments, during the presence of the European that hints the emergence of social inequality.
+We further examine the ceramic production that usually used as an indicator as the presence of social differentiation, such as the elite control [@Costin2001; @Junker1999].
+Although the difference was detected before and after the European presence, the pottery during the Chinese period show more standardized shape in general.
+This might indicate the appearance of specialized group.
+Given the decline of the indigenous population in the Chinese period might be a cause of a more standardized shape [@Chen2007], we examined the spatial distribution of pottery selected in this paper.
+The pottery distributed across the sampling area without any clusters could rule out the bias caused by the population dynamics.
+
+The nature and context of the interaction with the foreign presences in indigenous society might give some insights into the different degree of standardized pottery among those contact phases.
+
+It is interesting to find the significant difference is observed during the Chinese period that suggests a more standardized pottery shape compared with the previous two periods.
+
+When it comes to ceramic specialization, metric measurements and clay paste are two major indicators.
+Shape analysis could be viewed as a replacement of metric measurement that enables subtle changes to be detected.
+We observe the increasing standardization of the pottery shape, however, it is interesting to note that there is no obvious changes in the clay paste across different phases.
+More standardized shape without changes in raw materials indicate a
 
@@
 
-- cultural resistance
-- trade [@Alizadeh2018]
-- identity, social boundary, Raux2015, Boness2017
+# Conclusion
 
-The results confirm that the differences in pottery shapes can be detected using outline analysis approaches.
-It shows there is differences in pottery shapes between pre-European and post-European, and Pre-European and Chinese contact.
 An increasing standardization of shape of pottery during the Chinese contact period compared to the other two periods is observed.
@@
 This may suggest that an emergence of specialized groups or a consensus of the adoption of similar shape of pottery in the society.
-When it comes to ceramic specialization, metric measurements and clay paste are two major indicators.
-Shape analysis could be viewed as a replacement of metric measurement that enables subtle changes to be detected.
-We observe the increasing standardization of the pottery shape, however, it is interesting to note that there is no obvious changes in the clay paste across different phases.
-In the vitiation that the shape become more standardized without changes in raw materials,
-
-# Conclusion
-
-We find differences in shape and shape standardization of pottery that indicate changes in pottery production resulting from foreign contact, suggesting increasing craft specialization and changes in local social organization at Kiwulan.
 These results are important to understand the influence of culture contact on local indigenous societies and answer the anthropological question that concerns the the mechanisms for social changes.

### 2020-01-07 analysis/paper/paper.Rmd
work on discussion

@@
 Significant difference is also detected between pre-Eruopean and post-European (P-value < 0.05).
+Given the decline of the indigenous population in the Chinese period might be a cause of a more standardized shape [@Chen2007], we examined the spatial distribution of pottery selected in this paper.
+The pottery distributed across the sampling area without any clusters could rule out the bias caused by the population dynamics.
 
 # Discussion
+
+1.
+Did foreign influence have impacts on indigenous pottery production that can be detected in the shape of the vessels?
+2.
+How does pottery change in shape throughout 500 years from a single site that can represent the development of social organization in northeastern Taiwan?
+3. Did pottery become more homogeneous and standardized in shape after foreign contacts with the European colonizer or the Chinese immigrants?
 
@@
 
-The results confirm that the differences in pottery shapes can be detected using oEFA-based outline analysis approach.
-There are differences in pottery shapes detected when indigenous society encountered the foreign presence, including the European in the 17th century and the Chinese in the 19th century.
+Previous investigations at Kiwulan suggest an unequal distribution pattern of prestige goods, trade ornaments, during the presence of the European that hints the emergence of social inequality.
+Using an EFA-based outline approach, we examine the degree of the ceramic standardization as an indicator of the presence of ceramic specialization that is usually associate with social differentiation, such as the elite control [@Costin2001; @Junker1999].
+The results confirm that the differences in pottery shapes can be detected using EFA method.
+There are differences in pottery shapes detected in the indigenous society across different foreign contact phases, including the European in the 17th century and the Chinese in the 19th century.
 The significant difference especially presented between Chinese contact and either pre-European or post-European (P-value < 0.01).
-Previous investigations at Kiwulan suggest an unequal distribution pattern of prestige goods, trade ornaments, during the presence of the European that hints the emergence of social inequality.
-We further examine the ceramic production that usually used as an indicator as the presence of social differentiation, such as the elite control [@Costin2001; @Junker1999].
-Although the difference was detected before and after the European presence, the pottery during the Chinese period show more standardized shape in general.
+We also observe there is no obvious changes in the clay paste across different phases in relation to the increasing standardization of the pottery shape.
+More homogeneous and standardized in shape without changes in raw materials reflects the clay source might be the same but a preferred shape was emphasized during the Chinese contact period.
 This might indicate the appearance of specialized group.
-Given the decline of the indigenous population in the Chinese period might be a cause of a more standardized shape [@Chen2007], we examined the spatial distribution of pottery selected in this paper.
-The pottery distributed across the sampling area without any clusters could rule out the bias caused by the population dynamics.
+The pottery shape during the Chinese period tend to be short with oval-shaped body showing wider body diameter compared to previous two phases.
+It would need further analysis to understand if the changes in shape related to the function of pot.
 
-The nature and context of the interaction with the foreign presences in indigenous society might give some insights into the different degree of standardized pottery among those contact phases.
-
-It is interesting to find the significant difference is observed during the Chinese period that suggests a more standardized pottery shape compared with the previous two periods.
-
-When it comes to ceramic specialization, metric measurements and clay paste are two major indicators.
-Shape analysis could be viewed as a replacement of metric measurement that enables subtle changes to be detected.
-We observe the increasing standardization of the pottery shape, however, it is interesting to note that there is no obvious changes in the clay paste across different phases.
-More standardized shape without changes in raw materials indicate a
-
-Did foreign influence have impacts on indigenous pottery production that can be detected in the shape of the vessels?
-How does pottery change in shape throughout 500 years from a single site that can represent the development of social organization in northeastern Taiwan?
-Did pottery become more homogeneous and standardized in shape after foreign contacts with the European colonizer or the Chinese immigrants?
+The nature and context of the interaction with the foreign presences in indigenous society might give an insight into the different degree of standardized pottery among those contact phases.
 

### 2020-01-08 analysis/paper/paper.Rmd
work on discussion section

@@
 Finger impressions and seams usually on the interior indicate they were pinched using slabs of clay and shaped by hand.
-
 We examine 291 pots recovered from Kiwulan site.
@@
 
-The nature and context of the interaction with the foreign presences in indigenous society might give an insight into the different degree of standardized pottery among those contact phases.
+The nature and context of the interaction with the foreign presences might give an insight into the different degree of standardized pottery among those contact phases.
+Compared to other region in Taiwam, the European colonial control was weak in Yilan due to the relatively isolated location by mountains and the economic consideration of the Spanish and the Dutch who prefer northern Taiwan as trading base.
+Yilan could be seen as a place experienced indirect influences by the European trade network and their colonial activities in a pericolonial context [@Acabado2017].
+On the contrary, the interaction between indigenous people and the Chinese in the 19th century was more intense and direct since those Chinese immigrants settled in Yilan and lived closely with indigenous societies.
+This direct influence is reflected by the archaeological evidence of large amount of Chinese porcelains and distinctive architectural bricks and tiles used by Chinese [@Hsieh2009].
+Also, the finding of adoption of coffin in mortuary practice that viewed as an ethnic symbol for Chinese was observed [@Chen2007].
 

### 2020-01-09 analysis/paper/paper.Rmd
work on discussion

@@
 More homogeneous and standardized in shape without changes in raw materials reflects the clay source might be the same but a preferred shape was emphasized during the Chinese contact period.
-This might indicate the appearance of specialized group.
+The more standardized shape might indicate the appearance of specialized group.
 The pottery shape during the Chinese period tend to be short with oval-shaped body showing wider body diameter compared to previous two phases.
-It would need further analysis to understand if the changes in shape related to the function of pot.
+To figure out whether this change is related to the function of pot, we also attempt to conduct a preliminary lipid analysis for potsherds.
+However, we did not get useful results due to the low lipid yield for meaningful interpretation.
+
+The result of shape analysis suggests that there might be the emergence of specialized groups after the contact with the Europeans.
+This supports our hypothesis that foreign influences had impacts on the emergence of social inequality in local indigenous society.
+This inequality might develop based on the monopolies for trade between a small number of indigenous people and the social organization switch from a more corporate to a more network organization [@Feinman2000].
 

### 2020-01-10 analysis/paper/paper.Rmd
work on discussion, make a table

@@
 
-```{r CV-test-PCs, fig.cap= "P-values of the CV equality test of PC1 and PC2 between phases, red color indicates p-value is less than 0.05"}
+```{r CV-test-PCs-table}
 library(cvequality)
@@
 
+PC1_PC2_cv_tests <-
+bind_rows(PC1_cv_tests, 
+          PC2_cv_tests, .id = "PC") %>% 
+  mutate(PC = str_glue('PC{PC}')) %>% 
+  knitr::kable(PC1_PC2_cv_tests,
+               caption = "P-values of the CV equality test of PC1 and PC2 between phases"))
+```
+
+```{r CV-test-PCs_plot, fig.cap= "P-values of the CV equality test of PC1 and PC2 between phases, red color indicates p-value is less than 0.05"}
+
+PC1_PC2_cv_tests %>%
 bind_rows(PC1_cv_tests, 
@@
 ggsave(here("analysis/figures/cv-test-on-pcs.png"), h = 3, w = 12)
-
 ```
@@
 To see whether the differences in the distribution of PCs between any two phases is statistically significant, coefficients of variation statistic (CVs) among phase groups of PCs is examined with a modified signed-likelihood ratio test for the equality of CVs.
-P-values for the significant test of CVs of PC1 and PC2 show significant differences (P-value < 0.01) in the standardization of vessel shape across some periods, especially between Chinese contact with either pre-European or post-European Figure \@ref(fig:CV-test-PCs).
+P-values for the significant test of CVs of PC1 and PC2 show significant differences (P-value < 0.01) in the standardization of vessel shape across some periods, especially between Chinese contact with either pre-European or post-European Figure \@ref(fig:CV-test-PCs_plot).
 Significant difference is also detected between pre-Eruopean and post-European (P-value < 0.05).
@@
 The pottery distributed across the sampling area without any clusters could rule out the bias caused by the population dynamics.
+
+<---1. cv test for elliptic fourier coefficients 
+2. PCA - Anova - statistical tests - common strategies for testing PCA
+3. specific points, the outliers mean -->
 
@@
 - trade [@Alizadeh2018]
-- identity, social boundary, Raux2015, Boness2017
+- identity, social boundary, visible trait, Raux2015, Boness2017
+- thickness and diameter show difference but not reach the threshold of specialization, also the clay paste does not show any specific change throughout the time
+- shape as an symbol
 
-Previous investigations at Kiwulan suggest an unequal distribution pattern of prestige goods, trade ornaments, during the presence of the European that hints the emergence of social inequality.
-Using an EFA-based outline approach, we examine the degree of the ceramic standardization as an indicator of the presence of ceramic specialization that is usually associate with social differentiation, such as the elite control [@Costin2001; @Junker1999].
-The results confirm that the differences in pottery shapes can be detected using EFA method.
-There are differences in pottery shapes detected in the indigenous society across different foreign contact phases, including the European in the 17th century and the Chinese in the 19th century.
-The significant difference especially presented between Chinese contact and either pre-European or post-European (P-value < 0.01).
-We also observe there is no obvious changes in the clay paste across different phases in relation to the increasing standardization of the pottery shape.
-More homogeneous and standardized in shape without changes in raw materials reflects the clay source might be the same but a preferred shape was emphasized during the Chinese contact period.
-The more standardized shape might indicate the appearance of specialized group.
+Previous investigations at Kiwulan suggest an unequal distribution pattern of prestige goods, particularly trade ornaments, during the presence of the European that hints the emergence of social inequality.
+We examine the degree of the ceramic standardization as an indicator of the presence of ceramic specialization that is usually associate with social differentiation, such as the elite control [@Costin2001; @Junker1999].
+The results confirm that the differences in pottery shapes can be detected using EFA method, indicating the foreign influences on the indigenous society- the Europeans in the 17th century and the Chinese in the 19th century.
+There are significant differences especially between Chinese contact and either pre-European or post-European (P-value < 0.01).
+We also observe no obvious changes in the clay paste across different phases in relation to the increasing standardization of the pottery shape.
+The more standardized shape in the Chinese period might indicate the appearance of specialized group.
+However, when we look at the coef
+
 The pottery shape during the Chinese period tend to be short with oval-shaped body showing wider body diameter compared to previous two phases.
@@
 However, we did not get useful results due to the low lipid yield for meaningful interpretation.
+
+More homogeneous shape without changes in raw materials reflects the clay source might be the same but a preferred shape was emphasized during the Chinese contact period.
 

### 2020-01-10 analysis/paper/paper.Rmd
work on discussion

@@
 The results confirm that the differences in pottery shapes can be detected using EFA method, indicating the foreign influences on the indigenous society- the Europeans in the 17th century and the Chinese in the 19th century.
-There are significant differences especially between Chinese contact and either pre-European or post-European (P-value < 0.01).
-We also observe no obvious changes in the clay paste across different phases in relation to the increasing standardization of the pottery shape.
-The more standardized shape in the Chinese period might indicate the appearance of specialized group.
-However, when we look at the coef
+There are significant differences between pre-European period and post-European period (P-value < 0.05), and between Chinese contact and either pre-European or post-European (P-value < 0.01).
+The clay paste remains similar across different phases in relation to the increasing standardization of the pottery shape based on the petrographic analysis.
+More standardized shape in the Chinese period might indicate the appearance of specialized group.
+The pottery shape during the Chinese period tend to be short with oval-shaped body showing wider body diameter compared to previous two phases.
+To figure out whether this change is related to the function of pot, we made an attempt to conduct lipid analysis for potsherds samples to identify the source of residue absorbed into the fabric.
+However, we did not obtain useful results due to a low lipid yield, which might result from the thin and less porous fabric where not enough room for absorbing food residue [citation].
 
-The pottery shape during the Chinese period tend to be short with oval-shaped body showing wider body diameter compared to previous two phases.
-To figure out whether this change is related to the function of pot, we also attempt to conduct a preliminary lipid analysis for potsherds.
-However, we did not get useful results due to the low lipid yield for meaningful interpretation.
-
-More homogeneous shape without changes in raw materials reflects the clay source might be the same but a preferred shape was emphasized during the Chinese contact period.
-
-The result of shape analysis suggests that there might be the emergence of specialized groups after the contact with the Europeans.
-This supports our hypothesis that foreign influences had impacts on the emergence of social inequality in local indigenous society.
-This inequality might develop based on the monopolies for trade between a small number of indigenous people and the social organization switch from a more corporate to a more network organization [@Feinman2000].
-
-The nature and context of the interaction with the foreign presences might give an insight into the different degree of standardized pottery among those contact phases.
-Compared to other region in Taiwam, the European colonial control was weak in Yilan due to the relatively isolated location by mountains and the economic consideration of the Spanish and the Dutch who prefer northern Taiwan as trading base.
+Although there is differences in pottery shape between pre-European and post-European period, it is more significant in the Chinese period with more standardized pottery shape.
+The results support part of our hypothesis that foreign influences had impacts on the emergence of social inequality in local indigenous society, however, stronger in the Chinese period instead of the European period that we originally thought to be.
+The hypothesis of emergence of social inequality based on the monopolies for long-distance trade brought by the Europeans was not fully supported.
+The nature and context of the interaction with the foreign presences might give an insight into the different degree of standardized pottery.
+Compared to other region in Taiwan, the European colonial control was weak in Yilan due to the relatively isolated location by mountains and the economic consideration of the Spanish and the Dutch who prefer northern Taiwan as trading base.
 Yilan could be seen as a place experienced indirect influences by the European trade network and their colonial activities in a pericolonial context [@Acabado2017].

### 2020-01-10 analysis/paper/scrap.R
delete one script file

@@ -1,3 +1,0 @@
-# Hypotheses and expectations
-
-My hypothesis is that if foreign contacts including European and Chinese had impact on the emergence of social inequality in local indigenous society due to the monopolies for trade between a small number of indigenous people, then I expect to see the changes in social organization from a more corporate to a more network organization that can be reflected by pottery production at Kiwulan. The emergence of craft specialization, here pottery specialization, are usually related to the presence of elite control. In this case, if the competitions for foreign resources and being trade partners of European or Chinese colonizer among individuals gradually lead to the emergence of social inequality, then I expect the local pottery will show more homogeneous features after contact due to the pottery specialization caused by control of small group of individuals.

### 2020-01-11 analysis/paper/paper.Rmd
update for discussion

@@
 Ceramics is another archaeological materials that outline approaches can provide new insights into shape variation to answer anthropological questions related to production.
+@Topi2018 examine 89 photographs of globular jars from the Casas Grandes of northwest Mexico using semi-landmarks method to explore the relationship between social complexity and the emergence of craft specialization.
+Their results suggest that some ceramic types from Medio period (AD 1200-1450), were made by specialists.
+Among those high standardized jars, some were made by attached specialists, while others were made by independent specialists according to distribution patterns of jars.
 @Wilczek2014 evaluate the concordance between outline-based approaches, EFA and Discrete Cosine Transform (DCT), and traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France.
@@
 Furthermore, the EFA results help to understand the level of production standardization over time across the region.
-Similarly, @Topi2018 examine 89 photographs of globular jars from the Casas Grandes of northwest Mexico using semi-landmarks method to explore the relationship between social complexity and the emergence of craft specialization.
-Their results suggest that some ceramic types from Medio period (AD 1200-1450), were made by specialists.
-Among those high standardized jars, some were made by attached specialists, while others were made by independent specialists according to distribution patterns of jars.
-It shows that outline approaches can distinguish variation in ceramic shapes in a finer resolution to understand deeply the issue of craft specialisation.
+Those examples show that outline approaches can distinguish variation in ceramic shapes in a finer resolution to understand deeply the issue of craft specialisation.
 
@@
 Using pottery shape as a proxy to study craft specialization, we answer the questions: Did foreign influence have impacts on indigenous pottery production that can be detected in the shape of the vessels?
-How does pottery change in shape throughout 500 years from a single site that can represent the development of social organization in northeastern Taiwan?
+How does pottery change in shape in relation to foreign influences?
 Did pottery become more homogeneous and standardized in shape after foreign contacts with the European colonizer or the Chinese immigrants?
@@
 They were believed used for cooking because of the evidence of charred residues and carbon deposits commonly observed on the interior and soot on the exterior.
-Moveover, there is no other utilitarian earthenware ceramics found at this site suggests this type of pot was mainly used for daily cooking.
+Moreover, there is no other utilitarian earthenware ceramics found at this site suggests this type of pot was mainly used for daily cooking.
 Most pots are fired to orange to brownish color with a fully oxidized core or a reduced core with oxidized fringes.
@@
   mutate(PC = str_glue('PC{PC}')) %>% 
+  mutate(PC = as.character(PC))
+
 knitr::kable(PC1_PC2_cv_tests,
-               caption = "P-values of the CV equality test of PC1 and PC2 between phases"))
+             caption = "P-values of the CV equality test of PC1 and PC2 between phases")
 ```
@@
 PC1_PC2_cv_tests %>%
-bind_rows(PC1_cv_tests, 
-          PC2_cv_tests, .id = "PC") %>% 
-  mutate(PC = str_glue('PC{PC}')) %>% 
   mutate(colour = if_else(p_value < 0.05, "red", "black")) %>% 
@@
 To see whether the differences in the distribution of PCs between any two phases is statistically significant, coefficients of variation statistic (CVs) among phase groups of PCs is examined with a modified signed-likelihood ratio test for the equality of CVs.
-P-values for the significant test of CVs of PC1 and PC2 show significant differences (P-value < 0.01) in the standardization of vessel shape across some periods, especially between Chinese contact with either pre-European or post-European Figure \@ref(fig:CV-test-PCs_plot).
-Significant difference is also detected between pre-Eruopean and post-European (P-value < 0.05).
+P-values for the significant test of CVs of PC1 and PC2 show significant differences (P-value < 0.01) in the standardization of vessel shape across some periods, especially between Chinese contact with either pre-European or post-European(Figure \@ref(fig:CV-test-PCs_plot), Table \@ref(tab:CV-test-PCs-table)).
+Significant difference is also detected between pre-European and post-European (P-value < 0.05).
 Given the decline of the indigenous population in the Chinese period might be a cause of a more standardized shape [@Chen2007], we examined the spatial distribution of pottery selected in this paper.
@@
 
-1.
-Did foreign influence have impacts on indigenous pottery production that can be detected in the shape of the vessels?
-2.
-How does pottery change in shape throughout 500 years from a single site that can represent the development of social organization in northeastern Taiwan?
+2. How does pottery change in shape in relation to foreign influences?
 3. Did pottery become more homogeneous and standardized in shape after foreign contacts with the European colonizer or the Chinese immigrants?
@@
 
-Previous investigations at Kiwulan suggest an unequal distribution pattern of prestige goods, particularly trade ornaments, during the presence of the European that hints the emergence of social inequality.
-We examine the degree of the ceramic standardization as an indicator of the presence of ceramic specialization that is usually associate with social differentiation, such as the elite control [@Costin2001; @Junker1999].
+Previous investigations at Kiwulan suggest an unequal distribution pattern of prestige goods, trade ornaments specifically, after the presence of the European that hints the emergence of social inequality.
+We examine the degree of the ceramic standardization to measure ceramic specialization that is usually associate with social differentiation, such as the elite control [@Costin2001; @Junker1999].
 The results confirm that the differences in pottery shapes can be detected using EFA method, indicating the foreign influences on the indigenous society- the Europeans in the 17th century and the Chinese in the 19th century.
 There are significant differences between pre-European period and post-European period (P-value < 0.05), and between Chinese contact and either pre-European or post-European (P-value < 0.01).
-The clay paste remains similar across different phases in relation to the increasing standardization of the pottery shape based on the petrographic analysis.
+The average shape presents round body with wider rim and neck before the European contact, and turns to oval-shaped body with narrower rim and neck after the European presence and even more obvious during the Chinese contact period.
+In general, pottery shapes become shorter in hight over time that leads to oval-shaped body.
+The clay paste remains similar across different phases in relation to the increasing standardization of the pottery shape.
 More standardized shape in the Chinese period might indicate the appearance of specialized group.
-The pottery shape during the Chinese period tend to be short with oval-shaped body showing wider body diameter compared to previous two phases.
 To figure out whether this change is related to the function of pot, we made an attempt to conduct lipid analysis for potsherds samples to identify the source of residue absorbed into the fabric.
@@
 Also, the finding of adoption of coffin in mortuary practice that viewed as an ethnic symbol for Chinese was observed [@Chen2007].
+
+Other than appearance of specialized groups, more standardized form without changes in clay paste suggests a preferred shape for pottery production.
+Another explanation could be related to an act of resistance to the Chinese immigrant by emphasizing the transitional practice, especially shape is a visible trait.
+@Raux2015
 
@@
 Wen-Shan Chen in the Department of Geosciences, National Taiwan University for his invaluable guidance for petrographic analysis in his lab.
+This research used statistical consulting resources provided by the Center for Statistics and the Social Sciences, University of Washington.
 

### 2020-01-12 analysis/paper/paper.Rmd
updates for discussion section

@@
 To identify craft specialization, the variation of standardization of specialized product is commonly used as indicator based on the assumption that specialized mass production will lead to uniformity of the product due to routinization, increased skills, and fewer number of producers involved in production [@Stark1995; @Costin1991; @Costin2001].
-For ceramic standardization, several measurements have been proposed, such as metric, compositional, and technological approaches [@Arnold2000; @Rice1991; @Blackman1993; @Costin1991; @Tite1999; @Roux2015].
+For ceramic standardization, several measurements have been proposed, such as metric, compositional, and technological approaches [@Arnold2000; @Boness2015; @Rice1991; @Blackman1993; @Costin1991; @Tite1999; @Roux2015].
 Among those variables, metric measurement is widely applied in archaeological assemblages by calculating coefficient of variation for sets of data coupled with statistical comparison to identify lower value as an indicator of higher degree of standardization [@Eerkens2001; @Roux2003; @Stark1995].
@@
 Shape analysis based on geometric morphometrics method are new approaches for measuring the degree of standardization since it takes the overall shape as variable for comparison [@Slice2007].
-Using geometric morphomerics approaches, we study shape and standardization of locally made ceramics at Kiwulan (1350-1950 AD) [@Chen2007], a large multi-component archaeological site in northeastern Taiwan, to identify changes in ceramic production in cross-cultural interaction contexts.
+Using geometric morphometrics approaches, we study shape and standardization of locally made ceramics at Kiwulan (1350-1950 AD) [@Chen2007], a large multi-component archaeological site in northeastern Taiwan, to identify changes in ceramic production in cross-cultural interaction contexts.
 The aim of this paper is to investigate if there are increasing ceramic specialization resulting from interaction with the Europeans in the 17th century or the Chinese in the 19th century, two major foreign influences in early historical Taiwan, that might indicate social changes in the indigenous society.
@@
 Finger impressions and seams usually on the interior indicate they were pinched using slabs of clay and shaped by hand.
+
 We examine 291 pots recovered from Kiwulan site.
@@
 
-Petrographic analysis for thin sections was conducted at the Department of Geosciences, National Taiwan University.
-The results show high percentage of inclusions (15-50%), including argillite, quartz, feldspar, metasandstone, sandstone fragment, and slate.
+Petrographic analysis for 34 thin sections shows high percentage of inclusions (15-50%), including argillite, quartz, feldspar, metasandstone, sandstone fragment, and slate.
 The most common type of sherds, also the dominant one, has a high percentage of inclusion (25-40%), which is composed mainly by argillite (15-40%), followed by metasandstone (1-10%), sandstone (1-6%), and quartz (1-5%).
 The size of the particles ranges from 500 micron to 1300 micron (=1 mm).
-Overall, this type presents a mixture of fine, rounded argillite with a small part of rounded metasandstone and rounded to sub-angular monocrystalline quartz.
-Those minerals are believed to be local materials of the Yilan Plain.
+In general, this type presents a mixture of fine, rounded argillite with a small part of rounded metasandstone and rounded to sub-angular monocrystalline quartz.
+The composition matches the mineralogical composition of local raw materials in the Yilan Plain [\citation].
 
@@
 
-2. How does pottery change in shape in relation to foreign influences?
-3. Did pottery become more homogeneous and standardized in shape after foreign contacts with the European colonizer or the Chinese immigrants?
-
 - cultural resistance
@@
 The results confirm that the differences in pottery shapes can be detected using EFA method, indicating the foreign influences on the indigenous society- the Europeans in the 17th century and the Chinese in the 19th century.
-There are significant differences between pre-European period and post-European period (P-value < 0.05), and between Chinese contact and either pre-European or post-European (P-value < 0.01).
+The result of MANOVA shows there is significant difference in shapes between pre-European and Post-European period, and between the pre-European and the Chinese period.
 The average shape presents round body with wider rim and neck before the European contact, and turns to oval-shaped body with narrower rim and neck after the European presence and even more obvious during the Chinese contact period.
 In general, pottery shapes become shorter in hight over time that leads to oval-shaped body.
-The clay paste remains similar across different phases in relation to the increasing standardization of the pottery shape.
+For the standardization of pottery shape, the result of CV test indicates the significant differences between pre-European period and post-European period (P-value < 0.05), and between Chinese contact and either pre-European or post-European (P-value < 0.01).
+This suggests pottery shape becomes more homogeneous and standardized after foreign contacts with the European colonizer and even more with the Chinese immigrants.
+The compositional analysis shows clay pastes are similar, regardless of the increasing standardization of the pottery shape, reflecting the raw material sources remains unchanged.
 More standardized shape in the Chinese period might indicate the appearance of specialized group.
@@
 
-Although there is differences in pottery shape between pre-European and post-European period, it is more significant in the Chinese period with more standardized pottery shape.
-The results support part of our hypothesis that foreign influences had impacts on the emergence of social inequality in local indigenous society, however, stronger in the Chinese period instead of the European period that we originally thought to be.
+The results support part of our hypothesis that foreign influences had impacts on the emergence of social inequality in local indigenous society reflected by the degree of pottery standardization, however, stronger in the Chinese period instead of the European period originally thought to be.
 The hypothesis of emergence of social inequality based on the monopolies for long-distance trade brought by the Europeans was not fully supported.
@@
 
-Other than appearance of specialized groups, more standardized form without changes in clay paste suggests a preferred shape for pottery production.
-Another explanation could be related to an act of resistance to the Chinese immigrant by emphasizing the transitional practice, especially shape is a visible trait.
+More standardized pottery shape without controlling the clay paste or changes in technology suggests the shape as a preferred characteristic in pottery production.
+In addition to the appearance of specialized groups, another explanation could be related to an act of resistance to the Chinese immigrant by emphasizing the traditional pottery production practice, of which shape is a visible trait for.
 @Raux2015

### 2020-01-12 analysis/paper/paper.Rmd
updates for dicussion

@@
 
-- cultural resistance
-- trade [@Alizadeh2018]
-- identity, social boundary, visible trait, Raux2015, Boness2017
-- thickness and diameter show difference but not reach the threshold of specialization, also the clay paste does not show any specific change throughout the time
-- shape as an symbol
-
 Previous investigations at Kiwulan suggest an unequal distribution pattern of prestige goods, trade ornaments specifically, after the presence of the European that hints the emergence of social inequality.
@@
 To figure out whether this change is related to the function of pot, we made an attempt to conduct lipid analysis for potsherds samples to identify the source of residue absorbed into the fabric.
-However, we did not obtain useful results due to a low lipid yield, which might result from the thin and less porous fabric where not enough room for absorbing food residue [citation].
+However, we did not obtain useful results due to a low lipid yield, which might result from the thin and less porous fabric that offers limited spaces to protect organic molecules from microbiological degradation [@Evershed2008, pp.
+909].
 
@@
 Compared to other region in Taiwan, the European colonial control was weak in Yilan due to the relatively isolated location by mountains and the economic consideration of the Spanish and the Dutch who prefer northern Taiwan as trading base.
-Yilan could be seen as a place experienced indirect influences by the European trade network and their colonial activities in a pericolonial context [@Acabado2017].
+Yilan could be seen as a place experienced indirect influences by the European trade network and their colonial activities in a pericolonial context [cf.
+@Acabado2017].
 On the contrary, the interaction between indigenous people and the Chinese in the 19th century was more intense and direct since those Chinese immigrants settled in Yilan and lived closely with indigenous societies.
@@
 
-More standardized pottery shape without controlling the clay paste or changes in technology suggests the shape as a preferred characteristic in pottery production.
-In addition to the appearance of specialized groups, another explanation could be related to an act of resistance to the Chinese immigrant by emphasizing the traditional pottery production practice, of which shape is a visible trait for.
-@Raux2015
+More standardized pottery shape without controlling the clay paste or technology suggests the shape as a preferred characteristic in pottery production during the Chinese period.
+However, the standardization is not reflected by the metric attributes, specifically the rim, neck, and body thickness.
+There is higher variation in thickness attributes compared to diameter attributes, indicating shape is the only attribute that presents homogeneous.
+In addition to the emergence of specialized groups indicative of changes in social organization, another explanation is that producing homogeneous pottery is an act of resistance to the Chinese immigrant by emphasizing the practice of traditional pottery production.
+Among the pottery attributes, shape is the most  obvious and visible trait, which can be used as an expression of social boundary or group identity [cf.
+@Raux2015].
+
+- cultural resistance
+- trade [@Alizadeh2018]
+- identity, social boundary, visible trait, Raux2015, Boness2017
+- thickness and diameter show difference but not reach the threshold of specialization, also the clay paste does not show any specific change throughout the time
+- shape as an symbol
 
@@
 
-We would like to thank the Yilan County Cultural Affairs Bureau in Taiwan for permitting access to the pottery used in this study.
+We would like to thank the Yilan County Cultural Affairs Bureau in Taiwan for permitting access to the pottery used in this study and providing the shape images.
 We thank Dr.
-Wen-Shan Chen in the Department of Geosciences, National Taiwan University for his invaluable guidance for petrographic analysis in his lab.
+Wen-Shan Chen in the Department of Geosciences, National Taiwan University for his invaluable guidance of petrographic analysis at his lab.
 This research used statistical consulting resources provided by the Center for Statistics and the Social Sciences, University of Washington.

### 2020-01-12 analysis/paper/paper.Rmd
update the background

@@
 In addition to artifacts, features such as burials, middens, and post-holes with in-situ posts that explain Kiwulan was a large settlement site [@Chen2007].
+To compare different foreign influences, the deposits were classified into three phases, the pre-European, European, and Chinese periods.
+The principle of identification of three phases is based on 32 radiocarbon ages, excavation depth, consistency of contexts, and types of chronologically diagnostic artifacts, such as blue and white porcelains, light grey glazed jars, and large dark brown glazed stoneware jars prevailing in the 17th century, and bricks and tiles used by the Chinese in the 19th century [@Hsieh2009; @Wang2011].
+The deposit shows signs of continuous human occupation during each of the three phases across the sampling area.
+
 The predominant artefact found is locally made ceramics, which distributed throughout the sequence across the site.
@@
 
-We examine 291 pots recovered from Kiwulan site.
+We examine 291 reconstructed pots recovered from the 40 sampling units.
 The layer from 1 to 6 could be divided into 3 time periods.

### 2020-01-12 analysis/paper/pottery_metric.R
add script file for metric data

@@ -1,0 +1,78 @@
+# read metric data
+
+kwl_p_ave <- read.csv(here::here("analysis", "data", "raw_data",
+                                 "KWL-pottery-metrics-tidy-ave.csv"))
+
+# Select metric attributes for computing CV
+kwl_p_metrics_long <-
+  kwl_p_ave %>%
+  mutate(dia_rim_body = Dia_Rim_ave/Dia_Body_ave,
+         thi_rim_body = Thick_Rim_ave/Thick_Body_ave) %>%
+  select(period,
+         Dia_Rim_ave,
+         Dia_Neck_ave,
+         Dia_Body_ave,
+         Thick_Rim_ave,
+         Thick_Neck_ave,
+         Thick_Body_ave,
+         dia_rim_body,
+         thi_rim_body,
+         # H_Rim
+  ) %>%
+  rename(`Rim diameter (mm)` = Dia_Rim_ave,
+         `Neck diameter (mm)` = Dia_Neck_ave,
+         `Body diameter (mm)` = Dia_Body_ave,
+         `Rim thickness (mm)` = Thick_Rim_ave,
+         `Neck thickness (mm)` = Thick_Neck_ave,
+         `Body thickness (mm)` = Thick_Body_ave,
+         `Ratio of Rim/Body diameter` = dia_rim_body,
+         `Ratio of Rim/Body thickness` = thi_rim_body
+  ) %>%
+  gather(variable, value, -period) %>%
+  filter(!is.na(period),
+         !is.na(value))
+
+library(cvequality)
+mslr_test_set <- function(x, y){
+  set.seed(1)
+  mslr_test(nr = 1e4, x, y)
+}
+
+# CV test
+kiwulan_pottery_metrics_cvs_plot_cv_test_MSLRT <-
+  kwl_p_metrics_long %>%
+  nest(-variable) %>%  # or nest(period, value) %>%
+  mutate(mslr_test = map(data,
+                         ~bind_cols(mslr_test_set(
+                           .x$value,
+                           .x$period)))) %>%
+  unnest(mslr_test)
+
+# transform the data
+kwl_p_cvs <-
+  kwl_p_metrics_long %>%
+  group_by(period, variable) %>%
+  summarise(cvs = raster::cv(value, na.rm = TRUE))
+
+# join p-values to CVs
+kwl_p_cvs_and_pvalues <-
+  kwl_p_cvs %>%
[…]

### 2020-01-13 analysis/paper/paper.Rmd
add script file for metric data
update the metric code, and edit the background

@@
 The predominant artefact found is locally made ceramics, which distributed throughout the sequence across the site.
-More than 550 thousands of sherds were excavated from the upper component at Kiwulan, and around 1,200 vessels were reconstructed.
-It is interesting to note that there are only two forms of vessels recovered, one is cooking pot and another one is steamer made of two cooking pots stacked together with a filter clay layer in between.
+More than 550 thousands of sherds were excavated from the upper component, and around 1,200 vessels were reconstructed.
+It is interesting to note that there are only two forms of local vessels, one is cooking pot and another one is steamer made of two cooking pots stacked together with a filter clay layer in between.
 Those pots show high consistency in shape.
@@
 They were believed used for cooking because of the evidence of charred residues and carbon deposits commonly observed on the interior and soot on the exterior.
-Moreover, there is no other utilitarian earthenware ceramics found at this site suggests this type of pot was mainly used for daily cooking.
-Most pots are fired to orange to brownish color with a fully oxidized core or a reduced core with oxidized fringes.
+Moreover, a lack of evidence of other utilitarian earthenware ceramics found at this site suggests the globular pot was mainly used for daily cooking.
+They are fired to orange to brownish color with a fully oxidized core or a reduced core with oxidized fringes.
 Finger impressions and seams usually on the interior indicate they were pinched using slabs of clay and shaped by hand.
 
-We examine 291 reconstructed pots recovered from the 40 sampling units.
-The layer from 1 to 6 could be divided into 3 time periods.
-Layer 5 and 6 represent pre-contact period, layer 4 represents contact period, and layer1, 2, and 3 represent post-contact period.
-The amount of pots for each layer shows below.
-Although most pots are not complete, the thickness from rim, neck, to body can be measured.
-I have also measured the diameter of the rim, neck, and body.
-For those pots that are incomplete, the diameter is measured by its curvature.
-Since the height is incomplete for most pots, this preliminary analysis focuses more on the possible change in thickness and diameter of pot for different parts, and their ratio over time.
+The metric attributes for 362 reconstructed pots (pre-European = 153, European = 173, Chinese = 36) from the 40 sampling units were measure and compared using coefficient of variation (CV) with a significance test for the equality of CVs.
+The attributes include thickness of rim, neck, and body, and diameter of rim, neck, and body.
+Two measurements for thickness and diameter for each pot were took to obtain the average value.
+To understand more about the pottery shape, we also compare the ratio of Rim/Body thickness, and the ratio of Rim/Body diameter.
+The results show that there is a significant difference (P-value < 0.01) in the variation of two metric attributes, the rim diameter and the neck diameter across three phases.
+The CV values show that the variation of rim diameter is small (6.8%) before European contact, increases after the European presence (9.6%), and decreases after the Chinese presence (6.9%).
+For the neck diameter, the CV values indicate smaller variation after the European period (7.7%) and the Chinese period (7.1%) compared to pre-European period(10.6%).
 
@@
 Among the pottery attributes, shape is the most  obvious and visible trait, which can be used as an expression of social boundary or group identity [cf.
-@Raux2015].
+@Roux2015].
 

### 2020-01-13 analysis/paper/pottery_metric.R

@@ -1,9 +1,16 @@
 # read metric data
-
 kwl_p_ave <- read.csv(here::here("analysis", "data", "raw_data",
                                  "KWL-pottery-metrics-tidy-ave.csv"))
 
+count(kwl_p_ave, period)
+
+# subset, but does not work for later CV test
+kwl_p_ave_pre_post <-
+  kwl_p_ave %>%
+  filter(!period == "ch-con")
+
 # Select metric attributes for computing CV
+library(tidyverse)
 kwl_p_metrics_long <-
   kwl_p_ave %>%
   mutate(dia_rim_body = Dia_Rim_ave/Dia_Body_ave,
@@ -64,7 +71,8 @@ kwl_p_cvs_and_pvalues <-
 
 # making table for CV
 kwl_p_cvs_table <-
-  kwl_p_cvs %>%
+  kwl_p_cvs_and_pvalues %>%
+  select(-data, -significant) %>%
   pivot_wider(names_from = period, values_from = cvs) %>%
   mutate(variable = factor(variable, levels = c("Rim thickness (mm)",
                                                 "Neck thickness (mm)",
@@ -74,5 +82,9 @@ kwl_p_cvs_table <-
                                                 "Neck diameter (mm)",
                                                 "Body diameter (mm)",
                                                 "Ratio of Rim/Body diameter"))) %>%
-  arrange(variable)
-kable(kwl_p_cvs_table)
+  arrange(variable) %>%
+  select(variable, `pre-e`, `post-e`, `ch-con`, MSLRT, `p_value`)
+
+write_csv(kwl_p_cvs_table, here::here("analysis", "data", "raw_data", "Kwl_p_metrics_cv_test.csv"))
+
+knitr::kable(kwl_p_cvs_table)

### 2020-01-14 analysis/paper/paper.Rmd
updates for background and discussion

@@
 
+```{r metric-CVs-test}
+library(here)
+metric_cvs_with_phase <- readr::read_csv(here("analysis", "data", "raw_data", "Kwl_p_metrics_cv_test.csv"))
+
+knitr::kable(metric_cvs_with_phase,
+             caption = "Coefficient of Variation test for metric attributes by phases")
+```
+
 The metric attributes for 362 reconstructed pots (pre-European = 153, European = 173, Chinese = 36) from the 40 sampling units were measure and compared using coefficient of variation (CV) with a significance test for the equality of CVs.
 The attributes include thickness of rim, neck, and body, and diameter of rim, neck, and body.
-Two measurements for thickness and diameter for each pot were took to obtain the average value.
+For each pot, two measurements for thickness and diameter were took to compute an average value.
 To understand more about the pottery shape, we also compare the ratio of Rim/Body thickness, and the ratio of Rim/Body diameter.
-The results show that there is a significant difference (P-value < 0.01) in the variation of two metric attributes, the rim diameter and the neck diameter across three phases.
+The results (Table \@ref(tab:metric-CVs-test))show that there is a significant difference (P-value < 0.01) in the variation of two metric attributes, the rim diameter and the neck diameter across three phases.
 The CV values show that the variation of rim diameter is small (6.8%) before European contact, increases after the European presence (9.6%), and decreases after the Chinese presence (6.9%).
@@
 In general, this type presents a mixture of fine, rounded argillite with a small part of rounded metasandstone and rounded to sub-angular monocrystalline quartz.
-The composition matches the mineralogical composition of local raw materials in the Yilan Plain [\citation].
+The composition matches the mineralogical composition of local raw materials in the Yilan Plain [@Chen2016].
+It seems there is no obvious changes in the inclusions over time.
 
@@
 library(tidyverse)
-library(here)
 
@@
 More standardized pottery shape without controlling the clay paste or technology suggests the shape as a preferred characteristic in pottery production during the Chinese period.
-However, the standardization is not reflected by the metric attributes, specifically the rim, neck, and body thickness.
-There is higher variation in thickness attributes compared to diameter attributes, indicating shape is the only attribute that presents homogeneous.
-In addition to the emergence of specialized groups indicative of changes in social organization, another explanation is that producing homogeneous pottery is an act of resistance to the Chinese immigrant by emphasizing the practice of traditional pottery production.
-Among the pottery attributes, shape is the most  obvious and visible trait, which can be used as an expression of social boundary or group identity [cf.
+However, the standardization is not reflected by most of the metric attributes, such as the rim, neck, and body thickness.
+There is higher variation in thickness attributes compared to diameter attributes, indicating shape is the main attribute that presents homogeneity.
+In addition to the emergence of specialized groups indicative of changes in social organization, a further explanation could be that producing homogeneous pottery is an act of resistance to the Europeans intrusion and the the Chinese immigrants by emphasizing the practice of traditional pottery production.
+In a culture contact situation, social identity could be expressed through materials practices to show cultural homogeneous and distinction from other groups [@Voss2005].
+It is also important to recognize that social identity might be more complicated in a colonial context, not just a colonized–colonizer or local/foreign dichotomy [@Voss2005; @Voss2008].
+The case study of Shamaoshan cemetery (3BC- 4AD) in Southwest China showing the process of incorporation of Southeast China into Han Empire involved a century of conflicts, resistance, and acceptance among social groups with different identity, especially in the context of Han immigrants [@Wu2019].
+At Kiwulan, among the pottery attributes, shape is the most obvious and visible trait, which could be viewed as an expression of social boundary or identity [cf.
 @Roux2015].

### 2020-01-15 analysis/paper/paper.Rmd
fix some problems about code and edit text, make the word count right

@@
 To identify craft specialization, the variation of standardization of specialized product is commonly used as indicator based on the assumption that specialized mass production will lead to uniformity of the product due to routinization, increased skills, and fewer number of producers involved in production [@Stark1995; @Costin1991; @Costin2001].
-For ceramic standardization, several measurements have been proposed, such as metric, compositional, and technological approaches [@Arnold2000; @Boness2015; @Rice1991; @Blackman1993; @Costin1991; @Tite1999; @Roux2015].
+For ceramic standardization, several measurements have been proposed and discussed, such as metric, compositional, and technological approaches [@Arnold2000; @Boness2015; @Rice1991; @Blackman1993; @Costin1991; @Tite1999; @Roux2015].
 Among those variables, metric measurement is widely applied in archaeological assemblages by calculating coefficient of variation for sets of data coupled with statistical comparison to identify lower value as an indicator of higher degree of standardization [@Eerkens2001; @Roux2003; @Stark1995].
@@
 Furthermore, the EFA results help to understand the level of production standardization over time across the region.
-Those examples show that outline approaches can distinguish variation in ceramic shapes in a finer resolution to understand deeply the issue of craft specialisation.
+Those examples show that outline approaches can distinguish variation in ceramic shapes in a finer resolution to understand deeply the ceramic production.
 
 Taking the ceramics data from Kiwulan, northeaster Taiwan, we use EFA to evaluate the level of standardization of ceramics in relation to the European presence in the 17th century that might hint the emergence of ceramic specialization.
-In addition, we use significance test for the equality of coefficient of variation of shape variables to compare the vessel shapes from different periods in a statistical way.
+In addition, we use significance test for the equality of coefficient of variation of shape variables to compare the vessel standarization from different periods in a statistical way.
 Using pottery shape as a proxy to study craft specialization, we answer the questions: Did foreign influence have impacts on indigenous pottery production that can be detected in the shape of the vessels?
@@
 Ceramics analyzed in this paper come from 40 units (4m by 4m) sampled from four connected trenches at Kiwulan, northeastern Taiwan.
-Kiwulan is situated on a hill near a riverside at the northern margin of Yilan, which is characterized by a triangular alluvial plain facing eastwards the Pacific and mountains on three other sides.
-It is an ideal context to study peripheral colonial influence because it was isolated from intensive direct contact with the Spanish and Dutch colonial presence in other parts of Taiwan.
-The chronology of Kiwulan is divided into two cultural components, the upper and the lower component, with a sterile layer in between [@Chen2007].
-This paper focuses on the upper component dating from AD 1350 to AD 1950, which covers the late Iron Age and the historical period defined by the European presence in Taiwan in the early 17th century.
+Kiwulan is situated on a hill near a riverside at the northern margin of Yilan, which is characterized by a triangular alluvial plain facing eastwards the Pacific with high mountains on three other sides.
+Yilan is an ideal context to study peripheral colonial influence because it was relatively isolated by natural barrier that reduced frequency of direct contact with the Europeans presences, the Spanish and the Dutch settled in northern Taiwan.
+The chronology of Kiwulan consists of two cultural components, the upper component and the lower component, with a sterile layer in between [@Chen2007].
+This paper focuses on the upper component dated from AD 1350 to AD 1950, covering the late Iron Age and the historical period defined by the European presence in Taiwan in the early 17th century.
 The Dutch firstly came and occupied southern Taiwan in 1624 and then the Spanish occupied northern Taiwan in 1626.
@@
 
-The upper component experienced foreign contacts including the European colonial presence in the 17th century and great waves of Chinese immigrants since the 19th century.
-Imported ceramics from mainland China, stonewares, and ornaments elements such as beads were trade goods commonly found in the upper component that indicates the frequent long-distance trade activities in the 17th century during the European presence.
-In addition to artifacts, features such as burials, middens, and post-holes with in-situ posts that explain Kiwulan was a large settlement site [@Chen2007].
-To compare different foreign influences, the deposits were classified into three phases, the pre-European, European, and Chinese periods.
+The upper component experienced foreign contacts including the Europeans in the 17th century and great waves of Chinese immigrants in the 19th century.
+Imported ceramics from mainland China, stonewares, and ornaments elements such as beads were trade goods commonly found in the upper component that indicates the frequent long-distance trade activities in the 17th century during the European presence in northern Taiwan.
+In addition to artifacts, features such as burials, middens, and post-holes with in-situ posts widespread across 1-2 m thick deposit explain that Kiwulan was a continuously occupied and large settlement site [@Chen2007].
+To compare different foreign influences, the upper component were classified into three phases, the pre-European, European, and Chinese periods.
 The principle of identification of three phases is based on 32 radiocarbon ages, excavation depth, consistency of contexts, and types of chronologically diagnostic artifacts, such as blue and white porcelains, light grey glazed jars, and large dark brown glazed stoneware jars prevailing in the 17th century, and bricks and tiles used by the Chinese in the 19th century [@Hsieh2009; @Wang2011].
@@
 The predominant artefact found is locally made ceramics, which distributed throughout the sequence across the site.
-More than 550 thousands of sherds were excavated from the upper component, and around 1,200 vessels were reconstructed.
+More than 550 thousands of sherds were excavated, and around 1,200 vessels were reconstructed.
 It is interesting to note that there are only two forms of local vessels, one is cooking pot and another one is steamer made of two cooking pots stacked together with a filter clay layer in between.
@@
 The exterior surface below the neck is treated with a wide variety of impressed geometric motifs.
-They were believed used for cooking because of the evidence of charred residues and carbon deposits commonly observed on the interior and soot on the exterior.
-Moreover, a lack of evidence of other utilitarian earthenware ceramics found at this site suggests the globular pot was mainly used for daily cooking.
+They were believed used for cooking based on the evidence of charred residues and carbon deposits commonly observed on the interior and soot on the exterior.
+Moreover, a lack of evidence of other utilitarian earthenware ceramics suggests this globular pot was mainly used for daily cooking.
 They are fired to orange to brownish color with a fully oxidized core or a reduced core with oxidized fringes.
 Finger impressions and seams usually on the interior indicate they were pinched using slabs of clay and shaped by hand.
+In addition to Kiwulan site, this kind of pot was widely found at other archaeological sites during the same period in Yilan Plain.
 
-```{r metric-CVs-test}
+```{r metricCVstest}
 library(here)
@@
 To understand more about the pottery shape, we also compare the ratio of Rim/Body thickness, and the ratio of Rim/Body diameter.
-The results (Table \@ref(tab:metric-CVs-test))show that there is a significant difference (P-value < 0.01) in the variation of two metric attributes, the rim diameter and the neck diameter across three phases.
+The results (Table \@ref(tab:metricCVstest)) show that there is a significant difference (p-value less than 0.01) in the variation of two metric attributes, the rim diameter and the neck diameter across three phases.
 The CV values show that the variation of rim diameter is small (6.8%) before European contact, increases after the European presence (9.6%), and decreases after the Chinese presence (6.9%).
-For the neck diameter, the CV values indicate smaller variation after the European period (7.7%) and the Chinese period (7.1%) compared to pre-European period(10.6%).
+For the neck diameter, the CV values indicate smaller variation after the European period (7.7%) and the Chinese period (7.1%) compared to pre-European period (10.6%).
 
@@
 
-A total of `r total_no` vessels from pre-contact contexts (n = 32), post-European contexts (n = 27), and Chinese contact contexts (n = 14) were selected for this study because of their completeness mostly covering parts from rim to bottom.
+# Methods
 
-# Methods
+A total of `r total_no` vessels from pre-contact contexts (n = 32), post-European contexts (n = 27), and Chinese contact contexts (n = 14) were selected for this study due to their completeness covering most parts from rim to bottom.
 
@@
 Reliable pottery outline was captured by 13 harmonics that gather 99 % of the total harmonic power.
-The average shape of vessels from each phase was visualized with the mean of the standardized Fourier coefficients within each phase group (Figure \@ref(fig:EFA-analysis)).
+The average shape of vessels from each phase was visualized by computing the mean of the standardized Fourier coefficients within each phase group (Figure \@ref(fig:EFA-analysis)).
+Figure \@ref(fig:thin-plate-splines) shows the shape changes described using thin-plate spline warping for paired periods, pre and post European periods, and post-European and Chinese period in a time sequence.
 
@@
 
-```{r thin-plate-splines}
+```{r thin-plate-splines, fig.cap= "Compare the average shapes across phases using thin plate splines (TPS) with outline deformations required to pass from an extreme of the morphospace to the other."}
 # Thin plate spline analysis (TPS)
-# Thin plate splines (TPS) was used to compare average shapes and visualise the outline deformations required to pass from an extreme of the morphospace to the other.
+# visualize the outline deformations required to pass from an extreme of the morphospace to the other
 # Deformation grids
+
+par(mfrow=c(1,2)) # plot them side by side
 
@@
 png(here::here("analysis/figures/tps-iso.png"), res = 600, w = 1200, h = 1800)
-
-#? How to plot them next to each other, cowplot doesn't work when the plot not ggplot objects?
-# plot_grid(a, b, labels = c('A', 'B'))
 
@@
 
+```{r contribution-PC-hide, results="hide", include= FALSE}
+# the contributions of each PC to the shapes
+pc_contrib <-  PCcontrib(op.p, nax = 1:3) 
+```
+
 ```{r contribution-PCs, fig.cap= "Pottery shape variation based on scores along principal component axes.
@@
 Each column shows the reconstructed shape variable according to the score mean ±2 standard deviation (SD) with a 0.5 interval"}
-# the contributions of each PC to the shapes
-pc_contrib <- PCcontrib(op.p, nax = 1:3) 
-#?How to hide above plot, tried dev.off but seems it only stop to compute the following one
 pc_contrib$gg +  
@@
 # the first two principle components of the EFA 
-png(here::here("analysis/figures/pca-biplot.png"), res = 300, w = 1500, h = 1500)
-plot(op.p, 2, 
-     morpho = TRUE, 
-     ellipses = TRUE,
-     cex = 1, 
-     cex.labelsgroups = 1.5)
-dev.off()
+# png(here::here("analysis/figures/pca-biplot.png"), res = 300, w = 1500, h = 1500)
+# plot(op.p, 2, 
+#      morpho = TRUE, 
+#      ellipses = TRUE,
+#      cex = 1, 
+#      cex.labelsgroups = 1.5)
+# dev.off()
 
@@
 
-manv <- MANOVA_PW(op.p, "phase") #? how to hide the process, echo = FALSE?
+manv <- rrtools:::quietly ( MANOVA_PW(op.p, "phase") ) #? how to hide the process, echo = FALSE?
 su_manv <- as.data.frame(manv$summary)
 # how to reduce the number in the table to only 4 digits, tried mutate with signif but doesn't work?
-#su_manv %>% 
-  #mutate('Pr(>F)' = signif('Pr(>F)', 4))
+su_manv <- 
+  su_manv %>% 
+  mutate_all(round, 4)
 
@@
 # we don't use CV because that is not good for measurements with no absolute zero
+xx <- 
   pc_plot %>% 
@@
 
-```{r CV-test-PCs_plot, fig.cap= "P-values of the CV equality test of PC1 and PC2 between phases, red color indicates p-value is less than 0.05"}
+```{r CV-test-PCs-plot, fig.cap= "P-values of the CV equality test of PC1 and PC2 between phases, red color indicates p-value is less than 0.05"}
+
+add_line_format <- function(x, ...){
+  str_replace(x, " vs", "\nvs")
+}
 
@@
   theme_minimal(base_size = 12) +
+  scale_x_discrete(labels = add_line_format(unique(sort(PC1_PC2_cv_tests$phases)))) +
   coord_cartesian(clip = 'off') +
@@
 To see whether the differences in the distribution of PCs between any two phases is statistically significant, coefficients of variation statistic (CVs) among phase groups of PCs is examined with a modified signed-likelihood ratio test for the equality of CVs.
-P-values for the significant test of CVs of PC1 and PC2 show significant differences (P-value < 0.01) in the standardization of vessel shape across some periods, especially between Chinese contact with either pre-European or post-European(Figure \@ref(fig:CV-test-PCs_plot), Table \@ref(tab:CV-test-PCs-table)).
-Significant difference is also detected between pre-European and post-European (P-value < 0.05).
+P-values for the significant test of CVs of PC1 and PC2 show significant differences (P-value less than 0.01) in the standardization of vessel shape across some periods, especially between Chinese contact with either pre-European or post-European(Figure \@ref(fig:CV-test-PCs-plot), Table \@ref(tab:CV-test-PCs-table)).
+Significant difference is also detected between pre-European and post-European (P-value less than 0.05).
 Given the decline of the indigenous population in the Chinese period might be a cause of a more standardized shape [@Chen2007], we examined the spatial distribution of pottery selected in this paper.
 The pottery distributed across the sampling area without any clusters could rule out the bias caused by the population dynamics.
-
-<---1. cv test for elliptic fourier coefficients 
-2. PCA - Anova - statistical tests - common strategies for testing PCA
-3. specific points, the outliers mean -->
 
@@
 In general, pottery shapes become shorter in hight over time that leads to oval-shaped body.
-For the standardization of pottery shape, the result of CV test indicates the significant differences between pre-European period and post-European period (P-value < 0.05), and between Chinese contact and either pre-European or post-European (P-value < 0.01).
+For the standardization of pottery shape, the result of CV test indicates the significant differences between pre-European period and post-European period (P-value less than 0.05), and between Chinese contact and either pre-European or post-European (P-value less than 0.01).
 This suggests pottery shape becomes more homogeneous and standardized after foreign contacts with the European colonizer and even more with the Chinese immigrants.
@@
 More standardized pottery shape without controlling the clay paste or technology suggests the shape as a preferred characteristic in pottery production during the Chinese period.
-However, the standardization is not reflected by most of the metric attributes, such as the rim, neck, and body thickness.
+However, the standardization is not reflected by most metric attributes, such as the rim, neck, and body thickness.
 There is higher variation in thickness attributes compared to diameter attributes, indicating shape is the main attribute that presents homogeneity.
@@
 The case study of Shamaoshan cemetery (3BC- 4AD) in Southwest China showing the process of incorporation of Southeast China into Han Empire involved a century of conflicts, resistance, and acceptance among social groups with different identity, especially in the context of Han immigrants [@Wu2019].
-At Kiwulan, among the pottery attributes, shape is the most obvious and visible trait, which could be viewed as an expression of social boundary or identity [cf.
+Among the pottery attributes, shape is the most obvious and visible trait, which could be viewed as an expression of social boundary or identity [cf.
 @Roux2015].
 
-- cultural resistance
 - trade [@Alizadeh2018]

### 2020-01-16 analysis/paper/paper.Rmd
complete discussion and conclusion

@@
 More standardized pottery shape without controlling the clay paste or technology suggests the shape as a preferred characteristic in pottery production during the Chinese period.
-However, the standardization is not reflected by most metric attributes, such as the rim, neck, and body thickness.
+However, the standardization is not reflected by most metric attributes.
 There is higher variation in thickness attributes compared to diameter attributes, indicating shape is the main attribute that presents homogeneity.
@@
 It is also important to recognize that social identity might be more complicated in a colonial context, not just a colonized–colonizer or local/foreign dichotomy [@Voss2005; @Voss2008].
-The case study of Shamaoshan cemetery (3BC- 4AD) in Southwest China showing the process of incorporation of Southeast China into Han Empire involved a century of conflicts, resistance, and acceptance among social groups with different identity, especially in the context of Han immigrants [@Wu2019].
-Among the pottery attributes, shape is the most obvious and visible trait, which could be viewed as an expression of social boundary or identity [cf.
+The case study of Shamaoshan cemetery (3BC- 4AD) in Southwest China showing the process of incorporation of Southwest China into Han Empire involved a century of conflicts, resistance, and acceptance among social groups with different identity, especially in the historical context of Han immigrants [@Wu2019].
+At Kiwulan, the standardization of potter shape is significant in the 19th century when a a large number of Chinese immigrants migrated to Yilan.
+Shape can be viewed as a symbol as an expression the indigenous identity or social boundary because shape is the most obvious and visible trait compared with other features of pottery [cf.
 @Roux2015].
-
-- trade [@Alizadeh2018]
-- identity, social boundary, visible trait, Raux2015, Boness2017
-- thickness and diameter show difference but not reach the threshold of specialization, also the clay paste does not show any specific change throughout the time
-- shape as an symbol
+Although there is an increase in number of imported ceramics over time and reached high quantity and diversity in the 19th century, the production of the local pottery continued and became more standardized.
+This might imply not only the utilitarian function, but an deliberate action to emphasize the pottery tradition as their cultural custom.
 
@@
 
-An increasing standardization of shape of pottery during the Chinese contact period compared to the other two periods is observed.
-This study provide an comparable basis for our understanding of the standardized pottery as an indicator of social changes in a cultural contact conditions.
-Northeastern Taiwan had been encountered two major foreign contacts in the 17th century with the Europeans and 19th century with the Chines immigrants.
-It shows that the standardization of pottery shape appears during the European presence and increased in the Chinese period.
-This may suggest that an emergence of specialized groups or a consensus of the adoption of similar shape of pottery in the society.
-These results are important to understand the influence of culture contact on local indigenous societies and answer the anthropological question that concerns the the mechanisms for social changes.
-In addition, our case study, which includes an openly available research compendium of R code suitable for use with any other assemblage, will help to expand the use of shape-based quantitative methods to questions about craft specialization and standardization in prehistoric technologies.
+This case study demonstrates the application of Elliptic Fourier analysis on ceramic shape to explore the emergence of ceramic specialization and the underlying mechanism indicative of foreign culture influences.
+Compared with the traditional linear measurement approach, geometric morphometrics methods especially EFA-based outline approach is more effective to detect shape variations for artifact shape without obvious landmarks.
+Here, EFA approach combined with a significance test for the equality of CVs for shape variables provides a robust way to identify and statistically assess the difference in shape standardization.
+Lower variation in ceramic shape was identified after the European presence and even much lower variation during the Chinese presence.
+The direct relationship between foreign influences and highly standardized ceramic shape was tested here on Kiwulan pottery.
+This finding helps to understand the cause and factor that would lead to the standardization of pottery production in a cross-cultural interaction context.
+This gave an insight into the discussion of ceramic specialization which is usually associated with the changes in social organization.
+The more homogeneous shape during the contact period without any changes in clay paste or producing technique suggests that shape standardization is an outcome resulted from an intentional act.
+The results here suggest an expression of social identity or showing cultural boundary in indigenous societies through visible variable when encountering foreign intrusion.
+The analysis allows comparisons between sites with a similar historical trajectory in a pericolonial context.
+Our case study includes an openly available pipeline suitable for other assemblages, ensuring reproducibility and expanding the use of shape-based quantitative methods to questions about craft specialization and standardization in prehistoric technologies.
 

### 2020-01-16 analysis/paper/paper.Rmd
add kWL map

@@
 
-Ceramics analyzed in this paper come from 40 units (4m by 4m) sampled from four connected trenches at Kiwulan, northeastern Taiwan.
+Ceramics analyzed in this paper come from 40 units (4m by 4m) sampled from four connected trenches at Kiwulan, northeastern Taiwan (Figure \@ref(fig:KWL-map)).
 Kiwulan is situated on a hill near a riverside at the northern margin of Yilan, which is characterized by a triangular alluvial plain facing eastwards the Pacific with high mountains on three other sides.
@@
 In addition to Kiwulan site, this kind of pot was widely found at other archaeological sites during the same period in Yilan Plain.
+
+```{r, KWL-map, fig.cap = "Map showing the location of Kiwulan, and other places in northern Taiwan named in the text. Map data from naturalearthdata.com", fig.width=image_width_2_col_inch}
+knitr::include_graphics(here::here("analysis", "figures", "kiwulan-location-map.jpg"))
+```
 

### 2020-01-17 analysis/paper/paper.Rmd
add excavation map

@@
 
-Ceramics analyzed in this paper come from 40 units (4m by 4m) sampled from four connected trenches at Kiwulan, northeastern Taiwan (Figure \@ref(fig:KWL-map)).
+Ceramics analyzed in this paper come from 40 units (4m by 4m) sampled from four connected trenches at Kiwulan, northeastern Taiwan (Figure \@ref(fig:KWL-sam-area), Figure \@ref(fig:KWL-map)).
 Kiwulan is situated on a hill near a riverside at the northern margin of Yilan, which is characterized by a triangular alluvial plain facing eastwards the Pacific with high mountains on three other sides.
@@
 knitr::include_graphics(here::here("analysis", "figures", "kiwulan-location-map.jpg"))
+```
+
+```{r, KWL-sam-area, fig.cap = "Map showing the largest section of excavation areas at Kiwulan, and the distribution of forty squares sampled in this paper presented in red with square ID number.
+Small dots represent the location of post-holes.
+Each square is 4 x 4 m", fig.width=image_width_2_col_inch}
+knitr::include_graphics(here("analysis", "figures", "KWL-excavation-map.jpg"))
 ```

### 2020-01-18 analysis/paper/paper.Rmd
add one refenrence and some sentences for that paper

@@
 The applications in archaeology are seen on human remains or zooarchaeology to understand the biological variation and recently more on stone artifacts to explore stone tool technology [@Fox2015; @Hoggard2019; @Iovictua2010].
-Ceramics is another archaeological materials that outline approaches can provide new insights into shape variation to answer anthropological questions related to production.
-@Topi2018 examine 89 photographs of globular jars from the Casas Grandes of northwest Mexico using semi-landmarks method to explore the relationship between social complexity and the emergence of craft specialization.
-Their results suggest that some ceramic types from Medio period (AD 1200-1450), were made by specialists.
-Among those high standardized jars, some were made by attached specialists, while others were made by independent specialists according to distribution patterns of jars.
-@Wilczek2014 evaluate the concordance between outline-based approaches, EFA and Discrete Cosine Transform (DCT), and traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France.
+Ceramics is another archaeological materials that outline approaches can provide new insights into shape variation to answer anthropological questions related to ceramic specialization.
+Using semi-landmarks method, @Topi2018 examine 89 photographs of globular jars from the Casas Grandes of northwest Mexico.
+Their results suggest that some ceramic types from Medio period (AD 1200-1450), were made by specialists based on high standardized shape.
+Among those high standardized jars, some were made by attached specialists, while others were made by independent specialists according to their spatial distribution patterns.
+Similarity, @Selden2019 examines 28 Caddo bottle excavated at several sites (AD 500-1700) in northwest Louisiana from the Clarence H.
+Webb collections.
+The results indicate a significant integration between the different parts of the bottles including rim, neck, body, and base that vary in a coordinated manner.
+Also, two discrete base and body shapes represented north and south difference were identified.
+For the EFA method, @Wilczek2014 evaluate the concordance between EFA and Discrete Cosine Transform (DCT), and traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France.
 The results show the variation indicated by EFA and DCT matches traditional ceramic typology, which supports that outline-based approaches can be efficiently used for studying variations in ceramic shapes.
@@
 
-Ceramics analyzed in this paper come from 40 units (4m by 4m) sampled from four connected trenches at Kiwulan, northeastern Taiwan (Figure \@ref(fig:KWL-sam-area), Figure \@ref(fig:KWL-map)).
+Ceramics analyzed in this paper come from 40 units (4m by 4m) sampled from the largest excavation section at Kiwulan, northeastern Taiwan (Figure \@ref(fig:KWL-map), Figure \@ref(fig:KWL-sam-area)).
 Kiwulan is situated on a hill near a riverside at the northern margin of Yilan, which is characterized by a triangular alluvial plain facing eastwards the Pacific with high mountains on three other sides.

### 2020-01-18 analysis/paper/paper.Rmd
get different CV test result from another project, but the same data?

@@
 
-Ceramics analyzed in this paper come from 40 units (4m by 4m) sampled from the largest excavation section at Kiwulan, northeastern Taiwan (Figure \@ref(fig:KWL-map), Figure \@ref(fig:KWL-sam-area)).
-Kiwulan is situated on a hill near a riverside at the northern margin of Yilan, which is characterized by a triangular alluvial plain facing eastwards the Pacific with high mountains on three other sides.
+Ceramics analyzed in this paper come from 40 units (4m by 4m) sampled from the largest excavation section at Kiwulan, northeastern Taiwan (Figure \@ref(fig:KWL-map); Figure \@ref(fig:KWL-sam-area)).
+Kiwulan is situated on a hill near a riverside at the northern margin of Yilan, which is characterized by a triangular alluvial plain facing eastwards the Pacific with high mountains on three other sides .
 Yilan is an ideal context to study peripheral colonial influence because it was relatively isolated by natural barrier that reduced frequency of direct contact with the Europeans presences, the Spanish and the Dutch settled in northern Taiwan.
@@
 Each square is 4 x 4 m", fig.width=image_width_2_col_inch}
-knitr::include_graphics(here("analysis", "figures", "KWL-excavation-map.jpg"))
+knitr::include_graphics(here::here("analysis", "figures", "KWL-excavation-map.jpg"))
 ```
@@
   mutate(PC = str_glue('PC{PC}')) %>% 
-  mutate(PC = as.character(PC))
+  mutate(PC = as.character(PC)) %>% 
+  mutate_at(c("MSLRT", "p_value"), round, 4)
 

### 2020-01-18 analysis/paper/pottery_metric.R

@@ -2,15 +2,10 @@
 kwl_p_ave <- read.csv(here::here("analysis", "data", "raw_data",
                                  "KWL-pottery-metrics-tidy-ave.csv"))
 
+library(tidyverse)
 count(kwl_p_ave, period)
 
-# subset, but does not work for later CV test
-kwl_p_ave_pre_post <-
-  kwl_p_ave %>%
-  filter(!period == "ch-con")
-
-# Select metric attributes for computing CV
-library(tidyverse)
+# Select metric attributes and tidy data for computing CV
 kwl_p_metrics_long <-
   kwl_p_ave %>%
   mutate(dia_rim_body = Dia_Rim_ave/Dia_Body_ave,
@@ -39,39 +34,51 @@ kwl_p_metrics_long <-
   filter(!is.na(period),
          !is.na(value))
 
+# CV test function
 library(cvequality)
 mslr_test_set <- function(x, y){
   set.seed(1)
   mslr_test(nr = 1e4, x, y)
 }
 
-# CV test
-kiwulan_pottery_metrics_cvs_plot_cv_test_MSLRT <-
+# filter pre-e and post-e periods
+kwl_p_metrics_long_pre_post <-
   kwl_p_metrics_long %>%
-  nest(-variable) %>%  # or nest(period, value) %>%
-  mutate(mslr_test = map(data,
-                         ~bind_cols(mslr_test_set(
-                           .x$value,
-                           .x$period)))) %>%
-  unnest(mslr_test)
+  filter(!period == "ch-con")
 
-# transform the data
-kwl_p_cvs <-
+# filter post-e and chi periods
+kwl_p_metrics_long_post_chi <-
   kwl_p_metrics_long %>%
+  filter(!period == "pre-e")
+
+# CV test for pre-e and post-e pair
+# ? CV test results are different from the results tested for the LYdissertation project
+kiwulan_pottery_metrics_cvs_plot_cv_test_MSLRT_pre_post <-
+  kwl_p_metrics_long_pre_post %>%
+  nest(-variable) %>%  # or nest(period, value) %>%
+  mutate(asymptotic_test = map(data, # change from mslr_test to asymptotic_test
+                               ~bind_cols(asymptotic_test (
+                                 .x$value,
+                                 .x$period)))) %>%
[…]

### 2020-01-18 analysis/paper/paper.Rmd
edit and move the section of metric measurements, still use MSLRT because the result from asymptotic_test() is problematic

@@
 
-```{r metricCVstest}
-library(here)
-metric_cvs_with_phase <- readr::read_csv(here("analysis", "data", "raw_data", "Kwl_p_metrics_cv_test.csv"))
-
-knitr::kable(metric_cvs_with_phase,
-             caption = "Coefficient of Variation test for metric attributes by phases")
-```
-
-The metric attributes for 362 reconstructed pots (pre-European = 153, European = 173, Chinese = 36) from the 40 sampling units were measure and compared using coefficient of variation (CV) with a significance test for the equality of CVs.
-The attributes include thickness of rim, neck, and body, and diameter of rim, neck, and body.
-For each pot, two measurements for thickness and diameter were took to compute an average value.
-To understand more about the pottery shape, we also compare the ratio of Rim/Body thickness, and the ratio of Rim/Body diameter.
-The results (Table \@ref(tab:metricCVstest)) show that there is a significant difference (p-value less than 0.01) in the variation of two metric attributes, the rim diameter and the neck diameter across three phases.
-The CV values show that the variation of rim diameter is small (6.8%) before European contact, increases after the European presence (9.6%), and decreases after the Chinese presence (6.9%).
-For the neck diameter, the CV values indicate smaller variation after the European period (7.7%) and the Chinese period (7.1%) compared to pre-European period (10.6%).
-
 Petrographic analysis for 34 thin sections shows high percentage of inclusions (15-50%), including argillite, quartz, feldspar, metasandstone, sandstone fragment, and slate.
@@
 
-A total of `r total_no` vessels from pre-contact contexts (n = 32), post-European contexts (n = 27), and Chinese contact contexts (n = 14) were selected for this study due to their completeness covering most parts from rim to bottom.
+A total of `r total_no` vessels from pre-contact contexts (n = 32), post-European contexts (n = 27), and Chinese contact contexts (n = 14) were selected for Ellipic Fourier analysis due to their completeness covering most parts from rim to bottom.
+To fully explore the pottery standardization, we also take the metric measurements into account as a basis for comparison.
+The total of 362 reconstructed pots (pre-European = 153, European = 173, Chinese = 36) were measured.
+The metric attributes include thickness of rim, neck, and body, and diameter of rim, neck, and body.
+For each pot, two measurements for thickness and diameter were took to compute an average value.
+To obtain information about the pottery shape, we also compare the ratio of Rim/Body thickness, and the ratio of Rim/Body diameter.
 
@@
 Feltz and Miller (1996) note that the $D'AD$ value distributes as a central $\chi^2$ random variable with $k-1$ degrees of freedom, from which a p-value can be computed.
-
 We also include the Krishnamoorthy and Lee (2014) test as a more recent development with lower rates of type I error, better performance with uneven sample numbers, and more power across a range of conditions.
@@
 # Results
+
+```{r metricCVstest}
+library(here)
+metric_cvs_with_phase <- readr::read_csv(here::here("analysis", "data", "raw_data", "Kwl_p_metrics_cv_test.csv"))
+metric_cvs_with_phase %>% 
+  mutate_at(c("pre-e", "post-e", "MSLRT", "p_value"), round, 4) %>% 
+  rename(`pre-European` = "pre-e", `after European` = "post-e", Chinese = "ch-con")
+
+knitr::kable(metric_cvs_with_phase,
+             caption = "Coefficient of Variation test for metric attributes by phases")
+```
+
+Table \@ref(tab:metricCVstest)) shows the results of Modified signed-likelihood ratio test (MSLRT) for equality of CVs for for six measurements and two ratio values.
+There is a significant difference (p-value less than 0.01) between CV values for two metric measurements across phases, the rim diameter and the neck diameter.
+The CV values show that the variation of rim diameter is small (6.8%) before European contact, increases after the European presence (9.6%), and decreases after the Chinese presence (6.9%).
+For the neck diameter, the CV values indicate smaller variation after the European period (7.7%) and the Chinese period (7.1%) compared to pre-European period (10.6%).
 

### 2020-01-18 analysis/paper/paper.Rmd
delete some figures and fix knitting issue

@@
 ```{r, KWL-map, fig.cap = "Map showing the location of Kiwulan, and other places in northern Taiwan named in the text. Map data from naturalearthdata.com", fig.width=image_width_2_col_inch}
-knitr::include_graphics(here::here("analysis", "figures", "kiwulan-location-map.jpg"))
+library(here)
+knitr::include_graphics(here("analysis", "figures", "kiwulan-location-map.jpg"))
 ```
@@
 Each square is 4 x 4 m", fig.width=image_width_2_col_inch}
-knitr::include_graphics(here::here("analysis", "figures", "KWL-excavation-map.jpg"))
+knitr::include_graphics(here("analysis", "figures", "KWL-excavation-map.jpg"))
 ```
@@
 
-```{r metricCVstest}
-library(here)
-metric_cvs_with_phase <- readr::read_csv(here::here("analysis", "data", "raw_data", "Kwl_p_metrics_cv_test.csv"))
-metric_cvs_with_phase %>% 
-  mutate_at(c("pre-e", "post-e", "MSLRT", "p_value"), round, 4) %>% 
-  rename(`pre-European` = "pre-e", `after European` = "post-e", Chinese = "ch-con")
-
-knitr::kable(metric_cvs_with_phase,
-             caption = "Coefficient of Variation test for metric attributes by phases")
-```
-
-Table \@ref(tab:metricCVstest)) shows the results of Modified signed-likelihood ratio test (MSLRT) for equality of CVs for for six measurements and two ratio values.
+Table \@ref(tab:metricCVstest) shows the results of Modified signed-likelihood ratio test (MSLRT) for equality of CVs for for six measurements and two ratio values.
 There is a significant difference (p-value less than 0.01) between CV values for two metric measurements across phases, the rim diameter and the neck diameter.
@@
 For the neck diameter, the CV values indicate smaller variation after the European period (7.7%) and the Chinese period (7.1%) compared to pre-European period (10.6%).
+
+```{r metricCVstest}
+metric_cvs_with_phase <- readr::read_csv(here::here("analysis", "data", "raw_data", "Kwl_p_metrics_cv_test.csv"))
+
+metric_cvs_with_phase_tidy <-
+  metric_cvs_with_phase%>% 
+  mutate_at(c("pre-e", "post-e", "ch-con", "MSLRT", "p_value"), round, 4) %>% 
+  rename(`pre-European` = "pre-e", `after European` = "post-e", Chinese = "ch-con")
+
+knitr::kable(metric_cvs_with_phase_tidy,
+             caption = "Coefficient of Variation test for metric attributes by phases")
+```
 

### 2020-01-19 analysis/paper/paper.Rmd
add references for the CV statistical test

@@
 A total of `r total_no` vessels from pre-contact contexts (n = 32), post-European contexts (n = 27), and Chinese contact contexts (n = 14) were selected for Ellipic Fourier analysis due to their completeness covering most parts from rim to bottom.
-To fully explore the pottery standardization, we also take the metric measurements into account as a basis for comparison.
+To fully explore the pottery standardization, we also take the metric measurements into account for comparison.
 The total of 362 reconstructed pots (pre-European = 153, European = 173, Chinese = 36) were measured.
@@
 
-The scanned pottery drawings of those pots were acquired from Bureau of Cultural Affairs in Yilan.
+The scanned pottery drawings were acquired from Bureau of Cultural Affairs in Yilan.
 All drawing presents two-dimensional view of the section of a vessel with indications of metric measurements.
-The scanned drawings were imported into the Inkscape software for outlines tracing to remove additional information such as marks, lines, and measurements on the original drawings.
+The scanned drawings were imported into the Inkscape software for outlines tracing to remove additional information such as marks, lines, and measurements on the original drawings in order to obtain clear outline.
 Each traced half cross-section image was duplicated, flipped, and then joined with another one to create a 2D closed outline.
@@
 The principal components (PCs) scores were analysed with a multivariate analysis of variance (MANOVA) to test for significant effects of the groups of occupation context on shape variances.
-Finally, we computed coefficients of variation statistic (CVs) among multiple groups of PCs with a significance test for the equality of CVs, which enable us to compare CVs in a statistical way (Karl Pearson 1896).
+Finally, we computed coefficients of variation statistic (CVs) among multiple groups of PCs with a significance test for the equality of CVs, which enable us to compare CVs in a statistical way [@Pearson1896].
 The coefficient of variation is a common and widely-used statistical measure of the spread of a set of measurements of a sample.
@@
 To answer this question, we need convenient methods for testing the equality of $c_v$ values from different samples, which has been described in many publications and here we proposed a significant test for CVs using the R package cvequality [@Marwick2019] in the CRAN repository.
-The package cvequality includes two tests, the "Asymptotic test for the equality of coefficients of variation from k populations" (Feltz and Miller 1996) and the "Modified signed-likelihood ratio test (SLRT) for equality of CVs" (Krishnamoorthy and Lee 2014).
-Feltz and Miller (1996) test is widely cited as an authoritative test for the equality of $c_v$ values that performs better than many approximative tests.
+The package cvequality includes two tests, the "Asymptotic test for the equality of coefficients of variation from k populations" [@Feltz1996] and the "Modified signed-likelihood ratio test (SLRT) for equality of CVs" [@Krishnamoorthy2014].
+@Feltz1996 test is widely cited as an authoritative test for the equality of $c_v$ values that performs better than many approximative tests.
 It is a 'gold standard' test for comparing multiple $c_v$.
-Our implementation is based on the following from Feltz and Miller (1996):
+Our implementation is based on the following from @Feltz1996:
 
@@
 The $D'AD$ value measures how far each sample $c_v$ is from our estimate of the population $c_v$.
-Feltz and Miller (1996) note that the $D'AD$ value distributes as a central $\chi^2$ random variable with $k-1$ degrees of freedom, from which a p-value can be computed.
-We also include the Krishnamoorthy and Lee (2014) test as a more recent development with lower rates of type I error, better performance with uneven sample numbers, and more power across a range of conditions.
+@Feltz1996 note that the $D'AD$ value distributes as a central $\chi^2$ random variable with $k-1$ degrees of freedom, from which a p-value can be computed.
+We also include the @Krishnamoorthy2014 test as a more recent development with lower rates of type I error, better performance with uneven sample numbers, and more power across a range of conditions.
 

### 2020-01-19 analysis/paper/paper.Rmd
add distribution map for pottery

@@
 
+# read in pottery data
+pottery_data <- readr::read_csv(here("analysis/data/raw_data/KWL_Ceramic_final2018_with_phases.csv"))
+
 # get list of jpgs only
@@
 # join file names to phase
-pottery_data <- readr::read_csv(here("analysis/data/raw_data/KWL_Ceramic_final2018_with_phases.csv"))
 pottery_data <- pottery_data %>% 
@@
 
-A total of `r total_no` vessels from pre-contact contexts (n = 32), post-European contexts (n = 27), and Chinese contact contexts (n = 14) were selected for Ellipic Fourier analysis due to their completeness covering most parts from rim to bottom.
+A total of `r total_no` relativity intact vessels from pre-contact contexts (n = 32), post-European contexts (n = 27), and Chinese contact contexts (n = 14) were selected for Ellipic Fourier analysis.
+Most of them cover every components from rim to bottom.
+Given that the decline of indigenous population in the Chinese period might have impact on the interpretation, we examined the spatial distribution of pots analyzed in this paper.
+The pottery distributed randomly across the sampling area without any clusters could rule out the bias caused by the population dynamics (Figure \@ref(fig:pottery-spatial-pattern)).
 To fully explore the pottery standardization, we also take the metric measurements into account for comparison.
@@
 To obtain information about the pottery shape, we also compare the ratio of Rim/Body thickness, and the ratio of Rim/Body diameter.
+
+```{r pottery-spatial-pattern, fig.cap= "The spatial distribution of the pottery selected for shape analysis. The quantity is indicated by the color scale."}
+library(sf)
+library(viridis)
+
+# read in the spatial data
+AD_zone <- invisible(st_read(here("analysis", "data", "raw_data", "AD_zone.shp"), quiet = TRUE))
+units_40 <- invisible(read.csv(here("analysis", "data", "raw_data", "kwl-list-of-sampling-squares.csv")))
+
+AD_zone_tidy <-
+  AD_zone %>% 
+  mutate(Pit =paste0("P0",Id)) %>% 
+  inner_join(units_40, by = c("Pit" = "the_sq")) %>% 
+  select(Pit, geometry)
+
+# tidy pottery data
+pottery_count <-
+  pottery_data %>% 
+  select(Site, Context, Pit, Layer, No., phase) %>% 
+  count(Pit, phase)
+
+# join two dataset to get coordinate
+pottery_count_join <-
+  pottery_count %>%
+  mutate(phase = as.factor(phase)) %>% # seems need this line first
+  inner_join(AD_zone_tidy) %>% 
+  filter(!is.na(phase)) %>% 
+  mutate(phase = case_when(
+    phase %in% "pre-e"~ "Before European Contact",
+    phase %in% "post-e" ~ "European Presence",
+    phase %in% "ch-con" ~ "Chinese Presence")) %>% 
+  mutate(phase = factor(phase,
+                        level = c("Before European Contact",
+                                  "European Presence",
+                                  "Chinese Presence")))
+# plot spatial pattern for pottery
+plot_spatial_distribution_pottery_by_phase <- 
+  ggplot() +
+  geom_sf(data = AD_zone_tidy, fill = NA) +
+  geom_sf(data = pottery_count_join, 
+          aes(geometry = geometry, 
+              fill =n), 
+          alpha = 0.9)+
+  facet_wrap(~ phase) +
+  labs(fill = "frequency") +
+  scale_fill_viridis(direction = -1) +
+  theme_minimal() +
+  theme(plot.title = element_text(hjust = 0.5),
+        axis.text.x = element_blank(),
+        axis.text.y = element_blank(),
+        strip.text.x = element_text(size = 6),
+        legend.text = element_text(size = 6),
+        legend.title = element_text(size = 6),
+        legend.key.size =  unit(3, "mm"))
+
+plot_spatial_distribution_pottery_by_phase
+
+ggsave(plot = plot_spatial_distribution_pottery_by_phase,
+       here("analysis", "figures",
+       "plot-spatial-distribution-pottery-by-period.jpg"),
+       width = 135, 
+       height = 31,
+       dpi = 300,
+       units = "mm")
+```
 
@@
 Significant difference is also detected between pre-European and post-European (P-value less than 0.05).
-Given the decline of the indigenous population in the Chinese period might be a cause of a more standardized shape [@Chen2007], we examined the spatial distribution of pottery selected in this paper.
-The pottery distributed across the sampling area without any clusters could rule out the bias caused by the population dynamics.
 

### 2020-01-19 analysis/paper/paper.Rmd
first draft

@@
 keywords: |
-  keyword 1; keyword 2; keyword 3
+  Geometric Morphometrics; ceramic specialization; Iron age; culture contact; standardization
 highlights: |
@@
 
-The development of craft specialization is usually associated with the social changes since it reflects production systems that inform social, economic, or political conditions of a society.
+The development of craft specialization is usually associated with the social changes since it reflects production systems that could inform social, economic, or political conditions of a society.
 Specialization is defined as "differential participation in specific economic activities", in which a particular good is produced by fewer producers compared to the total population who consume it [@Costin1991; @Costin2001, pp.
@@
 The implication of the emergence of specialization for social structure or social process has been a theme in archaeological investigations.
-According to the political approaches, craft specialization, attached production specifically, is viewed as a strategy used by elites to control over production systems for the sake of power and authority [@Costin2001].
-The relationship between craft specialization and social differentiation is broadly discussed, and many cases show that specialized production can be a useful indicator of increasing socio-political complexity [@Hirshman2010; @Junker1999].
-Economic models consider production, exchange, and consumption as a whole to explain specialization as a subsistence strategy to accommodate uneven distribute of resources [@Costin2001; @Arnold1994], a reaction to the changes in subsistence practices [@Stark1995economic], or an outcome stimulated by long-distance trade and inter-regional exchange [@Alizadeh2018].
+From the perspective of the political approaches, craft specialization specifically attached production, is a strategy used by elites to control over production systems to maintain power and authority [@Costin2001].
+The relationship between craft specialization and social differentiation is broadly discussed, and many cases show that specialized production is a useful indicator of increasing socio-political complexity [@Hirshman2010; @Junker1999].
+One the other hand, economic models consider production, exchange, and consumption as a whole to explain specialization as a subsistence strategy to accommodate uneven distribute of resources [@Costin2001; @Arnold1994], a reaction to the changes in subsistence practices [@Stark1995economic], or an outcome result from by long-distance trade and inter-regional exchange [@Alizadeh2018].
 An present-day investigation of ceramic specialization in the Jodhpur region of India suggests that the adoption of standardized products across cultural boundaries is a reaction to a collapse of previous economic system, indicating the correlation with major socio-economic changes [@Roux2015].
-However, it is important to note that the explanation for the emergence of specialized production might not be exclusive or straightforward.
-Moreover, @Acabado2018's case study in lowland northern Philippines shows that specialization is not necessarily accompanied with social differentiation based on the result of a low degree of ceramic specialization in Ifugao ranked societies during the Spanish colonization.
-The development of specialization could be history dependent, where condition and contexts under which specialists organized are important for interpretation of changes in social organization [@Roux2015; @Costin2001].
+However, it is important to note that the explanation for the emergence of specialized production might not be straightforward or exclusive.
+@Acabado2018's case study in lowland northern Philippines shows that specialization is not necessarily accompanied with social differentiation based on the result of a low degree of ceramic specialization in Ifugao ranked societies during the Spanish colonization.
+The development of specialization could be history dependent, where condition and contexts under which specialists organized are important to make an inference for changes in social organization [@Roux2015; @Costin2001].
 
 The emergence of social inequality in the context of cross-cultural interaction especially in a colonial situation was observed in many parts of world when the introduction of foreign trade goods to local indigenous societies.
-The monopolization of long-distance trade goods usaully cause substantial transformations of indigenous economic, cultural, and socio-political systems [@Dietler2005; @Dietler1997; @Junker1993; @Silliman2005].
-Recently more studies discuss the indirect effects of colonialism or pericolonial archaeology that investigates areas where European colonial rule was limited or their conquests were unsuccessful, but their colonial activities had economically and politically impacts on indigenous peoples in the periphery of colonial control [@Trabert2017; @Acabado2017].
+The monopolization of long-distance trade goods usually cause substantial transformations of indigenous economic, cultural, and socio-political systems [@Dietler2005; @Dietler1997; @Junker1993; @Silliman2005].
+Recently indirect effects of colonialism or pericolonial archaeology were address when investigating areas where European colonial rule was limited or their conquests were unsuccessful, but their colonial activities had economically and politically impacts on indigenous peoples in the periphery of colonial control [@Trabert2017; @Acabado2017].
 For example, @Acabado2017's study on Ifugao society in the highland Philippines suggests economic and political intensification during the Spanish presence as the response of indigenous peoples to the Spanish cooptation.
 Indigenous societies might not just passively accept the colonial rules, but instead actively negotiated with the colonist, and accommodate or resist the foreign intrusion through their daily cultural practices, such as consumption patterns of foreign goods [@Dietler2015; @Given2004; @Mullins2011; @Scaramelli2005; @Silliman2001; @TorrenceandClarke2000].
+Northeastern Taiwan is an ideal example to explore the indirect impacts of the foreign occupations on local indigenous societies due to its isolated location.
 The archaeological evidence in northeastern Taiwan show evidence of increasing use of prestige goods such as trade ornaments and uneven distribution pattern in domestic area and burial contexts when the society encountered by the Spanish and the Dutch in the 17th century.
@@
 
-To identify craft specialization, the variation of standardization of specialized product is commonly used as indicator based on the assumption that specialized mass production will lead to uniformity of the product due to routinization, increased skills, and fewer number of producers involved in production [@Stark1995; @Costin1991; @Costin2001].
+To identify craft specialization, the variation of standardization of specialized product is commonly used as an indicator based on the assumption that specialized mass production will lead to uniformity of the product due to routinization, increased skills, and fewer number of producers involved in production [@Stark1995; @Costin1991; @Costin2001].
 For ceramic standardization, several measurements have been proposed and discussed, such as metric, compositional, and technological approaches [@Arnold2000; @Boness2015; @Rice1991; @Blackman1993; @Costin1991; @Tite1999; @Roux2015].
-Among those variables, metric measurement is widely applied in archaeological assemblages by calculating coefficient of variation for sets of data coupled with statistical comparison to identify lower value as an indicator of higher degree of standardization [@Eerkens2001; @Roux2003; @Stark1995].
+Among those variables, metric measurement is widely applied to archaeological assemblages.
+By comparing variations for sets of data using coefficient of variation coupled with statistical testing, the degree of standardization can be identified according to values [@Eerkens2001; @Roux2003; @Stark1995].
 @Roux2018's ethnographic study in the Jodhpur region of India suggests the number of artisans can be assessed by comparing coefficient of variation and the differences between artisans can be detected at both intra-individual level and inter-individual level in a region.
-Although this traditional typological and linear measurements are useful for identifying differences in archaeological assembles, it might be limited because dimensional measurements can be insensitive to subtle variations resulting from changes in assemblages.
+However, linear measurements might be limited for identifying differences in variations because they could be insensitive to subtle variations resulting from changes in assemblages.
 Shape analysis based on geometric morphometrics method are new approaches for measuring the degree of standardization since it takes the overall shape as variable for comparison [@Slice2007].
 Using geometric morphometrics approaches, we study shape and standardization of locally made ceramics at Kiwulan (1350-1950 AD) [@Chen2007], a large multi-component archaeological site in northeastern Taiwan, to identify changes in ceramic production in cross-cultural interaction contexts.
-The aim of this paper is to investigate if there are increasing ceramic specialization resulting from interaction with the Europeans in the 17th century or the Chinese in the 19th century, two major foreign influences in early historical Taiwan, that might indicate social changes in the indigenous society.
+The aim of this paper is to investigate if there is increasing ceramic specialization resulting from interaction with the Europeans in the 17th century or the Chinese in the 19th century, two major foreign influences in early historical Taiwan, that might indicate social changes in local indigenous societies.
 
@@
 The emergence of craft specialization, here pottery specialization, are usually related to changes in social organization towards a society with increasing status inequality.
-In this case, if the competitions for foreign resources and being trade partners of European or Chinese colonizer among individuals gradually lead to increasing social inequality, then we expect the local ceramic will show more homogeneous features after contact due to the craft specialization caused by control of small group of individuals.
+In this case, if the competitions for foreign resources and being trade partners of European or Chinese colonizer among individuals gradually lead to increasing social inequality, then we expect the local ceramic will show more homogeneous features after contact due to the emergence of craft specialization caused by control of small group of individuals.
 This study is important to understand the indirect influences of the European colonists on the local indigenous societies, which remains unclear in East Asia where the colonial power was not successful compared to other places in Southeast Asia.
-Also, it would help to understand the relationship among the use of prestige goods, the degree of ceramic specialization, and the influence of foreign colonizers in a pericolonial context [@Acabado2017].
+Also, it would help to understand the relationship among the use of prestige goods, the degree of ceramic specialization, and the influence of foreign colonizers in a pericolonial context [cf.
+@Acabado2017].
 
@@
 Geometric morphometric methods (GMM) has been increasingly applied in archaeology for shape analysis, which explores morphological variability and similarity of archaeological materials to address the questions of anthropological interests.
-Different from linear measurement approaches that capture shapes by measuring length, width or ratios of objects, geometric morphometrics methods use Cartesian coordinates of morphological structures to define shapes [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007].
+Different from linear measurement approaches that capture shapes by length, width or ratios of objects, geometric morphometrics methods use Cartesian coordinates of morphological structures to define shapes [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007].
 Landmarks, curves or outlines of objects can be represented by coordinates in terms of their unique point locations with respect to numerical values of coordinate axes.
@@
 
-However, one limitation of landmark approaches is that landmark data may not be able to capture the shape differences of a morphological structure where the curving outlines between landmarks are crucial for variation.
+However, one limitation of landmark approaches is that landmarks may not be able to capture the shape differences of a morphological structure where the curving outlines between landmarks are crucial for variation.
 Furthermore, representative and reliable landmark points may be unavailable for complex or rounded shaped structures.
@@
 The Dutch firstly came and occupied southern Taiwan in 1624 and then the Spanish occupied northern Taiwan in 1626.
-In 1642, the Spanish was expelled by the Dutch, who took over their forts in northern Taiwan.
+In 1642, the Spanish was expelled by the Dutch, who took over their forts at Helping dau in Keelung, and in Tamsui.
 Since then, western Taiwan was mostly under the colonial rule of the Dutch until 1662 when the Kingdom of Tungning in Taiwan was founded by Koxinga, a loyalist of the Ming dynasty of China [@Andrade2007].
@@
 The results confirm that the differences in pottery shapes can be detected using EFA method, indicating the foreign influences on the indigenous society- the Europeans in the 17th century and the Chinese in the 19th century.
-The result of MANOVA shows there is significant difference in shapes between pre-European and Post-European period, and between the pre-European and the Chinese period.
-The average shape presents round body with wider rim and neck before the European contact, and turns to oval-shaped body with narrower rim and neck after the European presence and even more obvious during the Chinese contact period.
-In general, pottery shapes become shorter in hight over time that leads to oval-shaped body.
+The result of MANOVA shows there is a significant difference in shapes between pre-European and Post-European period, and between the pre-European and the Chinese period.
+The average shape presents a round body with a wider rim and neck before the European contact, and turns to oval-shaped body with narrower rim and neck after the European presence and even more obvious during the Chinese contact period.
+In general, pottery shapes become shorter in hight over time that leads to the oval-shaped body.
 For the standardization of pottery shape, the result of CV test indicates the significant differences between pre-European period and post-European period (P-value less than 0.05), and between Chinese contact and either pre-European or post-European (P-value less than 0.01).
@@
 The compositional analysis shows clay pastes are similar, regardless of the increasing standardization of the pottery shape, reflecting the raw material sources remains unchanged.
-More standardized shape in the Chinese period might indicate the appearance of specialized group.
+More standardized shape in the Chinese period might indicate the appearance of a specialized group.
 To figure out whether this change is related to the function of pot, we made an attempt to conduct lipid analysis for potsherds samples to identify the source of residue absorbed into the fabric.
@@
 The results support part of our hypothesis that foreign influences had impacts on the emergence of social inequality in local indigenous society reflected by the degree of pottery standardization, however, stronger in the Chinese period instead of the European period originally thought to be.
-The hypothesis of emergence of social inequality based on the monopolies for long-distance trade brought by the Europeans was not fully supported.
-The nature and context of the interaction with the foreign presences might give an insight into the different degree of standardized pottery.
-Compared to other region in Taiwan, the European colonial control was weak in Yilan due to the relatively isolated location by mountains and the economic consideration of the Spanish and the Dutch who prefer northern Taiwan as trading base.
+The hypothesis of the emergence of social inequality based on the monopolies for long-distance trade brought by the Europeans was not fully supported.
+The nature and context of the interaction with the foreign presences might give an insight into the different degrees of standardized pottery.
+Compared to other region in Taiwan, the European colonial control was weak in Yilan due to the relatively isolated location by mountains and the economic consideration of the Spanish and the Dutch who prefer northern Taiwan as a trading base.
 Yilan could be seen as a place experienced indirect influences by the European trade network and their colonial activities in a pericolonial context [cf.
@@
 On the contrary, the interaction between indigenous people and the Chinese in the 19th century was more intense and direct since those Chinese immigrants settled in Yilan and lived closely with indigenous societies.
-This direct influence is reflected by the archaeological evidence of large amount of Chinese porcelains and distinctive architectural bricks and tiles used by Chinese [@Hsieh2009].
+This direct influence is reflected by the archaeological evidence of a large amount of Chinese porcelains and distinctive architectural bricks and tiles used by Chinese [@Hsieh2009].
 Also, the finding of adoption of coffin in mortuary practice that viewed as an ethnic symbol for Chinese was observed [@Chen2007].
@@
 There is higher variation in thickness attributes compared to diameter attributes, indicating shape is the main attribute that presents homogeneity.
-In addition to the emergence of specialized groups indicative of changes in social organization, a further explanation could be that producing homogeneous pottery is an act of resistance to the Europeans intrusion and the the Chinese immigrants by emphasizing the practice of traditional pottery production.
-In a culture contact situation, social identity could be expressed through materials practices to show cultural homogeneous and distinction from other groups [@Voss2005].
+In addition to the emergence of specialized groups indicative of changes in social organization, a further explanation could be that producing homogeneous pottery is an act of resistance to the Europeans intrusion and the Chinese immigrants by emphasizing the practice of traditional pottery production.
+In a culture contact situation, social identity could be expressed through materials practices to show cultural homogeneity and distinction from other groups [@Voss2005].
 It is also important to recognize that social identity might be more complicated in a colonial context, not just a colonized–colonizer or local/foreign dichotomy [@Voss2005; @Voss2008].
-The case study of Shamaoshan cemetery (3BC- 4AD) in Southwest China showing the process of incorporation of Southwest China into Han Empire involved a century of conflicts, resistance, and acceptance among social groups with different identity, especially in the historical context of Han immigrants [@Wu2019].
+The case study of Shamaoshan cemetery (3BC- 4AD) in Southwest China showing the process of incorporation of Southwest China into the Han Empire involved a century of conflicts, resistance, and acceptance among social groups with different identities, especially in the historical context of Han immigrants [@Wu2019].
 At Kiwulan, the standardization of potter shape is significant in the 19th century when a a large number of Chinese immigrants migrated to Yilan.
@@
 
-This case study demonstrates the application of Elliptic Fourier analysis on ceramic shape to explore the emergence of ceramic specialization and the underlying mechanism indicative of foreign culture influences.
+This case study demonstrates the application of Elliptic Fourier analysis on ceramic shapes to explore the emergence of ceramic specialization and the underlying mechanism indicative of foreign cultural influences.
 Compared with the traditional linear measurement approach, geometric morphometrics methods especially EFA-based outline approach is more effective to detect shape variations for artifact shape without obvious landmarks.
@@
 The more homogeneous shape during the contact period without any changes in clay paste or producing technique suggests that shape standardization is an outcome resulted from an intentional act.
-The results here suggest an expression of social identity or showing cultural boundary in indigenous societies through visible variable when encountering foreign intrusion.
+The results here suggest an expression of social identity or showing cultural boundaries in indigenous societies through visible variables when encountering foreign intrusion.
 The analysis allows comparisons between sites with a similar historical trajectory in a pericolonial context.

### 2020-01-23 analysis/paper/pottery_metric.R
solve the problem of CV test

@@ -1,11 +1,11 @@
 # read metric data
-kwl_p_ave <- read.csv(here::here("analysis", "data", "raw_data",
+kwl_p_ave <- readr::read_csv(here::here("analysis", "data", "raw_data",
                                  "KWL-pottery-metrics-tidy-ave.csv"))
 
 library(tidyverse)
 count(kwl_p_ave, period)
 
-# Select metric attributes and tidy data for computing CV
+# select metric attributes and tidy data for computing CV
 kwl_p_metrics_long <-
   kwl_p_ave %>%
   mutate(dia_rim_body = Dia_Rim_ave/Dia_Body_ave,

### 2020-01-23 analysis/paper/pottery_metric.R
updates for CV test code

@@ -51,8 +51,8 @@ kwl_p_metrics_long_post_chi <-
   kwl_p_metrics_long %>%
   filter(!period == "pre-e")
 
-# CV test for pre-e and post-e pair
-# ? CV test results are different from the results tested for the LYdissertation project
+#------------------------------------------CV test for pre-e and post-e pair
+
 kiwulan_pottery_metrics_cvs_plot_cv_test_MSLRT_pre_post <-
   kwl_p_metrics_long_pre_post %>%
   nest(-variable) %>%  # or nest(period, value) %>%
@@ -92,6 +92,89 @@ kwl_p_cvs_table_pre_post <-
   arrange(variable) %>%
   select(variable, `pre-e`, `post-e`, `D_AD`, `p_value`) %>%
   mutate_at(c("pre-e", "post-e", "D_AD", "p_value"), round, 4)
+
+#-----------------------------------CV test for post-e and ch-con pair
+
+kiwulan_pottery_metrics_cvs_plot_cv_test_MSLRT_post_chi <-
+  kwl_p_metrics_long_post_chi %>%
+  nest(-variable) %>%  # or nest(period, value) %>%
+  mutate(asymptotic_test = map(data, # change from mslr_test to asymptotic_test
+                               ~bind_cols(asymptotic_test (
+                                 .x$value,
+                                 .x$period)))) %>%
+  unnest(asymptotic_test)
+
+# compute CV
+kwl_p_cvs_pre_post <-
+  kwl_p_metrics_long_post_chi %>%
+  group_by(period, variable) %>%
+  summarise(cvs = raster::cv(value, na.rm = TRUE))
+
+# join p-values to CVs
+kwl_p_cvs_and_pvalues_post_chi <-
+  kwl_p_cvs_pre_post %>%
+  left_join(kiwulan_pottery_metrics_cvs_plot_cv_test_MSLRT_post_chi) %>%
+  mutate(significant = ifelse(p_value <= 0.05, 'yes', 'no')) %>%
+  mutate(variable = factor(variable, ordered = TRUE)) %>%
+  arrange(variable)
+
+# making table for CV
+kwl_p_cvs_table_post_chi <-
+  kwl_p_cvs_and_pvalues_post_chi %>%
+  select(-data, -significant) %>%
+  pivot_wider(names_from = period, values_from = cvs) %>%
+  mutate(variable = factor(variable, levels = c("Rim thickness (mm)",
+                                                "Neck thickness (mm)",
[…]

### 2020-01-26 analysis/paper/paper.Rmd
para 1

@@
 
-The development of craft specialization is usually associated with the social changes since it reflects production systems that could inform social, economic, or political conditions of a society.
-Specialization is defined as "differential participation in specific economic activities", in which a particular good is produced by fewer producers compared to the total population who consume it [@Costin1991; @Costin2001, pp.
+The development of craft specialization is sometimes associated with social changes since it reflects changes in production systems that could inform on the social, economic, or political conditions of a society.
+We define specialization as "differential participation in specific economic activities", in which a particular good is produced by fewer producers compared to the total population who consume it [@Costin1991; @Costin2001, pp.
 43].
-The implication of the emergence of specialization for social structure or social process has been a theme in archaeological investigations.
-From the perspective of the political approaches, craft specialization specifically attached production, is a strategy used by elites to control over production systems to maintain power and authority [@Costin2001].
-The relationship between craft specialization and social differentiation is broadly discussed, and many cases show that specialized production is a useful indicator of increasing socio-political complexity [@Hirshman2010; @Junker1999].
-One the other hand, economic models consider production, exchange, and consumption as a whole to explain specialization as a subsistence strategy to accommodate uneven distribute of resources [@Costin2001; @Arnold1994], a reaction to the changes in subsistence practices [@Stark1995economic], or an outcome result from by long-distance trade and inter-regional exchange [@Alizadeh2018].
-An present-day investigation of ceramic specialization in the Jodhpur region of India suggests that the adoption of standardized products across cultural boundaries is a reaction to a collapse of previous economic system, indicating the correlation with major socio-economic changes [@Roux2015].
-However, it is important to note that the explanation for the emergence of specialized production might not be straightforward or exclusive.
-@Acabado2018's case study in lowland northern Philippines shows that specialization is not necessarily accompanied with social differentiation based on the result of a low degree of ceramic specialization in Ifugao ranked societies during the Spanish colonization.
-The development of specialization could be history dependent, where condition and contexts under which specialists organized are important to make an inference for changes in social organization [@Roux2015; @Costin2001].
+The effects of the emergence of specialization in past societies on their social structures and social processes is an important question in archaeological investigations of small scale societies in the Late Holocene.
+Considered through a political lens, craft specialization may be a strategy used by elites to control production systems to maintain their power and authority [@Costin2001].
+This relationship between craft specialization and social differentiation has been broadly discussed, and many cases show that increasingly specialized production can be a useful indicator of increasing socio-political complexity [@Hirshman2010; @Junker1999].
+That said, more economic focus considers production, exchange, and consumption as a whole to explain specialization as part of a strategy to accommodate uneven distributions of resources [@Costin2001; @Arnold1994].
+Craft specialisation may appear as a reaction to changes in subsistence practices [@Stark1995economic], or as a result of long-distance trade and inter-regional exchange [@Alizadeh2018].
+Present-day investigation of ceramic specialization in the Jodhpur region of India suggests that the adoption of standardized products across cultural boundaries is a reaction to a collapse of previous economic system, indicating a correlation with major socio-economic changes [@Roux2015].
+However, it is important to note that explanations for the emergence of specialized production in past societies might not be so straightforward or singular.
+@Acabado2018's case study in lowland northern Philippines shows that specialization is not necessarily accompanied with social differentiation based on a low degree of ceramic specialization in Ifugao ranked societies during the time of Spanish colonization.
+The development of specialization could be dependent on unique historical factors, where the conditions and contexts under which specialists organized must be taken into account when making inferences about changes in social organization [@Roux2015; @Costin2001].
 

### 2020-01-26 analysis/paper/paper.Rmd
para 2 & 3

@@
 
-The emergence of social inequality in the context of cross-cultural interaction especially in a colonial situation was observed in many parts of world when the introduction of foreign trade goods to local indigenous societies.
-The monopolization of long-distance trade goods usually cause substantial transformations of indigenous economic, cultural, and socio-political systems [@Dietler2005; @Dietler1997; @Junker1993; @Silliman2005].
-Recently indirect effects of colonialism or pericolonial archaeology were address when investigating areas where European colonial rule was limited or their conquests were unsuccessful, but their colonial activities had economically and politically impacts on indigenous peoples in the periphery of colonial control [@Trabert2017; @Acabado2017].
-For example, @Acabado2017's study on Ifugao society in the highland Philippines suggests economic and political intensification during the Spanish presence as the response of indigenous peoples to the Spanish cooptation.
-Indigenous societies might not just passively accept the colonial rules, but instead actively negotiated with the colonist, and accommodate or resist the foreign intrusion through their daily cultural practices, such as consumption patterns of foreign goods [@Dietler2015; @Given2004; @Mullins2011; @Scaramelli2005; @Silliman2001; @TorrenceandClarke2000].
-Northeastern Taiwan is an ideal example to explore the indirect impacts of the foreign occupations on local indigenous societies due to its isolated location.
-The archaeological evidence in northeastern Taiwan show evidence of increasing use of prestige goods such as trade ornaments and uneven distribution pattern in domestic area and burial contexts when the society encountered by the Spanish and the Dutch in the 17th century.
-In addition to the differential distribution of trade ornaments, the locally made ceramic present a high consistency in form and shape compared to other pottery throughout northeastern Taiwan that hints the possibility of pottery specialization.
+The emergence of social inequality in the context of cross-cultural interaction, especially in a colonial situation, has been observed in many parts of world, for example, when foreign trade goods are introduced to local Indigenous societies.
+The monopolization of long-distance trade goods has caused substantial transformations of Indigenous economic, cultural, and socio-political systems [@Dietler2005; @Dietler1997; @Junker1993; @Silliman2005].
+Pericolonial archaeology is the study of these indirect effects of colonialism, investigating areas where European colonial rule was limited and their conquests were short-lived and unsuccessful, but their colonial activities had economic and political impacts on Indigenous peoples living on the periphery of colonial control and influence [@Trabert2017; @Acabado2017].
+@Acabado2017's study of Ifugao society in the  Philippines highland suggests economic and political intensification during the Spanish presence as the response of Indigenous peoples to the Spanish cooptation.
+Indigenous societies might not simply passively accept colonial rule, but instead actively negotiate with the colonist, and accommodate or resist foreign intrusion through their daily cultural practices, such as their consumption patterns of foreign goods [@Dietler2015; @Given2004; @Mullins2011; @Scaramelli2005; @Silliman2001; @TorrenceandClarke2000].
+In this paper we present a case study from Northeastern Taiwan as an example of these indirect impacts of foreign occupations on local Indigenous societies.
+Our aim is to investigate if there was increasing ceramic specialization resulting from Indigenous interaction with Europeans in the 17th century or Chinese in the 19th century, two major foreign influences in early historical Taiwan, that might indicate social changes in local indigenous societies.
+Previous reports of archaeological evidence in northeastern Taiwan descrbes increases in the use of prestige goods, such as trade ornaments, and uneven artefact distribution patterns in domestic areas and burial contexts at the time when the Indigenous communities encountered the Spanish and the Dutch in the 17th century.
+Here we show that in addition to the differential distribution of trade ornaments post-European contact, the locally-made ceramics become highly consistent in their form and shape, compared to other pottery throughout northeastern Taiwan.
+This shift to higher consistency hints at the possibility of craft specialization in pottery manufacture.
 
-To identify craft specialization, the variation of standardization of specialized product is commonly used as an indicator based on the assumption that specialized mass production will lead to uniformity of the product due to routinization, increased skills, and fewer number of producers involved in production [@Stark1995; @Costin1991; @Costin2001].
-For ceramic standardization, several measurements have been proposed and discussed, such as metric, compositional, and technological approaches [@Arnold2000; @Boness2015; @Rice1991; @Blackman1993; @Costin1991; @Tite1999; @Roux2015].
-Among those variables, metric measurement is widely applied to archaeological assemblages.
+To identify craft specialization in the archaeological record, variation in the standardization of a specialized product is often used as an indicator.
+This is based on the assumption that specialized mass production will lead to uniformity of the product due to routinization, higher skills, and fewer producers involved in production [@Stark1995; @Costin1991; @Costin2001].
+For investigating ceramic standardization, several measurements have been discussed and used, such as metric, compositional, and technological approaches [@Arnold2000; @Boness2015; @Rice1991; @Blackman1993; @Costin1991; @Tite1999; @Roux2015].
+Among those variables, metric measurements are most widely applied to archaeological assemblages because of the ease of collecting these data.
 By comparing variations for sets of data using coefficient of variation coupled with statistical testing, the degree of standardization can be identified according to values [@Eerkens2001; @Roux2003; @Stark1995].
-@Roux2018's ethnographic study in the Jodhpur region of India suggests the number of artisans can be assessed by comparing coefficient of variation and the differences between artisans can be detected at both intra-individual level and inter-individual level in a region.
+@Roux2018's ethnographic study in the Jodhpur region of India suggests the number of artisans can be statistically assessed by comparing coefficients of variation and differences between artisans can be detected at both intra-individual level and inter-individual level in a region.
 However, linear measurements might be limited for identifying differences in variations because they could be insensitive to subtle variations resulting from changes in assemblages.
@@
 Using geometric morphometrics approaches, we study shape and standardization of locally made ceramics at Kiwulan (1350-1950 AD) [@Chen2007], a large multi-component archaeological site in northeastern Taiwan, to identify changes in ceramic production in cross-cultural interaction contexts.
-The aim of this paper is to investigate if there is increasing ceramic specialization resulting from interaction with the Europeans in the 17th century or the Chinese in the 19th century, two major foreign influences in early historical Taiwan, that might indicate social changes in local indigenous societies.
 

### 2020-01-26 analysis/paper/paper.Rmd
para 4

@@
 By comparing variations for sets of data using coefficient of variation coupled with statistical testing, the degree of standardization can be identified according to values [@Eerkens2001; @Roux2003; @Stark1995].
-@Roux2018's ethnographic study in the Jodhpur region of India suggests the number of artisans can be statistically assessed by comparing coefficients of variation and differences between artisans can be detected at both intra-individual level and inter-individual level in a region.
-However, linear measurements might be limited for identifying differences in variations because they could be insensitive to subtle variations resulting from changes in assemblages.
-Shape analysis based on geometric morphometrics method are new approaches for measuring the degree of standardization since it takes the overall shape as variable for comparison [@Slice2007].
-Using geometric morphometrics approaches, we study shape and standardization of locally made ceramics at Kiwulan (1350-1950 AD) [@Chen2007], a large multi-component archaeological site in northeastern Taiwan, to identify changes in ceramic production in cross-cultural interaction contexts.
+@Roux2018's ethnographic study in the Jodhpur region of India suggests the number of artisans can be statistically assessed by comparing coefficients of variation of linear dimensions of pots, and differences between artisans can be detected at both intra-individual level and inter-individual level in a region.
+However, because pottery vessels are typically have curved shapes, linear measurements have limited sensitivity to many kinds of shape variations.
+Thus, to capture subtle shape variations that might be relevant to standardization, a method that is senstive to non-linear aspects of shape variation is necessary.
+Here we analyse pottery shape using geometric morphometric methods, a new approach for measuring the degree of standardization because it takes a statistical summary of the overall shape as variable for comparison [@Slice2007].
+Using geometric morphometry, we study shape and standardization of locally made ceramics at Kiwulan (1350-1950 AD) [@Chen2007], a large multi-component archaeological site in northeastern Taiwan, to identify changes in ceramic production resulting from cross-cultural interaction contexts.
 
-Our hypothesis is that if foreign influences had impacts on the emergence of social inequality in local indigenous society due to the monopolies for trade between a small number of indigenous people, then we expect to see the changes in social organization from a more corporate to a more network organization that would be reflected by pottery production at Kiwulan [@Feinman2000].
-The emergence of craft specialization, here pottery specialization, are usually related to changes in social organization towards a society with increasing status inequality.
-In this case, if the competitions for foreign resources and being trade partners of European or Chinese colonizer among individuals gradually lead to increasing social inequality, then we expect the local ceramic will show more homogeneous features after contact due to the emergence of craft specialization caused by control of small group of individuals.
-This study is important to understand the indirect influences of the European colonists on the local indigenous societies, which remains unclear in East Asia where the colonial power was not successful compared to other places in Southeast Asia.
-Also, it would help to understand the relationship among the use of prestige goods, the degree of ceramic specialization, and the influence of foreign colonizers in a pericolonial context [cf.
+Our hypothesis is that if foreign influences had impacts on the emergence of social inequality in local Indigenous society due to monopolies of trade arising among a small number of Indigenous people, then we expect to see changes in social organization from a more corporate to a more network organization [@Feinman2000].
+We expect this corporate-network shift to manifest as increased specialization pottery manufacture, indicated archaeologically by icreased standardisation of pottery production at Kiwulan.
+The emergence of craft specialization, here pottery specialization, is often related to changes in social organization towards a society with increasing status inequality.
+In this case, if the competition among Indigenous individual for foreign resources and trade partnership with European or Chinese colonizers gradually lead to increasing social inequality, then we expect the local ceramics will show more homogeneous shapes after contact due to the emergence of craft specialization caused by rgreater economic and social control by a small group of individuals at Kiwulan.
+Our results are important for understanding the indirect influences of  European colonists on local Indigenous societies, which remains unclear in East Asia where colonial power had limited success compared to other places in Southeast Asia.
+Our findings improve our understanding of the relationship between the use of prestige goods, the degree of ceramic specialization, and the influence of foreign colonizers in a pericolonial context [cf.
 @Acabado2017].
 
-## Geometric Morphometric approaches
+## Geometric Morphometry
 

### 2020-01-26 analysis/paper/paper.Rmd
up to # Materials and archaeological background

@@
 
-Geometric morphometric methods (GMM) has been increasingly applied in archaeology for shape analysis, which explores morphological variability and similarity of archaeological materials to address the questions of anthropological interests.
-Different from linear measurement approaches that capture shapes by length, width or ratios of objects, geometric morphometrics methods use Cartesian coordinates of morphological structures to define shapes [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007].
-Landmarks, curves or outlines of objects can be represented by coordinates in terms of their unique point locations with respect to numerical values of coordinate axes.
-According to different focuses and nature of original shape data, there are two common morphometric methods, landmark approaches and outline approaches [@Adams2004].
-Landmark approaches assign a set of landmarks or semilandmarks onto objects as reference points that can be specified on a coordinate system as x, y coordinates for two dimensions, or x, y, z coordinates for three dimensions.
-Based on distances between raw landmarks, objects are translated to a common centroid, rescaled into the same size, and rotated until the summed squared distances between corresponding landmarks are minimized using generalized Procrustes analysis (GPA) [@Bookstein1991].
-GPA is a method that superimposes sets of landmark configurations into a common coordinate system, where superimposed landmark coordinates are used as shape variables for further multivariate statistics analyses [@Slice2007].
-Landmark-based morphometrics are widely applied to archaeological objects with a obvious morphological signature that can provide ideal reference points for landmark placement, such as projectile point tips or biologically measurements points indicating osteological features [@Birch2019; @Buchanan2019; @Lycett2013; @Cardillo2010; @Haruda2019; @Meloro2015].
+Geometric morphometric methods (GMM) has been increasingly applied in archaeology for shape analysis, which explores morphological variability and similarity of archaeological materials to address the questions of anthropological interest.
+Geometric morphometrics differs from traditional linear measurements through its use of Cartesian coordinates of morphological structures to define shapes [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007].
+Landmarks, curves or outlines of objects can be represented by coordinates in terms of their unique point locations with respect to numerical values on coordinate axes.
+There are two common morphometric methods: landmark approaches and outline approaches [@Adams2004].
+Landmark approaches assign a set of landmarks or semilandmarks onto objects as reference points that can be specified on a coordinate system as xy coordinates for two dimensions, or xyz coordinates for three dimensions.
+In a typical analysis, starting with raw landmarks, a sample of objects are translated to a common centroid, rescaled into the same size, and rotated until the summed squared distances between corresponding landmarks are minimized using generalized Procrustes analysis (GPA) [@Bookstein1991].
+GPA is a method that superimposes sets of landmark configurations into a common coordinate system, where superimposed landmark coordinates are used as shape variables for further multivariate statistical analyses [@Slice2007].
+Landmark-based morphometrics are widely applied to archaeological objects with obvious morphological features that can provide unambiguous reference points for landmark placement, such as projectile point tips or biological measurements points indicating osteological features [@Birch2019; @Buchanan2019; @Lycett2013; @Cardillo2010; @Haruda2019; @Meloro2015].
 
-However, one limitation of landmark approaches is that landmarks may not be able to capture the shape differences of a morphological structure where the curving outlines between landmarks are crucial for variation.
-Furthermore, representative and reliable landmark points may be unavailable for complex or rounded shaped structures.
-In those cases, outline approaches are better way for analyzing the overall shape of an object.
-One of the outline approaches is the semi-landmarks method, also called sliding landmarks, which assigns points along the curve between two landmarks at defined intervals [@Bookstein1997; @Lawing2010].
+However, one limitation of landmark approaches is that landmarks may not be able to capture the shape differences of a morphological structure where the curving outlines between landmarks contains crucial variation.
+Furthermore, landmark points may be unavailable for structures that are mostly or entirely curves.
+In those cases, outline approaches are more suitable for analyzing the overall shape of an object.
+One outline approache is the semi-landmarks method, also called sliding landmarks, which assigns points along the curve between two landmarks at defined intervals [@Bookstein1997; @Lawing2010].
 Those semi-landmarks are allowed to slide along the curve to remove the effect of the arbitrary landmark spacing by minimizing either Procrustes distance or bending energy [@Bookstein1997; @Slice2007; @Gunz2013].
-Another approach is elliptic Fourier Analysis (EFA) that turns coordinates into coefficients using mathematical function, commonly applied to two-dimensional closed shape [@Kuhl1982].
-EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions, called harmonic coefficients [@Adams2004; @Claude2008; @Bonhomme2014].
-The number of harmonics determines the quality and precision of the geometry of an object.
-The harmonic power, a cumulated sum of squared harmonic coefficient, provides a robust rule for the desired number of harmonics [@Bonhomme2014].
+Another approach that is commonly applied to two-dimensional closed shapes is elliptic Fourier Analysis (EFA), that converts coordinates along a curve into Fourier function coefficients, called harmonic coefficients [@Kuhl1982].
+EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions [@Adams2004; @Claude2008; @Bonhomme2014].
+The number of harmonics determines the quality and precision of the geometric representation of an object.
+The harmonic power, a cumulated sum of squared harmonic coefficients, provides a robust rule for determining the desired number of harmonics [@Bonhomme2014].
 
-Outline approaches especially EFA is suitable for shapes lacking representative landmarks or curves convey more meaningful variations.
-The applications in archaeology are seen on human remains or zooarchaeology to understand the biological variation and recently more on stone artifacts to explore stone tool technology [@Fox2015; @Hoggard2019; @Iovictua2010].
+EFA is suitable for shapes lacking representative landmarks or where curves contain the most meaningful variation.
+Applications in archaeology include human remains and zooarchaeology to understand  biological variation, and more recently on stone artifacts to explore stone tool technology [@Fox2015; @Hoggard2019; @Iovictua2010].
 Ceramics is another archaeological materials that outline approaches can provide new insights into shape variation to answer anthropological questions related to ceramic specialization.
 Using semi-landmarks method, @Topi2018 examine 89 photographs of globular jars from the Casas Grandes of northwest Mexico.
-Their results suggest that some ceramic types from Medio period (AD 1200-1450), were made by specialists based on high standardized shape.
-Among those high standardized jars, some were made by attached specialists, while others were made by independent specialists according to their spatial distribution patterns.
-Similarity, @Selden2019 examines 28 Caddo bottle excavated at several sites (AD 500-1700) in northwest Louisiana from the Clarence H.
-Webb collections.
+Their results suggest that some ceramic types from the Medio period (AD 1200-1450), were made by specialists based on highly standardized shapes.
+Among those highly standardized jars, some were made by attached specialists <!-- what does this mean, 'attached' to what?
+please clarify -->, while others were made by independent specialists according to their spatial distribution patterns.
+Similarity, @Selden2019 examined 28 Caddo bottle excavated at several sites (AD 500-1700) in northwest Louisiana <!-- using semi-landmarks or something else?
+please add -->.
 The results indicate a significant integration between the different parts of the bottles including rim, neck, body, and base that vary in a coordinated manner.
-Also, two discrete base and body shapes represented north and south difference were identified.
-For the EFA method, @Wilczek2014 evaluate the concordance between EFA and Discrete Cosine Transform (DCT), and traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France.
-The results show the variation indicated by EFA and DCT matches traditional ceramic typology, which supports that outline-based approaches can be efficiently used for studying variations in ceramic shapes.
-Furthermore, the EFA results help to understand the level of production standardization over time across the region.
-Those examples show that outline approaches can distinguish variation in ceramic shapes in a finer resolution to understand deeply the ceramic production.
+Also, two discrete base and body shapes represented northern and southern differences were identified.
+For the EFA method, @Wilczek2014 evaluate the concordance between EFA and Discrete Cosine Transform (DCT), and a traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France.
+The results show the variation indicated by EFA and DCT matches the traditional ceramic typology, which supports that claim that outline-based approaches can be efficiently used for studying variations in ceramic shapes.
+Furthermore, @Wilczek201's EFA results help us understand the level of production standardization over time across the region.
+Those examples show that outline approaches can distinguish variation in ceramic shapes at high resolutions to understand ceramic production.
 
-Taking the ceramics data from Kiwulan, northeaster Taiwan, we use EFA to evaluate the level of standardization of ceramics in relation to the European presence in the 17th century that might hint the emergence of ceramic specialization.
-In addition, we use significance test for the equality of coefficient of variation of shape variables to compare the vessel standarization from different periods in a statistical way.
-Using pottery shape as a proxy to study craft specialization, we answer the questions: Did foreign influence have impacts on indigenous pottery production that can be detected in the shape of the vessels?
-How does pottery change in shape in relation to foreign influences?
-Did pottery become more homogeneous and standardized in shape after foreign contacts with the European colonizer or the Chinese immigrants?
-Starting from the historical background of colonial influences and its reflection in the material record, this study explore the relationship between the emergence of social inequality and craft specialization that can extend our understanding of the reactions of indigenous societies in the periphery of colonial control, providing a perspective that could be applied in other parts of the world in a similar context.
+Taking the ceramics data from Kiwulan, northeastern Taiwan, we use EFA to evaluate the level of standardization of ceramics in relation to the European presence in the 17th century to get insights into the emergence of ceramic specialization.
+We use a significance test for the equality of coefficient of variations of shape variables to statistically compare the vessel standarization from different periods.
+Using pottery shape as a proxy to study craft specialization, we address these questions: Did foreign influence have impacts on indigenous pottery production that can be detected in the shape of the vessels?
+Did pottery shape become more homogeneous and standardized after foreign contacts with European colonizers or Chinese immigrants?
 

### 2020-01-26 analysis/paper/paper.Rmd
working on methods section

@@
 
-# Materials and archaeological background
+# Archaeological background and materials
 
-Ceramics analyzed in this paper come from 40 units (4m by 4m) sampled from the largest excavation section at Kiwulan, northeastern Taiwan (Figure \@ref(fig:KWL-map); Figure \@ref(fig:KWL-sam-area)).
-Kiwulan is situated on a hill near a riverside at the northern margin of Yilan, which is characterized by a triangular alluvial plain facing eastwards the Pacific with high mountains on three other sides .
-Yilan is an ideal context to study peripheral colonial influence because it was relatively isolated by natural barrier that reduced frequency of direct contact with the Europeans presences, the Spanish and the Dutch settled in northern Taiwan.
-The chronology of Kiwulan consists of two cultural components, the upper component and the lower component, with a sterile layer in between [@Chen2007].
-This paper focuses on the upper component dated from AD 1350 to AD 1950, covering the late Iron Age and the historical period defined by the European presence in Taiwan in the early 17th century.
-The Dutch firstly came and occupied southern Taiwan in 1624 and then the Spanish occupied northern Taiwan in 1626.
-In 1642, the Spanish was expelled by the Dutch, who took over their forts at Helping dau in Keelung, and in Tamsui.
-Since then, western Taiwan was mostly under the colonial rule of the Dutch until 1662 when the Kingdom of Tungning in Taiwan was founded by Koxinga, a loyalist of the Ming dynasty of China [@Andrade2007].
+Ceramics analyzed in this paper come from 40 units (4m by 4m) sampled from the central, undisturbed area of archaeological excavations at Kiwulan (Figure \@ref(fig:KWL-map); Figure \@ref(fig:KWL-sam-area)).
+Kiwulan is situated on a hill near a riverside at the northern margin of Yilan Province, which is characterized by a triangular alluvial plain facing eastwards the Pacific with high mountains on three other sides.
+Yilan is an ideal context to study peripheral colonial influence because Indigenous communities there were isolated by natural barriers that limited the frequency of direct contact with the Spanish and the Dutch settlers in northern Taiwan.
+The chronology of the archaeological deposits at Kiwulan consists of two cultural components, the upper component and the lower component, with a sterile layer in between [@Chen2007].
+This paper focuses on the upper component, dated from AD 1350 to AD 1950, covering the late Iron Age and the historical period, which we define as the European presence in Taiwan in the early 17th century.
+The Dutch first occupied southern Taiwan in 1624 and then the Spanish occupied northern Taiwan in 1626 <!-- add citation for these dates -->.
+In 1642, the Spanish was expelled by the Dutch in northern Taiwan, who took over their forts at Helping Dau in Keelung, and in Tamsui.
+Since then, western Taiwan was mostly under Dutch colonial rule until 1662 when the Kingdom of Tungning in Taiwan was founded by Koxinga, a loyalist of the Ming dynasty of China [@Andrade2007].
 
-The upper component experienced foreign contacts including the Europeans in the 17th century and great waves of Chinese immigrants in the 19th century.
-Imported ceramics from mainland China, stonewares, and ornaments elements such as beads were trade goods commonly found in the upper component that indicates the frequent long-distance trade activities in the 17th century during the European presence in northern Taiwan.
-In addition to artifacts, features such as burials, middens, and post-holes with in-situ posts widespread across 1-2 m thick deposit explain that Kiwulan was a continuously occupied and large settlement site [@Chen2007].
-To compare different foreign influences, the upper component were classified into three phases, the pre-European, European, and Chinese periods.
-The principle of identification of three phases is based on 32 radiocarbon ages, excavation depth, consistency of contexts, and types of chronologically diagnostic artifacts, such as blue and white porcelains, light grey glazed jars, and large dark brown glazed stoneware jars prevailing in the 17th century, and bricks and tiles used by the Chinese in the 19th century [@Hsieh2009; @Wang2011].
+The archaeological record of Kiwulan's upper component shows traces of foreign contacts, including Europeans in the 17th century, and waves of Chinese immigrants in the 19th century.
+Imported ceramics from mainland China, stoneware, and ornaments such as beads have been found in the upper component, indicating frequent long-distance trade activities in the 17th century during the European presence in northern Taiwan.
+In addition to artifacts, archaeological features such as burials, middens, and post-holes with _in-situ_ posts are widespread across  the 1-2 m thick deposit of the upper component, and demonstrate that Kiwulan was a continuously occupied large settlement site [@Chen2007].
+To compare different foreign influences, we classified the upper component into three chronological phases: the pre-European, European, and Chinese periods.
+We identified these phases using 32 previously published radiocarbon ages <!-- add citaion to report with dates -->, excavation depth measurements, stratigraphic details reported by the excavators (colour, disturbance, etc.), and finds of chronologically diagnostic artifacts, such as blue and white porcelains, light grey glazed jars, and large dark brown glazed stoneware jars commonly used in the 17th century, and bricks and tiles used by the Chinese in the 19th century [@Hsieh2009; @Wang2011].
 The deposit shows signs of continuous human occupation during each of the three phases across the sampling area.
 
-The predominant artefact found is locally made ceramics, which distributed throughout the sequence across the site.
-More than 550 thousands of sherds were excavated, and around 1,200 vessels were reconstructed.
-It is interesting to note that there are only two forms of local vessels, one is cooking pot and another one is steamer made of two cooking pots stacked together with a filter clay layer in between.
-Those pots show high consistency in shape.
-They are globular body with short neck and wide mouth.
-The exterior surface below the neck is treated with a wide variety of impressed geometric motifs.
-They were believed used for cooking based on the evidence of charred residues and carbon deposits commonly observed on the interior and soot on the exterior.
-Moreover, a lack of evidence of other utilitarian earthenware ceramics suggests this globular pot was mainly used for daily cooking.
+The most abundant artefacts in the upper component are locally made ceramics, which are distributed throughout the sequence and across the site.
+More than 550,000 sherds were excavated, and around 1,200 vessels could be reconstructed.
+There are only two forms of local vessels, a cooking pot and a steamer made of two cooking pots stacked together with a clay filter layer in between.
+Those pots forms show high consistency in shape.
+They have a globular body with a short neck and wide mouth.
+The exterior surface below the neck is decorated with a wide variety of impressed geometric motifs.
+These vessels were probably used for cooking, with evidence of charred residues and carbon deposits frequently observed on vessel interiors and soot on their exteriors.
+Moreover, a lack of evidence of other utilitarian earthenware ceramics suggests these globular pots were mainly used for daily cooking.
 They are fired to orange to brownish color with a fully oxidized core or a reduced core with oxidized fringes.
 Finger impressions and seams usually on the interior indicate they were pinched using slabs of clay and shaped by hand.
-In addition to Kiwulan site, this kind of pot was widely found at other archaeological sites during the same period in Yilan Plain.
+This kind of pot has been widely found at archaeological sites during the late Iron Age and the historical period throughout the Yilan Plain.
 
@@
 
-Petrographic analysis for 34 thin sections shows high percentage of inclusions (15-50%), including argillite, quartz, feldspar, metasandstone, sandstone fragment, and slate.
-The most common type of sherds, also the dominant one, has a high percentage of inclusion (25-40%), which is composed mainly by argillite (15-40%), followed by metasandstone (1-10%), sandstone (1-6%), and quartz (1-5%).
-The size of the particles ranges from 500 micron to 1300 micron (=1 mm).
-In general, this type presents a mixture of fine, rounded argillite with a small part of rounded metasandstone and rounded to sub-angular monocrystalline quartz.
+Petrographic analysis for 34 thin sections shows high percentage of inclusions (15-50%), including argillite (15-40%), metasandstone (1-10%),  sandstone fragments (1-6%), quartz (1-5%), and trace amounts of feldspar and slate.
+The size of the particles ranges from 500 micron to 1300 micron.
+In general, the globular vessel farbric presents a mixture of fine, rounded argillite with a small part of rounded metasandstone and rounded to sub-angular monocrystalline quartz.
 The composition matches the mineralogical composition of local raw materials in the Yilan Plain [@Chen2016].
-It seems there is no obvious changes in the inclusions over time.
+There is no substantial changes in the inclusions over time, indicating continuity in pottery fabric composition across the three periods.
 
@@
 
-A total of `r total_no` relativity intact vessels from pre-contact contexts (n = 32), post-European contexts (n = 27), and Chinese contact contexts (n = 14) were selected for Ellipic Fourier analysis.
-Most of them cover every components from rim to bottom.
-Given that the decline of indigenous population in the Chinese period might have impact on the interpretation, we examined the spatial distribution of pots analyzed in this paper.
-The pottery distributed randomly across the sampling area without any clusters could rule out the bias caused by the population dynamics (Figure \@ref(fig:pottery-spatial-pattern)).
+A total of `r total_no` largely intact vessels from pre-contact contexts (n = 32), post-European contexts (n = 27), and Chinese contact contexts (n = 14) were selected for Ellipic Fourier analysis.
+Most of them cover every vessel component from rim to bottom.
 To fully explore the pottery standardization, we also take the metric measurements into account for comparison.
@@
 
-```{r pottery-spatial-pattern, fig.cap= "The spatial distribution of the pottery selected for shape analysis. The quantity is indicated by the color scale."}
-library(sf)
-library(viridis)
-
-# read in the spatial data
-AD_zone <- invisible(st_read(here("analysis", "data", "raw_data", "AD_zone.shp"), quiet = TRUE))
-units_40 <- invisible(read.csv(here("analysis", "data", "raw_data", "kwl-list-of-sampling-squares.csv")))
-
-AD_zone_tidy <-
-  AD_zone %>% 
-  mutate(Pit =paste0("P0",Id)) %>% 
-  inner_join(units_40, by = c("Pit" = "the_sq")) %>% 
-  select(Pit, geometry)
-
-# tidy pottery data
-pottery_count <-
-  pottery_data %>% 
-  select(Site, Context, Pit, Layer, No., phase) %>% 
-  count(Pit, phase)
-
-# join two dataset to get coordinate
-pottery_count_join <-
-  pottery_count %>%
-  mutate(phase = as.factor(phase)) %>% # seems need this line first
-  inner_join(AD_zone_tidy) %>% 
-  filter(!is.na(phase)) %>% 
-  mutate(phase = case_when(
-    phase %in% "pre-e"~ "Before European Contact",
-    phase %in% "post-e" ~ "European Presence",
-    phase %in% "ch-con" ~ "Chinese Presence")) %>% 
-  mutate(phase = factor(phase,
-                        level = c("Before European Contact",
-                                  "European Presence",
-                                  "Chinese Presence")))
-# plot spatial pattern for pottery
-plot_spatial_distribution_pottery_by_phase <- 
-  ggplot() +
-  geom_sf(data = AD_zone_tidy, fill = NA) +
-  geom_sf(data = pottery_count_join, 
-          aes(geometry = geometry, 
-              fill =n), 
-          alpha = 0.9)+
-  facet_wrap(~ phase) +
-  labs(fill = "frequency") +
-  scale_fill_viridis(direction = -1) +
-  theme_minimal() +
-  theme(plot.title = element_text(hjust = 0.5),
-        axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        strip.text.x = element_text(size = 6),
-        legend.text = element_text(size = 6),
-        legend.title = element_text(size = 6),
-        legend.key.size =  unit(3, "mm"))
-
-plot_spatial_distribution_pottery_by_phase
-
-ggsave(plot = plot_spatial_distribution_pottery_by_phase,
-       here("analysis", "figures",
-       "plot-spatial-distribution-pottery-by-period.jpg"),
-       width = 135, 
-       height = 31,
-       dpi = 300,
-       units = "mm")
-```
-
 ## Digitising and analysing by EFA
 
-The scanned pottery drawings were acquired from Bureau of Cultural Affairs in Yilan.
-All drawing presents two-dimensional view of the section of a vessel with indications of metric measurements.
-The scanned drawings were imported into the Inkscape software for outlines tracing to remove additional information such as marks, lines, and measurements on the original drawings in order to obtain clear outline.
-Each traced half cross-section image was duplicated, flipped, and then joined with another one to create a 2D closed outline.
-Geometric morphometric analyses were conducted in the R software (www.rproject.org, Core-Team, 2015) using the functions included in the Momocs, a R package intended to quantify the shape and compare its variation, especially for outline analysis [@Bonhomme2014].
+We used scanned pottery drawings acquired from Bureau of Cultural Affairs in Yilan.
+All drawings present a two-dimensional view of vessel cross-sections with indications of metric measurements.
+The scanned drawings were imported into Inkscape (http://inkscape.org) where outlines were traced manually.
+Where only half cross-section images were available, these were duplicated, flipped, and then joined with another one to create a 2D closed outline for each vessel.
+Geometric morphometric analyses were conducted using the R software [@RCoreTeam] using the functions included in the Momocs, a R package intended to quantify and analyse shapes, especially for outline analysis [@Bonhomme2014].
 The digitised outlines were converted into a list of successive x-y pixel coordinates for elliptic Fourier analysis (EFA), which assesses morphological differences among pottery shapes from three occupation contexts.

### 2020-01-26 analysis/paper/paper.Rmd
starting on results section

@@
 
-A total of `r total_no` largely intact vessels from pre-contact contexts (n = 32), post-European contexts (n = 27), and Chinese contact contexts (n = 14) were selected for Ellipic Fourier analysis.
+A total of `r total_no` largely intact vessels from pre-contact contexts (n = 32), post-European contexts (n = 27), and Chinese contact contexts (n = 14) were selected for ellipic Fourier analysis.
 Most of them cover every vessel component from rim to bottom.
@@
 Where only half cross-section images were available, these were duplicated, flipped, and then joined with another one to create a 2D closed outline for each vessel.
-Geometric morphometric analyses were conducted using the R software [@RCoreTeam] using the functions included in the Momocs, a R package intended to quantify and analyse shapes, especially for outline analysis [@Bonhomme2014].
-The digitised outlines were converted into a list of successive x-y pixel coordinates for elliptic Fourier analysis (EFA), which assesses morphological differences among pottery shapes from three occupation contexts.
-The harmonic coefficients generated by EFA were analysed by principal component analysis (PCA) to illustrate the diversity of the shape data and identify the major patterns of variation through dimensionality reduction.
+Geometric morphometric analyses were conducted using the R software [@Rlanguage2019] using the functions included in the Momocs, a R package intended to quantify and analyse shapes, especially for outline analysis [@Bonhomme2014].
+The digitised outlines were converted into a list of successive x-y pixel coordinates for elliptic Fourier analysis (EFA), to investigate morphological differences among pottery shapes from our three occupation phases.
+The harmonic coefficients generated by EFA were analysed by principal component analysis (PCA) for dimensionality reduction to illustrate the diversity of the shape data and identify major patterns of variation through .
 
-## Statistical testing
+## Statistical analysis
 
-The principal components (PCs) scores were analysed with a multivariate analysis of variance (MANOVA) to test for significant effects of the groups of occupation context on shape variances.
-Finally, we computed coefficients of variation statistic (CVs) among multiple groups of PCs with a significance test for the equality of CVs, which enable us to compare CVs in a statistical way [@Pearson1896].
+The principal component (PC) scores were analysed with a multivariate analysis of variance (MANOVA) test to identify significant differences in shapes between occupation phases.
+We also computed coefficients of variation values (CVs) for the PCs, treating the PCs as shape variables that are more informative than linear dimensions.
 The coefficient of variation is a common and widely-used statistical measure of the spread of a set of measurements of a sample.
-It is defined as the standard deviation divided by the mean in a ratio scale format.
+It is defined as the standard deviation divided by the mean in a ratio scale format:
 
@@
 
-By standardizing standard deviation in each data set, the coefficients of variation statistics allows us to directly compare variation in samples measured with different units or means.
-Although it is a useful method, there is a question that how should we make a decision that the differences among the multiple CVs are significant or not?
-To answer this question, we need convenient methods for testing the equality of $c_v$ values from different samples, which has been described in many publications and here we proposed a significant test for CVs using the R package cvequality [@Marwick2019] in the CRAN repository.
-The package cvequality includes two tests, the "Asymptotic test for the equality of coefficients of variation from k populations" [@Feltz1996] and the "Modified signed-likelihood ratio test (SLRT) for equality of CVs" [@Krishnamoorthy2014].
-@Feltz1996 test is widely cited as an authoritative test for the equality of $c_v$ values that performs better than many approximative tests.
-It is a 'gold standard' test for comparing multiple $c_v$.
+By standardizing standard deviation in each data set, the coefficients of variation values allows us to directly compare variation in samples measured with different units or means.
+We take this as our measurement of standardisation in vessel shape: lower CV values reflect higher standardisation, and thus increased craft specialisation in the community.
+To answer the question of whether CV values for vessels samples across our three occupational phases are significantly different or not, we used the "Asymptotic test for the equality of coefficients of variation from k populations" [@Feltz1996], a significance test for CVs provided by the R package cvequality [@Marwick2019].
 Our implementation is based on the following from @Feltz1996:
@@
 $$M=\sum _{j=1}^m m_j$$
+
 Then the test statistic can be computed with:
+
 $$\begin{aligned} D'A D= \widehat{\tau }_c^{-2}\left( 0.5+\widehat{\tau }_c^2\right) ^{-1}\left[ \sum _{i=1}^km_i\left(\frac{s_i}{\bar{x}_i} \right)^2-\frac{\widehat{\tau }_c^2}{M} \right] \end{aligned}$$
+
 The $D'AD$ value measures how far each sample $c_v$ is from our estimate of the population $c_v$.
 @Feltz1996 note that the $D'AD$ value distributes as a central $\chi^2$ random variable with $k-1$ degrees of freedom, from which a p-value can be computed.
-We also include the @Krishnamoorthy2014 test as a more recent development with lower rates of type I error, better performance with uneven sample numbers, and more power across a range of conditions.
 
@@
 
-To enable re-use of materials and improve reproducibility and transparency [@Marwick2017], the entire R code [@Rlanguage2019] used for all the analysis and visualizations contained in this paper is included in http:XXX.
+To enable re-use of materials and improve reproducibility and transparency [@Marwick2017], the entire R code [@Rlanguage2019] used for all the analysis and visualizations contained in this paper is openly available online at https://doi.org/10.17605/OSF.IO/ABVGF.
 Also in this version-controlled compendium [@Marwick2018] are the raw data for all the visualizations and tests reported here.
@@
 # Results
+
+To investigate the effect of the historically-documented decline of Indigenous population in the Chinese period <!-- add citation on population change during Chinese period --> on pottery production we examined the spatial distribution of pots in our sample in each period.
+Our concern is that the smaller population size during the Chinese period could confound our investigations into craft specialisation beause of reduced numbers of potters and pottery production shrinking a a few locations in the settlement during this time.
+Altough the overall number of vessels is smaller during the Chinese contact period, Figure \@ref(fig:pottery-spatial-pattern) shows that pottery is distributed randomly across the sampling area without any distinctive clusters.
+This suggests that differences in population across our three occupation phases are probably not driving variation in craft specialsation.
+
+```{r pottery-spatial-pattern, fig.cap= "The spatial distribution of the pottery selected for shape analysis. The quantity is indicated by the color scale."}
+library(sf)
+library(viridis)
+
+# read in the spatial data
+AD_zone <- invisible(st_read(here("analysis", "data", "raw_data", "AD_zone.shp"), quiet = TRUE))
+units_40 <- invisible(read.csv(here("analysis", "data", "raw_data", "kwl-list-of-sampling-squares.csv")))
+
+AD_zone_tidy <-
+  AD_zone %>% 
+  mutate(Pit =paste0("P0",Id)) %>% 
+  inner_join(units_40, by = c("Pit" = "the_sq")) %>% 
+  select(Pit, geometry)
+
+# tidy pottery data
+pottery_count <-
+  pottery_data %>% 
+  select(Site, Context, Pit, Layer, No., phase) %>% 
+  count(Pit, phase)
+
+# join two dataset to get coordinate
+pottery_count_join <-
+  pottery_count %>%
+  mutate(phase = as.factor(phase)) %>% # seems need this line first
+  inner_join(AD_zone_tidy) %>% 
+  filter(!is.na(phase)) %>% 
+  mutate(phase = case_when(
+    phase %in% "pre-e"~ "Before European Contact",
+    phase %in% "post-e" ~ "European Presence",
+    phase %in% "ch-con" ~ "Chinese Presence")) %>% 
+  mutate(phase = factor(phase,
+                        level = c("Before European Contact",
+                                  "European Presence",
+                                  "Chinese Presence")))
+# plot spatial pattern for pottery
+plot_spatial_distribution_pottery_by_phase <- 
+  ggplot() +
+  geom_sf(data = AD_zone_tidy, fill = NA) +
+  geom_sf(data = pottery_count_join, 
+          aes(geometry = geometry, 
+              fill =n), 
+          alpha = 0.9)+
+  facet_wrap(~ phase) +
+  labs(fill = "frequency") +
+  scale_fill_viridis(direction = -1) +
+  theme_minimal() +
+  theme(plot.title = element_text(hjust = 0.5),
+        axis.text.x = element_blank(),
+        axis.text.y = element_blank(),
+        strip.text.x = element_text(size = 6),
+        legend.text = element_text(size = 6),
+        legend.title = element_text(size = 6),
+        legend.key.size =  unit(3, "mm"))
+
+plot_spatial_distribution_pottery_by_phase
+
+ggsave(plot = plot_spatial_distribution_pottery_by_phase,
+       here("analysis", "figures",
+            "plot-spatial-distribution-pottery-by-period.jpg"),
+       width = 135, 
+       height = 31,
+       dpi = 300,
+       units = "mm")
+```
 
@@
 metric_cvs_with_phase_tidy <-
-  metric_cvs_with_phase%>% 
+  metric_cvs_with_phase %>% 
   mutate_at(c("pre-e", "post-e", "ch-con", "MSLRT", "p_value"), round, 4) %>% 

### 2020-01-26 analysis/paper/paper.Rmd
editing results: basic shape description and MANOVA

@@
 
-Table \@ref(tab:metricCVstest) shows the results of Modified signed-likelihood ratio test (MSLRT) for equality of CVs for for six measurements and two ratio values.
-There is a significant difference (p-value less than 0.01) between CV values for two metric measurements across phases, the rim diameter and the neck diameter.
-The CV values show that the variation of rim diameter is small (6.8%) before European contact, increases after the European presence (9.6%), and decreases after the Chinese presence (6.9%).
-For the neck diameter, the CV values indicate smaller variation after the European period (7.7%) and the Chinese period (7.1%) compared to pre-European period (10.6%).
-
-```{r metricCVstest}
-metric_cvs_with_phase <- readr::read_csv(here::here("analysis", "data", "raw_data", "Kwl_p_metrics_cv_test.csv"))
-
-metric_cvs_with_phase_tidy <-
-  metric_cvs_with_phase %>% 
-  mutate_at(c("pre-e", "post-e", "ch-con", "MSLRT", "p_value"), round, 4) %>% 
-  rename(`pre-European` = "pre-e", `after European` = "post-e", Chinese = "ch-con")
-
-knitr::kable(metric_cvs_with_phase_tidy,
-             caption = "Coefficient of Variation test for metric attributes by phases")
-```
-
 ```{r jpg-coordinates, eval=FALSE}
@@
 
+<!-- basic description of shape -->
+
 The elliptic Fourier coefficients of `r total_no` pottery from three phases were calculated.
-Reliable pottery outline was captured by 13 harmonics that gather 99 % of the total harmonic power.
+Reliable pottery outlines were captured by 13 harmonics that gather 99% of the total harmonic power.
 The average shape of vessels from each phase was visualized by computing the mean of the standardized Fourier coefficients within each phase group (Figure \@ref(fig:EFA-analysis)).
-Figure \@ref(fig:thin-plate-splines) shows the shape changes described using thin-plate spline warping for paired periods, pre and post European periods, and post-European and Chinese period in a time sequence.
+Figure \@ref(fig:thin-plate-splines) shows the shape changes described using thin-plate spline warping for paired periods, pre- and post-European periods, and post-European and Chinese periods.
 
@@
 
-The elliptic Fourier coefficients are examined by principal components analysis to explore the shape variations across three phases.
+<!-- further description of shape and differences between phases -->
+
+The elliptic Fourier coefficients were examined by Principal Components Analysis to explore the shape variations across three phases.
 The first two principal components (PCs) explain `r vari_two`% of the total variances, of which `r pc1`% is explained by the first principle component.
-With the third component, the first three principal components explain `r vari_three`% of the total variances.
+With the third component, the first three principal components explain `r vari_three`% of the total variance.
 Figure \@ref(fig:contribution-PCs) shows the shape variation associated with the first three principal components.
-From the negative to the positive scores, the PC1 represents the height of the vessels from tall to short and the roundness of the body from round to oval-shaped.
-The PC2 relates to the neck and mouth constriction from narrow to wide.
-The PC3 explains a smaller portion of the variance (`r pc3`%), which relates to the degree of the flare of the neck from curved to straight shape.
-The first two components account for most of the variance in relation to three phases were represented in Figure \@ref(fig:first-two-PCs-byplot).
-The result reflects a large overlap of three groupings especially for shapes in the pre-European and post-European periods that might show some extent of the similarity.
-However, the spread of shape distribution indications a wider variation in shapes in the pre-European and post-European periods compared to those in the Chinese period along both PC1 and PC2 axes.
-In other words, there is a decreasing in shape variance in the Chinese period that shares the features of shorter height and narrower mouth.
+PC1 captures the height of the vessels, from tall to short, and the roundness of the body from round to oval-shaped.
+PC2 relates to the neck and mouth constriction, from narrow to wide.
+PC3 explains a smaller portion of the variance (`r pc3`%), which relates to the degree of the flare of the neck, from curved to straight shape.
+The first two components account for most of the variance in relation to three phases (Figure \@ref(fig:first-two-PCs-byplot)).
+The results reflect a large overlap in shapes from three occupations phases, especially for shapes in the pre-European and post-European periods.
+However, the spread of shape distribution indicates a wider variation in shapes in the pre-European and post-European periods compared to those in the Chinese period along both PC1 and PC2 axes.
+In other words, we find a decrease in shape variance in the Chinese period that is evident in the shorter height and narrower mouth of vessels used in that period.
 
@@
 # PCA byplot
-plot(PCA(op), 2, pos.shp="range")
+plot_PCA(op.p, ~phase, chullfilled = TRUE)
 ```
@@
 
-manv <- rrtools:::quietly ( MANOVA_PW(op.p, "phase") ) #? how to hide the process, echo = FALSE?
+manv <- rrtools:::quietly ( MANOVA_PW(op.p, "phase") ) 
 su_manv <- as.data.frame(manv$summary)
-# how to reduce the number in the table to only 4 digits, tried mutate with signif but doesn't work?
-su_manv <- 
+
+su_manv_tab <- 
   su_manv %>% 
-  mutate_all(round, 4)
+  mutate_all(round, 4) %>% 
+  select(-Df, -`num Df`) %>% 
+  rename(`Pillai's trace` = Pillai,
+         `Approximate F value` = `approx F`,
+         `degrees of freedom` = `den Df`)
 
-pval_chi_pre <- signif(unname(su_manv[,'Pr(>F)'][2]), 4)
-pval_post_pre <- signif(unname(su_manv[,'Pr(>F)'][3]), 4)
+pval_chi_pre <- signif(unname(su_manv_tab[,'Pr(>F)'][2]), 4)
+pval_post_pre <- signif(unname(su_manv_tab[,'Pr(>F)'][3]), 4)
 
-knitr::kable(su_manv,
-             caption = "The statistics summary of MANOVA test for the PC scores")
+knitr::kable(su_manv_tab,
+             caption = "The statistics summary of MANOVA test for the PC scores. `Pr(>F)` is the p-value associated with the F statistic of the effect and test statistic. ")
 ```
 
-To test differences in the distribution of shape variables indicated by the PC scores, we use multivariate analysis of variance (MANOVA) to compare a pairwise combination between every sample means across three phases.
-The results in Table \@ref(tab:MANOVA-table) show significant differences in shape between Pre-European and Post-European (p = $`r pval_post_pre`$) and Pre-European and Chinese contact shapes (p = $`r pval_chi_pre`$), which corresponds to the previous observation of the different distribution in shapes in Chinese period according to PCA byplot.
-In addition, although there is a considerable overlap of shape variables between Pre-European and Post-European, their PC scores are significantly different from each other.
-There are not much differences between the shapes in the Post-European and Chinese contact periods.
+<!-- statistical test of difference in shape decsribed in previous paragraph -->
 
-```{r GAMM, eval = FALSE}
-#----------------------------------------------------
-# GAMMS from https://www-nature-com.offcampus.lib.washington.edu/articles/s41598-018-20122-9
+To test for differences in the distributions of shape variables indicated by the PC scores and biplot, we use the multivariate analysis of variance (MANOVA) test to compare pairwise combinations across our three occupation phases.
+Table \@ref(tab:MANOVA-table) shows significant differences in shape between Pre-European and Post-European phases (p = $`r pval_post_pre`$), and Pre-European and Chinese phases (p = $`r pval_chi_pre`$).
+Although there is considerable overlap of shape variables between the Pre-European and Post-European phases, their PC scores are significantly different from each other.
+There is little difference between the shapes in the Post-European and Chinese contact periods.
 
-# Generalised additive mixed models (GAMMs) were used to explain shape variance with respect to mean environmental parameters and shell size, and to compare between individual shape features (PCs).
-A GAMM is a generalised linear mixed model with a linear predictor involving a sum of smooth functions of covariates.
-The model allowed for flexible specification of the dependence of the response on the covariates by defining regression splines (smoothing functions) and estimating their optimal degree of smoothness, rather than calculating parametric relationships
+<!-- CV equality testing -->
 
-# metric data
-ggplot(pottery_data,
-       aes(phase,
-           as.numeric(Thick1_Neck))) +
-  geom_boxplot()
+Table \@ref(tab:metricCVstest) shows the results of Modified signed-likelihood ratio test (MSLRT) for equality of CVs for for six measurements and two ratio values.
+There is a significant difference (p-value less than 0.01) between CV values for two metric measurements across phases, the rim diameter and the neck diameter.
+The CV values show that the variation of rim diameter is small (6.8%) before European contact, increases after the European presence (9.6%), and decreases after the Chinese presence (6.9%).
+For the neck diameter, the CV values indicate smaller variation after the European period (7.7%) and the Chinese period (7.1%) compared to pre-European period (10.6%).
 
-pottery_data$thick_neck1 <- as.numeric(pottery_data$Thick1_Neck)
-pottery_data$thick_body1 <- as.numeric(pottery_data$Thick1_Body)
-pottery_data$height <- as.numeric(pottery_data$Hight)
+```{r metricCVstest}
+metric_cvs_with_phase <- readr::read_csv(here::here("analysis", "data", "raw_data", "Kwl_p_metrics_cv_test.csv"))
 
-plot(op.p$x[ , 2], pottery_data$thick_body1)
+metric_cvs_with_phase_tidy <-
+  metric_cvs_with_phase %>% 
+  mutate_at(c("pre-e", "post-e", "ch-con", "MSLRT", "p_value"), round, 4) %>% 
+  rename(`pre-European` = "pre-e", `after European` = "post-e", Chinese = "ch-con")
 
-gams_input <- data.frame(thick_body1 = pottery_data$thick_body1 , 
-                         height = pottery_data$height, 
-                         shape_pc1 = op.p$x[ , 1])
-
-# Bayesian Regression Models using 'Stan'
-library(brms)
-m2 <- brm(bf(shape_pc1 ~ s(thick_body1)),
-          data = gams_input, family = gaussian(), cores = 4, seed = 17,
-          iter = 4000, warmup = 1000, thin = 10, refresh = 0,
-          control = list(adapt_delta = 0.99))
-summary(m2) # so many zeros... not very promising 
-# plot marginal effects for each predictor
-plot(marginal_effects(m2), ask = FALSE)
-loo(m2)
-pp_check(m2)
+knitr::kable(metric_cvs_with_phase_tidy,
+             caption = "Coefficient of Variation test for metric attributes by phases")
 ```

### 2020-01-27 analysis/paper/paper.Rmd
trying to understand the text on CV equality testing in results section

@@
 ```{r first-two-PCs-byplot, fig.cap= "Pottery shape distribution by each occupation phase according to the first two PCs"}
-# the pots groups by phase, and their shape distribution according to 
-# the first two principle components of the EFA 
-# png(here::here("analysis/figures/pca-biplot.png"), res = 300, w = 1500, h = 1500)
-# plot(op.p, 2, 
-#      morpho = TRUE, 
-#      ellipses = TRUE,
-#      cex = 1, 
-#      cex.labelsgroups = 1.5)
-# dev.off()
 
@@
 ```
+
+<!-- statistical test of difference in shape decsribed in previous paragraph -->
+
+To test for differences in the distributions of shape variables indicated by the PC scores shown in Figure \ref(@fig:first-two-PCs-byplot)), we used a multivariate analysis of variance (MANOVA) test to compare pairwise combinations across our three occupation phases.
+Table \@ref(tab:MANOVA-table) shows significant differences in shape between Pre-European and Post-European phases (p = $`r pval_post_pre`$), and Pre-European and Chinese phases (p = $`r pval_chi_pre`$).
+Although there is considerable overlap of shape variables between the Pre-European and Post-European phases, their PC scores are significantly different from each other.
+There is little difference between vessel shapes in the Post-European and Chinese contact periods.
 
@@
 
-<!-- statistical test of difference in shape decsribed in previous paragraph -->
-
-To test for differences in the distributions of shape variables indicated by the PC scores and biplot, we use the multivariate analysis of variance (MANOVA) test to compare pairwise combinations across our three occupation phases.
-Table \@ref(tab:MANOVA-table) shows significant differences in shape between Pre-European and Post-European phases (p = $`r pval_post_pre`$), and Pre-European and Chinese phases (p = $`r pval_chi_pre`$).
-Although there is considerable overlap of shape variables between the Pre-European and Post-European phases, their PC scores are significantly different from each other.
-There is little difference between the shapes in the Post-European and Chinese contact periods.
-
 <!-- CV equality testing -->
@@
 For the neck diameter, the CV values indicate smaller variation after the European period (7.7%) and the Chinese period (7.1%) compared to pre-European period (10.6%).
+<!-- these numeric values in this paragraph should come from inline code, not copy-pasted in like this -->
+
+<!-- You need to put the code that computes the MSLRT results right in here, in the Rmd -->
 

### 2020-01-27 analysis/paper/paper.Rmd
editing para on cv equality test on shapes

@@
 <!-- CV equality testing -->
+To compare pottery shape standardization across the three phases we  investigate the distributions of the first three PC scores, taking the PC scores as proxy variables for vessel shape (Figure \@ref(fig:PCs-distribution)).
+The first PC, vessel height, shows higher variation in the pre-European period compared to the Chinese period.
+That indicates that shape standardization was higher in the Chinese period compared to the pre-European period.
+The second PC also presents a similar pattern of higher shape standardization in the Chinese period compared to the other two phases.
+To see whether the differences in the distribution of PCs between any two phases are substantive or due to chance, we used a modified signed-likelihood ratio test to assess the equality of CVs.
+P-values for this significance test of CVs for PC1 and PC2 show significant differences in the standardization of vessel shapes across periods, especially between Chinese contact with either pre-European or post-European(Figure \@ref(fig:CV-test-PCs-plot), Table \@ref(tab:CV-test-PCs-table)).
+A significant difference is also detected between the pre-European and post-European periods.
 
 Table \@ref(tab:metricCVstest) shows the results of Modified signed-likelihood ratio test (MSLRT) for equality of CVs for for six measurements and two ratio values.
-There is a significant difference (p-value less than 0.01) between CV values for two metric measurements across phases, the rim diameter and the neck diameter.
+There is a significant difference between CV values for two metric measurements across phases, the rim diameter and the neck diameter.
 The CV values show that the variation of rim diameter is small (6.8%)  before European contact, increases after the European presence (9.6%), and decreases after the Chinese presence (6.9%).
@@
 
-The variation of pottery shape for each phase is explored by the distributions of the first three PC scores to compare the extent of standardization across phases.
-Figure \@ref(fig:PCs-distribution) suggests variations of PC scores for each vessel by phase.
-The first PC, the hight of the vessels, shows more variations in the pre-European period compared to the Chinese period.
-That indicates standardization appears strong in the Chinese period but weak in the pre-European period.
-The second PC also presents a similar pattern that stronger standardization of shape in the Chinese period compared to the other two phases.
-To see whether the differences in the distribution of PCs between any two phases is statistically significant, coefficients of variation statistic (CVs) among phase groups of PCs is examined with a modified signed-likelihood ratio test for the equality of CVs.
-P-values for the significant test of CVs of PC1 and PC2 show significant differences (P-value less than 0.01) in the standardization of vessel shape across some periods, especially between Chinese contact with either pre-European or post-European(Figure \@ref(fig:CV-test-PCs-plot), Table \@ref(tab:CV-test-PCs-table)).
-Significant difference is also detected between pre-European and post-European (P-value less than 0.05).
-
 # Discussion

### 2020-01-27 analysis/paper/paper.Rmd
remove unused var.test code

@@
 knitr::kable(su_manv_tab,
-             caption = "The statistics summary of MANOVA test for the PC scores. `Pr(>F)` is the p-value associated with the F statistic of the effect and test statistic. ")
+             caption = "Summary statistics for the MANOVA test on the PC scores. `Pr(>F)` is the p-value associated with the F statistic of the effect and test statistic. ")
 ```
@@
        h = 10, w = 5)
-```
-
-```{r variance-test-PCs}
-# we don't use CV because that is not good for measurements with no absolute zero
-xx <- 
-  pc_plot %>% 
-  group_by(pcn, phase) %>% 
-  summarise(variance = var(pc))
-
-# statistical test to compare the variances of two samples
-pc_plot_tests <- 
-  pc_plot %>% 
-  group_by_at(vars(-pc)) %>% 
-  mutate(i = row_number()) %>% 
-  ungroup() %>% 
-  mutate(phase = as.character(phase),
-         pcn = as.character(pcn)) %>% 
-  spread(phase, pc) %>% 
-  select(-i) %>% 
-  nest_legacy(-pcn) %>% 
-  mutate(var_test = purrr::map(data, ~broom::tidy(var.test(.x$`Post-European` ,
-                                            .x$`Pre-European`)))) %>% 
-  unnest(var_test)
 ```

### 2020-01-27 analysis/paper/paper.Rmd
remove CV test on metric data (for now? for good? let's see)

@@
 
-Table \@ref(tab:metricCVstest) shows the results of Modified signed-likelihood ratio test (MSLRT) for equality of CVs for for six measurements and two ratio values.
-There is a significant difference between CV values for two metric measurements across phases, the rim diameter and the neck diameter.
-The CV values show that the variation of rim diameter is small (6.8%)  before European contact, increases after the European presence (9.6%), and decreases after the Chinese presence (6.9%).
-For the neck diameter, the CV values indicate smaller variation after the European period (7.7%) and the Chinese period (7.1%) compared to pre-European period (10.6%).
-<!-- these numeric values in this paragraph should come from inline code, not copy-pasted in like this -->
-
-<!-- You need to put the code that computes the MSLRT results right in here, in the Rmd -->
-
-```{r metricCVstest}
-metric_cvs_with_phase <- readr::read_csv(here::here("analysis", "data", "raw_data", "Kwl_p_metrics_cv_test.csv"))
-
-metric_cvs_with_phase_tidy <-
-  metric_cvs_with_phase %>% 
-  mutate_at(c("pre-e", "post-e", "ch-con", "MSLRT", "p_value"), round, 4) %>% 
-  rename(`pre-European` = "pre-e", `after European` = "post-e", Chinese = "ch-con")
-
-knitr::kable(metric_cvs_with_phase_tidy,
-             caption = "Coefficient of Variation test for metric attributes by phases")
-```
-
 ```{r PCs-distribution, fig.cap= "The distribution of PC scores by phases"}

### 2020-01-27 analysis/paper/paper.Rmd
remove CV-test-PCs-plot "P-values of the CV equality test of PC1 and PC2 between phases, red color indicates p-value is less than 0.05"

@@
 
-```{r CV-test-PCs-plot, fig.cap= "P-values of the CV equality test of PC1 and PC2 between phases, red color indicates p-value is less than 0.05"}
-
-add_line_format <- function(x, ...){
-  str_replace(x, " vs", "\nvs")
-}
-
-PC1_PC2_cv_tests %>%
-  mutate(colour = if_else(p_value < 0.05, "red", "black")) %>% 
-  ggplot(aes(str_wrap(phases, 40),
-             p_value,
-             colour = I(colour),
-             shape = PC)) +
-  geom_point(size = 8,  position = position_dodge(width = 0.5)) +
-  theme_minimal(base_size = 12) +
-  scale_x_discrete(labels = add_line_format(unique(sort(PC1_PC2_cv_tests$phases)))) +
-  coord_cartesian(clip = 'off') +
-  xlab("") +
-  ylab("CV equality test p-values")
-
-ggsave(here("analysis/figures/cv-test-on-pcs.png"), h = 3, w = 12)
-```
-
 # Discussion

### 2020-01-27 analysis/paper/paper.Rmd
add cv test code for metric data to rmd, and revise the para

--- 2020-01-27 analysis/paper/paper.Rmd
@@
 
-The development of craft specialization is sometimes associated with social changes since it reflects changes in production systems that could inform on the social, economic, or political conditions of a society.
-We define specialization as "differential participation in specific economic activities", in which a particular good is produced by fewer producers compared to the total population who consume it [@Costin1991; @Costin2001, pp.
+The development of craft specialization is usually associated with the social changes since it reflects production systems that could inform social, economic, or political conditions of a society.
+Specialization is defined as "differential participation in specific economic activities", in which a particular good is produced by fewer producers compared to the total population who consume it [@Costin1991; @Costin2001, pp.
 43].
-The effects of the emergence of specialization in past societies on their social structures and social processes is an important question in archaeological investigations of small scale societies in the Late Holocene.
-Considered through a political lens, craft specialization may be a strategy used by elites to control production systems to maintain their power and authority [@Costin2001].
-This relationship between craft specialization and social differentiation has been broadly discussed, and many cases show that increasingly specialized production can be a useful indicator of increasing socio-political complexity [@Hirshman2010; @Junker1999].
-That said, more economic focus considers production, exchange, and consumption as a whole to explain specialization as part of a strategy to accommodate uneven distributions of resources [@Costin2001; @Arnold1994].
-Craft specialisation may appear as a reaction to changes in subsistence practices [@Stark1995economic], or as a result of long-distance trade and inter-regional exchange [@Alizadeh2018].
-Present-day investigation of ceramic specialization in the Jodhpur region of India suggests that the adoption of standardized products across cultural boundaries is a reaction to a collapse of previous economic system, indicating a correlation with major socio-economic changes [@Roux2015].
-However, it is important to note that explanations for the emergence of specialized production in past societies might not be so straightforward or singular.
-@Acabado2018's case study in lowland northern Philippines shows that specialization is not necessarily accompanied with social differentiation based on a low degree of ceramic specialization in Ifugao ranked societies during the time of Spanish colonization.
-The development of specialization could be dependent on unique historical factors, where the conditions and contexts under which specialists organized must be taken into account when making inferences about changes in social organization [@Roux2015; @Costin2001].
+The implication of the emergence of specialization for social structure or social process has been a theme in archaeological investigations.
+From the perspective of the political approaches, craft specialization specifically attached production, is a strategy used by elites to control over production systems to maintain power and authority [@Costin2001].
+The relationship between craft specialization and social differentiation is broadly discussed, and many cases show that specialized production is a useful indicator of increasing socio-political complexity [@Hirshman2010; @Junker1999].
+One the other hand, economic models consider production, exchange, and consumption as a whole to explain specialization as a subsistence strategy to accommodate uneven distribute of resources [@Costin2001; @Arnold1994], a reaction to the changes in subsistence practices [@Stark1995economic], or an outcome result from by long-distance trade and inter-regional exchange [@Alizadeh2018].
+An present-day investigation of ceramic specialization in the Jodhpur region of India suggests that the adoption of standardized products across cultural boundaries is a reaction to a collapse of previous economic system, indicating the correlation with major socio-economic changes [@Roux2015].
+However, it is important to note that the explanation for the emergence of specialized production might not be straightforward or exclusive.
+@Acabado2018's case study in lowland northern Philippines shows that specialization is not necessarily accompanied with social differentiation based on the result of a low degree of ceramic specialization in Ifugao ranked societies during the Spanish colonization.
+The development of specialization could be history dependent, where condition and contexts under which specialists organized are important to make an inference for changes in social organization [@Roux2015; @Costin2001].
 
-The emergence of social inequality in the context of cross-cultural interaction, especially in a colonial situation, has been observed in many parts of world, for example, when foreign trade goods are introduced to local Indigenous societies.
-The monopolization of long-distance trade goods has caused substantial transformations of Indigenous economic, cultural, and socio-political systems [@Dietler2005; @Dietler1997; @Junker1993; @Silliman2005].
-Pericolonial archaeology is the study of these indirect effects of colonialism, investigating areas where European colonial rule was limited and their conquests were short-lived and unsuccessful, but their colonial activities had economic and political impacts on Indigenous peoples living on the periphery of colonial control and influence [@Trabert2017; @Acabado2017].
-@Acabado2017's study of Ifugao society in the  Philippines highland suggests economic and political intensification during the Spanish presence as the response of Indigenous peoples to the Spanish cooptation.
-Indigenous societies might not simply passively accept colonial rule, but instead actively negotiate with the colonist, and accommodate or resist foreign intrusion through their daily cultural practices, such as their consumption patterns of foreign goods [@Dietler2015; @Given2004; @Mullins2011; @Scaramelli2005; @Silliman2001; @TorrenceandClarke2000].
-In this paper we present a case study from Northeastern Taiwan as an example of these indirect impacts of foreign occupations on local Indigenous societies.
-Our aim is to investigate if there was increasing ceramic specialization resulting from Indigenous interaction with Europeans in the 17th century or Chinese in the 19th century, two major foreign influences in early historical Taiwan, that might indicate social changes in local indigenous societies.
-Previous reports of archaeological evidence in northeastern Taiwan descrbes increases in the use of prestige goods, such as trade ornaments, and uneven artefact distribution patterns in domestic areas and burial contexts at the time when the Indigenous communities encountered the Spanish and the Dutch in the 17th century.
-Here we show that in addition to the differential distribution of trade ornaments post-European contact, the locally-made ceramics become highly consistent in their form and shape, compared to other pottery throughout northeastern Taiwan.
-This shift to higher consistency hints at the possibility of craft specialization in pottery manufacture.
+The emergence of social inequality in the context of cross-cultural interaction especially in a colonial situation was observed in many parts of world when the introduction of foreign trade goods to local indigenous societies.
+The monopolization of long-distance trade goods usually cause substantial transformations of indigenous economic, cultural, and socio-political systems [@Dietler2005; @Dietler1997; @Junker1993; @Silliman2005].
+Recently indirect effects of colonialism or pericolonial archaeology were address when investigating areas where European colonial rule was limited or their conquests were unsuccessful, but their colonial activities had economically and politically impacts on indigenous peoples in the periphery of colonial control [@Trabert2017; @Acabado2017].
+For example, @Acabado2017's study on Ifugao society in the highland Philippines suggests economic and political intensification during the Spanish presence as the response of indigenous peoples to the Spanish cooptation.
+Indigenous societies might not just passively accept the colonial rules, but instead actively negotiated with the colonist, and accommodate or resist the foreign intrusion through their daily cultural practices, such as consumption patterns of foreign goods [@Dietler2015; @Given2004; @Mullins2011; @Scaramelli2005; @Silliman2001; @TorrenceandClarke2000].
+Northeastern Taiwan is an ideal example to explore the indirect impacts of the foreign occupations on local indigenous societies due to its isolated location.
+The archaeological evidence in northeastern Taiwan show evidence of increasing use of prestige goods such as trade ornaments and uneven distribution pattern in domestic area and burial contexts when the society encountered by the Spanish and the Dutch in the 17th century.
+In addition to the differential distribution of trade ornaments, the locally made ceramic present a high consistency in form and shape compared to other pottery throughout northeastern Taiwan that hints the possibility of pottery specialization.
 
-To identify craft specialization in the archaeological record, variation in the standardization of a specialized product is often used as an indicator.
-This is based on the assumption that specialized mass production will lead to uniformity of the product due to routinization, higher skills, and fewer producers involved in production [@Stark1995; @Costin1991; @Costin2001].
-For investigating ceramic standardization, several measurements have been discussed and used, such as metric, compositional, and technological approaches [@Arnold2000; @Boness2015; @Rice1991; @Blackman1993; @Costin1991; @Tite1999; @Roux2015].
-Among those variables, metric measurements are most widely applied to archaeological assemblages because of the ease of collecting these data.
+To identify craft specialization, the variation of standardization of specialized product is commonly used as an indicator based on the assumption that specialized mass production will lead to uniformity of the product due to routinization, increased skills, and fewer number of producers involved in production [@Stark1995; @Costin1991; @Costin2001].
+For ceramic standardization, several measurements have been proposed and discussed, such as metric, compositional, and technological approaches [@Arnold2000; @Boness2015; @Rice1991; @Blackman1993; @Costin1991; @Tite1999; @Roux2015].
+Among those variables, metric measurement is widely applied to archaeological assemblages.
 By comparing variations for sets of data using coefficient of variation coupled with statistical testing, the degree of standardization can be identified according to values [@Eerkens2001; @Roux2003; @Stark1995].
-@Roux2018's ethnographic study in the Jodhpur region of India suggests the number of artisans can be statistically assessed by comparing coefficients of variation of linear dimensions of pots, and differences between artisans can be detected at both intra-individual level and inter-individual level in a region.
-However, because pottery vessels are typically have curved shapes, linear measurements have limited sensitivity to many kinds of shape variations.
-Thus, to capture subtle shape variations that might be relevant to standardization, a method that is senstive to non-linear aspects of shape variation is necessary.
-Here we analyse pottery shape using geometric morphometric methods, a new approach for measuring the degree of standardization because it takes a statistical summary of the overall shape as variable for comparison [@Slice2007].
-Using geometric morphometry, we study shape and standardization of locally made ceramics at Kiwulan (1350-1950 AD) [@Chen2007], a large multi-component archaeological site in northeastern Taiwan, to identify changes in ceramic production resulting from cross-cultural interaction contexts.
+@Roux2018's ethnographic study in the Jodhpur region of India suggests the number of artisans can be assessed by comparing coefficient of variation and the differences between artisans can be detected at both intra-individual level and inter-individual level in a region.
+However, linear measurements might be limited for identifying differences in variations because they could be insensitive to subtle variations resulting from changes in assemblages.
+Shape analysis based on geometric morphometrics method are new approaches for measuring the degree of standardization since it takes the overall shape as variable for comparison [@Slice2007].
+Using geometric morphometrics approaches, we study shape and standardization of locally made ceramics at Kiwulan (1350-1950 AD) [@Chen2007], a large multi-component archaeological site in northeastern Taiwan, to identify changes in ceramic production in cross-cultural interaction contexts.
+The aim of this paper is to investigate if there is increasing ceramic specialization resulting from interaction with the Europeans in the 17th century or the Chinese in the 19th century, two major foreign influences in early historical Taiwan, that might indicate social changes in local indigenous societies.
 
-Our hypothesis is that if foreign influences had impacts on the emergence of social inequality in local Indigenous society due to monopolies of trade arising among a small number of Indigenous people, then we expect to see changes in social organization from a more corporate to a more network organization [@Feinman2000].
-We expect this corporate-network shift to manifest as increased specialization pottery manufacture, indicated archaeologically by icreased standardisation of pottery production at Kiwulan.
-The emergence of craft specialization, here pottery specialization, is often related to changes in social organization towards a society with increasing status inequality.
-In this case, if the competition among Indigenous individual for foreign resources and trade partnership with European or Chinese colonizers gradually lead to increasing social inequality, then we expect the local ceramics will show more homogeneous shapes after contact due to the emergence of craft specialization caused by rgreater economic and social control by a small group of individuals at Kiwulan.
-Our results are important for understanding the indirect influences of  European colonists on local Indigenous societies, which remains unclear in East Asia where colonial power had limited success compared to other places in Southeast Asia.
-Our findings improve our understanding of the relationship between the use of prestige goods, the degree of ceramic specialization, and the influence of foreign colonizers in a pericolonial context [cf.
+Our hypothesis is that if foreign influences had impacts on the emergence of social inequality in local indigenous society due to the monopolies for trade between a small number of indigenous people, then we expect to see the changes in social organization from a more corporate to a more network organization that would be reflected by pottery production at Kiwulan [@Feinman2000].
+The emergence of craft specialization, here pottery specialization, are usually related to changes in social organization towards a society with increasing status inequality.
+In this case, if the competitions for foreign resources and being trade partners of European or Chinese colonizer among individuals gradually lead to increasing social inequality, then we expect the local ceramic will show more homogeneous features after contact due to the emergence of craft specialization caused by control of small group of individuals.
+This study is important to understand the indirect influences of the European colonists on the local indigenous societies, which remains unclear in East Asia where the colonial power was not successful compared to other places in Southeast Asia.
+Also, it would help to understand the relationship among the use of prestige goods, the degree of ceramic specialization, and the influence of foreign colonizers in a pericolonial context [cf.
 @Acabado2017].
 
-## Geometric Morphometry
+## Geometric Morphometric approaches
 
-Geometric morphometric methods (GMM) has been increasingly applied in archaeology for shape analysis, which explores morphological variability and similarity of archaeological materials to address the questions of anthropological interest.
-Geometric morphometrics differs from traditional linear measurements through its use of Cartesian coordinates of morphological structures to define shapes [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007].
-Landmarks, curves or outlines of objects can be represented by coordinates in terms of their unique point locations with respect to numerical values on coordinate axes.
-There are two common morphometric methods: landmark approaches and outline approaches [@Adams2004].
-Landmark approaches assign a set of landmarks or semilandmarks onto objects as reference points that can be specified on a coordinate system as xy coordinates for two dimensions, or xyz coordinates for three dimensions.
-In a typical analysis, starting with raw landmarks, a sample of objects are translated to a common centroid, rescaled into the same size, and rotated until the summed squared distances between corresponding landmarks are minimized using generalized Procrustes analysis (GPA) [@Bookstein1991].
-GPA is a method that superimposes sets of landmark configurations into a common coordinate system, where superimposed landmark coordinates are used as shape variables for further multivariate statistical analyses [@Slice2007].
-Landmark-based morphometrics are widely applied to archaeological objects with obvious morphological features that can provide unambiguous reference points for landmark placement, such as projectile point tips or biological measurements points indicating osteological features [@Birch2019; @Buchanan2019; @Lycett2013; @Cardillo2010; @Haruda2019; @Meloro2015].
+Geometric morphometric methods (GMM) has been increasingly applied in archaeology for shape analysis, which explores morphological variability and similarity of archaeological materials to address the questions of anthropological interests.
+Different from linear measurement approaches that capture shapes by length, width or ratios of objects, geometric morphometrics methods use Cartesian coordinates of morphological structures to define shapes [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007].
+Landmarks, curves or outlines of objects can be represented by coordinates in terms of their unique point locations with respect to numerical values of coordinate axes.
+According to different focuses and nature of original shape data, there are two common morphometric methods, landmark approaches and outline approaches [@Adams2004].
+Landmark approaches assign a set of landmarks or semilandmarks onto objects as reference points that can be specified on a coordinate system as x, y coordinates for two dimensions, or x, y, z coordinates for three dimensions.
+Based on distances between raw landmarks, objects are translated to a common centroid, rescaled into the same size, and rotated until the summed squared distances between corresponding landmarks are minimized using generalized Procrustes analysis (GPA) [@Bookstein1991].
+GPA is a method that superimposes sets of landmark configurations into a common coordinate system, where superimposed landmark coordinates are used as shape variables for further multivariate statistics analyses [@Slice2007].
+Landmark-based morphometrics are widely applied to archaeological objects with a obvious morphological signature that can provide ideal reference points for landmark placement, such as projectile point tips or biologically measurements points indicating osteological features [@Birch2019; @Buchanan2019; @Lycett2013; @Cardillo2010; @Haruda2019; @Meloro2015].
 
-However, one limitation of landmark approaches is that landmarks may not be able to capture the shape differences of a morphological structure where the curving outlines between landmarks contains crucial variation.
-Furthermore, landmark points may be unavailable for structures that are mostly or entirely curves.
-In those cases, outline approaches are more suitable for analyzing the overall shape of an object.
-One outline approache is the semi-landmarks method, also called sliding landmarks, which assigns points along the curve between two landmarks at defined intervals [@Bookstein1997; @Lawing2010].
+However, one limitation of landmark approaches is that landmarks may not be able to capture the shape differences of a morphological structure where the curving outlines between landmarks are crucial for variation.
+Furthermore, representative and reliable landmark points may be unavailable for complex or rounded shaped structures.
+In those cases, outline approaches are better way for analyzing the overall shape of an object.
+One of the outline approaches is the semi-landmarks method, also called sliding landmarks, which assigns points along the curve between two landmarks at defined intervals [@Bookstein1997; @Lawing2010].
 Those semi-landmarks are allowed to slide along the curve to remove the effect of the arbitrary landmark spacing by minimizing either Procrustes distance or bending energy [@Bookstein1997; @Slice2007; @Gunz2013].
-Another approach that is commonly applied to two-dimensional closed shapes is elliptic Fourier Analysis (EFA), that converts coordinates along a curve into Fourier function coefficients, called harmonic coefficients [@Kuhl1982].
-EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions [@Adams2004; @Claude2008; @Bonhomme2014].
-The number of harmonics determines the quality and precision of the geometric representation of an object.
-The harmonic power, a cumulated sum of squared harmonic coefficients, provides a robust rule for determining the desired number of harmonics [@Bonhomme2014].
+Another approach is elliptic Fourier Analysis (EFA) that turns coordinates into coefficients using mathematical function, commonly applied to two-dimensional closed shape [@Kuhl1982].
+EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions, called harmonic coefficients [@Adams2004; @Claude2008; @Bonhomme2014].
+The number of harmonics determines the quality and precision of the geometry of an object.
+The harmonic power, a cumulated sum of squared harmonic coefficient, provides a robust rule for the desired number of harmonics [@Bonhomme2014].
 
-EFA is suitable for shapes lacking representative landmarks or where curves contain the most meaningful variation.
-Applications in archaeology include human remains and zooarchaeology to understand  biological variation, and more recently on stone artifacts to explore stone tool technology [@Fox2015; @Hoggard2019; @Iovictua2010].
+Outline approaches especially EFA is suitable for shapes lacking representative landmarks or curves convey more meaningful variations.
+The applications in archaeology are seen on human remains or zooarchaeology to understand the biological variation and recently more on stone artifacts to explore stone tool technology [@Fox2015; @Hoggard2019; @Iovictua2010].
 Ceramics is another archaeological materials that outline approaches can provide new insights into shape variation to answer anthropological questions related to ceramic specialization.
 Using semi-landmarks method, @Topi2018 examine 89 photographs of globular jars from the Casas Grandes of northwest Mexico.
-Their results suggest that some ceramic types from the Medio period (AD 1200-1450), were made by specialists based on highly standardized shapes.
-Among those highly standardized jars, some were made by attached specialists <!-- what does this mean, 'attached' to what?
-please clarify -->, while others were made by independent specialists according to their spatial distribution patterns.
-Similarity, @Selden2019 examined 28 Caddo bottle excavated at several sites (AD 500-1700) in northwest Louisiana <!-- using semi-landmarks or something else?
-please add -->.
+Their results suggest that some ceramic types from Medio period (AD 1200-1450), were made by specialists based on high standardized shape.
+Among those high standardized jars, some were made by attached specialists, while others were made by independent specialists according to their spatial distribution patterns.
+Similarity, @Selden2019 examines 28 Caddo bottle excavated at several sites (AD 500-1700) in northwest Louisiana from the Clarence H.
+Webb collections.
 The results indicate a significant integration between the different parts of the bottles including rim, neck, body, and base that vary in a coordinated manner.
-Also, two discrete base and body shapes represented northern and southern differences were identified.
-For the EFA method, @Wilczek2014 evaluate the concordance between EFA and Discrete Cosine Transform (DCT), and a traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France.
-The results show the variation indicated by EFA and DCT matches the traditional ceramic typology, which supports that claim that outline-based approaches can be efficiently used for studying variations in ceramic shapes.
-Furthermore, @Wilczek201's EFA results help us understand the level of production standardization over time across the region.
-Those examples show that outline approaches can distinguish variation in ceramic shapes at high resolutions to understand ceramic production.
+Also, two discrete base and body shapes represented north and south difference were identified.
+For the EFA method, @Wilczek2014 evaluate the concordance between EFA and Discrete Cosine Transform (DCT), and traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France.
+The results show the variation indicated by EFA and DCT matches traditional ceramic typology, which supports that outline-based approaches can be efficiently used for studying variations in ceramic shapes.
+Furthermore, the EFA results help to understand the level of production standardization over time across the region.
+Those examples show that outline approaches can distinguish variation in ceramic shapes in a finer resolution to understand deeply the ceramic production.
 
-Taking the ceramics data from Kiwulan, northeastern Taiwan, we use EFA to evaluate the level of standardization of ceramics in relation to the European presence in the 17th century to get insights into the emergence of ceramic specialization.
-We use a significance test for the equality of coefficient of variations of shape variables to statistically compare the vessel standarization from different periods.
-Using pottery shape as a proxy to study craft specialization, we address these questions: Did foreign influence have impacts on indigenous pottery production that can be detected in the shape of the vessels?
-Did pottery shape become more homogeneous and standardized after foreign contacts with European colonizers or Chinese immigrants?
+Taking the ceramics data from Kiwulan, northeaster Taiwan, we use EFA to evaluate the level of standardization of ceramics in relation to the European presence in the 17th century that might hint the emergence of ceramic specialization.
+In addition, we use significance test for the equality of coefficient of variation of shape variables to compare the vessel standarization from different periods in a statistical way.
+Using pottery shape as a proxy to study craft specialization, we answer the questions: Did foreign influence have impacts on indigenous pottery production that can be detected in the shape of the vessels?
+How does pottery change in shape in relation to foreign influences?
+Did pottery become more homogeneous and standardized in shape after foreign contacts with the European colonizer or the Chinese immigrants?
+Starting from the historical background of colonial influences and its reflection in the material record, this study explore the relationship between the emergence of social inequality and craft specialization that can extend our understanding of the reactions of indigenous societies in the periphery of colonial control, providing a perspective that could be applied in other parts of the world in a similar context.
 
-# Archaeological background and materials
+# Materials and archaeological background
 
-Ceramics analyzed in this paper come from 40 units (4m by 4m) sampled from the central, undisturbed area of archaeological excavations at Kiwulan (Figure \@ref(fig:KWL-map); Figure \@ref(fig:KWL-sam-area)).
-Kiwulan is situated on a hill near a riverside at the northern margin of Yilan Province, which is characterized by a triangular alluvial plain facing eastwards the Pacific with high mountains on three other sides.
-Yilan is an ideal context to study peripheral colonial influence because Indigenous communities there were isolated by natural barriers that limited the frequency of direct contact with the Spanish and the Dutch settlers in northern Taiwan.
-The chronology of the archaeological deposits at Kiwulan consists of two cultural components, the upper component and the lower component, with a sterile layer in between [@Chen2007].
-This paper focuses on the upper component, dated from AD 1350 to AD 1950, covering the late Iron Age and the historical period, which we define as the European presence in Taiwan in the early 17th century.
-The Dutch first occupied southern Taiwan in 1624 and then the Spanish occupied northern Taiwan in 1626 <!-- add citation for these dates -->.
-In 1642, the Spanish was expelled by the Dutch in northern Taiwan, who took over their forts at Helping Dau in Keelung, and in Tamsui.
-Since then, western Taiwan was mostly under Dutch colonial rule until 1662 when the Kingdom of Tungning in Taiwan was founded by Koxinga, a loyalist of the Ming dynasty of China [@Andrade2007].
+Ceramics analyzed in this paper come from 40 units (4m by 4m) sampled from the largest excavation section at Kiwulan, northeastern Taiwan (Figure \@ref(fig:KWL-map); Figure \@ref(fig:KWL-sam-area)).
+Kiwulan is situated on a hill near a riverside at the northern margin of Yilan, which is characterized by a triangular alluvial plain facing eastwards the Pacific with high mountains on three other sides .
+Yilan is an ideal context to study peripheral colonial influence because it was relatively isolated by natural barrier that reduced frequency of direct contact with the Europeans presences, the Spanish and the Dutch settled in northern Taiwan.
+The chronology of Kiwulan consists of two cultural components, the upper component and the lower component, with a sterile layer in between [@Chen2007].
+This paper focuses on the upper component dated from AD 1350 to AD 1950, covering the late Iron Age and the historical period defined by the European presence in Taiwan in the early 17th century.
+The Dutch firstly came and occupied southern Taiwan in 1624 and then the Spanish occupied northern Taiwan in 1626.
+In 1642, the Spanish was expelled by the Dutch, who took over their forts at Helping dau in Keelung, and in Tamsui.
+Since then, western Taiwan was mostly under the colonial rule of the Dutch until 1662 when the Kingdom of Tungning in Taiwan was founded by Koxinga, a loyalist of the Ming dynasty of China [@Andrade2007].
 
-The archaeological record of Kiwulan's upper component shows traces of foreign contacts, including Europeans in the 17th century, and waves of Chinese immigrants in the 19th century.
-Imported ceramics from mainland China, stoneware, and ornaments such as beads have been found in the upper component, indicating frequent long-distance trade activities in the 17th century during the European presence in northern Taiwan.
-In addition to artifacts, archaeological features such as burials, middens, and post-holes with _in-situ_ posts are widespread across  the 1-2 m thick deposit of the upper component, and demonstrate that Kiwulan was a continuously occupied large settlement site [@Chen2007].
-To compare different foreign influences, we classified the upper component into three chronological phases: the pre-European, European, and Chinese periods.
-We identified these phases using 32 previously published radiocarbon ages <!-- add citaion to report with dates -->, excavation depth measurements, stratigraphic details reported by the excavators (colour, disturbance, etc.), and finds of chronologically diagnostic artifacts, such as blue and white porcelains, light grey glazed jars, and large dark brown glazed stoneware jars commonly used in the 17th century, and bricks and tiles used by the Chinese in the 19th century [@Hsieh2009; @Wang2011].
+The upper component experienced foreign contacts including the Europeans in the 17th century and great waves of Chinese immigrants in the 19th century.
+Imported ceramics from mainland China, stonewares, and ornaments elements such as beads were trade goods commonly found in the upper component that indicates the frequent long-distance trade activities in the 17th century during the European presence in northern Taiwan.
+In addition to artifacts, features such as burials, middens, and post-holes with in-situ posts widespread across 1-2 m thick deposit explain that Kiwulan was a continuously occupied and large settlement site [@Chen2007].
+To compare different foreign influences, the upper component were classified into three phases, the pre-European, European, and Chinese periods.
+The principle of identification of three phases is based on 32 radiocarbon ages, excavation depth, consistency of contexts, and types of chronologically diagnostic artifacts, such as blue and white porcelains, light grey glazed jars, and large dark brown glazed stoneware jars prevailing in the 17th century, and bricks and tiles used by the Chinese in the 19th century [@Hsieh2009; @Wang2011].
 The deposit shows signs of continuous human occupation during each of the three phases across the sampling area.
 
-The most abundant artefacts in the upper component are locally made ceramics, which are distributed throughout the sequence and across the site.
-More than 550,000 sherds were excavated, and around 1,200 vessels could be reconstructed.
-There are only two forms of local vessels, a cooking pot and a steamer made of two cooking pots stacked together with a clay filter layer in between.
-Those pots forms show high consistency in shape.
-They have a globular body with a short neck and wide mouth.
-The exterior surface below the neck is decorated with a wide variety of impressed geometric motifs.
-These vessels were probably used for cooking, with evidence of charred residues and carbon deposits frequently observed on vessel interiors and soot on their exteriors.
-Moreover, a lack of evidence of other utilitarian earthenware ceramics suggests these globular pots were mainly used for daily cooking.
+The predominant artefact found is locally made ceramics, which distributed throughout the sequence across the site.
+More than 550 thousands of sherds were excavated, and around 1,200 vessels were reconstructed.
+It is interesting to note that there are only two forms of local vessels, one is cooking pot and another one is steamer made of two cooking pots stacked together with a filter clay layer in between.
+Those pots show high consistency in shape.
+They are globular body with short neck and wide mouth.
+The exterior surface below the neck is treated with a wide variety of impressed geometric motifs.
+They were believed used for cooking based on the evidence of charred residues and carbon deposits commonly observed on the interior and soot on the exterior.
+Moreover, a lack of evidence of other utilitarian earthenware ceramics suggests this globular pot was mainly used for daily cooking.
 They are fired to orange to brownish color with a fully oxidized core or a reduced core with oxidized fringes.
 Finger impressions and seams usually on the interior indicate they were pinched using slabs of clay and shaped by hand.
-This kind of pot has been widely found at archaeological sites during the late Iron Age and the historical period throughout the Yilan Plain.
+In addition to Kiwulan site, this kind of pot was widely found at other archaeological sites during the same period in Yilan Plain.
 
@@
 
-Petrographic analysis for 34 thin sections shows high percentage of inclusions (15-50%), including argillite (15-40%), metasandstone (1-10%),  sandstone fragments (1-6%), quartz (1-5%), and trace amounts of feldspar and slate.
-The size of the particles ranges from 500 micron to 1300 micron.
-In general, the globular vessel farbric presents a mixture of fine, rounded argillite with a small part of rounded metasandstone and rounded to sub-angular monocrystalline quartz.
+Petrographic analysis for 34 thin sections shows high percentage of inclusions (15-50%), including argillite, quartz, feldspar, metasandstone, sandstone fragment, and slate.
+The most common type of sherds, also the dominant one, has a high percentage of inclusion (25-40%), which is composed mainly by argillite (15-40%), followed by metasandstone (1-10%), sandstone (1-6%), and quartz (1-5%).
+The size of the particles ranges from 500 micron to 1300 micron (=1 mm).
+In general, this type presents a mixture of fine, rounded argillite with a small part of rounded metasandstone and rounded to sub-angular monocrystalline quartz.
 The composition matches the mineralogical composition of local raw materials in the Yilan Plain [@Chen2016].
-There is no substantial changes in the inclusions over time, indicating continuity in pottery fabric composition across the three periods.
+It seems there is no obvious changes in the inclusions over time.
 
@@
 
-A total of `r total_no` largely intact vessels from pre-contact contexts (n = 32), post-European contexts (n = 27), and Chinese contact contexts (n = 14) were selected for ellipic Fourier analysis.
-Most of them cover every vessel component from rim to bottom.
+A total of `r total_no` relativity intact vessels from pre-contact contexts (n = 32), post-European contexts (n = 27), and Chinese contact contexts (n = 14) were selected for Ellipic Fourier analysis.
+Most of them cover every components from rim to bottom.
+Given that the decline of indigenous population in the Chinese period might have impact on the interpretation, we examined the spatial distribution of pots analyzed in this paper.
+The pottery distributed randomly across the sampling area without any clusters could rule out the bias caused by the population dynamics (Figure \@ref(fig:pottery-spatial-pattern)).
 To fully explore the pottery standardization, we also take the metric measurements into account for comparison.
@@
 To obtain information about the pottery shape, we also compare the ratio of Rim/Body thickness, and the ratio of Rim/Body diameter.
-
-## Digitising and analysing by EFA
-
-We used scanned pottery drawings acquired from Bureau of Cultural Affairs in Yilan.
-All drawings present a two-dimensional view of vessel cross-sections with indications of metric measurements.
-The scanned drawings were imported into Inkscape (http://inkscape.org) where outlines were traced manually.
-Where only half cross-section images were available, these were duplicated, flipped, and then joined with another one to create a 2D closed outline for each vessel.
-Geometric morphometric analyses were conducted using the R software [@Rlanguage2019] using the functions included in the Momocs, a R package intended to quantify and analyse shapes, especially for outline analysis [@Bonhomme2014].
-The digitised outlines were converted into a list of successive x-y pixel coordinates for elliptic Fourier analysis (EFA), to investigate morphological differences among pottery shapes from our three occupation phases.
-The harmonic coefficients generated by EFA were analysed by principal component analysis (PCA) for dimensionality reduction to illustrate the diversity of the shape data and identify major patterns of variation through .
-
-## Statistical analysis
-
-The principal component (PC) scores were analysed with a multivariate analysis of variance (MANOVA) test to identify significant differences in shapes between occupation phases.
-We also computed coefficients of variation values (CVs) for the PCs, treating the PCs as shape variables that are more informative than linear dimensions.
-The coefficient of variation is a common and widely-used statistical measure of the spread of a set of measurements of a sample.
-It is defined as the standard deviation divided by the mean in a ratio scale format:
-
-$$c_v = \frac{\sigma}{\mu}$$
-
-By standardizing standard deviation in each data set, the coefficients of variation values allows us to directly compare variation in samples measured with different units or means.
-We take this as our measurement of standardisation in vessel shape: lower CV values reflect higher standardisation, and thus increased craft specialisation in the community.
-To answer the question of whether CV values for vessels samples across our three occupational phases are significantly different or not, we used the "Asymptotic test for the equality of coefficients of variation from k populations" [@Feltz1996], a significance test for CVs provided by the R package cvequality [@Marwick2019].
-Our implementation is based on the following from @Feltz1996:
-
-Let $s_i$ be the observed standard deviation of the $i$th sample (or group of measurements), $x_i$ be the observed mean of the $i$th sample, and let
-
-$$m_i=n_i-1$$
-
-and we do not know the population $c_v$, so let $\widehat{\tau }_c$ be an estimate as follows:
-
-$$\widehat{\tau }_c = \frac{\left( \sum _{j=1}^km_j\frac{S_j}{\bar{x}_j} \right)}{M}$$ 
-where 
-$$M=\sum _{j=1}^m m_j$$
-
-Then the test statistic can be computed with:
-
-$$\begin{aligned} D'A D= \widehat{\tau }_c^{-2}\left( 0.5+\widehat{\tau }_c^2\right) ^{-1}\left[ \sum _{i=1}^km_i\left(\frac{s_i}{\bar{x}_i} \right)^2-\frac{\widehat{\tau }_c^2}{M} \right] \end{aligned}$$
-
-The $D'AD$ value measures how far each sample $c_v$ is from our estimate of the population $c_v$.
-@Feltz1996 note that the $D'AD$ value distributes as a central $\chi^2$ random variable with $k-1$ degrees of freedom, from which a p-value can be computed.
-
-# Reproducibility and open source materials
-
-To enable re-use of materials and improve reproducibility and transparency [@Marwick2017], the entire R code [@Rlanguage2019] used for all the analysis and visualizations contained in this paper is openly available online at https://doi.org/10.17605/OSF.IO/ABVGF.
-Also in this version-controlled compendium [@Marwick2018] are the raw data for all the visualizations and tests reported here.
-All of the figures, tables, and statistical test results presented here can be independently reproduced with the code and data in this repository.
-The code is released under the MIT license, the data as CC-0, and figures as CC-BY, to enable maximum re-use.
-
-# Results
-
-To investigate the effect of the historically-documented decline of Indigenous population in the Chinese period <!-- add citation on population change during Chinese period --> on pottery production we examined the spatial distribution of pots in our sample in each period.
-Our concern is that the smaller population size during the Chinese period could confound our investigations into craft specialisation beause of reduced numbers of potters and pottery production shrinking a a few locations in the settlement during this time.
-Altough the overall number of vessels is smaller during the Chinese contact period, Figure \@ref(fig:pottery-spatial-pattern) shows that pottery is distributed randomly across the sampling area without any distinctive clusters.
-This suggests that differences in population across our three occupation phases are probably not driving variation in craft specialsation.
 
@@
 
+## Digitising and analysing by EFA
+
+The scanned pottery drawings were acquired from Bureau of Cultural Affairs in Yilan.
+All drawing presents two-dimensional view of the section of a vessel with indications of metric measurements.
+The scanned drawings were imported into the Inkscape software for outlines tracing to remove additional information such as marks, lines, and measurements on the original drawings in order to obtain clear outline.
+Each traced half cross-section image was duplicated, flipped, and then joined with another one to create a 2D closed outline.
+Geometric morphometric analyses were conducted in the R software (www.rproject.org, Core-Team, 2015) using the functions included in the Momocs, a R package intended to quantify the shape and compare its variation, especially for outline analysis [@Bonhomme2014].
+The digitised outlines were converted into a list of successive x-y pixel coordinates for elliptic Fourier analysis (EFA), which assesses morphological differences among pottery shapes from three occupation contexts.
+The harmonic coefficients generated by EFA were analysed by principal component analysis (PCA) to illustrate the diversity of the shape data and identify the major patterns of variation through dimensionality reduction.
+
+## Statistical testing
+
+The principal components (PCs) scores were analysed with a multivariate analysis of variance (MANOVA) to test for significant effects of the groups of occupation context on shape variances.
+Finally, we computed coefficients of variation statistic (CVs) among multiple groups of PCs with a significance test for the equality of CVs, which enable us to compare CVs in a statistical way [@Pearson1896].
+The coefficient of variation is a common and widely-used statistical measure of the spread of a set of measurements of a sample.
+It is defined as the standard deviation divided by the mean in a ratio scale format.
+
+$$c_v = \frac{\sigma}{\mu}$$
+
+By standardizing standard deviation in each data set, the coefficients of variation statistics allows us to directly compare variation in samples measured with different units or means.
+Although it is a useful method, there is a question that how should we make a decision that the differences among the multiple CVs are significant or not?
+To answer this question, we need convenient methods for testing the equality of $c_v$ values from different samples, which has been described in many publications and here we proposed a significant test for CVs using the R package cvequality [@Marwick2019] in the CRAN repository.
+The package cvequality includes two tests, the "Asymptotic test for the equality of coefficients of variation from k populations" [@Feltz1996] and the "Modified signed-likelihood ratio test (SLRT) for equality of CVs" [@Krishnamoorthy2014].
+@Feltz1996 test is widely cited as an authoritative test for the equality of $c_v$ values that performs better than many approximative tests.
+It is a 'gold standard' test for comparing multiple $c_v$.
+Our implementation is based on the following from @Feltz1996:
+
+Let $s_i$ be the observed standard deviation of the $i$th sample (or group of measurements), $x_i$ be the observed mean of the $i$th sample, and let
+
+$$m_i=n_i-1$$
+
+and we do not know the population $c_v$, so let $\widehat{\tau }_c$ be an estimate as follows:
+
+$$\widehat{\tau }_c = \frac{\left( \sum _{j=1}^km_j\frac{S_j}{\bar{x}_j} \right)}{M}$$ 
+where 
+$$M=\sum _{j=1}^m m_j$$ 
+Then the test statistic can be computed with: 
+$$\begin{aligned} D'A D= \widehat{\tau }_c^{-2}\left( 0.5+\widehat{\tau }_c^2\right) ^{-1}\left[ \sum _{i=1}^km_i\left(\frac{s_i}{\bar{x}_i} \right)^2-\frac{\widehat{\tau }_c^2}{M} \right] \end{aligned}$$
+The $D'AD$ value measures how far each sample $c_v$ is from our estimate of the population $c_v$.
+@Feltz1996 note that the $D'AD$ value distributes as a central $\chi^2$ random variable with $k-1$ degrees of freedom, from which a p-value can be computed.
+We also include the @Krishnamoorthy2014 test as a more recent development with lower rates of type I error, better performance with uneven sample numbers, and more power across a range of conditions.
+
+# Reproducibility and open source materials
+
+To enable re-use of materials and improve reproducibility and transparency [@Marwick2017], the entire R code [@Rlanguage2019] used for all the analysis and visualizations contained in this paper is included in http:XXX.
+Also in this version-controlled compendium [@Marwick2018] are the raw data for all the visualizations and tests reported here.
+All of the figures, tables, and statistical test results presented here can be independently reproduced with the code and data in this repository.
+The code is released under the MIT license, the data as CC-0, and figures as CC-BY, to enable maximum re-use.
+
+# Results
+
+Table \@ref(tab:metricCVstest-pre-post) and \@ref(tab:metricCVstest-post-chi) shows the results of Asymptotic test [@Feltz1996] for equality of CVs for for six measurements and two ratio values.
+There is a significant difference (p-value less than 0.01) between CV values for two metric measurements across phases, the rim diameter and the neck diameter before and after the European presnece.
+The CV values show that the variation of rim diameter is small (6.8%) before European contact, increases after the European presence (9.6%), and decreases after the Chinese presence (6.9%).
+For the neck diameter, the CV values indicate smaller variation after the European period (7.7%) and the Chinese period (7.1%) compared to pre-European period (10.6%).
+After the arrival of the Chinese, there is only one metric attribute, the rim diameter, showing significant difference from post-European period (p-value less than 0.05).
+
+```{r metricCVstest-pre-post, eval=FALSE}
+# read metric data
+kwl_p_ave <- readr::read_csv(here::here("analysis", "data", "raw_data",
+                                 "KWL-pottery-metrics-tidy-ave.csv"))
+library(cvequality)
+# select metric attributes and tidy data for computing CV
+kwl_p_metrics_long <-
+  kwl_p_ave %>%
+  mutate(dia_rim_body = Dia_Rim_ave/Dia_Body_ave,
+         thi_rim_body = Thick_Rim_ave/Thick_Body_ave) %>%
+  select(period,
+         Dia_Rim_ave,
+         Dia_Neck_ave,
+         Dia_Body_ave,
+         Thick_Rim_ave,
+         Thick_Neck_ave,
+         Thick_Body_ave,
+         dia_rim_body,
+         thi_rim_body,
+         # H_Rim
+  ) %>%
+  rename(`Rim diameter (mm)` = Dia_Rim_ave,
+         `Neck diameter (mm)` = Dia_Neck_ave,
+         `Body diameter (mm)` = Dia_Body_ave,
+         `Rim thickness (mm)` = Thick_Rim_ave,
+         `Neck thickness (mm)` = Thick_Neck_ave,
+         `Body thickness (mm)` = Thick_Body_ave,
+         `Ratio of Rim/Body diameter` = dia_rim_body,
+         `Ratio of Rim/Body thickness` = thi_rim_body
+  ) %>%
+  gather(variable, value, -period) %>%
+  filter(!is.na(period),
+         !is.na(value))
+
+# filter pre-e and post-e periods
+kwl_p_metrics_long_pre_post <-
+  kwl_p_metrics_long %>%
+  filter(!period == "ch-con")
+
+# filter post-e and chi periods
+kwl_p_metrics_long_post_chi <-
+  kwl_p_metrics_long %>%
+  filter(!period == "pre-e")
+
+#------------------------------------------CV test for pre-e and post-e pair
+
+kiwulan_pottery_metrics_cvs_plot_cv_test_asym_pre_post <-
+  kwl_p_metrics_long_pre_post %>%
+  nest(-variable) %>%  # or nest(period, value) %>%
+  mutate(asymptotic_test = map(data, # change from mslr_test to asymptotic_test
+                               ~bind_cols(asymptotic_test (
+                                 .x$value,
+                                 .x$period)))) %>%
+  unnest(asymptotic_test)
+
+# compute CV
+kwl_p_cvs_pre_post <-
+  kwl_p_metrics_long_pre_post %>%
+  group_by(period, variable) %>%
+  summarise(cvs = raster::cv(value, na.rm = TRUE))
+
+# join p-values to CVs
+kwl_p_cvs_and_pvalues_pre_post <-
+  kwl_p_cvs_pre_post %>%
+  left_join(kiwulan_pottery_metrics_cvs_plot_cv_test_asym_pre_post) %>%
+  mutate(significant = ifelse(p_value <= 0.05, 'yes', 'no')) %>%
+  mutate(variable = factor(variable, ordered = TRUE)) %>%
+  arrange(variable)
+
+# making table for CV
+kwl_p_cvs_table_pre_post <-
+  kwl_p_cvs_and_pvalues_pre_post %>%
+  select(-data, -significant) %>%
+  pivot_wider(names_from = period, values_from = cvs) %>%
+  mutate(variable = factor(variable, levels = c("Rim thickness (mm)",
+                                                "Neck thickness (mm)",
+                                                "Body thickness (mm)",
+                                                "Ratio of Rim/Body thickness",
+                                                "Rim diameter (mm)",
+                                                "Neck diameter (mm)",
+                                                "Body diameter (mm)",
+                                                "Ratio of Rim/Body diameter"))) %>%
+  arrange(variable) %>%
+  select(variable, `pre-e`, `post-e`, `D_AD`, `p_value`) %>%
+  mutate_at(c("pre-e", "post-e", "D_AD", "p_value"), round, 4)
+
+knitr::kable(kwl_p_cvs_table_pre_post,
+             caption = "Coefficient of Variation test for metric attributes before and after European presence")
+```
+
+```{r metricCVstest-post-chi, eval=FALSE}
+#-----------------------------------CV test for post-e and ch-con pair
+kiwulan_pottery_metrics_cvs_plot_cv_test_asym_post_chi <-
+  kwl_p_metrics_long_post_chi %>%
+  nest(-variable) %>%  # or nest(period, value) %>%
+  mutate(asymptotic_test = map(data, # change from mslr_test to asymptotic_test
+                               ~bind_cols(asymptotic_test (
+                                 .x$value,
+                                 .x$period)))) %>%
+  unnest(asymptotic_test)
+
+# compute CV
+kwl_p_cvs_pre_post <-
+  kwl_p_metrics_long_post_chi %>%
+  group_by(period, variable) %>%
+  summarise(cvs = raster::cv(value, na.rm = TRUE))
+
+# join p-values to CVs
+kwl_p_cvs_and_pvalues_post_chi <-
+  kwl_p_cvs_pre_post %>%
+  left_join(kiwulan_pottery_metrics_cvs_plot_cv_test_asym_post_chi) %>%
+  mutate(significant = ifelse(p_value <= 0.05, 'yes', 'no')) %>%
+  mutate(variable = factor(variable, ordered = TRUE)) %>%
+  arrange(variable)
+
+# making table for CV
+kwl_p_cvs_table_post_chi <-
+  kwl_p_cvs_and_pvalues_post_chi %>%
+  select(-data, -significant) %>%
+  pivot_wider(names_from = period, values_from = cvs) %>%
+  mutate(variable = factor(variable, levels = c("Rim thickness (mm)",
+                                                "Neck thickness (mm)",
+                                                "Body thickness (mm)",
+                                                "Ratio of Rim/Body thickness",
+                                                "Rim diameter (mm)",
+                                                "Neck diameter (mm)",
+                                                "Body diameter (mm)",
+                                                "Ratio of Rim/Body diameter"))) %>%
+  arrange(variable) %>%
+  select(variable, `post-e`, `ch-con`, `D_AD`, `p_value`) %>%
+  mutate_at(c("post-e", "ch-con", "D_AD", "p_value"), round, 4)
+
+knitr::kable(kwl_p_cvs_table_post_chi,
+             caption = "Coefficient of Variation test for metric attributes between European and Chinese presence")
+```
+
 ```{r jpg-coordinates, eval=FALSE}
@@
 
-<!-- basic description of shape -->
-
 The elliptic Fourier coefficients of `r total_no` pottery from three phases were calculated.
-Reliable pottery outlines were captured by 13 harmonics that gather 99% of the total harmonic power.
+Reliable pottery outline was captured by 13 harmonics that gather 99 % of the total harmonic power.
 The average shape of vessels from each phase was visualized by computing the mean of the standardized Fourier coefficients within each phase group (Figure \@ref(fig:EFA-analysis)).
-Figure \@ref(fig:thin-plate-splines) shows the shape changes described using thin-plate spline warping for paired periods, pre- and post-European periods, and post-European and Chinese periods.
+Figure \@ref(fig:thin-plate-splines) shows the shape changes described using thin-plate spline warping for paired periods, pre and post European periods, and post-European and Chinese period in a time sequence.
 
@@
 
-<!-- further description of shape and differences between phases -->
-
-The elliptic Fourier coefficients were examined by Principal Components Analysis to explore the shape variations across three phases.
+The elliptic Fourier coefficients are examined by principal components analysis to explore the shape variations across three phases.
 The first two principal components (PCs) explain `r vari_two`% of the total variances, of which `r pc1`% is explained by the first principle component.
-With the third component, the first three principal components explain `r vari_three`% of the total variance.
+With the third component, the first three principal components explain `r vari_three`% of the total variances.
 Figure \@ref(fig:contribution-PCs) shows the shape variation associated with the first three principal components.
-PC1 captures the height of the vessels, from tall to short, and the roundness of the body from round to oval-shaped.
-PC2 relates to the neck and mouth constriction, from narrow to wide.
-PC3 explains a smaller portion of the variance (`r pc3`%), which relates to the degree of the flare of the neck, from curved to straight shape.
-The first two components account for most of the variance in relation to three phases (Figure \@ref(fig:first-two-PCs-byplot)).
-The results reflect a large overlap in shapes from three occupations phases, especially for shapes in the pre-European and post-European periods.
-However, the spread of shape distribution indicates a wider variation in shapes in the pre-European and post-European periods compared to those in the Chinese period along both PC1 and PC2 axes.
-In other words, we find a decrease in shape variance in the Chinese period that is evident in the shorter height and narrower mouth of vessels used in that period.
+From the negative to the positive scores, the PC1 represents the height of the vessels from tall to short and the roundness of the body from round to oval-shaped.
+The PC2 relates to the neck and mouth constriction from narrow to wide.
+The PC3 explains a smaller portion of the variance (`r pc3`%), which relates to the degree of the flare of the neck from curved to straight shape.
+The first two components account for most of the variance in relation to three phases were represented in Figure \@ref(fig:first-two-PCs-byplot).
+The result reflects a large overlap of three groupings especially for shapes in the pre-European and post-European periods that might show some extent of the similarity.
+However, the spread of shape distribution indications a wider variation in shapes in the pre-European and post-European periods compared to those in the Chinese period along both PC1 and PC2 axes.
+In other words, there is a decreasing in shape variance in the Chinese period that shares the features of shorter height and narrower mouth.
 
@@
 ```{r first-two-PCs-byplot, fig.cap= "Pottery shape distribution by each occupation phase according to the first two PCs"}
+# the pots groups by phase, and their shape distribution according to 
+# the first two principle components of the EFA 
+# png(here::here("analysis/figures/pca-biplot.png"), res = 300, w = 1500, h = 1500)
+# plot(op.p, 2, 
+#      morpho = TRUE, 
+#      ellipses = TRUE,
+#      cex = 1, 
+#      cex.labelsgroups = 1.5)
+# dev.off()
 
 # PCA byplot
-plot_PCA(op.p, ~phase, chullfilled = TRUE)
+plot(PCA(op), 2, pos.shp="range")
 ```
-
-<!-- statistical test of difference in shape decsribed in previous paragraph -->
-
-To test for differences in the distributions of shape variables indicated by the PC scores shown in Figure \ref(@fig:first-two-PCs-byplot)), we used a multivariate analysis of variance (MANOVA) test to compare pairwise combinations across our three occupation phases.
-Table \@ref(tab:MANOVA-table) shows significant differences in shape between Pre-European and Post-European phases (p = $`r pval_post_pre`$), and Pre-European and Chinese phases (p = $`r pval_chi_pre`$).
-Although there is considerable overlap of shape variables between the Pre-European and Post-European phases, their PC scores are significantly different from each other.
-There is little difference between vessel shapes in the Post-European and Chinese contact periods.
 
@@
 
-manv <- rrtools:::quietly ( MANOVA_PW(op.p, "phase") ) 
+manv <- rrtools:::quietly ( MANOVA_PW(op.p, "phase") ) #? how to hide the process, echo = FALSE?
 su_manv <- as.data.frame(manv$summary)
+# how to reduce the number in the table to only 4 digits, tried mutate with signif but doesn't work?
+su_manv <- 
+  su_manv %>% 
+  mutate_all(round, 4)
 
-su_manv_tab <- 
-  su_manv %>% 
-  mutate_all(round, 4) %>% 
-  select(-Df, -`num Df`) %>% 
-  rename(`Pillai's trace` = Pillai,
-         `Approximate F value` = `approx F`,
-         `degrees of freedom` = `den Df`)
+pval_chi_pre <- signif(unname(su_manv[,'Pr(>F)'][2]), 4)
+pval_post_pre <- signif(unname(su_manv[,'Pr(>F)'][3]), 4)
 
-pval_chi_pre <- signif(unname(su_manv_tab[,'Pr(>F)'][2]), 4)
-pval_post_pre <- signif(unname(su_manv_tab[,'Pr(>F)'][3]), 4)
-
-knitr::kable(su_manv_tab,
-             caption = "Summary statistics for the MANOVA test on the PC scores. `Pr(>F)` is the p-value associated with the F statistic of the effect and test statistic. ")
+knitr::kable(su_manv,
+             caption = "The statistics summary of MANOVA test for the PC scores")
 ```
 
-<!-- CV equality testing -->
-To compare pottery shape standardization across the three phases we  investigate the distributions of the first three PC scores, taking the PC scores as proxy variables for vessel shape (Figure \@ref(fig:PCs-distribution)).
-The first PC, vessel height, shows higher variation in the pre-European period compared to the Chinese period.
-That indicates that shape standardization was higher in the Chinese period compared to the pre-European period.
-The second PC also presents a similar pattern of higher shape standardization in the Chinese period compared to the other two phases.
-To see whether the differences in the distribution of PCs between any two phases are substantive or due to chance, we used a modified signed-likelihood ratio test to assess the equality of CVs.
-P-values for this significance test of CVs for PC1 and PC2 show significant differences in the standardization of vessel shapes across periods, especially between Chinese contact with either pre-European or post-European(Figure \@ref(fig:CV-test-PCs-plot), Table \@ref(tab:CV-test-PCs-table)).
-A significant difference is also detected between the pre-European and post-European periods.
+To test differences in the distribution of shape variables indicated by the PC scores, we use multivariate analysis of variance (MANOVA) to compare a pairwise combination between every sample means across three phases.
+The results in Table \@ref(tab:MANOVA-table) show significant differences in shape between Pre-European and Post-European (p = $`r pval_post_pre`$) and Pre-European and Chinese contact shapes (p = $`r pval_chi_pre`$), which corresponds to the previous observation of the different distribution in shapes in Chinese period according to PCA byplot.
+In addition, although there is a considerable overlap of shape variables between Pre-European and Post-European, their PC scores are significantly different from each other.
+There are not much differences between the shapes in the Post-European and Chinese contact periods.
+
+```{r GAMM, eval = FALSE}
+#----------------------------------------------------
+# GAMMS from https://www-nature-com.offcampus.lib.washington.edu/articles/s41598-018-20122-9
+
+# Generalised additive mixed models (GAMMs) were used to explain shape variance with respect to mean environmental parameters and shell size, and to compare between individual shape features (PCs).
+A GAMM is a generalised linear mixed model with a linear predictor involving a sum of smooth functions of covariates.
+The model allowed for flexible specification of the dependence of the response on the covariates by defining regression splines (smoothing functions) and estimating their optimal degree of smoothness, rather than calculating parametric relationships
+
+# metric data
+ggplot(pottery_data,
+       aes(phase,
+           as.numeric(Thick1_Neck))) +
+  geom_boxplot()
+
+pottery_data$thick_neck1 <- as.numeric(pottery_data$Thick1_Neck)
+pottery_data$thick_body1 <- as.numeric(pottery_data$Thick1_Body)
+pottery_data$height <- as.numeric(pottery_data$Hight)
+
+plot(op.p$x[ , 2], pottery_data$thick_body1)
+
+gams_input <- data.frame(thick_body1 = pottery_data$thick_body1 , 
+                         height = pottery_data$height, 
+                         shape_pc1 = op.p$x[ , 1])
+
+# Bayesian Regression Models using 'Stan'
+library(brms)
+m2 <- brm(bf(shape_pc1 ~ s(thick_body1)),
+          data = gams_input, family = gaussian(), cores = 4, seed = 17,
+          iter = 4000, warmup = 1000, thin = 10, refresh = 0,
+          control = list(adapt_delta = 0.99))
+summary(m2) # so many zeros... not very promising 
+# plot marginal effects for each predictor
+plot(marginal_effects(m2), ask = FALSE)
+loo(m2)
+pp_check(m2)
+```
 
@@
        h = 10, w = 5)
+```
+
+```{r variance-test-PCs}
+# we don't use CV because that is not good for measurements with no absolute zero
+xx <- 
+  pc_plot %>% 
+  group_by(pcn, phase) %>% 
+  summarise(variance = var(pc))
+
+# statistical test to compare the variances of two samples
+pc_plot_tests <- 
+  pc_plot %>% 
+  group_by_at(vars(-pc)) %>% 
+  mutate(i = row_number()) %>% 
+  ungroup() %>% 
+  mutate(phase = as.character(phase),
+         pcn = as.character(pcn)) %>% 
+  spread(phase, pc) %>% 
+  select(-i) %>% 
+  nest_legacy(-pcn) %>% 
+  mutate(var_test = purrr::map(data, ~broom::tidy(var.test(.x$`Post-European` ,
+                                            .x$`Pre-European`)))) %>% 
+  unnest(var_test)
 ```
@@
 
+The variation of pottery shape for each phase is explored by the distributions of the first three PC scores to compare the extent of standardization across phases.
+Figure \@ref(fig:PCs-distribution) suggests variations of PC scores for each vessel by phase.
+The first PC, the hight of the vessels, shows more variations in the pre-European period compared to the Chinese period.
+That indicates standardization appears strong in the Chinese period but weak in the pre-European period.
+The second PC also presents a similar pattern that stronger standardization of shape in the Chinese period compared to the other two phases.
+To see whether the differences in the distribution of PCs between any two phases is statistically significant, coefficients of variation statistic (CVs) among phase groups of PCs is examined with a modified signed-likelihood ratio test for the equality of CVs.
+P-values for the significant test of CVs of PC1 and PC2 show significant differences (P-value less than 0.01) in the standardization of vessel shape across some periods, especially between Chinese contact with either pre-European or post-European(Figure \@ref(fig:CV-test-PCs-plot), Table \@ref(tab:CV-test-PCs-table)).
+Significant difference is also detected between pre-European and post-European (P-value less than 0.05).
+
 # Discussion

### 2020-01-27 analysis/paper/paper.Rmd
starting to edit the discussion

--- 2020-01-27 analysis/paper/paper.Rmd
@@
 
-Previous investigations at Kiwulan suggest an unequal distribution pattern of prestige goods, trade ornaments specifically, after the presence of the European that hints the emergence of social inequality.
-We examine the degree of the ceramic standardization to measure ceramic specialization that is usually associate with social differentiation, such as the elite control [@Costin2001; @Junker1999].
+Previous investigations at Kiwulan have suggested an unequal distribution of prestige goods, trade ornaments specifically, following the appearance of Europeans <!-- citation to that previous investigation -->, hinting at the emergence of social inequality in the Indigenous community.
+Here we have examined ceramic vessel shape standardization to measure craft specialization as a proxy for social differentiation, such as elite control [@Costin2001; @Junker1999].
 The results confirm that the differences in pottery shapes can be detected using EFA method, indicating the foreign influences on the indigenous society- the Europeans in the 17th century and the Chinese in the 19th century.

### 2020-01-27 analysis/paper/paper.Rmd
more edits on discussion section

@@
 
-Previous investigations at Kiwulan have suggested an unequal distribution of prestige goods, trade ornaments specifically, following the appearance of Europeans <!-- citation to that previous investigation -->, hinting at the emergence of social inequality in the Indigenous community.
+Previous investigations at Kiwulan have suggested an unequal distribution of prestige goods, trade ornaments specifically, following the appearance of Europeans <!-- add citation to that previous investigation -->, hinting at the emergence of social inequality in the Indigenous community.
 Here we have examined ceramic vessel shape standardization to measure craft specialization as a proxy for social differentiation, such as elite control [@Costin2001; @Junker1999].
-The results confirm that the differences in pottery shapes can be detected using EFA method, indicating the foreign influences on the indigenous society- the Europeans in the 17th century and the Chinese in the 19th century.
-The result of MANOVA shows there is a significant difference in shapes between pre-European and Post-European period, and between the pre-European and the Chinese period.
-The average shape presents a round body with a wider rim and neck before the European contact, and turns to oval-shaped body with narrower rim and neck after the European presence and even more obvious during the Chinese contact period.
-In general, pottery shapes become shorter in hight over time that leads to the oval-shaped body.
-For the standardization of pottery shape, the result of CV test indicates the significant differences between pre-European period and post-European period (P-value less than 0.05), and between Chinese contact and either pre-European or post-European (P-value less than 0.01).
-This suggests pottery shape becomes more homogeneous and standardized after foreign contacts with the European colonizer and even more with the Chinese immigrants.
-The compositional analysis shows clay pastes are similar, regardless of the increasing standardization of the pottery shape, reflecting the raw material sources remains unchanged.
-More standardized shape in the Chinese period might indicate the appearance of a specialized group.
-To figure out whether this change is related to the function of pot, we made an attempt to conduct lipid analysis for potsherds samples to identify the source of residue absorbed into the fabric.
-However, we did not obtain useful results due to a low lipid yield, which might result from the thin and less porous fabric that offers limited spaces to protect organic molecules from microbiological degradation [@Evershed2008, pp.
+Our results indicate that differences in pottery shape can be detected using EFA.
+The result of our MANOVA test shows there is a significant difference in shapes between the pre-European and Post-European periods, and between the pre-European and the Chinese periods.
+The average shape presents a round body with a wider rim and neck before European contact, and turns to a more oval-shaped body with narrower rim and neck after the European presence.
+These changes even more pronounced during the Chinese contact period.
+In general, pottery vessels become shorter in height over time, that leading to an oval-shaped body.
+
+Our CV tests indicates that there are significant differences in shape standardisation between the pre-European period and post-European period, and between the Chinese contact and either pre-European or post-European periods.
+This suggests pottery shape became more homogeneous and standardized after foreign contacts with European colonizers and even moreso after contact with Chinese immigrants.
+Compositional analysis shows that the clay pastes are similar, regardless of the increasing standardization of the pottery shape, reflecting continuity in the raw material sources.
+We can thus rule out changes in clay fabric as a factor in changes in vessel shape.
+To determine if shape changes might be related to changes in the function of pots at Kiwulan, we attempted to identify lipids trapped in the fabric of potsherds to identify foods that may have contributed residues absorbed into the clay <!-- add citation of work that we are trying to follow here, eg.
+Seung's paper in JIPA -->.
+Unfortunately, we did not obtain useful results due to very low lipid yields, which might result from the very thin, dense, and low porosity fabric of rhte Kiwulan pottery.
+These physical characteristics of the clay offers limited spaces to trap and protect organic molecules from microbiological degradation [cf.
+@Evershed2008, pp.
 909].
 
-The results support part of our hypothesis that foreign influences had impacts on the emergence of social inequality in local indigenous society reflected by the degree of pottery standardization, however, stronger in the Chinese period instead of the European period originally thought to be.
+Our results support the hypothesis that foreign influences at Kiwulan influenced emergence of social inequality in the local Indigenous society.
+We find that , stronger in the Chinese period instead of the European period originally thought to be.
 The hypothesis of the emergence of social inequality based on the monopolies for long-distance trade brought by the Europeans was not fully supported.

### 2020-01-27 analysis/paper/paper.Rmd
repair the methods section, we only want to mention the MSLR test for now

@@
 
-Ceramics analyzed in this paper come from 40 units (4m by 4m) sampled from the central, undisturbed area of archaeological excavations at Kiwulan (Figure \@ref(fig:KWL-map); Figure \@ref(fig:KWL-sam-area)).
+Ceramics analyzed in this paper come from 40 units (4m by 4m each) sampled from the central, undisturbed area of archaeological excavations at Kiwulan (Figure \@ref(fig:KWL-map); Figure \@ref(fig:KWL-sam-area)).
 Kiwulan is situated on a hill near a riverside at the northern margin of Yilan Province, which is characterized by a triangular alluvial plain facing eastwards the Pacific with high mountains on three other sides.
@@
 The composition matches the mineralogical composition of local raw materials in the Yilan Plain [@Chen2016].
-There is no substantial changes in the inclusions over time, indicating continuity in pottery fabric composition across the three periods.
+There are no substantial changes in the inclusions over time, indicating continuity in pottery fabric composition across the three periods.
 
@@
 
-We used scanned pottery drawings acquired from Bureau of Cultural Affairs in Yilan.
+We used scans of pottery drawings acquired from the Bureau of Cultural Affairs in Yilan.
 All drawings present a two-dimensional view of vessel cross-sections with indications of metric measurements.
@@
 We take this as our measurement of standardisation in vessel shape: lower CV values reflect higher standardisation, and thus increased craft specialisation in the community.
-To answer the question of whether CV values for vessels samples across our three occupational phases are significantly different or not, we used the "Asymptotic test for the equality of coefficients of variation from k populations" [@Feltz1996], a significance test for CVs provided by the R package cvequality [@Marwick2019].
-Our implementation is based on the following from @Feltz1996:
-
-Let $s_i$ be the observed standard deviation of the $i$th sample (or group of measurements), $x_i$ be the observed mean of the $i$th sample, and let
-
-$$m_i=n_i-1$$
-
-and we do not know the population $c_v$, so let $\widehat{\tau }_c$ be an estimate as follows:
-
-$$\widehat{\tau }_c = \frac{\left( \sum _{j=1}^km_j\frac{S_j}{\bar{x}_j} \right)}{M}$$ 
-where 
-$$M=\sum _{j=1}^m m_j$$
-
-Then the test statistic can be computed with:
-
-$$\begin{aligned} D'A D= \widehat{\tau }_c^{-2}\left( 0.5+\widehat{\tau }_c^2\right) ^{-1}\left[ \sum _{i=1}^km_i\left(\frac{s_i}{\bar{x}_i} \right)^2-\frac{\widehat{\tau }_c^2}{M} \right] \end{aligned}$$
-
-The $D'AD$ value measures how far each sample $c_v$ is from our estimate of the population $c_v$.
-@Feltz1996 note that the $D'AD$ value distributes as a central $\chi^2$ random variable with $k-1$ degrees of freedom, from which a p-value can be computed.
+To answer the question of whether CV values for vessels samples across our three occupational phases are significantly different or not, we used the the Modified signed-likelihood ratio (MSLR) test for equality of CVs [@Krishnamoorthy2014].
+While previous work has used the @Feltz1996 Asymptotic test for the equality of coefficients of variation from k populations, we prefer the MSLR test as a more recent development with lower rates of type I error, better performance with uneven sample numbers, and more power across a range of conditions [@Krishnamoorthy2014].
 
@@
 <!-- CV equality testing -->
+
 To compare pottery shape standardization across the three phases we investigate the distributions of the first three PC scores, taking the PC scores as proxy variables for vessel shape (Figure \@ref(fig:PCs-distribution)).

### 2020-01-27 analysis/paper/paper.Rmd
turning to continue editing the discussion section

@@
 To answer the question of whether CV values for vessels samples across our three occupational phases are significantly different or not, we used the the Modified signed-likelihood ratio (MSLR) test for equality of CVs [@Krishnamoorthy2014].
-While previous work has used the @Feltz1996 Asymptotic test for the equality of coefficients of variation from k populations, we prefer the MSLR test as a more recent development with lower rates of type I error, better performance with uneven sample numbers, and more power across a range of conditions [@Krishnamoorthy2014].
+While previous work has used the @Feltz1996 asymptotic test for the equality of coefficients of variation from k populations <!-- add citation to archaeology papers using the asymptotic test  -->, we prefer the MSLR test as a more recent development with lower rates of type I error, better performance with uneven sample numbers, and more power across a range of conditions [@Krishnamoorthy2014].
 

### 2020-01-27 analysis/paper/paper.Rmd
edits on discussion

@@
 Our CV tests indicates that there are significant differences in shape standardisation between the pre-European period and post-European period, and between the Chinese contact and either pre-European or post-European periods.
-This suggests pottery shape became more homogeneous and standardized after foreign contacts with European colonizers and even moreso after contact with Chinese immigrants.
+This suggests pottery shape became more homogeneous and standardized after foreign contacts with European colonizers and even more so after contact with Chinese immigrants.
 Compositional analysis shows that the clay pastes are similar, regardless of the increasing standardization of the pottery shape, reflecting continuity in the raw material sources.
-We can thus rule out changes in clay fabric as a factor in changes in vessel shape.
-To determine if shape changes might be related to changes in the function of pots at Kiwulan, we attempted to identify lipids trapped in the fabric of potsherds to identify foods that may have contributed residues absorbed into the clay <!-- add citation of work that we are trying to follow here, eg.
+We can thus rule out changes in clay fabric as a factor in explaining changes in vessel shape.
+To determine if shape changes might be related to changes in the function of pots at Kiwulan, we used geochemical methods to extract and identify lipids trapped in the fabric of potsherds to identify foods that may have contributed residues absorbed into the clay <!-- add citation of work that we are trying to follow here, eg.
 Seung's paper in JIPA -->.
-Unfortunately, we did not obtain useful results due to very low lipid yields, which might result from the very thin, dense, and low porosity fabric of rhte Kiwulan pottery.
+Unfortunately, we did not obtain useful results due to extremely low lipid yields, which is probably due to the very thin, dense, and low porosity fabric of Kiwulan pottery.
 These physical characteristics of the clay offers limited spaces to trap and protect organic molecules from microbiological degradation [cf.
@@
 Our results support the hypothesis that foreign influences at Kiwulan influenced emergence of social inequality in the local Indigenous society.
-We find that , stronger in the Chinese period instead of the European period originally thought to be.
-The hypothesis of the emergence of social inequality based on the monopolies for long-distance trade brought by the Europeans was not fully supported.
-The nature and context of the interaction with the foreign presences might give an insight into the different degrees of standardized pottery.
-Compared to other region in Taiwan, the European colonial control was weak in Yilan due to the relatively isolated location by mountains and the economic consideration of the Spanish and the Dutch who prefer northern Taiwan as a trading base.
-Yilan could be seen as a place experienced indirect influences by the European trade network and their colonial activities in a pericolonial context [cf.
+We find that foreign influence may have been stronger in the Chinese period than the European period, contrary to previous work <!-- add citation to previous work -->.
+The hypothesis of an emergence of social inequality based on the monopolies for long-distance trade brought by the Europeans was not fully supported.
+<!-- add more detail about why this hypothesis was not fully supported --> Compared to other regions in Taiwan, European colonial influence was weak in Yilan due to the isolated of the surrounding mountains and the economic focus of the Spanish and  Dutch who prefered northern Taiwan as their trading base.
+Indigenous communities Yilan experienced indirect influence from European trade networks and their colonial activities in a pericolonial context [cf.
 @Acabado2017].
-On the contrary, the interaction between indigenous people and the Chinese in the 19th century was more intense and direct since those Chinese immigrants settled in Yilan and lived closely with indigenous societies.
-This direct influence is reflected by the archaeological evidence of a large amount of Chinese porcelains and distinctive architectural bricks and tiles used by Chinese [@Hsieh2009].
-Also, the finding of adoption of coffin in mortuary practice that viewed as an ethnic symbol for Chinese was observed [@Chen2007].
+In contrast to the Indigenous-European interactions at Kiwulan, the interaction between indigenous people and  Chinese in the 19th century appears to have been more intense and direct.
+Historical records indicate that  Chinese immigrants settled in Yilan and lived closely with Kiwulan Indigenous societies <!-- add citation that is the source of this information --> .
+This direct influence is reflected by the archaeological evidence of large amounts of Chinese porcelains and distinctive architectural bricks and tiles used by Chinese [@Hsieh2009].
+Similarly, burials at Kiwulan in this later phase show the adoption of coffins in mortuary practices, which @Chen2007 interprets as the adoption of a symbol of ethnic Chinese [@Chen2007].
 

### 2020-01-27 analysis/paper/paper.Rmd
lediting last para of intro. check that Rmd can knit, it can

@@
 
-<!-- statistical test of difference in shape decsribed in previous paragraph -->
-
-To test for differences in the distributions of shape variables indicated by the PC scores shown in Figure \ref(@fig:first-two-PCs-byplot)), we used a multivariate analysis of variance (MANOVA) test to compare pairwise combinations across our three occupation phases.
-Table \@ref(tab:MANOVA-table) shows significant differences in shape between Pre-European and Post-European phases (p = $`r pval_post_pre`$), and Pre-European and Chinese phases (p = $`r pval_chi_pre`$).
-Although there is considerable overlap of shape variables between the Pre-European and Post-European phases, their PC scores are significantly different from each other.
-There is little difference between vessel shapes in the Post-European and Chinese contact periods.
-
 ```{r MANOVA-table}
@@
 This direct influence is reflected by the archaeological evidence of large amounts of Chinese porcelains and distinctive architectural bricks and tiles used by Chinese [@Hsieh2009].
-Similarly, burials at Kiwulan in this later phase show the adoption of coffins in mortuary practices, which @Chen2007 interprets as the adoption of a symbol of ethnic Chinese [@Chen2007].
+Similarly, burials at Kiwulan in this later phase show the adoption of coffins in mortuary practices, which @Chen2007 interprets as the adoption of a symbol of ethnic Chinese.
 
-More standardized pottery shape without controlling the clay paste or technology suggests the shape as a preferred characteristic in pottery production during the Chinese period.
-However, the standardization is not reflected by most metric attributes.
-There is higher variation in thickness attributes compared to diameter attributes, indicating shape is the main attribute that presents homogeneity.
 In addition to the emergence of specialized groups indicative of changes in social organization, a further explanation could be that producing homogeneous pottery is an act of resistance to the Europeans intrusion and the Chinese immigrants by emphasizing the practice of traditional pottery production.

### 2020-01-27 analysis/paper/paper.Rmd
fix typos and edit introduction section based on comments

--- 2020-01-27 analysis/paper/paper.Rmd
@@
 Our aim is to investigate if there was increasing ceramic specialization resulting from Indigenous interaction with Europeans in the 17th century or Chinese in the 19th century, two major foreign influences in early historical Taiwan, that might indicate social changes in local indigenous societies.
-Previous reports of archaeological evidence in northeastern Taiwan descrbes increases in the use of prestige goods, such as trade ornaments, and uneven artefact distribution patterns in domestic areas and burial contexts at the time when the Indigenous communities encountered the Spanish and the Dutch in the 17th century.
+Previous reports of archaeological evidence in northeastern Taiwan describes increases in the use of prestige goods, such as trade ornaments, and uneven artefact distribution patterns in domestic areas and burial contexts at the time when the Indigenous communities encountered the Spanish and the Dutch in the 17th century.
 Here we show that in addition to the differential distribution of trade ornaments post-European contact, the locally-made ceramics become highly consistent in their form and shape, compared to other pottery throughout northeastern Taiwan.
@@
 However, because pottery vessels are typically have curved shapes, linear measurements have limited sensitivity to many kinds of shape variations.
-Thus, to capture subtle shape variations that might be relevant to standardization, a method that is senstive to non-linear aspects of shape variation is necessary.
+Thus, to capture subtle shape variations that might be relevant to standardization, a method that is sensitive to non-linear aspects of shape variation is necessary.
 Here we analyse pottery shape using geometric morphometric methods, a new approach for measuring the degree of standardization because it takes a statistical summary of the overall shape as variable for comparison [@Slice2007].
@@
 Our hypothesis is that if foreign influences had impacts on the emergence of social inequality in local Indigenous society due to monopolies of trade arising among a small number of Indigenous people, then we expect to see changes in social organization from a more corporate to a more network organization [@Feinman2000].
-We expect this corporate-network shift to manifest as increased specialization pottery manufacture, indicated archaeologically by icreased standardisation of pottery production at Kiwulan.
+We expect this corporate-network shift to manifest as increased specialization pottery manufacture, indicated archaeologically by increased standardisation of pottery production at Kiwulan.
 The emergence of craft specialization, here pottery specialization, is often related to changes in social organization towards a society with increasing status inequality.
-In this case, if the competition among Indigenous individual for foreign resources and trade partnership with European or Chinese colonizers gradually lead to increasing social inequality, then we expect the local ceramics will show more homogeneous shapes after contact due to the emergence of craft specialization caused by rgreater economic and social control by a small group of individuals at Kiwulan.
+In this case, if the competition among Indigenous individual for foreign resources and trade partnership with European or Chinese colonizers gradually lead to increasing social inequality, then we expect the local ceramics will show more homogeneous shapes after contact due to the emergence of craft specialization caused by greater economic and social control by a small group of individuals at Kiwulan.
 Our results are important for understanding the indirect influences of  European colonists on local Indigenous societies, which remains unclear in East Asia where colonial power had limited success compared to other places in Southeast Asia.
@@
 In those cases, outline approaches are more suitable for analyzing the overall shape of an object.
-One outline approache is the semi-landmarks method, also called sliding landmarks, which assigns points along the curve between two landmarks at defined intervals [@Bookstein1997; @Lawing2010].
+One outline approach is the semi-landmarks method, also called sliding landmarks, which assigns points along the curve between two landmarks at defined intervals [@Bookstein1997; @Lawing2010].
 Those semi-landmarks are allowed to slide along the curve to remove the effect of the arbitrary landmark spacing by minimizing either Procrustes distance or bending energy [@Bookstein1997; @Slice2007; @Gunz2013].
@@
 Their results suggest that some ceramic types from the Medio period (AD 1200-1450), were made by specialists based on highly standardized shapes.
-Among those highly standardized jars, some were made by attached specialists <!-- what does this mean, 'attached' to what?
-please clarify -->, while others were made by independent specialists according to their spatial distribution patterns.
-Similarity, @Selden2019 examined 28 Caddo bottle excavated at several sites (AD 500-1700) in northwest Louisiana <!-- using semi-landmarks or something else?
-please add -->.
+Among those highly standardized jars, some were made by attached specialists controlled by elites indicating the presence of social inequality, while others were made by independent specialists according to their spatial distribution patterns.
+Similarity, @Selden2019 examined 28 Caddo bottle excavated at several sites (AD 500-1700) in northwest Louisiana using semi-landmarks method.
 The results indicate a significant integration between the different parts of the bottles including rim, neck, body, and base that vary in a coordinated manner.
@@
 This paper focuses on the upper component, dated from AD 1350 to AD 1950, covering the late Iron Age and the historical period, which we define as the European presence in Taiwan in the early 17th century.
-The Dutch first occupied southern Taiwan in 1624 and then the Spanish occupied northern Taiwan in 1626 <!-- add citation for these dates -->.
+The Dutch first occupied southern Taiwan in 1624 and then the Spanish occupied northern Taiwan in 1626 [@Andrade2007].
 In 1642, the Spanish was expelled by the Dutch in northern Taiwan, who took over their forts at Helping Dau in Keelung, and in Tamsui.
@@
 The size of the particles ranges from 500 micron to 1300 micron.
-In general, the globular vessel farbric presents a mixture of fine, rounded argillite with a small part of rounded metasandstone and rounded to sub-angular monocrystalline quartz.
+In general, the globular vessel fabric presents a mixture of fine, rounded argillite with a small part of rounded metasandstone and rounded to sub-angular monocrystalline quartz.
 The composition matches the mineralogical composition of local raw materials in the Yilan Plain [@Chen2016].
@@
 To answer the question of whether CV values for vessels samples across our three occupational phases are significantly different or not, we used the the Modified signed-likelihood ratio (MSLR) test for equality of CVs [@Krishnamoorthy2014].
-While previous work has used the @Feltz1996 asymptotic test for the equality of coefficients of variation from k populations <!-- add citation to archaeology papers using the asymptotic test  -->, we prefer the MSLR test as a more recent development with lower rates of type I error, better performance with uneven sample numbers, and more power across a range of conditions [@Krishnamoorthy2014].
+While previous work has used the @Feltz1996 Asymptotic test for the equality of coefficients of variation from k populations, we prefer the MSLR test as a more recent development with lower rates of type I error, better performance with uneven sample numbers, and more power across a range of conditions [@Krishnamoorthy2014].
 
@@
 
-To investigate the effect of the historically-documented decline of Indigenous population in the Chinese period <!-- add citation on population change during Chinese period --> on pottery production we examined the spatial distribution of pots in our sample in each period.
-Our concern is that the smaller population size during the Chinese period could confound our investigations into craft specialisation beause of reduced numbers of potters and pottery production shrinking a a few locations in the settlement during this time.
-Altough the overall number of vessels is smaller during the Chinese contact period, Figure \@ref(fig:pottery-spatial-pattern) shows that pottery is distributed randomly across the sampling area without any distinctive clusters.
+To investigate the effect of the historically-documented decline of Indigenous population in the Chinese period [@Chen2007; @Hsieh2009] on pottery production we examined the spatial distribution of pots in our sample in each period.
+Our concern is that the smaller population size during the Chinese period could confound our investigations into craft specialisation because of reduced numbers of potters and pottery production shrinking a a few locations in the settlement during this time.
+Although the overall number of vessels is smaller during the Chinese contact period, Figure \@ref(fig:pottery-spatial-pattern) shows that pottery is distributed randomly across the sampling area without any distinctive clusters.
 This suggests that differences in population across our three occupation phases are probably not driving variation in craft specialsation.

### 2020-01-27 R/rich-diffs-rmd-on-GitHub.R

@@ -21,8 +21,9 @@ browseURL(file.path(
 
 # Edit these to get the full SHA for the commits you want to compare, get
 # the full SHA from the GitHub webpage that was opened in the previous step
-commit_1_sha <- "8249ffc832acfb4839dfb9e5f838d45b82ca0063" # a recent commit
-commit_2_sha <- "a05cad1b678a31c9def1860cbceeba4cbfb98577" # an earlier commit
+commit_1_sha <- "3c25822c902819e6db7b88dee767b0b3d0235499" # a recent commit
+commit_2_sha <- "db390f98c0b4589c02f07006823b7c20ecad2856" # an earlier commit
+
 
 # Don't change this
 gh_raw <- "https://raw.githubusercontent.com"

### 2020-01-27 analysis/paper/paper.Rmd

@@
 Our results support the hypothesis that foreign influences at Kiwulan influenced emergence of social inequality in the local Indigenous society.
-We find that , stronger in the Chinese period instead of the European period originally thought to be.
-The hypothesis of the emergence of social inequality based on the monopolies for long-distance trade brought by the Europeans was not fully supported.
-The nature and context of the interaction with the foreign presences might give an insight into the different degrees of standardized pottery.
-Compared to other region in Taiwan, the European colonial control was weak in Yilan due to the relatively isolated location by mountains and the economic consideration of the Spanish and the Dutch who prefer northern Taiwan as a trading base.
-Yilan could be seen as a place experienced indirect influences by the European trade network and their colonial activities in a pericolonial context [cf.
+We find that foreign influence may have been stronger in the Chinese period than the European period, contrary to previous work <!-- add citation to previous work -->.
+The hypothesis of an emergence of social inequality based on the monopolies for long-distance trade brought by the Europeans was not fully supported.
+<!-- add more detail about why this hypothesis was not fully supported --> Compared to other regions in Taiwan, European colonial influence was weak in Yilan due to the isolated of the surrounding mountains and the economic focus of the Spanish and  Dutch who prefered northern Taiwan as their trading base.
+Indigenous communities Yilan experienced indirect influence from European trade networks and their colonial activities in a pericolonial context [cf.
 @Acabado2017].
-On the contrary, the interaction between indigenous people and the Chinese in the 19th century was more intense and direct since those Chinese immigrants settled in Yilan and lived closely with indigenous societies.
-This direct influence is reflected by the archaeological evidence of a large amount of Chinese porcelains and distinctive architectural bricks and tiles used by Chinese [@Hsieh2009].
-Also, the finding of adoption of coffin in mortuary practice that viewed as an ethnic symbol for Chinese was observed [@Chen2007].
+In contrast to the Indigenous-European interactions at Kiwulan, the interaction between indigenous people and  Chinese in the 19th century appears to have been more intense and direct.
+Historical records indicate that  Chinese immigrants settled in Yilan and lived closely with Kiwulan Indigenous societies <!-- add citation that is the source of this information --> .
+This direct influence is reflected by the archaeological evidence of large amounts of Chinese porcelains and distinctive architectural bricks and tiles used by Chinese [@Hsieh2009].
+Similarly, burials at Kiwulan in this later phase show the adoption of coffins in mortuary practices, which @Chen2007 interprets as the adoption of a symbol of ethnic Chinese.
 
-More standardized pottery shape without controlling the clay paste or technology suggests the shape as a preferred characteristic in pottery production during the Chinese period.
-However, the standardization is not reflected by most metric attributes.
-There is higher variation in thickness attributes compared to diameter attributes, indicating shape is the main attribute that presents homogeneity.
 In addition to the emergence of specialized groups indicative of changes in social organization, a further explanation could be that producing homogeneous pottery is an act of resistance to the Europeans intrusion and the Chinese immigrants by emphasizing the practice of traditional pottery production.

### 2020-01-27 analysis/paper/paper.Rmd
finished editing conclusion

--- 2020-01-27 analysis/paper/paper.Rmd
@@
 In contrast to the Indigenous-European interactions at Kiwulan, the interaction between indigenous people and  Chinese in the 19th century appears to have been more intense and direct.
-Historical records indicate that  Chinese immigrants settled in Yilan and lived closely with Kiwulan Indigenous societies <!-- add citation that is the source of this information --> .
+Historical records indicate that  Chinese immigrants settled in Yilan and lived closely with Kiwulan Indigenous societies <!-- add citation that is the source of this information -->.
 This direct influence is reflected by the archaeological evidence of large amounts of Chinese porcelains and distinctive architectural bricks and tiles used by Chinese [@Hsieh2009].
@@
 
-In addition to the emergence of specialized groups indicative of changes in social organization, a further explanation could be that producing homogeneous pottery is an act of resistance to the Europeans intrusion and the Chinese immigrants by emphasizing the practice of traditional pottery production.
-In a culture contact situation, social identity could be expressed through materials practices to show cultural homogeneity and distinction from other groups [@Voss2005].
-It is also important to recognize that social identity might be more complicated in a colonial context, not just a colonized–colonizer or local/foreign dichotomy [@Voss2005; @Voss2008].
-The case study of Shamaoshan cemetery (3BC- 4AD) in Southwest China showing the process of incorporation of Southwest China into the Han Empire involved a century of conflicts, resistance, and acceptance among social groups with different identities, especially in the historical context of Han immigrants [@Wu2019].
-At Kiwulan, the standardization of potter shape is significant in the 19th century when a a large number of Chinese immigrants migrated to Yilan.
-Shape can be viewed as a symbol as an expression the indigenous identity or social boundary because shape is the most obvious and visible trait compared with other features of pottery [cf.
+We recognise that the shape variation reported here is subtle, and this invites consideration of the possibility that the absence of major changes in vessel shape at Kiwulan may have been an act of resistance to foreign.
+Continuities in vessel shape over time draws our attention to the endurance of traditional pottery production practices amid intrusions from Europeans and Chinese.
+In a culture contact situation, social identity may be expressed through material practices to show cultural homogeneity and distinction from other groups [@Voss2005].
+It is also important to recognize that social identity might be more complicated in a colonial context, and may be more than a colonized–colonizer or local/foreign dichotomy [@Voss2005; @Voss2008].
+Shamaoshan cemetery (3BC- 4AD) in Southwest China shows that the process of the incorporation of Southwest China into the Han Empire involved a century of conflicts, resistance, and acceptance among social groups with different identities, especially in the historical context of Han immigrants [@Wu2019].
+A similar dynamic may have occured at Kiwulan, with vessel shape indicating both acceptance of forign influence through increased shape standardisation, and resistence through the overall continuity in vessel shape.
+Vessle shape can be viewed as a symbol as an expression the Indigenous identity and social boundaries because shape is a highly visible trait compared with other features of pottery [cf.
 @Roux2015].
-Although there is an increase in number of imported ceramics over time and reached high quantity and diversity in the 19th century, the production of the local pottery continued and became more standardized.
-This might imply not only the utilitarian function, but an deliberate action to emphasize the pottery tradition as their cultural custom.
+Although there is an increase in number of imported ceramics over time at Kiwulan, the production of the local pottery continued and became more standardized.
+This might imply not only a utilitarian function of the local vessels, but a deliberate action to emphasize the local pottery tradition as their cultural custom as foreign contact intensified.
 
@@
 
-This case study demonstrates the application of Elliptic Fourier analysis on ceramic shapes to explore the emergence of ceramic specialization and the underlying mechanism indicative of foreign cultural influences.
-Compared with the traditional linear measurement approach, geometric morphometrics methods especially EFA-based outline approach is more effective to detect shape variations for artifact shape without obvious landmarks.
-Here, EFA approach combined with a significance test for the equality of CVs for shape variables provides a robust way to identify and statistically assess the difference in shape standardization.
-Lower variation in ceramic shape was identified after the European presence and even much lower variation during the Chinese presence.
-The direct relationship between foreign influences and highly standardized ceramic shape was tested here on Kiwulan pottery.
-This finding helps to understand the cause and factor that would lead to the standardization of pottery production in a cross-cultural interaction context.
-This gave an insight into the discussion of ceramic specialization which is usually associated with the changes in social organization.
-The more homogeneous shape during the contact period without any changes in clay paste or producing technique suggests that shape standardization is an outcome resulted from an intentional act.
-The results here suggest an expression of social identity or showing cultural boundaries in indigenous societies through visible variables when encountering foreign intrusion.
-The analysis allows comparisons between sites with a similar historical trajectory in a pericolonial context.
-Our case study includes an openly available pipeline suitable for other assemblages, ensuring reproducibility and expanding the use of shape-based quantitative methods to questions about craft specialization and standardization in prehistoric technologies.
+This study demonstrates an application of elliptic Fourier analysis on ceramic shapes to explore the emergence of ceramic specialization as indicative of foreign influences.
+Here, EFA is combined with a significance test for the equality of CVs of shape variables to provide a robust way to identify and statistically assess differences in shape standardization.
+The direct relationship between foreign influences and standardization of ceramic shape was tested on pottery from Kiwulan, and large Iron Age Indienous settlement in northeast Taiwan.
+Lower variation in ceramic shape was identified after European presence began, and even lower variation during the period of Chinese presence.
+Our findings help to understand the factors that may lead to standardization of pottery production in a cross-cultural interaction context.
+More homogeneous shapes during the contact periods, without any changes in clay paste composition or production technique, suggests that shape standardization was intentional.
+The results here further suggest that expressions of social identity or cultural boundaries in Indigenous societies through highly visible vessel qualities, such as shape, may be heightened during periods of foreign intrusion in pericolonial contexts.
+Our analysis, with its openly available methods and data, is readily extensible to other pottery assemblages in the region to further explore related questions about craft specialization and standardization in Iron Age ceramic technologies.
 
@@
 This research used statistical consulting resources provided by the Center for Statistics and the Social Sciences, University of Washington.
+<!-- add thanks to QRC for funding for residue analysis, give lots of detail -->
 

### 2020-01-27 analysis/paper/paper.Rmd
add mention of corporate/network in discussion

@@
 This relationship between craft specialization and social differentiation has been broadly discussed, and many cases show that increasingly specialized production can be a useful indicator of increasing socio-political complexity [@Hirshman2010; @Junker1999].
-That said, more economic focus considers production, exchange, and consumption as a whole to explain specialization as part of a strategy to accommodate uneven distributions of resources [@Costin2001; @Arnold1994].
+That said, a more economic focus considers production, exchange, and consumption as a whole to explain specialization as part of a strategy to accommodate uneven distributions of resources [@Costin2001; @Arnold1994].
 Craft specialisation may appear as a reaction to changes in subsistence practices [@Stark1995economic], or as a result of long-distance trade and inter-regional exchange [@Alizadeh2018].
@@
 
-Our results support the hypothesis that foreign influences at Kiwulan influenced emergence of social inequality in the local Indigenous society.
-We find that foreign influence may have been stronger in the Chinese period than the European period, contrary to previous work <!-- add citation to previous work -->.
-The hypothesis of an emergence of social inequality based on the monopolies for long-distance trade brought by the Europeans was not fully supported.
-<!-- add more detail about why this hypothesis was not fully supported --> Compared to other regions in Taiwan, European colonial influence was weak in Yilan due to the isolated of the surrounding mountains and the economic focus of the Spanish and  Dutch who prefered northern Taiwan as their trading base.
+Our results offer tentative support for the hypothesis that foreign influences at Kiwulan influenced emergence of social inequality in the local Indigenous society.
+If increased shape standardisation is a reliable indicator of craft specialisation, then we may be seeing evidence of a shift from corporate to network organization [@Feinman2000].
+However, strong claims for an emergence of social complexity resulting from foreign contact at Kiwulan will need support from multiple and diverse sources of evidence.
+This should include the other two main datasets at Kiwulan, the ornaments assemblages, and the burial goods.
+We find vessel shapes were more standardised in the Chinese period than the European period, contrary to previous work that has downplayed the effect of Chinese settlement in Yilan <!-- add citation to previous work -->.
+Compared to other regions in Taiwan, European colonial influence was weak in Yilan due to the isolated of the surrounding mountains and the economic focus of the Spanish and  Dutch who prefered northern Taiwan as their trading base.
 Indigenous communities Yilan experienced indirect influence from European trade networks and their colonial activities in a pericolonial context [cf.

### 2020-01-27 analysis/paper/paper.Rmd
edit introduction and add references to method section

--- 2020-01-27 analysis/paper/paper.Rmd
@@
 Their results suggest that some ceramic types from the Medio period (AD 1200-1450), were made by specialists based on highly standardized shapes.
-Among those highly standardized jars, some were made by attached specialists controlled by elites indicating the presence of social inequality, while others were made by independent specialists according to their spatial distribution patterns.
+Among those highly standardized jars, some were made by specialists attached to elites indicating the presence of social inequality, while others were made by independent specialists according to their spatial distribution patterns.
 Similarity, @Selden2019 examined 28 Caddo bottle excavated at several sites (AD 500-1700) in northwest Louisiana using semi-landmarks method.
@@
 To compare different foreign influences, we classified the upper component into three chronological phases: the pre-European, European, and Chinese periods.
-We identified these phases using 32 previously published radiocarbon ages <!-- add citaion to report with dates -->, excavation depth measurements, stratigraphic details reported by the excavators (colour, disturbance, etc.), and finds of chronologically diagnostic artifacts, such as blue and white porcelains, light grey glazed jars, and large dark brown glazed stoneware jars commonly used in the 17th century, and bricks and tiles used by the Chinese in the 19th century [@Hsieh2009; @Wang2011].
+We identified these phases using 32 previously published radiocarbon ages [@Chen2007], excavation depth measurements, stratigraphic details reported by the excavators (colour, disturbance, etc.), and finds of chronologically diagnostic artifacts, such as blue and white porcelains, light grey glazed jars, and large dark brown glazed stoneware jars commonly used in the 17th century, and bricks and tiles used by the Chinese in the 19th century [@Hsieh2009; @Wang2011].
 The deposit shows signs of continuous human occupation during each of the three phases across the sampling area.
@@
 To answer the question of whether CV values for vessels samples across our three occupational phases are significantly different or not, we used the the Modified signed-likelihood ratio (MSLR) test for equality of CVs [@Krishnamoorthy2014].
-While previous work has used the @Feltz1996 asymptotic test for the equality of coefficients of variation from k populations <!-- add citation to archaeology papers using the asymptotic test  -->, we prefer the MSLR test as a more recent development with lower rates of type I error, better performance with uneven sample numbers, and more power across a range of conditions [@Krishnamoorthy2014].
+While previous work has used the @Feltz1996 asymptotic test for the equality of coefficients of variation from k populations [@Lycett2008; @Okumura2014; @Eerkens2000; @Eerkens2001; @Hoggard2017], we prefer the MSLR test as a more recent development with lower rates of type I error, better performance with uneven sample numbers, and more power across a range of conditions [@Krishnamoorthy2014].
 
@@
 Although the overall number of vessels is smaller during the Chinese contact period, Figure \@ref(fig:pottery-spatial-pattern) shows that pottery is distributed randomly across the sampling area without any distinctive clusters.
-This suggests that differences in population across our three occupation phases are probably not driving variation in craft specialsation.
+This suggests that differences in population across our three occupation phases are probably not driving variation in craft specialization.
 
@@
 In contrast to the Indigenous-European interactions at Kiwulan, the interaction between indigenous people and  Chinese in the 19th century appears to have been more intense and direct.
-Historical records indicate that  Chinese immigrants settled in Yilan and lived closely with Kiwulan Indigenous societies <!-- add citation that is the source of this information -->.
+Historical records indicate that  Chinese immigrants settled in Yilan and lived closely with Kiwulan Indigenous societies <!-- add citation that is the source of this information --> .
 This direct influence is reflected by the archaeological evidence of large amounts of Chinese porcelains and distinctive architectural bricks and tiles used by Chinese [@Hsieh2009].
@@
 
-We recognise that the shape variation reported here is subtle, and this invites consideration of the possibility that the absence of major changes in vessel shape at Kiwulan may have been an act of resistance to foreign.
-Continuities in vessel shape over time draws our attention to the endurance of traditional pottery production practices amid intrusions from Europeans and Chinese.
-In a culture contact situation, social identity may be expressed through material practices to show cultural homogeneity and distinction from other groups [@Voss2005].
-It is also important to recognize that social identity might be more complicated in a colonial context, and may be more than a colonized–colonizer or local/foreign dichotomy [@Voss2005; @Voss2008].
-Shamaoshan cemetery (3BC- 4AD) in Southwest China shows that the process of the incorporation of Southwest China into the Han Empire involved a century of conflicts, resistance, and acceptance among social groups with different identities, especially in the historical context of Han immigrants [@Wu2019].
-A similar dynamic may have occured at Kiwulan, with vessel shape indicating both acceptance of forign influence through increased shape standardisation, and resistence through the overall continuity in vessel shape.
-Vessle shape can be viewed as a symbol as an expression the Indigenous identity and social boundaries because shape is a highly visible trait compared with other features of pottery [cf.
+In addition to the emergence of specialized groups indicative of changes in social organization, a further explanation could be that producing homogeneous pottery is an act of resistance to the Europeans intrusion and the Chinese immigrants by emphasizing the practice of traditional pottery production.
+In a culture contact situation, social identity could be expressed through materials practices to show cultural homogeneity and distinction from other groups [@Voss2005].
+It is also important to recognize that social identity might be more complicated in a colonial context, not just a colonized–colonizer or local/foreign dichotomy [@Voss2005; @Voss2008].
+The case study of Shamaoshan cemetery (3BC- 4AD) in Southwest China showing the process of incorporation of Southwest China into the Han Empire involved a century of conflicts, resistance, and acceptance among social groups with different identities, especially in the historical context of Han immigrants [@Wu2019].
+At Kiwulan, the standardization of potter shape is significant in the 19th century when a a large number of Chinese immigrants migrated to Yilan.
+Shape can be viewed as a symbol as an expression the indigenous identity or social boundary because shape is the most obvious and visible trait compared with other features of pottery [cf.
 @Roux2015].
-Although there is an increase in number of imported ceramics over time at Kiwulan, the production of the local pottery continued and became more standardized.
-This might imply not only a utilitarian function of the local vessels, but a deliberate action to emphasize the local pottery tradition as their cultural custom as foreign contact intensified.
+Although there is an increase in number of imported ceramics over time and reached high quantity and diversity in the 19th century, the production of the local pottery continued and became more standardized.
+This might imply not only the utilitarian function, but an deliberate action to emphasize the pottery tradition as their cultural custom.
 
@@
 
-This study demonstrates an application of elliptic Fourier analysis on ceramic shapes to explore the emergence of ceramic specialization as indicative of foreign influences.
-Here, EFA is combined with a significance test for the equality of CVs of shape variables to provide a robust way to identify and statistically assess differences in shape standardization.
-The direct relationship between foreign influences and standardization of ceramic shape was tested on pottery from Kiwulan, and large Iron Age Indienous settlement in northeast Taiwan.
-Lower variation in ceramic shape was identified after European presence began, and even lower variation during the period of Chinese presence.
-Our findings help to understand the factors that may lead to standardization of pottery production in a cross-cultural interaction context.
-More homogeneous shapes during the contact periods, without any changes in clay paste composition or production technique, suggests that shape standardization was intentional.
-The results here further suggest that expressions of social identity or cultural boundaries in Indigenous societies through highly visible vessel qualities, such as shape, may be heightened during periods of foreign intrusion in pericolonial contexts.
-Our analysis, with its openly available methods and data, is readily extensible to other pottery assemblages in the region to further explore related questions about craft specialization and standardization in Iron Age ceramic technologies.
+This case study demonstrates the application of Elliptic Fourier analysis on ceramic shapes to explore the emergence of ceramic specialization and the underlying mechanism indicative of foreign cultural influences.
+Compared with the traditional linear measurement approach, geometric morphometrics methods especially EFA-based outline approach is more effective to detect shape variations for artifact shape without obvious landmarks.
+Here, EFA approach combined with a significance test for the equality of CVs for shape variables provides a robust way to identify and statistically assess the difference in shape standardization.
+Lower variation in ceramic shape was identified after the European presence and even much lower variation during the Chinese presence.
+The direct relationship between foreign influences and highly standardized ceramic shape was tested here on Kiwulan pottery.
+This finding helps to understand the cause and factor that would lead to the standardization of pottery production in a cross-cultural interaction context.
+This gave an insight into the discussion of ceramic specialization which is usually associated with the changes in social organization.
+The more homogeneous shape during the contact period without any changes in clay paste or producing technique suggests that shape standardization is an outcome resulted from an intentional act.
+The results here suggest an expression of social identity or showing cultural boundaries in indigenous societies through visible variables when encountering foreign intrusion.
+The analysis allows comparisons between sites with a similar historical trajectory in a pericolonial context.
+Our case study includes an openly available pipeline suitable for other assemblages, ensuring reproducibility and expanding the use of shape-based quantitative methods to questions about craft specialization and standardization in prehistoric technologies.
 
@@
 This research used statistical consulting resources provided by the Center for Statistics and the Social Sciences, University of Washington.
-<!-- add thanks to QRC for funding for residue analysis, give lots of detail -->
 

### 2020-01-27 analysis/paper/paper.Rmd
add references to discussion session and edit acknowledgment

@@
 
-Previous investigations at Kiwulan have suggested an unequal distribution of prestige goods, trade ornaments specifically, following the appearance of Europeans <!-- add citation to that previous investigation -->, hinting at the emergence of social inequality in the Indigenous community.
+Previous investigations at Kiwulan have suggested an unequal distribution of prestige goods, trade ornaments specifically, following the appearance of Europeans [@Cheng2008; @Wang2011], hinting at the emergence of social inequality in the Indigenous community.
 Here we have examined ceramic vessel shape standardization to measure craft specialization as a proxy for social differentiation, such as elite control [@Costin2001; @Junker1999].
@@
 We can thus rule out changes in clay fabric as a factor in explaining changes in vessel shape.
-To determine if shape changes might be related to changes in the function of pots at Kiwulan, we used geochemical methods to extract and identify lipids trapped in the fabric of potsherds to identify foods that may have contributed residues absorbed into the clay <!-- add citation of work that we are trying to follow here, eg.
-Seung's paper in JIPA -->.
+To determine if shape changes might be related to changes in the function of pots at Kiwulan, we used geochemical methods to extract and identify lipids trapped in the fabric of potsherds to identify foods that may have contributed residues absorbed into the clay [cf.
+@Kwak2015].
 Unfortunately, we did not obtain useful results due to extremely low lipid yields, which is probably due to the very thin, dense, and low porosity fabric of Kiwulan pottery.
@@
 This should include the other two main datasets at Kiwulan, the ornaments assemblages, and the burial goods.
-We find vessel shapes were more standardised in the Chinese period than the European period, contrary to previous work that has downplayed the effect of Chinese settlement in Yilan <!-- add citation to previous work -->.
-Compared to other regions in Taiwan, European colonial influence was weak in Yilan due to the isolated of the surrounding mountains and the economic focus of the Spanish and  Dutch who prefered northern Taiwan as their trading base.
+We find vessel shapes were more standardised in the Chinese period than the European period, contrary to previous work that has downplayed the effect of Chinese settlement in Yilan [@Wang2011].
+Compared to other regions in Taiwan, European colonial influence was weak in Yilan due to the isolated of the surrounding mountains and the economic focus of the Spanish and  Dutch who preferred northern Taiwan as their trading base.
 Indigenous communities Yilan experienced indirect influence from European trade networks and their colonial activities in a pericolonial context [cf.
@@
 In contrast to the Indigenous-European interactions at Kiwulan, the interaction between indigenous people and Chinese in the 19th century appears to have been more intense and direct.
-Historical records indicate that  Chinese immigrants settled in Yilan and lived closely with Kiwulan Indigenous societies <!-- add citation that is the source of this information -->.
+Historical records indicate that Chinese immigrants settled in Yilan and lived closely with Kiwulan Indigenous societies [@Chen1963; @Ke1993].
 This direct influence is reflected by the archaeological evidence of large amounts of Chinese porcelains and distinctive architectural bricks and tiles used by Chinese [@Hsieh2009].
@@
 Shamaoshan cemetery (3BC- 4AD) in Southwest China shows that the process of the incorporation of Southwest China into the Han Empire involved a century of conflicts, resistance, and acceptance among social groups with different identities, especially in the historical context of Han immigrants [@Wu2019].
-A similar dynamic may have occured at Kiwulan, with vessel shape indicating both acceptance of forign influence through increased shape standardisation, and resistence through the overall continuity in vessel shape.
-Vessle shape can be viewed as a symbol as an expression the Indigenous identity and social boundaries because shape is a highly visible trait compared with other features of pottery [cf.
+A similar dynamic may have occurred at Kiwulan, with vessel shape indicating both acceptance of foreign influence through increased shape standardisation, and resistance through the overall continuity in vessel shape.
+Vessel shape can be viewed as a symbol as an expression the Indigenous identity and social boundaries because shape is a highly visible trait compared with other features of pottery [cf.
 @Roux2015].
@@
 Wen-Shan Chen in the Department of Geosciences, National Taiwan University for his invaluable guidance of petrographic analysis at his lab.
+We thank the Quaternary Research Center funding for supporting the organic residue analysis in this project.
+We thank Dr.
+Julian Sachs in the Department of Oceanography, University of Washington for his supports and providing his lab for us to conduct geochemical analysis of potsherds.
+We thank Dr.
+Matthew Wolhowe for his help in developing protocols for lipid extraction and his assistance with the GC-MS, GC-FID, and GC-C-IRMS analyses.
 This research used statistical consulting resources provided by the Center for Statistics and the Social Sciences, University of Washington.
-<!-- add thanks to QRC for funding for residue analysis, give lots of detail -->
 

### 2020-01-27 analysis/paper/paper.Rmd
remove figure 4 and 6, edit the relevant text

@@
 The results show the variation indicated by EFA and DCT matches the traditional ceramic typology, which supports that claim that outline-based approaches can be efficiently used for studying variations in ceramic shapes.
-Furthermore, @Wilczek201's EFA results help us understand the level of production standardization over time across the region.
+Furthermore, @Wilczek2014's EFA results help us understand the level of production standardization over time across the region.
 Those examples show that outline approaches can distinguish variation in ceramic shapes at high resolutions to understand ceramic production.
@@
 Reliable pottery outlines were captured by 13 harmonics that gather 99% of the total harmonic power.
-The average shape of vessels from each phase was visualized by computing the mean of the standardized Fourier coefficients within each phase group (Figure \@ref(fig:EFA-analysis)).
 Figure \@ref(fig:thin-plate-splines) shows the shape changes described using thin-plate spline warping for paired periods, pre- and post-European periods, and post-European and Chinese periods.
+The average shape of vessels from each phase was visualized and it shows more obvious differences between post-European and Chinese periods.
 
-```{r EFA-analysis, fig.cap= "Mean pottery shapes of three phases"}
+```{r thin-plate-splines, fig.cap= "Compare the average shapes across phases using thin plate splines (TPS) with outline deformations required to pass from an extreme of the morphospace to the other."}
+# Thin plate spline analysis (TPS)
+# visualize the outline deformations required to pass from an extreme of the morphospace to the other
+# Deformation grids
+
+# Mean pottery shapes of three phases
 library(Momocs)
@@
 
-# default is 13 harmanics that gather 99% of the harmonic power 
+# default is 13 harmonics that gather 99% of the harmonic power
 op <- efourier(jpgs_imported_coo, norm = FALSE, 13)
@@
 
-`Post-European` <- mshp$poe %T>% coo_plot(border="blue")
-`Pre-European` <-  mshp$pre %T>% coo_plot(border="red", plot.new=FALSE)
-`Chinese` <-       mshp$chi %T>% coo_plot(border="green", plot.new=FALSE)
-legend("topright", 
-       lwd=1,
-       col=c("red", "blue", "green"), 
-       legend=c( "Pre-European", "Post-European", "Chinese")) 
-#? How to put the legend out of the plot, tried change the margin using par(), but doesn't work
-```
-
-```{r thin-plate-splines, fig.cap= "Compare the average shapes across phases using thin plate splines (TPS) with outline deformations required to pass from an extreme of the morphospace to the other."}
-# Thin plate spline analysis (TPS)
-# visualize the outline deformations required to pass from an extreme of the morphospace to the other
-# Deformation grids
+`Post-European` <- mshp$poe 
+`Pre-European` <-  mshp$pre 
+`Chinese` <-       mshp$chi
 
@@
 With the third component, the first three principal components explain `r vari_three`% of the total variance.
-Figure \@ref(fig:contribution-PCs) shows the shape variation associated with the first three principal components.
-PC1 captures the height of the vessels, from tall to short, and the roundness of the body from round to oval-shaped.
+According to Figure \@ref(fig:first-two-PCs-byplot), PC1 captures the height of the vessels, from tall to short, and the roundness of the body from round to oval-shaped.
 PC2 relates to the neck and mouth constriction, from narrow to wide.
 PC3 explains a smaller portion of the variance (`r pc3`%), which relates to the degree of the flare of the neck, from curved to straight shape.
-The first two components account for most of the variance in relation to three phases (Figure \@ref(fig:first-two-PCs-byplot)).
+The first two components account for most of the variance in relation to three phases.
 The results reflect a large overlap in shapes from three occupations phases, especially for shapes in the pre-European and post-European periods.
@@
 In other words, we find a decrease in shape variance in the Chinese period that is evident in the shorter height and narrower mouth of vessels used in that period.
-
-```{r contribution-PC-hide, results="hide", include= FALSE}
-# the contributions of each PC to the shapes
-pc_contrib <-  PCcontrib(op.p, nax = 1:3) 
-```
-
-```{r contribution-PCs, fig.cap= "Pottery shape variation based on scores along principal component axes.
-The rows from the top to the bottom represent the first to third principal component respectively.
-Each column shows the reconstructed shape variable according to the score mean ±2 standard deviation (SD) with a 0.5 interval"}
-pc_contrib$gg +  
-  geom_polygon(fill="slategrey", col="black") + 
-  theme_minimal(base_size = 6)
-
-ggsave(here::here("analysis/figures/pca-contrib.png"))
-
-# PC1 is height of vessel
-# PC2 is neck constriction
-# PC3 is the degree of flare of neck
-```
 
@@
 
-To test for differences in the distributions of shape variables indicated by the PC scores shown in Figure \ref(@fig:first-two-PCs-byplot)), we used a multivariate analysis of variance (MANOVA) test to compare pairwise combinations across our three occupation phases.
+To test for differences in the distributions of shape variables indicated by the PC scores shown in Figure \@ref(fig:first-two-PCs-byplot), we used a multivariate analysis of variance (MANOVA) test to compare pairwise combinations across our three occupation phases.
 Table \@ref(tab:MANOVA-table) shows significant differences in shape between Pre-European and Post-European phases (p = $`r pval_post_pre`$), and Pre-European and Chinese phases (p = $`r pval_chi_pre`$).
@@
 We thank Dr.
-Julian Sachs in the Department of Oceanography, University of Washington for his supports and providing his lab for us to conduct geochemical analysis of potsherds.
+Julian Sachs in the Department of Oceanography, University of Washington for his supports and providing his lab for us to conduct organic geochemistry analysis of potsherds.
 We thank Dr.

### 2020-01-27 analysis/paper/shapes.R

@@ -27,6 +27,48 @@ tps_iso(`Pre-European`, `Chinese`,  shp.lwd = c(2, 2), shp.border = c("red", "bl
         palette = col_summer, grid = F)
 dev.off()
 
+# Mean pottery shapes of three phases
+library(Momocs)
+library(cowplot)
+jpgs_imported_coo <- readr::read_rds(here::here("analysis/data/derived_data/jpgs_imported_coo.rds"))
+
+# default is 13 harmonics that gather 99% of the harmonic power
+op <- efourier(jpgs_imported_coo, norm = FALSE, 13)
+
+# mean shape, per group
+pot.ms <- MSHAPES(op, 'phase')
+mshp <- pot.ms$shp
+names(mshp) <- c("chi", "poe", "pre")
+
+`Post-European` <- mshp$poe %T>% coo_plot(border="blue")
+`Pre-European` <-  mshp$pre %T>% coo_plot(border="red", plot.new=FALSE)
+`Chinese` <-       mshp$chi %T>% coo_plot(border="green", plot.new=FALSE)
+legend("topright",
+       lwd=1,
+       col=c("red", "blue", "green"),
+       legend=c( "Pre-European", "Post-European", "Chinese"))
+# The average shape of vessels from each phase was visualized by computing the mean of
+# the standardized Fourier coefficients within each phase group (Figure 4).
+#? How to put the legend out of the plot, tried change the margin using par(), but doesn't work
+
+
+# contribution-PC-hide
+# the contributions of each PC to the shapes
+pc_contrib <-  PCcontrib(op.p, nax = 1:3)
+
+# contribution-PCs, fig.cap= "Pottery shape variation based on scores along principal component axes.
+# The rows from the top to the bottom represent the first to third principal component respectively.
+# Each column shows the reconstructed shape variable according to the score mean ±2 standard deviation (SD) with a 0.5 interval"}
+pc_contrib$gg +
+  geom_polygon(fill="slategrey", col="black") +
+  theme_minimal(base_size = 6)
+
+ggsave(here::here("analysis/figures/pca-contrib.png"))
+
+# PC1 is height of vessel
+# PC2 is neck constriction
+# PC3 is the degree of flare of neck
+
 # PCA byplot, different visualization
 plot(PCA(op), 2, pos.shp="full")
 plot(PCA(op), 2, pos.shp="range")

### 2020-01-29 analysis/paper/paper.Rmd
Some more comments from BM

@@
 
-A total of `r total_no` largely intact vessels from pre-contact contexts (n = 32), post-European contexts (n = 27), and Chinese contact contexts (n = 14) were selected for ellipic Fourier analysis.
+A total of `r total_no` largely intact vessels from pre-contact contexts (n = 32), post-European contexts (n = 27), and Chinese contact contexts (n = 14) were selected for elliptic Fourier analysis.
 Most of them cover every vessel component from rim to bottom.
@@
 To obtain information about the pottery shape, we also compare the ratio of Rim/Body thickness, and the ratio of Rim/Body diameter.
+<!-- we also look at spatial distribution to see if clustering changes over time...
+...as craft specialization increases, we might expect spatial clustering to increase...
+as possible indicator of craft specialization -->
 
@@
 We take this as our measurement of standardisation in vessel shape: lower CV values reflect higher standardisation, and thus increased craft specialisation in the community.
+<!-- note that our samples are different in size,, say something about how the CV values are robust to different sample sizes, see if there is a citation that is relevant here -->
+
 To answer the question of whether CV values for vessels samples across our three occupational phases are significantly different or not, we used the the Modified signed-likelihood ratio (MSLR) test for equality of CVs [@Krishnamoorthy2014].
@@
 To investigate the effect of the historically-documented decline of Indigenous population in the Chinese period [@Chen2007; @Hsieh2009] on pottery production we examined the spatial distribution of pots in our sample in each period.
-Our concern is that the smaller population size during the Chinese period could confound our investigations into craft specialisation because of reduced numbers of potters and pottery production shrinking a a few locations in the settlement during this time.
+Our concern is that the smaller population size during the Chinese period could confound our investigations into craft specialisation because of reduced numbers of potters and pottery production shrinking a few locations in the settlement during this time.
 Although the overall number of vessels is smaller during the Chinese contact period, Figure \@ref(fig:pottery-spatial-pattern) shows that pottery is distributed randomly across the sampling area without any distinctive clusters.
@@
 
-<!-- statistical test of difference in shape decsribed in previous paragraph -->
+<!-- statistical test of difference in shape described in previous paragraph -->
 
@@
 
+<!-- a few lines to compare metric to shape... main point to make clear is the substantive result about the main research questions -->
+
 Our results offer tentative support for the hypothesis that foreign influences at Kiwulan influenced emergence of social inequality in the local Indigenous society.
@@
 
-We recognise that the shape variation reported here is subtle, and this invites consideration of the possibility that the absence of major changes in vessel shape at Kiwulan may have been an act of resistance to foreign.
+We recognise that the shape variation reported here is subtle, and this invites consideration of the possibility that the absence of major changes in vessel shape at Kiwulan may have been an act of resistance to foreign influence.
 Continuities in vessel shape over time draws our attention to the endurance of traditional pottery production practices amid intrusions from Europeans and Chinese.

### 2020-01-29 R/rich-diffs-rmd-in-local-git-repo.R

@@ -1,0 +1,42 @@
+# How to see rich diffs of two commits of a single R Markdown document in a local Git repo
+
+# https://github.com/lorenzwalthert/gitsum
+library("gitsum")
+library("tidyverse")
+
+# To browse the commits locally, open an RStudio
+# project that is using Git version control, then ...
+
+# Set the path within the project to the Rmd file
+# in the project that we want to compare versions of
+file_path <- "analysis/paper/paper.Rmd"
+
+# Make a data frame that stores the git commit history,
+# one row per commit
+library(gitsum)
+gitsum::init_gitsum()
+tbl_out <- parse_log_detailed() %>%
+  select(date, author_name, message, hash)
+
+# View this dataframe to pick the specific commits to compare
+View(tbl_out)
+
+# choose two commits to compare, here I've got the most recent
+# one (the last row of the table), and the one immediately before that
+i <- nrow(tbl_out) - 1
+j <- nrow(tbl_out) - 2
+
+# consrtruct the terminal commants using the SHA of the commits we want to
+# compare, and the file path
+git_show_str_1 <- str_glue('git show ', tbl_out[i, ]$hash, ":", file_path)
+git_show_str_2 <- str_glue('git show ', tbl_out[j, ]$hash, ":", file_path)
+
+# get the contents of the file at each commit and
+# store in a vector
+earlier_commit <- system(git_show_str_1, intern = TRUE)
+later_commit <- system(git_show_str_2, intern = TRUE)
+
+# compute differences and visualise
+diffobj::diffChr(earlier_commit,
+                 later_commit,
+                 mode = "sidebyside")

### 2020-01-29 analysis/paper/paper.Rmd
edit method section

@@
 However, because pottery vessels are typically have curved shapes, linear measurements have limited sensitivity to many kinds of shape variations.
-Thus, to capture subtle shape variations that might be relevant to standardization, a method that is sensitive to non-linear aspects of shape variation is necessary.
+Thus, to capture subtle shape variations that might be relevant to standardization, a method that is senstive to non-linear aspects of shape variation is necessary.
 Here we analyse pottery shape using geometric morphometric methods, a new approach for measuring the degree of standardization because it takes a statistical summary of the overall shape as variable for comparison [@Slice2007].
@@
 Our hypothesis is that if foreign influences had impacts on the emergence of social inequality in local Indigenous society due to monopolies of trade arising among a small number of Indigenous people, then we expect to see changes in social organization from a more corporate to a more network organization [@Feinman2000].
-We expect this corporate-network shift to manifest as increased specialization pottery manufacture, indicated archaeologically by increased standardisation of pottery production at Kiwulan.
+We expect this corporate-network shift to manifest as increased specialization pottery manufacture, indicated archaeologically by icreased standardisation of pottery production at Kiwulan.
 The emergence of craft specialization, here pottery specialization, is often related to changes in social organization towards a society with increasing status inequality.
-In this case, if the competition among Indigenous individual for foreign resources and trade partnership with European or Chinese colonizers gradually lead to increasing social inequality, then we expect the local ceramics will show more homogeneous shapes after contact due to the emergence of craft specialization caused by greater economic and social control by a small group of individuals at Kiwulan.
+In this case, if the competition among Indigenous individual for foreign resources and trade partnership with European or Chinese colonizers gradually lead to increasing social inequality, then we expect the local ceramics will show more homogeneous shapes after contact due to the emergence of craft specialization caused by rgreater economic and social control by a small group of individuals at Kiwulan.
 Our results are important for understanding the indirect influences of  European colonists on local Indigenous societies, which remains unclear in East Asia where colonial power had limited success compared to other places in Southeast Asia.
@@
 For each pot, two measurements for thickness and diameter were took to compute an average value.
-To obtain information about the pottery shape, we also compare the ratio of Rim/Body thickness, and the ratio of Rim/Body diameter.
+To obtain information about the pottery shape, we compare the ratio of Rim/Body thickness, and the ratio of Rim/Body diameter.
+Also, we explore spatial distribution of pottery to examine if clustering changes over time across the sampling area.
+As craft specialization increases, pottery distribution might present a shift from dispersed clusters to concentrated clusters that hints the emergence of specialized group working in workshop.
+
 <!-- we also look at spatial distribution to see if clustering changes over time...

### 2020-01-29 R/rich-diffs-rmd-in-local-git-repo.R

@@ -13,8 +13,7 @@ file_path <- "analysis/paper/paper.Rmd"
 
 # Make a data frame that stores the git commit history,
 # one row per commit
-library(gitsum)
-gitsum::init_gitsum()
+init_gitsum()
 tbl_out <- parse_log_detailed() %>%
   select(date, author_name, message, hash)
 
@@ -23,8 +22,8 @@ View(tbl_out)
 
 # choose two commits to compare, here I've got the most recent
 # one (the last row of the table), and the one immediately before that
-i <- nrow(tbl_out) - 1
-j <- nrow(tbl_out) - 2
+i <- nrow(tbl_out) - 30
+j <- nrow(tbl_out) - 31
 
 # consrtruct the terminal commants using the SHA of the commits we want to
 # compare, and the file path
@@ -33,8 +32,8 @@ git_show_str_2 <- str_glue('git show ', tbl_out[j, ]$hash, ":", file_path)
 
 # get the contents of the file at each commit and
 # store in a vector
-earlier_commit <- system(git_show_str_1, intern = TRUE)
-later_commit <- system(git_show_str_2, intern = TRUE)
+earlier_commit <- system(git_show_str_2, intern = TRUE)
+later_commit <- system(git_show_str_1, intern = TRUE)
 
 # compute differences and visualise
 diffobj::diffChr(earlier_commit,

### 2020-01-31 analysis/paper/paper.Rmd
edit statistical analysis section

@@
 
-By standardizing standard deviation in each data set, the coefficients of variation values allows us to directly compare variation in samples measured with different units or means.
-We take this as our measurement of standardisation in vessel shape: lower CV values reflect higher standardisation, and thus increased craft specialisation in the community.
-<!-- note that our samples are different in size,, say something about how the CV values are robust to different sample sizes, see if there is a citation that is relevant here -->
+By standardizing standard deviation in each data set, the coefficients of variation (CV) values allows us to directly compare variation in samples measured with different units or means.
+CV values are robust to different sample sizes in the condition that the mean-standard deviation relationship is linear and positive since standard deviation is scaled to the mean [@Eerkens2001, pp.496].
+We take this as our measurement of standardisation in vessel shape variables: lower CV values reflect higher standardisation, and thus increased craft specialisation in the community.
 

### 2020-01-31 analysis/paper/paper.Rmd
paste the deleted para about metric data to scraps.R

@@
 Here, EFA is combined with a significance test for the equality of CVs of shape variables to provide a robust way to identify and statistically assess differences in shape standardization.
-The direct relationship between foreign influences and standardization of ceramic shape was tested on pottery from Kiwulan, and large Iron Age Indienous settlement in northeast Taiwan.
+The direct relationship between foreign influences and standardization of ceramic shape was tested on pottery from Kiwulan, and large Iron Age Indigenous settlement in northeast Taiwan.
 Lower variation in ceramic shape was identified after European presence began, and even lower variation during the period of Chinese presence.

### 2020-01-31 analysis/paper/scraps.Rmd
delete one script file
paste the deleted para about metric data to scraps.R

@@
 
-# Part of Introdution
+# Part of Introduction
 
@@
 To test the prediction of a shift in the level of social inequality in this case study in northeastern Taiwan influenced by the colonial European presence, we use standardization of pottery in shapes as an index for craft specialization to identify the presence of a few individuals controlling large scale pottery production [@Costin1995; @Hirshman2010] using outline-based geometric morphometrics couples with traditional metric measurements to compare and evaluate the use of outline approaches for pottery with high consistency in shapes.
+
+#--------------original metric data analysis------------
+Table \@ref(tab:metricCVstest) shows the results of Modified signed-likelihood ratio test (MSLRT) for equality of CVs for for six measurements and two ratio values.
+There is a significant difference between CV values for two metric measurements across phases, the rim diameter and the neck diameter.
+The CV values show that the variation of rim diameter is small (6.8%)  before European contact, increases after the European presence (9.6%), and decreases after the Chinese presence (6.9%).
+For the neck diameter, the CV values indicate smaller variation after the European period (7.7%) and the Chinese period (7.1%) compared to pre-European period (10.6%).
+<!-- these numeric values in this paragraph should come from inline code, not copy-pasted in like this -->
+<!-- You need to put the code that computes the MSLRT results right in here, in the Rmd -->
+
+```{r metricCVstest}
+metric_cvs_with_phase <- readr::read_csv(here::here("analysis", "data", "raw_data", "Kwl_p_metrics_cv_test.csv"))
+metric_cvs_with_phase_tidy <-
+  metric_cvs_with_phase %>% 
+  mutate_at(c("pre-e", "post-e", "ch-con", "MSLRT", "p_value"), round, 4) %>% 
+  rename(`pre-European` = "pre-e", `after European` = "post-e", Chinese = "ch-con")
+knitr::kable(metric_cvs_with_phase_tidy,
+             caption = "Coefficient of Variation test for metric attributes by phases")
+```
+
+#------------revised metric data analysis---------------
+
+Table \@ref(tab:metricCVstest-pre-post) and \@ref(tab:metricCVstest-post-chi) shows the results of Asymptotic test [@Feltz1996] for equality of CVs for for six measurements and two ratio values.
+There is a significant difference (p-value less than 0.01) between CV values for two metric measurements across phases, the rim diameter and the neck diameter before and after the European presnece.
+The CV values show that the variation of rim diameter is small (6.8%) before European contact, increases after the European presence (9.6%), and decreases after the Chinese presence (6.9%).
+For the neck diameter, the CV values indicate smaller variation after the European period (7.7%) and the Chinese period (7.1%) compared to pre-European period (10.6%).
+After the arrival of the Chinese, there is only one metric attribute, the rim diameter, showing significant difference from post-European period (p-value less than 0.05).
+
+```{r metricCVstest-pre-post, eval=FALSE}
+# read metric data
+kwl_p_ave <- readr::read_csv(here::here("analysis", "data", "raw_data",
+                                 "KWL-pottery-metrics-tidy-ave.csv"))
+library(cvequality)
+# select metric attributes and tidy data for computing CV
+kwl_p_metrics_long <-
+  kwl_p_ave %>%
+  mutate(dia_rim_body = Dia_Rim_ave/Dia_Body_ave,
+         thi_rim_body = Thick_Rim_ave/Thick_Body_ave) %>%
+  select(period,
+         Dia_Rim_ave,
+         Dia_Neck_ave,
+         Dia_Body_ave,
+         Thick_Rim_ave,
+         Thick_Neck_ave,
+         Thick_Body_ave,
+         dia_rim_body,
+         thi_rim_body,
+         # H_Rim
+  ) %>%
+  rename(`Rim diameter (mm)` = Dia_Rim_ave,
+         `Neck diameter (mm)` = Dia_Neck_ave,
+         `Body diameter (mm)` = Dia_Body_ave,
+         `Rim thickness (mm)` = Thick_Rim_ave,
+         `Neck thickness (mm)` = Thick_Neck_ave,
+         `Body thickness (mm)` = Thick_Body_ave,
+         `Ratio of Rim/Body diameter` = dia_rim_body,
+         `Ratio of Rim/Body thickness` = thi_rim_body
+  ) %>%
+  gather(variable, value, -period) %>%
+  filter(!is.na(period),
+         !is.na(value))
+# filter pre-e and post-e periods
+kwl_p_metrics_long_pre_post <-
+  kwl_p_metrics_long %>%
+  filter(!period == "ch-con")
+# filter post-e and chi periods
+kwl_p_metrics_long_post_chi <-
+  kwl_p_metrics_long %>%
+  filter(!period == "pre-e")
+#------------------------------------------CV test for pre-e and post-e pair
+kiwulan_pottery_metrics_cvs_plot_cv_test_asym_pre_post <-
+  kwl_p_metrics_long_pre_post %>%
+  nest(-variable) %>%  # or nest(period, value) %>%
+  mutate(asymptotic_test = map(data, # change from mslr_test to asymptotic_test
+                               ~bind_cols(asymptotic_test (
+                                 .x$value,
+                                 .x$period)))) %>%
+  unnest(asymptotic_test)
+# compute CV
+kwl_p_cvs_pre_post <-
+  kwl_p_metrics_long_pre_post %>%
+  group_by(period, variable) %>%
+  summarise(cvs = raster::cv(value, na.rm = TRUE))
+# join p-values to CVs
+kwl_p_cvs_and_pvalues_pre_post <-
+  kwl_p_cvs_pre_post %>%
+  left_join(kiwulan_pottery_metrics_cvs_plot_cv_test_asym_pre_post) %>%
+  mutate(significant = ifelse(p_value <= 0.05, 'yes', 'no')) %>%
+  mutate(variable = factor(variable, ordered = TRUE)) %>%
+  arrange(variable)
+# making table for CV
+kwl_p_cvs_table_pre_post <-
+  kwl_p_cvs_and_pvalues_pre_post %>%
+  select(-data, -significant) %>%
+  pivot_wider(names_from = period, values_from = cvs) %>%
+  mutate(variable = factor(variable, levels = c("Rim thickness (mm)",
+                                                "Neck thickness (mm)",
+                                                "Body thickness (mm)",
+                                                "Ratio of Rim/Body thickness",
+                                                "Rim diameter (mm)",
+                                                "Neck diameter (mm)",
+                                                "Body diameter (mm)",
+                                                "Ratio of Rim/Body diameter"))) %>%
+  arrange(variable) %>%
+  select(variable, `pre-e`, `post-e`, `D_AD`, `p_value`) %>%
+  mutate_at(c("pre-e", "post-e", "D_AD", "p_value"), round, 4)
+knitr::kable(kwl_p_cvs_table_pre_post,
+             caption = "Coefficient of Variation test for metric attributes before and after European presence")
+```
+knitr::kable(metric_cvs_with_phase_tidy,
+             caption = "Coefficient of Variation test for metric attributes by phases")
+```{r metricCVstest-post-chi, eval=FALSE}
+#-----------------------------------CV test for post-e and ch-con pair
+kiwulan_pottery_metrics_cvs_plot_cv_test_asym_post_chi <-
+  kwl_p_metrics_long_post_chi %>%
+  nest(-variable) %>%  # or nest(period, value) %>%
+  mutate(asymptotic_test = map(data, # change from mslr_test to asymptotic_test
+                               ~bind_cols(asymptotic_test (
+                                 .x$value,
+                                 .x$period)))) %>%
+  unnest(asymptotic_test)
+# compute CV
+kwl_p_cvs_pre_post <-
+  kwl_p_metrics_long_post_chi %>%
+  group_by(period, variable) %>%
+  summarise(cvs = raster::cv(value, na.rm = TRUE))
+# join p-values to CVs
+kwl_p_cvs_and_pvalues_post_chi <-
+  kwl_p_cvs_pre_post %>%
+  left_join(kiwulan_pottery_metrics_cvs_plot_cv_test_asym_post_chi) %>%
+  mutate(significant = ifelse(p_value <= 0.05, 'yes', 'no')) %>%
+  mutate(variable = factor(variable, ordered = TRUE)) %>%
+  arrange(variable)
+# making table for CV
+kwl_p_cvs_table_post_chi <-
+  kwl_p_cvs_and_pvalues_post_chi %>%
+  select(-data, -significant) %>%
+  pivot_wider(names_from = period, values_from = cvs) %>%
+  mutate(variable = factor(variable, levels = c("Rim thickness (mm)",
+                                                "Neck thickness (mm)",
+                                                "Body thickness (mm)",
+                                                "Ratio of Rim/Body thickness",
+                                                "Rim diameter (mm)",
+                                                "Neck diameter (mm)",
+                                                "Body diameter (mm)",
+                                                "Ratio of Rim/Body diameter"))) %>%
+  arrange(variable) %>%
+  select(variable, `post-e`, `ch-con`, `D_AD`, `p_value`) %>%
+  mutate_at(c("post-e", "ch-con", "D_AD", "p_value"), round, 4)
+knitr::kable(kwl_p_cvs_table_post_chi,
+             caption = "Coefficient of Variation test for metric attributes between European and Chinese presence")

### 2020-01-31 analysis/paper/paper.Rmd
add text of metric data to discussion section

@@
 
-<!-- a few lines to compare metric to shape... main point to make clear is the substantive result about the main research questions -->
+When we examine metric data for six measurements and two ratio values using CV tests (\table here), it shows there is a significant difference (p-value less than 0.01) between CV values for two metric measurements across phases, the rim diameter and the neck diameter before and after the European presence.
+After the arrival of the Chinese, there is only one metric attribute, the rim diameter, showing significant difference compared with post-European period (p-value less than 0.05).
+Although these results shows the differences between different contact phases that corresponds to part of the results of shape variable, elliptic Fourier analysis captures more subtle differences in a substantially way reflected by overall shapes of pottery, not only limited to linear aspects.
 

### 2020-01-31 analysis/paper/scraps.Rmd
add text of metric data to discussion section

--- 2020-01-04 analysis/paper/scraps.Rmd
@@
 To test the prediction of a shift in the level of social inequality in this case study in northeastern Taiwan influenced by the colonial European presence, we use standardization of pottery in shapes as an index for craft specialization to identify the presence of a few individuals controlling large scale pottery production [@Costin1995; @Hirshman2010] using outline-based geometric morphometrics couples with traditional metric measurements to compare and evaluate the use of outline approaches for pottery with high consistency in shapes.
+
+#--------------original metric data analysis------------
+Table \@ref(tab:metricCVstest) shows the results of Modified signed-likelihood ratio test (MSLRT) for equality of CVs for for six measurements and two ratio values.
+There is a significant difference between CV values for two metric measurements across phases, the rim diameter and the neck diameter.
+The CV values show that the variation of rim diameter is small (6.8%)  before European contact, increases after the European presence (9.6%), and decreases after the Chinese presence (6.9%).
+For the neck diameter, the CV values indicate smaller variation after the European period (7.7%) and the Chinese period (7.1%) compared to pre-European period (10.6%).
+<!-- these numeric values in this paragraph should come from inline code, not copy-pasted in like this -->
+<!-- You need to put the code that computes the MSLRT results right in here, in the Rmd -->
+
+```{r metricCVstest}
+metric_cvs_with_phase <- readr::read_csv(here::here("analysis", "data", "raw_data", "Kwl_p_metrics_cv_test.csv"))
+metric_cvs_with_phase_tidy <-
+  metric_cvs_with_phase %>% 
+  mutate_at(c("pre-e", "post-e", "ch-con", "MSLRT", "p_value"), round, 4) %>% 
+  rename(`pre-European` = "pre-e", `after European` = "post-e", Chinese = "ch-con")
+knitr::kable(metric_cvs_with_phase_tidy,
+             caption = "Coefficient of Variation test for metric attributes by phases")
+```
+
+#------------revised metric data analysis---------------
+
+Table \@ref(tab:metricCVstest-pre-post) and \@ref(tab:metricCVstest-post-chi) shows the results of Asymptotic test [@Feltz1996] for equality of CVs for for six measurements and two ratio values.
+There is a significant difference (p-value less than 0.01) between CV values for two metric measurements across phases, the rim diameter and the neck diameter before and after the European presence.
+The CV values show that the variation of rim diameter is small (6.8%) before European contact, increases after the European presence (9.6%), and decreases after the Chinese presence (6.9%).
+For the neck diameter, the CV values indicate smaller variation after the European period (7.7%) and the Chinese period (7.1%) compared to pre-European period (10.6%).
+After the arrival of the Chinese, there is only one metric attribute, the rim diameter, showing significant difference from post-European period (p-value less than 0.05).
+
+```{r metricCVstest-pre-post, eval=FALSE}
+# read metric data
+kwl_p_ave <- readr::read_csv(here::here("analysis", "data", "raw_data",
+                                 "KWL-pottery-metrics-tidy-ave.csv"))
+library(cvequality)
+library(tidyverse)
+# select metric attributes and tidy data for computing CV
+kwl_p_metrics_long <-
+  kwl_p_ave %>%
+  mutate(dia_rim_body = Dia_Rim_ave/Dia_Body_ave,
+         thi_rim_body = Thick_Rim_ave/Thick_Body_ave) %>%
+  select(period,
+         Dia_Rim_ave,
+         Dia_Neck_ave,
+         Dia_Body_ave,
+         Thick_Rim_ave,
+         Thick_Neck_ave,
+         Thick_Body_ave,
+         dia_rim_body,
+         thi_rim_body,
+         # H_Rim
+  ) %>%
+  rename(`Rim diameter (mm)` = Dia_Rim_ave,
+         `Neck diameter (mm)` = Dia_Neck_ave,
+         `Body diameter (mm)` = Dia_Body_ave,
+         `Rim thickness (mm)` = Thick_Rim_ave,
+         `Neck thickness (mm)` = Thick_Neck_ave,
+         `Body thickness (mm)` = Thick_Body_ave,
+         `Ratio of Rim/Body diameter` = dia_rim_body,
+         `Ratio of Rim/Body thickness` = thi_rim_body
+  ) %>%
+  gather(variable, value, -period) %>%
+  filter(!is.na(period),
+         !is.na(value))
+# filter pre-e and post-e periods
+kwl_p_metrics_long_pre_post <-
+  kwl_p_metrics_long %>%
+  filter(!period == "ch-con")
+# filter post-e and chi periods
+kwl_p_metrics_long_post_chi <-
+  kwl_p_metrics_long %>%
+  filter(!period == "pre-e")
+#------------------------------------------CV test for pre-e and post-e pair
+kiwulan_pottery_metrics_cvs_plot_cv_test_asym_pre_post <-
+  kwl_p_metrics_long_pre_post %>%
+  nest(-variable) %>%  # or nest(period, value) %>%
+  mutate(asymptotic_test = map(data, # change from mslr_test to asymptotic_test
+                               ~bind_cols(asymptotic_test (
+                                 .x$value,
+                                 .x$period)))) %>%
+  unnest(asymptotic_test)
+# compute CV
+kwl_p_cvs_pre_post <-
+  kwl_p_metrics_long_pre_post %>%
+  group_by(period, variable) %>%
+  summarise(cvs = raster::cv(value, na.rm = TRUE))
+# join p-values to CVs
+kwl_p_cvs_and_pvalues_pre_post <-
+  kwl_p_cvs_pre_post %>%
+  left_join(kiwulan_pottery_metrics_cvs_plot_cv_test_asym_pre_post) %>%
+  mutate(significant = ifelse(p_value <= 0.05, 'yes', 'no')) %>%
+  mutate(variable = factor(variable, ordered = TRUE)) %>%
+  arrange(variable)
+# making table for CV
+kwl_p_cvs_table_pre_post <-
+  kwl_p_cvs_and_pvalues_pre_post %>%
+  select(-data, -significant) %>%
+  pivot_wider(names_from = period, values_from = cvs) %>%
+  mutate(variable = factor(variable, levels = c("Rim thickness (mm)",
+                                                "Neck thickness (mm)",
+                                                "Body thickness (mm)",
+                                                "Ratio of Rim/Body thickness",
+                                                "Rim diameter (mm)",
+                                                "Neck diameter (mm)",
+                                                "Body diameter (mm)",
+                                                "Ratio of Rim/Body diameter"))) %>%
+  arrange(variable) %>%
+  select(variable, `pre-e`, `post-e`, `D_AD`, `p_value`) %>%
+  mutate_at(c("pre-e", "post-e", "D_AD", "p_value"), round, 4)
+knitr::kable(kwl_p_cvs_table_pre_post,
+             caption = "Coefficient of Variation for metric attributes before and after European presence using Asymptotic test [@Feltz1996]")
+```
+
+```{r metricCVstest-post-chi, eval=FALSE}
+#-----------------------------------CV test for post-e and ch-con pair
+kiwulan_pottery_metrics_cvs_plot_cv_test_asym_post_chi <-
+  kwl_p_metrics_long_post_chi %>%
+  nest(-variable) %>%  # or nest(period, value) %>%
+  mutate(asymptotic_test = map(data, # change from mslr_test to asymptotic_test
+                               ~bind_cols(asymptotic_test (
+                                 .x$value,
+                                 .x$period)))) %>%
+  unnest(asymptotic_test)
+# compute CV
+kwl_p_cvs_pre_post <-
+  kwl_p_metrics_long_post_chi %>%
+  group_by(period, variable) %>%
+  summarise(cvs = raster::cv(value, na.rm = TRUE))
+# join p-values to CVs
+kwl_p_cvs_and_pvalues_post_chi <-
+  kwl_p_cvs_pre_post %>%
+  left_join(kiwulan_pottery_metrics_cvs_plot_cv_test_asym_post_chi) %>%
+  mutate(significant = ifelse(p_value <= 0.05, 'yes', 'no')) %>%
+  mutate(variable = factor(variable, ordered = TRUE)) %>%
+  arrange(variable)
+# making table for CV
+kwl_p_cvs_table_post_chi <-
+  kwl_p_cvs_and_pvalues_post_chi %>%
+  select(-data, -significant) %>%
+  pivot_wider(names_from = period, values_from = cvs) %>%
+  mutate(variable = factor(variable, levels = c("Rim thickness (mm)",
+                                                "Neck thickness (mm)",
+                                                "Body thickness (mm)",
+                                                "Ratio of Rim/Body thickness",
+                                                "Rim diameter (mm)",
+                                                "Neck diameter (mm)",
+                                                "Body diameter (mm)",
+                                                "Ratio of Rim/Body diameter"))) %>%
+  arrange(variable) %>%
+  select(variable, `post-e`, `ch-con`, `D_AD`, `p_value`) %>%
+  mutate_at(c("post-e", "ch-con", "D_AD", "p_value"), round, 4)
+knitr::kable(kwl_p_cvs_table_post_chi,
+             caption = "Coefficient of Variation for metric attributes between European and Chinese presence using Asymptotic test [@Feltz1996]")
+```

### 2020-01-31 analysis/paper/scraps.Rmd
combine two metric data tables

@@
              caption = "Coefficient of Variation for metric attributes between European and Chinese presence using Asymptotic test [@Feltz1996]")
+
+# join two table together
+kwl_p_cvs_table_all <-
+  kwl_p_cvs_table_pre_post %>% 
+  left_join(kwl_p_cvs_table_post_chi, "variable") %>% 
+  select(-`post-e.y`) %>% 
+  rename(`before European` = "pre-e", `European` = "post-e.x", Chinese = "ch-con",
+         `D_AD.1` = "D_AD.x", `D_AD.2` = "D_AD.y",
+         `P value.1` = `p_value.x`, `P value.2` = `p_value.y`) %>% 
+  select(variable, `before European`, European, Chinese,
+         `D_AD.1`, `P value.1`, `D_AD.2`, `P value.2`)
 ```

### 2020-01-31 analysis/paper/paper.Rmd
fix typo and use American spelling

@@
 Outline analysis is an important method because, unlike more commonly-used landmark analysis methods, it can effectively quantify shape differences for objects that lack distinctive measurement points needed for landmark analysis.
-We demonstrate this method using pottery from Kiwulan, a large multi-component Iron Age site (AD 1350-1850) in northeast Taiwan.
+We demonstrate this method using pottery from Kiwulan, a large multi-component Iron Age site (AD 1350-1850) in northeastern Taiwan.
 To measure ceramic specialization, we quantified pottery standardization by analyzing shape variables with reproducible geometric morphometric methods.
 We computed coefficients of variation (CVs) for shape coefficients obtained by elliptical Fourier analysis to test for shape standardization.
-We found significant differences in pottery shape and shape standardization that indicate changes in pottery production resulting from contact with mainland Han Chinese groups in northeast Taiwan.
+We found significant differences in pottery shape and shape standardization that indicate changes in pottery production resulting from contact with mainland Han Chinese groups in northeastern Taiwan.
 We infer increasing craft specialization and changes in social organization.
@@
 That said, a more economic focus considers production, exchange, and consumption as a whole to explain specialization as part of a strategy to accommodate uneven distributions of resources [@Costin2001; @Arnold1994].
-Craft specialisation may appear as a reaction to changes in subsistence practices [@Stark1995economic], or as a result of long-distance trade and inter-regional exchange [@Alizadeh2018].
+Craft specialization may appear as a reaction to changes in subsistence practices [@Stark1995economic], or as a result of long-distance trade and inter-regional exchange [@Alizadeh2018].
 Present-day investigation of ceramic specialization in the Jodhpur region of India suggests that the adoption of standardized products across cultural boundaries is a reaction to a collapse of previous economic system, indicating a correlation with major socio-economic changes [@Roux2015].
@@
 However, because pottery vessels are typically have curved shapes, linear measurements have limited sensitivity to many kinds of shape variations.
-Thus, to capture subtle shape variations that might be relevant to standardization, a method that is senstive to non-linear aspects of shape variation is necessary.
-Here we analyse pottery shape using geometric morphometric methods, a new approach for measuring the degree of standardization because it takes a statistical summary of the overall shape as variable for comparison [@Slice2007].
+Thus, to capture subtle shape variations that might be relevant to standardization, a method that is sensitive to non-linear aspects of shape variation is necessary.
+Here we analyze pottery shape using geometric morphometric methods, a new approach for measuring the degree of standardization because it takes a statistical summary of the overall shape as variable for comparison [@Slice2007].
 Using geometric morphometry, we study shape and standardization of locally made ceramics at Kiwulan (1350-1950 AD) [@Chen2007], a large multi-component archaeological site in northeastern Taiwan, to identify changes in ceramic production resulting from cross-cultural interaction contexts.
@@
 Our hypothesis is that if foreign influences had impacts on the emergence of social inequality in local Indigenous society due to monopolies of trade arising among a small number of Indigenous people, then we expect to see changes in social organization from a more corporate to a more network organization [@Feinman2000].
-We expect this corporate-network shift to manifest as increased specialization pottery manufacture, indicated archaeologically by icreased standardisation of pottery production at Kiwulan.
+We expect this corporate-network shift to manifest as increased specialization pottery manufacture, indicated archaeologically by increased standardization of pottery production at Kiwulan.
 The emergence of craft specialization, here pottery specialization, is often related to changes in social organization towards a society with increasing status inequality.
@@
 Where only half cross-section images were available, these were duplicated, flipped, and then joined with another one to create a 2D closed outline for each vessel.
-Geometric morphometric analyses were conducted using the R software [@Rlanguage2019] using the functions included in the Momocs, a R package intended to quantify and analyse shapes, especially for outline analysis [@Bonhomme2014].
+Geometric morphometric analyses were conducted using the R software [@Rlanguage2019] using the functions included in the Momocs, a R package intended to quantify and analyze shapes, especially for outline analysis [@Bonhomme2014].
 The digitised outlines were converted into a list of successive x-y pixel coordinates for elliptic Fourier analysis (EFA), to investigate morphological differences among pottery shapes from our three occupation phases.
-The harmonic coefficients generated by EFA were analysed by principal component analysis (PCA) for dimensionality reduction to illustrate the diversity of the shape data and identify major patterns of variation through .
+The harmonic coefficients generated by EFA were analyzed by principal component analysis (PCA) for dimensionality reduction to illustrate the diversity of the shape data and identify major patterns of variation through .
 
@@
 
-The principal component (PC) scores were analysed with a multivariate analysis of variance (MANOVA) test to identify significant differences in shapes between occupation phases.
+The principal component (PC) scores were analyzed with a multivariate analysis of variance (MANOVA) test to identify significant differences in shapes between occupation phases.
 We also computed coefficients of variation values (CVs) for the PCs, treating the PCs as shape variables that are more informative than linear dimensions.
@@
 CV values are robust to different sample sizes in the condition that the mean-standard deviation relationship is linear and positive since standard deviation is scaled to the mean [@Eerkens2001, pp.496].
-We take this as our measurement of standardisation in vessel shape variables: lower CV values reflect higher standardisation, and thus increased craft specialisation in the community.
+We take this as our measurement of standardization in vessel shape variables: lower CV values reflect higher standardization, and thus increased craft specialization in the community.
 
@@
 To investigate the effect of the historically-documented decline of Indigenous population in the Chinese period [@Chen2007; @Hsieh2009] on pottery production we examined the spatial distribution of pots in our sample in each period.
-Our concern is that the smaller population size during the Chinese period could confound our investigations into craft specialisation because of reduced numbers of potters and pottery production shrinking a few locations in the settlement during this time.
+Our concern is that the smaller population size during the Chinese period could confound our investigations into craft specialization because of reduced numbers of potters and pottery production shrinking a few locations in the settlement during this time.
 Although the overall number of vessels is smaller during the Chinese contact period, Figure \@ref(fig:pottery-spatial-pattern) shows that pottery is distributed randomly across the sampling area without any distinctive clusters.
@@
    fgProcrustes %>% 
-  # Normalisation of the starting points
+  # Normalization of the starting points
    coo_slidedirection
@@
 # what are the variences for each PC in each period?
-# CVs on shape PCs for standardisation
+# CVs on shape PCs for standardization
 len <- length(op.p$x[,1])
@@
 
-Our CV tests indicates that there are significant differences in shape standardisation between the pre-European period and post-European period, and between the Chinese contact and either pre-European or post-European periods.
+Our CV tests indicates that there are significant differences in shape standardization between the pre-European period and post-European period, and between the Chinese contact and either pre-European or post-European periods.
 This suggests pottery shape became more homogeneous and standardized after foreign contacts with European colonizers and even more so after contact with Chinese immigrants.
@@
 Our results offer tentative support for the hypothesis that foreign influences at Kiwulan influenced emergence of social inequality in the local Indigenous society.
-If increased shape standardisation is a reliable indicator of craft specialisation, then we may be seeing evidence of a shift from corporate to network organization [@Feinman2000].
+If increased shape standardization is a reliable indicator of craft specialization, then we may be seeing evidence of a shift from corporate to network organization [@Feinman2000].
 However, strong claims for an emergence of social complexity resulting from foreign contact at Kiwulan will need support from multiple and diverse sources of evidence.
@@
 Shamaoshan cemetery (3BC- 4AD) in Southwest China shows that the process of the incorporation of Southwest China into the Han Empire involved a century of conflicts, resistance, and acceptance among social groups with different identities, especially in the historical context of Han immigrants [@Wu2019].
-A similar dynamic may have occurred at Kiwulan, with vessel shape indicating both acceptance of foreign influence through increased shape standardisation, and resistance through the overall continuity in vessel shape.
+A similar dynamic may have occurred at Kiwulan, with vessel shape indicating both acceptance of foreign influence through increased shape standardization, and resistance through the overall continuity in vessel shape.
 Vessel shape can be viewed as a symbol as an expression the Indigenous identity and social boundaries because shape is a highly visible trait compared with other features of pottery [cf.
@@
 Here, EFA is combined with a significance test for the equality of CVs of shape variables to provide a robust way to identify and statistically assess differences in shape standardization.
-The direct relationship between foreign influences and standardization of ceramic shape was tested on pottery from Kiwulan, and large Iron Age Indigenous settlement in northeast Taiwan.
+The direct relationship between foreign influences and standardization of ceramic shape was tested on pottery from Kiwulan, and large Iron Age Indigenous settlement in northeastern Taiwan.
 Lower variation in ceramic shape was identified after European presence began, and even lower variation during the period of Chinese presence.

### 2020-02-01 analysis/paper/paper.Rmd
add metric CV test to Rmd file

@@
 
-When we examine metric data for six measurements and two ratio values using CV tests (\table here), it shows there is a significant difference (p-value less than 0.01) between CV values for two metric measurements across phases, the rim diameter and the neck diameter before and after the European presence.
+We compare the result of shape variable with the result of metric data for six measurements and two ratio values using CV test (Table \@ref(tab:metricCVstest)).
+Metric measurements show there is a significant difference (p-value less than 0.01) between CV values for two metric measurements across phases, the rim diameter and the neck diameter before and after the European presence.
 After the arrival of the Chinese, there is only one metric attribute, the rim diameter, showing significant difference compared with post-European period (p-value less than 0.05).
-Although these results shows the differences between different contact phases that corresponds to part of the results of shape variable, elliptic Fourier analysis captures more subtle differences in a substantially way reflected by overall shapes of pottery, not only limited to linear aspects.
+Although these results shows the differences between different contact phases that corresponds to part of the results of shape variable, elliptic Fourier analysis captures more subtle differences in a substantially way reflected by overall shapes of pottery, not only limited to diameter variables.
+
+```{r metricCVstest}
+# read metric data
+kwl_p_ave <- readr::read_csv(here::here("analysis", "data", "raw_data",
+                                        "KWL-pottery-metrics-tidy-ave.csv"))
+
+# select metric attributes and tidy data for computing CV
+kwl_p_metrics_long <-
+  kwl_p_ave %>%
+  mutate(dia_rim_body = Dia_Rim_ave/Dia_Body_ave,
+         thi_rim_body = Thick_Rim_ave/Thick_Body_ave) %>%
+  select(period,
+         Dia_Rim_ave,
+         Dia_Neck_ave,
+         Dia_Body_ave,
+         Thick_Rim_ave,
+         Thick_Neck_ave,
+         Thick_Body_ave,
+         dia_rim_body,
+         thi_rim_body,) %>%
+  rename(`Rim diameter (mm)` = Dia_Rim_ave,
+         `Neck diameter (mm)` = Dia_Neck_ave,
+         `Body diameter (mm)` = Dia_Body_ave,
+         `Rim thickness (mm)` = Thick_Rim_ave,
+         `Neck thickness (mm)` = Thick_Neck_ave,
+         `Body thickness (mm)` = Thick_Body_ave,
+         `Ratio of Rim/Body diameter` = dia_rim_body,
+         `Ratio of Rim/Body thickness` = thi_rim_body) %>%
+  gather(variable, value, -period) %>%
+  filter(!is.na(period), !is.na(value)) %>% 
+  mutate(variable = factor(variable, levels = c("Rim thickness (mm)",
+                                                "Neck thickness (mm)",
+                                                "Body thickness (mm)",
+                                                "Ratio of Rim/Body thickness",
+                                                "Rim diameter (mm)",
+                                                "Neck diameter (mm)",
+                                                "Body diameter (mm)",
+                                                "Ratio of Rim/Body diameter"))) %>%
+  arrange(variable)
+# filter pre-e and post-e periods
+kwl_p_metrics_long_pre_post <-
+  kwl_p_metrics_long %>%
+  filter(!period == "ch-con")
+# filter post-e and chi periods
+kwl_p_metrics_long_post_chi <-
+  kwl_p_metrics_long %>%
+  filter(!period == "pre-e")
+
+#------------------------------------------CV test for pre-e and post-e pair
+kiwulan_pottery_metrics_cvs_plot_cv_test_asym_pre_post <-
+  kwl_p_metrics_long_pre_post %>%
+  nest(-variable) %>%  # or nest(period, value) %>%
+  mutate(asymptotic_test = map(data, # change from mslr_test to asymptotic_test
+                               ~bind_cols(asymptotic_test (
+                                 .x$value,
+                                 .x$period)))) %>%
+  unnest(asymptotic_test)
+# compute CV
+kwl_p_cvs_pre_post <-
+  kwl_p_metrics_long_pre_post %>%
+  group_by(period, variable) %>%
+  summarise(cvs = raster::cv(value, na.rm = TRUE))
+# join p-values to CVs
+kwl_p_cvs_and_pvalues_pre_post <-
+  kwl_p_cvs_pre_post %>%
+  left_join(kiwulan_pottery_metrics_cvs_plot_cv_test_asym_pre_post) %>%
+  mutate(significant = ifelse(p_value <= 0.05, 'yes', 'no')) %>%
+  mutate(variable = factor(variable, ordered = TRUE)) %>%
+  arrange(variable)
+# making table for CV
+kwl_p_cvs_table_pre_post <-
+  kwl_p_cvs_and_pvalues_pre_post %>%
+  select(-data, -significant) %>%
+  pivot_wider(names_from = period, values_from = cvs) %>%
+  select(variable, `pre-e`, `post-e`, `D_AD`, `p_value`) %>%
+  mutate_at(c("pre-e", "post-e", "D_AD", "p_value"), round, 4)
+
+#-----------------------------------CV test for post-e and ch-con pair
+kiwulan_pottery_metrics_cvs_plot_cv_test_asym_post_chi <-
+  kwl_p_metrics_long_post_chi %>%
+  nest(-variable) %>%  # or nest(period, value) %>%
+  mutate(asymptotic_test = map(data, # change from mslr_test to asymptotic_test
+                               ~bind_cols(asymptotic_test (
+                                 .x$value,
+                                 .x$period)))) %>%
+  unnest(asymptotic_test)
+# compute CV
+kwl_p_cvs_pre_post <-
+  kwl_p_metrics_long_post_chi %>%
+  group_by(period, variable) %>%
+  summarise(cvs = raster::cv(value, na.rm = TRUE))
+# join p-values to CVs
+kwl_p_cvs_and_pvalues_post_chi <-
+  kwl_p_cvs_pre_post %>%
+  left_join(kiwulan_pottery_metrics_cvs_plot_cv_test_asym_post_chi) %>%
+  mutate(significant = ifelse(p_value <= 0.05, 'yes', 'no')) %>%
+  mutate(variable = factor(variable, ordered = TRUE)) %>%
+  arrange(variable)
+# making table for CV
+kwl_p_cvs_table_post_chi <-
+  kwl_p_cvs_and_pvalues_post_chi %>%
+  select(-data, -significant) %>%
+  pivot_wider(names_from = period, values_from = cvs) %>%
+  select(variable, `post-e`, `ch-con`, `D_AD`, `p_value`) %>%
+  mutate_at(c("post-e", "ch-con", "D_AD", "p_value"), round, 4)
+
+# join two table together
+kwl_p_cvs_table_all <-
+  kwl_p_cvs_table_pre_post %>% 
+  left_join(kwl_p_cvs_table_post_chi, "variable") %>% 
+  select(-`post-e.y`) %>% 
+  rename(`before European` = "pre-e", `European` = "post-e.x", Chinese = "ch-con",
+         `D_AD-1` = "D_AD.x", `D_AD-2` = "D_AD.y",
+         `P value-1` = `p_value.x`, `P value-2` = `p_value.y`) %>% 
+  select(variable, `before European`, European, Chinese,
+         `D_AD-1`, `P value-1`, `D_AD-2`, `P value-2`)
+
+knitr::kable(kwl_p_cvs_table_all,
+             caption = "Coefficient of Variation for metric attributes by three phases using Asymptotic test [@Feltz1996].
+The first set of P-value represents the comparison between before European and European periods, while the second represents the comparison between European and Chinese periods ")
+```
 

### 2020-02-03 analysis/paper/scraps.Rmd
fix typo and use American spelling
create a script file and use point pattern analysis for pottery

@@
 
-#--------------original metric data analysis------------
+#--------------original metric data analysis that looks at phases all together------------
 Table \@ref(tab:metricCVstest) shows the results of Modified signed-likelihood ratio test (MSLRT) for equality of CVs for for six measurements and two ratio values.

### 2020-02-03 analysis/paper/spatial_pattern.R

@@ -1,0 +1,120 @@
+library(sf)
+library(viridis)
+library(tidyverse)
+
+# read in pottery data
+pottery_data <- readr::read_csv(here::here("analysis/data/raw_data/KWL_Ceramic_final2018_with_phases.csv"))
+
+# read in the spatial data
+AD_zone <- invisible(st_read(here::here("analysis", "data", "raw_data", "AD_zone.shp"), quiet = TRUE))
+units_40 <- invisible(read.csv(here::here("analysis", "data", "raw_data", "kwl-list-of-sampling-squares.csv")))
+
+AD_zone_tidy <-
+  AD_zone %>%
+  mutate(Pit =paste0("P0",Id)) %>%
+  inner_join(units_40, by = c("Pit" = "the_sq")) %>%
+  select(Pit, geometry)
+
+# tidy pottery data
+pottery_count <-
+  pottery_data %>%
+  select(Site, Context, Pit, Layer, No., phase) %>%
+  count(Pit, phase)
+
+# join two dataset to get coordinate
+pottery_count_join <-
+  pottery_count %>%
+  mutate(phase = as.factor(phase)) %>% # seems need this line first
+  inner_join(AD_zone_tidy) %>%
+  filter(!is.na(phase)) %>%
+  mutate(phase = case_when(
+    phase %in% "pre-e"~ "Before European Contact",
+    phase %in% "post-e" ~ "European Presence",
+    phase %in% "ch-con" ~ "Chinese Presence")) %>%
+  mutate(phase = factor(phase,
+                        level = c("Before European Contact",
+                                  "European Presence",
+                                  "Chinese Presence")))
+# plot spatial pattern for pottery
+plot_spatial_distribution_pottery_by_phase <-
+  ggplot() +
+  geom_sf(data = AD_zone_tidy, fill = NA) +
+  geom_sf(data = pottery_count_join,
+          aes(geometry = geometry,
+              fill =n),
+          alpha = 0.9)+
+  facet_wrap(~ phase) +
+  labs(fill = "frequency") +
+  scale_fill_viridis(direction = -1) +
+  theme_minimal() +
+  theme(plot.title = element_text(hjust = 0.5),
+        axis.text.x = element_blank(),
+        axis.text.y = element_blank(),
+        strip.text.x = element_text(size = 6),
+        legend.text = element_text(size = 6),
+        legend.title = element_text(size = 6),
+        legend.key.size =  unit(3, "mm"))
+
[…]

### 2020-02-03 analysis/paper/spatial_pattern.R
add hypothesis testing for pottery spatial pattern

@@ -86,15 +86,15 @@ density_analysis <- function(x, ...) {
   # create ppp and sp(spacial points) format
   lon_tmp <- x$lon
   lat_tmp <- x$lat
-  ornaments_point_ppp_tmp <- ppp(lon_tmp, lat_tmp, xrange, yrange)
-  ornaments_point_sp_tmp <- as(ornaments_point_ppp_tmp, "SpatialPoints")
+  pottery_point_ppp_tmp <- ppp(lon_tmp, lat_tmp, xrange, yrange)
+  pottery_point_sp_tmp <- as(pottery_point_ppp_tmp, "SpatialPoints")
   # kernel density estimation with contour
-  kernel_tmp <- density(ornaments_point_ppp_tmp)
+  kernel_tmp <- density(pottery_point_ppp_tmp)
   kernel_to_plot <-
     plot(kernel_tmp, main = "")
   contour(kernel_tmp, add = TRUE)
   title(..., line = -0.3, adj = 0.4, cex.main = 0.8)
-  return(list(ornaments_point_ppp = ornaments_point_ppp_tmp,
+  return(list(pottery_point_ppp = pottery_point_ppp_tmp,
               kernel_to_plot = kernel_to_plot))
 }
 
@@ -118,3 +118,50 @@ invisible(dev.off())
 knitr::include_graphics(here::here("analysis",
                              "figures",
                              "plot-kde-maps.jpg"))
+
+#--------------------------hypothesis testing for spatial distribution--------------
+
+# function for hypothesis testing
+nndist_hypo_testing <- function(x, n = 1e3L,  ... ) {
+
+  ann_tmp <- mean(nndist(x$pottery_point_ppp, k = 1))
+  # object for storing simulated ANN values
+  ann_r_tmp <- vector(length = n)
+  for (i in 1:n) {
+    # Generate random point locations
+    rand_tmp <-  rpoint(n = x$pottery_point_ppp$n,
+                        win = as.owin(as_Spatial(AD_zone_tidy)))
+    ann_r_tmp[i] <-
+      mean(nndist(rand_tmp, k = 1))  # Tally the ANN values
+  }
+  the_plot <-
+    hist(
+      ann_r_tmp,
+      xlab = NULL,
+      las = 1,
+      breaks = 40,
+      col = "bisque",
+      xlim = range(ann_tmp, ann_r_tmp),
+      ...)
+  abline(v = ann_tmp, col = "blue")
+  return(the_plot)
[…]

### 2020-02-05 analysis/paper/paper.Rmd
replace numbers with inline code for method section

--- 2020-01-31 analysis/paper/paper.Rmd
@@
   pottery_data %>%  
-  select(filename, phase) %>% 
+  dplyr::select(filename, phase) %>% 
   filter(filename %in% isect) %>% 
@@
 total_no <- sum(phase_fre$n)
+pre_no <- phase_fre$n[1]
+Euro_no <- phase_fre$n[2]
+Chi_no <- phase_fre$n[3]
+```
+
+```{r metricCVstest}
+library(cvequality)
+# read metric data
+kwl_p_ave <- readr::read_csv(here::here("analysis", "data", "raw_data",
+                                        "KWL-pottery-metrics-tidy-ave.csv"))
+# number for each phase 
+kwl_p_ave_total <-
+  kwl_p_ave %>% 
+  count(period)
+# inline code
+metric_all <- sum(kwl_p_ave_total$n)
+pre <- kwl_p_ave_total$n[3]
+euro <- kwl_p_ave_total$n[2]
+chi <- kwl_p_ave_total$n[1]
+
+# select metric attributes and tidy data for computing CV
+kwl_p_metrics_long <-
+  kwl_p_ave %>%
+  mutate(dia_rim_body = Dia_Rim_ave/Dia_Body_ave,
+         thi_rim_body = Thick_Rim_ave/Thick_Body_ave) %>%
+  select(period, 
+         Dia_Rim_ave, Dia_Neck_ave, 
+         Dia_Body_ave, Thick_Rim_ave, 
+         Thick_Neck_ave, Thick_Body_ave,
+         dia_rim_body, thi_rim_body,) %>%
+  rename(`Rim diameter (mm)` = Dia_Rim_ave,
+         `Neck diameter (mm)` = Dia_Neck_ave,
+         `Body diameter (mm)` = Dia_Body_ave,
+         `Rim thickness (mm)` = Thick_Rim_ave,
+         `Neck thickness (mm)` = Thick_Neck_ave,
+         `Body thickness (mm)` = Thick_Body_ave,
+         `Ratio of Rim/Body diameter` = dia_rim_body,
+         `Ratio of Rim/Body thickness` = thi_rim_body) %>%
+  gather(variable, value, -period) %>%
+  filter(!is.na(period), !is.na(value)) %>% 
+  mutate(variable = factor(variable, levels = c("Rim thickness (mm)",
+                                                "Neck thickness (mm)",
+                                                "Body thickness (mm)",
+                                                "Ratio of Rim/Body thickness",
+                                                "Rim diameter (mm)",
+                                                "Neck diameter (mm)",
+                                                "Body diameter (mm)",
+                                                "Ratio of Rim/Body diameter"))) %>%
+  arrange(variable)
+
+# filter pre-e and post-e periods
+kwl_p_metrics_long_pre_post <-
+  kwl_p_metrics_long %>%
+  filter(!period == "ch-con")
+# filter post-e and chi periods
+kwl_p_metrics_long_post_chi <-
+  kwl_p_metrics_long %>%
+  filter(!period == "pre-e")
+
+#---------------------------CV test for pre-e and post-e pair
+kiwulan_pottery_metrics_cvs_plot_cv_test_asym_pre_post <-
+  kwl_p_metrics_long_pre_post %>%
+  nest(-variable) %>%  # or nest(period, value) %>%
+  mutate(asymptotic_test = map(data, # change from mslr_test to asymptotic_test
+                               ~bind_cols(asymptotic_test (
+                                 .x$value,
+                                 .x$period)))) %>%
+  unnest(asymptotic_test)
+# compute CV
+kwl_p_cvs_pre_post <-
+  kwl_p_metrics_long_pre_post %>%
+  group_by(period, variable) %>%
+  summarise(cvs = raster::cv(value, na.rm = TRUE))
+# join p-values to CVs
+kwl_p_cvs_and_pvalues_pre_post <-
+  kwl_p_cvs_pre_post %>%
+  left_join(kiwulan_pottery_metrics_cvs_plot_cv_test_asym_pre_post) %>%
+  mutate(significant = ifelse(p_value <= 0.05, 'yes', 'no')) %>%
+  mutate(variable = factor(variable, ordered = TRUE)) %>%
+  arrange(variable)
+# making table for CV
+kwl_p_cvs_table_pre_post <-
+  kwl_p_cvs_and_pvalues_pre_post %>%
+  select(-data, -significant) %>%
+  pivot_wider(names_from = period, values_from = cvs) %>%
+  select(variable, `pre-e`, `post-e`, `D_AD`, `p_value`) %>%
+  mutate_at(c("pre-e", "post-e", "D_AD", "p_value"), round, 4)
+
+#---------------------------CV test for post-e and ch-con pair
+kiwulan_pottery_metrics_cvs_plot_cv_test_asym_post_chi <-
+  kwl_p_metrics_long_post_chi %>%
+  nest(-variable) %>%  # or nest(period, value) %>%
+  mutate(asymptotic_test = map(data, # change from mslr_test to asymptotic_test
+                               ~bind_cols(asymptotic_test (
+                                 .x$value,
+                                 .x$period)))) %>%
+  unnest(asymptotic_test)
+# compute CV
+kwl_p_cvs_pre_post <-
+  kwl_p_metrics_long_post_chi %>%
+  group_by(period, variable) %>%
+  summarise(cvs = raster::cv(value, na.rm = TRUE))
+# join p-values to CVs
+kwl_p_cvs_and_pvalues_post_chi <-
+  kwl_p_cvs_pre_post %>%
+  left_join(kiwulan_pottery_metrics_cvs_plot_cv_test_asym_post_chi) %>%
+  mutate(significant = ifelse(p_value <= 0.05, 'yes', 'no')) %>%
+  mutate(variable = factor(variable, ordered = TRUE)) %>%
+  arrange(variable)
+# making table for CV
+kwl_p_cvs_table_post_chi <-
+  kwl_p_cvs_and_pvalues_post_chi %>%
+  select(-data, -significant) %>%
+  pivot_wider(names_from = period, values_from = cvs) %>%
+  select(variable, `post-e`, `ch-con`, `D_AD`, `p_value`) %>%
+  mutate_at(c("post-e", "ch-con", "D_AD", "p_value"), round, 4)
+
+# join two table together
+kwl_p_cvs_table_all <-
+  kwl_p_cvs_table_pre_post %>% 
+  left_join(kwl_p_cvs_table_post_chi, "variable") %>% 
+  select(-`post-e.y`) %>% 
+  rename(`before European` = "pre-e", `European` = "post-e.x", Chinese = "ch-con",
+         `D_AD-1` = "D_AD.x", `D_AD-2` = "D_AD.y",
+         `P value-1` = `p_value.x`, `P value-2` = `p_value.y`) %>% 
+  select(variable, `before European`, European, Chinese,
+         `D_AD-1`, `P value-1`, `D_AD-2`, `P value-2`)
+
+knitr::kable(kwl_p_cvs_table_all,
+             caption = "Coefficient of Variation for metric attributes by three phases using Asymptotic test [@Feltz1996].
+The first set of P-value represents the comparison between before European and European periods, while the second represents the comparison between European and Chinese periods ")
 ```
@@
 
-A total of `r total_no` largely intact vessels from pre-contact contexts (n = 32), post-European contexts (n = 27), and Chinese contact contexts (n = 14) were selected for elliptic Fourier analysis.
+A total of `r total_no` largely intact vessels from pre-contact contexts (n = `r pre_no`), post-European contexts (n = `r Euro_no`), and Chinese contact contexts (n = `r Chi_no`) were selected for elliptic Fourier analysis.
 Most of them cover every vessel component from rim to bottom.
 To fully explore the pottery standardization, we also take the metric measurements into account for comparison.
-The total of 362 reconstructed pots (pre-European = 153, European = 173, Chinese = 36) were measured.
+The total of `r metric_all` reconstructed pots (pre-European = `r pre`, European = `r post_e`, Chinese = `r chi`) were measured.
 The metric attributes include thickness of rim, neck, and body, and diameter of rim, neck, and body.
@@
 ```{r CV-test-PCs-table}
-library(cvequality)
 pc1 <- data.frame(
@@
 
-When we examine metric data for six measurements and two ratio values using CV tests (\table here), it shows there is a significant difference (p-value less than 0.01) between CV values for two metric measurements across phases, the rim diameter and the neck diameter before and after the European presence.
+We compare the result of shape variable with the result of metric data for six measurements and two ratio values using CV test (Table \@ref(tab:metricCVstest)).
+Metric measurements show there is a significant difference (p-value less than 0.01) between CV values for two metric measurements across phases, the rim diameter and the neck diameter before and after the European presence.
 After the arrival of the Chinese, there is only one metric attribute, the rim diameter, showing significant difference compared with post-European period (p-value less than 0.05).
-Although these results shows the differences between different contact phases that corresponds to part of the results of shape variable, elliptic Fourier analysis captures more subtle differences in a substantially way reflected by overall shapes of pottery, not only limited to linear aspects.
+Although these results shows the differences between different contact phases that corresponds to part of the results of shape variable, elliptic Fourier analysis captures more subtle differences in a substantially way reflected by overall shapes of pottery, not only limited to diameter variables.
 

### 2020-02-05 analysis/paper/paper.Rmd
add sentences and a figure for spatial analysis in dicussion section and method section

@@
 # read metric data
-kwl_p_ave <- readr::read_csv(here::here("analysis", "data", "raw_data",
-                                        "KWL-pottery-metrics-tidy-ave.csv"))
+kwl_p_ave <- readr::read_csv(here("analysis", "data", "raw_data", "KWL-pottery-metrics-tidy-ave.csv"))
 # number for each phase 
@@
 Most of them cover every vessel component from rim to bottom.
-To fully explore the pottery standardization, we also take the metric measurements into account for comparison.
+The shape of vessel is the major object analyzed in this paper, however, to fully explore the pottery standardization, we also take the metric measurements into account for comparison.
 The total of `r metric_all` reconstructed pots (pre-European = `r pre`, European = `r post_e`, Chinese = `r chi`) were measured.
@@
 To obtain information about the pottery shape, we compare the ratio of Rim/Body thickness, and the ratio of Rim/Body diameter.
-Also, we explore spatial distribution of pottery to examine if clustering changes over time across the sampling area.
-As craft specialization increases, pottery distribution might present a shift from dispersed clusters to concentrated clusters that hints the emergence of specialized group working in workshop.
-
-<!-- we also look at spatial distribution to see if clustering changes over time...
-...as craft specialization increases, we might expect spatial clustering to increase...
-as possible indicator of craft specialization -->
+In addition, we use Monte Carlo method to test for randomness in spatial locations of pots to explore whether the distribution is meaningful indicated by clustered or dispersed pattern.
+As craft specialization increases, pottery distribution might present a shift from dispersed pattern to concentrated clusters that hints the emergence of specialized groups.
 
@@
 saveRDS(jpgs_imported_coo,
-        here::here("analysis/data/derived_data/jpgs_imported_coo.rds"))
+        here("analysis/data/derived_data/jpgs_imported_coo.rds"))
 ```
@@
 library(cowplot)
-jpgs_imported_coo <- readr::read_rds(here::here("analysis/data/derived_data/jpgs_imported_coo.rds"))
+jpgs_imported_coo <- readr::read_rds(here("analysis/data/derived_data/jpgs_imported_coo.rds"))
 
@@
 # Isodeformation lines
-png(here::here("analysis/figures/tps-iso.png"), res = 600, w = 1200, h = 1800)
+png(here("analysis/figures/tps-iso.png"), res = 600, w = 1200, h = 1800)
 
@@
 Although these results shows the differences between different contact phases that corresponds to part of the results of shape variable, elliptic Fourier analysis captures more subtle differences in a substantially way reflected by overall shapes of pottery, not only limited to diameter variables.
+In addition, results of spatial analysis for pottery shape samples show a non-randomly dispersed distribution before European contact and more extreme dispersed after European presence.
+However, the distribution of pottery is more similar to the random distributions during the Chinese period.
+This contradicts to our expectation that clustered patten will be observed with an increase in pottery standardization since the emergence of specialized groups.
+Or, this indicates that there are other factors lead to pottery standardization instead of a small number of artisans.
+
+```{r Monte-Carlo-spatial-pattern, fig.cap= "Histograms of simulated average nearest-neighbour distances (ANN) values from 1000 simulations for three phases.
+X-axis values based on meteres represent ANN expected value.
+Each sample distribution presents the null hypothesis with the blue line indicating the observed ANN value"}
+knitr::include_graphics(here("analysis", "figures", "plot-kde-ann-histograms.jpg"))
+```
 

### 2020-02-06 analysis/paper/paper.Rmd
fix typos

@@
 keywords: |
-  Geometric Morphometrics; ceramic specialization; Iron age; culture contact; standardization
+  Geometric Morphometrics; craft specialization; Iron age; culture contact; ceramic standardization
 highlights: |
@@
 
-The emergence of social inequality in the context of cross-cultural interaction, especially in a colonial situation, has been observed in many parts of world, for example, when foreign trade goods are introduced to local Indigenous societies.
+The emergence of social inequality in the context of cross-cultural interaction, especially in a colonial situation, has been observed in many parts of the world, for example, when foreign trade goods are introduced to local Indigenous societies.
 The monopolization of long-distance trade goods has caused substantial transformations of Indigenous economic, cultural, and socio-political systems [@Dietler2005; @Dietler1997; @Junker1993; @Silliman2005].
@@
 Our aim is to investigate if there was increasing ceramic specialization resulting from Indigenous interaction with Europeans in the 17th century or Chinese in the 19th century, two major foreign influences in early historical Taiwan, that might indicate social changes in local indigenous societies.
-Previous reports of archaeological evidence in northeastern Taiwan describes increases in the use of prestige goods, such as trade ornaments, and uneven artefact distribution patterns in domestic areas and burial contexts at the time when the Indigenous communities encountered the Spanish and the Dutch in the 17th century.
+Previous reports of archaeological evidence in northeastern Taiwan describes increases in the use of prestige goods, such as trade ornaments, and uneven artifact distribution patterns in domestic areas and burial contexts at the time when the Indigenous communities encountered the Spanish and the Dutch in the 17th century.
 Here we show that in addition to the differential distribution of trade ornaments post-European contact, the locally-made ceramics become highly consistent in their form and shape, compared to other pottery throughout northeastern Taiwan.
@@
 @Roux2018's ethnographic study in the Jodhpur region of India suggests the number of artisans can be statistically assessed by comparing coefficients of variation of linear dimensions of pots, and differences between artisans can be detected at both intra-individual level and inter-individual level in a region.
-However, because pottery vessels are typically have curved shapes, linear measurements have limited sensitivity to many kinds of shape variations.
+However, because pottery vessels typically have curved shapes, linear measurements have limited sensitivity to many kinds of shape variations.
 Thus, to capture subtle shape variations that might be relevant to standardization, a method that is sensitive to non-linear aspects of shape variation is necessary.
@@
 The emergence of craft specialization, here pottery specialization, is often related to changes in social organization towards a society with increasing status inequality.
-In this case, if the competition among Indigenous individual for foreign resources and trade partnership with European or Chinese colonizers gradually lead to increasing social inequality, then we expect the local ceramics will show more homogeneous shapes after contact due to the emergence of craft specialization caused by rgreater economic and social control by a small group of individuals at Kiwulan.
+In this case, if the competition among Indigenous individual for foreign resources and trade partnership with European or Chinese colonizers gradually lead to increasing social inequality, then we expect the local ceramics will show more homogeneous shapes after contact due to the emergence of craft specialization caused by greater economic and social control by a small group of individuals at Kiwulan.
 Our results are important for understanding the indirect influences of  European colonists on local Indigenous societies, which remains unclear in East Asia where colonial power had limited success compared to other places in Southeast Asia.
@@
 
-However, one limitation of landmark approaches is that landmarks may not be able to capture the shape differences of a morphological structure where the curving outlines between landmarks contains crucial variation.
+However, one limitation of landmark approaches is that landmarks may not be able to capture the shape differences of a morphological structure where the curving outlines between landmarks contain crucial variation.
 Furthermore, landmark points may be unavailable for structures that are mostly or entirely curves.
@@
 Among those highly standardized jars, some were made by specialists attached to elites indicating the presence of social inequality, while others were made by independent specialists according to their spatial distribution patterns.
-Similarity, @Selden2019 examined 28 Caddo bottle excavated at several sites (AD 500-1700) in northwest Louisiana using semi-landmarks method.
+Similarity, @Selden2019 examined 28 Caddo bottles excavated at several sites (AD 500-1700) in northwest Louisiana using semi-landmarks method.
 The results indicate a significant integration between the different parts of the bottles including rim, neck, body, and base that vary in a coordinated manner.
@@
 Taking the ceramics data from Kiwulan, northeastern Taiwan, we use EFA to evaluate the level of standardization of ceramics in relation to the European presence in the 17th century to get insights into the emergence of ceramic specialization.
-We use a significance test for the equality of coefficient of variations of shape variables to statistically compare the vessel standarization from different periods.
+We use a significance test for the equality of coefficient of variations of shape variables to statistically compare the vessel standardization from different periods.
 Using pottery shape as a proxy to study craft specialization, we address these questions: Did foreign influence have impacts on indigenous pottery production that can be detected in the shape of the vessels?
@@
 To compare different foreign influences, we classified the upper component into three chronological phases: the pre-European, European, and Chinese periods.
-We identified these phases using 32 previously published radiocarbon ages [@Chen2007], excavation depth measurements, stratigraphic details reported by the excavators (colour, disturbance, etc.), and finds of chronologically diagnostic artifacts, such as blue and white porcelains, light grey glazed jars, and large dark brown glazed stoneware jars commonly used in the 17th century, and bricks and tiles used by the Chinese in the 19th century [@Hsieh2009; @Wang2011].
+We identified these phases using 32 previously published radiocarbon ages [@Chen2007], excavation depth measurements, stratigraphic details reported by the excavators (color, disturbance, etc.), and finds of chronologically diagnostic artifacts, such as blue and white porcelains, light grey glazed jars, and large dark brown glazed stoneware jars commonly used in the 17th century, and bricks and tiles used by the Chinese in the 19th century [@Hsieh2009; @Wang2011].
 The deposit shows signs of continuous human occupation during each of the three phases across the sampling area.
 
-The most abundant artefacts in the upper component are locally made ceramics, which are distributed throughout the sequence and across the site.
+The most abundant artifacts in the upper component are locally made ceramics, which are distributed throughout the sequence and across the site.
 More than 550,000 sherds were excavated, and around 1,200 vessels could be reconstructed.
@@
 The metric attributes include thickness of rim, neck, and body, and diameter of rim, neck, and body.
-For each pot, two measurements for thickness and diameter were took to compute an average value.
+For each pot, two measurements for thickness and diameter were taken to compute an average value.
 To obtain information about the pottery shape, we compare the ratio of Rim/Body thickness, and the ratio of Rim/Body diameter.
@@
 Geometric morphometric analyses were conducted using the R software [@Rlanguage2019] using the functions included in the Momocs, a R package intended to quantify and analyze shapes, especially for outline analysis [@Bonhomme2014].
-The digitised outlines were converted into a list of successive x-y pixel coordinates for elliptic Fourier analysis (EFA), to investigate morphological differences among pottery shapes from our three occupation phases.
-The harmonic coefficients generated by EFA were analyzed by principal component analysis (PCA) for dimensionality reduction to illustrate the diversity of the shape data and identify major patterns of variation through .
+The digilized outlines were converted into a list of successive x-y pixel coordinates for elliptic Fourier analysis (EFA), to investigate morphological differences among pottery shapes from our three occupation phases.
+The harmonic coefficients generated by EFA were analyzed by principal component analysis (PCA) for dimensionality reduction to illustrate the diversity of the shape data and identify major patterns of variation.
 
@@
 
-To answer the question of whether CV values for vessels samples across our three occupational phases are significantly different or not, we used the the Modified signed-likelihood ratio (MSLR) test for equality of CVs [@Krishnamoorthy2014].
+To answer the question of whether CV values for vessel samples across our three occupational phases are significantly different or not, we used the Modified signed-likelihood ratio (MSLR) test for equality of CVs [@Krishnamoorthy2014].
 While previous work has used the @Feltz1996 asymptotic test for the equality of coefficients of variation from k populations [@Lycett2008; @Okumura2014; @Eerkens2000; @Eerkens2001; @Hoggard2017], we prefer the MSLR test as a more recent development with lower rates of type I error, better performance with uneven sample numbers, and more power across a range of conditions [@Krishnamoorthy2014].
@@
 To see whether the differences in the distribution of PCs between any two phases are substantive or due to chance, we used a modified signed-likelihood ratio test to assess the equality of CVs.
-P-values for this significance test of CVs for PC1 and PC2 show significant differences in the standardization of vessel shapes across periods, especially between Chinese contact with either pre-European or post-European(Figure \@ref(fig:CV-test-PCs-plot), Table \@ref(tab:CV-test-PCs-table)).
+P-values for this significance test of CVs for PC1 and PC2 show significant differences in the standardization of vessel shapes across periods, especially between Chinese contact with either pre-European or post-European (Figure \@ref(fig:CV-test-PCs-plot), Table \@ref(tab:CV-test-PCs-table)).
 A significant difference is also detected between the pre-European and post-European periods.
@@
 These changes even more pronounced during the Chinese contact period.
-In general, pottery vessels become shorter in height over time, that leading to an oval-shaped body.
+In general, pottery vessels become shorter in height over time, leading to an oval-shaped body.
 
-Our CV tests indicates that there are significant differences in shape standardization between the pre-European period and post-European period, and between the Chinese contact and either pre-European or post-European periods.
+Our CV tests indicate that there are significant differences in shape standardization between the pre-European period and post-European period, and between the Chinese contact and either pre-European or post-European periods.
 This suggests pottery shape became more homogeneous and standardized after foreign contacts with European colonizers and even more so after contact with Chinese immigrants.
@@
 Unfortunately, we did not obtain useful results due to extremely low lipid yields, which is probably due to the very thin, dense, and low porosity fabric of Kiwulan pottery.
-These physical characteristics of the clay offers limited spaces to trap and protect organic molecules from microbiological degradation [cf.
+These physical characteristics of the clay offer limited spaces to trap and protect organic molecules from microbiological degradation [cf.
 @Evershed2008, pp.
@@
 After the arrival of the Chinese, there is only one metric attribute, the rim diameter, showing significant difference compared with post-European period (p-value less than 0.05).
-Although these results shows the differences between different contact phases that corresponds to part of the results of shape variable, elliptic Fourier analysis captures more subtle differences in a substantially way reflected by overall shapes of pottery, not only limited to diameter variables.
+Although these results show the differences between different contact phases that corresponds to part of the results of shape variable, elliptic Fourier analysis captures more subtle differences in a substantial way reflected by overall shapes of pottery, not only limited to diameter variables.
 In addition, results of spatial analysis for pottery shape samples show a non-randomly dispersed distribution before European contact and more extreme dispersed after European presence.
 However, the distribution of pottery is more similar to the random distributions during the Chinese period.
-This contradicts to our expectation that clustered patten will be observed with an increase in pottery standardization since the emergence of specialized groups.
+This contradicts our expectation that clustered pattern will be observed with an increase in pottery standardization since the emergence of specialized groups.
 Or, this indicates that there are other factors lead to pottery standardization instead of a small number of artisans.
@@
 This should include the other two main datasets at Kiwulan, the ornaments assemblages, and the burial goods.
-We find vessel shapes were more standardised in the Chinese period than the European period, contrary to previous work that has downplayed the effect of Chinese settlement in Yilan [@Wang2011].
+We find vessel shapes were more standardized in the Chinese period than the European period, contrary to previous work that has downplayed the effect of Chinese settlement in Yilan [@Wang2011].
 Compared to other regions in Taiwan, European colonial influence was weak in Yilan due to the isolated of the surrounding mountains and the economic focus of the Spanish and  Dutch who preferred northern Taiwan as their trading base.
@@
 
-We recognise that the shape variation reported here is subtle, and this invites consideration of the possibility that the absence of major changes in vessel shape at Kiwulan may have been an act of resistance to foreign influence.
+We recognize that the shape variation reported here is subtle, and this invites consideration of the possibility that the absence of major changes in vessel shape at Kiwulan may have been an act of resistance to foreign influence.
 Continuities in vessel shape over time draws our attention to the endurance of traditional pottery production practices amid intrusions from Europeans and Chinese.

### 2020-02-06 analysis/paper/paper.Rmd
add cross reference and one figure

@@
 Although these results show the differences between different contact phases that corresponds to part of the results of shape variable, elliptic Fourier analysis captures more subtle differences in a substantial way reflected by overall shapes of pottery, not only limited to diameter variables.
-In addition, results of spatial analysis for pottery shape samples show a non-randomly dispersed distribution before European contact and more extreme dispersed after European presence.
+In addition, results of spatial analysis (Figure \@ref(fig:kernel-plot); Figure \@ref(fig:CV-test-PCs-plot)) for pottery shape samples show a non-randomly dispersed distribution before European contact and more extreme dispersed after European presence.
 However, the distribution of pottery is more similar to the random distributions during the Chinese period.
@@
 Or, this indicates that there are other factors lead to pottery standardization instead of a small number of artisans.
+
+```{r kernel-plot, fig.cap= "Kernel density map visulizes the probability of the density of pottery across space.
+The map shows  major core area during the pre-European period, multiple core areas during the European period, and a single core during the Chinese period.
+Used the bandwidth based on Silverman (1986)'s rule of thumb"}
+knitr::include_graphics(here("analysis", "figures", "plot-kde-maps.jpg"))
+```
 

### 2020-02-06 analysis/paper/paper.Rmd
work on discussion section

@@
 Although these results show the differences between different contact phases that corresponds to part of the results of shape variable, elliptic Fourier analysis captures more subtle differences in a substantial way reflected by overall shapes of pottery, not only limited to diameter variables.
-In addition, results of spatial analysis (Figure \@ref(fig:kernel-plot); Figure \@ref(fig:CV-test-PCs-plot)) for pottery shape samples show a non-randomly dispersed distribution before European contact and more extreme dispersed after European presence.
-However, the distribution of pottery is more similar to the random distributions during the Chinese period.
+In addition, the result of spatial analysis (Figure \@ref(fig:kernel-plot)) for pottery shape samples presents multiple clusters with high density of pottery during the European presence.
+However, the hypothesis testing on spatial randomness indicates a non-randomly dispersed distribution before European contact and more extreme dispersed after European presence.
+In contrast, the distribution of pottery is more similar to the random distributions during the Chinese period.
 This contradicts our expectation that clustered pattern will be observed with an increase in pottery standardization since the emergence of specialized groups.
-Or, this indicates that there are other factors lead to pottery standardization instead of a small number of artisans.
+Or, this may indicate that there are other factors lead to pottery standardization instead of a small number of artisans.
 

### 2020-02-10 analysis/paper/paper.Rmd
proof from introduction to method section

@@
 However, it is important to note that explanations for the emergence of specialized production in past societies might not be so straightforward or singular.
-@Acabado2018's case study in lowland northern Philippines shows that specialization is not necessarily accompanied with social differentiation based on a low degree of ceramic specialization in Ifugao ranked societies during the time of Spanish colonization.
+@Acabado2018's case study in lowland northern Philippines shows that specialization is not necessarily accompanied with social differentiation based on a low degree of ceramic specialization in Ifugao ranked societies during the time of Spanish occupation.
 The development of specialization could be dependent on unique historical factors, where the conditions and contexts under which specialists organized must be taken into account when making inferences about changes in social organization [@Roux2015; @Costin2001].
@@
 The monopolization of long-distance trade goods has caused substantial transformations of Indigenous economic, cultural, and socio-political systems [@Dietler2005; @Dietler1997; @Junker1993; @Silliman2005].
-Pericolonial archaeology is the study of these indirect effects of colonialism, investigating areas where European colonial rule was limited and their conquests were short-lived and unsuccessful, but their colonial activities had economic and political impacts on Indigenous peoples living on the periphery of colonial control and influence [@Trabert2017; @Acabado2017].
+Pericolonial archaeology is the study of these indirect effects of colonialism, investigating areas where European colonial rule was limited and their conquests were short-lived and unsuccessful, but their colonial activities had economic and political impacts on Indigenous peoples living on the periphery of colonial control [@Trabert2017; @Acabado2017].
 @Acabado2017's study of Ifugao society in the  Philippines highland suggests economic and political intensification during the Spanish presence as the response of Indigenous peoples to the Spanish cooptation.
 Indigenous societies might not simply passively accept colonial rule, but instead actively negotiate with the colonist, and accommodate or resist foreign intrusion through their daily cultural practices, such as their consumption patterns of foreign goods [@Dietler2015; @Given2004; @Mullins2011; @Scaramelli2005; @Silliman2001; @TorrenceandClarke2000].
-In this paper we present a case study from Northeastern Taiwan as an example of these indirect impacts of foreign occupations on local Indigenous societies.
+In this paper we present a case study from northeastern Taiwan as an example of these indirect impacts of foreign occupations on local Indigenous societies.
 Our aim is to investigate if there was increasing ceramic specialization resulting from Indigenous interaction with Europeans in the 17th century or Chinese in the 19th century, two major foreign influences in early historical Taiwan, that might indicate social changes in local indigenous societies.
 Previous reports of archaeological evidence in northeastern Taiwan describes increases in the use of prestige goods, such as trade ornaments, and uneven artifact distribution patterns in domestic areas and burial contexts at the time when the Indigenous communities encountered the Spanish and the Dutch in the 17th century.
-Here we show that in addition to the differential distribution of trade ornaments post-European contact, the locally-made ceramics become highly consistent in their form and shape, compared to other pottery throughout northeastern Taiwan.
+Here we show that in addition to the differential distribution of trade ornaments post-European contact, the locally-made ceramics become highly consistent in their form and shape, compared to other pottery throughout Taiwan.
 This shift to higher consistency hints at the possibility of craft specialization in pottery manufacture.
@@
 Our results are important for understanding the indirect influences of  European colonists on local Indigenous societies, which remains unclear in East Asia where colonial power had limited success compared to other places in Southeast Asia.
-Our findings improve our understanding of the relationship between the use of prestige goods, the degree of ceramic specialization, and the influence of foreign colonizers in a pericolonial context [cf.
+Our findings improve our understanding of the relationship between the use of imported prestige goods, the degree of ceramic specialization, and the influence of foreign colonizers in a pericolonial context [cf.
 @Acabado2017].
@@
 Ceramics analyzed in this paper come from 40 units (4m by 4m each) sampled from the central, undisturbed area of archaeological excavations at Kiwulan (Figure \@ref(fig:KWL-map); Figure \@ref(fig:KWL-sam-area)).
-Kiwulan is situated on a hill near a riverside at the northern margin of Yilan Province, which is characterized by a triangular alluvial plain facing eastwards the Pacific with high mountains on three other sides.
+Kiwulan is situated on a hill near a riverside at the northern margin of Yilan County, which is characterized by a triangular alluvial plain facing eastwards the Pacific with high mountains on three other sides.
 Yilan is an ideal context to study peripheral colonial influence because Indigenous communities there were isolated by natural barriers that limited the frequency of direct contact with the Spanish and the Dutch settlers in northern Taiwan.
 The chronology of the archaeological deposits at Kiwulan consists of two cultural components, the upper component and the lower component, with a sterile layer in between [@Chen2007].
-This paper focuses on the upper component, dated from AD 1350 to AD 1950, covering the late Iron Age and the historical period, which we define as the European presence in Taiwan in the early 17th century.
+This paper focuses on the upper component, dated from AD 1350 to AD 1950, covering the late Iron Age and the historical period, which we define as the European occupation in Taiwan in the early 17th century.
 The Dutch first occupied southern Taiwan in 1624 and then the Spanish occupied northern Taiwan in 1626 [@Andrade2007].
@@
 
+```{r, KWL-map, fig.cap = "Map showing the location of Kiwulan, and other places in northern Taiwan named in the text. Map data from naturalearthdata.com", fig.width=image_width_2_col_inch}
+library(here)
+knitr::include_graphics(here("analysis", "figures", "kiwulan-location-map.jpg"))
+```
+
+```{r, KWL-sam-area, fig.cap = "Map showing the largest section of excavation areas at Kiwulan, and the distribution of forty squares sampled in this paper presented in red with square ID number.
+Small dots represent the location of post-holes.
+Each square is 4 x 4 m", fig.width=image_width_2_col_inch}
+knitr::include_graphics(here("analysis", "figures", "KWL-excavation-map.jpg"))
+```
+
 The archaeological record of Kiwulan's upper component shows traces of foreign contacts, including Europeans in the 17th century, and waves of Chinese immigrants in the 19th century.
-Imported ceramics from mainland China, stoneware, and ornaments such as beads have been found in the upper component, indicating frequent long-distance trade activities in the 17th century during the European presence in northern Taiwan.
+Imported ceramics from mainland China, stoneware, and ornaments such as beads have been found in the upper component, indicating frequent long-distance trade activities with Europeans and Chinese merchants.
 In addition to artifacts, archaeological features such as burials, middens, and post-holes with _in-situ_ posts are widespread across the 1-2 m thick deposit of the upper component, and demonstrate that Kiwulan was a continuously occupied large settlement site [@Chen2007].
 To compare different foreign influences, we classified the upper component into three chronological phases: the pre-European, European, and Chinese periods.
-We identified these phases using 32 previously published radiocarbon ages [@Chen2007], excavation depth measurements, stratigraphic details reported by the excavators (color, disturbance, etc.), and finds of chronologically diagnostic artifacts, such as blue and white porcelains, light grey glazed jars, and large dark brown glazed stoneware jars commonly used in the 17th century, and bricks and tiles used by the Chinese in the 19th century [@Hsieh2009; @Wang2011].
+We identified these phases using 32 previously published radiocarbon ages [@Chen2007], excavation depth measurements, stratigraphic details reported by the excavators (color, texture, disturbance, etc.), and finds of chronologically diagnostic artifacts, such as blue and white porcelains, light grey glazed jars, and large dark brown glazed stoneware jars commonly used in the 17th century, and bricks and tiles used by the Chinese in the 19th century [@Chen2007; @Hsieh2009; @Wang2011].
 The deposit shows signs of continuous human occupation during each of the three phases across the sampling area.
@@
 This kind of pot has been widely found at archaeological sites during the late Iron Age and the historical period throughout the Yilan Plain.
-
-```{r, KWL-map, fig.cap = "Map showing the location of Kiwulan, and other places in northern Taiwan named in the text. Map data from naturalearthdata.com", fig.width=image_width_2_col_inch}
-library(here)
-knitr::include_graphics(here("analysis", "figures", "kiwulan-location-map.jpg"))
-```
-
-```{r, KWL-sam-area, fig.cap = "Map showing the largest section of excavation areas at Kiwulan, and the distribution of forty squares sampled in this paper presented in red with square ID number.
-Small dots represent the location of post-holes.
-Each square is 4 x 4 m", fig.width=image_width_2_col_inch}
-knitr::include_graphics(here("analysis", "figures", "KWL-excavation-map.jpg"))
-```
 
@@
 
-```{r metricCVstest}
-library(cvequality)
-# read metric data
-kwl_p_ave <- readr::read_csv(here("analysis", "data", "raw_data", "KWL-pottery-metrics-tidy-ave.csv"))
-# number for each phase 
-kwl_p_ave_total <-
-  kwl_p_ave %>% 
-  count(period)
-# inline code
-metric_all <- sum(kwl_p_ave_total$n)
-pre <- kwl_p_ave_total$n[3]
-euro <- kwl_p_ave_total$n[2]
-chi <- kwl_p_ave_total$n[1]
-
-# select metric attributes and tidy data for computing CV
-kwl_p_metrics_long <-
-  kwl_p_ave %>%
-  mutate(dia_rim_body = Dia_Rim_ave/Dia_Body_ave,
-         thi_rim_body = Thick_Rim_ave/Thick_Body_ave) %>%
-  select(period, 
-         Dia_Rim_ave, Dia_Neck_ave, 
-         Dia_Body_ave, Thick_Rim_ave, 
-         Thick_Neck_ave, Thick_Body_ave,
-         dia_rim_body, thi_rim_body,) %>%
-  rename(`Rim diameter (mm)` = Dia_Rim_ave,
-         `Neck diameter (mm)` = Dia_Neck_ave,
-         `Body diameter (mm)` = Dia_Body_ave,
-         `Rim thickness (mm)` = Thick_Rim_ave,
-         `Neck thickness (mm)` = Thick_Neck_ave,
-         `Body thickness (mm)` = Thick_Body_ave,
-         `Ratio of Rim/Body diameter` = dia_rim_body,
-         `Ratio of Rim/Body thickness` = thi_rim_body) %>%
-  gather(variable, value, -period) %>%
-  filter(!is.na(period), !is.na(value)) %>% 
-  mutate(variable = factor(variable, levels = c("Rim thickness (mm)",
-                                                "Neck thickness (mm)",
-                                                "Body thickness (mm)",
-                                                "Ratio of Rim/Body thickness",
-                                                "Rim diameter (mm)",
-                                                "Neck diameter (mm)",
-                                                "Body diameter (mm)",
-                                                "Ratio of Rim/Body diameter"))) %>%
-  arrange(variable)
-
-# filter pre-e and post-e periods
-kwl_p_metrics_long_pre_post <-
-  kwl_p_metrics_long %>%
-  filter(!period == "ch-con")
-# filter post-e and chi periods
-kwl_p_metrics_long_post_chi <-
-  kwl_p_metrics_long %>%
-  filter(!period == "pre-e")
-
-#---------------------------CV test for pre-e and post-e pair
-kiwulan_pottery_metrics_cvs_plot_cv_test_asym_pre_post <-
-  kwl_p_metrics_long_pre_post %>%
-  nest(-variable) %>%  # or nest(period, value) %>%
-  mutate(asymptotic_test = map(data, # change from mslr_test to asymptotic_test
-                               ~bind_cols(asymptotic_test (
-                                 .x$value,
-                                 .x$period)))) %>%
-  unnest(asymptotic_test)
-# compute CV
-kwl_p_cvs_pre_post <-
-  kwl_p_metrics_long_pre_post %>%
-  group_by(period, variable) %>%
-  summarise(cvs = raster::cv(value, na.rm = TRUE))
-# join p-values to CVs
-kwl_p_cvs_and_pvalues_pre_post <-
-  kwl_p_cvs_pre_post %>%
-  left_join(kiwulan_pottery_metrics_cvs_plot_cv_test_asym_pre_post) %>%
-  mutate(significant = ifelse(p_value <= 0.05, 'yes', 'no')) %>%
-  mutate(variable = factor(variable, ordered = TRUE)) %>%
-  arrange(variable)
-# making table for CV
-kwl_p_cvs_table_pre_post <-
-  kwl_p_cvs_and_pvalues_pre_post %>%
-  select(-data, -significant) %>%
-  pivot_wider(names_from = period, values_from = cvs) %>%
-  select(variable, `pre-e`, `post-e`, `D_AD`, `p_value`) %>%
-  mutate_at(c("pre-e", "post-e", "D_AD", "p_value"), round, 4)
-
-#---------------------------CV test for post-e and ch-con pair
-kiwulan_pottery_metrics_cvs_plot_cv_test_asym_post_chi <-
-  kwl_p_metrics_long_post_chi %>%
-  nest(-variable) %>%  # or nest(period, value) %>%
-  mutate(asymptotic_test = map(data, # change from mslr_test to asymptotic_test
-                               ~bind_cols(asymptotic_test (
-                                 .x$value,
-                                 .x$period)))) %>%
-  unnest(asymptotic_test)
-# compute CV
-kwl_p_cvs_pre_post <-
-  kwl_p_metrics_long_post_chi %>%
-  group_by(period, variable) %>%
-  summarise(cvs = raster::cv(value, na.rm = TRUE))
-# join p-values to CVs
-kwl_p_cvs_and_pvalues_post_chi <-
-  kwl_p_cvs_pre_post %>%
-  left_join(kiwulan_pottery_metrics_cvs_plot_cv_test_asym_post_chi) %>%
-  mutate(significant = ifelse(p_value <= 0.05, 'yes', 'no')) %>%
-  mutate(variable = factor(variable, ordered = TRUE)) %>%
-  arrange(variable)
-# making table for CV
-kwl_p_cvs_table_post_chi <-
-  kwl_p_cvs_and_pvalues_post_chi %>%
-  select(-data, -significant) %>%
-  pivot_wider(names_from = period, values_from = cvs) %>%
-  select(variable, `post-e`, `ch-con`, `D_AD`, `p_value`) %>%
-  mutate_at(c("post-e", "ch-con", "D_AD", "p_value"), round, 4)
-
-# join two table together
-kwl_p_cvs_table_all <-
-  kwl_p_cvs_table_pre_post %>% 
-  left_join(kwl_p_cvs_table_post_chi, "variable") %>% 
-  select(-`post-e.y`) %>% 
-  rename(`before European` = "pre-e", `European` = "post-e.x", Chinese = "ch-con",
-         `D_AD-1` = "D_AD.x", `D_AD-2` = "D_AD.y",
-         `P value-1` = `p_value.x`, `P value-2` = `p_value.y`) %>% 
-  select(variable, `before European`, European, Chinese,
-         `D_AD-1`, `P value-1`, `D_AD-2`, `P value-2`)
-
-knitr::kable(kwl_p_cvs_table_all,
-             caption = "Coefficient of Variation for metric attributes by three phases using Asymptotic test [@Feltz1996].
-The first set of P-value represents the comparison between before European and European periods, while the second represents the comparison between European and Chinese periods ")
-```
-
 # Methods
@@
 The shape of vessel is the major object analyzed in this paper, however, to fully explore the pottery standardization, we also take the metric measurements into account for comparison.
-The total of `r metric_all` reconstructed pots (pre-European = `r pre`, European = `r post_e`, Chinese = `r chi`) were measured.
+The total of `r metric_all` reconstructed pots (pre-European = `r pre`, European = `r euro`, Chinese = `r chi`) were measured.
 The metric attributes include thickness of rim, neck, and body, and diameter of rim, neck, and body.
@@
 To obtain information about the pottery shape, we compare the ratio of Rim/Body thickness, and the ratio of Rim/Body diameter.
-In addition, we use Monte Carlo method to test for randomness in spatial locations of pots to explore whether the distribution is meaningful indicated by clustered or dispersed pattern.
-As craft specialization increases, pottery distribution might present a shift from dispersed pattern to concentrated clusters that hints the emergence of specialized groups.
+In addition, we use Monte Carlo test for randomness in spatial locations of pots to explore whether their distribution is meaningful, such as clustered or dispersed pattern.
+As craft specialization increases, pottery distribution might shift from a dispersed pattern to concentrated clusters that reflects the location of production [@Costin2001].
 
@@
 Where only half cross-section images were available, these were duplicated, flipped, and then joined with another one to create a 2D closed outline for each vessel.
-Geometric morphometric analyses were conducted using the R software [@Rlanguage2019] using the functions included in the Momocs, a R package intended to quantify and analyze shapes, especially for outline analysis [@Bonhomme2014].
-The digilized outlines were converted into a list of successive x-y pixel coordinates for elliptic Fourier analysis (EFA), to investigate morphological differences among pottery shapes from our three occupation phases.
+Geometric morphometric analyses were conducted using the R software [@Rlanguage2019] and the functions included in the Momocs, a R package intended to quantify and analyze shapes, especially for outline analysis [@Bonhomme2014].
+The digilized outlines were converted into a list of successive xy pixel coordinates for elliptic Fourier analysis (EFA), to investigate morphological differences among pottery shapes from our three occupation phases.
 The harmonic coefficients generated by EFA were analyzed by principal component analysis (PCA) for dimensionality reduction to illustrate the diversity of the shape data and identify major patterns of variation.
@@
 By standardizing standard deviation in each data set, the coefficients of variation (CV) values allows us to directly compare variation in samples measured with different units or means.
-CV values are robust to different sample sizes in the condition that the mean-standard deviation relationship is linear and positive since standard deviation is scaled to the mean [@Eerkens2001, pp.496].
+This is useful to compare the degree of standardization for archaeological assembles since CV scales standard deviation to the mean that enables comparison of variation across different sample sizes [@Eerkens2001, pp.498].
 We take this as our measurement of standardization in vessel shape variables: lower CV values reflect higher standardization, and thus increased craft specialization in the community.
@@
 To answer the question of whether CV values for vessel samples across our three occupational phases are significantly different or not, we used the Modified signed-likelihood ratio (MSLR) test for equality of CVs [@Krishnamoorthy2014].
-While previous work has used the @Feltz1996 asymptotic test for the equality of coefficients of variation from k populations [@Lycett2008; @Okumura2014; @Eerkens2000; @Eerkens2001; @Hoggard2017], we prefer the MSLR test as a more recent development with lower rates of type I error, better performance with uneven sample numbers, and more power across a range of conditions [@Krishnamoorthy2014].
+While previous work has used the @Feltz1996 asymptotic test for the equality of coefficients of variation from k populations [@Lycett2008; @Okumura2014; @Eerkens2000; @Eerkens2001; @Hoggard2017], we prefer the MSLR test for shape variables as a more recent development with lower rates of type I error, better performance with uneven sample numbers, and more power across a range of conditions [@Krishnamoorthy2014].
 
@@
 
+```{r metricCVstest}
+library(cvequality)
+# read metric data
+kwl_p_ave <- readr::read_csv(here("analysis", "data", "raw_data", "KWL-pottery-metrics-tidy-ave.csv"))
+# number for each phase 
+kwl_p_ave_total <-
+  kwl_p_ave %>% 
+  count(period)
+# inline code
+metric_all <- sum(kwl_p_ave_total$n)
+pre <- kwl_p_ave_total$n[3]
+euro <- kwl_p_ave_total$n[2]
+chi <- kwl_p_ave_total$n[1]
+
+# select metric attributes and tidy data for computing CV
+kwl_p_metrics_long <-
+  kwl_p_ave %>%
+  mutate(dia_rim_body = Dia_Rim_ave/Dia_Body_ave,
+         thi_rim_body = Thick_Rim_ave/Thick_Body_ave) %>%
+  select(period, 
+         Dia_Rim_ave, Dia_Neck_ave, 
+         Dia_Body_ave, Thick_Rim_ave, 
+         Thick_Neck_ave, Thick_Body_ave,
+         dia_rim_body, thi_rim_body,) %>%
+  rename(`Rim diameter (mm)` = Dia_Rim_ave,
+         `Neck diameter (mm)` = Dia_Neck_ave,
+         `Body diameter (mm)` = Dia_Body_ave,
+         `Rim thickness (mm)` = Thick_Rim_ave,
+         `Neck thickness (mm)` = Thick_Neck_ave,
+         `Body thickness (mm)` = Thick_Body_ave,
+         `Ratio of Rim/Body diameter` = dia_rim_body,
+         `Ratio of Rim/Body thickness` = thi_rim_body) %>%
+  gather(variable, value, -period) %>%
+  filter(!is.na(period), !is.na(value)) %>% 
+  mutate(variable = factor(variable, levels = c("Rim thickness (mm)",
+                                                "Neck thickness (mm)",
+                                                "Body thickness (mm)",
+                                                "Ratio of Rim/Body thickness",
+                                                "Rim diameter (mm)",
+                                                "Neck diameter (mm)",
+                                                "Body diameter (mm)",
+                                                "Ratio of Rim/Body diameter"))) %>%
+  arrange(variable)
+
+# filter pre-e and post-e periods
+kwl_p_metrics_long_pre_post <-
+  kwl_p_metrics_long %>%
+  filter(!period == "ch-con")
+# filter post-e and chi periods
+kwl_p_metrics_long_post_chi <-
+  kwl_p_metrics_long %>%
+  filter(!period == "pre-e")
+
+#---------------------------CV test for pre-e and post-e pair
+kiwulan_pottery_metrics_cvs_plot_cv_test_asym_pre_post <-
+  kwl_p_metrics_long_pre_post %>%
+  nest(-variable) %>%  # or nest(period, value) %>%
+  mutate(asymptotic_test = map(data, # change from mslr_test to asymptotic_test
+                               ~bind_cols(asymptotic_test (
+                                 .x$value,
+                                 .x$period)))) %>%
+  unnest(asymptotic_test)
+# compute CV
+kwl_p_cvs_pre_post <-
+  kwl_p_metrics_long_pre_post %>%
+  group_by(period, variable) %>%
+  summarise(cvs = raster::cv(value, na.rm = TRUE))
+# join p-values to CVs
+kwl_p_cvs_and_pvalues_pre_post <-
+  kwl_p_cvs_pre_post %>%
+  left_join(kiwulan_pottery_metrics_cvs_plot_cv_test_asym_pre_post) %>%
+  mutate(significant = ifelse(p_value <= 0.05, 'yes', 'no')) %>%
+  mutate(variable = factor(variable, ordered = TRUE)) %>%
+  arrange(variable)
+# making table for CV
+kwl_p_cvs_table_pre_post <-
+  kwl_p_cvs_and_pvalues_pre_post %>%
+  select(-data, -significant) %>%
+  pivot_wider(names_from = period, values_from = cvs) %>%
+  select(variable, `pre-e`, `post-e`, `D_AD`, `p_value`) %>%
+  mutate_at(c("pre-e", "post-e", "D_AD", "p_value"), round, 4)
+
+#---------------------------CV test for post-e and ch-con pair
+kiwulan_pottery_metrics_cvs_plot_cv_test_asym_post_chi <-
+  kwl_p_metrics_long_post_chi %>%
+  nest(-variable) %>%  # or nest(period, value) %>%
+  mutate(asymptotic_test = map(data, # change from mslr_test to asymptotic_test
+                               ~bind_cols(asymptotic_test (
+                                 .x$value,
+                                 .x$period)))) %>%
+  unnest(asymptotic_test)
+# compute CV
+kwl_p_cvs_pre_post <-
+  kwl_p_metrics_long_post_chi %>%
+  group_by(period, variable) %>%
+  summarise(cvs = raster::cv(value, na.rm = TRUE))
+# join p-values to CVs
+kwl_p_cvs_and_pvalues_post_chi <-
+  kwl_p_cvs_pre_post %>%
+  left_join(kiwulan_pottery_metrics_cvs_plot_cv_test_asym_post_chi) %>%
+  mutate(significant = ifelse(p_value <= 0.05, 'yes', 'no')) %>%
+  mutate(variable = factor(variable, ordered = TRUE)) %>%
+  arrange(variable)
+# making table for CV
+kwl_p_cvs_table_post_chi <-
+  kwl_p_cvs_and_pvalues_post_chi %>%
+  select(-data, -significant) %>%
+  pivot_wider(names_from = period, values_from = cvs) %>%
+  select(variable, `post-e`, `ch-con`, `D_AD`, `p_value`) %>%
+  mutate_at(c("post-e", "ch-con", "D_AD", "p_value"), round, 4)
+
+# join two table together
+kwl_p_cvs_table_all <-
+  kwl_p_cvs_table_pre_post %>% 
+  left_join(kwl_p_cvs_table_post_chi, "variable") %>% 
+  select(-`post-e.y`) %>% 
+  rename(`before European` = "pre-e", `European` = "post-e.x", Chinese = "ch-con",
+         `D_AD-1` = "D_AD.x", `D_AD-2` = "D_AD.y",
+         `P value-1` = `p_value.x`, `P value-2` = `p_value.y`) %>% 
+  select(variable, `before European`, European, Chinese,
+         `D_AD-1`, `P value-1`, `D_AD-2`, `P value-2`)
+
+knitr::kable(kwl_p_cvs_table_all,
+             caption = "Coefficient of Variation for metric attributes by three phases using Asymptotic test [@Feltz1996].
+The first set of P-value represents the comparison between before European and European periods, while the second represents the comparison between European and Chinese periods ")
+```
+
 ```{r kernel-plot, fig.cap= "Kernel density map visulizes the probability of the density of pottery across space.

### 2020-02-12 analysis/paper/paper.Rmd
add sentence to methods and test relationship between mean and sd

@@
 We take this as our measurement of standardization in vessel shape variables: lower CV values reflect higher standardization, and thus increased craft specialization in the community.
+Given that CV is robust for positive values due to the representation by ratio, we normalized PC scores to a range between 0 and 1 before computation of CV.
 
@@
 ```{r PCs-distribution, fig.cap= "The distribution of PC scores by phases"}
+
+# function for normalized data
+normalize <- function(x) (x-min(x))/(max(x)-min(x))
+
+# for one matrix
+op.p_tmp <- 
+op.p$x %>% 
+  as_tibble() %>% 
+  mutate_all(normalize) %>% 
+  as.matrix()
+
+row.names(op.p_tmp) <- row.names(op.p$x )
+
+op.p$x <- op.p_tmp
+
 # what are the variences for each PC in each period?
-# CVs on shape PCs for standardization
 len <- length(op.p$x[,1])
@@
 ```{r CV-test-PCs-table}
+# CVs on shape PCs for standardization
 pc1 <- data.frame(

### 2020-02-12 analysis/paper/scraps.Rmd
proof from introduction to method section
add sentence to methods and test relationship between mean and sd

@@
 ```
+
+CV is more robust in the condition that the mean-standard deviation relationship is linear and positive [@Eerkens2001, pp.498]
+
+```{r mean-sd-relationship, eval = FALSE}
+# mean and sd by phases for PC scores
+pc1_mean_sd_by_phase <-
+  pc1 %>% 
+  group_by(y) %>% 
+  summarize(mean = mean(x),
+            sd = sd(x))
+
+pc1_mean_sd_plot <-
+  pc1_mean_sd_by_phase %>% 
+  ggplot(aes(mean, sd)) +
+  geom_point()
+
+pc2_mean_sd_by_phase <-
+  pc2 %>% 
+  group_by(y) %>% 
+  summarize(mean = mean(x),
+            sd = sd(x))
+
+pc2_mean_sd_plot <-
+  pc2_mean_sd_by_phase %>% 
+  ggplot(aes(mean, sd)) +
+  geom_point()
+
+pc3_mean_sd_by_phase <-
+  pc3 %>% 
+  group_by(y) %>% 
+  summarize(mean = mean(x),
+            sd = sd(x))
+
+pc3_mean_sd_plot <-
+  pc3_mean_sd_by_phase %>% 
+  ggplot(aes(mean, sd)) +
+  geom_point()
+```

### 2020-02-15 analysis/paper/paper.Rmd

@@
 Our case study, which includes an openly available research compendium of R code, represents an innovative application of outline-based methods in geometric morphometry to answer the anthropological questions of craft specialization.
-This study implies that craft specialization can be a deliberate act of resistance to show indigenous identity and ethnicity in a direct culture contact situation.
+This study implies that craft specialization can be a deliberate act of resistance to show Indigenous identity and ethnicity in a direct culture contact situation.
 keywords: |
@@
 In this paper we present a case study from northeastern Taiwan as an example of these indirect impacts of foreign occupations on local Indigenous societies.
-Our aim is to investigate if there was increasing ceramic specialization resulting from Indigenous interaction with Europeans in the 17th century or Chinese in the 19th century, two major foreign influences in early historical Taiwan, that might indicate social changes in local indigenous societies.
+Our aim is to investigate if there was increasing ceramic specialization resulting from Indigenous interaction with Europeans in the 17th century or Chinese in the 19th century, two major foreign influences in early historical Taiwan, that might indicate social changes in local Indigenous societies.
 Previous reports of archaeological evidence in northeastern Taiwan describes increases in the use of prestige goods, such as trade ornaments, and uneven artifact distribution patterns in domestic areas and burial contexts at the time when the Indigenous communities encountered the Spanish and the Dutch in the 17th century.
@@
 We use a significance test for the equality of coefficient of variations of shape variables to statistically compare the vessel standardization from different periods.
-Using pottery shape as a proxy to study craft specialization, we address these questions: Did foreign influence have impacts on indigenous pottery production that can be detected in the shape of the vessels?
+Using pottery shape as a proxy to study craft specialization, we address these questions: Did foreign influence have impacts on Indigenous pottery production that can be detected in the shape of the vessels?
 Did pottery shape become more homogeneous and standardized after foreign contacts with European colonizers or Chinese immigrants?
@@
 This contradicts our expectation that clustered pattern will be observed with an increase in pottery standardization since the emergence of specialized groups.
-Or, this may indicate that there are other factors lead to pottery standardization instead of a small number of artisans.
+Or, this may indicate that there are other factors lead to pottery shape standardization instead of a small number of artisans.
 
@@
 @Acabado2017].
-In contrast to the Indigenous-European interactions at Kiwulan, the interaction between indigenous people and Chinese in the 19th century appears to have been more intense and direct.
+In contrast to the Indigenous-European interactions at Kiwulan, the interaction between Indigenous people and Chinese in the 19th century appears to have been more intense and direct.
 Historical records indicate that Chinese immigrants settled in Yilan and lived closely with Kiwulan Indigenous societies [@Chen1963; @Ke1993].

### 2020-02-15 analysis/paper/scraps.Rmd

--- 2020-02-03 analysis/paper/scraps.Rmd
@@
 ```
+
+CV is more robust in the condition that the mean-standard deviation relationship is linear and positive [@Eerkens2001, pp.498].
+We examine the relationship between mean and sd for our observations for PC scores, three points referring to three phases here.
+The results show that some of them have negative relationships.
+However, it might be not reliable because we only have three data points.
+
+```{r mean-sd-relationship, eval = FALSE}
+# mean and sd by phases for PC scores
+pc1_mean_sd_by_phase <-
+  pc1 %>% 
+  group_by(y) %>% 
+  summarize(mean = mean(x),
+            sd = sd(x))
+
+pc1_mean_sd_plot <-
+  pc1_mean_sd_by_phase %>% 
+  ggplot(aes(mean, sd)) +
+  geom_point() +
+  geom_text(aes(label=y), 
+            hjust= 0, vjust= 0)
+
+pc2_mean_sd_by_phase <-
+  pc2 %>% 
+  group_by(y) %>% 
+  summarize(mean = mean(x),
+            sd = sd(x))
+
+pc2_mean_sd_plot <-
+  pc2_mean_sd_by_phase %>% 
+  ggplot(aes(mean, sd)) +
+  geom_point() +
+  geom_text(aes(label=y), 
+            hjust= 0, vjust= 0)
+
+pc3_mean_sd_by_phase <-
+  pc3 %>% 
+  group_by(y) %>% 
+  summarize(mean = mean(x),
+            sd = sd(x))
+
+pc3_mean_sd_plot <-
+  pc3_mean_sd_by_phase %>% 
+  ggplot(aes(mean, sd)) +
+  geom_point() +
+  geom_text(aes(label=y), 
+            hjust= 0, vjust= 0)
+```

### 2020-02-15 analysis/paper/paper.Rmd
fix two errors that prevented knitting

@@
 Chi_no <- phase_fre$n[3]
+
+# read metric data
+kwl_p_ave <- readr::read_csv(here("analysis", "data", "raw_data", "KWL-pottery-metrics-tidy-ave.csv"))
+
+# number for each phase 
+kwl_p_ave_total <-
+  kwl_p_ave %>% 
+  count(period)
+
+# inline code later in the paper
+metric_all <- sum(kwl_p_ave_total$n)
+pre <- kwl_p_ave_total$n[3]
+euro <- kwl_p_ave_total$n[2]
+chi <- kwl_p_ave_total$n[1]
 ```
@@
 ```{r CV-test-PCs-table}
+library(cvequality)
 # CVs on shape PCs for standardization
@@
 ```{r metricCVstest}
-library(cvequality)
-# read metric data
-kwl_p_ave <- readr::read_csv(here("analysis", "data", "raw_data", "KWL-pottery-metrics-tidy-ave.csv"))
-# number for each phase 
-kwl_p_ave_total <-
-  kwl_p_ave %>% 
-  count(period)
-# inline code
-metric_all <- sum(kwl_p_ave_total$n)
-pre <- kwl_p_ave_total$n[3]
-euro <- kwl_p_ave_total$n[2]
-chi <- kwl_p_ave_total$n[1]
 

### 2020-02-15 analysis/paper/rich-diffs-rmd-in-local-git-repo.R
Delete rich-diffs-rmd-in-local-git-repo.R

@@ -1,41 +1,0 @@
-# How to see rich diffs of two commits of a single R Markdown document in a local Git repo
-
-# https://github.com/lorenzwalthert/gitsum
-library("gitsum")
-library("tidyverse")
-
-# To browse the commits locally, open an RStudio
-# project that is using Git version control, then ...
-
-# Set the path within the project to the Rmd file
-# in the project that we want to compare versions of
-file_path <- "analysis/paper/paper.Rmd"
-
-# Make a data frame that stores the git commit history,
-# one row per commit
-init_gitsum()
-tbl_out <- parse_log_detailed() %>%
-  select(date, author_name, message, hash)
-
-# View this dataframe to pick the specific commits to compare
-View(tbl_out)
-
-# choose two commits to compare, here I've got the most recent
-# one (the last row of the table), and the one immediately before that
-i <- nrow(tbl_out) - 30
-j <- nrow(tbl_out) - 31
-
-# consrtruct the terminal commants using the SHA of the commits we want to
-# compare, and the file path
-git_show_str_1 <- str_glue('git show ', tbl_out[i, ]$hash, ":", file_path)
-git_show_str_2 <- str_glue('git show ', tbl_out[j, ]$hash, ":", file_path)
-
-# get the contents of the file at each commit and
-# store in a vector
-earlier_commit <- system(git_show_str_2, intern = TRUE)
-later_commit <- system(git_show_str_1, intern = TRUE)
-
-# compute differences and visualise
-diffobj::diffChr(earlier_commit,
-                 later_commit,
-                 mode = "sidebyside")

### 2020-02-15 analysis/paper/rich-diffs-rmd-on-GitHub.R
Delete rich-diffs-rmd-on-GitHub.R

@@ -1,39 +1,0 @@
-# How to see rich diffs of two commits of a single R Markdown document on a GitHub repo (on the master branch only)
-
-# Edit this to use the repo that has the file you want to inspect
-gh_account_and_repo <- "LiYingWang/kwl.pottery"
-
-# Edit this to use the single file that you want to inspect
-# start the path here from the top level of the github repo
-gh_file_path <- "analysis/paper/paper.Rmd"
-
-# The next line opens a web page to see the commit history
-# of this file so we can
-# copy the SHA ids to use in the next step. Look for the
-# little clipboard icon to click on to copy the full SHA,
-# and past them in the step below
-browseURL(file.path(
-  "https://github.com",
-  gh_account_and_repo,
-  "commits/master",
-  gh_file_path
-))
-
-# Edit these to get the full SHA for the commits you want to compare, get
-# the full SHA from the GitHub webpage that was opened in the previous step
-commit_1_sha <- "3c25822c902819e6db7b88dee767b0b3d0235499" # a recent commit
-commit_2_sha <- "db390f98c0b4589c02f07006823b7c20ecad2856" # an earlier commit
-
-
-# Don't change this
-gh_raw <- "https://raw.githubusercontent.com"
-
-# run these lines...
-later_commit <- file.path(gh_raw, gh_account_and_repo, commit_1_sha, gh_file_path)
-earlier_commit <- file.path(gh_raw, gh_account_and_repo, commit_2_sha, gh_file_path)
-
-# run these lines to view the rich diff:
-diffobj::diffFile(earlier_commit,
-                  later_commit,
-                  mode = "sidebyside")
-@benmarwick

### 2020-02-15 README.Rmd
typo

@@
 
-[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/LiYingWang/kwl.pottery/master?urlpath=rstudio)
+[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/LiYingWang/kwl.pottery/master?urlpath=rstudio) ![R-CMD-check-compendium](https://github.com/LiYingWang/kwl.pottery/workflows/R-CMD-check-compendium/badge.svg)
 

### 2020-02-15 README.Rmd
Update README.Rmd

--- 2019-12-19 README.Rmd
@@
 
-[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/LiYingWang/kwl.pottery/master?urlpath=rstudio)
+[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/LiYingWang/kwl.pottery/master?urlpath=rstudio) [![R-CMD-check-compendium](https://github.com/LiYingWang/kwl.pottery/workflows/R-CMD-check-compendium/badge.svg)](https://github.com/LiYingWang/kwl.pottery/actions)
 

### 2020-02-17 README.Rmd
Update README.Rmd

--- 2019-12-19 README.Rmd
@@
 
-[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/LiYingWang/kwl.pottery/master?urlpath=rstudio)
+[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/LiYingWang/kwl.pottery/master?urlpath=rstudio) [![run-docker](https://github.com/LiYingWang/kwl.pottery/workflows/.github/workflows/run-docker.yml/badge.svg)](https://github.com/LiYingWang/kwl.pottery/actions)
 

### 2020-04-10 analysis/paper/paper.Rmd
editing... abstract

@@
 ---
-title: "Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeastern Taiwan"
+title: "Investigating shape standardisation using geometric morphometry of pottery from Iron Age northeastern Taiwan"
 author:
@@
 Here we demonstrate how to investigate ceramic standardization by analyzing outlines of ceramic vessels.
-Outline analysis is an important method because, unlike more commonly-used landmark analysis methods, it can effectively quantify shape differences for objects that lack distinctive measurement points needed for landmark analysis.
+Outline analysis is useful because, unlike more commonly-used landmark analysis methods, it can effectively quantify shape differences for objects that lack distinctive measurement points needed for landmark analysis.
 We demonstrate this method using pottery from Kiwulan, a large multi-component Iron Age site (AD 1350-1850) in northeastern Taiwan.
@@
 We found significant differences in pottery shape and shape standardization that indicate changes in pottery production resulting from contact with mainland Han Chinese groups in northeastern Taiwan.
-We infer increasing craft specialization and changes in social organization.
 Our case study, which includes an openly available research compendium of R code, represents an innovative application of outline-based methods in geometric morphometry to answer the anthropological questions of craft specialization.
-This study implies that craft specialization can be a deliberate act of resistance to show Indigenous identity and ethnicity in a direct culture contact situation.
 keywords: |

### 2020-04-10 analysis/paper/paper.Rmd
edit intro section

@@
 
-The development of craft specialization is sometimes associated with social changes since it reflects changes in production systems that could inform on the social, economic, or political conditions of a society.
-We define specialization as "differential participation in specific economic activities", in which a particular good is produced by fewer producers compared to the total population who consume it [@Costin1991; @Costin2001, pp.
-43].
-The effects of the emergence of specialization in past societies on their social structures and social processes is an important question in archaeological investigations of small scale societies in the Late Holocene.
-Considered through a political lens, craft specialization may be a strategy used by elites to control production systems to maintain their power and authority [@Costin2001].
-This relationship between craft specialization and social differentiation has been broadly discussed, and many cases show that increasingly specialized production can be a useful indicator of increasing socio-political complexity [@Hirshman2010; @Junker1999].
-That said, a more economic focus considers production, exchange, and consumption as a whole to explain specialization as part of a strategy to accommodate uneven distributions of resources [@Costin2001; @Arnold1994].
-Craft specialization may appear as a reaction to changes in subsistence practices [@Stark1995economic], or as a result of long-distance trade and inter-regional exchange [@Alizadeh2018].
-Present-day investigation of ceramic specialization in the Jodhpur region of India suggests that the adoption of standardized products across cultural boundaries is a reaction to a collapse of previous economic system, indicating a correlation with major socio-economic changes [@Roux2015].
-However, it is important to note that explanations for the emergence of specialized production in past societies might not be so straightforward or singular.
-@Acabado2018's case study in lowland northern Philippines shows that specialization is not necessarily accompanied with social differentiation based on a low degree of ceramic specialization in Ifugao ranked societies during the time of Spanish occupation.
-The development of specialization could be dependent on unique historical factors, where the conditions and contexts under which specialists organized must be taken into account when making inferences about changes in social organization [@Roux2015; @Costin2001].
+A major historical factor that is often linked to social change in small-scale societies is colonial settlers introducing foreign trade goods to local Indigenous societies.
+The monopolization of long-distance trade goods has caused substantial transformations of Indigenous economic, cultural, and socio-political systems (Dietler, 2005, 1997; Junker, 1993; Silliman, 2005).
+Pericolonial archaeology is the study of these indirect effects of colonialism, investigating areas where direct European colonial rule was limited, their conquests often short-lived and unsuccessful, but their commercial activities had economic and political impacts on Indigenous peoples living on the periphery of colonial control (Acabado, 2017; Trabert, 2017).
+Pericolonial situations were common during the seventeenth to nineteenth centuries in East and Southeast Asia where European trading activity was extensive, but direct European rule less widespread.
+An emerging priority in archaeological research in this region is identifying the indirect influences are apparent on Indigenous communities during this time.
+For example, Acabado (2017)’s study of Ifugao society in the Philippines highland suggests economic and political intensification during the Spanish presence as the response of Indigenous peoples to the Spanish cooptation.
 
-The emergence of social inequality in the context of cross-cultural interaction, especially in a colonial situation, has been observed in many parts of the world, for example, when foreign trade goods are introduced to local Indigenous societies.
-The monopolization of long-distance trade goods has caused substantial transformations of Indigenous economic, cultural, and socio-political systems [@Dietler2005; @Dietler1997; @Junker1993; @Silliman2005].
-Pericolonial archaeology is the study of these indirect effects of colonialism, investigating areas where European colonial rule was limited and their conquests were short-lived and unsuccessful, but their colonial activities had economic and political impacts on Indigenous peoples living on the periphery of colonial control [@Trabert2017; @Acabado2017].
-@Acabado2017's study of Ifugao society in the  Philippines highland suggests economic and political intensification during the Spanish presence as the response of Indigenous peoples to the Spanish cooptation.
-Indigenous societies might not simply passively accept colonial rule, but instead actively negotiate with the colonist, and accommodate or resist foreign intrusion through their daily cultural practices, such as their consumption patterns of foreign goods [@Dietler2015; @Given2004; @Mullins2011; @Scaramelli2005; @Silliman2001; @TorrenceandClarke2000].
-In this paper we present a case study from northeastern Taiwan as an example of these indirect impacts of foreign occupations on local Indigenous societies.
-Our aim is to investigate if there was increasing ceramic specialization resulting from Indigenous interaction with Europeans in the 17th century or Chinese in the 19th century, two major foreign influences in early historical Taiwan, that might indicate social changes in local Indigenous societies.
-Previous reports of archaeological evidence in northeastern Taiwan describes increases in the use of prestige goods, such as trade ornaments, and uneven artifact distribution patterns in domestic areas and burial contexts at the time when the Indigenous communities encountered the Spanish and the Dutch in the 17th century.
-Here we show that in addition to the differential distribution of trade ornaments post-European contact, the locally-made ceramics become highly consistent in their form and shape, compared to other pottery throughout Taiwan.
-This shift to higher consistency hints at the possibility of craft specialization in pottery manufacture.
+Indigenous societies’ responses to colonial contact range from passive acceptance through to active negotiation with the colonists, and accommodation or resistance of foreign intrusion.
+These responses can be identified through their daily cultural practices, such as their consumption patterns of foreign goods (Dietler, 2015; Given, 2004; Mullins, 2011; Scaramelli and Scaramelli, 2005; Silliman, 2001; Torrence and Clarke, 2000).
+In this paper we investigate the archaeology of a pericolonial situation at Kiwulan (1350-1950 AD), a large multi-component archaeological site in Yilan Province, northeastern Taiwan (Chen, 2007), to identify the indirect impacts of colonial settler activity on local Indigenous societies.
+Yilan is an ideal context to study peripheral colonial influences because Indigenous communities there were isolated by geographical barriers that limited the frequency of direct contact with the Spanish and the Dutch settlers in northern Taiwan.
+Kiwulan is situated on a hill near a riverside at the northern margin of Yilan County, which is characterized by a triangular alluvial plain facing eastwards the Pacific with high mountains on three other sides.
 
-To identify craft specialization in the archaeological record, variation in the standardization of a specialized product is often used as an indicator.
-This is based on the assumption that specialized mass production will lead to uniformity of the product due to routinization, higher skills, and fewer producers involved in production [@Stark1995; @Costin1991; @Costin2001].
-For investigating ceramic standardization, several measurements have been discussed and used, such as metric, compositional, and technological approaches [@Arnold2000; @Boness2015; @Rice1991; @Blackman1993; @Costin1991; @Tite1999; @Roux2015].
+Our specific aim is to investigate if there was increasing ceramic specialization resulting from Indigenous interaction with Europeans in the 17th century or Chinese in the 19th century, two major foreign influences in early historical Taiwan, that might indicate social changes in local Indigenous societies.
+We predict that competition within the Indigenous community at Kiwulan for foreign resources and trade partnerships with European or Chinese colonizers may have led to the emergence of craft specialization caused by greater economic and social control of pottery production by a small group of individuals.
+Using pottery shape as a proxy to study craft specialization, we ask: Did colonial trade impact the shape of locally-produced Indigenous pottery vessels?
+Did pottery shape become more homogeneous and standardized after foreign contacts with European colonizers or Chinese immigrants?
+
+Several measurements have been used for investigating ceramic standardization, such as metric, compositional, and technological variables (Arnold, 2000; Blackman et al., 1993; Boness et al., 2015; Costin, 1991; Rice, 1991; Roux, 2015; Tite, 1999).
 Among those variables, metric measurements are most widely applied to archaeological assemblages because of the ease of collecting these data.
-By comparing variations for sets of data using coefficient of variation coupled with statistical testing, the degree of standardization can be identified according to values [@Eerkens2001; @Roux2003; @Stark1995].
-@Roux2018's ethnographic study in the Jodhpur region of India suggests the number of artisans can be statistically assessed by comparing coefficients of variation of linear dimensions of pots, and differences between artisans can be detected at both intra-individual level and inter-individual level in a region.
+The coefficient of variation statistic is commonly used to quantify the degree of standardization in ceramic assemblages (Eerkens and Bettinger, 2001; Roux, 2003; Stark, 1995b).
 However, because pottery vessels typically have curved shapes, linear measurements have limited sensitivity to many kinds of shape variations.
-Thus, to capture subtle shape variations that might be relevant to standardization, a method that is sensitive to non-linear aspects of shape variation is necessary.
-Here we analyze pottery shape using geometric morphometric methods, a new approach for measuring the degree of standardization because it takes a statistical summary of the overall shape as variable for comparison [@Slice2007].
-Using geometric morphometry, we study shape and standardization of locally made ceramics at Kiwulan (1350-1950 AD) [@Chen2007], a large multi-component archaeological site in northeastern Taiwan, to identify changes in ceramic production resulting from cross-cultural interaction contexts.
-
-Our hypothesis is that if foreign influences had impacts on the emergence of social inequality in local Indigenous society due to monopolies of trade arising among a small number of Indigenous people, then we expect to see changes in social organization from a more corporate to a more network organization [@Feinman2000].
-We expect this corporate-network shift to manifest as increased specialization pottery manufacture, indicated archaeologically by increased standardization of pottery production at Kiwulan.
-The emergence of craft specialization, here pottery specialization, is often related to changes in social organization towards a society with increasing status inequality.
-In this case, if the competition among Indigenous individual for foreign resources and trade partnership with European or Chinese colonizers gradually lead to increasing social inequality, then we expect the local ceramics will show more homogeneous shapes after contact due to the emergence of craft specialization caused by greater economic and social control by a small group of individuals at Kiwulan.
-Our results are important for understanding the indirect influences of  European colonists on local Indigenous societies, which remains unclear in East Asia where colonial power had limited success compared to other places in Southeast Asia.
-Our findings improve our understanding of the relationship between the use of imported prestige goods, the degree of ceramic specialization, and the influence of foreign colonizers in a pericolonial context [cf.
-@Acabado2017].
+Thus, to capture subtle shape variations that might also be relevant to standardization, we analyze pottery shape using geometric morphometric methods.
 

### 2020-04-10 analysis/paper/paper.Rmd
edits to gmm section

@@
 
-Geometric morphometric methods (GMM) has been increasingly applied in archaeology for shape analysis, which explores morphological variability and similarity of archaeological materials to address the questions of anthropological interest.
-Geometric morphometrics differs from traditional linear measurements through its use of Cartesian coordinates of morphological structures to define shapes [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007].
+Geometric morphometrics differs from traditional linear measurements through its use of Cartesian coordinates of morphological structures to define shapes (Adams et al., 2004; Bookstein, 1997; Lawing and Polly, 2010; Slice, 2007).
 Landmarks, curves or outlines of objects can be represented by coordinates in terms of their unique point locations with respect to numerical values on coordinate axes.
-There are two common morphometric methods: landmark approaches and outline approaches [@Adams2004].
-Landmark approaches assign a set of landmarks or semilandmarks onto objects as reference points that can be specified on a coordinate system as xy coordinates for two dimensions, or xyz coordinates for three dimensions.
-In a typical analysis, starting with raw landmarks, a sample of objects are translated to a common centroid, rescaled into the same size, and rotated until the summed squared distances between corresponding landmarks are minimized using generalized Procrustes analysis (GPA) [@Bookstein1991].
-GPA is a method that superimposes sets of landmark configurations into a common coordinate system, where superimposed landmark coordinates are used as shape variables for further multivariate statistical analyses [@Slice2007].
-Landmark-based morphometrics are widely applied to archaeological objects with obvious morphological features that can provide unambiguous reference points for landmark placement, such as projectile point tips or biological measurements points indicating osteological features [@Birch2019; @Buchanan2019; @Lycett2013; @Cardillo2010; @Haruda2019; @Meloro2015].
+There are two common morphometric methods: landmark approaches and outline approaches (Adams et al., 2004).
+Landmark approaches assign a set of landmarks or semilandmarks onto objects as reference points that can be specified on a coordinate system.
+Landmark-based morphometrics are widely applied to archaeological objects with obvious morphological features that can provide unambiguous reference points for landmark placement, such as projectile point tips or visually distinctive osteological features (Birch and Martinón-Torres, 2019; Buchanan et al., 2019; Cardillo, 2010; Haruda et al., 2019; Lycett and Cramon-Taubadel, 2013; Meloro et al., 2015).
 
-However, one limitation of landmark approaches is that landmarks may not be able to capture the shape differences of a morphological structure where the curving outlines between landmarks contain crucial variation.
-Furthermore, landmark points may be unavailable for structures that are mostly or entirely curves.
+However, a key limitation of the landmark approach is that landmarks may be difficult to reproducibly locate for structures that are mostly or entirely curves.
 In those cases, outline approaches are more suitable for analyzing the overall shape of an object.
-One outline approach is the semi-landmarks method, also called sliding landmarks, which assigns points along the curve between two landmarks at defined intervals [@Bookstein1997; @Lawing2010].
-Those semi-landmarks are allowed to slide along the curve to remove the effect of the arbitrary landmark spacing by minimizing either Procrustes distance or bending energy [@Bookstein1997; @Slice2007; @Gunz2013].
-Another approach that is commonly applied to two-dimensional closed shapes is elliptic Fourier Analysis (EFA), that converts coordinates along a curve into Fourier function coefficients, called harmonic coefficients [@Kuhl1982].
-EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions [@Adams2004; @Claude2008; @Bonhomme2014].
+One outline approach is the semi-landmarks method, also called sliding landmarks, which assigns points along the curve between two landmarks at defined intervals (Bookstein, 1997; Lawing and Polly, 2010).
+Those semi-landmarks are allowed to slide along the curve to remove the effect of the arbitrary landmark spacing by minimizing either Procrustes distance or bending energy (Bookstein, 1997; Gunz and Mitteroecker, 2013; Slice, 2007).
+Another approach that is commonly applied to two-dimensional closed shapes is elliptic Fourier Analysis (EFA), that converts coordinates along a curve into Fourier function coefficients, called harmonic coefficients (Kuhl and Giardina, 1982).
+EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions (Adams et al., 2004; Bonhomme et al., 2014; Claude, 2008).
 The number of harmonics determines the quality and precision of the geometric representation of an object.
-The harmonic power, a cumulated sum of squared harmonic coefficients, provides a robust rule for determining the desired number of harmonics [@Bonhomme2014].
-
+The harmonic power, a cumulated sum of squared harmonic coefficients, provides a robust rule for determining the desired number of harmonics (Bonhomme et al., 2014).
 EFA is suitable for shapes lacking representative landmarks or where curves contain the most meaningful variation.
-Applications in archaeology include human remains and zooarchaeology to understand biological variation, and more recently on stone artifacts to explore stone tool technology [@Fox2015; @Hoggard2019; @Iovictua2010].
-Ceramics is another archaeological materials that outline approaches can provide new insights into shape variation to answer anthropological questions related to ceramic specialization.
-Using semi-landmarks method, @Topi2018 examine 89 photographs of globular jars from the Casas Grandes of northwest Mexico.
-Their results suggest that some ceramic types from the Medio period (AD 1200-1450), were made by specialists based on highly standardized shapes.
-Among those highly standardized jars, some were made by specialists attached to elites indicating the presence of social inequality, while others were made by independent specialists according to their spatial distribution patterns.
-Similarity, @Selden2019 examined 28 Caddo bottles excavated at several sites (AD 500-1700) in northwest Louisiana using semi-landmarks method.
-The results indicate a significant integration between the different parts of the bottles including rim, neck, body, and base that vary in a coordinated manner.
-Also, two discrete base and body shapes represented northern and southern differences were identified.
-For the EFA method, @Wilczek2014 evaluate the concordance between EFA and Discrete Cosine Transform (DCT), and a traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France.
-The results show the variation indicated by EFA and DCT matches the traditional ceramic typology, which supports that claim that outline-based approaches can be efficiently used for studying variations in ceramic shapes.
-Furthermore, @Wilczek2014's EFA results help us understand the level of production standardization over time across the region.
-Those examples show that outline approaches can distinguish variation in ceramic shapes at high resolutions to understand ceramic production.
-
+Applications in archaeology include human remains and zooarchaeology, stone artifacts, and ceramics (Fox, 2015; Hoggard et al., 2019; Ioviţă, 2010; Topi et al.
+2018; Selden Jr 2019; Wilczek et al.
+2014).
 Taking the ceramics data from Kiwulan, northeastern Taiwan, we use EFA to evaluate the level of standardization of ceramics in relation to the European presence in the 17th century to get insights into the emergence of ceramic specialization.
 We use a significance test for the equality of coefficient of variations of shape variables to statistically compare the vessel standardization from different periods.
-Using pottery shape as a proxy to study craft specialization, we address these questions: Did foreign influence have impacts on Indigenous pottery production that can be detected in the shape of the vessels?
-Did pottery shape become more homogeneous and standardized after foreign contacts with European colonizers or Chinese immigrants?
 

### 2020-04-10 analysis/paper/paper.Rmd
editing archaeological background

@@
 
-Ceramics analyzed in this paper come from 40 units (4m by 4m each) sampled from the central, undisturbed area of archaeological excavations at Kiwulan (Figure \@ref(fig:KWL-map); Figure \@ref(fig:KWL-sam-area)).
-Kiwulan is situated on a hill near a riverside at the northern margin of Yilan County, which is characterized by a triangular alluvial plain facing eastwards the Pacific with high mountains on three other sides.
-Yilan is an ideal context to study peripheral colonial influence because Indigenous communities there were isolated by natural barriers that limited the frequency of direct contact with the Spanish and the Dutch settlers in northern Taiwan.
-The chronology of the archaeological deposits at Kiwulan consists of two cultural components, the upper component and the lower component, with a sterile layer in between [@Chen2007].
+Ceramics analyzed in this paper come from 40 units (4m by 4m each) sampled from the central, undisturbed area of archaeological excavations at Kiwulan (Figure 1; Figure 2).
+The chronology of the archaeological deposits at Kiwulan consists of two cultural components, the upper component and the lower component, with a sterile layer in between (Chen, 2007).
 This paper focuses on the upper component, dated from AD 1350 to AD 1950, covering the late Iron Age and the historical period, which we define as the European occupation in Taiwan in the early 17th century.
-The Dutch first occupied southern Taiwan in 1624 and then the Spanish occupied northern Taiwan in 1626 [@Andrade2007].
+The Dutch first occupied southern Taiwan in 1624 and then the Spanish occupied northern Taiwan in 1626 (Andrade, 2007).
 In 1642, the Spanish was expelled by the Dutch in northern Taiwan, who took over their forts at Helping Dau in Keelung, and in Tamsui.
-Since then, western Taiwan was mostly under Dutch colonial rule until 1662 when the Kingdom of Tungning in Taiwan was founded by Koxinga, a loyalist of the Ming dynasty of China [@Andrade2007].
+Since then, western Taiwan was mostly under Dutch colonial rule until 1662 when the Kingdom of Tungning in Taiwan was founded by Koxinga, a loyalist of the Ming dynasty of China (Andrade, 2007).
 
@@
 The most abundant artifacts in the upper component are locally made ceramics, which are distributed throughout the sequence and across the site.
-More than 550,000 sherds were excavated, and around 1,200 vessels could be reconstructed.
+More than 550,000 sherds were excavated, and around 1,200 vessels could be partially reconstructed (i.e.
+complete rim, or complete base).
 There are only two forms of local vessels, a cooking pot and a steamer made of two cooking pots stacked together with a clay filter layer in between.
@@
 These vessels were probably used for cooking, with evidence of charred residues and carbon deposits frequently observed on vessel interiors and soot on their exteriors.
-Moreover, a lack of evidence of other utilitarian earthenware ceramics suggests these globular pots were mainly used for daily cooking.
 They are fired to orange to brownish color with a fully oxidized core or a reduced core with oxidized fringes.

### 2020-04-10 analysis/paper/paper.Rmd
editing start of methods section

@@
 
-A total of `r total_no` largely intact vessels from pre-contact contexts (n = `r pre_no`), post-European contexts (n = `r Euro_no`), and Chinese contact contexts (n = `r Chi_no`) were selected for elliptic Fourier analysis.
-Most of them cover every vessel component from rim to bottom.
-The shape of vessel is the major object analyzed in this paper, however, to fully explore the pottery standardization, we also take the metric measurements into account for comparison.
-The total of `r metric_all` reconstructed pots (pre-European = `r pre`, European = `r euro`, Chinese = `r chi`) were measured.
-The metric attributes include thickness of rim, neck, and body, and diameter of rim, neck, and body.
-For each pot, two measurements for thickness and diameter were taken to compute an average value.
-To obtain information about the pottery shape, we compare the ratio of Rim/Body thickness, and the ratio of Rim/Body diameter.
-In addition, we use Monte Carlo test for randomness in spatial locations of pots to explore whether their distribution is meaningful, such as clustered or dispersed pattern.
-As craft specialization increases, pottery distribution might shift from a dispersed pattern to concentrated clusters that reflects the location of production [@Costin2001].
+We used all `r total_no` vessels that could be reconstructed with rim, body and base parts and securely provenanced to pre-contact contexts (n = `r pre_no`), post-European contexts (n = `r Euro_no`), and Chinese contact contexts (n = `r Chi_no`) for elliptic Fourier analysis.
 

### 2020-04-10 analysis/paper/paper.Rmd
editing EFA methods

@@
 
-We used scans of pottery drawings acquired from the Bureau of Cultural Affairs in Yilan.
-All drawings present a two-dimensional view of vessel cross-sections with indications of metric measurements.
+drawings present a two-dimensional view of vessel cross-sections with indications of metric measurements.
 The scanned drawings were imported into Inkscape (http://inkscape.org) where outlines were traced manually.
 Where only half cross-section images were available, these were duplicated, flipped, and then joined with another one to create a 2D closed outline for each vessel.
-Geometric morphometric analyses were conducted using the R software [@Rlanguage2019] and the functions included in the Momocs, a R package intended to quantify and analyze shapes, especially for outline analysis [@Bonhomme2014].
-The digilized outlines were converted into a list of successive xy pixel coordinates for elliptic Fourier analysis (EFA), to investigate morphological differences among pottery shapes from our three occupation phases.
+Geometric morphometric analyses were conducted using the R software (R Core Team, 2019) and the functions included in the Momocs, a R package intended to quantify and analyze shapes (Bonhomme et al., 2014).
+The digilized outlines were converted into a list of successive xy pixel coordinates for elliptic Fourier analysis (EFA).
 The harmonic coefficients generated by EFA were analyzed by principal component analysis (PCA) for dimensionality reduction to illustrate the diversity of the shape data and identify major patterns of variation.

### 2020-04-10 analysis/paper/paper.Rmd
editing results: basic description of shape

@@
 The coefficient of variation is a common and widely-used statistical measure of the spread of a set of measurements of a sample.
-It is defined as the standard deviation divided by the mean in a ratio scale format:
+It is defined as the standard deviation divided by the mean:
 
@@
 
-By standardizing standard deviation in each data set, the coefficients of variation (CV) values allows us to directly compare variation in samples measured with different units or means.
-This is useful to compare the degree of standardization for archaeological assembles since CV scales standard deviation to the mean that enables comparison of variation across different sample sizes [@Eerkens2001, pp.498].
-We take this as our measurement of standardization in vessel shape variables: lower CV values reflect higher standardization, and thus increased craft specialization in the community.
-Given that CV is robust for positive values due to the representation by ratio, we normalized PC scores to a range between 0 and 1 before computation of CV.
+As a standardized measure of the spread of data, coefficients of variation (CV) allow us to directly compare variation in samples measured with different units or means.
+This is useful to compare the degree of standardization for archaeological assembles and enables comparison of variation across different sample sizes (Eerkens and Bettinger, 2001, p.
+498).
+Following [citations of previous work that has used CVs like this] we take this as our measurement of standardization in vessel shape variables: lower CV values reflect higher standardization, and thus increased craft specialization in the community.
+Given that CV is robust for positive values due to the representation by ratio, we normalized PC scores to a range between 0 and 1 before computation of CV
 
@@
 While previous work has used the @Feltz1996 asymptotic test for the equality of coefficients of variation from k populations [@Lycett2008; @Okumura2014; @Eerkens2000; @Eerkens2001; @Hoggard2017], we prefer the MSLR test for shape variables as a more recent development with lower rates of type I error, better performance with uneven sample numbers, and more power across a range of conditions [@Krishnamoorthy2014].
+
+To complement our investigation of craft specialization through shape standardization, we investigated spatial patterns of ceramic vessels at Kiwulan.
+As craft specialization increases, pottery distribution we expect a shift from a pattern of vessels dispersed across the site to a pattern of clusters that reflects the loci of production (Costin, 2001).
+We used a Monte Carlo test for randomness in spatial locations of pots to explore whether their distribution is significantly clustered or dispersed.
 
@@
 # Results
-
-To investigate the effect of the historically-documented decline of Indigenous population in the Chinese period [@Chen2007; @Hsieh2009] on pottery production we examined the spatial distribution of pots in our sample in each period.
-Our concern is that the smaller population size during the Chinese period could confound our investigations into craft specialization because of reduced numbers of potters and pottery production shrinking a few locations in the settlement during this time.
-Although the overall number of vessels is smaller during the Chinese contact period, Figure \@ref(fig:pottery-spatial-pattern) shows that pottery is distributed randomly across the sampling area without any distinctive clusters.
-This suggests that differences in population across our three occupation phases are probably not driving variation in craft specialization.
 
@@
 
-The elliptic Fourier coefficients of `r total_no` pottery from three phases were calculated.
-Reliable pottery outlines were captured by 13 harmonics that gather 99% of the total harmonic power.
-Figure \@ref(fig:thin-plate-splines) shows the shape changes described using thin-plate spline warping for paired periods, pre- and post-European periods, and post-European and Chinese periods.
-The average shape of vessels from each phase was visualized and it shows more obvious differences between post-European and Chinese periods.
+We found that 13 harmonics captured 99% of the total harmonic power in our elliptic Fourier coefficients of `r total_no` vessels from three phases.
+Figure \@ref(fig:thin-plate-splines) shows vessel shape changes described using thin-plate spline warping for paired periods, pre- and post-European periods, and post-European and Chinese periods, with the most greatest differences between pre-European and Chinese periods.
 
-```{r thin-plate-splines, fig.cap= "Compare the average shapes across phases using thin plate splines (TPS) with outline deformations required to pass from an extreme of the morphospace to the other."}
+```{r thin-plate-splines, fig.cap= "Comparison of average vessel shapes across phases using thin plate splines (TPS) with outline deformations required to pass from an extreme of one morphospace to another."}
 # Thin plate spline analysis (TPS)

### 2020-04-10 analysis/paper/paper.Rmd
editing results: pca

@@
 
-The elliptic Fourier coefficients were examined by Principal Components Analysis to explore the shape variations across three phases.
-The first two principal components (PCs) explain `r vari_two`% of the total variances, of which `r pc1`% is explained by the first principle component.
+The first two principal components (PCs) of the PCA on the elliptic Fourier coefficients explain `r vari_two`% of the total variance, of which `r pc1`% is explained by the first principle component.
 With the third component, the first three principal components explain `r vari_three`% of the total variance.
-According to Figure \@ref(fig:first-two-PCs-byplot), PC1 captures the height of the vessels, from tall to short, and the roundness of the body from round to oval-shaped.
+PC1 captures the height of the vessels, from tall to short, and the roundness of the body from round to oval-shaped (Figure \@ref(fig:first-two-PCs-byplot)).
 PC2 relates to the neck and mouth constriction, from narrow to wide.
-PC3 explains a smaller portion of the variance (`r pc3`%), which relates to the degree of the flare of the neck, from curved to straight shape.
-The first two components account for most of the variance in relation to three phases.
+PC3 explains a smaller portion of the variance (`r pc3`%), which relates to the degree of the flare of the neck, from a curved to straight shape.
 The results reflect a large overlap in shapes from three occupations phases, especially for shapes in the pre-European and post-European periods.

### 2020-04-10 analysis/paper/paper.Rmd
editing discussion: remove metrics

@@
 To compare pottery shape standardization across the three phases we investigate the distributions of the first three PC scores, taking the PC scores as proxy variables for vessel shape (Figure \@ref(fig:PCs-distribution)).
-The first PC, vessel height, shows higher variation in the pre-European period compared to the Chinese period.
+The first PC, capturing vessel height and roundness, shows higher variation in the pre-European period compared to the Chinese period.
 That indicates that shape standardization was higher in the Chinese period compared to the pre-European period.
 The second PC also presents a similar pattern of higher shape standardization in the Chinese period compared to the other two phases.
-To see whether the differences in the distribution of PCs between any two phases are substantive or due to chance, we used a modified signed-likelihood ratio test to assess the equality of CVs.
+
+To see whether the differences in the distribution of PCs between any two phases are substantive or due to chance, we used a modified signed-likelihood ratio test to assess the equality of CVs [citation needed: article and pkg].
 P-values for this significance test of CVs for PC1 and PC2 show significant differences in the standardization of vessel shapes across periods, especially between Chinese contact with either pre-European or post-European (Figure \@ref(fig:CV-test-PCs-plot), Table \@ref(tab:CV-test-PCs-table)).
-A significant difference is also detected between the pre-European and post-European periods.
+A significant difference was also detected between the pre-European and post-European periods.
 
@@
 
-Previous investigations at Kiwulan have suggested an unequal distribution of prestige goods, trade ornaments specifically, following the appearance of Europeans [@Cheng2008; @Wang2011], hinting at the emergence of social inequality in the Indigenous community.
-Here we have examined ceramic vessel shape standardization to measure craft specialization as a proxy for social differentiation, such as elite control [@Costin2001; @Junker1999].
-Our results indicate that differences in pottery shape can be detected using EFA.
+Previous investigations at Kiwulan have suggested an unequal distribution of prestige goods, trade ornaments specifically, following the appearance of Europeans (Cheng, 2008; Wang, 2011), hinting at an emergence of social inequality in the Indigenous community.
+We have examined ceramic vessel shape standardization to measure craft specialization as a proxy for social differentiation (Costin, 2001; Junker, 1999).
 The result of our MANOVA test shows there is a significant difference in shapes between the pre-European and Post-European periods, and between the pre-European and the Chinese periods.
@@
 Our CV tests indicate that there are significant differences in shape standardization between the pre-European period and post-European period, and between the Chinese contact and either pre-European or post-European periods.
-This suggests pottery shape became more homogeneous and standardized after foreign contacts with European colonizers and even more so after contact with Chinese immigrants.
+Pottery shape became more homogeneous and standardized after contact with Europeans and even more so after contact with the Chinese.
 Compositional analysis shows that the clay pastes are similar, regardless of the increasing standardization of the pottery shape, reflecting continuity in the raw material sources.
 We can thus rule out changes in clay fabric as a factor in explaining changes in vessel shape.
-To determine if shape changes might be related to changes in the function of pots at Kiwulan, we used geochemical methods to extract and identify lipids trapped in the fabric of potsherds to identify foods that may have contributed residues absorbed into the clay [cf.
-@Kwak2015].
+To determine if shape changes might be related to changes in the function of pots at Kiwulan, we used geochemical methods to extract and identify lipids trapped in the fabric of potsherds to identify foods that may have contributed residues absorbed into the clay (cf.
+Kwak and Marwick, 2015).
 Unfortunately, we did not obtain useful results due to extremely low lipid yields, which is probably due to the very thin, dense, and low porosity fabric of Kiwulan pottery.
-These physical characteristics of the clay offer limited spaces to trap and protect organic molecules from microbiological degradation [cf.
-@Evershed2008, pp.
-909].
+These physical characteristics of the clay offer limited spaces to trap and protect organic molecules from microbiological degradation (cf.
+Evershed, 2008, p.
+909).
 
-We compare the result of shape variable with the result of metric data for six measurements and two ratio values using CV test (Table \@ref(tab:metricCVstest)).
-Metric measurements show there is a significant difference (p-value less than 0.01) between CV values for two metric measurements across phases, the rim diameter and the neck diameter before and after the European presence.
-After the arrival of the Chinese, there is only one metric attribute, the rim diameter, showing significant difference compared with post-European period (p-value less than 0.05).
-Although these results show the differences between different contact phases that corresponds to part of the results of shape variable, elliptic Fourier analysis captures more subtle differences in a substantial way reflected by overall shapes of pottery, not only limited to diameter variables.
-In addition, the result of spatial analysis (Figure \@ref(fig:kernel-plot)) for pottery shape samples presents multiple clusters with high density of pottery during the European presence.
-However, the hypothesis testing on spatial randomness indicates a non-randomly dispersed distribution before European contact and more extreme dispersed after European presence.
-In contrast, the distribution of pottery is more similar to the random distributions during the Chinese period.
+The spatial analysis (Figure \@ref(fig:kernel-plot)) for pottery samples presents multiple clusters with high densities of pottery during European presence.
+Hypothesis testing on spatial randomness indicates a non-randomly dispersed distribution before European contact and more extreme dispersed distribution after European presence.
+In contrast, the distribution of pottery is more similar to random distributions during the Chinese period.
 This contradicts our expectation that clustered pattern will be observed with an increase in pottery standardization since the emergence of specialized groups.
-Or, this may indicate that there are other factors lead to pottery shape standardization instead of a small number of artisans.
-
-```{r metricCVstest}
-
-# select metric attributes and tidy data for computing CV
-kwl_p_metrics_long <-
-  kwl_p_ave %>%
-  mutate(dia_rim_body = Dia_Rim_ave/Dia_Body_ave,
-         thi_rim_body = Thick_Rim_ave/Thick_Body_ave) %>%
-  select(period, 
-         Dia_Rim_ave, Dia_Neck_ave, 
-         Dia_Body_ave, Thick_Rim_ave, 
-         Thick_Neck_ave, Thick_Body_ave,
-         dia_rim_body, thi_rim_body,) %>%
-  rename(`Rim diameter (mm)` = Dia_Rim_ave,
-         `Neck diameter (mm)` = Dia_Neck_ave,
-         `Body diameter (mm)` = Dia_Body_ave,
-         `Rim thickness (mm)` = Thick_Rim_ave,
-         `Neck thickness (mm)` = Thick_Neck_ave,
-         `Body thickness (mm)` = Thick_Body_ave,
-         `Ratio of Rim/Body diameter` = dia_rim_body,
-         `Ratio of Rim/Body thickness` = thi_rim_body) %>%
-  gather(variable, value, -period) %>%
-  filter(!is.na(period), !is.na(value)) %>% 
-  mutate(variable = factor(variable, levels = c("Rim thickness (mm)",
-                                                "Neck thickness (mm)",
-                                                "Body thickness (mm)",
-                                                "Ratio of Rim/Body thickness",
-                                                "Rim diameter (mm)",
-                                                "Neck diameter (mm)",
-                                                "Body diameter (mm)",
-                                                "Ratio of Rim/Body diameter"))) %>%
-  arrange(variable)
-
-# filter pre-e and post-e periods
-kwl_p_metrics_long_pre_post <-
-  kwl_p_metrics_long %>%
-  filter(!period == "ch-con")
-# filter post-e and chi periods
-kwl_p_metrics_long_post_chi <-
-  kwl_p_metrics_long %>%
-  filter(!period == "pre-e")
-
-#---------------------------CV test for pre-e and post-e pair
-kiwulan_pottery_metrics_cvs_plot_cv_test_asym_pre_post <-
-  kwl_p_metrics_long_pre_post %>%
-  nest(-variable) %>%  # or nest(period, value) %>%
-  mutate(asymptotic_test = map(data, # change from mslr_test to asymptotic_test
-                               ~bind_cols(asymptotic_test (
-                                 .x$value,
-                                 .x$period)))) %>%
-  unnest(asymptotic_test)
-# compute CV
-kwl_p_cvs_pre_post <-
-  kwl_p_metrics_long_pre_post %>%
-  group_by(period, variable) %>%
-  summarise(cvs = raster::cv(value, na.rm = TRUE))
-# join p-values to CVs
-kwl_p_cvs_and_pvalues_pre_post <-
-  kwl_p_cvs_pre_post %>%
-  left_join(kiwulan_pottery_metrics_cvs_plot_cv_test_asym_pre_post) %>%
-  mutate(significant = ifelse(p_value <= 0.05, 'yes', 'no')) %>%
-  mutate(variable = factor(variable, ordered = TRUE)) %>%
-  arrange(variable)
-# making table for CV
-kwl_p_cvs_table_pre_post <-
-  kwl_p_cvs_and_pvalues_pre_post %>%
-  select(-data, -significant) %>%
-  pivot_wider(names_from = period, values_from = cvs) %>%
-  select(variable, `pre-e`, `post-e`, `D_AD`, `p_value`) %>%
-  mutate_at(c("pre-e", "post-e", "D_AD", "p_value"), round, 4)
-
-#---------------------------CV test for post-e and ch-con pair
-kiwulan_pottery_metrics_cvs_plot_cv_test_asym_post_chi <-
-  kwl_p_metrics_long_post_chi %>%
-  nest(-variable) %>%  # or nest(period, value) %>%
-  mutate(asymptotic_test = map(data, # change from mslr_test to asymptotic_test
-                               ~bind_cols(asymptotic_test (
-                                 .x$value,
-                                 .x$period)))) %>%
-  unnest(asymptotic_test)
-# compute CV
-kwl_p_cvs_pre_post <-
-  kwl_p_metrics_long_post_chi %>%
-  group_by(period, variable) %>%
-  summarise(cvs = raster::cv(value, na.rm = TRUE))
-# join p-values to CVs
-kwl_p_cvs_and_pvalues_post_chi <-
-  kwl_p_cvs_pre_post %>%
-  left_join(kiwulan_pottery_metrics_cvs_plot_cv_test_asym_post_chi) %>%
-  mutate(significant = ifelse(p_value <= 0.05, 'yes', 'no')) %>%
-  mutate(variable = factor(variable, ordered = TRUE)) %>%
-  arrange(variable)
-# making table for CV
-kwl_p_cvs_table_post_chi <-
-  kwl_p_cvs_and_pvalues_post_chi %>%
-  select(-data, -significant) %>%
-  pivot_wider(names_from = period, values_from = cvs) %>%
-  select(variable, `post-e`, `ch-con`, `D_AD`, `p_value`) %>%
-  mutate_at(c("post-e", "ch-con", "D_AD", "p_value"), round, 4)
-
-# join two table together
-kwl_p_cvs_table_all <-
-  kwl_p_cvs_table_pre_post %>% 
-  left_join(kwl_p_cvs_table_post_chi, "variable") %>% 
-  select(-`post-e.y`) %>% 
-  rename(`before European` = "pre-e", `European` = "post-e.x", Chinese = "ch-con",
-         `D_AD-1` = "D_AD.x", `D_AD-2` = "D_AD.y",
-         `P value-1` = `p_value.x`, `P value-2` = `p_value.y`) %>% 
-  select(variable, `before European`, European, Chinese,
-         `D_AD-1`, `P value-1`, `D_AD-2`, `P value-2`)
-
-knitr::kable(kwl_p_cvs_table_all,
-             caption = "Coefficient of Variation for metric attributes by three phases using Asymptotic test [@Feltz1996].
-The first set of P-value represents the comparison between before European and European periods, while the second represents the comparison between European and Chinese periods ")
-```
+The absence of clusters in the Chinese period is notable because this was a time of a historically-documented decline of Indigenous population (Chen, 2007; Hsieh, 2009).
+We might expect reduced numbers of potters to result in pottery production shrinking to a few locations in the settlement during this time.
+Although the overall number of vessels is smaller during the Chinese contact period, Figure 7) shows that pottery is distributed randomly across the sampling area without any distinctive clusters during this time.
+This suggests that differences in population across our three occupation phases are probably not driving variation in craft specialization.
 

### 2020-04-10 analysis/paper/paper.Rmd
editing conclusion

@@
 Our results offer tentative support for the hypothesis that foreign influences at Kiwulan influenced emergence of social inequality in the local Indigenous society.
-If increased shape standardization is a reliable indicator of craft specialization, then we may be seeing evidence of a shift from corporate to network organization [@Feinman2000].
-However, strong claims for an emergence of social complexity resulting from foreign contact at Kiwulan will need support from multiple and diverse sources of evidence.
-This should include the other two main datasets at Kiwulan, the ornaments assemblages, and the burial goods.
-We find vessel shapes were more standardized in the Chinese period than the European period, contrary to previous work that has downplayed the effect of Chinese settlement in Yilan [@Wang2011].
-Compared to other regions in Taiwan, European colonial influence was weak in Yilan due to the isolated of the surrounding mountains and the economic focus of the Spanish and  Dutch who preferred northern Taiwan as their trading base.
-Indigenous communities Yilan experienced indirect influence from European trade networks and their colonial activities in a pericolonial context [cf.
-@Acabado2017].
-In contrast to the Indigenous-European interactions at Kiwulan, the interaction between Indigenous people and Chinese in the 19th century appears to have been more intense and direct.
-Historical records indicate that Chinese immigrants settled in Yilan and lived closely with Kiwulan Indigenous societies [@Chen1963; @Ke1993].
-This direct influence is reflected by the archaeological evidence of large amounts of Chinese porcelains and distinctive architectural bricks and tiles used by Chinese [@Hsieh2009].
-Similarly, burials at Kiwulan in this later phase show the adoption of coffins in mortuary practices, which @Chen2007 interprets as the adoption of a symbol of ethnic Chinese.
+If increased shape standardization is a reliable indicator of craft specialization, then we may be seeing evidence of a shift from corporate (group-based, distributed, collective, cooperative) to network (individual-based, competitive) organization (Feinman, 2000).
+However, strong claims for an emergence of social complexity resulting from foreign contact at Kiwulan will need support from multiple and diverse sources of evidence that are beyond the scope of this paper.
+We find vessel shapes were more standardized in the Chinese period than the European period, contrary to previous work that has downplayed the effect of Chinese settlement in Yilan (Wang, 2011).
+
+Compared to other regions in Taiwan, European colonial influence was weak in Yilan due to isolation by the surrounding mountains, and the economic focus of the Spanish and Dutch who preferred northern and northwest Taiwan as their trading base.
+Indigenous communities in Yilan experienced indirect influence from European trade networks and their colonial activities in a pericolonial context (cf.
+Acabado, 2017).
+In contrast to the Indigenous-European interactions at Kiwulan, the interaction between Indigenous people and Chinese immigrants in the 19th century appears to have been more intense and direct.
+Historical records indicate that Chinese groups settled in Yilan and lived closely with Kiwulan Indigenous societies (Chen, 1963; Ke, 1993).
+This direct influence is reflected by the archaeological evidence of large amounts of Chinese porcelains and distinctive architectural bricks and tiles used by Chinese (Hsieh, 2009).
+Similarly, burials at Kiwulan in this later phase show the adoption of coffins in mortuary practices, which Chen (2007) interprets as the adoption of a symbol of ethnic Chinese.
 
@@
 Here, EFA is combined with a significance test for the equality of CVs of shape variables to provide a robust way to identify and statistically assess differences in shape standardization.
-The direct relationship between foreign influences and standardization of ceramic shape was tested on pottery from Kiwulan, and large Iron Age Indigenous settlement in northeastern Taiwan.
+The direct relationship between foreign influences and standardization of ceramic shape was tested on pottery from Kiwulan, a large Iron Age Indigenous settlement in northeastern Taiwan.
 Lower variation in ceramic shape was identified after European presence began, and even lower variation during the period of Chinese presence.
-Our findings help to understand the factors that may lead to standardization of pottery production in a cross-cultural interaction context.
+Our findings help to understand the factors that may lead to standardization of pottery production in a pericolonial interaction context.
 More homogeneous shapes during the contact periods, without any changes in clay paste composition or production technique, suggests that shape standardization was intentional.
-The results here further suggest that expressions of social identity or cultural boundaries in Indigenous societies through highly visible vessel qualities, such as shape, may be heightened during periods of foreign intrusion in pericolonial contexts.
+The results here further suggest that expressions of social identity or cultural boundaries in Indigenous societies through highly visible vessel qualities, such as shape, may be heightened during periods of foreign contact in pericolonial contexts.
 Our analysis, with its openly available methods and data, is readily extensible to other pottery assemblages in the region to further explore related questions about craft specialization and standardization in Iron Age ceramic technologies.

### 2020-04-10 analysis/paper/paper.Rmd
editing methods efa

@@
 
-drawings present a two-dimensional view of vessel cross-sections with indications of metric measurements.
+We used scans of pottery drawings acquired from the Bureau of Cultural Affairs in Yilan.
+All drawings present a two-dimensional view of vessel cross-sections with indications of metric measurements.
 The scanned drawings were imported into Inkscape (http://inkscape.org) where outlines were traced manually.

### 2020-04-10 analysis/paper/paper.Rmd
edit title and intro section

@@
 ---
-title: "Investigating shape standardisation using geometric morphometry of pottery from Iron Age northeastern Taiwan"
+title: "Investigating shape standardization using geometric morphometry of pottery from Iron Age northeastern Taiwan"
 author:
@@
 A major historical factor that is often linked to social change in small-scale societies is colonial settlers introducing foreign trade goods to local Indigenous societies.
-The monopolization of long-distance trade goods has caused substantial transformations of Indigenous economic, cultural, and socio-political systems (Dietler, 2005, 1997; Junker, 1993; Silliman, 2005).
-Pericolonial archaeology is the study of these indirect effects of colonialism, investigating areas where direct European colonial rule was limited, their conquests often short-lived and unsuccessful, but their commercial activities had economic and political impacts on Indigenous peoples living on the periphery of colonial control (Acabado, 2017; Trabert, 2017).
+The monopolization of long-distance trade goods has caused substantial transformations of Indigenous economic, cultural, and socio-political systems [@Dietler2005; @Dietler1997; @Junker1993; @Silliman2005].
+Pericolonial archaeology is the study of these indirect effects of colonialism, investigating areas where direct European colonial rule was limited, their conquests often short-lived and unsuccessful, but their commercial activities had economic and political impacts on Indigenous peoples living on the periphery of colonial control [@Acabado2017; @Trabert2017].
 Pericolonial situations were common during the seventeenth to nineteenth centuries in East and Southeast Asia where European trading activity was extensive, but direct European rule less widespread.
 An emerging priority in archaeological research in this region is identifying the indirect influences are apparent on Indigenous communities during this time.
-For example, Acabado (2017)’s study of Ifugao society in the Philippines highland suggests economic and political intensification during the Spanish presence as the response of Indigenous peoples to the Spanish cooptation.
+For example, @Acabado2017’s study of Ifugao society in the Philippines highland suggests economic and political intensification during the Spanish presence as the response of Indigenous peoples to the Spanish cooptation.
 
 Indigenous societies’ responses to colonial contact range from passive acceptance through to active negotiation with the colonists, and accommodation or resistance of foreign intrusion.
-These responses can be identified through their daily cultural practices, such as their consumption patterns of foreign goods (Dietler, 2015; Given, 2004; Mullins, 2011; Scaramelli and Scaramelli, 2005; Silliman, 2001; Torrence and Clarke, 2000).
-In this paper we investigate the archaeology of a pericolonial situation at Kiwulan (1350-1950 AD), a large multi-component archaeological site in Yilan Province, northeastern Taiwan (Chen, 2007), to identify the indirect impacts of colonial settler activity on local Indigenous societies.
+These responses can be identified through their daily cultural practices, such as their consumption patterns of foreign goods [@Dietler2015; @Given2004; @Mullins2011; @Scaramelli2005; @Silliman2001; @TorrenceandClarke2000].
+In this paper we investigate the archaeology of a pericolonial situation at Kiwulan (1350-1850 AD), a large multi-component archaeological site in Yilan County, northeastern Taiwan [@Chen2007], to identify the indirect impacts of colonial settler activity on local Indigenous societies.
 Yilan is an ideal context to study peripheral colonial influences because Indigenous communities there were isolated by geographical barriers that limited the frequency of direct contact with the Spanish and the Dutch settlers in northern Taiwan.
@@
 
-Several measurements have been used for investigating ceramic standardization, such as metric, compositional, and technological variables (Arnold, 2000; Blackman et al., 1993; Boness et al., 2015; Costin, 1991; Rice, 1991; Roux, 2015; Tite, 1999).
+Several measurements have been used for investigating ceramic standardization, such as metric, compositional, and technological variables [@Arnold2000; @Blackman1993; @Boness2015; @Costin1991; @Rice1991; @Roux2015; @Tite1999].
 Among those variables, metric measurements are most widely applied to archaeological assemblages because of the ease of collecting these data.
-The coefficient of variation statistic is commonly used to quantify the degree of standardization in ceramic assemblages (Eerkens and Bettinger, 2001; Roux, 2003; Stark, 1995b).
+The coefficient of variation statistic is commonly used to quantify the degree of standardization in ceramic assemblages [@Eerkens2001; @Roux2003; @Stark1995].
 However, because pottery vessels typically have curved shapes, linear measurements have limited sensitivity to many kinds of shape variations.

### 2020-04-10 analysis/paper/paper.Rmd
edit GM section

@@
 
-Geometric morphometrics differs from traditional linear measurements through its use of Cartesian coordinates of morphological structures to define shapes (Adams et al., 2004; Bookstein, 1997; Lawing and Polly, 2010; Slice, 2007).
+Geometric morphometrics differs from traditional linear measurements through its use of Cartesian coordinates of morphological structures to define shapes [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007].
 Landmarks, curves or outlines of objects can be represented by coordinates in terms of their unique point locations with respect to numerical values on coordinate axes.
-There are two common morphometric methods: landmark approaches and outline approaches (Adams et al., 2004).
+There are two common morphometric methods: landmark approaches and outline approaches [@Adams2004].
 Landmark approaches assign a set of landmarks or semilandmarks onto objects as reference points that can be specified on a coordinate system.
-Landmark-based morphometrics are widely applied to archaeological objects with obvious morphological features that can provide unambiguous reference points for landmark placement, such as projectile point tips or visually distinctive osteological features (Birch and Martinón-Torres, 2019; Buchanan et al., 2019; Cardillo, 2010; Haruda et al., 2019; Lycett and Cramon-Taubadel, 2013; Meloro et al., 2015).
+Landmark-based morphometrics are widely applied to archaeological objects with obvious morphological features that can provide unambiguous reference points for landmark placement, such as projectile point tips or visually distinctive osteological features [@Birch2019; @Buchanan2019; @Cardillo2010; @Haruda2019; @Lycett2013; @Meloro2015].
 
@@
 In those cases, outline approaches are more suitable for analyzing the overall shape of an object.
-One outline approach is the semi-landmarks method, also called sliding landmarks, which assigns points along the curve between two landmarks at defined intervals (Bookstein, 1997; Lawing and Polly, 2010).
-Those semi-landmarks are allowed to slide along the curve to remove the effect of the arbitrary landmark spacing by minimizing either Procrustes distance or bending energy (Bookstein, 1997; Gunz and Mitteroecker, 2013; Slice, 2007).
-Another approach that is commonly applied to two-dimensional closed shapes is elliptic Fourier Analysis (EFA), that converts coordinates along a curve into Fourier function coefficients, called harmonic coefficients (Kuhl and Giardina, 1982).
-EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions (Adams et al., 2004; Bonhomme et al., 2014; Claude, 2008).
+One outline approach is the semi-landmarks method, also called sliding landmarks, which assigns points along the curve between two landmarks at defined intervals [@Bookstein1997; @Lawing2010].
+Those semi-landmarks are allowed to slide along the curve to remove the effect of the arbitrary landmark spacing by minimizing either Procrustes distance or bending energy [@Bookstein1997; @Gunz2013; @Slice2007].
+Another approach that is commonly applied to two-dimensional closed shapes is elliptic Fourier Analysis (EFA), that converts coordinates along a curve into Fourier function coefficients, called harmonic coefficients [@Kuhl1982].
+EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions [@Adams2004; @Bonhomme2014; @Claude2008].
 The number of harmonics determines the quality and precision of the geometric representation of an object.
-The harmonic power, a cumulated sum of squared harmonic coefficients, provides a robust rule for determining the desired number of harmonics (Bonhomme et al., 2014).
+The harmonic power, a cumulative sum of squared harmonic coefficients, provides a robust rule for determining the desired number of harmonics [@Bonhomme2014].
 EFA is suitable for shapes lacking representative landmarks or where curves contain the most meaningful variation.
-Applications in archaeology include human remains and zooarchaeology, stone artifacts, and ceramics (Fox, 2015; Hoggard et al., 2019; Ioviţă, 2010; Topi et al.
-2018; Selden Jr 2019; Wilczek et al.
-2014).
+Applications in archaeology include human remains and zooarchaeology, stone artifacts, and ceramics [@Fox2015; @Hoggard2019; @Iovictua2010; @Topi2018; @Selden2019; @Wilczek2014].
 Taking the ceramics data from Kiwulan, northeastern Taiwan, we use EFA to evaluate the level of standardization of ceramics in relation to the European presence in the 17th century to get insights into the emergence of ceramic specialization.

### 2020-04-10 analysis/paper/paper.Rmd
edit Archaeological background and materials

@@
 
-Ceramics analyzed in this paper come from 40 units (4m by 4m each) sampled from the central, undisturbed area of archaeological excavations at Kiwulan (Figure 1; Figure 2).
-The chronology of the archaeological deposits at Kiwulan consists of two cultural components, the upper component and the lower component, with a sterile layer in between (Chen, 2007).
-This paper focuses on the upper component, dated from AD 1350 to AD 1950, covering the late Iron Age and the historical period, which we define as the European occupation in Taiwan in the early 17th century.
-The Dutch first occupied southern Taiwan in 1624 and then the Spanish occupied northern Taiwan in 1626 (Andrade, 2007).
+Ceramics analyzed in this paper come from 40 units (4m by 4m each) sampled from the central, undisturbed area of archaeological excavations at Kiwulan (Figure \@ref(fig:KWL-map); Figure \@ref(fig:KWL-sam-area)).
+The chronology of the archaeological deposits at Kiwulan consists of two cultural components, the upper component and the lower component, with a sterile layer in between [@Chen2007].
+This paper focuses on the upper component, dated from AD 1350 to AD 1850, covering the late Iron Age and the historical period, which we define as the European occupation in Taiwan in the early 17th century.
+The Dutch first occupied southern Taiwan in 1624 and then the Spanish occupied northern Taiwan in 1626 [@Andrade2007].
 In 1642, the Spanish was expelled by the Dutch in northern Taiwan, who took over their forts at Helping Dau in Keelung, and in Tamsui.
-Since then, western Taiwan was mostly under Dutch colonial rule until 1662 when the Kingdom of Tungning in Taiwan was founded by Koxinga, a loyalist of the Ming dynasty of China (Andrade, 2007).
+Since then, western Taiwan was mostly under Dutch colonial rule until 1662 when the Kingdom of Tungning in Taiwan was founded by Koxinga, a loyalist of the Ming dynasty of China [@Andrade2007].
 

### 2020-04-10 analysis/paper/paper.Rmd
edit method section

@@
 Among those variables, metric measurements are most widely applied to archaeological assemblages because of the ease of collecting these data.
-The coefficient of variation statistic is commonly used to quantify the degree of standardization in ceramic assemblages [@Eerkens2001; @Roux2003; @Stark1995].
+The coefficient of variation statistic is commonly used to quantify the degree of standardization in ceramic assemblages [@Eerkens2001; @Roux2003; @Stark1995; @Junker1999].
 However, because pottery vessels typically have curved shapes, linear measurements have limited sensitivity to many kinds of shape variations.
@@
 
-## Digitising and analysing by EFA
+## Digitizing and analyzing by EFA
 
@@
 Where only half cross-section images were available, these were duplicated, flipped, and then joined with another one to create a 2D closed outline for each vessel.
-Geometric morphometric analyses were conducted using the R software (R Core Team, 2019) and the functions included in the Momocs, a R package intended to quantify and analyze shapes (Bonhomme et al., 2014).
+Geometric morphometric analyses were conducted using the R software (R Core Team, 2019) and the functions included in the Momocs, a R package intended to quantify and analyze shapes ([@Bonhomme2014].
 The digilized outlines were converted into a list of successive xy pixel coordinates for elliptic Fourier analysis (EFA).
@@
 
-As a standardized measure of the spread of data, coefficients of variation (CV) allow us to directly compare variation in samples measured with different units or means.
-This is useful to compare the degree of standardization for archaeological assembles and enables comparison of variation across different sample sizes (Eerkens and Bettinger, 2001, p.
-498).
-Following [citations of previous work that has used CVs like this] we take this as our measurement of standardization in vessel shape variables: lower CV values reflect higher standardization, and thus increased craft specialization in the community.
-Given that CV is robust for positive values due to the representation by ratio, we normalized PC scores to a range between 0 and 1 before computation of CV
+As a standardized measure of the spread of data, coefficients of variation (CV) allows us to directly compare variation in samples measured with different units or means.
+This is useful to compare the degree of standardization for archaeological assembles and enables comparison of variation across different sample sizes [@Eerkens2001, pp.498].
+Following @Eerkens2001 and @Roux2003, we take this as our measurement of standardization in vessel shape variables: lower CV values reflect higher standardization, and thus increased craft specialization in the community.
+Given that CV is robust for positive values due to the representation by ratio, we normalized PC scores to a range between 0 and 1 before computation of CV.
 
 To answer the question of whether CV values for vessel samples across our three occupational phases are significantly different or not, we used the Modified signed-likelihood ratio (MSLR) test for equality of CVs [@Krishnamoorthy2014].
-While previous work has used the @Feltz1996 asymptotic test for the equality of coefficients of variation from k populations [@Lycett2008; @Okumura2014; @Eerkens2000; @Eerkens2001; @Hoggard2017], we prefer the MSLR test for shape variables as a more recent development with lower rates of type I error, better performance with uneven sample numbers, and more power across a range of conditions [@Krishnamoorthy2014].
+While previous work has used the @Feltz1996's asymptotic test for the equality of coefficients of variation from k populations [@Lycett2008; @Okumura2014; @Eerkens2000; @Eerkens2001; @Hoggard2017], we prefer the MSLR test for shape variables as a more recent development with lower rates of type I error, better performance with uneven sample numbers, and more power across a range of conditions [@Krishnamoorthy2014].
 
 To complement our investigation of craft specialization through shape standardization, we investigated spatial patterns of ceramic vessels at Kiwulan.
-As craft specialization increases, pottery distribution we expect a shift from a pattern of vessels dispersed across the site to a pattern of clusters that reflects the loci of production (Costin, 2001).
+As craft specialization increases, pottery distribution we expect a shift from a pattern of vessels dispersed across the site to a pattern of clusters that reflects the loci of production [@Costin2001].
 We used a Monte Carlo test for randomness in spatial locations of pots to explore whether their distribution is significantly clustered or dispersed.

### 2020-04-10 analysis/paper/paper.Rmd
edit results section

@@
 
-To see whether the differences in the distribution of PCs between any two phases are substantive or due to chance, we used a modified signed-likelihood ratio test to assess the equality of CVs [citation needed: article and pkg].
+To see whether the differences in the distribution of PCs between any two phases are substantive or due to chance, we used a modified signed-likelihood ratio test to assess the equality of CVs [@Krishnamoorthy2014; @Marwick2019].
 P-values for this significance test of CVs for PC1 and PC2 show significant differences in the standardization of vessel shapes across periods, especially between Chinese contact with either pre-European or post-European (Figure \@ref(fig:CV-test-PCs-plot), Table \@ref(tab:CV-test-PCs-table)).

### 2020-04-10 analysis/paper/paper.Rmd
edit first half of discussion section

@@
 
-Previous investigations at Kiwulan have suggested an unequal distribution of prestige goods, trade ornaments specifically, following the appearance of Europeans (Cheng, 2008; Wang, 2011), hinting at an emergence of social inequality in the Indigenous community.
-We have examined ceramic vessel shape standardization to measure craft specialization as a proxy for social differentiation (Costin, 2001; Junker, 1999).
+Previous investigations at Kiwulan have suggested an unequal distribution of prestige goods, trade ornaments specifically, following the appearance of Europeans [@Cheng2008; @Wang2011], hinting at an emergence of social inequality in the Indigenous community.
+We have examined ceramic vessel shape standardization to measure craft specialization as a proxy for social differentiation [@Costin2001; @Junker1999].
 The result of our MANOVA test shows there is a significant difference in shapes between the pre-European and Post-European periods, and between the pre-European and the Chinese periods.
@@
 We can thus rule out changes in clay fabric as a factor in explaining changes in vessel shape.
-To determine if shape changes might be related to changes in the function of pots at Kiwulan, we used geochemical methods to extract and identify lipids trapped in the fabric of potsherds to identify foods that may have contributed residues absorbed into the clay (cf.
-Kwak and Marwick, 2015).
+To determine if shape changes might be related to changes in the function of pots at Kiwulan, we used geochemical methods to extract and identify lipids trapped in the fabric of potsherds to identify foods that may have contributed residues absorbed into the clay [cf.
+@Kwak2015].
 Unfortunately, we did not obtain useful results due to extremely low lipid yields, which is probably due to the very thin, dense, and low porosity fabric of Kiwulan pottery.
-These physical characteristics of the clay offer limited spaces to trap and protect organic molecules from microbiological degradation (cf.
-Evershed, 2008, p.
-909).
+These physical characteristics of the clay offer limited spaces to trap and protect organic molecules from microbiological degradation [cf.
+@Evershed2008, pp.
+909].
 
@@
 This contradicts our expectation that clustered pattern will be observed with an increase in pottery standardization since the emergence of specialized groups.
-The absence of clusters in the Chinese period is notable because this was a time of a historically-documented decline of Indigenous population (Chen, 2007; Hsieh, 2009).
+The absence of clusters in the Chinese period is notable because this was a time of a historically-documented decline of Indigenous population [@Chen2007; @Hsieh2009].
 We might expect reduced numbers of potters to result in pottery production shrinking to a few locations in the settlement during this time.
-Although the overall number of vessels is smaller during the Chinese contact period, Figure 7) shows that pottery is distributed randomly across the sampling area without any distinctive clusters during this time.
+Although the overall number of vessels is smaller during the Chinese contact period, (Figure \@ref(fig:kernel-plot)) shows that pottery is distributed randomly across the sampling area without any distinctive clusters during this time.
 This suggests that differences in population across our three occupation phases are probably not driving variation in craft specialization.

### 2020-04-11 analysis/paper/paper.Rmd
remove code for metric analysis

--- 2020-04-10 analysis/paper/paper.Rmd
@@
 ---
-title: "Investigating shape standardization using geometric morphometry of pottery from Iron Age northeastern Taiwan"
+title: "Investigating shape standardisation using geometric morphometry of pottery from Iron Age northeastern Taiwan"
 author:
@@
 A major historical factor that is often linked to social change in small-scale societies is colonial settlers introducing foreign trade goods to local Indigenous societies.
-The monopolization of long-distance trade goods has caused substantial transformations of Indigenous economic, cultural, and socio-political systems [@Dietler2005; @Dietler1997; @Junker1993; @Silliman2005].
-Pericolonial archaeology is the study of these indirect effects of colonialism, investigating areas where direct European colonial rule was limited, their conquests often short-lived and unsuccessful, but their commercial activities had economic and political impacts on Indigenous peoples living on the periphery of colonial control [@Acabado2017; @Trabert2017].
+The monopolization of long-distance trade goods has caused substantial transformations of Indigenous economic, cultural, and socio-political systems (Dietler, 2005, 1997; Junker, 1993; Silliman, 2005).
+Pericolonial archaeology is the study of these indirect effects of colonialism, investigating areas where direct European colonial rule was limited, their conquests often short-lived and unsuccessful, but their commercial activities had economic and political impacts on Indigenous peoples living on the periphery of colonial control (Acabado, 2017; Trabert, 2017).
 Pericolonial situations were common during the seventeenth to nineteenth centuries in East and Southeast Asia where European trading activity was extensive, but direct European rule less widespread.
 An emerging priority in archaeological research in this region is identifying the indirect influences are apparent on Indigenous communities during this time.
-For example, @Acabado2017’s study of Ifugao society in the Philippines highland suggests economic and political intensification during the Spanish presence as the response of Indigenous peoples to the Spanish cooptation.
+For example, Acabado (2017)’s study of Ifugao society in the Philippines highland suggests economic and political intensification during the Spanish presence as the response of Indigenous peoples to the Spanish cooptation.
 
 Indigenous societies’ responses to colonial contact range from passive acceptance through to active negotiation with the colonists, and accommodation or resistance of foreign intrusion.
-These responses can be identified through their daily cultural practices, such as their consumption patterns of foreign goods [@Dietler2015; @Given2004; @Mullins2011; @Scaramelli2005; @Silliman2001; @TorrenceandClarke2000].
-In this paper we investigate the archaeology of a pericolonial situation at Kiwulan (1350-1850 AD), a large multi-component archaeological site in Yilan County, northeastern Taiwan [@Chen2007], to identify the indirect impacts of colonial settler activity on local Indigenous societies.
+These responses can be identified through their daily cultural practices, such as their consumption patterns of foreign goods (Dietler, 2015; Given, 2004; Mullins, 2011; Scaramelli and Scaramelli, 2005; Silliman, 2001; Torrence and Clarke, 2000).
+In this paper we investigate the archaeology of a pericolonial situation at Kiwulan (1350-1950 AD), a large multi-component archaeological site in Yilan Province, northeastern Taiwan (Chen, 2007), to identify the indirect impacts of colonial settler activity on local Indigenous societies.
 Yilan is an ideal context to study peripheral colonial influences because Indigenous communities there were isolated by geographical barriers that limited the frequency of direct contact with the Spanish and the Dutch settlers in northern Taiwan.
@@
 
-Several measurements have been used for investigating ceramic standardization, such as metric, compositional, and technological variables [@Arnold2000; @Blackman1993; @Boness2015; @Costin1991; @Rice1991; @Roux2015; @Tite1999].
+Several measurements have been used for investigating ceramic standardization, such as metric, compositional, and technological variables (Arnold, 2000; Blackman et al., 1993; Boness et al., 2015; Costin, 1991; Rice, 1991; Roux, 2015; Tite, 1999).
 Among those variables, metric measurements are most widely applied to archaeological assemblages because of the ease of collecting these data.
-The coefficient of variation statistic is commonly used to quantify the degree of standardization in ceramic assemblages [@Eerkens2001; @Roux2003; @Stark1995; @Junker1999].
+The coefficient of variation statistic is commonly used to quantify the degree of standardization in ceramic assemblages (Eerkens and Bettinger, 2001; Roux, 2003; Stark, 1995b).
 However, because pottery vessels typically have curved shapes, linear measurements have limited sensitivity to many kinds of shape variations.
@@
 
-Geometric morphometrics differs from traditional linear measurements through its use of Cartesian coordinates of morphological structures to define shapes [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007].
+Geometric morphometrics differs from traditional linear measurements through its use of Cartesian coordinates of morphological structures to define shapes (Adams et al., 2004; Bookstein, 1997; Lawing and Polly, 2010; Slice, 2007).
 Landmarks, curves or outlines of objects can be represented by coordinates in terms of their unique point locations with respect to numerical values on coordinate axes.
-There are two common morphometric methods: landmark approaches and outline approaches [@Adams2004].
+There are two common morphometric methods: landmark approaches and outline approaches (Adams et al., 2004).
 Landmark approaches assign a set of landmarks or semilandmarks onto objects as reference points that can be specified on a coordinate system.
-Landmark-based morphometrics are widely applied to archaeological objects with obvious morphological features that can provide unambiguous reference points for landmark placement, such as projectile point tips or visually distinctive osteological features [@Birch2019; @Buchanan2019; @Cardillo2010; @Haruda2019; @Lycett2013; @Meloro2015].
+Landmark-based morphometrics are widely applied to archaeological objects with obvious morphological features that can provide unambiguous reference points for landmark placement, such as projectile point tips or visually distinctive osteological features (Birch and Martinón-Torres, 2019; Buchanan et al., 2019; Cardillo, 2010; Haruda et al., 2019; Lycett and Cramon-Taubadel, 2013; Meloro et al., 2015).
 
@@
 In those cases, outline approaches are more suitable for analyzing the overall shape of an object.
-One outline approach is the semi-landmarks method, also called sliding landmarks, which assigns points along the curve between two landmarks at defined intervals [@Bookstein1997; @Lawing2010].
-Those semi-landmarks are allowed to slide along the curve to remove the effect of the arbitrary landmark spacing by minimizing either Procrustes distance or bending energy [@Bookstein1997; @Gunz2013; @Slice2007].
-Another approach that is commonly applied to two-dimensional closed shapes is elliptic Fourier Analysis (EFA), that converts coordinates along a curve into Fourier function coefficients, called harmonic coefficients [@Kuhl1982].
-EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions [@Adams2004; @Bonhomme2014; @Claude2008].
+One outline approach is the semi-landmarks method, also called sliding landmarks, which assigns points along the curve between two landmarks at defined intervals (Bookstein, 1997; Lawing and Polly, 2010).
+Those semi-landmarks are allowed to slide along the curve to remove the effect of the arbitrary landmark spacing by minimizing either Procrustes distance or bending energy (Bookstein, 1997; Gunz and Mitteroecker, 2013; Slice, 2007).
+Another approach that is commonly applied to two-dimensional closed shapes is elliptic Fourier Analysis (EFA), that converts coordinates along a curve into Fourier function coefficients, called harmonic coefficients (Kuhl and Giardina, 1982).
+EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions (Adams et al., 2004; Bonhomme et al., 2014; Claude, 2008).
 The number of harmonics determines the quality and precision of the geometric representation of an object.
-The harmonic power, a cumulative sum of squared harmonic coefficients, provides a robust rule for determining the desired number of harmonics [@Bonhomme2014].
+The harmonic power, a cumulated sum of squared harmonic coefficients, provides a robust rule for determining the desired number of harmonics (Bonhomme et al., 2014).
 EFA is suitable for shapes lacking representative landmarks or where curves contain the most meaningful variation.
-Applications in archaeology include human remains and zooarchaeology, stone artifacts, and ceramics [@Fox2015; @Hoggard2019; @Iovictua2010; @Topi2018; @Selden2019; @Wilczek2014].
+Applications in archaeology include human remains and zooarchaeology, stone artifacts, and ceramics (Fox, 2015; Hoggard et al., 2019; Ioviţă, 2010; Topi et al.
+2018; Selden Jr 2019; Wilczek et al.
+2014).
 Taking the ceramics data from Kiwulan, northeastern Taiwan, we use EFA to evaluate the level of standardization of ceramics in relation to the European presence in the 17th century to get insights into the emergence of ceramic specialization.
@@
 
-Ceramics analyzed in this paper come from 40 units (4m by 4m each) sampled from the central, undisturbed area of archaeological excavations at Kiwulan (Figure \@ref(fig:KWL-map); Figure \@ref(fig:KWL-sam-area)).
-The chronology of the archaeological deposits at Kiwulan consists of two cultural components, the upper component and the lower component, with a sterile layer in between [@Chen2007].
-This paper focuses on the upper component, dated from AD 1350 to AD 1850, covering the late Iron Age and the historical period, which we define as the European occupation in Taiwan in the early 17th century.
-The Dutch first occupied southern Taiwan in 1624 and then the Spanish occupied northern Taiwan in 1626 [@Andrade2007].
+Ceramics analyzed in this paper come from 40 units (4m by 4m each) sampled from the central, undisturbed area of archaeological excavations at Kiwulan (Figure 1; Figure 2).
+The chronology of the archaeological deposits at Kiwulan consists of two cultural components, the upper component and the lower component, with a sterile layer in between (Chen, 2007).
+This paper focuses on the upper component, dated from AD 1350 to AD 1950, covering the late Iron Age and the historical period, which we define as the European occupation in Taiwan in the early 17th century.
+The Dutch first occupied southern Taiwan in 1624 and then the Spanish occupied northern Taiwan in 1626 (Andrade, 2007).
 In 1642, the Spanish was expelled by the Dutch in northern Taiwan, who took over their forts at Helping Dau in Keelung, and in Tamsui.
-Since then, western Taiwan was mostly under Dutch colonial rule until 1662 when the Kingdom of Tungning in Taiwan was founded by Koxinga, a loyalist of the Ming dynasty of China [@Andrade2007].
+Since then, western Taiwan was mostly under Dutch colonial rule until 1662 when the Kingdom of Tungning in Taiwan was founded by Koxinga, a loyalist of the Ming dynasty of China (Andrade, 2007).
 
@@
 Chi_no <- phase_fre$n[3]
-
-# read metric data
-kwl_p_ave <- readr::read_csv(here("analysis", "data", "raw_data", "KWL-pottery-metrics-tidy-ave.csv"))
-
-# number for each phase 
-kwl_p_ave_total <-
-  kwl_p_ave %>% 
-  count(period)
-
-# inline code later in the paper
-metric_all <- sum(kwl_p_ave_total$n)
-pre <- kwl_p_ave_total$n[3]
-euro <- kwl_p_ave_total$n[2]
-chi <- kwl_p_ave_total$n[1]
 ```
@@
 
-## Digitizing and analyzing by EFA
+## Digitising and analysing by EFA
 
@@
 Where only half cross-section images were available, these were duplicated, flipped, and then joined with another one to create a 2D closed outline for each vessel.
-Geometric morphometric analyses were conducted using the R software (R Core Team, 2019) and the functions included in the Momocs, a R package intended to quantify and analyze shapes ([@Bonhomme2014].
+Geometric morphometric analyses were conducted using the R software (R Core Team, 2019) and the functions included in the Momocs, a R package intended to quantify and analyze shapes (Bonhomme et al., 2014).
 The digilized outlines were converted into a list of successive xy pixel coordinates for elliptic Fourier analysis (EFA).
@@
 
-As a standardized measure of the spread of data, coefficients of variation (CV) allows us to directly compare variation in samples measured with different units or means.
-This is useful to compare the degree of standardization for archaeological assembles and enables comparison of variation across different sample sizes [@Eerkens2001, pp.498].
-Following @Eerkens2001 and @Roux2003, we take this as our measurement of standardization in vessel shape variables: lower CV values reflect higher standardization, and thus increased craft specialization in the community.
-Given that CV is robust for positive values due to the representation by ratio, we normalized PC scores to a range between 0 and 1 before computation of CV.
+As a standardized measure of the spread of data, coefficients of variation (CV) allow us to directly compare variation in samples measured with different units or means.
+This is useful to compare the degree of standardization for archaeological assembles and enables comparison of variation across different sample sizes (Eerkens and Bettinger, 2001, p.
+498).
+Following [citations of previous work that has used CVs like this] we take this as our measurement of standardization in vessel shape variables: lower CV values reflect higher standardization, and thus increased craft specialization in the community.
+Given that CV is robust for positive values due to the representation by ratio, we normalized PC scores to a range between 0 and 1 before computation of CV
 
 To answer the question of whether CV values for vessel samples across our three occupational phases are significantly different or not, we used the Modified signed-likelihood ratio (MSLR) test for equality of CVs [@Krishnamoorthy2014].
-While previous work has used the @Feltz1996's asymptotic test for the equality of coefficients of variation from k populations [@Lycett2008; @Okumura2014; @Eerkens2000; @Eerkens2001; @Hoggard2017], we prefer the MSLR test for shape variables as a more recent development with lower rates of type I error, better performance with uneven sample numbers, and more power across a range of conditions [@Krishnamoorthy2014].
+While previous work has used the @Feltz1996 asymptotic test for the equality of coefficients of variation from k populations [@Lycett2008; @Okumura2014; @Eerkens2000; @Eerkens2001; @Hoggard2017], we prefer the MSLR test for shape variables as a more recent development with lower rates of type I error, better performance with uneven sample numbers, and more power across a range of conditions [@Krishnamoorthy2014].
 
 To complement our investigation of craft specialization through shape standardization, we investigated spatial patterns of ceramic vessels at Kiwulan.
-As craft specialization increases, pottery distribution we expect a shift from a pattern of vessels dispersed across the site to a pattern of clusters that reflects the loci of production [@Costin2001].
+As craft specialization increases, pottery distribution we expect a shift from a pattern of vessels dispersed across the site to a pattern of clusters that reflects the loci of production (Costin, 2001).
 We used a Monte Carlo test for randomness in spatial locations of pots to explore whether their distribution is significantly clustered or dispersed.

### 2020-04-11 analysis/paper/paper.Rmd
edit the later part of discussion section

--- 2020-04-10 analysis/paper/paper.Rmd
@@
 Our results offer tentative support for the hypothesis that foreign influences at Kiwulan influenced emergence of social inequality in the local Indigenous society.
-If increased shape standardization is a reliable indicator of craft specialization, then we may be seeing evidence of a shift from corporate (group-based, distributed, collective, cooperative) to network (individual-based, competitive) organization (Feinman, 2000).
+If increased shape standardization is a reliable indicator of craft specialization, then we may be seeing evidence of a shift from corporate (group-based, distributed, collective, cooperative) to network (individual-based, competitive) organization [@Feinman2000].
 However, strong claims for an emergence of social complexity resulting from foreign contact at Kiwulan will need support from multiple and diverse sources of evidence that are beyond the scope of this paper.
-We find vessel shapes were more standardized in the Chinese period than the European period, contrary to previous work that has downplayed the effect of Chinese settlement in Yilan (Wang, 2011).
+We find vessel shapes were more standardized in the Chinese period than the European period.
 
 Compared to other regions in Taiwan, European colonial influence was weak in Yilan due to isolation by the surrounding mountains, and the economic focus of the Spanish and Dutch who preferred northern and northwest Taiwan as their trading base.
-Indigenous communities in Yilan experienced indirect influence from European trade networks and their colonial activities in a pericolonial context (cf.
-Acabado, 2017).
+Indigenous communities in Yilan experienced indirect influence from European trade networks and their colonial activities in a pericolonial context [cf.
+@Acabado2017].
 In contrast to the Indigenous-European interactions at Kiwulan, the interaction between Indigenous people and Chinese immigrants in the 19th century appears to have been more intense and direct.
-Historical records indicate that Chinese groups settled in Yilan and lived closely with Kiwulan Indigenous societies (Chen, 1963; Ke, 1993).
-This direct influence is reflected by the archaeological evidence of large amounts of Chinese porcelains and distinctive architectural bricks and tiles used by Chinese (Hsieh, 2009).
-Similarly, burials at Kiwulan in this later phase show the adoption of coffins in mortuary practices, which Chen (2007) interprets as the adoption of a symbol of ethnic Chinese.
+Historical records indicate that Chinese groups settled in Yilan and lived closely with Kiwulan Indigenous societies [@Chen1963; @Ke1993].
+This direct influence is reflected by the archaeological evidence of large amounts of Chinese porcelains and distinctive architectural bricks and tiles used by Chinese [@Hsieh2009].
+Similarly, burials at Kiwulan in this later phase show the adoption of coffins in mortuary practices, which @Chiu2004 interprets as the adoption of a symbol of ethnic Chinese.
 
@@
 
-# Acknowledgements
+# Acknowledgments
 

### 2020-04-11 analysis/paper/paper.Rmd
combine tps and PCA plots into one figure

--- 2020-04-11 analysis/paper/paper.Rmd
@@
 
-```{r pottery-spatial-pattern, fig.cap= "The spatial distribution of the pottery selected for shape analysis. The quantity is indicated by the color scale."}
-library(sf)
-library(viridis)
-
-# read in the spatial data
-AD_zone <- invisible(st_read(here("analysis", "data", "raw_data", "AD_zone.shp"), quiet = TRUE))
-units_40 <- invisible(read.csv(here("analysis", "data", "raw_data", "kwl-list-of-sampling-squares.csv")))
-
-AD_zone_tidy <-
-  AD_zone %>% 
-  mutate(Pit =paste0("P0",Id)) %>% 
-  inner_join(units_40, by = c("Pit" = "the_sq")) %>% 
-  select(Pit, geometry)
-
-# tidy pottery data
-pottery_count <-
-  pottery_data %>% 
-  select(Site, Context, Pit, Layer, No., phase) %>% 
-  count(Pit, phase)
-
-# join two dataset to get coordinate
-pottery_count_join <-
-  pottery_count %>%
-  mutate(phase = as.factor(phase)) %>% # seems need this line first
-  inner_join(AD_zone_tidy) %>% 
-  filter(!is.na(phase)) %>% 
-  mutate(phase = case_when(
-    phase %in% "pre-e"~ "Before European Contact",
-    phase %in% "post-e" ~ "European Presence",
-    phase %in% "ch-con" ~ "Chinese Presence")) %>% 
-  mutate(phase = factor(phase,
-                        level = c("Before European Contact",
-                                  "European Presence",
-                                  "Chinese Presence")))
-# plot spatial pattern for pottery
-plot_spatial_distribution_pottery_by_phase <- 
-  ggplot() +
-  geom_sf(data = AD_zone_tidy, fill = NA) +
-  geom_sf(data = pottery_count_join, 
-          aes(geometry = geometry, 
-              fill =n), 
-          alpha = 0.9)+
-  facet_wrap(~ phase) +
-  labs(fill = "frequency") +
-  scale_fill_viridis(direction = -1) +
-  theme_minimal() +
-  theme(plot.title = element_text(hjust = 0.5),
-        axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        strip.text.x = element_text(size = 6),
-        legend.text = element_text(size = 6),
-        legend.title = element_text(size = 6),
-        legend.key.size =  unit(3, "mm"))
-
-plot_spatial_distribution_pottery_by_phase
-
-ggsave(plot = plot_spatial_distribution_pottery_by_phase,
-       here("analysis", "figures",
-            "plot-spatial-distribution-pottery-by-period.jpg"),
-       width = 135, 
-       height = 31,
-       dpi = 300,
-       units = "mm")
-```
-
 ```{r jpg-coordinates, eval=FALSE}
+# this takes some time, so we don't run the code each time we knit, we run it once and save the output
 # start shape exploration
@@
 We found that 13 harmonics captured 99% of the total harmonic power in our elliptic Fourier coefficients of `r total_no` vessels from three phases.
-Figure \@ref(fig:thin-plate-splines) shows vessel shape changes described using thin-plate spline warping for paired periods, pre- and post-European periods, and post-European and Chinese periods, with the most greatest differences between pre-European and Chinese periods.
+Figure \@ref(fig:first-two-PCs-byplot-and-tps-plot) shows vessel shape changes described using thin-plate spline warping for paired periods, pre- and post-European periods, and post-European and Chinese periods, with the most greatest differences between pre-European and Chinese periods.
 
-```{r thin-plate-splines, fig.cap= "Comparison of average vessel shapes across phases using thin plate splines (TPS) with outline deformations required to pass from an extreme of one morphospace to another."}
+```{r thin-plate-splines}
 # Thin plate spline analysis (TPS)
@@
 
-par(mfrow=c(1,2)) # plot them side by side
+# save figure
+source(here("analysis/paper/my_tps_grid.R"))
+png(here("analysis/figures/tps-iso.png"), res = 300, w = 1800, h = 1000)
+par(mfrow=c(2,1)) # plot them side by side
 
-tps_grid(`Pre-European`, `Post-European`, shp.lwd = c(2, 2), shp.border = c("red", "blue"),
-           amp = 1, grid.size = 18, legend = T)
+my_tps_grid(`Pre-European`, 
+         `Post-European`, 
+         shp.lwd = c(2, 2), 
+         shp.border = c("red", "blue"),
+           amp = 1, 
+         grid.size = 18, 
+         legend = T)
 
-tps_grid(`Pre-European`, `Chinese`, shp.lwd = c(2, 2), shp.border = c("red", "blue"),
-           amp = 1, grid.size = 18, legend = T)
+my_tps_grid(`Pre-European`, 
+         `Chinese`, 
+         shp.lwd = c(2, 2), 
+         shp.border = c("red", "blue"),
+           amp = 1, 
+         grid.size = 18, 
+         legend = T)
 
 # Isodeformation lines
-png(here("analysis/figures/tps-iso.png"), res = 600, w = 1200, h = 1800)
+dev.off()
 
@@
 With the third component, the first three principal components explain `r vari_three`% of the total variance.
-PC1 captures the height of the vessels, from tall to short, and the roundness of the body from round to oval-shaped (Figure \@ref(fig:first-two-PCs-byplot)).
+PC1 captures the height of the vessels, from tall to short, and the roundness of the body from round to oval-shaped (Figure \@ref(fig:first-two-PCs-byplot-and-tps-plot)).
 PC2 relates to the neck and mouth constriction, from narrow to wide.
@@
 
-```{r first-two-PCs-byplot, fig.cap= "Pottery shape distribution by each occupation phase according to the first two PCs"}
+```{r}
+# PCA byplot, we save it to disk to use in the next chunk
+png(here("analysis/figures/pca-chull.png"), res = 300, w = 1800, h = 1000)
+plot_PCA(op.p, ~phase, chullfilled = TRUE)
+dev.off()
+```
 
-# PCA byplot
-plot_PCA(op.p, ~phase, chullfilled = TRUE)
+```{r}
+# composite composite image of thin plate spline and PCA output
+iso_plot <- image_read(here("analysis/figures/tps-iso.png"))
+iso_plot <- image_trim(iso_plot)
+pca_plot <- image_read(here("analysis/figures/pca-chull.png"))
+
+iso_plot_back <- image_scale(image_background(iso_plot, "none"), 360)
+iso_pca <- image_composite(pca_plot, iso_plot_back, offset = "+50+50")
+iso_pca_crop <- image_crop(iso_pca, geometry = geometry_area(1800, 750, 0, 0))
+image_write(iso_pca_crop, here("analysis/figures/iso_pca_composit_crop.png"))
+```
+
+```{r, first-two-PCs-byplot-and-tps-plot, echo=FALSE, fig.cap="caption", fig.align='Left: Comparison of average vessel shapes across phases using thin plate splines (TPS) with outline deformations required to pass from an extreme of one morphospace to another.
+Right: Pottery shape distribution by each occupation phase according to the first two PCs.'}
+knitr::include_graphics(here("analysis/figures/iso_pca_composit_crop.png"))
 ```
@@
 
+```{r pottery-spatial-pattern, fig.cap= "The spatial distribution of the pottery selected for shape analysis. The quantity is indicated by the color scale."}
+library(sf)
+library(viridis)
+
+# read in the spatial data
+AD_zone <- invisible(st_read(here("analysis", "data", "raw_data", "AD_zone.shp"), quiet = TRUE))
+units_40 <- invisible(read.csv(here("analysis", "data", "raw_data", "kwl-list-of-sampling-squares.csv")))
+
+AD_zone_tidy <-
+  AD_zone %>% 
+  mutate(Pit =paste0("P0",Id)) %>% 
+  inner_join(units_40, by = c("Pit" = "the_sq")) %>% 
+  select(Pit, geometry)
+
+# tidy pottery data
+pottery_count <-
+  pottery_data %>% 
+  select(Site, Context, Pit, Layer, No., phase) %>% 
+  count(Pit, phase)
+
+# join two dataset to get coordinate
+pottery_count_join <-
+  pottery_count %>%
+  mutate(phase = as.factor(phase)) %>% # seems need this line first
+  inner_join(AD_zone_tidy) %>% 
+  filter(!is.na(phase)) %>% 
+  mutate(phase = case_when(
+    phase %in% "pre-e"~ "Before European Contact",
+    phase %in% "post-e" ~ "European Presence",
+    phase %in% "ch-con" ~ "Chinese Presence")) %>% 
+  mutate(phase = factor(phase,
+                        level = c("Before European Contact",
+                                  "European Presence",
+                                  "Chinese Presence")))
+# plot spatial pattern for pottery
+plot_spatial_distribution_pottery_by_phase <- 
+  ggplot() +
+  geom_sf(data = AD_zone_tidy, fill = NA) +
+  geom_sf(data = pottery_count_join, 
+          aes(geometry = geometry, 
+              fill =n), 
+          alpha = 0.9)+
+  facet_wrap(~ phase) +
+  labs(fill = "frequency") +
+  scale_fill_viridis(direction = -1) +
+  theme_minimal() +
+  theme(plot.title = element_text(hjust = 0.5),
+        axis.text.x = element_blank(),
+        axis.text.y = element_blank(),
+        strip.text.x = element_text(size = 6),
+        legend.text = element_text(size = 6),
+        legend.title = element_text(size = 6),
+        legend.key.size =  unit(3, "mm"))
+
+plot_spatial_distribution_pottery_by_phase
+
+ggsave(plot = plot_spatial_distribution_pottery_by_phase,
+       here("analysis", "figures",
+            "plot-spatial-distribution-pottery-by-period.jpg"),
+       width = 135, 
+       height = 31,
+       dpi = 300,
+       units = "mm")
+```
+
 ```{r kernel-plot, fig.cap= "Kernel density map visulizes the probability of the density of pottery across space.

### 2020-04-11 analysis/paper/my_tps_grid.R

@@ -1,0 +1,43 @@
+my_tps_grid <- function (fr, to, amp = 1, over = 1.2, grid.size = 15, grid.col = "grey80",
+          poly = TRUE, shp = TRUE, shp.col = rep(NA, 2), shp.border = col_qual(2),
+          shp.lwd = c(1, 1), shp.lty = c(1, 1), legend = TRUE, legend.text,
+          ...)
+{
+  fr.n <- substitute(fr)
+  to.n <- substitute(to)
+  if (!missing(amp)) {
+    to <- to + (to - fr) * amp
+  }
+  grid0 <- .grid.sample(fr, to, nside = round(grid.size), over = over)
+  grid1 <- tps2d(grid0, fr, to)
+  dim.grid <- c(length(unique(grid0[, 1])), length(unique(grid0[,
+                                                                2])))
+  op <- par(mar = rep(0, 4))
+  on.exit(par(op))
+  plot(NA, xlim = range(grid1[, 1]) * over, ylim = range(grid1[,
+                                                               2]) * over, asp = 1, ann = FALSE, axes = FALSE, mar = rep(0,
+                                                                                                                         4))
+  for (i in 1:dim.grid[2]) {
+    lines(grid1[(1:dim.grid[1]) + (i - 1) * dim.grid[1],
+                ], col = grid.col)
+  }
+  for (i in 1:dim.grid[1]) {
+    lines(grid1[(1:dim.grid[2]) * dim.grid[1] - i + 1, ],
+          col = grid.col)
+  }
+  if (shp) {
+    points <- ifelse(poly, FALSE, TRUE)
+    coo_draw(fr, border = shp.border[1], col = NA, lwd = shp.lwd[1],
+             lty = shp.lty[1], points = points, first.point = FALSE,
+             centroid = FALSE, ...)
+    coo_draw(to, border = shp.border[2], col = NA, lwd = shp.lwd[2],
+             lty = shp.lty[2], points = points, first.point = FALSE,
+             centroid = FALSE, ...)
+    if (legend | !missing(legend.text)) {
+      if (missing(legend.text))
+        legend.text <- c(fr.n, to.n)
+      legend("center", legend = stringr::str_remove_all(legend.text, "`"), col = shp.border,
+             lwd = shp.lwd, bty = "n", cex = .45)
+    }
+  }
+}

### 2020-04-11 analysis/paper/paper.Rmd
add phase column to MANOVA table

@@
 manv <- rrtools:::quietly ( MANOVA_PW(op.p, "phase") ) 
-su_manv <- as.data.frame(manv$summary)
+su_manv <- as.data.frame(manv$summary) %>% 
+  rownames_to_column()
 
@@
   su_manv %>% 
-  mutate_all(round, 4) %>% 
+  mutate_if(is.numeric, round, 4) %>% 
   select(-Df, -`num Df`) %>% 
-  rename(`Pillai's trace` = Pillai,
+  rename(Comnparison = rowname,
+         `Pillai's trace` = Pillai,
          `Approximate F value` = `approx F`,

### 2020-04-11 analysis/paper/paper.Rmd
editing figs

@@
 # Isodeformation lines
-dev.off()
-
-#knitr::include_graphics(here::here("analysis", "figures", "tps-pre-Euro-to-post-Euro.png"))
+invisible(dev.off())
 ```
@@
 plot_PCA(op.p, ~phase, chullfilled = TRUE)
-dev.off()
+invisible(dev.off())
 ```
@@
 ```{r}
+library(magick)
 # composite composite image of thin plate spline and PCA output
@@
 
-```{r, first-two-PCs-byplot-and-tps-plot, echo=FALSE, fig.cap="caption", fig.align='Left: Comparison of average vessel shapes across phases using thin plate splines (TPS) with outline deformations required to pass from an extreme of one morphospace to another.
+```{r, first-two-PCs-byplot-and-tps-plot, echo=FALSE, fig.cap='Left: Comparison of average vessel shapes across phases using thin plate splines (TPS) with outline deformations required to pass from an extreme of one morphospace to another.
 Right: Pottery shape distribution by each occupation phase according to the first two PCs.'}
@@
   mutate(phase = case_when(
-    phase %in% "pre-e"~ "Before European Contact",
+    phase %in% "pre-e"~ "Before European\nContact",
     phase %in% "post-e" ~ "European Presence",
@@
   mutate(phase = factor(phase,
-                        level = c("Before European Contact",
+                        level = c("Before European\nContact",
                                   "European Presence",
@@
   theme_minimal() +
-  theme(plot.title = element_text(hjust = 0.5),
+  theme(panel.spacing = unit(3.5, "lines"),
+        plot.title = element_text(hjust = 0.5),
         axis.text.x = element_blank(),
         axis.text.y = element_blank(),
-        strip.text.x = element_text(size = 6),
+        strip.text.x = element_text(size = 11),
        legend.text = element_text(size = 6),
-        legend.title = element_text(size = 6),
-        legend.key.size =  unit(3, "mm"))
-
-plot_spatial_distribution_pottery_by_phase
+       legend.title = element_text(size = 6, vjust = 0.75),
+       legend.key.size =  unit(4, "mm"),
+        legend.position = "top")
 
@@
             "plot-spatial-distribution-pottery-by-period.jpg"),
-       width = 135, 
-       height = 31,
+       width = 275, 
+       height = 61,
        dpi = 300,
@@
 
-```{r kernel-plot, fig.cap= "Kernel density map visulizes the probability of the density of pottery across space.
-The map shows  major core area during the pre-European period, multiple core areas during the European period, and a single core during the Chinese period.
-Used the bandwidth based on Silverman (1986)'s rule of thumb"}
-knitr::include_graphics(here("analysis", "figures", "plot-kde-maps.jpg"))
+```{r kernel-plot, fig.cap= "A: The spatial distribution of the pottery selected for shape analysis.
+The quantity is indicated by the color scale.
+B: Kernel density map visulizes the probability of the density of pottery across space.
+The maps show a major core area during the pre-European period, multiple core areas during the European period, and a single core during the Chinese period.
+The bandwidth is based on Silverman (1986)"}
+
+p1 <- ggdraw(clip = 'on') + 
+  draw_image(here("analysis", "figures", "plot-spatial-distribution-pottery-by-period.jpg"), 
+             scale = 1.6,
+             x = -0.025)
+
+# code for this is in the spatial_pattern.R script file
+p2 <- ggdraw() + 
+  draw_image(here("analysis", "figures", "plot-kde-maps.jpg"), 
+             scale = 1)
+
+plot_grid(p1, p2, ncol = 1, labels="AUTO")
+ggsave(here("analysis/figures/kde-and-pottery-distribution-composite-plot.png"))
 ```
@@
 X-axis values based on meteres represent ANN expected value.
-Each sample distribution presents the null hypothesis with the blue line indicating the observed ANN value"}
-knitr::include_graphics(here("analysis", "figures", "plot-kde-ann-histograms.jpg"))
+Each sample distribution presents the null hypothesis with the blue line indicating the observed ANN value", fig.width=3}
+
+  image_resize(image_read( here("analysis", 
+                                "figures", 
+                                "plot-kde-ann-histograms.jpg")), 
+               geometry_area(width = 1000))
+
 ```

### 2020-04-11 analysis/paper/spatial_pattern.R

@@ -35,27 +35,6 @@ pottery_count_join <-
                         level = c("Before European Contact",
                                   "European Presence",
                                   "Chinese Presence")))
-# plot spatial pattern for pottery
-plot_spatial_distribution_pottery_by_phase <-
-  ggplot() +
-  geom_sf(data = AD_zone_tidy, fill = NA) +
-  geom_sf(data = pottery_count_join,
-          aes(geometry = geometry,
-              fill =n),
-          alpha = 0.9)+
-  facet_wrap(~ phase) +
-  labs(fill = "frequency") +
-  scale_fill_viridis(direction = -1) +
-  theme_minimal() +
-  theme(plot.title = element_text(hjust = 0.5),
-        axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        strip.text.x = element_text(size = 6),
-        legend.text = element_text(size = 6),
-        legend.title = element_text(size = 6),
-        legend.key.size =  unit(3, "mm"))
-
-plot_spatial_distribution_pottery_by_phase
 
 # --------------------------------point pattern analysis-------------------------------------------
 # get centroid and make a list of three periods
@@ -91,9 +70,12 @@ density_analysis <- function(x, ...) {
   # kernel density estimation with contour
   kernel_tmp <- density(pottery_point_ppp_tmp)
   kernel_to_plot <-
-    plot(kernel_tmp, main = "")
-  contour(kernel_tmp, add = TRUE)
-  title(..., line = -0.3, adj = 0.4, cex.main = 0.8)
+    graphics::plot(kernel_tmp,
+                   main = "",
+                   ribbon=TRUE,
+                   ribside=c("bottom"))
+    graphics::contour(kernel_tmp, add = TRUE)
+   #title(..., line = -0.3, adj = 0.4, cex.main = 1.5)
   return(list(pottery_point_ppp = pottery_point_ppp_tmp,
               kernel_to_plot = kernel_to_plot))
 }
@@ -102,8 +84,8 @@ titles <- names(pottery_period_join_cen)
 jpeg(
   here::here("analysis", "figures",
              "plot-kde-maps.jpg"),
-  width = 135,
-  height = 40,
+  width = 265,
+  height = 80,
   units = "mm",
   res = 300)
 par(mfrow = c(1, 3),
@@ -113,11 +95,8 @@ kernels_for_plotting <-
   map2(pottery_period_join_cen,
        titles,
        ~density_analysis(.x, main = .y))
-
[…]

### 2020-04-17 analysis/paper/paper.Rmd
merge and order the citation in the main text

@@
 To answer the question of whether CV values for vessel samples across our three occupational phases are significantly different or not, we used the Modified signed-likelihood ratio (MSLR) test for equality of CVs [@Krishnamoorthy2014].
-While previous work has used the @Feltz1996's asymptotic test for the equality of coefficients of variation from k populations [@Lycett2008; @Okumura2014; @Eerkens2000; @Eerkens2001; @Hoggard2017], we prefer the MSLR test for shape variables as a more recent development with lower rates of type I error, better performance with uneven sample numbers, and more power across a range of conditions [@Krishnamoorthy2014].
+While previous work has used the @Feltz1996's asymptotic test for the equality of coefficients of variation from k populations [@Eerkens2000; @Eerkens2001; @Hoggard2017; @Lycett2008; @Okumura2014], we prefer the MSLR test for shape variables as a more recent development with lower rates of type I error, better performance with uneven sample numbers, and more power across a range of conditions [@Krishnamoorthy2014].
 

### 2020-04-20 analysis/paper/paper.Rmd
add references

@@
 Our results offer tentative support for the hypothesis that foreign influences at Kiwulan influenced emergence of social inequality in the local Indigenous society.
-If increased shape standardization is a reliable indicator of craft specialization, then we may be seeing evidence of a shift from corporate (group-based, distributed, collective, cooperative) to network (individual-based, competitive) organization [@Feinman2000].
+If increased shape standardization is a reliable indicator of craft specialization, then we may be seeing evidence of a shift from corporate (group-based, distributed, collective, cooperative) to network (individual-based, competitive) organization [@Blanton1996; @Feinman1995; @Feinman2000; @Feinman2000political; @Feinman2010].
 However, strong claims for an emergence of social complexity resulting from foreign contact at Kiwulan will need support from multiple and diverse sources of evidence that are beyond the scope of this paper.

### 2020-04-20 analysis/paper/paper.Rmd
knit and fix cross-references

@@
 
-To test for differences in the distributions of shape variables indicated by the PC scores shown in Figure \@ref(fig:first-two-PCs-byplot), we used a multivariate analysis of variance (MANOVA) test to compare pairwise combinations across our three occupation phases.
+To test for differences in the distributions of shape variables indicated by the PC scores shown in Figure \@ref(fig:first-two-PCs-byplot-and-tps-plot), we used a multivariate analysis of variance (MANOVA) test to compare pairwise combinations across our three occupation phases.
 Table \@ref(tab:MANOVA-table) shows significant differences in shape between Pre-European and Post-European phases (p = $`r pval_post_pre`$), and Pre-European and Chinese phases (p = $`r pval_chi_pre`$).
@@
 To see whether the differences in the distribution of PCs between any two phases are substantive or due to chance, we used a modified signed-likelihood ratio test to assess the equality of CVs [@Krishnamoorthy2014; @Marwick2019].
-P-values for this significance test of CVs for PC1 and PC2 show significant differences in the standardization of vessel shapes across periods, especially between Chinese contact with either pre-European or post-European (Figure \@ref(fig:CV-test-PCs-plot), Table \@ref(tab:CV-test-PCs-table)).
+P-values for this significance test of CVs for PC1 and PC2 show significant differences in the standardization of vessel shapes across periods, especially between Chinese contact with either pre-European or post-European (Figure \@ref(fig:PCs-distribution), Table \@ref(tab:CV-test-PCs-table)).
 A significant difference was also detected between the pre-European and post-European periods.
@@
 We might expect reduced numbers of potters to result in pottery production shrinking to a few locations in the settlement during this time.
-Although the overall number of vessels is smaller during the Chinese contact period, (Figure \@ref(fig:kernel-plot)) shows that pottery is distributed randomly across the sampling area without any distinctive clusters during this time.
+Although the overall number of vessels is smaller during the Chinese contact period, Figure \@ref(fig:kernel-plot) shows that pottery is distributed randomly across the sampling area without any distinctive clusters during this time.
 This suggests that differences in population across our three occupation phases are probably not driving variation in craft specialization.

### 2020-04-20 analysis/paper/paper.Rmd
add highlights with BM's edits

@@
 highlights: |
-  These are the highlights. 
+  1. Iron Age pottery shapes from Kiwulan, Taiwan, were analyzed using elliptical fourier analysis
+  2. Statistical test for the equality of coefficients of variation of shape coefficients was applied
+  3. Significant differences in pottery shape were identified between culture contact phases
+  4. Pottery shapes show high standardization after contact with mainland Han Chinese groups
+
 ---

### 2020-04-23 analysis/paper/paper.Rmd
add image of pot

@@
 Those pots forms show high consistency in shape.
-They have a globular body with a short neck and wide mouth.
+They have a globular body with a short neck and wide mouth (Figure \@ref(fig:KWL-pots)).
 The exterior surface below the neck is decorated with a wide variety of impressed geometric motifs.
@@
 There are no substantial changes in the inclusions over time, indicating continuity in pottery fabric composition across the three periods.
+
+```{r, KWL-pots, fig.cap = "The Kiwulan pot (left) and an example of pottery drawing used in this paper (right)", fig.width=image_width_2_col_inch}
+knitr::include_graphics(here("analysis", "figures", "pots.png"))
+```
 
@@
 
-We used scans of pottery drawings acquired from the Bureau of Cultural Affairs in Yilan.
+We used scans of pottery drawings (Figure \@ref(fig:KWL-pots)) acquired from the Bureau of Cultural Affairs in Yilan.
 All drawings present a two-dimensional view of vessel cross-sections with indications of metric measurements.

### 2020-04-23 analysis/paper/paper.Rmd
revise the caption for pot image

@@
 
-```{r, KWL-pots, fig.cap = "The Kiwulan pot (left) and an example of pottery drawing used in this paper (right)", fig.width=image_width_2_col_inch}
+```{r, KWL-pots, fig.cap = "A typical pot from Kiwulan (left) and an example of a pottery drawing used for outline analysis (right)", fig.width=image_width_2_col_inch}
 knitr::include_graphics(here("analysis", "figures", "pots.png"))
@@
 Where only half cross-section images were available, these were duplicated, flipped, and then joined with another one to create a 2D closed outline for each vessel.
-Geometric morphometric analyses were conducted using the R software (R Core Team, 2019) and the functions included in the Momocs, a R package intended to quantify and analyze shapes ([@Bonhomme2014].
+Geometric morphometric analyses were conducted using the R software (R Core Team, 2019) and the functions included in the Momocs, a R package intended to quantify and analyze shapes [@Bonhomme2014].
 The digilized outlines were converted into a list of successive xy pixel coordinates for elliptic Fourier analysis (EFA).

### 2020-04-25 analysis/paper/paper.Rmd
submit to JASR

@@
 highlights: |
-  1. Iron Age pottery shapes from Kiwulan, Taiwan, were analyzed using elliptical fourier analysis
+  1. Iron Age pottery shapes from Kiwulan, Taiwan, were analyzed using elliptical Fourier analysis
   2. Statistical test for the equality of coefficients of variation of shape coefficients was applied
@@
   select(-Df, -`num Df`) %>% 
-  rename(Comnparison = rowname,
+  rename(Comparison = rowname,
          `Pillai's trace` = Pillai,

### 2020-06-25 analysis/paper/paper.Rmd
revise intro and edit description for Figure 4 based on R1 comments

@@
 ---
-title: "Investigating shape standardization using geometric morphometry of pottery from Iron Age northeastern Taiwan"
+title: "Standardization of ceramic shape: A case studay from the Iron Age pottery from northeastern Taiwan"
+
 author:
@@
 abstract: |
-  The emergence of ceramic specialization in prehistoric societies is often linked to shifts in the complexity of social structures, because specialized pottery production can reflect craft specialization and the presence of elite control.
+  The emergence of ceramic specialization in prehistoric societies is often linked to shifts in the complexity of social structures, because standardized ceramic production can reflect craft specialization and the presence of elite control.
 Previous work on identifying specialization relies on typological or linear metric analysis.
@@
 
-A major historical factor that is often linked to social change in small-scale societies is colonial settlers introducing foreign trade goods to local Indigenous societies.
-The monopolization of long-distance trade goods has caused substantial transformations of Indigenous economic, cultural, and socio-political systems [@Dietler2005; @Dietler1997; @Junker1993; @Silliman2005].
-Pericolonial archaeology is the study of these indirect effects of colonialism, investigating areas where direct European colonial rule was limited, their conquests often short-lived and unsuccessful, but their commercial activities had economic and political impacts on Indigenous peoples living on the periphery of colonial control [@Acabado2017; @Trabert2017].
+A major historical factor often linked to social change in small-scale societies is the introduction of foreign or exotic trade goods to local Indigenous societies [@Mullins2011].
+Monopolization of long-distance trade goods has caused substantial transformations in Indigenous economic, cultural, and socio-political systems [@Dietler2005; @Dietler1997; @Junker1993; @Silliman2005].
+Pericolonial archaeology is the study of these indirect effects of colonialism, investigating areas where direct European colonial rule was limited, their conquests were often short-lived and unsuccessful, but commercial activities yielded economic and political impacts on Indigenous peoples living on the periphery of colonial control [@Acabado2017; @Trabert2017].
 Pericolonial situations were common during the seventeenth to nineteenth centuries in East and Southeast Asia where European trading activity was extensive, but direct European rule less widespread.
-An emerging priority in archaeological research in this region is identifying the indirect influences are apparent on Indigenous communities during this time.
-For example, @Acabado2017’s study of Ifugao society in the Philippines highland suggests economic and political intensification during the Spanish presence as the response of Indigenous peoples to the Spanish cooptation.
+An emerging priority in archaeological research in Asia is identifying the indirect influences are apparent on Indigenous communities during the colonial period.
+For example, @Acabado2017’s study of Ifugao society in the Philippines highland suggests economic and political intensification during the Spanish presence in the lowlands as a strategy of Indigenous peoples to resist Spanish conquests.
 
-Indigenous societies’ responses to colonial contact range from passive acceptance through to active negotiation with the colonists, and accommodation or resistance of foreign intrusion.
-These responses can be identified through their daily cultural practices, such as their consumption patterns of foreign goods [@Dietler2015; @Given2004; @Mullins2011; @Scaramelli2005; @Silliman2001; @TorrenceandClarke2000].
-In this paper we investigate the archaeology of a pericolonial situation at Kiwulan (1350-1850 AD), a large multi-component archaeological site in Yilan County, northeastern Taiwan [@Chen2007], to identify the indirect impacts of colonial settler activity on local Indigenous societies.
-Yilan is an ideal context to study peripheral colonial influences because Indigenous communities there were isolated by geographical barriers that limited the frequency of direct contact with the Spanish and the Dutch settlers in northern Taiwan.
-Kiwulan is situated on a hill near a riverside at the northern margin of Yilan County, which is characterized by a triangular alluvial plain facing eastwards the Pacific with high mountains on three other sides.
+Indigenous societies' responses to colonial contact ranges from passive acceptance to active negotiation with the colonists, and accommodation or resistance of foreign intrusion, which can be identified through their daily cultural practices, such as their consumption patterns of foreign goods [@Dietler2015; @Given2004; @Mullins2011; @Scaramelli2005; @Silliman2001; @TorrenceandClarke2000].
+In this paper we investigate the archaeology of a pericolonial situation at Kiwulan (ca.
+AD 1350-1850), a large multi-component archaeological site in Yilan County, northeastern Taiwan [@Chen2007], to identify the indirect impacts of colonial settler activity on local Indigenous societies.
+Yilan is an ideal context to study peripheral colonial influences because the Indigenous communities were isolated by geographical barriers, limiting the frequency of direct contact with the Spanish and the Dutch settlers in northern Taiwan.
+Kiwulan is situated on a hill near a riverside at the northern margin of Yilan County, which is characterized by a triangular alluvial plain facing east toward the Pacific with high mountains on three other sides.
 
-Our specific aim is to investigate if there was increasing ceramic specialization resulting from Indigenous interaction with Europeans in the 17th century or Chinese in the 19th century, two major foreign influences in early historical Taiwan, that might indicate social changes in local Indigenous societies.
-We predict that competition within the Indigenous community at Kiwulan for foreign resources and trade partnerships with European or Chinese colonizers may have led to the emergence of craft specialization caused by greater economic and social control of pottery production by a small group of individuals.
-Using pottery shape as a proxy to study craft specialization, we ask: Did colonial trade impact the shape of locally-produced Indigenous pottery vessels?
+This research investigates if there was increasing ceramic specialization resulting from Indigenous interaction with Europeans in the 17th century or Chinese in the 19th century, two major foreign influences in early historical Taiwan, which may indicate social changes in local Indigenous societies.
+We predict that competition within the Indigenous community at Kiwulan for foreign resources and trade partnerships with European or Chinese colonizers may have led to the emergence of craft specialization caused by greater economic and social control of ceramic production by a small group of individuals.
+Using standardization in ceramic shapes as a proxy for craft specialization, we ask: Did colonial trade impact the shape of locally-produced Indigenous pottery vessels?
 Did pottery shape become more homogeneous and standardized after foreign contacts with European colonizers or Chinese immigrants?
 
-Several measurements have been used for investigating ceramic standardization, such as metric, compositional, and technological variables [@Arnold2000; @Blackman1993; @Boness2015; @Costin1991; @Rice1991; @Roux2015; @Tite1999].
-Among those variables, metric measurements are most widely applied to archaeological assemblages because of the ease of collecting these data.
-The coefficient of variation statistic is commonly used to quantify the degree of standardization in ceramic assemblages [@Eerkens2001; @Roux2003; @Stark1995; @Junker1999].
+Several measurements have been used for investigating ceramic standardization that include metric, compositional, and technological variables [@Arnold2000; @Blackman1993; @Boness2015; @Costin1991; @Rice1991; @Roux2015; @Tite1999].
+Among those variables, metric measurements are most widely applied to archaeological assemblages.
+The coefficient of variation (CV) statistic is regularly used to quantify the degree of standardization in ceramic assemblages [@Eerkens2001; @Roux2003; @Stark1995; @Junker1999].
 However, because pottery vessels typically have curved shapes, linear measurements have limited sensitivity to many kinds of shape variations.
-Thus, to capture subtle shape variations that might also be relevant to standardization, we analyze pottery shape using geometric morphometric methods.
+Thus, to capture subtle shape variations that might also be relevant to standardization, we analyze ceramic shapes using geometric morphometric (GM) methods.
 
@@
 
-```{r, first-two-PCs-byplot-and-tps-plot, echo=FALSE, fig.cap='Left: Comparison of average vessel shapes across phases using thin plate splines (TPS) with outline deformations required to pass from an extreme of one morphospace to another.
+```{r, first-two-PCs-byplot-and-tps-plot, echo=FALSE, fig.cap='Left: Significant differences in average vessel shapes across phases using thin plate splines (TPS) with outline deformations required to pass from an extreme of one morphospace to another.
 Right: Pottery shape distribution by each occupation phase according to the first two PCs.'}
@@
 Table \@ref(tab:MANOVA-table) shows significant differences in shape between Pre-European and Post-European phases (p = $`r pval_post_pre`$), and Pre-European and Chinese phases (p = $`r pval_chi_pre`$).
+These results match the differences in the visualization of average shapes between the occupations in the thin plate splines (Figure \@ref(fig:first-two-PCs-byplot-and-tps-plot), see left).
 Although there is considerable overlap of shape variables between the Pre-European and Post-European phases, their PC scores are significantly different from each other.

### 2020-06-26 analysis/paper/paper.Rmd
revised GM section based on R1's comments

@@
 
-## Geometric Morphometry
+## Geometric Morphometrics
 
-Geometric morphometrics differs from traditional linear measurements through its use of Cartesian coordinates of morphological structures to define shapes [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007].
+Geometric morphometrics differs from traditional linear measurements through its use of Cartesian coordinates of morphological structures to define and analyze shape [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007].
 Landmarks, curves or outlines of objects can be represented by coordinates in terms of their unique point locations with respect to numerical values on coordinate axes.
-There are two common morphometric methods: landmark approaches and outline approaches [@Adams2004].
-Landmark approaches assign a set of landmarks or semilandmarks onto objects as reference points that can be specified on a coordinate system.
-Landmark-based morphometrics are widely applied to archaeological objects with obvious morphological features that can provide unambiguous reference points for landmark placement, such as projectile point tips or visually distinctive osteological features [@Birch2019; @Buchanan2019; @Cardillo2010; @Haruda2019; @Lycett2013; @Meloro2015].
+There are two common morphometric methods: landmark and outline approaches [@Adams2004].
+Landmark GM approaches assign a set of landmarks and/or semilandmarks onto objects as reference points that can be specified on a coordinate system.
+Landmark-based morphometrics are widely applied to archaeological objects with obvious morphological features that provide unambiguous reference points for landmark placement, such as projectile point tips or visually distinctive osteological features [@Birch2019; @Buchanan2019; @Cardillo2010; @Haruda2019; @Lycett2013; @Meloro2015; @Selden2019].
 
-However, a key limitation of the landmark approach is that landmarks may be difficult to reproducibly locate for structures that are mostly or entirely curves.
-In those cases, outline approaches are more suitable for analyzing the overall shape of an object.
-One outline approach is the semi-landmarks method, also called sliding landmarks, which assigns points along the curve between two landmarks at defined intervals [@Bookstein1997; @Lawing2010].
-Those semi-landmarks are allowed to slide along the curve to remove the effect of the arbitrary landmark spacing by minimizing either Procrustes distance or bending energy [@Bookstein1997; @Gunz2013; @Slice2007].
-Another approach that is commonly applied to two-dimensional closed shapes is elliptic Fourier Analysis (EFA), that converts coordinates along a curve into Fourier function coefficients, called harmonic coefficients [@Kuhl1982].
+However, a key limitation of the landmark approach is that landmarks may be difficult to reproducibly locate for structures that are mostly or entirely curves if not mathematically-defined.
+In those cases, Elliptic Fourier Analysis (EFA) is an excellent mean for assessing morphological variation in whole structure of two-dimensional closed shapes.
+EFA has a long history in archaeology [@Gero1984], and has been increasingly used for the shapes of artifacts, such as stone artifacts and ceramics, and human remains and animal bones [@Fox2015; @Hoggard2019; @Iovictua2010; @Topi2018; @Wilczek2014].
 EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions [@Adams2004; @Bonhomme2014; @Claude2008].
+To be specific, EFA converts coordinates along a curve into Fourier function coefficients, called harmonic coefficients or harmonics [@Kuhl1982].
 The number of harmonics determines the quality and precision of the geometric representation of an object.
 The harmonic power, a cumulative sum of squared harmonic coefficients, provides a robust rule for determining the desired number of harmonics [@Bonhomme2014].
-EFA is suitable for shapes lacking representative landmarks or where curves contain the most meaningful variation.
-Applications in archaeology include human remains and zooarchaeology, stone artifacts, and ceramics [@Fox2015; @Hoggard2019; @Iovictua2010; @Topi2018; @Selden2019; @Wilczek2014].
-Taking the ceramics data from Kiwulan, northeastern Taiwan, we use EFA to evaluate the level of standardization of ceramics in relation to the European presence in the 17th century to get insights into the emergence of ceramic specialization.
+EFA is suitable for analyzing shapes lacking representative landmarks, or where curves contain the most meaningful variation.
+In such cases, the focus is the whole shape instead of the different components of a shape.
+We use EFA to evaluate the level of standardization of ceramics data from Kiwulan, northeastern Taiwan, in relation to the European presence in the 17th century to gain insights into the emergence of ceramic specialization.
+The reason of using EFA is because there is no sufficient landmarks could be identified straightforwardly due to the globular body of ceramics.
 We use a significance test for the equality of coefficient of variations of shape variables to statistically compare the vessel standardization from different periods.

### 2020-06-27 analysis/paper/paper.Rmd
finished small edits based on R1's comments

@@
 
+<!--discussion of standardization in material culture using GM approaches-->
+
 # Archaeological background and materials
 
-Ceramics analyzed in this paper come from 40 units (4m by 4m each) sampled from the central, undisturbed area of archaeological excavations at Kiwulan (Figure \@ref(fig:KWL-map); Figure \@ref(fig:KWL-sam-area)).
-The chronology of the archaeological deposits at Kiwulan consists of two cultural components, the upper component and the lower component, with a sterile layer in between [@Chen2007].
-This paper focuses on the upper component, dated from AD 1350 to AD 1850, covering the late Iron Age and the historical period, which we define as the European occupation in Taiwan in the early 17th century.
-The Dutch first occupied southern Taiwan in 1624 and then the Spanish occupied northern Taiwan in 1626 [@Andrade2007].
-In 1642, the Spanish was expelled by the Dutch in northern Taiwan, who took over their forts at Helping Dau in Keelung, and in Tamsui.
-Since then, western Taiwan was mostly under Dutch colonial rule until 1662 when the Kingdom of Tungning in Taiwan was founded by Koxinga, a loyalist of the Ming dynasty of China [@Andrade2007].
+Ceramics analyzed in this study come from 40 units (4m by 4m each) sampled from the central, undisturbed area of archaeological excavations at Kiwulan (Figure \@ref(fig:KWL-map); Figure \@ref(fig:KWL-sam-area)).
+The chronology of the archaeological deposits consist of two cultural components, the upper and the lower, with a sterile layer in between [@Chen2007].
+We focus on the upper component, dated from AD 1350 to 1850, spans the late Iron Age and the historical period, of which the start is defined as the European occupation in Taiwan in the early 17th century.
+The Dutch first occupied southern Taiwan in 1624, followed by the Spanish in northern Taiwan in 1626 [@Andrade2007].
+In 1642, the Spanish was expelled by the Dutch, who then occupied the previous Spanish forts at Helping Dau in Keelung, and in Tamsui.
+Western Taiwan remained under Dutch colonial rule until 1662 when the Kingdom of Tungning in Taiwan was founded by Koxinga, a loyalist of the Ming dynasty of China [@Andrade2007].
 
-```{r, KWL-map, fig.cap = "Map showing the location of Kiwulan, and other places in northern Taiwan named in the text. Map data from naturalearthdata.com", fig.width=image_width_2_col_inch}
+```{r, KWL-map, fig.cap = "Map illustrating the location of Kiwulan, and other locations in northern Taiwan that are named in the text.
+Map data is from naturalearthdata.com", fig.width=image_width_2_col_inch}
 library(here)
@@
 
-The archaeological record of Kiwulan's upper component shows traces of foreign contacts, including Europeans in the 17th century, and waves of Chinese immigrants in the 19th century.
-Imported ceramics from mainland China, stoneware, and ornaments such as beads have been found in the upper component, indicating frequent long-distance trade activities with Europeans and Chinese merchants.
-In addition to artifacts, archaeological features such as burials, middens, and post-holes with _in-situ_ posts are widespread across the 1-2 m thick deposit of the upper component, and demonstrate that Kiwulan was a continuously occupied large settlement site [@Chen2007].
-To compare different foreign influences, we classified the upper component into three chronological phases: the pre-European, European, and Chinese periods.
-We identified these phases using 32 previously published radiocarbon ages [@Chen2007], excavation depth measurements, stratigraphic details reported by the excavators (color, texture, disturbance, etc.), and finds of chronologically diagnostic artifacts, such as blue and white porcelains, light grey glazed jars, and large dark brown glazed stoneware jars commonly used in the 17th century, and bricks and tiles used by the Chinese in the 19th century [@Chen2007; @Hsieh2009; @Wang2011].
-The deposit shows signs of continuous human occupation during each of the three phases across the sampling area.
+The archaeological record of Kiwulan's upper component shows traces of foreign contact, including Europeans in the 17th century, and waves of Chinese immigrants in the 19th century.
+Imported ceramics from mainland China, stoneware, and ornaments such as beads have been recovered in the upper component, indicating frequent long-distance trade activities with Europeans and Chinese merchants.
+Archaeological features such as burials, middens, and post-holes with _in-situ_ posts are widespread across the 1-2 m thick deposit of the upper component, and demonstrate that Kiwulan was a continuously occupied large settlement site [@Chen2007].
+To compare different foreign influences, we classified the upper component into three chronological phases: pre-European, European, and Chinese.
+We identified these phases based on 32 previously published radiocarbon ages [@Chen2007], excavation depth measurements, stratigraphic details reported by the excavators (color, texture, disturbance, etc.), and finds of chronologically diagnostic artifacts, such as blue and white porcelains, light grey glazed jars, and large dark brown glazed stoneware jars commonly used in the 17th century, and bricks and tiles employed by the Chinese in the 19th century [@Chen2007; @Hsieh2009; @Wang2011].
+The deposit exhibited signs of continuous human occupation in each of the three phases with no apparent breaks.
 
-The most abundant artifacts in the upper component are locally made ceramics, which are distributed throughout the sequence and across the site.
+The most abundant artifacts in the upper component are locally manufactured ceramics, which are distributed throughout the temporal sequence, and across the study site.
 More than 550,000 sherds were excavated, and around 1,200 vessels could be partially reconstructed (i.e.
-complete rim, or complete base).
-There are only two forms of local vessels, a cooking pot and a steamer made of two cooking pots stacked together with a clay filter layer in between.
-Those pots forms show high consistency in shape.
-They have a globular body with a short neck and wide mouth (Figure \@ref(fig:KWL-pots)).
-The exterior surface below the neck is decorated with a wide variety of impressed geometric motifs.
-These vessels were probably used for cooking, with evidence of charred residues and carbon deposits frequently observed on vessel interiors and soot on their exteriors.
-They are fired to orange to brownish color with a fully oxidized core or a reduced core with oxidized fringes.
-Finger impressions and seams usually on the interior indicate they were pinched using slabs of clay and shaped by hand.
-This kind of pot has been widely found at archaeological sites during the late Iron Age and the historical period throughout the Yilan Plain.
+complete rim or base).
+There are two shapes of locally-manufactured vessels; a cooking pot and a steamer made of two cooking pots stacked together with a clay filter between.
+Those vessel shapes demonstrate suites of standard morphological components.
+Each has a globular body with a short neck and wide mouth (Figure \@ref(fig:KWL-pots)).
+The exterior surface below the neck is decorated with a variety of impressed geometric motifs.
+These vessels were potentially used for cooking indicated by evidence of charred residues and carbon deposits frequently observed on vessel interiors, and soot on vessel exteriors.
+Firing resulted in orange and brownish color with a fully oxidized core, or a reduced core with oxidized fringes [@Chen2007].
+Finger impressions and seams, usually on the interior, indicate that the vessels were shaped by hand, and that the seems between the slabs were pinched.
+This kind of vessel has been widely found at archaeological sites during the late Iron Age and the historical period throughout the Yilan Plain [@NMTH2005].
 
-Petrographic analysis for 34 thin sections shows high percentage of inclusions (15-50%), including argillite (15-40%), metasandstone (1-10%),  sandstone fragments (1-6%), quartz (1-5%), and trace amounts of feldspar and slate.
-The size of the particles ranges from 500 micron to 1300 micron.
-In general, the globular vessel fabric presents a mixture of fine, rounded argillite with a small part of rounded metasandstone and rounded to sub-angular monocrystalline quartz.
-The composition matches the mineralogical composition of local raw materials in the Yilan Plain [@Chen2016].
+Petrographic analysis for 34 thin sections presents a high percentage of inclusions (15-50%), including argillite (15-40%), metasandstone (1-10%),  sandstone fragments (1-6%), quartz (1-5%), and trace amounts of feldspar and slate.
+Particle sizes of range from 500 to 1300 microns.
+In general, the globular vessel fabric presents a mixture of fine, rounded argillite with a small amounts of rounded metasandstone and rounded to sub-angular monocrystalline quartz.
+This composition is consistent with the mineralogical composition of local raw materials found in the Yilan Plain [@Chen2016].
 There are no substantial changes in the inclusions over time, indicating continuity in pottery fabric composition across the three periods.
+<!--Was this tested using statistics?
+If so, then state that there are no significant changes in the inclusions over time, and include that script in your supplementary materials.-->
 
@@
 
-We used all `r total_no` vessels that could be reconstructed with rim, body and base parts and securely provenanced to pre-contact contexts (n = `r pre_no`), post-European contexts (n = `r Euro_no`), and Chinese contact contexts (n = `r Chi_no`) for elliptic Fourier analysis.
+The sample consists of `r total_no` reconstructed vessels with rim, body and base parts, and we securely provenanced to pre-contact (n = `r pre_no`), post-European (n = `r Euro_no`), and Chinese contact contexts (n = `r Chi_no`).
 
@@
 
-We used scans of pottery drawings (Figure \@ref(fig:KWL-pots)) acquired from the Bureau of Cultural Affairs in Yilan.
-All drawings present a two-dimensional view of vessel cross-sections with indications of metric measurements.
-The scanned drawings were imported into Inkscape (http://inkscape.org) where outlines were traced manually.
-Where only half cross-section images were available, these were duplicated, flipped, and then joined with another one to create a 2D closed outline for each vessel.
-Geometric morphometric analyses were conducted using the R software (R Core Team, 2019) and the functions included in the Momocs, a R package intended to quantify and analyze shapes [@Bonhomme2014].
-The digilized outlines were converted into a list of successive xy pixel coordinates for elliptic Fourier analysis (EFA).
-The harmonic coefficients generated by EFA were analyzed by principal component analysis (PCA) for dimensionality reduction to illustrate the diversity of the shape data and identify major patterns of variation.
+We used scans of pottery drawings acquired from the Bureau of Cultural Affairs in Yilan (Figure \@ref(fig:KWL-pots)).
+All drawings provide a two-dimensional view of vessel cross-sections, and metric measurements.
+The scanned drawings were imported into Inkscape (http://inkscape.org) for digitization where outlines were manually traced.
+In those instances where only one side of the cross-section image was available, these were mirrored and joined to create a closed outline for each vessel.
+Analyses were conducted in R software (R Core Team, 2019) using functions included in the Momocs package for quantifying and analyzing shapes [@Bonhomme2014].
+Outlines were converted into a list of successive x, y pixel coordinates for EFA.
+We analyzed harmonic coefficients by principal component analysis (PCA) for dimensionality reduction to illustrate the diversity of the shape data and identify major patterns of variation.
 
@@
 
-We found that 13 harmonics captured 99% of the total harmonic power in our elliptic Fourier coefficients of `r total_no` vessels from three phases.
-Figure \@ref(fig:first-two-PCs-byplot-and-tps-plot) shows vessel shape changes described using thin-plate spline warping for paired periods, pre- and post-European periods, and post-European and Chinese periods, with the most greatest differences between pre-European and Chinese periods.
+We found that 13 harmonics captured 99% of the total harmonic power in the elliptic Fourier coefficients of `r total_no` vessels from three phases.
+Figure \@ref(fig:first-two-PCs-byplot-and-tps-plot) illustrates significant differences in vessel shapes described using thin-plate spline warping for paired periods, pre- and post-European periods, and post-European and Chinese periods, with the greatest differences between pre-European and Chinese periods.
 
@@
 PC2 relates to the neck and mouth constriction, from narrow to wide.
-PC3 explains a smaller portion of the variance (`r pc3`%), which relates to the degree of the flare of the neck, from a curved to straight shape.
+PC3 explains a smaller portion of the variance (`r pc3`%), which relates to the degree of the flare in the neck, from a curved to straight shape.
 The results reflect a large overlap in shapes from three occupations phases, especially for shapes in the pre-European and post-European periods.
@@
 
-To test for differences in the distributions of shape variables indicated by the PC scores shown in Figure \@ref(fig:first-two-PCs-byplot-and-tps-plot), we used a multivariate analysis of variance (MANOVA) test to compare pairwise combinations across our three occupation phases.
-Table \@ref(tab:MANOVA-table) shows significant differences in shape between Pre-European and Post-European phases (p = $`r pval_post_pre`$), and Pre-European and Chinese phases (p = $`r pval_chi_pre`$).
-These results match the differences in the visualization of average shapes between the occupations in the thin plate splines (Figure \@ref(fig:first-two-PCs-byplot-and-tps-plot), see left).
-Although there is considerable overlap of shape variables between the Pre-European and Post-European phases, their PC scores are significantly different from each other.
-There is little difference between vessel shapes in the Post-European and Chinese contact periods.
+To test for differences in the distributions of shape variables indicated by the PC scores shown in Figure \@ref(fig:first-two-PCs-byplot-and-tps-plot), we used a multivariate analysis of variance (MANOVA) test to compare pairwise combinations across the three occupation phases.
+Table \@ref(tab:MANOVA-table) demonstrated the significant differences in shape between Pre-European and Post-European phases (p = $`r pval_post_pre`$), and Pre-European and Chinese phases (p = $`r pval_chi_pre`$).
+These results are consistent with the differences in the visualization of average shapes between the occupations in the thin plate splines (Figure \@ref(fig:first-two-PCs-byplot-and-tps-plot), see left).
+Although there is considerable overlap of shape variables between the Pre-European and Post-European phases, their PC scores differ significantly.
+There is not a significant difference in vessel shapes between the Post-European and Chinese contact periods.
 
@@
 
-Previous investigations at Kiwulan have suggested an unequal distribution of prestige goods, trade ornaments specifically, following the appearance of Europeans [@Cheng2008; @Wang2011], hinting at an emergence of social inequality in the Indigenous community.
-We have examined ceramic vessel shape standardization to measure craft specialization as a proxy for social differentiation [@Costin2001; @Junker1999].
-The result of our MANOVA test shows there is a significant difference in shapes between the pre-European and Post-European periods, and between the pre-European and the Chinese periods.
-The average shape presents a round body with a wider rim and neck before European contact, and turns to a more oval-shaped body with narrower rim and neck after the European presence.
-These changes even more pronounced during the Chinese contact period.
-In general, pottery vessels become shorter in height over time, leading to an oval-shaped body.
+Previous investigations at Kiwulan suggested an unequal distribution of prestige goods, trade ornaments in particular, following the appearance of Europeans [@Cheng2008; @Wang2011], which hinted at the emergence of social inequality within the Indigenous community.
+Ceramic vessel shape standardization was examined to measure craft specialization as a proxy for social differentiation [@Costin2001; @Junker1999].
+The result of our MANOVA demonstrates a significant difference in shapes between the pre-European and Post-European periods, and between the pre-European and the Chinese periods.
+The average shape presents as a round body with a wide rim and neck before European contact, which shifts to a more oval-shaped body with narrower rim and neck after the European presence.
+Shape changes are more pronounced during the Chinese contact period.
+In general, vessels become shorter in height over time, leading to an oval-shaped body.
 
@@
 However, strong claims for an emergence of social complexity resulting from foreign contact at Kiwulan will need support from multiple and diverse sources of evidence that are beyond the scope of this paper.
-We find vessel shapes were more standardized in the Chinese period than the European period.
+We find that vessel shapes were more standardized during the Chinese period than the European period.
 
@@
 
-We recognize that the shape variation reported here is subtle, and this invites consideration of the possibility that the absence of major changes in vessel shape at Kiwulan may have been an act of resistance to foreign influence.
+The shape variation reported here is subtle, and invites consideration of the possibility that the absence of major changes in vessel shape at Kiwulan may have been an act of resistance to foreign influence.
 Continuities in vessel shape over time draws our attention to the endurance of traditional pottery production practices amid intrusions from Europeans and Chinese.
-In a culture contact situation, social identity may be expressed through material practices to show cultural homogeneity and distinction from other groups [@Voss2005].
-It is also important to recognize that social identity might be more complicated in a colonial context, and may be more than a colonized–colonizer or local/foreign dichotomy [@Voss2005; @Voss2008].
-Shamaoshan cemetery (3BC- 4AD) in Southwest China shows that the process of the incorporation of Southwest China into the Han Empire involved a century of conflicts, resistance, and acceptance among social groups with different identities, especially in the historical context of Han immigrants [@Wu2019].
+In a culture contact situation, social identity may be expressed through material practices as a means of expressing cultural homogeneity and distinction from other groups [@Voss2005].
+It is also important to recognize that social identity might be more complicated in a colonial context, and may be representative of more than a colonized–colonizer or local/foreign dichotomy [@Voss2005; @Voss2008].
+Shamaoshan cemetery dating from BC 3 to AD 4 in Southwest China suggests that the process of the incorporation of Southwest China into the Han Empire involved a century of conflicts, resistance, and acceptance among social groups with different identities, especially in the historical context of Han immigrants [@Wu2019].
 A similar dynamic may have occurred at Kiwulan, with vessel shape indicating both acceptance of foreign influence through increased shape standardization, and resistance through the overall continuity in vessel shape.
-Vessel shape can be viewed as a symbol as an expression the Indigenous identity and social boundaries because shape is a highly visible trait compared with other features of pottery [cf.
+Vessel shape can be viewed as a symbolic expression the Indigenous identity and social boundaries because shape is a highly visible trait compared with other features of pottery [cf.
 @Roux2015].
-Although there is an increase in number of imported ceramics over time at Kiwulan, the production of the local pottery continued and became more standardized.
-This might imply not only a utilitarian function of the local vessels, but a deliberate action to emphasize the local pottery tradition as their cultural custom as foreign contact intensified.
+Although there is an increase in number of imported ceramics through time at Kiwulan, production was continuous, and increasingly standardized.
+This might imply not only the utilitarian function, but a deliberate and increased emphasis on the local ceramic tradition, their cultural custom, as a response to intensified foreign contact [cf.
+@Acabado2017].
 
@@
 
-This study demonstrates an application of elliptic Fourier analysis on ceramic shapes to explore the emergence of ceramic specialization as indicative of foreign influences.
-Here, EFA is combined with a significance test for the equality of CVs of shape variables to provide a robust way to identify and statistically assess differences in shape standardization.
-The direct relationship between foreign influences and standardization of ceramic shape was tested on pottery from Kiwulan, a large Iron Age Indigenous settlement in northeastern Taiwan.
+This study demonstrated an application of EFA on ceramic shapes to explore the emergence of ceramic specialization as indicative of foreign influences.
+Here, EFA is combined with a significance test for the equality of CVs of shape variables to provide a robust method of identifying and statistically assessing differences in shape standardization.
+The direct relationship between foreign influences and standardization of ceramic shape was tested on ceramics from Kiwulan, a large Iron Age Indigenous settlement in northeastern Taiwan.
 Lower variation in ceramic shape was identified after European presence began, and even lower variation during the period of Chinese presence.
-Our findings help to understand the factors that may lead to standardization of pottery production in a pericolonial interaction context.
-More homogeneous shapes during the contact periods, without any changes in clay paste composition or production technique, suggests that shape standardization was intentional.
-The results here further suggest that expressions of social identity or cultural boundaries in Indigenous societies through highly visible vessel qualities, such as shape, may be heightened during periods of foreign contact in pericolonial contexts.
+Our findings help to expand upon those factors that may lead to the standardization of ceramic production in a pericolonial interaction context.
+More homogeneous shapes during the contact periods, without any changes in clay paste composition or production technique, suggest that shape standardization was intentional.
+The results further suggest that expressions of social identity or cultural boundaries in Indigenous societies through highly visible vessel qualities, such as shape, may be heightened during periods of foreign contact in pericolonial contexts.
 Our analysis, with its openly available methods and data, is readily extensible to other pottery assemblages in the region to further explore related questions about craft specialization and standardization in Iron Age ceramic technologies.

### 2020-07-03 analysis/paper/paper.Rmd
edits on EFA methods

@@
 In those cases, Elliptic Fourier Analysis (EFA) is an excellent mean for assessing morphological variation in whole structure of two-dimensional closed shapes.
-EFA has a long history in archaeology [@Gero1984], and has been increasingly used for the shapes of artifacts, such as stone artifacts and ceramics, and human remains and animal bones [@Fox2015; @Hoggard2019; @Iovictua2010; @Topi2018; @Wilczek2014].
+EFA was first introduced to study shapes of artifacts in archaeology in 1980s [@Gero1984].
+Most applications are associated with human remains and animal bones to study biological variation, and has been increasing for stone artifacts to explore stone tool topology and technology [@Fox2015; @Hoggard2019; @Iovictua2010].
+Despite the EFA applications for ceramics is rare, they demonstrate the potential of this promising approach for answering questions related to ceramic taxonomy [@Topi2018; @Wilczek2014].
+For example, @Wilczek2014 evaluate the concordance between EFA and Discrete Cosine Transform (DCT), and a traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France.
+The results show the variation indicated by EFA and DCT matches the traditional ceramic typology, which supports that claim that outline-based approaches can be efficiently used for studying variations in ceramic shapes.
+Furthermore, @Wilczek2014's EFA results help us understand the level of production standardization over time across the region.
+
 EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions [@Adams2004; @Bonhomme2014; @Claude2008].

### 2020-07-10 analysis/paper/paper.Rmd
working on GMM background and added references

@@
 Landmark GM approaches assign a set of landmarks and/or semilandmarks onto objects as reference points that can be specified on a coordinate system.
-Landmark-based morphometrics are widely applied to archaeological objects with obvious morphological features that provide unambiguous reference points for landmark placement, such as projectile point tips or visually distinctive osteological features [@Birch2019; @Buchanan2019; @Cardillo2010; @Haruda2019; @Lycett2013; @Meloro2015; @Selden2019].
+Landmark-based morphometrics are widely applied to archaeological objects with obvious morphological features that provide unambiguous reference points for landmark placement, such as projectile point tips, visually distinctive osteological features, or ceramic assemblages with clear distinct components [@Birch2019; @Buchanan2019; @Cardillo2010; @Haruda2019; @Lycett2013; @Meloro2015; @Selden2018; @Selden2019].
+
+Geometric morphometrics methods can provide a useful framework for statistical evaluation on the standardization of archaeological assemblages across geographical or chronological divisions.
+The shape variables generated by GMM could be further analyzed by Principle Component Analysis (PCA) to identify any decreases or increases in the variation of a certain group of shapes, such as points shapes, to discuss the standardization reflecting the selection and reduction intensity [@Archer2015], or specialized function [@Buchanan2018; @Doyon2019].
+
+<!--discussion of standardization in material culture using GM approaches-->
 
@@
 We use a significance test for the equality of coefficient of variations of shape variables to statistically compare the vessel standardization from different periods.
-
-<!--discussion of standardization in material culture using GM approaches-->
 

### 2020-07-17 analysis/paper/paper.Rmd
work on GM background, and add the metric analysis back

--- 2020-07-03 analysis/paper/paper.Rmd
@@
 Landmark GM approaches assign a set of landmarks and/or semilandmarks onto objects as reference points that can be specified on a coordinate system.
-Landmark-based morphometrics are widely applied to archaeological objects with obvious morphological features that provide unambiguous reference points for landmark placement, such as projectile point tips or visually distinctive osteological features [@Birch2019; @Buchanan2019; @Cardillo2010; @Haruda2019; @Lycett2013; @Meloro2015; @Selden2019].
+Landmark-based morphometrics are widely applied to archaeological objects with obvious morphological features that provide unambiguous reference points for landmark placement, such as tips and edges of stone tools or iron weapons[@Birch2019; @Buchanan2019; @Cardillo2010; @Lycett2013; @Selden2018], visually distinctive bone features [@Haruda2019; @Meloro2015], or ceramic assemblages with clear distinct components [@Selden2019; @Topi2017].
+
+Geometric morphometrics methods can provide a useful framework for further statistical evaluation on shape variability of archaeological assemblages to explore how they change over time or across geographic areas to answer questions related to variation in morphology and standardization.
+A common procedure is using multivariate statistical methods, such as principal component analysis or canonical variate analysis, to capture the important variances of the overall variation in shape variables generated by GM and then visualize the variances to identify any clustered analytical groups that suggest less variability or high standardization.
+For example, the application of GM to stone tools coupled with statistical tests, such as coefficient of variation on principal component scores [@Archer2015], and on a series of metric variables including length, width, or thickness [@Buchanan2018; @Okumura2014; @Smith2016], or Kruskal-Wallis test on flake scar patterns [@Eren2015], which allows further discussion of causes of standardization and variability that may relate to degree of tool reduction, specialized function, or lack of innovation in a small group.
+In addition, cluster analysis such as k-mean aggregation for shape variables generated by GM is another way to discuss the variability among shape groups [@Doyon2019].
+
+<!--discussion of standardization in material culture using GM approaches-->
 
@@
 Most applications are associated with human remains and animal bones to study biological variation, and has been increasing for stone artifacts to explore stone tool topology and technology [@Fox2015; @Hoggard2019; @Iovictua2010].
-Despite the EFA applications for ceramics is rare, they demonstrate the potential of this promising approach for answering questions related to ceramic taxonomy [@Topi2018; @Wilczek2014].
+Despite the EFA applications for ceramics is rare, this approach is promising for answering questions related to ceramic taxonomy [@Topi2017; @Wilczek2014].
 For example, @Wilczek2014 evaluate the concordance between EFA and Discrete Cosine Transform (DCT), and a traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France.
@@
 
-The first two principal components (PCs) of the PCA on the elliptic Fourier coefficients explain `r vari_two`% of the total variance, of which `r pc1`% is explained by the first principle component.
+The first two principal components (PCs) of the PCA on the elliptic Fourier coefficients explain `r vari_two`% of the total variance, of which `r pc1`% is explained by the first principal component.
 With the third component, the first three principal components explain `r vari_three`% of the total variance.
@@
 
+We compare the result of shape variable with the result of metric data for six measurements and two ratio values using CV test (Table \@ref(tab:metricCVstest)).
+Metric measurements show there is a significant difference (p-value less than 0.01) between CV values for two metric measurements across phases, the rim diameter and the neck diameter before and after the European presence.
+After the arrival of the Chinese, there is only one metric attribute, the rim diameter, showing significant difference compared with post-European period (p-value less than 0.05).
+Although these results show the differences between different contact phases that corresponds to part of the results of shape variable, elliptic Fourier analysis captures more subtle differences in a substantial way reflected by overall shapes of pottery, not only limited to diameter variables.
+In addition, the result of spatial analysis (Figure \@ref(fig:kernel-plot)) for pottery shape samples presents multiple clusters with high density of pottery during the European presence.
+However, the hypothesis testing on spatial randomness indicates a non-randomly dispersed distribution before European contact and more extreme dispersed after European presence.
+In contrast, the distribution of pottery is more similar to the random distributions during the Chinese period.
+This contradicts our expectation that clustered pattern will be observed with an increase in pottery standardization since the emergence of specialized groups.
+Or, this may indicate that there are other factors lead to pottery shape standardization instead of a small number of artisans.
+
+```{r metricCVstest}
+# select metric attributes and tidy data for computing CV
+kwl_p_metrics_long <-
+  kwl_p_ave %>%
+  mutate(dia_rim_body = Dia_Rim_ave/Dia_Body_ave,
+         thi_rim_body = Thick_Rim_ave/Thick_Body_ave) %>%
+  select(period, 
+         Dia_Rim_ave, Dia_Neck_ave, 
+         Dia_Body_ave, Thick_Rim_ave, 
+         Thick_Neck_ave, Thick_Body_ave,
+         dia_rim_body, thi_rim_body,) %>%
+  rename(`Rim diameter (mm)` = Dia_Rim_ave,
+         `Neck diameter (mm)` = Dia_Neck_ave,
+         `Body diameter (mm)` = Dia_Body_ave,
+         `Rim thickness (mm)` = Thick_Rim_ave,
+         `Neck thickness (mm)` = Thick_Neck_ave,
+         `Body thickness (mm)` = Thick_Body_ave,
+         `Ratio of Rim/Body diameter` = dia_rim_body,
+         `Ratio of Rim/Body thickness` = thi_rim_body) %>%
+  gather(variable, value, -period) %>%
+  filter(!is.na(period), !is.na(value)) %>% 
+  mutate(variable = factor(variable, levels = c("Rim thickness (mm)",
+                                                "Neck thickness (mm)",
+                                                "Body thickness (mm)",
+                                                "Ratio of Rim/Body thickness",
+                                                "Rim diameter (mm)",
+                                                "Neck diameter (mm)",
+                                                "Body diameter (mm)",
+                                                "Ratio of Rim/Body diameter"))) %>%
+  arrange(variable)
+# filter pre-e and post-e periods
+kwl_p_metrics_long_pre_post <-
+  kwl_p_metrics_long %>%
+  filter(!period == "ch-con")
+# filter post-e and chi periods
+kwl_p_metrics_long_post_chi <-
+  kwl_p_metrics_long %>%
+  filter(!period == "pre-e")
+#---------------------------CV test for pre-e and post-e pair
+kiwulan_pottery_metrics_cvs_plot_cv_test_asym_pre_post <-
+  kwl_p_metrics_long_pre_post %>%
+  nest(-variable) %>%  # or nest(period, value) %>%
+  mutate(asymptotic_test = map(data, # change from mslr_test to asymptotic_test
+                               ~bind_cols(asymptotic_test (
+                                 .x$value,
+                                 .x$period)))) %>%
+  unnest(asymptotic_test)
+# compute CV
+kwl_p_cvs_pre_post <-
+  kwl_p_metrics_long_pre_post %>%
+  group_by(period, variable) %>%
+  summarise(cvs = raster::cv(value, na.rm = TRUE))
+# join p-values to CVs
+kwl_p_cvs_and_pvalues_pre_post <-
+  kwl_p_cvs_pre_post %>%
+  left_join(kiwulan_pottery_metrics_cvs_plot_cv_test_asym_pre_post) %>%
+  mutate(significant = ifelse(p_value <= 0.05, 'yes', 'no')) %>%
+  mutate(variable = factor(variable, ordered = TRUE)) %>%
+  arrange(variable)
+# making table for CV
+kwl_p_cvs_table_pre_post <-
+  kwl_p_cvs_and_pvalues_pre_post %>%
+  select(-data, -significant) %>%
+  pivot_wider(names_from = period, values_from = cvs) %>%
+  select(variable, `pre-e`, `post-e`, `D_AD`, `p_value`) %>%
+  mutate_at(c("pre-e", "post-e", "D_AD", "p_value"), round, 4)
+#---------------------------CV test for post-e and ch-con pair
+kiwulan_pottery_metrics_cvs_plot_cv_test_asym_post_chi <-
+  kwl_p_metrics_long_post_chi %>%
+  nest(-variable) %>%  # or nest(period, value) %>%
+  mutate(asymptotic_test = map(data, # change from mslr_test to asymptotic_test
+                               ~bind_cols(asymptotic_test (
+                                 .x$value,
+                                 .x$period)))) %>%
+  unnest(asymptotic_test)
+# compute CV
+kwl_p_cvs_pre_post <-
+  kwl_p_metrics_long_post_chi %>%
+  group_by(period, variable) %>%
+  summarise(cvs = raster::cv(value, na.rm = TRUE))
+# join p-values to CVs
+kwl_p_cvs_and_pvalues_post_chi <-
+  kwl_p_cvs_pre_post %>%
+  left_join(kiwulan_pottery_metrics_cvs_plot_cv_test_asym_post_chi) %>%
+  mutate(significant = ifelse(p_value <= 0.05, 'yes', 'no')) %>%
+  mutate(variable = factor(variable, ordered = TRUE)) %>%
+  arrange(variable)
+# making table for CV
+kwl_p_cvs_table_post_chi <-
+  kwl_p_cvs_and_pvalues_post_chi %>%
+  select(-data, -significant) %>%
+  pivot_wider(names_from = period, values_from = cvs) %>%
+  select(variable, `post-e`, `ch-con`, `D_AD`, `p_value`) %>%
+  mutate_at(c("post-e", "ch-con", "D_AD", "p_value"), round, 4)
+# join two table together
+kwl_p_cvs_table_all <-
+  kwl_p_cvs_table_pre_post %>% 
+  left_join(kwl_p_cvs_table_post_chi, "variable") %>% 
+  select(-`post-e.y`) %>% 
+  rename(`before European` = "pre-e", `European` = "post-e.x", Chinese = "ch-con",
+         `D_AD-1` = "D_AD.x", `D_AD-2` = "D_AD.y",
+         `P value-1` = `p_value.x`, `P value-2` = `p_value.y`) %>% 
+  select(variable, `before European`, European, Chinese,
+         `D_AD-1`, `P value-1`, `D_AD-2`, `P value-2`)
+knitr::kable(kwl_p_cvs_table_all,
+             caption = "Coefficient of Variation for metric attributes by three phases using Asymptotic test [@Feltz1996].
+The first set of P-value represents the comparison between before European and European periods, while the second represents the comparison between European and Chinese periods ")
+```
+
 # Discussion

### 2020-07-21 analysis/paper/paper.Rmd
udpate on method section

--- 2020-07-03 analysis/paper/paper.Rmd
@@
 However, because pottery vessels typically have curved shapes, linear measurements have limited sensitivity to many kinds of shape variations.
-Thus, to capture subtle shape variations that might also be relevant to standardization, we analyze ceramic shapes using geometric morphometric (GM) methods.
+Thus, to capture subtle shape variations that might also be relevant to standardization, we analyze ceramic shapes using geometric morphometric methods (GMM).
 
@@
 
-Geometric morphometrics differs from traditional linear measurements through its use of Cartesian coordinates of morphological structures to define and analyze shape [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007].
+Geometric morphometrics differs from traditional linear measurements through its use of Cartesian coordinates of morphological structures to quantify and analyze shape [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007].
 Landmarks, curves or outlines of objects can be represented by coordinates in terms of their unique point locations with respect to numerical values on coordinate axes.
 There are two common morphometric methods: landmark and outline approaches [@Adams2004].
-Landmark GM approaches assign a set of landmarks and/or semilandmarks onto objects as reference points that can be specified on a coordinate system.
-Landmark-based morphometrics are widely applied to archaeological objects with obvious morphological features that provide unambiguous reference points for landmark placement, such as projectile point tips or visually distinctive osteological features [@Birch2019; @Buchanan2019; @Cardillo2010; @Haruda2019; @Lycett2013; @Meloro2015; @Selden2019].
+Landmark GMM approaches assign a set of landmarks and/or semilandmarks onto objects as reference points that can be specified on a coordinate system.
+Landmark-based morphometrics are widely applied to archaeological objects with obvious morphological features that provide unambiguous reference points for landmark placement, such as tips and edges of stone tools or iron weapons[@Birch2019; @Buchanan2019; @Cardillo2010; @Lycett2013; @Selden2018], visually distinctive bone features [@Haruda2019; @Meloro2015], or ceramic assemblages with clear distinct components [@Selden2019; @Topi2017].
+The applications of GMM couple with multivariate statistics for shape variables of artifacts are commonly associated with topics of topological evolution or variations in assemblages to identify groups [@Doyon2019; @Eren2015; @Selden2018].
+
+In addition to vibrations, shape standardization of artifacts and its causes is another topic.
+A common procedure is using dimensional reduction techniques, such as principal components analysis or canonical variate analysis, to capture the essential features that can represent the overall shape.
+This enables the identification of clustered analytical groups through visualization for the results, followed by statistical tests on the shape variables.
+In this way, we can explore whether there is less variations or high standardization in the same assemblage over time or between assemblages over larger geographical areas.
+For example, @Archer2015's case study of Still Bay points in Southern Africa suggests an increase in point shape standardization over time that may relate to a maintenance of finished points.
+For comparison between several sites, @Buchanan2018 identified a more uniform base-shape of Folsom points than Clovis points across the western US, which indicates the specific function for hafting.
+Similarly, @Smith2016 suggested that the humongous fluted point base in Alaska and northern Yukon is a risk management strategy to ensure the ease of replacement during long-distance travel.
+There are also other factors could lead to an increase in standardization of point shapes such as low level of cultural innovation in a small group [@Okumura2014].
 
 However, a key limitation of the landmark approach is that landmarks may be difficult to reproducibly locate for structures that are mostly or entirely curves if not mathematically-defined.
-In those cases, Elliptic Fourier Analysis (EFA) is an excellent mean for assessing morphological variation in whole structure of two-dimensional closed shapes.
-EFA was first introduced to study shapes of artifacts in archaeology in 1980s [@Gero1984].
-Most applications are associated with human remains and animal bones to study biological variation, and has been increasing for stone artifacts to explore stone tool topology and technology [@Fox2015; @Hoggard2019; @Iovictua2010].
-Despite the EFA applications for ceramics is rare, they demonstrate the potential of this promising approach for answering questions related to ceramic taxonomy [@Topi2018; @Wilczek2014].
-For example, @Wilczek2014 evaluate the concordance between EFA and Discrete Cosine Transform (DCT), and a traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France.
-The results show the variation indicated by EFA and DCT matches the traditional ceramic typology, which supports that claim that outline-based approaches can be efficiently used for studying variations in ceramic shapes.
-Furthermore, @Wilczek2014's EFA results help us understand the level of production standardization over time across the region.
-
+In those cases, outline approach, Elliptic Fourier Analysis (EFA), is a better mean for assessing morphological variation in whole structure of two-dimensional closed shapes.
 EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions [@Adams2004; @Bonhomme2014; @Claude2008].
@@
 The harmonic power, a cumulative sum of squared harmonic coefficients, provides a robust rule for determining the desired number of harmonics [@Bonhomme2014].
+EFA was first introduced to study shapes of artifacts in archaeology in 1980s [@Gero1984], and has been increasing applied to different archaeological materials.
+Most applications are associated with human remains and animal bones to examine biological variation, and also stone artifacts to explore stone tool topology and technological evolution [@Fox2015; @Hoggard2019; @Iovictua2010].
+
+Despite the EFA applications for ceramics is rare, this approach is promising for answering questions related to ceramic taxonomy [@Topi2017; @Wilczek2014].
+For example, @Wilczek2014 evaluate the concordance between EFA and Discrete Cosine Transform (DCT), and a traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France.
+The results show the variation indicated by EFA and DCT matches the traditional ceramic typology, which supports that claim that outline-based approaches can be efficiently used for studying variations in ceramic shapes.
+Furthermore, @Wilczek2014's EFA results help us understand the level of production standardization over time across the region.
 EFA is suitable for analyzing shapes lacking representative landmarks, or where curves contain the most meaningful variation.

### 2020-07-22 analysis/paper/paper.Rmd
added references and finished tthe review of GMM studies

@@
 Landmark GMM approaches assign a set of landmarks and/or semilandmarks onto objects as reference points that can be specified on a coordinate system.
-Landmark-based morphometrics are widely applied to archaeological objects with obvious morphological features that provide unambiguous reference points for landmark placement, such as tips and edges of stone tools or iron weapons[@Birch2019; @Buchanan2019; @Cardillo2010; @Lycett2013; @Selden2018], visually distinctive bone features [@Haruda2019; @Meloro2015], or ceramic assemblages with clear distinct components [@Selden2019; @Topi2017].
-The applications of GMM couple with multivariate statistics for shape variables of artifacts are commonly associated with topics of topological evolution or variations in assemblages to identify groups [@Doyon2019; @Eren2015; @Selden2018].
+Landmark-based morphometrics are widely applied to archaeological objects with obvious morphological features that provide unambiguous reference points for landmark placement, such as tips and edges of stone tools or iron weapons [@Birch2019; @Lycett2013], visually distinctive bone features [@Haruda2019; @Meloro2015], or ceramic assemblages with clear distinct components [@Selden2019; @Topi2017].
+With univariate or multivariate statistics for shape data generated by GMM, the application to artifacts can answer research questions related to topological evolution or variations in assemblages to define groups [@Doyon2019; @Eren2015; @Selden2018lithic; @Perez2007; @Presnyakova].
+A common procedure is using dimensional reduction techniques, such as principal components analysis or canonical variate analysis, to capture the essential features that can represent the overall shape.
+This enables the identification of clustered analytical groups through the visualization of results, followed by statistical tests to robustly distinguish them.
 
-In addition to vibrations, shape standardization of artifacts and its causes is another topic.
-A common procedure is using dimensional reduction techniques, such as principal components analysis or canonical variate analysis, to capture the essential features that can represent the overall shape.
-This enables the identification of clustered analytical groups through visualization for the results, followed by statistical tests on the shape variables.
-In this way, we can explore whether there is less variations or high standardization in the same assemblage over time or between assemblages over larger geographical areas.
-For example, @Archer2015's case study of Still Bay points in Southern Africa suggests an increase in point shape standardization over time that may relate to a maintenance of finished points.
-For comparison between several sites, @Buchanan2018 identified a more uniform base-shape of Folsom points than Clovis points across the western US, which indicates the specific function for hafting.
+One of the interesting topics is to explore whether and why there is shape standardization over time, or between assemblages across geographical areas.
+There is many discussion about lithic technology using multivariate analysis on shape variables or coefficient of variations on associated metric data.
+For example, @Archer2015's case study of stone points in Southern Africa suggests an increase in shape standardization over time that may relate to a maintenance of finished points.
+With considerations of a set of metric data, @Buchanan2018; @Buchanan2019 identified a more uniform base-shape of Folsom points than Clovis points across the western US, which indicates the specific function for hafting.
 Similarly, @Smith2016 suggested that the humongous fluted point base in Alaska and northern Yukon is a risk management strategy to ensure the ease of replacement during long-distance travel.
-There are also other factors could lead to an increase in standardization of point shapes such as low level of cultural innovation in a small group [@Okumura2014].
+Other factor, such as low level of cultural innovation in a small group could also lead to an increase in standardization of point shapes [@Okumura2014].
+For ceramics, @Topi2017 found a standardized vessel shape from the Casas Grandes culture in northwest Mexico using coefficient of variations for the positions of semi-landmarks across shape groups.
+They suggest the standardization might indicate the pretense of specialized producers reflecting social complexity.
 
-However, a key limitation of the landmark approach is that landmarks may be difficult to reproducibly locate for structures that are mostly or entirely curves if not mathematically-defined.
-In those cases, outline approach, Elliptic Fourier Analysis (EFA), is a better mean for assessing morphological variation in whole structure of two-dimensional closed shapes.
+Another way to examine standardization is using morphological disparity to perform pairwise comparison of correlations between shape groups [@Wills2001].
+In this manner, @Selden2018ceramic2 examined Caddo ceramics in northeast Texas region and found a gradual shape standardization over time that provides a basis for further discussion of craft specialization or group identity.
+Similar applications to bifaces in central Texas presented the evidence of grater diversity in morphology [@Selden2020].
+More other applications, such as cranial deformation, demonstrate that landmark approach with multivariate analyses on shape variances or symmetry is an effective method to evaluate shape standardization [@Kuzminsky2016; @Natahi2019; @Perez2007]
+
+Despite the wide use of landmark approach in archaeology, a key limitation is that landmarks may be difficult to reproducibly locate for structures that are mostly or entirely curves if not mathematically-defined.
+In those cases, outline approach, such as Elliptic Fourier Analysis (EFA), is a better mean for assessing morphological variation in whole structure of two-dimensional closed shapes [@Cardillo2010].
 EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions [@Adams2004; @Bonhomme2014; @Claude2008].
@@
 The harmonic power, a cumulative sum of squared harmonic coefficients, provides a robust rule for determining the desired number of harmonics [@Bonhomme2014].
-EFA was first introduced to study shapes of artifacts in archaeology in 1980s [@Gero1984], and has been increasing applied to different archaeological materials.
-Most applications are associated with human remains and animal bones to examine biological variation, and also stone artifacts to explore stone tool topology and technological evolution [@Fox2015; @Hoggard2019; @Iovictua2010].
+The first systematic use of EFA in archaeology was @Gero1984's study of lithic flakes in Peru that demonstrated the changes in forms from a more angular to rounded shape over time.
+Since then, EFA has been applied to different archaeological materials.
+Similar to landmark approach, most applications are associated with human remains and animal bones to examine biological variation, and stone artifacts to explore topology and technological evolution [@Fox2015; @Hoggard2019; @Iovictua2010].
 
-Despite the EFA applications for ceramics is rare, this approach is promising for answering questions related to ceramic taxonomy [@Topi2017; @Wilczek2014].
+The EFA applications for ceramics is rare, however, this approach is promising for answering questions related to ceramic taxonomy and standardization.
 For example, @Wilczek2014 evaluate the concordance between EFA and Discrete Cosine Transform (DCT), and a traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France.

### 2020-07-23 analysis/paper/paper.Rmd
added linear regression model for size

@@
 One of the interesting topics is to explore whether and why there is shape standardization over time, or between assemblages across geographical areas.
-There is many discussion about lithic technology using multivariate analysis on shape variables or coefficient of variations on associated metric data.
+It is common to use multivariate analysis on shape variables or coefficient of variations on associated metric data to discuss standardization, especially for lithic assemblages.
 For example, @Archer2015's case study of stone points in Southern Africa suggests an increase in shape standardization over time that may relate to a maintenance of finished points.
@@
 They suggest the standardization might indicate the pretense of specialized producers reflecting social complexity.
-
 Another way to examine standardization is using morphological disparity to perform pairwise comparison of correlations between shape groups [@Wills2001].
-In this manner, @Selden2018ceramic2 examined Caddo ceramics in northeast Texas region and found a gradual shape standardization over time that provides a basis for further discussion of craft specialization or group identity.
+In this manner, @Selden2018ceramic2; @Selden2019 examined Caddo ceramics in northeast Texas region and found a gradual shape standardization over time that provides a basis for further discussion of craft specialization or group identity.
 Similar applications to bifaces in central Texas presented the evidence of grater diversity in morphology [@Selden2020].
-More other applications, such as cranial deformation, demonstrate that landmark approach with multivariate analyses on shape variances or symmetry is an effective method to evaluate shape standardization [@Kuzminsky2016; @Natahi2019; @Perez2007]
+More other applications, such as cranial deformation, demonstrate that landmark approach with multivariate analyses on shape variances or symmetry is an effective method to evaluate shape standardization [@Kuzminsky2016; @Natahi2019; @Perez2007].
 
@@
 The harmonic power, a cumulative sum of squared harmonic coefficients, provides a robust rule for determining the desired number of harmonics [@Bonhomme2014].
-The first systematic use of EFA in archaeology was @Gero1984's study of lithic flakes in Peru that demonstrated the changes in forms from a more angular to rounded shape over time.
-Since then, EFA has been applied to different archaeological materials.
-Similar to landmark approach, most applications are associated with human remains and animal bones to examine biological variation, and stone artifacts to explore topology and technological evolution [@Fox2015; @Hoggard2019; @Iovictua2010].
+The first systematic use of EFA in archaeology was @Gero1984's study of lithic flakes in Peru that suggests the changes in forms from a more angular to rounded shape over time.
+Later, @Saragusti2005 introduced several functions for the further calculation of the specific shape attributes, including symmetry, roughness, and deformation, which demonstrates the potential for analysis of curves.
+To compare resharpening trajectories in the European Middle Paleolithic stone tools, @Iovictua2009 presented a protocol from outline digitization to EFA, with considerations of size using multivariate linear regression.
+More recent applications to stone artifacts seek to answer questions about topological classification of Late Woodland points [@Fox2015], cultural taxonomies for the European Late Palaeolithic [@Ivanovaite2020], and shape and symmetry standardization of British Acheulean [@Hoggard2019].
 
@@
 
+```{r size-regression}
+# calculate centroid size
+jpgs_imported_coo_size <-
+  jpgs_imported_coo %>% 
+  mutate(size =  coo_centsize(jpgs_imported_coo))
+
+# create pcs dataframe for joining
+pc1_df_with_filename <-
+  bind_cols(filename = rownames(pc1), pc1) %>% 
+  rename(c("pc1" = "x")) %>% 
+  mutate(filename = parse_number(filename))
+
+pc2_df_with_filename <-
+  bind_cols(filename = rownames(pc2), pc2) %>% 
+  rename(c("pc2" = "x")) %>% 
+  mutate(filename = parse_number(filename))
+
+# make a dataframe for centroid size and join with pc1
+jpgs_imported_coo_size_df <-
+  as.data.frame(jpgs_imported_coo_size$fac) %>% 
+  mutate(filename = parse_number(as.character(filename))) %>% 
+  left_join(pc1_df_with_filename) %>% 
+  left_join(pc2_df_with_filename)
+
+# size boxplot by period
+ggplot(jpgs_imported_coo_size_df,
+       aes(size, fct_rev(phase))) +
+  coord_flip() +
+  geom_boxplot() +
+  labs(x = "phase",
+       y = "centroid size")
+
+# build regression model
+linearMod <- lm( size ~ pc1, data = jpgs_imported_coo_size_df)
+
+ggplot(jpgs_imported_coo_size_df, 
+       aes(size, pc1)) +
+  geom_point() +
+  geom_smooth(method = "lm") +
+  facet_wrap(~ fct_rev(phase)) +
+  labs(x = "centroid size")
+```
+
 # Discussion

### 2020-07-23 analysis/paper/paper.Rmd
update regression plot for size

@@
 
-```{r size-regression}
-# calculate centroid size
-jpgs_imported_coo_size <-
-  jpgs_imported_coo %>% 
-  mutate(size =  coo_centsize(jpgs_imported_coo))
+```{r size-regression, fig.cap= "Measure "}
+library(ggpubr)
+library(geomorph)
 
-# create pcs dataframe for joining
+jpgs_imported_coo_size <- coo_centsize(jpgs_imported_coo)
+jpgs_imported_coo_size_df <- as.data.frame(jpgs_imported_coo_size) 
+jpgs_imported_coo_size_df <-
+  bind_cols(filename = rownames(jpgs_imported_coo_size_df), 
+            jpgs_imported_coo_size_df) %>% 
+  mutate(filename = parse_number(filename))
+
+# prepare metric data for joining
+pottery_data_tidy <-
+  pottery_data %>%
+  mutate_at(vars(starts_with(c("Thick", "Diameter"))), as.numeric) %>% 
+  mutate(Height = as.numeric(Hight)) %>% 
+  mutate(filename = parse_number(filename))
+
+# prepare pcs dataframe for joining
 pc1_df_with_filename <-
@@
 # make a dataframe for centroid size and join with pc1
-jpgs_imported_coo_size_df <-
-  as.data.frame(jpgs_imported_coo_size$fac) %>% 
+jpgs_imported_coo_size_pcs_metric <-
+  jpgs_imported_coo_size_df %>% 
   mutate(filename = parse_number(as.character(filename))) %>% 
   left_join(pc1_df_with_filename) %>% 
-  left_join(pc2_df_with_filename)
+  left_join(pc2_df_with_filename) %>% 
+  left_join(pottery_data_tidy)
 
 # size boxplot by period
-ggplot(jpgs_imported_coo_size_df,
-       aes(size, fct_rev(phase))) +
+boxplot_size <-
+  ggplot(jpgs_imported_coo_size_pcs_metric,
+         aes(Diameter1_Body, fct_rev(y))) +
+  geom_boxplot() +
   coord_flip() +
-  geom_boxplot() +
-  labs(x = "phase",
-       y = "centroid size")
+  labs(x = "body diameter (mm)", y = "") +
+  scale_y_discrete(breaks = c("Pre-European", "Post-European", "Chinese"),
+                   labels = c("Pre-\nEuropean", "Post-\nEuropean", "Chinese")) +
+  theme(axis.text = element_text(size = 8))
 
-# build regression model
-linearMod <- lm( size ~ pc1, data = jpgs_imported_coo_size_df)
+# can pc1 predict size by period?
+pc1_size_reg <- 
+  ggscatter(jpgs_imported_coo_size_pcs_metric, 
+            x = "Diameter1_Body", 
+            y = "pc1",
+            add = "reg.line",
+            size = 1) +
+    facet_wrap(~ fct_rev(y)) +
+    stat_cor(label.y = 1.25, label.sep = "\n") +
+    #stat_regline_equation(label.y = 1) +
+    labs(x = "") +
+    theme(axis.text = element_text(size = 8)) +
+    #scale_x_continuous(breaks = c(75, 200)) +
+    scale_y_continuous(limits = c(0, 1.5),
+                       breaks = c(seq(0, 1, 0.5)),
+                       labels = c(seq(0, 1, 0.5)))
 
-ggplot(jpgs_imported_coo_size_df, 
-       aes(size, pc1)) +
-  geom_point() +
-  geom_smooth(method = "lm") +
-  facet_wrap(~ fct_rev(phase)) +
-  labs(x = "centroid size")
+# can pc1 predict size by period?
+pc2_size_reg <- 
+  ggscatter(jpgs_imported_coo_size_pcs_metric, 
+            x = "Diameter1_Body", 
+            y = "pc2",
+            add = "reg.line",
+            size = 1) +
+    facet_wrap(~ fct_rev(y)) +
+    stat_cor(label.y = 1.25, label.sep = "\n") +
+    #stat_regline_equation(label.y = 1) +
+    labs(x = "body diameter (mm)") +
+    theme(axis.text = element_text(size = 8)) +
+    #scale_x_continuous(breaks = c(75, 200)) +
+    scale_y_continuous(limits = c(0, 1.5),
+                       breaks = c(seq(0, 1, 0.5)),
+                       labels = c(seq(0, 1, 0.5)))
+
+# plot boxplot and regression together 
+right_column <-
+  plot_grid(pc1_size_reg, pc2_size_reg, ncol =1)
+
+size_boxplot_regression <-
+  plot_grid(boxplot_size,
+            right_column,
+            labels = c("A", "B"),
+            label_size = 12,
+            rel_widths = c(1.8, 3.7))
+
+ggsave(plot = size_boxplot_regression,
+       here("analysis", "figures",
+            "size-boxplot-regression.jpg"),
+       width = 180, 
+       height = 100,
+       dpi = 300,
+       units = "mm")
 ```

### 2020-07-23 analysis/paper/size_correlation .R

@@ -1,0 +1,9 @@
+pottery_data <- readr::read_csv(here("analysis/data/raw_data/KWL_Ceramic_final2018_with_phases.csv"))
+
+pottery_data_tidy <-
+  pottery_data %>%
+  mutate_at(vars(starts_with(c("Thick", "Diameter"))), as.numeric) %>%
+  mutate(Height = as.numeric(Hight))
+
+cor(pottery_data_tidy$Height,
+    pottery_data_tidy$Diameter1_Body, use = "pairwise.complete.obs")

### 2020-07-24 analysis/paper/paper.Rmd
added para about size analysis

@@
 
-```{r size-regression, fig.cap= "Measure "}
+<!-- size analysis -->
+
+We also conducted size analysis by examining the body diameter of the vessels to explore any differences in vessel form reflected by size and shape.
+Body diameter is a suitable metric measurement to represent size for ceramics since it is associated with volume as a utilitarian function, and could present differences in the situation of similar shape.
+The body diameter of the vessels from Chinese period is larger than those from the two earlier periods, and the vessels before European contact have the smallest body diameter on average (Figure \@ref(fig:size-regression): A).
+It is interesting that linear regression model (Figure \@ref(fig:size-regression): B) predicts that shape, captured by PC1 and PC2, changes with size across three phases, except PC1 in Chinese period.
+That is, the shorter vessels are larger in body diameter according to the significant positive correlation between PC1 and size; however, the vessel from Chinese period does not show this pattern.
+On the contrary, the negative relationship between PC2 and size suggests that the vessels with narrower neck and mouth tend to have a larger body diameter.
+
+```{r size-regression, fig.cap= "A: Distribution of body diamater of vessels across three phases.
+B: Scatter plots of shape variable, pc1 at the top and pc2 at the bottom, as a function of body dimater"}
 library(ggpubr)
@@
             x = "Diameter1_Body", 
-            y = "pc1",
+            y = "PC1",
             add = "reg.line",
@@
 
-# can pc1 predict size by period?
+# can pc2 predict size by period?
 pc2_size_reg <- 
@@
             x = "Diameter1_Body", 
-            y = "pc2",
+            y = "PC2",
             add = "reg.line",
@@
 In general, vessels become shorter in height over time, leading to an oval-shaped body.
+Moreover, size significantly varies with shape for vessels before and after the European presence, which presents that shorter vessels with narrower opening tend to have a larger body diameter.
+This correlation between narrower opening and larger diameter can be also observed during the Chinese period without the effect of height.
 
@@
 We can thus rule out changes in clay fabric as a factor in explaining changes in vessel shape.
-To determine if shape changes might be related to changes in the function of pots at Kiwulan, we used geochemical methods to extract and identify lipids trapped in the fabric of potsherds to identify foods that may have contributed residues absorbed into the clay [cf.
+To determine if form changes might be related to changes in the function of pots at Kiwulan, we used geochemical methods to extract and identify lipids trapped in the fabric of potsherds to identify foods that may have contributed residues absorbed into the clay [cf.
 @Kwak2015].

### 2020-07-24 analysis/paper/size_correlation .R

@@ -5,5 +5,16 @@ pottery_data_tidy <-
   mutate_at(vars(starts_with(c("Thick", "Diameter"))), as.numeric) %>%
   mutate(Height = as.numeric(Hight))
 
+# correlation between height and body diameter
 cor(pottery_data_tidy$Height,
     pottery_data_tidy$Diameter1_Body, use = "pairwise.complete.obs")
+
+# run the code in the paper rmd first
+# does the linear regression model work well for all data? Yes
+linearMod <- lm(pc1 ~ Diameter1_Body,
+                data =jpgs_imported_coo_size_pcs_metric)
+plot(linearMod)
+summary(linearMod)
+
+# are residuals normally distributed? Yes
+qqnorm(jpgs_imported_coo_size_pcs_metric$Diameter1_Body)

### 2020-07-24 analysis/paper/paper.Rmd
update description file

@@
 
-```{r size-regression, fig.cap= "A: Distribution of body diamater of vessels across three phases.
-B: Scatter plots of shape variable, pc1 at the top and pc2 at the bottom, as a function of body dimater"}
+```{r size-regression, fig.cap= "A: Distribution of the body diameter of the vessels across three phases.
+B: Scatter plots of shape variable, PC1 at the top and PC2 at the bottom, as a function of body diamater"}
+
 library(ggpubr)
@@
 Moreover, size significantly varies with shape for vessels before and after the European presence, which presents that shorter vessels with narrower opening tend to have a larger body diameter.
-This correlation between narrower opening and larger diameter can be also observed during the Chinese period without the effect of height.
+This correlation between size and shape can be also observed during the Chinese period without the consideration of height.
 

### 2020-07-24 analysis/paper/size_correlation .R

@@ -18,3 +18,35 @@ summary(linearMod)
 
 # are residuals normally distributed? Yes
 qqnorm(jpgs_imported_coo_size_pcs_metric$Diameter1_Body)
+
+# assign landmarks and GPA
+jpgs_imported_coo_land <-
+  jpgs_imported %>%
+  Out(., fac = phases_isect)  %>%
+  # Centring, scaling and sampling pseudo-landmarks
+  coo_center %>%
+  coo_scale %>%
+  coo_sample(., 50) %>%
+  # Procrustes superimposition for pseudo-landmarks
+  fgProcrustes
+
+# geomorph data frame
+geomorph_df <- geomorph.data.frame(shape = jpgs_imported_coo_land$coo,
+                                   size = jpgs_imported_coo_size_pcs_metric$jpgs_imported_coo_size,
+                                   phase = jpgs_imported_coo_size_pcs_metric$y)
+
+# modeling shape according
+fit.shapephase <-
+  procD.lm(shape ~ phase,
+           data = geomorph_df,
+           print.progress = FALSE,
+           iter = 9999)
+
+# morphological disparity
+morphol.disparity(fit.shapephase,
+                  groups = jpgs_imported_coo_size_pcs_metric$y,
+                  data = geomorph_df,
+                  print.progress = FALSE,
+                  iter = 9999)
+
+

### 2020-07-24 analysis/paper/paper.Rmd
added sentences about keywords in GMM in conclusion

@@
 Kiwulan is situated on a hill near a riverside at the northern margin of Yilan County, which is characterized by a triangular alluvial plain facing east toward the Pacific with high mountains on three other sides.
-
 This research investigates if there was increasing ceramic specialization resulting from Indigenous interaction with Europeans in the 17th century or Chinese in the 19th century, two major foreign influences in early historical Taiwan, which may indicate social changes in local Indigenous societies.
@@
 Landmark-based morphometrics are widely applied to archaeological objects with obvious morphological features that provide unambiguous reference points for landmark placement, such as tips and edges of stone tools or iron weapons [@Birch2019; @Lycett2013], visually distinctive bone features [@Haruda2019; @Meloro2015], or ceramic assemblages with clear distinct components [@Selden2019; @Topi2017].
-With univariate or multivariate statistics for shape data generated by GMM, the application to artifacts can answer research questions related to topological evolution or variations in assemblages to define groups [@Doyon2019; @Eren2015; @Selden2018lithic; @Perez2007; @Presnyakova].
+With univariate or multivariate statistics, the GMM applications can answer research questions related to lithic topological evolution and variations [@Doyon2019; @Eren2015; @Selden2018lithic; @Perez2007; @Presnyakova2018], animal domestication or mobility [@Haruda2019; @Owen2014], hominid activities through cutmarks and taphonomic traces [@Aramendi2019; @Courtenay2019].
 A common procedure is using dimensional reduction techniques, such as principal components analysis or canonical variate analysis, to capture the essential features that can represent the overall shape.
@@
 <!-- size analysis -->
-
 We also conducted size analysis by examining the body diameter of the vessels to explore any differences in vessel form reflected by size and shape.
 Body diameter is a suitable metric measurement to represent size for ceramics since it is associated with volume as a utilitarian function, and could present differences in the situation of similar shape.
-The body diameter of the vessels from Chinese period is larger than those from the two earlier periods, and the vessels before European contact have the smallest body diameter on average (Figure \@ref(fig:size-regression): A).
-It is interesting that linear regression model (Figure \@ref(fig:size-regression): B) predicts that shape, captured by PC1 and PC2, changes with size across three phases, except PC1 in Chinese period.
+The body diameter of the vessels from Chinese period is larger than those from the two earlier periods, and the vessels before European contact have the smallest body diameter on average (Figure \@ref(fig:size-shape-regression-plots): A).
+It is interesting that linear regression model (Figure \@ref(fig:size-shape-regression-plots): B) predicts that shape, captured by PC1 and PC2, changes with size across three phases, except PC1 in Chinese period.
 That is, the shorter vessels are larger in body diameter according to the significant positive correlation between PC1 and size; however, the vessel from Chinese period does not show this pattern.
@@
 
-```{r size-regression, fig.cap= "A: Distribution of the body diameter of the vessels across three phases.
-B: Scatter plots of shape variable, PC1 at the top and PC2 at the bottom, as a function of body diamater"}
+```{r size-analysis}
 
@@
             x = "Diameter1_Body", 
-            y = "PC1",
+            y = "pc1",
             add = "reg.line",
@@
     #stat_regline_equation(label.y = 1) +
-    labs(x = "") +
+    labs(x = "", y = "PC1") +
     theme(axis.text = element_text(size = 8)) +
@@
             x = "Diameter1_Body", 
-            y = "PC2",
+            y = "pc2",
             add = "reg.line",
@@
     #stat_regline_equation(label.y = 1) +
-    labs(x = "body diameter (mm)") +
+    labs(x = "body diameter (mm)", y = "PC2") +
     theme(axis.text = element_text(size = 8)) +
@@
        units = "mm")
+```
+
+```{r, size-shape-regression-plots, echo=FALSE, fig.cap= "A: Distribution of the body diameter of the vessels across three phases.
+B: Scatter plots of shape variable, PC1 at the top and PC2 at the bottom, as a function of body diamater"}
+knitr::include_graphics(here("analysis/figures/size-boxplot-regression.jpg"))
 ```
@@
 The results further suggest that expressions of social identity or cultural boundaries in Indigenous societies through highly visible vessel qualities, such as shape, may be heightened during periods of foreign contact in pericolonial contexts.
+Current GMM method are mostly applied to study lithics topology, variations in bone shape in the filed of bioarchaeology or zooarchaeology, and some new attempts for understanding cutmarks and tophomic processes.
+We demonstrate that how we can use EFA-based GMM to identify the differences in ceramics to tackle the issue of specialized production.
 Our analysis, with its openly available methods and data, is readily extensible to other pottery assemblages in the region to further explore related questions about craft specialization and standardization in Iron Age ceramic technologies.

### 2020-07-24 analysis/paper/paper.Rmd
editing new text on size

@@
 Landmark GMM approaches assign a set of landmarks and/or semilandmarks onto objects as reference points that can be specified on a coordinate system.
-Landmark-based morphometrics are widely applied to archaeological objects with obvious morphological features that provide unambiguous reference points for landmark placement, such as tips and edges of stone tools or iron weapons [@Birch2019; @Lycett2013], visually distinctive bone features [@Haruda2019; @Meloro2015], or ceramic assemblages with clear distinct components [@Selden2019; @Topi2017].
-With univariate or multivariate statistics, the GMM applications can answer research questions related to lithic topological evolution and variations [@Doyon2019; @Eren2015; @Selden2018lithic; @Perez2007; @Presnyakova2018], animal domestication or mobility [@Haruda2019; @Owen2014], hominid activities through cutmarks and taphonomic traces [@Aramendi2019; @Courtenay2019].
-A common procedure is using dimensional reduction techniques, such as principal components analysis or canonical variate analysis, to capture the essential features that can represent the overall shape.
-This enables the identification of clustered analytical groups through the visualization of results, followed by statistical tests to robustly distinguish them.
+Landmark-based morphometrics have been widely applied to archaeological objects with obvious morphological features that provide unambiguous reference points for landmark placement, such as tips and edges of stone tools or iron weapons [@Birch2019; @Lycett2013], visually distinctive bone features [@Haruda2019; @Meloro2015], or ceramic assemblages with distinct components [@Selden2019; @Topi2017].
+With univariate or multivariate statistics, the GMM applications can answer research questions related to lithic typological and technological change [@Doyon2019; @Eren2015; @Selden2018lithic; @Perez2007; @Presnyakova2018], animal domestication or mobility [@Haruda2019; @Owen2014], and hominid activities through cutmarks and taphonomic traces [@Aramendi2019; @Courtenay2019].
+A common procedure is using dimensional reduction techniques, such as Principal Components Analysis or Canonical Variate Analysis, to capture the key features that represent overall shape.
+Visualisation of the reduced data enables the identification of groups, followed by statistical tests to robustly distinguish them.
 
-One of the interesting topics is to explore whether and why there is shape standardization over time, or between assemblages across geographical areas.
-It is common to use multivariate analysis on shape variables or coefficient of variations on associated metric data to discuss standardization, especially for lithic assemblages.
-For example, @Archer2015's case study of stone points in Southern Africa suggests an increase in shape standardization over time that may relate to a maintenance of finished points.
-With considerations of a set of metric data, @Buchanan2018; @Buchanan2019 identified a more uniform base-shape of Folsom points than Clovis points across the western US, which indicates the specific function for hafting.
-Similarly, @Smith2016 suggested that the humongous fluted point base in Alaska and northern Yukon is a risk management strategy to ensure the ease of replacement during long-distance travel.
-Other factor, such as low level of cultural innovation in a small group could also lead to an increase in standardization of point shapes [@Okumura2014].
-For ceramics, @Topi2017 found a standardized vessel shape from the Casas Grandes culture in northwest Mexico using coefficient of variations for the positions of semi-landmarks across shape groups.
-They suggest the standardization might indicate the pretense of specialized producers reflecting social complexity.
-Another way to examine standardization is using morphological disparity to perform pairwise comparison of correlations between shape groups [@Wills2001].
-In this manner, @Selden2018ceramic2; @Selden2019 examined Caddo ceramics in northeast Texas region and found a gradual shape standardization over time that provides a basis for further discussion of craft specialization or group identity.
-Similar applications to bifaces in central Texas presented the evidence of grater diversity in morphology [@Selden2020].
-More other applications, such as cranial deformation, demonstrate that landmark approach with multivariate analyses on shape variances or symmetry is an effective method to evaluate shape standardization [@Kuzminsky2016; @Natahi2019; @Perez2007].
+Key questions in archaeological shape analysis often involve measuring shape standardization over time, or between geographical areas.
+Standardization is often investigated using multivariate analysis on shape variables computed from landmark data <!-- make sure it's clear that this is a common element in the studies in this para -->, or coefficients of variation on associated metric data, especially for lithic assemblages.
+For example, @Archer2015's case study of stone points in Southern Africa suggests an increase in shape standardization over time that may relate to increased maintenance of finished points.
+@Buchanan2018 studied metric data and GMM to identify a more uniform base-shape of Folsom points than Clovis points across the western US [@Buchanan2019].
+Similarly, @Smith2016 suggested that the fluted point base in Alaska and northern Yukon is a risk management strategy to ensure the ease of replacement during long-distance travel <!-- not clear how this fits in this paragraph--> .
+Other factors, such as low levels of cultural innovation in a small group, could also lead to an increase in standardization of point shapes [@Okumura2014].
+For ceramics, @Topi2017 found standardized vessel shapes from the Casas Grandes culture in northwest Mexico using coefficients of variation for the positions of semi-landmarks across shape groups.
+They suggest standardization might indicate the presence of specialized producers, reflecting social complexity.
+Another way to examine standardization is using morphological disparity to perform pairwise comparisons of correlations between shape groups [@Wills2001] <!-- this is vague --> .
+In this manner, @Selden2018ceramic2; @Selden2019 examined Caddo ceramics in northeast Texas and found a gradual increase in shape standardization over time that provides a basis for further discussion of craft specialization or group identity <!-- make it clear he did something different to you -->.
+Similar applications to bifaces in central Texas show greater diversity in morphology <!-- any interpretation?
+--> [@Selden2020].
+Other applications, such as the study of cranial deformation, demonstrate that landmark approaches with multivariate analyses of shape variances is an effective method to evaluate shape standardization [@Kuzminsky2016; @Natahi2019; @Perez2007].
 
-Despite the wide use of landmark approach in archaeology, a key limitation is that landmarks may be difficult to reproducibly locate for structures that are mostly or entirely curves if not mathematically-defined.
+Despite the wide use of landmark approaches in archaeology, a key limitation is that landmarks may be difficult to reproducibly locate for structures that are mostly or entirely curves if not mathematically-defined.
 In those cases, outline approach, such as Elliptic Fourier Analysis (EFA), is a better mean for assessing morphological variation in whole structure of two-dimensional closed shapes [@Cardillo2010].
@@
 The first systematic use of EFA in archaeology was @Gero1984's study of lithic flakes in Peru that suggests the changes in forms from a more angular to rounded shape over time.
-Later, @Saragusti2005 introduced several functions for the further calculation of the specific shape attributes, including symmetry, roughness, and deformation, which demonstrates the potential for analysis of curves.
-To compare resharpening trajectories in the European Middle Paleolithic stone tools, @Iovictua2009 presented a protocol from outline digitization to EFA, with considerations of size using multivariate linear regression.
-More recent applications to stone artifacts seek to answer questions about topological classification of Late Woodland points [@Fox2015], cultural taxonomies for the European Late Palaeolithic [@Ivanovaite2020], and shape and symmetry standardization of British Acheulean [@Hoggard2019].
+Later, @Saragusti2005 introduced several functions for the further calculation of specific shape attributes, including symmetry, roughness, and deformation, which demonstrates the potential for analysis of curves.
+To compare resharpening trajectories in European Middle Paleolithic stone tools, @Iovictua2009 presented a protocol from outline digitization to EFA and multivariate linear regression.
+More recent applications to stone artifacts seek to answer questions about typological classification of Late Woodland points [@Fox2015], cultural taxonomies for the European Late Palaeolithic [@Ivanovaite2020], and shape and symmetry standardization of the British Acheulean [@Hoggard2019].
 
@@
 The scanned drawings were imported into Inkscape (http://inkscape.org) for digitization where outlines were manually traced.
-In those instances where only one side of the cross-section image was available, these were mirrored and joined to create a closed outline for each vessel.
+In those instances where only one side of the cross-section image was available, or some sections were missing, we interpolated the curves and then mirrored and joined to create a closed outline for each vessel.
 Analyses were conducted in R software (R Core Team, 2019) using functions included in the Momocs package for quantifying and analyzing shapes [@Bonhomme2014].
@@
 <!-- size analysis -->
+
 We also conducted size analysis by examining the body diameter of the vessels to explore any differences in vessel form reflected by size and shape.
-Body diameter is a suitable metric measurement to represent size for ceramics since it is associated with volume as a utilitarian function, and could present differences in the situation of similar shape.
-The body diameter of the vessels from Chinese period is larger than those from the two earlier periods, and the vessels before European contact have the smallest body diameter on average (Figure \@ref(fig:size-shape-regression-plots): A).
-It is interesting that linear regression model (Figure \@ref(fig:size-shape-regression-plots): B) predicts that shape, captured by PC1 and PC2, changes with size across three phases, except PC1 in Chinese period.
-That is, the shorter vessels are larger in body diameter according to the significant positive correlation between PC1 and size; however, the vessel from Chinese period does not show this pattern.
-On the contrary, the negative relationship between PC2 and size suggests that the vessels with narrower neck and mouth tend to have a larger body diameter.
+We measured body diameter directly from each physical vessel in the collection, and we focus on this metric because it is available for more vessels than any other metric.
+The body diameter of vessels from the Chinese period is larger than those from the two earlier periods, and the vessels before European contact have the smallest body diameter on average (Figure \@ref(fig:size-shape-regression-plots): A).
+Our linear regression models (Figure \@ref(fig:size-shape-regression-plots): B) show that shape, captured by PC1 and PC2, changes with size across the three phases, except for PC1 in the Chinese period.
+That is, the shorter vessels are larger in body diameter according to the significant positive correlation between PC1 and size.
+However, vessels from the Chinese period does not show this pattern.
+On the contrary, the negative relationship between PC2 and size suggests that vessels with a narrower neck and mouth tend to have a larger body diameter.
 
@@
 In general, vessels become shorter in height over time, leading to an oval-shaped body.
-Moreover, size significantly varies with shape for vessels before and after the European presence, which presents that shorter vessels with narrower opening tend to have a larger body diameter.
-This correlation between size and shape can be also observed during the Chinese period without the consideration of height.
+Moreover, size significantly varies with shape for vessels before and after the European presence.
+Shorter vessels with a narrower opening tended to have a larger body diameter.
+This correlation between size and shape can be also observed during the Chinese period, but there is no relationship with height.
 

### 2020-07-27 analysis/paper/paper.Rmd
edit GMM section

@@
 
-Geometric morphometrics differs from traditional linear measurements through its use of Cartesian coordinates of morphological structures to quantify and analyze shape [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007].
+Geometric morphometrics (GMM) differs from traditional linear measurements through its use of Cartesian coordinates of morphological structures to quantify and analyze shape [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007].
 Landmarks, curves or outlines of objects can be represented by coordinates in terms of their unique point locations with respect to numerical values on coordinate axes.
 There are two common morphometric methods: landmark and outline approaches [@Adams2004].
-Landmark GMM approaches assign a set of landmarks and/or semilandmarks onto objects as reference points that can be specified on a coordinate system.
+Landmark GMM approaches assign a set of landmarks and/or semilandmarks onto objects as reference points.
+Generalized Procrustes analysis (GPA) is used to superimpose landmark data on a common coordinate system by translating, rotating, and scaling [@Bookstein1991].
+After the procedure of GPA, superimposed landmark coordinates become shape variables that allow further statistical analyses [@Slice2007].
 Landmark-based morphometrics have been widely applied to archaeological objects with obvious morphological features that provide unambiguous reference points for landmark placement, such as tips and edges of stone tools or iron weapons [@Birch2019; @Lycett2013], visually distinctive bone features [@Haruda2019; @Meloro2015], or ceramic assemblages with distinct components [@Selden2019; @Topi2017].
@@
 A common procedure is using dimensional reduction techniques, such as Principal Components Analysis or Canonical Variate Analysis, to capture the key features that represent overall shape.
-Visualisation of the reduced data enables the identification of groups, followed by statistical tests to robustly distinguish them.
+Visualization of the reduced data enables the identification of groups, followed by statistical tests to robustly distinguish them.
 
-Key questions in archaeological shape analysis often involve measuring shape standardization over time, or between geographical areas.
-Standardization is often investigated using multivariate analysis on shape variables computed from landmark data <!-- make sure it's clear that this is a common element in the studies in this para -->, or coefficients of variation on associated metric data, especially for lithic assemblages.
+Key questions in archaeological shape analysis normally involve measuring shape standardization over time, or between geographical areas.
+Standardization is often investigated using multivariate analysis of shape variables computed from landmark data, or coefficients of variation on associated metric data, especially for lithic assemblages.
 For example, @Archer2015's case study of stone points in Southern Africa suggests an increase in shape standardization over time that may relate to increased maintenance of finished points.
 @Buchanan2018 studied metric data and GMM to identify a more uniform base-shape of Folsom points than Clovis points across the western US [@Buchanan2019].
-Similarly, @Smith2016 suggested that the fluted point base in Alaska and northern Yukon is a risk management strategy to ensure the ease of replacement during long-distance travel <!-- not clear how this fits in this paragraph--> .
+Also using metric data and GMM, @Smith2016 found standardized bases of fluted point in Alaska and northern Yukon that might indicate a risk management strategy to ensure the ease of replacement during long-distance travel.
 Other factors, such as low levels of cultural innovation in a small group, could also lead to an increase in standardization of point shapes [@Okumura2014].
-For ceramics, @Topi2017 found standardized vessel shapes from the Casas Grandes culture in northwest Mexico using coefficients of variation for the positions of semi-landmarks across shape groups.
-They suggest standardization might indicate the presence of specialized producers, reflecting social complexity.
-Another way to examine standardization is using morphological disparity to perform pairwise comparisons of correlations between shape groups [@Wills2001] <!-- this is vague --> .
-In this manner, @Selden2018ceramic2; @Selden2019 examined Caddo ceramics in northeast Texas and found a gradual increase in shape standardization over time that provides a basis for further discussion of craft specialization or group identity <!-- make it clear he did something different to you -->.
-Similar applications to bifaces in central Texas show greater diversity in morphology <!-- any interpretation?
---> [@Selden2020].
-Other applications, such as the study of cranial deformation, demonstrate that landmark approaches with multivariate analyses of shape variances is an effective method to evaluate shape standardization [@Kuzminsky2016; @Natahi2019; @Perez2007].
+For ceramics, @Topi2017 identified standardized vessel shapes from the Casas Grandes culture in northwest Mexico using coefficients of variation for the positions of semi-landmarks across shape groups.
+They suggested standardization might hint the presence of specialized producers, reflecting social complexity.
+Another way to explore standardization is pairwise testing of variations in morphological disparity among groups of shape variables based on the calculation of distances between groups in morphospace they occupy [@Wills2001].
+In this manner, @Selden2018ceramic2 examined Caddo ceramics using semi-landmark approaches in northeast Texas and found a gradual increase in shape standardization over time that provides a basis for further discussion of craft specialization or group identity [@Selden2019].
+Similar applications to lithic assemblages suggests that the Gahagan bifaces from the central Texas exhibit less size standardization than those from the southern Caddo area, indicative of different uses  [@Selden2020].
+Other applications, such as the study of cranial deformation, demonstrate that landmark approaches with multivariate analyses of shape variances are an effective method to evaluate shape standardization [@Kuzminsky2016; @Natahi2019; @Perez2007].
 
 Despite the wide use of landmark approaches in archaeology, a key limitation is that landmarks may be difficult to reproducibly locate for structures that are mostly or entirely curves if not mathematically-defined.
-In those cases, outline approach, such as Elliptic Fourier Analysis (EFA), is a better mean for assessing morphological variation in whole structure of two-dimensional closed shapes [@Cardillo2010].
+In those cases, outline approaches, such as Elliptic Fourier Analysis (EFA), is better for assessing morphological variations in the whole structure of two-dimensional closed shapes [@Cardillo2010].
 EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions [@Adams2004; @Bonhomme2014; @Claude2008].
-To be specific, EFA converts coordinates along a curve into Fourier function coefficients, called harmonic coefficients or harmonics [@Kuhl1982].
+In other words, EFA converts coordinates along a curve into Fourier function coefficients, called harmonic coefficients or harmonics [@Kuhl1982].
 The number of harmonics determines the quality and precision of the geometric representation of an object.
 The harmonic power, a cumulative sum of squared harmonic coefficients, provides a robust rule for determining the desired number of harmonics [@Bonhomme2014].
-The first systematic use of EFA in archaeology was @Gero1984's study of lithic flakes in Peru that suggests the changes in forms from a more angular to rounded shape over time.
-Later, @Saragusti2005 introduced several functions for the further calculation of specific shape attributes, including symmetry, roughness, and deformation, which demonstrates the potential for analysis of curves.
-To compare resharpening trajectories in European Middle Paleolithic stone tools, @Iovictua2009 presented a protocol from outline digitization to EFA and multivariate linear regression.
-More recent applications to stone artifacts seek to answer questions about typological classification of Late Woodland points [@Fox2015], cultural taxonomies for the European Late Palaeolithic [@Ivanovaite2020], and shape and symmetry standardization of the British Acheulean [@Hoggard2019].
+The first systematic use of  Fourier series to analyze shapes of artifacts in archaeology was @Gero1984's study of lithic flakes in Peru that suggests the changes in shape from a more angular to rounded tool over time.
+Later, @Saragusti2005 introduced more functions for the further calculation of specific shape attributes, including symmetry, roughness, and deformation, which demonstrates the potential of EFA for analysis of curves.
+To compare resharpening trajectories in European Middle Paleolithic stone tools, @Iovictua2009 presented a protocol from outline digitization, EFA procedure, to multivariate linear regression for shape variables.
+Recent lithic studies desmotrate the wide range of application of EFA to answer questions about typological classification of Late Woodland points [@Fox2015], the function of flaked obsidian tools in Easter Island [@Lipo2016], shape and symmetry standardization of the British Acheulean [@Hoggard2019], and cultural taxonomies for the European Late Palaeolithic [@Ivanovaite2020].
 
-The EFA applications for ceramics is rare, however, this approach is promising for answering questions related to ceramic taxonomy and standardization.
+Ceramic studies using EFA is rare, however, this approach is promising for answering questions related to ceramic taxonomy and standardization.
 For example, @Wilczek2014 evaluate the concordance between EFA and Discrete Cosine Transform (DCT), and a traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France.
@@
 We use EFA to evaluate the level of standardization of ceramics data from Kiwulan, northeastern Taiwan, in relation to the European presence in the 17th century to gain insights into the emergence of ceramic specialization.
-The reason of using EFA is because there is no sufficient landmarks could be identified straightforwardly due to the globular body of ceramics.
-We use a significance test for the equality of coefficient of variations of shape variables to statistically compare the vessel standardization from different periods.
+The reason for using EFA is because there are no sufficient landmarks that could be identified straightforwardly due to the globular body of ceramics.
+We use a significance test for the equality of coefficients of variation of shape variables to statistically compare the vessel standardization from different periods.
 
@@
 The results further suggest that expressions of social identity or cultural boundaries in Indigenous societies through highly visible vessel qualities, such as shape, may be heightened during periods of foreign contact in pericolonial contexts.
-Current GMM method are mostly applied to study lithics topology, variations in bone shape in the filed of bioarchaeology or zooarchaeology, and some new attempts for understanding cutmarks and tophomic processes.
-We demonstrate that how we can use EFA-based GMM to identify the differences in ceramics to tackle the issue of specialized production.
+Current GMM studies are mostly used for lithic topology, variations in bone shape in the filed of bioarchaeology or zooarchaeology, and some new applications for understanding cutmarks and tophomic processes.
+We demonstrated how we can use EFA-based GMM to identify the differences in ceramics to tackle the issue of specialized production.
 Our analysis, with its openly available methods and data, is readily extensible to other pottery assemblages in the region to further explore related questions about craft specialization and standardization in Iron Age ceramic technologies.

### 2020-07-27 analysis/paper/size_correlation .R

@@ -42,11 +42,20 @@ fit.shapephase <-
            print.progress = FALSE,
            iter = 9999)
 
-# morphological disparity
+# morphological disparity, only works for landmarks
 morphol.disparity(fit.shapephase,
                   groups = jpgs_imported_coo_size_pcs_metric$y,
                   data = geomorph_df,
                   print.progress = FALSE,
                   iter = 9999)
 
+# using dispRity example:
+
+phase_group <- crown.stem(op.p)
+
+subset <- custom.subsets(data = op.p$x, group= op.p$phase)
+sum_of_variances <- dispRity(data = op.p$x,
+                             group= op.p$phase,
+                             metric = c(sum, variances))
+summary(sum_of_variances)
 

### 2020-07-28 analysis/paper/paper.Rmd
add annotation of CV on PCs plot

@@
 len <- length(op.p$x[,1])
+
 # plot PC1-3 distribution by phase
@@
 
+# transform the data to wider format
+pc_all_wide  <- 
+  pc_plot %>% 
+  group_by(pcn) %>% 
+  mutate(row = row_number()) %>%
+  pivot_wider(names_from = pcn, 
+              values_from = pc) %>% 
+  select(-row)
+
+# calculate cv for each PC
+cv_pcs_all <-
+  pc_all_wide %>% 
+  group_by(phase) %>%
+  mutate(cv_pc1 = round(sd(PC1)/ mean(PC1), digits = 2)*100,
+         cv_pc2 = round(sd(PC2)/ mean(PC2), digits = 2)*100,
+         cv_pc3 = round(sd(PC3)/ mean(PC3), digits = 2)*100)%>% 
+  distinct(cv_pc1, cv_pc2, cv_pc3)
+
+# making the cv label
+cv_label <-
+  cv_pcs_all %>% 
+  mutate(cv_pc1 = paste("CV=", cv_pc1, "%"),
+         cv_pc2 = paste("CV=", cv_pc2, "%"),
+         cv_pc3 = paste("CV=", cv_pc3, "%"))
+
+cv_label <- c(cv_label$cv_pc1, cv_label$cv_pc2, cv_label$cv_pc3)
+
 ggplot(pc_plot,
@@
   facet_wrap(~ pcn, ncol = 1) +
+  geom_text(data = cv_pcs_all,
+            mapping = aes(x = c(2.1, 1.1, 3.1, 2.1, 1.1, 3.1, 2.1, 1.1, 3.1), 
+                          y = 0.8, 
+                          label = cv_label),
+            size = 3,
+            color= "red",
+            hjust = -0.1,
+            vjust= -1)+
   theme_minimal()
+
 ggsave(here("analysis/figures/pcs-boxplot-for-shape-variation.png"),
-       h = 10, w = 5)
+       h = 4, w = 5)
 ```
@@
        units = "mm")
+```
+
+```{r boxplot-pc1-ave-size}
+size_pcs_cv_join <-
+  jpgs_imported_coo_size_pcs_metric %>% 
+  left_join(cv_pcs_all, by = c("y", "phase"))
+
 ```

### 2020-07-28 analysis/paper/paper.Rmd
change normalization scale

@@
 Following @Eerkens2001 and @Roux2003, we take this as our measurement of standardization in vessel shape variables: lower CV values reflect higher standardization, and thus increased craft specialization in the community.
-Given that CV is robust for positive values due to the representation by ratio, we normalized PC scores to a range between 0 and 1 before computation of CV.
+Given that CV is robust for positive values due to the representation by ratio, we normalized PC scores to a range between 1 and 2 before computation of CV.
 
@@
 The second PC also presents a similar pattern of higher shape standardization in the Chinese period compared to the other two phases.
+The CVs calculated of the three PC also support a general trend toward a more standardized shape over time, especially for first PC.
 
@@
 # function for normalized data 
-normalize <- function(x) (x-min(x))/(max(x)-min(x))
+# normalize <- function(x) (x-min(x))/(max(x)-min(x)) # between 0-1
+normalize <- function(x) (1 + ((x-min(x))* 2))/(max(x)-min(x))
 
@@
             mapping = aes(x = c(2.1, 1.1, 3.1, 2.1, 1.1, 3.1, 2.1, 1.1, 3.1), 
-                          y = 0.8, 
+                          y = 2.1, 
                           label = cv_label),
@@
   stringsAsFactors = F)
+
+pc2_chinese <-
+  pc2 %>% 
+  filter(y == "Chinese") %>% 
+  mutate(cv = round(sd(x)/ mean(x), digits = 2)*100)
 
@@
 ```{r size-analysis}
-
 library(ggpubr)

### 2020-07-29 analysis/paper/paper.Rmd
adjust number in regression plot, and make boxplot for cv of size vs shape

@@
 Kiwulan is situated on a hill near a riverside at the northern margin of Yilan County, which is characterized by a triangular alluvial plain facing east toward the Pacific with high mountains on three other sides.
+
 This research investigates if there was increasing ceramic specialization resulting from Indigenous interaction with Europeans in the 17th century or Chinese in the 19th century, two major foreign influences in early historical Taiwan, which may indicate social changes in local Indigenous societies.
@@
 After the procedure of GPA, superimposed landmark coordinates become shape variables that allow further statistical analyses [@Slice2007].
-Landmark-based morphometrics have been widely applied to archaeological objects with obvious morphological features that provide unambiguous reference points for landmark placement, such as tips and edges of stone tools or iron weapons [@Birch2019; @Lycett2013], visually distinctive bone features [@Haruda2019; @Meloro2015], or ceramic assemblages with distinct components [@Selden2019; @Topi2017].
-With univariate or multivariate statistics, the GMM applications can answer research questions related to lithic typological and technological change [@Doyon2019; @Eren2015; @Selden2018lithic; @Perez2007; @Presnyakova2018], animal domestication or mobility [@Haruda2019; @Owen2014], and hominid activities through cutmarks and taphonomic traces [@Aramendi2019; @Courtenay2019].
 A common procedure is using dimensional reduction techniques, such as Principal Components Analysis or Canonical Variate Analysis, to capture the key features that represent overall shape.
 Visualization of the reduced data enables the identification of groups, followed by statistical tests to robustly distinguish them.
+Landmark-based morphometrics have been widely applied to archaeological objects with obvious morphological features that provide unambiguous reference points for landmark placement, such as tips and edges of stone tools or iron weapons [@Birch2019; @Lycett2013], visually distinctive bone features [@Haruda2019; @Meloro2015], or ceramic assemblages with distinct components [@Selden2019; @Topi2017].
+GMM applications can answer research questions related to lithic typological and technological change [@Doyon2019; @Eren2015; @Selden2018lithic; @Perez2007; @Presnyakova2018], animal domestication or mobility [@Haruda2019; @Owen2014], or hominid activities through cutmarks and taphonomic traces [@Aramendi2019; @Courtenay2019].
 
@@
 Other factors, such as low levels of cultural innovation in a small group, could also lead to an increase in standardization of point shapes [@Okumura2014].
+
 For ceramics, @Topi2017 identified standardized vessel shapes from the Casas Grandes culture in northwest Mexico using coefficients of variation for the positions of semi-landmarks across shape groups.
@@
 
-```{r PCs-distribution, fig.cap= "The distribution of PC scores by phases"}
+```{r PCs-distribution, fig.cap= "The distribution of mormalized PC scores by phases with CV values (%)"}
 
@@
 # normalize <- function(x) (x-min(x))/(max(x)-min(x)) # between 0-1
-normalize <- function(x) (1 + ((x-min(x))* 2))/(max(x)-min(x))
+normalize <- function(x) (1 + (x-min(x)))/(max(x)-min(x))
 
@@
             mapping = aes(x = c(2.1, 1.1, 3.1, 2.1, 1.1, 3.1, 2.1, 1.1, 3.1), 
-                          y = 2.1, 
+                          y = 1.8, 
                           label = cv_label),
@@
   stringsAsFactors = F)
-
-pc2_chinese <-
-  pc2 %>% 
-  filter(y == "Chinese") %>% 
-  mutate(cv = round(sd(x)/ mean(x), digits = 2)*100)
 
@@
     facet_wrap(~ fct_rev(y)) +
-    stat_cor(label.y = 1.25, label.sep = "\n") +
+    stat_cor(label.y = 2.25, label.sep = "\n") +
     #stat_regline_equation(label.y = 1) +
@@
     #scale_x_continuous(breaks = c(75, 200)) +
-    scale_y_continuous(limits = c(0, 1.5),
-                       breaks = c(seq(0, 1, 0.5)),
-                       labels = c(seq(0, 1, 0.5)))
+    scale_y_continuous(limits = c(1, 2.5),
+                       breaks = c(seq(1, 2.5, 0.5)),
+                       labels = c(seq(1, 2.5, 0.5)))
 
@@
   jpgs_imported_coo_size_pcs_metric %>% 
-  left_join(cv_pcs_all, by = c("y", "phase"))
+  group_by(y) %>% 
+  mutate(Diameter1_Body_ave = round(mean(Diameter1_Body), 2),
+         cv_body_diameter = round(sd(Diameter1_Body)/mean(Diameter1_Body), 
+                                  digits = 2)*100) %>%
+  mutate(phase = as.factor(y)) %>% 
+  left_join(cv_pcs_all) %>% 
+  left_join(pc1_df_with_filename)
+  #distinct(phase, cv_pc1, cv_pc2, cv_pc3, Diameter1_Body_ave, cv_body_diameter)
 
+boxplot_ave_size_cv_pc <-
+  size_pcs_cv_join %>% 
+  ggplot(aes(cv_body_diameter, cv_pc1, shape = phase)) +
+  geom_point() +
+  coord_flip() +
+  labs(x = "CV of Body diameter (%)",
+       y = "CV of PC1 (%)")
 ```

### 2020-07-29 analysis/paper/paper.Rmd
change PCs to scale of 1-10, and update plots

@@
 
-# function for normalized data 
-# normalize <- function(x) (x-min(x))/(max(x)-min(x)) # between 0-1
-normalize <- function(x) (1 + (x-min(x)))/(max(x)-min(x))
-
 # for one matrix
@@
   as_tibble() %>% 
-  mutate_all(normalize) %>% 
+  mutate_all(~scales::rescale(., to = c(1, 10))) %>% # normalize to 1-10
   as.matrix()
@@
          cv_pc3 = round(sd(PC3)/ mean(PC3), digits = 2)*100)%>% 
-  distinct(cv_pc1, cv_pc2, cv_pc3)
+  mutate(mean_pc1 = mean(PC1),
+         mean_pc2 = mean(PC2),
+         mean_pc3 = mean(PC3)) %>% 
+  distinct(cv_pc1, cv_pc2, cv_pc3, mean_pc1, mean_pc2, mean_pc3)
 
@@
             mapping = aes(x = c(2.1, 1.1, 3.1, 2.1, 1.1, 3.1, 2.1, 1.1, 3.1), 
-                          y = 1.8, 
+                          y = 8.5, 
                           label = cv_label),
@@
 
-```{r size-analysis}
+```{r size-analysis-prep}
 library(ggpubr)
-library(geomorph)
 
@@
   left_join(pottery_data_tidy)
+```
 
+```{r boxplot-size-by-phase}
 # size boxplot by period
@@
   theme(axis.text = element_text(size = 8))
+```
 
+```{r pcs-size-regression}
 # can pc1 predict size by period?
@@
     facet_wrap(~ fct_rev(y)) +
-    stat_cor(label.y = 2.25, label.sep = "\n") +
+    stat_cor(label.y = 10, label.sep = "\n") +
     #stat_regline_equation(label.y = 1) +
@@
     #scale_x_continuous(breaks = c(75, 200)) +
-    scale_y_continuous(limits = c(1, 2.5),
-                       breaks = c(seq(1, 2.5, 0.5)),
-                       labels = c(seq(1, 2.5, 0.5)))
+    scale_y_continuous(limits = c(0, 12),
+                       breaks = c(seq(0, 10, 2)),
+                       labels = c(seq(0, 10, 2)))
 
@@
     #scale_x_continuous(breaks = c(75, 200)) +
-    scale_y_continuous(limits = c(1, 2.5),
-                       breaks = c(seq(1, 2.5, 0.5)),
-                       labels = c(seq(1, 2.5, 0.5)))
+    scale_y_continuous(limits = c(0, 12),
+                       breaks = c(seq(0, 10, 2)),
+                       labels = c(seq(0, 10, 2)))
+```
 
+```{r size-regression-plot-together}
 # plot boxplot and regression together 
 right_column <-
-  plot_grid(pc1_size_reg, pc2_size_reg, ncol =1)
+  plot_grid(pc1_size_reg, pc2_size_reg, ncol = 1)
 

### 2020-07-29 analysis/paper/paper.Rmd
stash

--- 2020-07-28 analysis/paper/paper.Rmd
@@
 
-Geometric morphometrics (GMM) differs from traditional linear measurements through its use of Cartesian coordinates of morphological structures to quantify and analyze shape [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007].
+Geometric morphometrics differs from traditional linear measurements through its use of Cartesian coordinates of morphological structures to quantify and analyze shape [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007].
 Landmarks, curves or outlines of objects can be represented by coordinates in terms of their unique point locations with respect to numerical values on coordinate axes.
 There are two common morphometric methods: landmark and outline approaches [@Adams2004].
-Landmark GMM approaches assign a set of landmarks and/or semilandmarks onto objects as reference points.
-Generalized Procrustes analysis (GPA) is used to superimpose landmark data on a common coordinate system by translating, rotating, and scaling [@Bookstein1991].
-After the procedure of GPA, superimposed landmark coordinates become shape variables that allow further statistical analyses [@Slice2007].
+Landmark GMM approaches assign a set of landmarks and/or semilandmarks onto objects as reference points that can be specified on a coordinate system.
 Landmark-based morphometrics have been widely applied to archaeological objects with obvious morphological features that provide unambiguous reference points for landmark placement, such as tips and edges of stone tools or iron weapons [@Birch2019; @Lycett2013], visually distinctive bone features [@Haruda2019; @Meloro2015], or ceramic assemblages with distinct components [@Selden2019; @Topi2017].
@@
 A common procedure is using dimensional reduction techniques, such as Principal Components Analysis or Canonical Variate Analysis, to capture the key features that represent overall shape.
-Visualization of the reduced data enables the identification of groups, followed by statistical tests to robustly distinguish them.
+Visualisation of the reduced data enables the identification of groups, followed by statistical tests to robustly distinguish them.
 
-Key questions in archaeological shape analysis normally involve measuring shape standardization over time, or between geographical areas.
-Standardization is often investigated using multivariate analysis of shape variables computed from landmark data, or coefficients of variation on associated metric data, especially for lithic assemblages.
+Key questions in archaeological shape analysis often involve measuring shape standardization over time, or between geographical areas.
+Standardization is often investigated using multivariate analysis on shape variables computed from landmark data <!-- make sure it's clear that this is a common element in the studies in this para -->, or coefficients of variation on associated metric data, especially for lithic assemblages.
 For example, @Archer2015's case study of stone points in Southern Africa suggests an increase in shape standardization over time that may relate to increased maintenance of finished points.
 @Buchanan2018 studied metric data and GMM to identify a more uniform base-shape of Folsom points than Clovis points across the western US [@Buchanan2019].
-Also using metric data and GMM, @Smith2016 found standardized bases of fluted point in Alaska and northern Yukon that might indicate a risk management strategy to ensure the ease of replacement during long-distance travel.
+Similarly, @Smith2016 suggested that the fluted point base in Alaska and northern Yukon is a risk management strategy to ensure the ease of replacement during long-distance travel <!-- not clear how this fits in this paragraph--> .
 Other factors, such as low levels of cultural innovation in a small group, could also lead to an increase in standardization of point shapes [@Okumura2014].
-For ceramics, @Topi2017 identified standardized vessel shapes from the Casas Grandes culture in northwest Mexico using coefficients of variation for the positions of semi-landmarks across shape groups.
-They suggested standardization might hint the presence of specialized producers, reflecting social complexity.
-Another way to explore standardization is pairwise testing of variations in morphological disparity among groups of shape variables based on the calculation of distances between groups in morphospace they occupy [@Wills2001].
-In this manner, @Selden2018ceramic2 examined Caddo ceramics using semi-landmark approaches in northeast Texas and found a gradual increase in shape standardization over time that provides a basis for further discussion of craft specialization or group identity [@Selden2019].
-Similar applications to lithic assemblages suggests that the Gahagan bifaces from the central Texas exhibit less size standardization than those from the southern Caddo area, indicative of different uses  [@Selden2020].
-Other applications, such as the study of cranial deformation, demonstrate that landmark approaches with multivariate analyses of shape variances are an effective method to evaluate shape standardization [@Kuzminsky2016; @Natahi2019; @Perez2007].
+For ceramics, @Topi2017 found standardized vessel shapes from the Casas Grandes culture in northwest Mexico using coefficients of variation for the positions of semi-landmarks across shape groups.
+They suggest standardization might indicate the presence of specialized producers, reflecting social complexity.
+Another way to examine standardization is using morphological disparity to perform pairwise comparisons of correlations between shape groups [@Wills2001] <!-- this is vague --> .
+In this manner, @Selden2018ceramic2; @Selden2019 examined Caddo ceramics in northeast Texas and found a gradual increase in shape standardization over time that provides a basis for further discussion of craft specialization or group identity <!-- make it clear he did something different to you -->.
+Similar applications to bifaces in central Texas show greater diversity in morphology <!-- any interpretation?
+--> [@Selden2020].
+Other applications, such as the study of cranial deformation, demonstrate that landmark approaches with multivariate analyses of shape variances is an effective method to evaluate shape standardization [@Kuzminsky2016; @Natahi2019; @Perez2007].
 
 Despite the wide use of landmark approaches in archaeology, a key limitation is that landmarks may be difficult to reproducibly locate for structures that are mostly or entirely curves if not mathematically-defined.
-In those cases, outline approaches, such as Elliptic Fourier Analysis (EFA), is better for assessing morphological variations in the whole structure of two-dimensional closed shapes [@Cardillo2010].
+In those cases, outline approach, such as Elliptic Fourier Analysis (EFA), is a better mean for assessing morphological variation in whole structure of two-dimensional closed shapes [@Cardillo2010].
 EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions [@Adams2004; @Bonhomme2014; @Claude2008].
-In other words, EFA converts coordinates along a curve into Fourier function coefficients, called harmonic coefficients or harmonics [@Kuhl1982].
+To be specific, EFA converts coordinates along a curve into Fourier function coefficients, called harmonic coefficients or harmonics [@Kuhl1982].
 The number of harmonics determines the quality and precision of the geometric representation of an object.
 The harmonic power, a cumulative sum of squared harmonic coefficients, provides a robust rule for determining the desired number of harmonics [@Bonhomme2014].
-The first systematic use of  Fourier series to analyze shapes of artifacts in archaeology was @Gero1984's study of lithic flakes in Peru that suggests the changes in shape from a more angular to rounded tool over time.
-Later, @Saragusti2005 introduced more functions for the further calculation of specific shape attributes, including symmetry, roughness, and deformation, which demonstrates the potential of EFA for analysis of curves.
-To compare resharpening trajectories in European Middle Paleolithic stone tools, @Iovictua2009 presented a protocol from outline digitization, EFA procedure, to multivariate linear regression for shape variables.
-Recent lithic studies desmotrate the wide range of application of EFA to answer questions about typological classification of Late Woodland points [@Fox2015], the function of flaked obsidian tools in Easter Island [@Lipo2016], shape and symmetry standardization of the British Acheulean [@Hoggard2019], and cultural taxonomies for the European Late Palaeolithic [@Ivanovaite2020].
+The first systematic use of EFA in archaeology was @Gero1984's study of lithic flakes in Peru that suggests the changes in forms from a more angular to rounded shape over time.
+Later, @Saragusti2005 introduced several functions for the further calculation of specific shape attributes, including symmetry, roughness, and deformation, which demonstrates the potential for analysis of curves.
+To compare resharpening trajectories in European Middle Paleolithic stone tools, @Iovictua2009 presented a protocol from outline digitization to EFA and multivariate linear regression.
+More recent applications to stone artifacts seek to answer questions about typological classification of Late Woodland points [@Fox2015], cultural taxonomies for the European Late Palaeolithic [@Ivanovaite2020], and shape and symmetry standardization of the British Acheulean [@Hoggard2019].
 
-Ceramic studies using EFA is rare, however, this approach is promising for answering questions related to ceramic taxonomy and standardization.
+The EFA applications for ceramics is rare, however, this approach is promising for answering questions related to ceramic taxonomy and standardization.
 For example, @Wilczek2014 evaluate the concordance between EFA and Discrete Cosine Transform (DCT), and a traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France.
@@
 We use EFA to evaluate the level of standardization of ceramics data from Kiwulan, northeastern Taiwan, in relation to the European presence in the 17th century to gain insights into the emergence of ceramic specialization.
-The reason for using EFA is because there are no sufficient landmarks that could be identified straightforwardly due to the globular body of ceramics.
-We use a significance test for the equality of coefficients of variation of shape variables to statistically compare the vessel standardization from different periods.
+The reason of using EFA is because there is no sufficient landmarks could be identified straightforwardly due to the globular body of ceramics.
+We use a significance test for the equality of coefficient of variations of shape variables to statistically compare the vessel standardization from different periods.
 
@@
 Following @Eerkens2001 and @Roux2003, we take this as our measurement of standardization in vessel shape variables: lower CV values reflect higher standardization, and thus increased craft specialization in the community.
-Given that CV is robust for positive values due to the representation by ratio, we normalized PC scores to a range between 1 and 2 before computation of CV.
+Given that CV is robust for positive values due to the representation by ratio, we normalized PC scores to a range between 0 and 1 before computation of CV.
 
@@
 The second PC also presents a similar pattern of higher shape standardization in the Chinese period compared to the other two phases.
-The CVs calculated of the three PC also support a general trend toward a more standardized shape over time, especially for first PC.
 
@@
 # function for normalized data
-# normalize <- function(x) (x-min(x))/(max(x)-min(x)) # between 0-1
-normalize <- function(x) (1 + ((x-min(x))* 2))/(max(x)-min(x))
+# o scale your variable 𝑥 into a range (a, b) you can use:
+normalize <- function(a, b, x) (b - a) * (x - min(x)) / (max(x) - min(x)) + a
+
+scales::rescale(1:10, to = c(5, 7))
 
@@
   as_tibble() %>% 
-  mutate_all(normalize) %>% 
+  # mutate_all(~normalize(a = 1, b = 10, x = .)) %>% 
+  mutate_all(~scales::rescale(., to = c(1, 10))) %>% 
   as.matrix()
@@
 
-# what are the variences for each PC in each period?
+# what are the variances for each PC in each period?
 len <- length(op.p$x[,1])
-
 # plot PC1-3 distribution by phase
@@
 
+pc_plot %>% 
+  group_by(phase, pcn) %>% 
+  summarise(cvs = raster::cv(pc)) %>% 
+  pivot_wider(names_from = pcn, 
+              values_from = cvs)
+
 pc_plot$phase <- fct_relevel(pc_plot$phase, 
@@
                               "Post-European"))
-
-# transform the data to wider format
-pc_all_wide  <- 
-  pc_plot %>% 
-  group_by(pcn) %>% 
-  mutate(row = row_number()) %>%
-  pivot_wider(names_from = pcn, 
-              values_from = pc) %>% 
-  select(-row)
-
-# calculate cv for each PC
-cv_pcs_all <-
-  pc_all_wide %>% 
-  group_by(phase) %>%
-  mutate(cv_pc1 = round(sd(PC1)/ mean(PC1), digits = 2)*100,
-         cv_pc2 = round(sd(PC2)/ mean(PC2), digits = 2)*100,
-         cv_pc3 = round(sd(PC3)/ mean(PC3), digits = 2)*100)%>% 
-  distinct(cv_pc1, cv_pc2, cv_pc3)
-
-# making the cv label
-cv_label <-
-  cv_pcs_all %>% 
-  mutate(cv_pc1 = paste("CV=", cv_pc1, "%"),
-         cv_pc2 = paste("CV=", cv_pc2, "%"),
-         cv_pc3 = paste("CV=", cv_pc3, "%"))
-
-cv_label <- c(cv_label$cv_pc1, cv_label$cv_pc2, cv_label$cv_pc3)
 
@@
   facet_wrap(~ pcn, ncol = 1) +
-  geom_text(data = cv_pcs_all,
-            mapping = aes(x = c(2.1, 1.1, 3.1, 2.1, 1.1, 3.1, 2.1, 1.1, 3.1), 
-                          y = 2.1, 
-                          label = cv_label),
-            size = 3,
-            color= "red",
-            hjust = -0.1,
-            vjust= -1)+
   theme_minimal()
-
 ggsave(here("analysis/figures/pcs-boxplot-for-shape-variation.png"),
-       h = 4, w = 5)
+       h = 10, w = 5)
 ```
@@
 ```{r size-analysis}
+
 library(ggpubr)
@@
        units = "mm")
-```
-
-```{r boxplot-pc1-ave-size}
-size_pcs_cv_join <-
-  jpgs_imported_coo_size_pcs_metric %>% 
-  left_join(cv_pcs_all, by = c("y", "phase"))
-
 ```
@@
 The results further suggest that expressions of social identity or cultural boundaries in Indigenous societies through highly visible vessel qualities, such as shape, may be heightened during periods of foreign contact in pericolonial contexts.
-Current GMM studies are mostly used for lithic topology, variations in bone shape in the filed of bioarchaeology or zooarchaeology, and some new applications for understanding cutmarks and tophomic processes.
-We demonstrated how we can use EFA-based GMM to identify the differences in ceramics to tackle the issue of specialized production.
+Current GMM method are mostly applied to study lithics topology, variations in bone shape in the filed of bioarchaeology or zooarchaeology, and some new attempts for understanding cutmarks and tophomic processes.
+We demonstrate that how we can use EFA-based GMM to identify the differences in ceramics to tackle the issue of specialized production.
 Our analysis, with its openly available methods and data, is readily extensible to other pottery assemblages in the region to further explore related questions about craft specialization and standardization in Iron Age ceramic technologies.

### 2020-07-29 analysis/paper/paper.Rmd
resolve merge conflict
polish some of the plots a little bit

--- 2020-07-29 analysis/paper/paper.Rmd
@@
 
-# what are the variences for each PC in each period?
+# what are the variances for each PC in each period?
 len <- length(op.p$x[,1])
@@
                   pcn = str_glue('PC{sort(rep(1:3, len))}'))
+
+pc_cv_table <- 
+pc_plot %>% 
+  group_by(phase, pcn) %>% 
+  summarise(cvs = raster::cv(pc)) %>% 
+  pivot_wider(names_from = pcn, 
+              values_from = cvs)
 
@@
   cv_pcs_all %>% 
-  mutate(cv_pc1 = paste("CV=", cv_pc1, "%"),
-         cv_pc2 = paste("CV=", cv_pc2, "%"),
-         cv_pc3 = paste("CV=", cv_pc3, "%"))
+  mutate(cv_pc1 = paste("CV = ", cv_pc1, "%"),
+         cv_pc2 = paste("CV = ", cv_pc2, "%"),
+         cv_pc3 = paste("CV = ", cv_pc3, "%"))
 
@@
                           label = cv_label),
-            size = 3,
+            size = 2,
             color= "red",
@@
                    labels = c("Pre-\nEuropean", "Post-\nEuropean", "Chinese")) +
-  theme(axis.text = element_text(size = 8))
+  theme(axis.text = element_text(size = 8)) +
+  theme_bw()
 ```
@@
     facet_wrap(~ fct_rev(y)) +
-    stat_cor(label.y = 10, label.sep = "\n") +
+    stat_cor(label.y = 11, 
+             label.sep = ", ",
+             size = 2) +
     #stat_regline_equation(label.y = 1) +

### 2020-07-29 analysis/paper/paper.Rmd
pull all size analysis plots together

--- 2020-07-29 analysis/paper/paper.Rmd
@@
 
-```{r size-analysis-prep}
+```{r size-analysis-prep-1}
 library(ggpubr)
@@
 
+```{r size-analysis-prep-2}
+# combine with cv information 
+size_pcs_cv_join <-
+  jpgs_imported_coo_size_pcs_metric %>% 
+  group_by(y) %>% 
+  mutate(Diameter1_Body_ave = round(mean(Diameter1_Body), 2),
+         cv_body_diameter = round(sd(Diameter1_Body)/mean(Diameter1_Body), 
+                                  digits = 2)*100) %>%
+  mutate(phase = as.factor(y)) %>% 
+  left_join(cv_pcs_all) %>% 
+  left_join(pc1_df_with_filename)
+  #distinct(phase, cv_pc1, cv_pc2, cv_pc3, Diameter1_Body_ave, cv_body_diameter)
+```
+
 ```{r boxplot-size-by-phase}
+size_cv_df <-
+  size_pcs_cv_join %>% 
+  distinct(phase, Diameter1_Body_ave, cv_body_diameter, cv_pc1) %>% 
+  mutate(cv_body_diameter_label = paste("CV=", cv_body_diameter, "%"),
+         Diameter1_Body_ave_label = paste("mean=", Diameter1_Body_ave))
+
+cv_diameter_label <- size_cv_df$cv_body_diameter_label
+
 # size boxplot by period
@@
   coord_flip() +
-  labs(x = "body diameter (mm)", y = "") +
+  labs(x = "body diameter\n(mm)", y = "") +
   scale_y_discrete(breaks = c("Pre-European", "Post-European", "Chinese"),
                    labels = c("Pre-\nEuropean", "Post-\nEuropean", "Chinese")) +
-  theme(axis.text = element_text(size = 8))
+  theme(axis.text = element_text(size = 8)) +
+  annotate("text", x = c(126,122,145), y = c(2,1,3), 
+           size = 3, label = cv_diameter_label) +
+  theme_minimal()
 ```
+
+```{r boxplot-pc1-ave-size}
+boxplot_ave_size_cv_pc <-
+  size_cv_df %>% 
+  ggplot(aes(cv_body_diameter, cv_pc1)) +
+  geom_point() +
+  geom_text(label = size_cv_df$phase,
+            size = 3.5,
+            nudge_x = - 0.1, nudge_y = 0.5,
+            vjust="inward",hjust="inward") +
+  labs(x = "CV of body diameter (%)",
+       y = "CV of PC1\n(%)") +
+  theme(legend.position = "none") +
+  theme_minimal()
+```
+
+  scale_y_continuous(limits = c(10, 35),
+                       breaks = c(seq(10, 35, 5)),
+                       labels = c(seq(10, 35, 5))) +
 
@@
 # plot boxplot and regression together 
+left_column <-
+  plot_grid(boxplot_size, boxplot_ave_size_cv_pc, 
+            ncol = 1, rel_heights = c(1.2,1), 
+            labels = c("A", "B"), label_size = 12 )
+
 right_column <-
@@
 size_boxplot_regression <-
-  plot_grid(boxplot_size,
+  plot_grid(left_column,
             right_column,
-            labels = c("A", "B"),
+            labels = c("", "C", "B"),
             label_size = 12,
@@
        units = "mm")
-```
-
-```{r boxplot-pc1-ave-size}
-size_pcs_cv_join <-
-  jpgs_imported_coo_size_pcs_metric %>% 
-  group_by(y) %>% 
-  mutate(Diameter1_Body_ave = round(mean(Diameter1_Body), 2),
-         cv_body_diameter = round(sd(Diameter1_Body)/mean(Diameter1_Body), 
-                                  digits = 2)*100) %>%
-  mutate(phase = as.factor(y)) %>% 
-  left_join(cv_pcs_all) %>% 
-  left_join(pc1_df_with_filename)
-  #distinct(phase, cv_pc1, cv_pc2, cv_pc3, Diameter1_Body_ave, cv_body_diameter)
-
-boxplot_ave_size_cv_pc <-
-  size_pcs_cv_join %>% 
-  ggplot(aes(cv_body_diameter, cv_pc1, shape = phase)) +
-  geom_point() +
-  coord_flip() +
-  labs(x = "CV of Body diameter (%)",
-       y = "CV of PC1 (%)")
 ```

### 2020-07-29 analysis/paper/paper.Rmd
resolve merge conflict and adjust the plots

--- 2020-07-29 analysis/paper/paper.Rmd
@@
 
+```{r size-analysis-prep-2}
+# combine with cv information 
+size_pcs_cv_join <-
+  jpgs_imported_coo_size_pcs_metric %>% 
+  group_by(y) %>% 
+  mutate(Diameter1_Body_ave = round(mean(Diameter1_Body), 2),
+         cv_body_diameter = round(sd(Diameter1_Body)/mean(Diameter1_Body), 
+                                  digits = 2)*100) %>%
+  mutate(phase = as.factor(y)) %>% 
+  left_join(cv_pcs_all) %>% 
+  left_join(pc1_df_with_filename)
+  #distinct(phase, cv_pc1, cv_pc2, cv_pc3, Diameter1_Body_ave, cv_body_diameter)
+```
+
 ```{r boxplot-size-by-phase}
+size_cv_df <-
+  size_pcs_cv_join %>% 
+  distinct(phase, Diameter1_Body_ave, cv_body_diameter, cv_pc1) %>% 
+  mutate(cv_body_diameter_label = paste("CV =", cv_body_diameter, "%"),
+         Diameter1_Body_ave_label = paste("mean=", Diameter1_Body_ave))
+
+cv_diameter_label <- size_cv_df$cv_body_diameter_label
+
 # size boxplot by period
@@
   coord_flip() +
-  labs(x = "body diameter (mm)", y = "") +
+  labs(x = "body diameter\n(mm)", y = "") +
   scale_y_discrete(breaks = c("Pre-European", "Post-European", "Chinese"),
                    labels = c("Pre-\nEuropean", "Post-\nEuropean", "Chinese")) +
-  theme(axis.text = element_text(size = 8)) +
-  theme_bw()
+  annotate("text", x = c(126,122,144), y = c(2,1,3), 
+           size = 2, label = cv_diameter_label) +
+  theme_bw() +
+  theme(axis.text.x = element_text(size = rel(0.9)),
+        axis.text.y = element_text(size = rel(0.9)))
 ```
+
+```{r boxplot-pc1-ave-size}
+library(ggrepel)
+y_limits <- c(NA, 32)
+boxplot_ave_size_cv_pc <-
+  size_cv_df %>% 
+  ggplot(aes(cv_body_diameter, cv_pc1)) +
+  geom_point() +
+  geom_text_repel(label = size_cv_df$phase, 
+                  size = 2.5) + 
+  labs(x = "CV of body diameter (%)",
+       y = "CV of PC1\n(%)") +
+  theme(legend.position = "none") +
+  theme_bw() +
+  theme(axis.text.x = element_text(size = rel(0.9)),
+        axis.text.y = element_text(size = rel(0.9)))
+```
+
+label = size_cv_df$phase,
+                   size = 2.5, ylim = y_limits
 
@@
     theme(axis.text = element_text(size = 8)) +
-    #scale_x_continuous(breaks = c(75, 200)) +
     scale_y_continuous(limits = c(0, 12),
@@
 # plot boxplot and regression together 
+left_column <-
+  plot_grid(boxplot_size, boxplot_ave_size_cv_pc, 
+            ncol = 1, rel_heights = c(1.1,1), 
+            labels = c("A", "B"), label_size = 12 )
+
 right_column <-

### 2020-07-31 analysis/paper/paper.Rmd
edit the results and discussion of shape analysis

@@
 Other factors, such as low levels of cultural innovation in a small group, could also lead to an increase in standardization of point shapes [@Okumura2014].
+To test the effectiveness of measuring standardization, @Birch2019 compared landmark-based GMM to traditional metric analysis with CV using European iron weapons as an example.
+They demonstrated that landmark-based GMM can capture more variation in not only overall shape, but also bilateral symmetry.
 
@@
 Following @Eerkens2001 and @Roux2003, we take this as our measurement of standardization in vessel shape variables: lower CV values reflect higher standardization, and thus increased craft specialization in the community.
-Given that CV is robust for positive values due to the representation by ratio, we normalized PC scores to a range between 1 and 2 before computation of CV.
+Given that CV is robust for positive values due to the representation by a ratio scale, we normalized PC scores to a range between 1 and 10 before computation of CV.
 
@@
 
-We also conducted size analysis by examining the body diameter of the vessels to explore any differences in vessel form reflected by size and shape.
+We also conducted size analysis by examining the body diameter of vessels to examine any differences in vessel size and their relationships with vessel shape.
 We measured body diameter directly from each physical vessel in the collection, and we focus on this metric because it is available for more vessels than any other metric.
 The body diameter of vessels from the Chinese period is larger than those from the two earlier periods, and the vessels before European contact have the smallest body diameter on average (Figure \@ref(fig:size-shape-regression-plots): A).
-Our linear regression models (Figure \@ref(fig:size-shape-regression-plots): B) show that shape, captured by PC1 and PC2, changes with size across the three phases, except for PC1 in the Chinese period.
-That is, the shorter vessels are larger in body diameter according to the significant positive correlation between PC1 and size.
-However, vessels from the Chinese period does not show this pattern.
-On the contrary, the negative relationship between PC2 and size suggests that vessels with a narrower neck and mouth tend to have a larger body diameter.
+To investigate standardization of vessel form reflected by both shape and size, we calculated CV for PC1 (shape) in relation to CV for body diameter (size).
+The result (Figure \@ref(fig:size-shape-regression-plots): B) presents a higher form standardization in the Chinese period according to smaller CV values compared to CV values from the other two phases.
+However, there is no big differences in form standardization between and after the European presence.
+To understand the relationship between shape and size, we created linear regression models for PCs and the measurement of body diameter, with an examination of Pearson's correlation coefficient and p-value (Figure \@ref(fig:size-shape-regression-plots): C).
+The results demonstrate that shape changes significantly with size in all phases indicated by moderate positive relationships (0.3 ≤ r ≤ 0.7) and small p-values (≤ 0.05), except for PC1 in the Chinese period and PC2 in the pre-European period.
+That is, the shorter vessels are larger in body diameter according to the positive correlation between PC1 and body diameter.
+However, vessels from the Chinese period do not show this pattern.
+On the contrary, the negative relationship between PC2 and body diameter suggests that vessels with a narrower neck and mouth tend to have a larger body diameter.
 
@@
 
-```{r size-analysis-prep-2}
+```{r cv-metric-table-with-pc1-prep}
 # combine with cv information 
@@
   group_by(y) %>% 
-  mutate(Diameter1_Body_ave = round(mean(Diameter1_Body), 2),
-         cv_body_diameter = round(sd(Diameter1_Body)/mean(Diameter1_Body), 
-                                  digits = 2)*100) %>%
+  mutate(cv_body_diameter = round(raster::cv(Diameter1_Body), 2),
+         cv_neck_diameter = round(raster::cv(Diameter1_Neck), 2),
+         cv_rim_diameter = round(raster::cv(Diameter1_Rim), 2),
+         Diameter1_Body_mean = round(mean(Diameter1_Body), 2)) %>% 
   mutate(phase = as.factor(y)) %>% 
   left_join(cv_pcs_all) %>% 
-  left_join(pc1_df_with_filename)
-  #distinct(phase, cv_pc1, cv_pc2, cv_pc3, Diameter1_Body_ave, cv_body_diameter)
+  left_join(pc1_df_with_filename) %>% 
+  select(-y)
+  #distinct(phase, cv_pc1, cv_pc2, cv_pc3, Diameter1_Body_mean, cv_body_diameter)
 ```
@@
 ```{r boxplot-size-by-phase}
+# prepare label
 size_cv_df <-
   size_pcs_cv_join %>% 
-  distinct(phase, Diameter1_Body_ave, cv_body_diameter, cv_pc1) %>% 
+  distinct(phase, Diameter1_Body_mean, cv_body_diameter, cv_pc1) %>% 
   mutate(cv_body_diameter_label = paste("CV =", cv_body_diameter, "%"),
-         Diameter1_Body_ave_label = paste("mean=", Diameter1_Body_ave))
+         Diameter1_Body_mean_label = paste("mean=", Diameter1_Body_mean))
 
@@
 
-```{r boxplot-pc1-ave-size}
+```{r byplot-cvs-pc1-size}
 library(ggrepel)
 y_limits <- c(NA, 32)
-boxplot_ave_size_cv_pc <-
+byplot_cvs_pc1_diameter <-
   size_cv_df %>% 
@@
 ```
-
-label = size_cv_df$phase,
-                   size = 2.5, ylim = y_limits
 
@@
 
-```{r size-regression-plot-together}
+```{r plot-size-regression-together}
 # plot boxplot and regression together 
 left_column <-
-  plot_grid(boxplot_size, boxplot_ave_size_cv_pc, 
+  plot_grid(boxplot_size, byplot_cvs_pc1_diameter, 
             ncol = 1, rel_heights = c(1.1,1), 
@@
        here("analysis", "figures",
-            "size-boxplot-regression.jpg"),
+            "plot-size-boxplot-regression.jpg"),
        width = 180, 
@@
 
-```{r, size-shape-regression-plots, echo=FALSE, fig.cap= "A: Distribution of the body diameter of the vessels across three phases.
-B: Scatter plots of shape variable, PC1 at the top and PC2 at the bottom, as a function of body diamater"}
+```{r, size-shape-regression-plots, echo=FALSE, fig.cap= "A: Distribution of the body diameter of vessels with coefficients of variation by phases.
+B: Coefficients of variation of vessel shape represented by PC1 in relation to vessel size represetnted by body diameter, showing a more standardization in vessel form in the Chinese period.
+C: Correlation between shape variables (PC1 and PC2) and body diameter of vessles with Person's r and P-values by phases"}
 knitr::include_graphics(here("analysis/figures/size-boxplot-regression.jpg"))
@@
 In general, vessels become shorter in height over time, leading to an oval-shaped body.
-Moreover, size significantly varies with shape for vessels before and after the European presence.
-Shorter vessels with a narrower opening tended to have a larger body diameter.
-This correlation between size and shape can be also observed during the Chinese period, but there is no relationship with height.
+This can be also supported by an increase in body diameter of vessels.
+Also, the correlation between shape and body diameter suggests that size significantly varies with vessel shape.
+Shorter vessels with a narrower opening tended to have a larger body diameter, indicative of a change in overall vessel form.
 
@@
 Pottery shape became more homogeneous and standardized after contact with Europeans and even more so after contact with the Chinese.
+In addition, we found a larger and more standardized body diameter of vessels in the Chinese period.
+We notice the possibility of more human errors could be involved in hand-crafting when the size of artifacts increases [@Eerkens2001].
+However, it is opposite for the ceramics in the Chinese period when we examining body diameter.
+This might support the intent to achieve a homogeneous form of vessels, but we still need to compare it with other metric variables to validate this claim.
 Compositional analysis shows that the clay pastes are similar, regardless of the increasing standardization of the pottery shape, reflecting continuity in the raw material sources.

### 2020-07-31 analysis/paper/size_correlation .R

@@ -59,3 +59,22 @@ sum_of_variances <- dispRity(data = op.p$x,
                              metric = c(sum, variances))
 summary(sum_of_variances)
 
+
+library(ggrepel)
+byplot_cv_pc_metric_diameter <-
+  size_pcs_cv_join %>%
+  rename(body = Diameter1_Body, neck =Diameter1_Neck, rim = Diameter1_Rim) %>%
+  select(body, neck, rim, cv_pc1, phase) %>%
+  pivot_longer(-c(cv_pc1, y, phase), names_to = "diameter", values_to = "width") %>%
+  ggplot(aes(x = width, y = cv_pc1, color = diameter)) +
+  geom_point() +
+  annotate("text", x = c(126, 120, 145), y = c(32, 30, 16),
+           label = c("Pre-European", "Post-European", "Chinese")) +
+  labs(x = "diameter (mm)",
+       y = "CV of PC1 (%)") +
+  scale_color_viridis(discrete = TRUE, option = "D") +
+  scale_fill_viridis(discrete = TRUE) +
+  theme_bw() +
+  theme(axis.text.x = element_text(size = rel(0.9)),
+        axis.text.y = element_text(size = rel(0.9)))
+```

### 2020-07-31 analysis/paper/paper.Rmd
edit conclusion and add reference

@@
 
-This research investigates if there was increasing ceramic specialization resulting from Indigenous interaction with Europeans in the 17th century or Chinese in the 19th century, two major foreign influences in early historical Taiwan, which may indicate social changes in local Indigenous societies.
+This research investigates if there was increasing ceramic specialization resulting from Indigenous interaction with Europeans in the 17th century or Chinese in the 19th century, two major foreign influences in early historical Taiwan, which may indicate social changes in local Indigenous societies [cf.
+@Berrocal2020].
 We predict that competition within the Indigenous community at Kiwulan for foreign resources and trade partnerships with European or Chinese colonizers may have led to the emergence of craft specialization caused by greater economic and social control of ceramic production by a small group of individuals.
@@
 The results show the variation indicated by EFA and DCT matches the traditional ceramic typology, which supports that claim that outline-based approaches can be efficiently used for studying variations in ceramic shapes.
-Furthermore, @Wilczek2014's EFA results help us understand the level of production standardization over time across the region.
+Furthermore, @Wilczek2014 point out the potential of EFA for detecting the level of ceramic standardization.
 EFA is suitable for analyzing shapes lacking representative landmarks, or where curves contain the most meaningful variation.
@@
 However, it is opposite for the ceramics in the Chinese period when we examining body diameter.
-This might support the intent to achieve a homogeneous form of vessels, but we still need to compare it with other metric variables to validate this claim.
+This might support the intent to achieve a homogeneous form of vessels, but we still need to compare it with other metric variables to assess the variation in size.
 Compositional analysis shows that the clay pastes are similar, regardless of the increasing standardization of the pottery shape, reflecting continuity in the raw material sources.
@@
 
-This study demonstrated an application of EFA on ceramic shapes to explore the emergence of ceramic specialization as indicative of foreign influences.
+This study demonstrated the first use of EFA on ceramic shapes to explore the emergence of ceramic specialization as indicative of foreign influences.
 Here, EFA is combined with a significance test for the equality of CVs of shape variables to provide a robust method of identifying and statistically assessing differences in shape standardization.
@@
 The results further suggest that expressions of social identity or cultural boundaries in Indigenous societies through highly visible vessel qualities, such as shape, may be heightened during periods of foreign contact in pericolonial contexts.
-Current GMM studies are mostly used for lithic topology, variations in bone shape in the filed of bioarchaeology or zooarchaeology, and some new applications for understanding cutmarks and tophomic processes.
-We demonstrated how we can use EFA-based GMM to identify the differences in ceramics to tackle the issue of specialized production.
 Our analysis, with its openly available methods and data, is readily extensible to other pottery assemblages in the region to further explore related questions about craft specialization and standardization in Iron Age ceramic technologies.
+Furthermore, this study broadens the GMM field, where the common topics are lithic topology, bone morphology, and patterns of cutmarks or taphonomy.
+We suggest that EFA-based GMM is an effective method for detecting morphological differences in ceramics assemblages.
 

### 2020-07-31 analysis/paper/size_correlation .R

@@ -19,6 +19,8 @@ summary(linearMod)
 # are residuals normally distributed? Yes
 qqnorm(jpgs_imported_coo_size_pcs_metric$Diameter1_Body)
 
+
+-----------trying morphological disparity------------
 # assign landmarks and GPA
 jpgs_imported_coo_land <-
   jpgs_imported %>%
@@ -50,7 +52,7 @@ morphol.disparity(fit.shapephase,
                   iter = 9999)
 
 # using dispRity example:
-
+package(dispRity)
 phase_group <- crown.stem(op.p)
 
 subset <- custom.subsets(data = op.p$x, group= op.p$phase)

### 2020-07-31 analysis/paper/paper.Rmd
update for discussion

@@
 We can thus rule out changes in clay fabric as a factor in explaining changes in vessel shape.
-To determine if form changes might be related to changes in the function of pots at Kiwulan, we used geochemical methods to extract and identify lipids trapped in the fabric of potsherds to identify foods that may have contributed residues absorbed into the clay [cf.
+To determine if shape changes might be related to changes in the function of pots at Kiwulan, we used geochemical methods to extract and identify lipids trapped in the fabric of potsherds to identify foods that may have contributed residues absorbed into the clay [cf.
 @Kwak2015].
@@
 This suggests that differences in population across our three occupation phases are probably not driving variation in craft specialization.
+We also note that a smaller sample size in the Chinese period may have effects on a more standardized shape.
+The effect can be reduced using CV that offers a way to robustly compare variations across uneven sample sizes, regardless of a small sample size in the whole dataset [@Eerkens2001, pp.498].
+Moreover, the MSLR test for equality of CVs can deal with different sample sizes as well.
+Also, the spatial pattern shows that Chinese pottery samples were collected from different units acorss research area.
 

### 2020-07-31 disparity.R

@@ -1,0 +1,47 @@
+require(geomorph)
+require(dispRity)
+
+## Loading the plethodon dataset
+data(plethodon)
+
+str(plethodon)
+
+## Performing a Procrustes transform on the landmarks
+procrustes <- gpagen(plethodon$land, PrinAxes = FALSE,
+                     print.progress = FALSE)
+
+## Ordinating this data
+geomorph.ordination(procrustes)[1:5,1:5]
+
+## Using a geomorph.data.frame
+geomorph_df <- geomorph.data.frame(procrustes,
+                                   species = plethodon$species,
+                                   site = plethodon$site)
+
+## Ordinating this data and making a dispRity object
+geomorph.ordination(geomorph_df)
+
+# https://rawgit.com/TGuillerme/dispRity/master/inst/gitbook/_book/getting-started-with-disprity.html#what-sort-of-data-does-disprity-work-with
+plethodon$species
+
+species_list <-
+  list("Jord" = which(plethodon$species == "Jord" )     ,
+       "Teyah" = which(plethodon$species == "Teyah" )
+  )
+
+phases_list <-
+  list("Pre-European" = which(op.p$phase == "Pre-European"),
+       "Post-European" = which(op.p$phase == "Post-European"),
+       "Chinese" = which(op.p$phase == "Chinese"))
+
+disparity_data <-
+  dispRity.per.group(op.p$x,
+                     group = phases_list,
+                     metric = c(sum, variances))
+
+summary(disparity_data)
+plot(disparity_data)
+test.dispRity(disparity_data, test = wilcox.test,
+              correction = "bonferroni", details = TRUE)
+
+

### 2020-08-02 analysis/paper/paper.Rmd
added morphological disparity test and updated the discussion section

@@
 
-```{r, first-two-PCs-byplot-and-tps-plot, echo=FALSE, fig.cap='Left: Significant differences in average vessel shapes across phases using thin plate splines (TPS) with outline deformations required to pass from an extreme of one morphospace to another.
+```{r, first-two-PCs-byplot-and-tps-plot, echo=FALSE, fig.cap='Left: Significant differences in average vessel shapes between the Chinese and the post-European period using thin plate splines (TPS) with outline deformations required to pass from an extreme of one morphospace to another.
 Right: Pottery shape distribution by each occupation phase according to the first two PCs.'}
@@
 
-To compare pottery shape standardization across the three phases we investigate the distributions of the first three PC scores, taking the PC scores as proxy variables for vessel shape (Figure \@ref(fig:PCs-distribution)).
-The first PC, capturing vessel height and roundness, shows higher variation in the pre-European period compared to the Chinese period.
-That indicates that shape standardization was higher in the Chinese period compared to the pre-European period.
-The second PC also presents a similar pattern of higher shape standardization in the Chinese period compared to the other two phases.
-The CVs calculated of the three PC also support a general trend toward a more standardized shape over time, especially for first PC.
+To compare pottery shape standardization across the three phases we investigated the distributions of the first three PC scores, taking the PC scores as proxy variables for vessel shape (Figure \@ref(fig:PCs-distribution)).
+The CVs calculated of the three PC support a general trend toward a more standardized shape over time, especially the shape identified by PC1.
+The PC1, capturing vessel height and roundness, shows higher variation in the pre-European period and post-European period compared to the Chinese period.
+In other words, shape standardization was higher in the Chinese period.
+However, the PC2 presents a similar diversity in ceramics assemblages across three phases, while the PC3 demonstrates a slightly standardized shape in the Chinese period.
 
 To see whether the differences in the distribution of PCs between any two phases are substantive or due to chance, we used a modified signed-likelihood ratio test to assess the equality of CVs [@Krishnamoorthy2014; @Marwick2019].
-P-values for this significance test of CVs for PC1 and PC2 show significant differences in the standardization of vessel shapes across periods, especially between Chinese contact with either pre-European or post-European (Figure \@ref(fig:PCs-distribution), Table \@ref(tab:CV-test-PCs-table)).
-A significant difference was also detected between the pre-European and post-European periods.
+P-values for this significance test of CVs for PC1 show significant differences in the standardization of vessel shapes across periods, between Chinese contact with either pre-European or post-European (Table \@ref(tab:CV-test-PCs-table)).
+This result statistically supports the observation of a higher standardized shape in Chinese period.
 
@@
 Ceramic vessel shape standardization was examined to measure craft specialization as a proxy for social differentiation [@Costin2001; @Junker1999].
-The result of our MANOVA demonstrates a significant difference in shapes between the pre-European and Post-European periods, and between the pre-European and the Chinese periods.
+The result of our MANOVA demonstrates a significant difference in shapes between the pre-European and the Chinese periods, and between the pre-European and Post-European periods.
 The average shape presents as a round body with a wide rim and neck before European contact, which shifts to a more oval-shaped body with narrower rim and neck after the European presence.
@@
 
-Our CV tests indicate that there are significant differences in shape standardization between the pre-European period and post-European period, and between the Chinese contact and either pre-European or post-European periods.
-Pottery shape became more homogeneous and standardized after contact with Europeans and even more so after contact with the Chinese.
-In addition, we found a larger and more standardized body diameter of vessels in the Chinese period.
+For the degree of shape standardization across periods, our CV tests on PC1 indicate that there are significant differences between Chinese contact and either pre-European or post-European periods.
+That means pottery shape became more homogeneous and standardized after contact with the Chinese.
+A test of morphological disparity we conducted as a cross validation that also supports this result.
+In addition, we found a larger and more standardized size, indicated by body diameter of vessels, in the Chinese period.
 We notice the possibility of more human errors could be involved in hand-crafting when the size of artifacts increases [@Eerkens2001].
-However, it is opposite for the ceramics in the Chinese period when we examining body diameter.
-This might support the intent to achieve a homogeneous form of vessels, but we still need to compare it with other metric variables to assess the variation in size.
+However, it is opposite for the ceramics in the Chinese period using body diameter as a proxy variable of size.
+This might hint the intent to achieve a homogeneous form of vessels, but we still need to obtain more metric variables to assess the variation in size.
 Compositional analysis shows that the clay pastes are similar, regardless of the increasing standardization of the pottery shape, reflecting continuity in the raw material sources.
@@
 We also note that a smaller sample size in the Chinese period may have effects on a more standardized shape.
-The effect can be reduced using CV that offers a way to robustly compare variations across uneven sample sizes, regardless of a small sample size in the whole dataset [@Eerkens2001, pp.498].
+However, the effect can be reduced using CV, which offers a way to robustly compare variations across uneven sample sizes, regardless of a smaller sample size in the whole dataset [@Eerkens2001, pp.498].
 Moreover, the MSLR test for equality of CVs can deal with different sample sizes as well.
-Also, the spatial pattern shows that Chinese pottery samples were collected from different units acorss research area.
 

### 2020-08-02 analysis/paper/size_correlation .R

@@ -20,48 +20,6 @@ summary(linearMod)
 qqnorm(jpgs_imported_coo_size_pcs_metric$Diameter1_Body)
 
 
------------trying morphological disparity------------
-# assign landmarks and GPA
-jpgs_imported_coo_land <-
-  jpgs_imported %>%
-  Out(., fac = phases_isect)  %>%
-  # Centring, scaling and sampling pseudo-landmarks
-  coo_center %>%
-  coo_scale %>%
-  coo_sample(., 50) %>%
-  # Procrustes superimposition for pseudo-landmarks
-  fgProcrustes
-
-# geomorph data frame
-geomorph_df <- geomorph.data.frame(shape = jpgs_imported_coo_land$coo,
-                                   size = jpgs_imported_coo_size_pcs_metric$jpgs_imported_coo_size,
-                                   phase = jpgs_imported_coo_size_pcs_metric$y)
-
-# modeling shape according
-fit.shapephase <-
-  procD.lm(shape ~ phase,
-           data = geomorph_df,
-           print.progress = FALSE,
-           iter = 9999)
-
-# morphological disparity, only works for landmarks
-morphol.disparity(fit.shapephase,
-                  groups = jpgs_imported_coo_size_pcs_metric$y,
-                  data = geomorph_df,
-                  print.progress = FALSE,
-                  iter = 9999)
-
-# using dispRity example:
-package(dispRity)
-phase_group <- crown.stem(op.p)
-
-subset <- custom.subsets(data = op.p$x, group= op.p$phase)
-sum_of_variances <- dispRity(data = op.p$x,
-                             group= op.p$phase,
-                             metric = c(sum, variances))
-summary(sum_of_variances)
-
-
 library(ggrepel)
 byplot_cv_pc_metric_diameter <-
   size_pcs_cv_join %>%

### 2020-08-02 disparity.R

@@ -1,33 +1,6 @@
-require(geomorph)
 require(dispRity)
 
-## Loading the plethodon dataset
-data(plethodon)
-
-str(plethodon)
-
-## Performing a Procrustes transform on the landmarks
-procrustes <- gpagen(plethodon$land, PrinAxes = FALSE,
-                     print.progress = FALSE)
-
-## Ordinating this data
-geomorph.ordination(procrustes)[1:5,1:5]
-
-## Using a geomorph.data.frame
-geomorph_df <- geomorph.data.frame(procrustes,
-                                   species = plethodon$species,
-                                   site = plethodon$site)
-
-## Ordinating this data and making a dispRity object
-geomorph.ordination(geomorph_df)
-
-# https://rawgit.com/TGuillerme/dispRity/master/inst/gitbook/_book/getting-started-with-disprity.html#what-sort-of-data-does-disprity-work-with
-plethodon$species
-
-species_list <-
-  list("Jord" = which(plethodon$species == "Jord" )     ,
-       "Teyah" = which(plethodon$species == "Teyah" )
-  )
+# https://rawgit.com/TGuillerme/dispRity/master/inst/gitbook/_book/getting-started-with-disprity.html#simpleanalysis
 
 phases_list <-
   list("Pre-European" = which(op.p$phase == "Pre-European"),
@@ -40,8 +13,8 @@ disparity_data <-
                      metric = c(sum, variances))
 
 summary(disparity_data)
+
 plot(disparity_data)
+
 test.dispRity(disparity_data, test = wilcox.test,
               correction = "bonferroni", details = TRUE)
-
-

### 2020-08-06 analysis/paper/paper.Rmd
edit discussion section

@@
 
-```{r PCs-distribution, fig.cap= "The distribution of mormalized PC scores by phases with CV values (%)"}
+```{r PCs-distribution, fig.cap= "The distribution of normalized PC scores by phases with CV values (%)"}
 
@@
 
-Previous investigations at Kiwulan suggested an unequal distribution of prestige goods, trade ornaments in particular, following the appearance of Europeans [@Cheng2008; @Wang2011], which hinted at the emergence of social inequality within the Indigenous community.
-Ceramic vessel shape standardization was examined to measure craft specialization as a proxy for social differentiation [@Costin2001; @Junker1999].
-The result of our MANOVA demonstrates a significant difference in shapes between the pre-European and the Chinese periods, and between the pre-European and Post-European periods.
+Previous investigations at Kiwulan suggested an unequal distribution of prestige goods, trade ornaments in particular, following the appearance of Europeans with a high diversity of types [@Cheng2008; @Wang2011].
+This hinted at the emergence of social inequality within the Indigenous community.
+To understand the relationship between social inequality and foreign presences, we examined ceramic vessel shape standardization to measure craft specialization as a proxy for social change [@Costin2001; @Junker1999].
+The result of our MANOVA demonstrates a significant difference in shapes between the pre-European and Post-European periods, and the pre-European and Chinese periods.
 The average shape presents as a round body with a wide rim and neck before European contact, which shifts to a more oval-shaped body with narrower rim and neck after the European presence.
-Shape changes are more pronounced during the Chinese contact period.
-In general, vessels become shorter in height over time, leading to an oval-shaped body.
-This can be also supported by an increase in body diameter of vessels.
-Also, the correlation between shape and body diameter suggests that size significantly varies with vessel shape.
-Shorter vessels with a narrower opening tended to have a larger body diameter, indicative of a change in overall vessel form.
+Such shape is more pronounced in the Chinese period.
+In general, vessels become oval-shaped with a restricted mouth over time, which corresponds with an increase in body diameter of vessels.
+The correlation between shape and body diameter suggests that body diameter significantly varies with vessel shape.
+Oval-shaped vessels with a narrower opening tended to have a larger body diameter, indicative of a change in overall vessel form.
 
 For the degree of shape standardization across periods, our CV tests on PC1 indicate that there are significant differences between Chinese contact and either pre-European or post-European periods.
-That means pottery shape became more homogeneous and standardized after contact with the Chinese.
-A test of morphological disparity we conducted as a cross validation that also supports this result.
-In addition, we found a larger and more standardized size, indicated by body diameter of vessels, in the Chinese period.
-We notice the possibility of more human errors could be involved in hand-crafting when the size of artifacts increases [@Eerkens2001].
-However, it is opposite for the ceramics in the Chinese period using body diameter as a proxy variable of size.
-This might hint the intent to achieve a homogeneous form of vessels, but we still need to obtain more metric variables to assess the variation in size.
-Compositional analysis shows that the clay pastes are similar, regardless of the increasing standardization of the pottery shape, reflecting continuity in the raw material sources.
+A test of morphological disparity also demonstrates a similar result as well (See Online Supplementary Materials), suggesting that pottery shape became more standardized after contact with the Chinese.
+In addition, we found a more homogeneous shape accompanied by standardized size, represented by the body diameter of vessels, in the Chinese period.
+Humans tend to make mistakes in hand-crafting when the size of an object increases, leading to higher variations in larger artifacts [@Eerkens2001].
+However, it is opposite for the ceramics in the Chinese period when we use body diameter as our proxy variable of size.
+This might hint an intentional behavior to achieve a homogeneous form of vessels, but we still need more metric variables to assess the variation in size.
+Mineral composition shows that the clay pastes are similar throughout three phases, regardless of the increasing standardization of the pottery shape, reflecting continuity in the raw material sources.
 We can thus rule out changes in clay fabric as a factor in explaining changes in vessel shape.
+We note that a small sample size in the Chinese period may lead to a more standardized shape.
+However, this effect can be reduced using CV that scales variation to magnitude, allowing reliable comparison across uneven sample numbers, even at small sample size [@Eerkens2001].
+Moreover, the MSLR test for equality of CVs enables a robust test between different sample numbers [@Krishnamoorthy2014].
+
+Whether standardization reflects an emergence of specialized production is associated with the number of producers, which distinguishes mechanical standardization from intentional standardization [@Costin2001; @Costin1995].
+In other words, mechanical standardization depends on potter's skills and experiences that increased uniformity normally reflects fewer producers.
+
 To determine if shape changes might be related to changes in the function of pots at Kiwulan, we used geochemical methods to extract and identify lipids trapped in the fabric of potsherds to identify foods that may have contributed residues absorbed into the clay [cf.
@@
 This suggests that differences in population across our three occupation phases are probably not driving variation in craft specialization.
-We also note that a smaller sample size in the Chinese period may have effects on a more standardized shape.
-However, the effect can be reduced using CV, which offers a way to robustly compare variations across uneven sample sizes, regardless of a smaller sample size in the whole dataset [@Eerkens2001, pp.498].
-Moreover, the MSLR test for equality of CVs can deal with different sample sizes as well.
 

### 2020-08-07 analysis/paper/paper.Rmd
update discussion

@@
 
+```{r poisson-linear-model}
+total_potsherds <- read.csv(here("analysis", "data", "raw_data", "Kwl_total_potsherds.csv"))
+
+# poisson model
+fit <- glm(formula = sum_pot ~ cv_pc1,
+            family = "poisson",
+            data = total_potsherds)
+
+pval <- unname(coef(summary(fit))[,'Pr(>|z|)'][2])
+# how to print with a reasonable number of significant figures:
+pval_print <- signif(pval, 4)
+
+beta <- unname(coef(summary(fit))[,'Estimate'][2])
+# how to print with a reasonable number of significant figures:
+beta_print <- signif(beta, 4)
+```
+
 # Discussion
@@
 
-Whether standardization reflects an emergence of specialized production is associated with the number of producers, which distinguishes mechanical standardization from intentional standardization [@Costin2001; @Costin1995].
-In other words, mechanical standardization depends on potter's skills and experiences that increased uniformity normally reflects fewer producers.
+Whether standardization reflects an emergence of specialized production is associated with the number of producers, which distinguishes mechanical standardization from intentional standardization expressed by stylish and functional attributes [@Costin2001; @Costin1995].
+In other words, mechanical standardization depends on potter's skills and experiences that increased morphological uniformity normally reflects fewer producers.
+The population could be an indicator to estimate potential number of producer.
+According to the historical records [@Nakamura1938, pp.
+12], the population in Kiwulan reached higher number in the 17th century as the European presence and then decreased in the Chinese period due to a movement of indigenous people to the south [@Chen2007].
+The decline in population is supported by ceramic abundance in total, which indicates less population might be a factor leading to higher standardization in the Chinese period.
 

### 2020-08-07 analysis/paper/paper.Rmd
edit discussion

@@
 
-Whether standardization reflects an emergence of specialized production is associated with the number of producers, which distinguishes mechanical standardization from intentional standardization expressed by stylish and functional attributes [@Costin2001; @Costin1995].
-In other words, mechanical standardization depends on potter's skills and experiences that increased morphological uniformity normally reflects fewer producers.
-The population could be an indicator to estimate potential number of producer.
+Whether standardization suggests an emergence of specialized production depends on the number of producers, which distinguishes mechanical standardization from intentional standardization reflected by stylish and functional attributes [@Costin2001; @Costin1995].
+In other words, mechanical standardization is defined by potter's skills and experiences that increased morphological uniformity normally relates to fewer producers.
+Population could be an indicator to estimate the potential number of producers.
 According to the historical records [@Nakamura1938, pp.
-12], the population in Kiwulan reached higher number in the 17th century as the European presence and then decreased in the Chinese period due to a movement of indigenous people to the south [@Chen2007].
-The decline in population is supported by ceramic abundance in total, which indicates less population might be a factor leading to higher standardization in the Chinese period.
-
+12], the population in Kiwulan reached a higher number in the 17th century as the European presence and decreased in the Chinese period due to movement of indigenous people to the south [@Chen2007].
+The population dynamics in historical documents corresponds to an increase in ceramic abundance in the European presence and a decrease in the Chinese period.
+Thus, we model cv values as a function of the mass of ceramics using Poisson GLM with a link function.
+The result suggests that ceramic abundance strongly predicts the cv values (β = $`r beta_print`$, p = $`r pval_print`$).
+This indicates that standardized ceramics in the Chinese period may be influenced by a small population leading to fewer producers if ceramic abundance can reflect the population at Kiwulan.
+
+In order to ensure a mechanical standardization caused by fewer producers or specialized organization, we need to investigate other factors that may contribute to intentional standardization, such as the function of ceramics.
 To determine if shape changes might be related to changes in the function of pots at Kiwulan, we used geochemical methods to extract and identify lipids trapped in the fabric of potsherds to identify foods that may have contributed residues absorbed into the clay [cf.
@@
 
-The spatial analysis (Figure \@ref(fig:kernel-plot)) for pottery samples presents multiple clusters with high densities of pottery during European presence.
+The spatial analysis (Figure \@ref(fig:kernel-plot)) can also provide information about the intensity and location of manufacturing.
+The result reveals that pottery samples presents multiple clusters with high densities of pottery during European presence.
 Hypothesis testing on spatial randomness indicates a non-randomly dispersed distribution before European contact and more extreme dispersed distribution after European presence.

### 2020-08-08 analysis/paper/paper.Rmd
update code and summary of disparity test

@@
 
-To enable re-use of materials and improve reproducibility and transparency [@Marwick2017], the entire R code [@Rlanguage2019] used for all the analysis and visualizations contained in this paper is openly available online at https://doi.org/10.17605/OSF.IO/ABVGF.
+To enable re-use of materials and improve reproducibility and transparency [@Marwick2017], the entire R code [@Rlanguage2019] used for all the analysis and visualizations contained in this paper is openly available online at https://doi.org/10.17605/OSF.IO/ABVGF [@WangandMarwick2020].
 Also in this version-controlled compendium [@Marwick2018] are the raw data for all the visualizations and tests reported here.
@@
 
+```{r morphological-disparity}
+require(dispRity)
+phases_list <-
+  list("Pre-European" = which(op.p$phase == "Pre-European"),
+       "Post-European" = which(op.p$phase == "Post-European"),
+       "Chinese" = which(op.p$phase == "Chinese"))
+
+disparity_data <-
+  dispRity.per.group(op.p$x,
+                     group = phases_list,
+                     metric = c(sum, variances))
+
+disparity <-
+  test.dispRity(disparity_data,
+                test = wilcox.test,
+                correction = "bonferroni",
+                details = TRUE)
+
+pre_chi <- signif(disparity[[2]][[1]]$p.value, 4)
+post_chi <- signif(disparity[[3]][[1]]$p.value, 4)
+```
+
 ```{r poisson-linear-model}
@@
 
-For the degree of shape standardization across periods, our CV tests on PC1 indicate that there are significant differences between Chinese contact and either pre-European or post-European periods.
-A test of morphological disparity also demonstrates a similar result as well (See Online Supplementary Materials), suggesting that pottery shape became more standardized after contact with the Chinese.
+For the degree of shape standardization, our CV tests on PC1 indicate a significant difference between the Chinese period and either pre-European or post-European periods.
+A test of morphological disparity also demonstrates similar results between Chinese contact and either pre-European (p = $`pre_chi`$) or post-European periods (p= $`post_chi`$) [@WangandMarwick2020], suggesting that a more standardized pottery shape after contact with the Chinese.
 In addition, we found a more homogeneous shape accompanied by standardized size, represented by the body diameter of vessels, in the Chinese period.
@@
 We note that a small sample size in the Chinese period may lead to a more standardized shape.
-However, this effect can be reduced using CV that scales variation to magnitude, allowing reliable comparison across uneven sample numbers, even at small sample size [@Eerkens2001].
+However, this effect can be reduced using CV that scales variation to magnitude, allowing reliable comparison across uneven sample numbers, even for small sample size [@Eerkens2001].
 Moreover, the MSLR test for equality of CVs enables a robust test between different sample numbers [@Krishnamoorthy2014].
@@
 According to the historical records [@Nakamura1938, pp.
-12], the population in Kiwulan reached a higher number in the 17th century as the European presence and decreased in the Chinese period due to movement of indigenous people to the south [@Chen2007].
-The population dynamics in historical documents corresponds to an increase in ceramic abundance in the European presence and a decrease in the Chinese period.
+12], the population in Kiwulan reached a higher number in the 17th century as the European presence and declined in the Chinese period due to movement of indigenous people to the south [@Chen2007].
+The changes in population size described in documents corresponds to the observation of ceramic abundance at Kiwulan.
 Thus, we model cv values as a function of the mass of ceramics using Poisson GLM with a link function.
-The result suggests that ceramic abundance strongly predicts the cv values (β = $`r beta_print`$, p = $`r pval_print`$).
-This indicates that standardized ceramics in the Chinese period may be influenced by a small population leading to fewer producers if ceramic abundance can reflect the population at Kiwulan.
+The model suggests that ceramic abundance strongly predicts the cv values (β = $`r beta_print`$, p = $`r pval_print`$).
+This indicates that a more standardized shape in the Chinese period may be influenced by a small population if ceramic abundance can reflect the population size.
 

### 2020-08-08 disparity.R

@@ -16,5 +16,13 @@ summary(disparity_data)
 
 plot(disparity_data)
 
-test.dispRity(disparity_data, test = wilcox.test,
-              correction = "bonferroni", details = TRUE)
+disparity <-
+  test.dispRity(disparity_data,
+                test = wilcox.test,
+                correction = "bonferroni",
+                details = TRUE)
+
+pre_post <- signif(disparity[[1]][[1]]$p.value, 4)
+pre_chi <- signif(disparity[[2]][[1]]$p.value, 4)
+post_chi <- signif(disparity[[3]][[1]]$p.value, 4)
+

### 2020-08-09 analysis/paper/paper.Rmd
added code and discussion of decoration

@@
 
+```{r decoration-by-phase}
+pottery_deco_table <-
+  pottery_data_tidy %>%
+  select(Pit, Layer, `No.`, phase, `紋飾/腹片`) %>%
+  mutate(`紋飾/腹片`= str_remove_all(`紋飾/腹片`, "[[:punct:]]")) %>%
+  count(phase, `紋飾/腹片`, name = "count") %>%
+  group_by(phase) %>%
+  mutate(sum = sum(count))
+
+pottery_deco_count <-
+  pottery_deco_table %>%
+  count(phase, sum, name = "decoration_types") %>%
+  mutate(ratio = round(decoration_types/sum, 2))
+
+pre_deco <- pottery_deco_count$ratio[3]
+post_deco <- pottery_deco_count$ratio[2]
+chinese_deco <- pottery_deco_count$ratio[1]
+```
+
 # Discussion
@@
 
-Whether standardization suggests an emergence of specialized production depends on the number of producers, which distinguishes mechanical standardization from intentional standardization reflected by stylish and functional attributes [@Costin2001; @Costin1995].
-In other words, mechanical standardization is defined by potter's skills and experiences that increased morphological uniformity normally relates to fewer producers.
+Whether standardization suggests an emergence of specialized production depends on the number of producers, which distinguishes mechanical standardization from intentional standardization defined by stylish and functional attributes [@Costin2001; @Costin1995].
+In other words, mechanical standardization is associated with potter's skills and experiences that increased morphological uniformity normally reflects fewer producers.
 Population could be an indicator to estimate the potential number of producers.
 According to the historical records [@Nakamura1938, pp.
-12], the population in Kiwulan reached a higher number in the 17th century as the European presence and declined in the Chinese period due to movement of indigenous people to the south [@Chen2007].
+12], the population in Kiwulan reached a higher number in the 17th century as the European presence and declined in the Chinese period due to movement of Indigenous people to the south [@Chen2007].
 The changes in population size described in documents corresponds to the observation of ceramic abundance at Kiwulan.
@@
 
-In order to ensure a mechanical standardization caused by fewer producers or specialized organization, we need to investigate other factors that may contribute to intentional standardization, such as the function of ceramics.
-To determine if shape changes might be related to changes in the function of pots at Kiwulan, we used geochemical methods to extract and identify lipids trapped in the fabric of potsherds to identify foods that may have contributed residues absorbed into the clay [cf.
+However, intentional standardization could also contribute to shape standardization in our case.
+To ensure the standardization is caused by fewer producers or specialized organization, we investigate other factors that may contribute to shape standardization, such as the function and decoration of ceramics.
+To determine the function of pots in relation to shape changes, we used geochemical methods to extract and identify lipids trapped in the fabric of potsherds to identify foods that may have contributed residues absorbed into the clay [cf.
 @Kwak2015].
@@
 909].
+Stylish attributes present a wide variation in geometric motifs from pot to pot across three phases.
+In general, the ceramics in the Chinese period have less types of motif according to the ratio of distinct motif to the total number for each phase (Chinese = `r chinese_deco`, post-European = `r post_deco`, pre-European = `r pre_deco`) [@WangandMarwick2020].
 

### 2020-08-09 analysis/paper/paper.Rmd
edit discussion and conclusion

@@
 Stylish attributes present a wide variation in geometric motifs from pot to pot across three phases.
-In general, the ceramics in the Chinese period have less types of motif according to the ratio of distinct motif to the total number for each phase (Chinese = `r chinese_deco`, post-European = `r post_deco`, pre-European = `r pre_deco`) [@WangandMarwick2020].
-
-The spatial analysis (Figure \@ref(fig:kernel-plot)) can also provide information about the intensity and location of manufacturing.
-The result reveals that pottery samples presents multiple clusters with high densities of pottery during European presence.
+In general, the ceramics in the Chinese period have slightly fewer types of motifs according to the ratio of distinct motifs to the total number for each phase (Chinese = `r chinese_deco`, post-European = `r post_deco`, pre-European = `r pre_deco`) [@WangandMarwick2020].
+The evidence shows that we could not completely rule out the intentional standardization in our sample.
+
+The spatial pattern of ceramics could provide important information about the intensity and location of manufacturing or consumption.
+The spatial analysis (Figure \@ref(fig:kernel-plot)) reveals that pottery samples presents multiple clusters with high densities of pottery during European presence.
 Hypothesis testing on spatial randomness indicates a non-randomly dispersed distribution before European contact and more extreme dispersed distribution after European presence.
@@
 We might expect reduced numbers of potters to result in pottery production shrinking to a few locations in the settlement during this time.
-Although the overall number of vessels is smaller during the Chinese contact period, Figure \@ref(fig:kernel-plot) shows that pottery is distributed randomly across the sampling area without any distinctive clusters during this time.
-This suggests that differences in population across our three occupation phases are probably not driving variation in craft specialization.
+Despite small number of vessels during the Chinese contact period, Figure \@ref(fig:kernel-plot) shows that pottery is distributed randomly across the sampling area without any distinctive clusters during this time.
+This suggests that differences in population across our three occupation phases are probably not driving variation in shape standardization.
+Also, the pottery manufacture inclines to be household-produced since no workshops have been found.
 
@@
 
-Our results offer tentative support for the hypothesis that foreign influences at Kiwulan influenced emergence of social inequality in the local Indigenous society.
+Our results offer tentative support for the hypothesis that foreign presence at Kiwulan influenced emergence of social inequality in the local Indigenous society.
 If increased shape standardization is a reliable indicator of craft specialization, then we may be seeing evidence of a shift from corporate (group-based, distributed, collective, cooperative) to network (individual-based, competitive) organization [@Blanton1996; @Feinman1995; @Feinman2000; @Feinman2000political; @Feinman2010].
@@
 
-Compared to other regions in Taiwan, European colonial influence was weak in Yilan due to isolation by the surrounding mountains, and the economic focus of the Spanish and Dutch who preferred northern and northwest Taiwan as their trading base.
+Compared to other regions in Taiwan, European colonial influence was weak in Yilan due to isolation by the surrounding mountains, and the economic focus of the Spanish and Dutch who preferred northern and northwest Taiwan as their trading base [cf.
+@Berrocal2020].
 Indigenous communities in Yilan experienced indirect influence from European trade networks and their colonial activities in a pericolonial context [cf.
@@
 
-The shape variation reported here is subtle, and invites consideration of the possibility that the absence of major changes in vessel shape at Kiwulan may have been an act of resistance to foreign influence.
-Continuities in vessel shape over time draws our attention to the endurance of traditional pottery production practices amid intrusions from Europeans and Chinese.
+The shape variation reported here is subtle, and invites consideration of another possible scenario that the absence of major changes in vessel shape at Kiwulan may have been an act of resistance to foreign influence.
+We recognize the decline in population may lead to a standardized shape, however, the evidence of ceramic spatial distribution shows a dispersed or random pattern across the whole area throughout three phases.
+This indicates manufacturing activities may be limited to household scale and reflect a common practice in the community.
+The standardization in vessel shape over time draws our attention to the endurance of traditional pottery production practices amid intrusions from Europeans and Chinese.
 In a culture contact situation, social identity may be expressed through material practices as a means of expressing cultural homogeneity and distinction from other groups [@Voss2005].
@@
 @Roux2015].
-Although there is an increase in number of imported ceramics through time at Kiwulan, production was continuous, and increasingly standardized.
-This might imply not only the utilitarian function, but a deliberate and increased emphasis on the local ceramic tradition, their cultural custom, as a response to intensified foreign contact [cf.
+Although there is an increase in number of imported ceramics through time at Kiwulan, production of local ceramics was continuous, and increasingly standardized.
+This might imply not only the utilitarian function, but a intentional and increased emphasis on the local ceramic tradition, their cultural custom, as a response to intensified foreign contact [cf.
 @Acabado2017].
+However, we still need more lines of evidence to confirm this speculation of an act of resistance.
 
@@
 Our findings help to expand upon those factors that may lead to the standardization of ceramic production in a pericolonial interaction context.
-More homogeneous shapes during the contact periods, without any changes in clay paste composition or production technique, suggest that shape standardization was intentional.
-The results further suggest that expressions of social identity or cultural boundaries in Indigenous societies through highly visible vessel qualities, such as shape, may be heightened during periods of foreign contact in pericolonial contexts.
+More homogeneous shapes during the Chinese period, without any changes in clay paste composition, production technique, and spatial distribution, suggest that shape standardization may relate more to an intentional behavior instead of the emergence of specialized groups.
+The results further suggest that expressions of social identity or cultural boundaries in Indigenous societies through highly visible vessel qualities, such as shape, maybe heightened during periods of foreign contact in pericolonial contexts.
 Our analysis, with its openly available methods and data, is readily extensible to other pottery assemblages in the region to further explore related questions about craft specialization and standardization in Iron Age ceramic technologies.
-Furthermore, this study broadens the GMM field, where the common topics are lithic topology, bone morphology, and patterns of cutmarks or taphonomy.
-We suggest that EFA-based GMM is an effective method for detecting morphological differences in ceramics assemblages.
+Furthermore, this study broadens the GMM field to include the discussion of ceramics, other than the common topics of lithic topology, bone morphology, and patterns of cutmarks or taphonomy.
+We suggest that EFA-based GMM is an effective method for detecting morphological differences in ceramics assemblage
 

### 2020-08-09 analysis/paper/paper.Rmd
added petrography data and analysis

@@
 To compare different foreign influences, we classified the upper component into three chronological phases: pre-European, European, and Chinese.
-We identified these phases based on 32 previously published radiocarbon ages [@Chen2007], excavation depth measurements, stratigraphic details reported by the excavators (color, texture, disturbance, etc.), and finds of chronologically diagnostic artifacts, such as blue and white porcelains, light grey glazed jars, and large dark brown glazed stoneware jars commonly used in the 17th century, and bricks and tiles employed by the Chinese in the 19th century [@Chen2007; @Hsieh2009; @Wang2011].
+These phases were identified based using 32 previously published radiocarbon ages [@Chen2007], and enlisted excavation depth measurements, stratigraphic details reported by the excavators (color, texture, disturbance, etc.), and finds of chronologically diagnostic artifacts, such as blue and white porcelains, light grey glazed jars, and large dark brown glazed stoneware jars commonly used in the 17th century, and bricks and tiles employed by the Chinese in the 19th century [@Chen2007; @Hsieh2009; @Wang2011].
 The deposit exhibited signs of continuous human occupation in each of the three phases with no apparent breaks.
@@
 
-We used scans of pottery drawings acquired from the Bureau of Cultural Affairs in Yilan (Figure \@ref(fig:KWL-pots)).
+We used scans of pottery drawings at 300 dpi acquired from the Bureau of Cultural Affairs in Yilan (Figure \@ref(fig:KWL-pots)).
 All drawings provide a two-dimensional view of vessel cross-sections, and metric measurements.
@@
 For the degree of shape standardization, our CV tests on PC1 indicate a significant difference between the Chinese period and either pre-European or post-European periods.
-A test of morphological disparity also demonstrates similar results between Chinese contact and either pre-European (p = $`pre_chi`$) or post-European periods (p= $`post_chi`$) [@WangandMarwick2020], suggesting that a more standardized pottery shape after contact with the Chinese.
+A test of morphological disparity also demonstrates similar results between Chinese contact and either pre-European (p = $`r pre_chi`$) or post-European periods (p= $`r post_chi`$) [@WangandMarwick2020], suggesting that a more standardized pottery shape after contact with the Chinese.
 In addition, we found a more homogeneous shape accompanied by standardized size, represented by the body diameter of vessels, in the Chinese period.

### 2020-08-09 analysis/code/thin_section.R

@@ -1,0 +1,76 @@
+# read data
+library(readxl)
+thin_section <- read_excel(here::here("analysis", "data", "raw_data", "kwl_petrography.xlsx"), 2)
+sherd <- read_excel(here::here("analysis", "data", "raw_data", "kwl_petrography.xlsx"), 1)
+
+# sum the columns of similar minerals
+thin_section_combine <-
+  thin_section %>%
+  mutate(Quartz = `Monocrystalline quartz` + `Polycrystalline quartz`,
+         Argillite_all = `Argillite` + `Reddish-brown Argillite`)
+
+# prepare data for pca
+thin_section_combine_pca <-
+  thin_section_combine %>%
+  select(4:5, 8:12, 17:18)
+
+# PCA
+pca1 <- prcomp(thin_section_combine_pca, scale. = TRUE)
+# sqrt of eigenvalues
+pca1$sdev
+# loadings
+head(pca1$rotation)
+# PCs (aka scores)
+head(pca1$x)
+
+library(ggplot2)
+# create data frame with scores
+scores <- as.data.frame(pca1$x)
+
+# plot of observations
+ggplot(data = scores, aes(x = PC1, y = PC2, label = rownames(scores))) +
+  geom_hline(yintercept = 0, colour = "gray65") +
+  geom_vline(xintercept = 0, colour = "gray65") +
+  geom_text(colour = "tomato", alpha = 0.8, size = 4) +
+  ggtitle("PCA plot of thin sections - Mineral")
+
+
+# convex hulls for time periods by argillite and quartz
+kwl_th_hulls <-
+  thin_section_combine %>%
+  filter(!is.na(Quartz), !is.na(Argillite_all)) %>%
+  nest(-phase) %>%
+  mutate(hulls = map(data, ~.[chull(.$Quartz, .$Argillite_all),])) %>%
+  select(phase, hulls) %>%
+  unnest(hulls)
+
+ggplot(thin_section_combine,
+       aes(Quartz,
+           Argillite_all,
+           colour = phase,
+           fill = phase)) +
+  geom_point() +
+  geom_polygon(data = kwl_th_hulls,
+               alpha = 0.1,
+               color  = NA) +
+  theme_minimal()
+
+# convex hulls for time periods by argillite and metasandstone
+kwl_th_hulls2 <-
+  thin_section_combine %>%
+  filter(!is.na(Metasandstone), !is.na(Argillite_all)) %>%
+  nest(-phase) %>%
[…]

### 2020-08-10 analysis/paper/paper.Rmd
add MANOVA test

@@
 
+```{r}
+library(readxl)
+thin_section <- read_excel(here::here("analysis", "data", "raw_data", "kwl_petrography.xlsx"), 2)
+sherd <- read_excel(here::here("analysis", "data", "raw_data", "kwl_petrography.xlsx"), 1)
+
+# sum the columns of similar minerals
+thin_section_combine <-
+  thin_section %>%
+  mutate(Quartz = `Monocrystalline quartz` + `Polycrystalline quartz`,
+         Argillite_all = `Argillite` + `Reddish-brown Argillite`)
+
+# prepare data for pca
+thin_section_combine_pca <-
+  thin_section_combine %>%
+  select(4:5, 8:12, 17:18)
+
+# PCA
+pca1 <- prcomp(thin_section_combine_pca, scale. = TRUE)
+# add phase
+pca1$phase <- as.factor(thin_section_combine$phase)
+
+# MANOVA
+man_pca <- manova(pca1$x ~pca1$phase)
+sum_man_all <- summary(man_pca, test="Pillai")
+
+# get p value
+sum_man_all_df <-
+  as.data.frame(sum_man_all[4]) %>%
+  mutate_if(is.numeric, round, 4) %>%
+  select(-`stats.Df`, -`stats.num.Df`) %>%
+  rename(`Pillai's trace` = `stats.Pillai`,
+         `Approximate F value` = `stats.approx.F`,
+         `degrees of freedom` = `stats.den.Df`,
+         `Pr(>F)` = `stats.Pr..F.`)
+
+pval_all <- signif(unname(sum_man_all_df[,'Pr(>F)'][1]), 4)
+```
+
 The archaeological record of Kiwulan's upper component shows traces of foreign contact, including Europeans in the 17th century, and waves of Chinese immigrants in the 19th century.
@@
 This composition is consistent with the mineralogical composition of local raw materials found in the Yilan Plain [@Chen2016].
-There are no substantial changes in the inclusions over time, indicating continuity in pottery fabric composition across the three periods.
-<!--Was this tested using statistics?
-If so, then state that there are no significant changes in the inclusions over time, and include that script in your supplementary materials.-->
+There are no significant changes in the inclusions over time, indicating continuity in pottery fabric composition across the three periods (p = $`r pval_all`$) [@WangandMarwick2020].
 

### 2020-08-10 analysis/code/thin_section.R

@@ -14,27 +14,6 @@ thin_section_combine_pca <-
   thin_section_combine %>%
   select(4:5, 8:12, 17:18)
 
-# PCA
-pca1 <- prcomp(thin_section_combine_pca, scale. = TRUE)
-# sqrt of eigenvalues
-pca1$sdev
-# loadings
-head(pca1$rotation)
-# PCs (aka scores)
-head(pca1$x)
-
-library(ggplot2)
-# create data frame with scores
-scores <- as.data.frame(pca1$x)
-
-# plot of observations
-ggplot(data = scores, aes(x = PC1, y = PC2, label = rownames(scores))) +
-  geom_hline(yintercept = 0, colour = "gray65") +
-  geom_vline(xintercept = 0, colour = "gray65") +
-  geom_text(colour = "tomato", alpha = 0.8, size = 4) +
-  ggtitle("PCA plot of thin sections - Mineral")
-
-
 # convex hulls for time periods by argillite and quartz
 kwl_th_hulls <-
   thin_section_combine %>%
@@ -44,6 +23,7 @@ kwl_th_hulls <-
   select(phase, hulls) %>%
   unnest(hulls)
 
+library(ggplot2)
 ggplot(thin_section_combine,
        aes(Quartz,
            Argillite_all,
@@ -74,3 +54,64 @@ ggplot(thin_section_combine,
                alpha = 0.1,
                color  = NA) +
   theme_minimal()
+
+# PCA
+pca1 <- prcomp(thin_section_combine_pca, scale. = TRUE)
+# sqrt of eigenvalues
+pca1$sdev
+# loadings
+head(pca1$rotation)
+# PCs (aka scores)
+head(pca1$x)
+pca1$phase <- as.factor(thin_section_combine$phase)
+
+# create data frame with scores
+scores <- as.data.frame(pca1$x)
+# add phase
+scores$phase <- as.factor(thin_section_combine$phase)
+
+pca_hulls <-
+  scores %>%
+  nest(-phase) %>%
+  mutate(hulls = map(data, ~.[chull(.$PC1, .$PC2),])) %>%
+  select(phase, hulls) %>%
+  unnest(hulls)
+
+# plot of PC values
+ggplot(data = scores,
+       aes(x = PC1, y = PC2, color = phase, fill = phase)) +
+  geom_hline(yintercept = 0, color = "gray65") +
[…]

### 2020-08-11 analysis/paper/paper.Rmd
edit chronology section

@@
 Pericolonial archaeology is the study of these indirect effects of colonialism, investigating areas where direct European colonial rule was limited, their conquests were often short-lived and unsuccessful, but commercial activities yielded economic and political impacts on Indigenous peoples living on the periphery of colonial control [@Acabado2017; @Trabert2017].
-Pericolonial situations were common during the seventeenth to nineteenth centuries in East and Southeast Asia where European trading activity was extensive, but direct European rule less widespread.
+Pericolonial situations were common during the 17th to 19th centuries in East and Southeast Asia where European trading activity was extensive, but direct European rule less widespread.
 An emerging priority in archaeological research in Asia is identifying the indirect influences are apparent on Indigenous communities during the colonial period.
@@
 
-Indigenous societies' responses to colonial contact ranges from passive acceptance to active negotiation with the colonists, and accommodation or resistance of foreign intrusion, which can be identified through their daily cultural practices, such as their consumption patterns of foreign goods [@Dietler2015; @Given2004; @Mullins2011; @Scaramelli2005; @Silliman2001; @TorrenceandClarke2000].
+Indigenous societies' responses to colonial contact ranges from passive acceptance to active negotiation with the colonists, and accommodation or resistance of foreign intrusion [@TorrenceandClarke2000].
+The responses can be identified through their daily cultural practices, such as their consumption patterns of foreign goods [@Dietler2015; @Given2004; @Mullins2011; @Scaramelli2005; @Silliman2001].
 In this paper we investigate the archaeology of a pericolonial situation at Kiwulan (ca.
@@
 After the procedure of GPA, superimposed landmark coordinates become shape variables that allow further statistical analyses [@Slice2007].
-A common procedure is using dimensional reduction techniques, such as Principal Components Analysis or Canonical Variate Analysis, to capture the key features that represent overall shape.
+A common procedure is using dimensional reduction techniques, such as Principal Components Analysis or Canonical Variate Analysis, to capture the key features that represent the overall shape.
 Visualization of the reduced data enables the identification of groups, followed by statistical tests to robustly distinguish them.
 Landmark-based morphometrics have been widely applied to archaeological objects with obvious morphological features that provide unambiguous reference points for landmark placement, such as tips and edges of stone tools or iron weapons [@Birch2019; @Lycett2013], visually distinctive bone features [@Haruda2019; @Meloro2015], or ceramic assemblages with distinct components [@Selden2019; @Topi2017].
-GMM applications can answer research questions related to lithic typological and technological change [@Doyon2019; @Eren2015; @Selden2018lithic; @Perez2007; @Presnyakova2018], animal domestication or mobility [@Haruda2019; @Owen2014], or hominid activities through cutmarks and taphonomic traces [@Aramendi2019; @Courtenay2019].
+GMM applications often answer research questions related to lithic typological and technological change [@Doyon2019; @Eren2015; @Selden2018lithic; @Perez2007; @Presnyakova2018], animal domestication or mobility [@Haruda2019; @Owen2014], or hominid activities through cutmarks and taphonomic traces [@Aramendi2019; @Courtenay2019].
 
@@
 The harmonic power, a cumulative sum of squared harmonic coefficients, provides a robust rule for determining the desired number of harmonics [@Bonhomme2014].
-The first systematic use of  Fourier series to analyze shapes of artifacts in archaeology was @Gero1984's study of lithic flakes in Peru that suggests the changes in shape from a more angular to rounded tool over time.
-Later, @Saragusti2005 introduced more functions for the further calculation of specific shape attributes, including symmetry, roughness, and deformation, which demonstrates the potential of EFA for analysis of curves.
-To compare resharpening trajectories in European Middle Paleolithic stone tools, @Iovictua2009 presented a protocol from outline digitization, EFA procedure, to multivariate linear regression for shape variables.
-Recent lithic studies desmotrate the wide range of application of EFA to answer questions about typological classification of Late Woodland points [@Fox2015], the function of flaked obsidian tools in Easter Island [@Lipo2016], shape and symmetry standardization of the British Acheulean [@Hoggard2019], and cultural taxonomies for the European Late Palaeolithic [@Ivanovaite2020].
-
-Ceramic studies using EFA is rare, however, this approach is promising for answering questions related to ceramic taxonomy and standardization.
+The first systematic use of Fourier series to analyze shapes of artifacts in archaeology was @Gero1984's study of lithic flakes in Peru that suggested the changes in shape from a more angular to rounded tool over time.
+Later, @Saragusti2005 introduced more functions for further calculation of specific shape attributes, including symmetry, roughness, and deformation, which demonstrates the potential of EFA for analysis of curves.
+@Iovictua2009 presented a protocol from outline digitization, EFA procedure, to multivariate linear regression to compare resharpening trajectories of European Middle Paleolithic stone tools, and found resharpening can be independent of morphology.
+Recent lithic studies demonstrate the usefulness of EFA to answer questions about the typological classification of Late Woodland points [@Fox2015], the function of flaked obsidian tools in Easter Island [@Lipo2016], the shape and symmetry standardization of the British Acheulean [@Hoggard2019], and the cultural taxonomies for the European Late Palaeolithic [@Ivanovaite2020].
+
+Although ceramic studies using EFA is rare, this approach is promising for examining ceramic taxonomy and standardization.
 For example, @Wilczek2014 evaluate the concordance between EFA and Discrete Cosine Transform (DCT), and a traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France.
@@
 library(readxl)
+library(tidyverse)
 thin_section <- read_excel(here::here("analysis", "data", "raw_data", "kwl_petrography.xlsx"), 2)
@@
 To compare different foreign influences, we classified the upper component into three chronological phases: pre-European, European, and Chinese.
-These phases were identified based using 32 previously published radiocarbon ages [@Chen2007], and enlisted excavation depth measurements, stratigraphic details reported by the excavators (color, texture, disturbance, etc.), and finds of chronologically diagnostic artifacts, such as blue and white porcelains, light grey glazed jars, and large dark brown glazed stoneware jars commonly used in the 17th century, and bricks and tiles employed by the Chinese in the 19th century [@Chen2007; @Hsieh2009; @Wang2011].
+These phases were identified according to chronologically diagnostic artifacts due to the fined-grained time scale compared to radiocarbon ages, which tend to have long time ranges that can only serve as a cross-validation if applicable.
+The diagnostic artifacts include blue and white porcelains, light grey glazed jars, and large dark brown glazed stoneware jars commonly used in the 17th century, and bricks and tiles employed by the Chinese in the 19th century [@Chen2007; @Hsieh2009; @Wang2011].
+We also enlisted excavation depth measurements, stratigraphic details reported by the excavators (color, texture, disturbance, etc.) to reliably separate three phases.
 The deposit exhibited signs of continuous human occupation in each of the three phases with no apparent breaks.
+More details of the assignment of the time periods are in the Online Supplementary Materials.
 
@@
 Firing resulted in orange and brownish color with a fully oxidized core, or a reduced core with oxidized fringes [@Chen2007].
-Finger impressions and seams, usually on the interior, indicate that the vessels were shaped by hand, and that the seems between the slabs were pinched.
+Finger impressions and seams, usually on the interior, indicating that the vessels were shaped by hand, and that the seems between the slabs were pinched.
 This kind of vessel has been widely found at archaeological sites during the late Iron Age and the historical period throughout the Yilan Plain [@NMTH2005].
@@
 Petrographic analysis for 34 thin sections presents a high percentage of inclusions (15-50%), including argillite (15-40%), metasandstone (1-10%), sandstone fragments (1-6%), quartz (1-5%), and trace amounts of feldspar and slate.
-Particle sizes of range from 500 to 1300 microns.
-In general, the globular vessel fabric presents a mixture of fine, rounded argillite with a small amounts of rounded metasandstone and rounded to sub-angular monocrystalline quartz.
+Particle sizes range from 500 to 1300 microns.
+In general, the globular vessel fabric presents a mixture of fine, rounded argillite with a small amount of rounded metasandstone and rounded sub-angular monocrystalline quartz.
 This composition is consistent with the mineralogical composition of local raw materials found in the Yilan Plain [@Chen2016].
@@
 # Once we have the jpgs, start from here
-library(tidyverse)
-
 # read in pottery data
@@
 Following @Eerkens2001 and @Roux2003, we take this as our measurement of standardization in vessel shape variables: lower CV values reflect higher standardization, and thus increased craft specialization in the community.
-Given that CV is robust for positive values due to the representation by a ratio scale, we normalized PC scores to a range between 1 and 10 before computation of CV.
+Given that CV is robust for positive values due to the representation by a ratio scale, we normalized PC scores to a range between 1 and 10 before the computation of CV.
 
@@
 # save figure
-source(here("analysis/paper/my_tps_grid.R"))
+source(here("analysis/code/my_tps_grid.R"))
 png(here("analysis/figures/tps-iso.png"), res = 300, w = 1800, h = 1000)
@@
 PC2 relates to the neck and mouth constriction, from narrow to wide.
-PC3 explains a smaller portion of the variance (`r pc3`%), which relates to the degree of the flare in the neck, from a curved to straight shape.
+PC3 explains a smaller portion of the variance (`r pc3`%), which relates to the degree of the flare in the neck, from a curved to a straight shape.
 The results reflect a large overlap in shapes from three occupations phases, especially for shapes in the pre-European and post-European periods.
@@
 The CVs calculated of the three PC support a general trend toward a more standardized shape over time, especially the shape identified by PC1.
-The PC1, capturing vessel height and roundness, shows higher variation in the pre-European period and post-European period compared to the Chinese period.
+The PC1, capturing vessel height and roundness, shows a higher variation in the pre-European period and post-European period compared to the Chinese period.
 In other words, shape standardization was higher in the Chinese period.
@@
 The body diameter of vessels from the Chinese period is larger than those from the two earlier periods, and the vessels before European contact have the smallest body diameter on average (Figure \@ref(fig:size-shape-regression-plots): A).
-To investigate standardization of vessel form reflected by both shape and size, we calculated CV for PC1 (shape) in relation to CV for body diameter (size).
-The result (Figure \@ref(fig:size-shape-regression-plots): B) presents a higher form standardization in the Chinese period according to smaller CV values compared to CV values from the other two phases.
-However, there is no big differences in form standardization between and after the European presence.
+To investigate vessel form standardization reflected by both shape and size, we calculated CV for PC1 (shape) in relation to CV for body diameter (size).
+The result (Figure \@ref(fig:size-shape-regression-plots): B) presents a higher standardization in vessel form in the Chinese period according to smaller CV values compared to CV values from the other two phases.
+However, there are no big differences in form standardization between and after the European presence.
 To understand the relationship between shape and size, we created linear regression models for PCs and the measurement of body diameter, with an examination of Pearson's correlation coefficient and p-value (Figure \@ref(fig:size-shape-regression-plots): C).
@@
 
+<!-- a summary of results -->
+
 Previous investigations at Kiwulan suggested an unequal distribution of prestige goods, trade ornaments in particular, following the appearance of Europeans with a high diversity of types [@Cheng2008; @Wang2011].
@@
 Oval-shaped vessels with a narrower opening tended to have a larger body diameter, indicative of a change in overall vessel form.
+
+<!-- standardization and the issue of small sample size -->
 
@@
 
+<!-- does standardization reflect specialization? -->
+
 Whether standardization suggests an emergence of specialized production depends on the number of producers, which distinguishes mechanical standardization from intentional standardization defined by stylish and functional attributes [@Costin2001; @Costin1995].
 In other words, mechanical standardization is associated with potter's skills and experiences that increased morphological uniformity normally reflects fewer producers.
-Population could be an indicator to estimate the potential number of producers.
+The population at Kiwulan could be an indicator to estimate the potential number of producers.
 According to the historical records [@Nakamura1938, pp.
 12], the population in Kiwulan reached a higher number in the 17th century as the European presence and declined in the Chinese period due to movement of Indigenous people to the south [@Chen2007].
-The changes in population size described in documents corresponds to the observation of ceramic abundance at Kiwulan.
+The changes in population size described in documents correspond to the observation of ceramic abundance at Kiwulan.
 Thus, we model cv values as a function of the mass of ceramics using Poisson GLM with a link function.
@@
 In contrast, the distribution of pottery is more similar to random distributions during the Chinese period.
-This contradicts our expectation that clustered pattern will be observed with an increase in pottery standardization since the emergence of specialized groups.
+This contradicts our expectation that a clustered pattern will be observed with an increase in pottery standardization since the emergence of specialized groups.
 The absence of clusters in the Chinese period is notable because this was a time of a historically-documented decline of Indigenous population [@Chen2007; @Hsieh2009].
 We might expect reduced numbers of potters to result in pottery production shrinking to a few locations in the settlement during this time.
-Despite small number of vessels during the Chinese contact period, Figure \@ref(fig:kernel-plot) shows that pottery is distributed randomly across the sampling area without any distinctive clusters during this time.
+Despite the small number of vessels during the Chinese contact period, Figure \@ref(fig:kernel-plot) shows that pottery is distributed randomly across the sampling area without any distinctive clusters during this time.
 This suggests that differences in population across our three occupation phases are probably not driving variation in shape standardization.
-Also, the pottery manufacture inclines to be household-produced since no workshops have been found.
+Also, the pottery manufacture inclines to be household-produced since no workshops have been found [@Chen2007].
 
@@
 
-Our results offer tentative support for the hypothesis that foreign presence at Kiwulan influenced emergence of social inequality in the local Indigenous society.
+Our results offer tentative support for the hypothesis that foreign presence at Kiwulan influenced the emergence of social inequality in the local Indigenous society.
 If increased shape standardization is a reliable indicator of craft specialization, then we may be seeing evidence of a shift from corporate (group-based, distributed, collective, cooperative) to network (individual-based, competitive) organization [@Blanton1996; @Feinman1995; @Feinman2000; @Feinman2000political; @Feinman2010].
@@
 In a culture contact situation, social identity may be expressed through material practices as a means of expressing cultural homogeneity and distinction from other groups [@Voss2005].
-It is also important to recognize that social identity might be more complicated in a colonial context, and may be representative of more than a colonized–colonizer or local/foreign dichotomy [@Voss2005; @Voss2008].
+
+It is also important to recognize that social identity might be more complicated in a colonial context, and maybe representative of more than a colonized–colonizer or local/foreign dichotomy [@Voss2005; @Voss2008].
 Shamaoshan cemetery dating from BC 3 to AD 4 in Southwest China suggests that the process of the incorporation of Southwest China into the Han Empire involved a century of conflicts, resistance, and acceptance among social groups with different identities, especially in the historical context of Han immigrants [@Wu2019].
 A similar dynamic may have occurred at Kiwulan, with vessel shape indicating both acceptance of foreign influence through increased shape standardization, and resistance through the overall continuity in vessel shape.
-Vessel shape can be viewed as a symbolic expression the Indigenous identity and social boundaries because shape is a highly visible trait compared with other features of pottery [cf.
+Vessel shape can be viewed as a symbolic expression of the Indigenous identity and social boundaries because shape is a highly visible trait compared with other features of pottery [cf.
 @Roux2015].
 Although there is an increase in number of imported ceramics through time at Kiwulan, production of local ceramics was continuous, and increasingly standardized.
-This might imply not only the utilitarian function, but a intentional and increased emphasis on the local ceramic tradition, their cultural custom, as a response to intensified foreign contact [cf.
+This might imply not only the utilitarian function, but an intentional and increased emphasis on the local ceramic tradition, their cultural custom, as a response to intensified foreign contact [cf.
 @Acabado2017].
@@
 This study demonstrated the first use of EFA on ceramic shapes to explore the emergence of ceramic specialization as indicative of foreign influences.
-Here, EFA is combined with a significance test for the equality of CVs of shape variables to provide a robust method of identifying and statistically assessing differences in shape standardization.
+Here, EFA is combined with a significance test for the equality of CVs of shape variables to provide a robust method for assessing differences in shape standardization.
 The direct relationship between foreign influences and standardization of ceramic shape was tested on ceramics from Kiwulan, a large Iron Age Indigenous settlement in northeastern Taiwan.
-Lower variation in ceramic shape was identified after European presence began, and even lower variation during the period of Chinese presence.
+Much lower variation in ceramic shape was identified during the period of Chinese presence.
 Our findings help to expand upon those factors that may lead to the standardization of ceramic production in a pericolonial interaction context.
-More homogeneous shapes during the Chinese period, without any changes in clay paste composition, production technique, and spatial distribution, suggest that shape standardization may relate more to an intentional behavior instead of the emergence of specialized groups.
-The results further suggest that expressions of social identity or cultural boundaries in Indigenous societies through highly visible vessel qualities, such as shape, maybe heightened during periods of foreign contact in pericolonial contexts.
-Our analysis, with its openly available methods and data, is readily extensible to other pottery assemblages in the region to further explore related questions about craft specialization and standardization in Iron Age ceramic technologies.
-Furthermore, this study broadens the GMM field to include the discussion of ceramics, other than the common topics of lithic topology, bone morphology, and patterns of cutmarks or taphonomy.
-We suggest that EFA-based GMM is an effective method for detecting morphological differences in ceramics assemblage
+More homogeneous vessel shapes and sizes during the Chinese period, without any changes in clay paste composition, production technique, and spatial distribution, suggest that shape standardization may relate more to intentional behavior instead of the emergence of specialized groups.
+Despite a small sample size in the Chinese period, the random distribution of ceramics does not support clustered patterns of manufacturing location, such as workshops.
+Instead, the production of ceramic is likely to be produced in households.
+The homogeneous appearance may suggest expressions of social identity or cultural boundaries in Indigenous societies through highly visible vessel qualities, such as shape, maybe heightened during periods of foreign contact in pericolonial contexts.
+Our analysis, with its openly available methods and data, is readily extensible to other pottery assemblages in the region to further explore related questions about craft specialization and standardization in Iron Age ceramic assemblages.
+This study also broadens the GMM field by focusing on ceramic technologies, which could motivate more ceramic studies and become a promising branch parallel to current applications to lithic topology and bone morphology.
 
@@
 We thank Dr.
-Julian Sachs in the Department of Oceanography, University of Washington for his supports and providing his lab for us to conduct organic geochemistry analysis of potsherds.
+Julian Sachs in the Department of Oceanography, University of Washington for his supports and suggestions for conducting organic geochemistry analysis of potsherds.
 We thank Dr.

### 2020-08-12 analysis/paper/paper.Rmd
edit intro and GMM

@@
 
-A major historical factor often linked to social change in small-scale societies is the introduction of foreign or exotic trade goods to local Indigenous societies [@Mullins2011].
+A major historical factor of social change in small-scale societies is often linked to the introduction of foreign or exotic trade goods to local Indigenous societies [@Mullins2011].
 Monopolization of long-distance trade goods has caused substantial transformations in Indigenous economic, cultural, and socio-political systems [@Dietler2005; @Dietler1997; @Junker1993; @Silliman2005].
@@
 Pericolonial situations were common during the 17th to 19th centuries in East and Southeast Asia where European trading activity was extensive, but direct European rule less widespread.
-An emerging priority in archaeological research in Asia is identifying the indirect influences are apparent on Indigenous communities during the colonial period.
+An emerging priority in archaeological research in Asia is identifying the indirect influences that are apparent on Indigenous communities during the colonial period.
 For example, @Acabado2017’s study of Ifugao society in the Philippines highland suggests economic and political intensification during the Spanish presence in the lowlands as a strategy of Indigenous peoples to resist Spanish conquests.
@@
 In this paper we investigate the archaeology of a pericolonial situation at Kiwulan (ca.
-AD 1350-1850), a large multi-component archaeological site in Yilan County, northeastern Taiwan [@Chen2007], to identify the indirect impacts of colonial settler activity on local Indigenous societies.
-Yilan is an ideal context to study peripheral colonial influences because the Indigenous communities were isolated by geographical barriers, limiting the frequency of direct contact with the Spanish and the Dutch settlers in northern Taiwan.
+AD 1350-1850) [@Chen2007], a large multi-component archaeological site in Yilan County, northeastern Taiwan, to identify the indirect impacts of colonial settler activity on local Indigenous societies.
+Yilan is an ideal context to study peripheral colonial influences because the Indigenous communities were isolated by geographical barriers, limiting the frequency of direct contact with the Spanish and the Dutch settlers in northern Taiwan [cf.
+@Berrocal2020].
 Kiwulan is situated on a hill near a riverside at the northern margin of Yilan County, which is characterized by a triangular alluvial plain facing east toward the Pacific with high mountains on three other sides.
 
-This research investigates if there was increasing ceramic specialization resulting from Indigenous interaction with Europeans in the 17th century or Chinese in the 19th century, two major foreign influences in early historical Taiwan, which may indicate social changes in local Indigenous societies [cf.
-@Berrocal2020].
+This research investigates if there was increasing ceramic specialization resulting from Indigenous interaction with Europeans in the 17th century or Chinese in the 19th century, two major foreign influences in early historical Taiwan, which may indicate social changes in Indigenous societies.
 We predict that competition within the Indigenous community at Kiwulan for foreign resources and trade partnerships with European or Chinese colonizers may have led to the emergence of craft specialization caused by greater economic and social control of ceramic production by a small group of individuals.
 Using standardization in ceramic shapes as a proxy for craft specialization, we ask: Did colonial trade impact the shape of locally-produced Indigenous pottery vessels?
-Did pottery shape become more homogeneous and standardized after foreign contacts with European colonizers or Chinese immigrants?
+Did pottery shape become more statistically homogeneous after foreign contacts with European colonizers or Chinese immigrants?
 
@@
 Visualization of the reduced data enables the identification of groups, followed by statistical tests to robustly distinguish them.
-Landmark-based morphometrics have been widely applied to archaeological objects with obvious morphological features that provide unambiguous reference points for landmark placement, such as tips and edges of stone tools or iron weapons [@Birch2019; @Lycett2013], visually distinctive bone features [@Haruda2019; @Meloro2015], or ceramic assemblages with distinct components [@Selden2019; @Topi2017].
-GMM applications often answer research questions related to lithic typological and technological change [@Doyon2019; @Eren2015; @Selden2018lithic; @Perez2007; @Presnyakova2018], animal domestication or mobility [@Haruda2019; @Owen2014], or hominid activities through cutmarks and taphonomic traces [@Aramendi2019; @Courtenay2019].
+Landmark-based morphometrics have been widely applied to archaeological objects with obvious morphological features that provide unambiguous reference points for landmark placement, such as tips and edges of stone or metal tools [@Birch2019; @Lycett2013], visually distinctive bone features [@Haruda2019; @Meloro2015], or ceramic assemblages with distinct components [@Selden2019; @Topi2017].
+GMM is often used to answer research questions related to lithic typological and technological change [@Doyon2019; @Eren2015; @Selden2018lithic; @Perez2007; @Presnyakova2018], animal domestication or mobility [@Haruda2019; @Owen2014], or hominid activities through cutmarks and taphonomic traces [@Aramendi2019; @Courtenay2019].
 
 Key questions in archaeological shape analysis normally involve measuring shape standardization over time, or between geographical areas.
-Standardization is often investigated using multivariate analysis of shape variables computed from landmark data, or coefficients of variation on associated metric data, especially for lithic assemblages.
+Standardization is often investigated using multivariate analysis of shape variables computed from landmark data, coupled with coefficients of variation on associated metric data, especially for lithic assemblages.
 For example, @Archer2015's case study of stone points in Southern Africa suggests an increase in shape standardization over time that may relate to increased maintenance of finished points.
-@Buchanan2018 studied metric data and GMM to identify a more uniform base-shape of Folsom points than Clovis points across the western US [@Buchanan2019].
-Also using metric data and GMM, @Smith2016 found standardized bases of fluted point in Alaska and northern Yukon that might indicate a risk management strategy to ensure the ease of replacement during long-distance travel.
+@Buchanan2018 analyzed lithic morphology with metric data and identified a more uniform base-shape of Folsom points compared to Clovis points across the western US [@Buchanan2019].
+With similar methods, @Smith2016 found standardized bases of fluted point in Alaska and northern Yukon that might indicate a risk management strategy to ensure the ease of replacement during long-distance travel.
 Other factors, such as low levels of cultural innovation in a small group, could also lead to an increase in standardization of point shapes [@Okumura2014].
@@
 
-For ceramics, @Topi2017 identified standardized vessel shapes from the Casas Grandes culture in northwest Mexico using coefficients of variation for the positions of semi-landmarks across shape groups.
+For ceramics, @Topi2017 identified that two types of the Casas Grandes vessels in northwest Mexico tend to have standardized shapes using coefficients of variation for the positions of semi-landmarks across shape groups.
 They suggested standardization might hint the presence of specialized producers, reflecting social complexity.
-Another way to explore standardization is pairwise testing of variations in morphological disparity among groups of shape variables based on the calculation of distances between groups in morphospace they occupy [@Wills2001].
-In this manner, @Selden2018ceramic2 examined Caddo ceramics using semi-landmark approaches in northeast Texas and found a gradual increase in shape standardization over time that provides a basis for further discussion of craft specialization or group identity [@Selden2019].
-Similar applications to lithic assemblages suggests that the Gahagan bifaces from the central Texas exhibit less size standardization than those from the southern Caddo area, indicative of different uses [@Selden2020].
-Other applications, such as the study of cranial deformation, demonstrate that landmark approaches with multivariate analyses of shape variances are an effective method to evaluate shape standardization [@Kuzminsky2016; @Natahi2019; @Perez2007].
-
-Despite the wide use of landmark approaches in archaeology, a key limitation is that landmarks may be difficult to reproducibly locate for structures that are mostly or entirely curves if not mathematically-defined.
-In those cases, outline approaches, such as Elliptic Fourier Analysis (EFA), is better for assessing morphological variations in the whole structure of two-dimensional closed shapes [@Cardillo2010].
+Another way to explore standardization is pairwise testing of variations in morphological disparity between shape groups by calculating their distances in morphospace, an n-dimensional space where shape groups occupy [@Wills2001].
+In this manner, @Selden2018ceramic2 examined Caddo ceramics in northeast Texas using semi-landmark approaches and found am increase in shape standardization over time, providing a basis for further discussion of craft specialization or group identity [@Selden2019].
+Similarly, a lithic case study suggests that the Gahagan bifaces from the central Texas exhibit less size standardization than those from the southern Caddo area, indicating different uses or tool types [@Selden2020].
+Other applications, such as the study of cranial deformation, demonstrate that landmark approaches with multivariate analyses of shape variances are useful to evaluate shape standardization [@Kuzminsky2016; @Natahi2019; @Perez2007].
+
+A key limitation of landmark approaches in archaeology is that landmarks may be difficult to reproducibly locate for structures that are mostly or entirely curves if not mathematically-defined.
+In those cases, outline approaches, such as Elliptic Fourier Analysis (EFA), is more effective for assessing morphological variations in the whole structure of two-dimensional closed shapes [@Cardillo2010].
 EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions [@Adams2004; @Bonhomme2014; @Claude2008].
-In other words, EFA converts coordinates along a curve into Fourier function coefficients, called harmonic coefficients or harmonics [@Kuhl1982].
+That is, coordinates along a curve are converted into Fourier function coefficients, called harmonic coefficients or harmonics [@Kuhl1982].
 The number of harmonics determines the quality and precision of the geometric representation of an object.
 The harmonic power, a cumulative sum of squared harmonic coefficients, provides a robust rule for determining the desired number of harmonics [@Bonhomme2014].
-The first systematic use of Fourier series to analyze shapes of artifacts in archaeology was @Gero1984's study of lithic flakes in Peru that suggested the changes in shape from a more angular to rounded tool over time.
-Later, @Saragusti2005 introduced more functions for further calculation of specific shape attributes, including symmetry, roughness, and deformation, which demonstrates the potential of EFA for analysis of curves.
-@Iovictua2009 presented a protocol from outline digitization, EFA procedure, to multivariate linear regression to compare resharpening trajectories of European Middle Paleolithic stone tools, and found resharpening can be independent of morphology.
-Recent lithic studies demonstrate the usefulness of EFA to answer questions about the typological classification of Late Woodland points [@Fox2015], the function of flaked obsidian tools in Easter Island [@Lipo2016], the shape and symmetry standardization of the British Acheulean [@Hoggard2019], and the cultural taxonomies for the European Late Palaeolithic [@Ivanovaite2020].
-
-Although ceramic studies using EFA is rare, this approach is promising for examining ceramic taxonomy and standardization.
-For example, @Wilczek2014 evaluate the concordance between EFA and Discrete Cosine Transform (DCT), and a traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France.
-The results show the variation indicated by EFA and DCT matches the traditional ceramic typology, which supports that claim that outline-based approaches can be efficiently used for studying variations in ceramic shapes.
-Furthermore, @Wilczek2014 point out the potential of EFA for detecting the level of ceramic standardization.
-EFA is suitable for analyzing shapes lacking representative landmarks, or where curves contain the most meaningful variation.
-In such cases, the focus is the whole shape instead of the different components of a shape.
-We use EFA to evaluate the level of standardization of ceramics data from Kiwulan, northeastern Taiwan, in relation to the European presence in the 17th century to gain insights into the emergence of ceramic specialization.
-The reason for using EFA is because there are no sufficient landmarks that could be identified straightforwardly due to the globular body of ceramics.
-We use a significance test for the equality of coefficients of variation of shape variables to statistically compare the vessel standardization from different periods.
+The first systematic use of Fourier series to analyze shapes of artifacts in archaeology was @Gero1984's study of lithic flakes in Peru.
+They successfully identified the changes in tool shape from a more angular to rounded shape over time.
+Later, @Saragusti2005 introduced more functions allowing the calculation of the specific shape attributes, such as symmetry, roughness, and deformation.
+This presents the potential of EFA for the analysis of curves in detail.
+@Iovictua2009 demonstrated a protocol, including outline digitization, EFA procedure, and multivariate linear regression, to compare resharpening trajectories of European Middle Paleolithic stone tools.
+They found that resharpening can be independent of morphology, suggesting functional attributes should be studied separately.
+Recent case studies further support that EFA is applicable and effective to examine lithic assemblages, such as typological classification of Late Woodland points [@Fox2015], the function of flaked obsidian tools in Easter Island [@Lipo2016], the shape and symmetry standardization of the British Acheulean [@Hoggard2019], and the cultural taxonomies for the European Late Palaeolithic [@Ivanovaite2020].
+
+Despite rare ceramic studies using EFA, this approach is promising for analyzing ceramic taxonomy and standardization.
+For example, @Wilczek2014 evaluated the concordance between EFA and Discrete Cosine Transform (DCT), and a traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France.
+The results present that the variation indicated by EFA and DCT matches the traditional ceramic typology, which supports the claim that outline-based approaches can be efficiently used for studying variations in ceramic shapes.
+Furthermore, @Wilczek2014 pointed out the potential of EFA for detecting the level of ceramic standardization.
+In this paper, we use EFA to evaluate the level of standardization of ceramics data from Kiwulan, northeastern Taiwan, in relation to the foreign colonial presences to gain insights into the emergence of ceramic specialization.
+Since our ceramic sample lacks sufficient landmarks that could be identified straightforwardly due to their globular body, EFA is a suitable method by focusing the overall shape of an artifact.
+In addition, we use a significance test for the equality of coefficients of variation of shape variables to statistically compare the vessel standardization from different periods.
 
@@
 The deposit exhibited signs of continuous human occupation in each of the three phases with no apparent breaks.
-More details of the assignment of the time periods are in the Online Supplementary Materials.
+More details for the assignment of the time periods are in the Online Supplementary Materials.
 

### 2020-08-12 analysis/paper/paper.Rmd
edit background and result

@@
 Furthermore, @Wilczek2014 pointed out the potential of EFA for detecting the level of ceramic standardization.
-In this paper, we use EFA to evaluate the level of standardization of ceramics data from Kiwulan, northeastern Taiwan, in relation to the foreign colonial presences to gain insights into the emergence of ceramic specialization.
+In this paper we use EFA to evaluate the level of standardization of ceramics data from Kiwulan, northeastern Taiwan, in relation to the foreign colonial presences to gain insights into the emergence of ceramic specialization.
 Since our ceramic sample lacks sufficient landmarks that could be identified straightforwardly due to their globular body, EFA is a suitable method by focusing the overall shape of an artifact.
@@
 Ceramics analyzed in this study come from 40 units (4m by 4m each) sampled from the central, undisturbed area of archaeological excavations at Kiwulan (Figure \@ref(fig:KWL-map); Figure \@ref(fig:KWL-sam-area)).
-The chronology of the archaeological deposits consist of two cultural components, the upper and the lower, with a sterile layer in between [@Chen2007].
-We focus on the upper component, dated from AD 1350 to 1850, spans the late Iron Age and the historical period, of which the start is defined as the European occupation in Taiwan in the early 17th century.
+The chronology of the archaeological deposits consists of two cultural components, the upper and the lower, with a sterile layer in between [@Chen2007].
+We focus on the upper component, dated from AD 1350 to 1850, spans the late Iron Age and the historical period.
+The historical period in Taiwan started with the presence of the Europeans in the early 17th century.
 The Dutch first occupied southern Taiwan in 1624, followed by the Spanish in northern Taiwan in 1626 [@Andrade2007].
@@
 To compare different foreign influences, we classified the upper component into three chronological phases: pre-European, European, and Chinese.
-These phases were identified according to chronologically diagnostic artifacts due to the fined-grained time scale compared to radiocarbon ages, which tend to have long time ranges that can only serve as a cross-validation if applicable.
+These phases were identified according to chronologically diagnostic artifacts based on their fined-grained time scale.
+Despite the presence of 17 radiocarbon ages [@Chen2007], they tend to have long time ranges that can only serve as a cross-validation if applicable.
 The diagnostic artifacts include blue and white porcelains, light grey glazed jars, and large dark brown glazed stoneware jars commonly used in the 17th century, and bricks and tiles employed by the Chinese in the 19th century [@Chen2007; @Hsieh2009; @Wang2011].
-We also enlisted excavation depth measurements, stratigraphic details reported by the excavators (color, texture, disturbance, etc.) to reliably separate three phases.
+We also examined excavation depth measurement and stratigraphic details reported by the excavators (color, texture, disturbance, etc.) to reliably separate three phases.
 The deposit exhibited signs of continuous human occupation in each of the three phases with no apparent breaks.
-More details for the assignment of the time periods are in the Online Supplementary Materials.
-
-The most abundant artifacts in the upper component are locally manufactured ceramics, which are distributed throughout the temporal sequence, and across the study site.
-More than 550,000 sherds were excavated, and around 1,200 vessels could be partially reconstructed (i.e.
+More details for the assignment of different phases are in the Online Supplementary Materials.
+
+The most abundant artifacts in the upper component are locally manufactured ceramics, which are distributed throughout the temporal sequence, and across the study area.
+More than 550,000 sherds were revealed, and around 1,200 vessels could be completely or partially reconstructed (i.e.
 complete rim or base).
@@
 The exterior surface below the neck is decorated with a variety of impressed geometric motifs.
-These vessels were potentially used for cooking indicated by evidence of charred residues and carbon deposits frequently observed on vessel interiors, and soot on vessel exteriors.
+These vessels were potentially used for cooking indicated by evidence of the frequent presence of charred residues and carbon deposits on vessel interiors, and soot on vessel exteriors.
 Firing resulted in orange and brownish color with a fully oxidized core, or a reduced core with oxidized fringes [@Chen2007].
-Finger impressions and seams, usually on the interior, indicating that the vessels were shaped by hand, and that the seems between the slabs were pinched.
+The vessels were believed to be made with pinching technique according to the hand-shaped traces on vessel interiors, such as finger impressions and seams.
 This kind of vessel has been widely found at archaeological sites during the late Iron Age and the historical period throughout the Yilan Plain [@NMTH2005].
@@
 Particle sizes range from 500 to 1300 microns.
-In general, the globular vessel fabric presents a mixture of fine, rounded argillite with a small amount of rounded metasandstone and rounded sub-angular monocrystalline quartz.
+In general, the vessel fabric presents a mixture of fine, rounded argillite with a small amount of rounded metasandstone and rounded sub-angular monocrystalline quartz.
 This composition is consistent with the mineralogical composition of local raw materials found in the Yilan Plain [@Chen2016].
@@
 We used scans of pottery drawings at 300 dpi acquired from the Bureau of Cultural Affairs in Yilan (Figure \@ref(fig:KWL-pots)).
-All drawings provide a two-dimensional view of vessel cross-sections, and metric measurements.
+All drawings provide a two-dimensional view of vessel cross-sections based on metric measurements.
 The scanned drawings were imported into Inkscape (http://inkscape.org) for digitization where outlines were manually traced.
@@
 
-The principal component (PC) scores were analyzed with a multivariate analysis of variance (MANOVA) test to identify significant differences in shapes between occupation phases.
+The principal component (PC) scores were analyzed with a multivariate analysis of variance (MANOVA) to test significant differences in shapes between occupation phases.
 We also computed coefficients of variation values (CVs) for the PCs, treating the PCs as shape variables that are more informative than linear dimensions.
@@
 
-As a standardized measure of the spread of data, coefficients of variation (CV) allows us to directly compare variation in samples measured with different units or means.
-This is useful to compare the degree of standardization for archaeological assembles and enables comparison of variation across different sample sizes [@Eerkens2001, pp.498].
+As a standardized measure of the spread of data, coefficients of variation (CV) allows a direct comparison for variation in samples measured with different units or means.
+This is useful to examine the degree of standardization for archaeological assemblages and enables comparison of variation across different sample sizes [@Eerkens2001, pp.498].
 Following @Eerkens2001 and @Roux2003, we take this as our measurement of standardization in vessel shape variables: lower CV values reflect higher standardization, and thus increased craft specialization in the community.
-Given that CV is robust for positive values due to the representation by a ratio scale, we normalized PC scores to a range between 1 and 10 before the computation of CV.
-
-To answer the question of whether CV values for vessel samples across our three occupational phases are significantly different or not, we used the Modified signed-likelihood ratio (MSLR) test for equality of CVs [@Krishnamoorthy2014].
+Given that CV is robust for positive values due to the representation by a ratio scale, we normalized PC scores to a range between 1 and 10 for the computation of CV.
+
+To answer the question of whether CV values of vessel samples across our three occupational phases are significantly different or not, we used the modified signed-likelihood ratio (MSLR) test for equality of CVs [@Krishnamoorthy2014].
 While previous work has used the @Feltz1996's asymptotic test for the equality of coefficients of variation from k populations [@Eerkens2000; @Eerkens2001; @Hoggard2017; @Lycett2008; @Okumura2014], we prefer the MSLR test for shape variables as a more recent development with lower rates of type I error, better performance with uneven sample numbers, and more power across a range of conditions [@Krishnamoorthy2014].
@@
 As craft specialization increases, pottery distribution we expect a shift from a pattern of vessels dispersed across the site to a pattern of clusters that reflects the loci of production [@Costin2001].
-We used a Monte Carlo test for randomness in spatial locations of pots to explore whether their distribution is significantly clustered or dispersed.
+We used a Monte Carlo test for randomness in spatial locations of ceramics to robustly test whether their distribution is significantly clustered or dispersed.
 
@@
 
-We found that 13 harmonics captured 99% of the total harmonic power in the elliptic Fourier coefficients of `r total_no` vessels from three phases.
-Figure \@ref(fig:first-two-PCs-byplot-and-tps-plot) illustrates significant differences in vessel shapes described using thin-plate spline warping for paired periods, pre- and post-European periods, and post-European and Chinese periods, with the greatest differences between pre-European and Chinese periods.
+Thirteen harmonics captured 99% of the total harmonic power in the elliptic Fourier coefficients of `r total_no` vessels from three discrete phases.
+Figure \@ref(fig:first-two-PCs-byplot-and-tps-plot) illustrates differences in vessel shapes described using thin-plate spline warping for paired periods, pre- and post-European periods, and post-European and Chinese periods, with the greatest differences evidenced between pre-European and Chinese periods.
 
@@
 However, the spread of shape distribution indicates a wider variation in shapes in the pre-European and post-European periods compared to those in the Chinese period along both PC1 and PC2 axes.
-In other words, we find a decrease in shape variance in the Chinese period that is evident in the shorter height and narrower mouth of vessels used in that period.
+In other words, we find a decrease in shape variance in the Chinese period evidenced in the shorter height and narrower mouth of vessels used in that period.
 
@@
 To test for differences in the distributions of shape variables indicated by the PC scores shown in Figure \@ref(fig:first-two-PCs-byplot-and-tps-plot), we used a multivariate analysis of variance (MANOVA) test to compare pairwise combinations across the three occupation phases.
-Table \@ref(tab:MANOVA-table) demonstrated the significant differences in shape between Pre-European and Post-European phases (p = $`r pval_post_pre`$), and Pre-European and Chinese phases (p = $`r pval_chi_pre`$).
-These results are consistent with the differences in the visualization of average shapes between the occupations in the thin plate splines (Figure \@ref(fig:first-two-PCs-byplot-and-tps-plot), see left).
+Table \@ref(tab:MANOVA-table) demonstrates the significant differences in shape between Pre-European and Post-European phases (p = $`r pval_post_pre`$), and Pre-European and Chinese phases (p = $`r pval_chi_pre`$).
+These results are consistent with the differences in the visualization of average shapes between the phases (Figure \@ref(fig:first-two-PCs-byplot-and-tps-plot), see left).
 Although there is considerable overlap of shape variables between the Pre-European and Post-European phases, their PC scores differ significantly.
-There is not a significant difference in vessel shapes between the Post-European and Chinese contact periods.
+There is no significant difference in vessel shapes between the Post-European and Chinese contact periods.
 
@@
 To compare pottery shape standardization across the three phases we investigated the distributions of the first three PC scores, taking the PC scores as proxy variables for vessel shape (Figure \@ref(fig:PCs-distribution)).
-The CVs calculated of the three PC support a general trend toward a more standardized shape over time, especially the shape identified by PC1.
-The PC1, capturing vessel height and roundness, shows a higher variation in the pre-European period and post-European period compared to the Chinese period.
-In other words, shape standardization was higher in the Chinese period.
+The CVs calculated of the three PC support a general trend toward a more standardized shape over time, especially the shape identified by PC1 that represents vessel height and roundness.
+The PC1 shows a higher variation in the pre-European period and post-European period compared to the Chinese period.
+That is, a more standardized shape found in the Chinese period.
 However, the PC2 presents a similar diversity in ceramics assemblages across three phases, while the PC3 demonstrates a slightly standardized shape in the Chinese period.
 
-To see whether the differences in the distribution of PCs between any two phases are substantive or due to chance, we used a modified signed-likelihood ratio test to assess the equality of CVs [@Krishnamoorthy2014; @Marwick2019].
-P-values for this significance test of CVs for PC1 show significant differences in the standardization of vessel shapes across periods, between Chinese contact with either pre-European or post-European (Table \@ref(tab:CV-test-PCs-table)).
-This result statistically supports the observation of a higher standardized shape in Chinese period.
+To see whether the differences in the distribution of PCs between any two phases are substantive or due to chance, we assessed the equality of CVs for PC1 and PC2 with a modified signed-likelihood ratio test  [@Krishnamoorthy2014; @Marwick2019].
+P-values for PC1 show significant differences in shape standardization across periods, between Chinese contact with either pre-European or post-European (Table \@ref(tab:CV-test-PCs-table)).
+This result statistically supports the observation of a higher standardized shape in the Chinese period.
 
@@
 
-We also conducted size analysis by examining the body diameter of vessels to examine any differences in vessel size and their relationships with vessel shape.
+Vessel size is another important variable for detecting standardization, we used the body diameter of vessels as a proxy of size to examine their variations and relationships with vessel shape.
 We measured body diameter directly from each physical vessel in the collection, and we focus on this metric because it is available for more vessels than any other metric.
 The body diameter of vessels from the Chinese period is larger than those from the two earlier periods, and the vessels before European contact have the smallest body diameter on average (Figure \@ref(fig:size-shape-regression-plots): A).
-To investigate vessel form standardization reflected by both shape and size, we calculated CV for PC1 (shape) in relation to CV for body diameter (size).
-The result (Figure \@ref(fig:size-shape-regression-plots): B) presents a higher standardization in vessel form in the Chinese period according to smaller CV values compared to CV values from the other two phases.
-However, there are no big differences in form standardization between and after the European presence.
-To understand the relationship between shape and size, we created linear regression models for PCs and the measurement of body diameter, with an examination of Pearson's correlation coefficient and p-value (Figure \@ref(fig:size-shape-regression-plots): C).
-The results demonstrate that shape changes significantly with size in all phases indicated by moderate positive relationships (0.3 ≤ r ≤ 0.7) and small p-values (≤ 0.05), except for PC1 in the Chinese period and PC2 in the pre-European period.
-That is, the shorter vessels are larger in body diameter according to the positive correlation between PC1 and body diameter.
+To investigate vessel form standardization, represented by shape and size, we compared CV for PC1 (shape) in relation to CV for body diameter (size).
+The result (Figure \@ref(fig:size-shape-regression-plots): B) presents a higher standardization in vessel form in the Chinese period according to smaller CV values compared to those from the other two phases.
+However, there are no obvious differences in form standardization before and after the European presence.
+To understand the relationship between shape and size, we modeled linear regression for PCs and the measurement of body diameter, with an examination of Pearson's correlation coefficient and p-value (Figure \@ref(fig:size-shape-regression-plots): C).
+The results demonstrate that shape changes with size in all phases indicated by moderate positive relationships (0.3 ≤ r ≤ 0.7) and small p-values (≤ 0.05), except for PC1 in the Chinese period and PC2 in the pre-European period.
+In general, the shorter vessels are larger in body diameter according to the significant positive correlation.
 However, vessels from the Chinese period do not show this pattern.
-On the contrary, the negative relationship between PC2 and body diameter suggests that vessels with a narrower neck and mouth tend to have a larger body diameter.
+For the relationship between PC2 and body diameter, the negative relationship suggests that vessels with a narrower neck and mouth tend to have a larger body diameter.
 

### 2020-08-12 analysis/paper/paper.Rmd
edit discussion

@@
 
-Previous investigations at Kiwulan suggested an unequal distribution of prestige goods, trade ornaments in particular, following the appearance of Europeans with a high diversity of types [@Cheng2008; @Wang2011].
+Previous investigations at Kiwulan suggested an unequal distribution of prestige goods with high diversity in types, trade ornaments in particular, following the appearance of Europeans [@Cheng2008; @Wang2011].
 This hinted at the emergence of social inequality within the Indigenous community.
-To understand the relationship between social inequality and foreign presences, we examined ceramic vessel shape standardization to measure craft specialization as a proxy for social change [@Costin2001; @Junker1999].
-The result of our MANOVA demonstrates a significant difference in shapes between the pre-European and Post-European periods, and the pre-European and Chinese periods.
+To understand the relationship between social inequality and foreign presences, we examined shape standardization of ceramics to measure craft specialization as a proxy for social change [@Costin2001; @Junker1999].
+The result of our MANOVA demonstrates significant differences in shapes between the pre-European and Post-European periods, and pre-European and Chinese periods.
 The average shape presents as a round body with a wide rim and neck before European contact, which shifts to a more oval-shaped body with narrower rim and neck after the European presence.
@@
 For the degree of shape standardization, our CV tests on PC1 indicate a significant difference between the Chinese period and either pre-European or post-European periods.
-A test of morphological disparity also demonstrates similar results between Chinese contact and either pre-European (p = $`r pre_chi`$) or post-European periods (p= $`r post_chi`$) [@WangandMarwick2020], suggesting that a more standardized pottery shape after contact with the Chinese.
-In addition, we found a more homogeneous shape accompanied by standardized size, represented by the body diameter of vessels, in the Chinese period.
+A test of morphological disparity also demonstrates similar results between Chinese contact and either pre-European (p = $`r pre_chi`$) or post-European periods (p= $`r post_chi`$) [@WangandMarwick2020], suggesting a more standardized shape after contact with the Chinese.
+In addition, we found a more homogeneous shape accompanied by a more standardized size, represented by the body diameter of vessels, in the Chinese period.
 Humans tend to make mistakes in hand-crafting when the size of an object increases, leading to higher variations in larger artifacts [@Eerkens2001].
@@
 
-Whether standardization suggests an emergence of specialized production depends on the number of producers, which distinguishes mechanical standardization from intentional standardization defined by stylish and functional attributes [@Costin2001; @Costin1995].
-In other words, mechanical standardization is associated with potter's skills and experiences that increased morphological uniformity normally reflects fewer producers.
+Whether the shape standardization we found suggests craft specialization depends on the number of producers, which distinguishes mechanical standardization from intentional standardization defined by stylish and functional attributes [@Costin2001; @Costin1995].
+Mechanical standardization is used to access the appearance of specialized production based on the assumption that increased skills, routinization, and lower diversity of producers will lead to morphological uniformity [@Arnold2000].
 The population at Kiwulan could be an indicator to estimate the potential number of producers.
-According to the historical records [@Nakamura1938, pp.
-12], the population in Kiwulan reached a higher number in the 17th century as the European presence and declined in the Chinese period due to movement of Indigenous people to the south [@Chen2007].
-The changes in population size described in documents correspond to the observation of ceramic abundance at Kiwulan.
-Thus, we model cv values as a function of the mass of ceramics using Poisson GLM with a link function.
-The model suggests that ceramic abundance strongly predicts the cv values (β = $`r beta_print`$, p = $`r pval_print`$).
+According to the Dutch census in 1648 [@Nakamura1938, pp.
+12], the population at Kiwulan was large but declined in the Chinese period due to the movement of Indigenous people to the south [@Chen2007].
+The change in population size corresponds with the observation of ceramic abundance at Kiwulan.
+Thus, we model CV values as a function of the mass of ceramics using Poisson GLM with a link function.
+The model suggests that ceramic abundance strongly predicts the CV values (β = $`r beta_print`$, p = $`r pval_print`$).
 This indicates that a more standardized shape in the Chinese period may be influenced by a small population if ceramic abundance can reflect the population size.
@@
 However, intentional standardization could also contribute to shape standardization in our case.
-To ensure the standardization is caused by fewer producers or specialized organization, we investigate other factors that may contribute to shape standardization, such as the function and decoration of ceramics.
+To ensure the standardization is caused by a small number of specialized potters, we investigate other factors that may contribute to shape standardization, including the function and decoration of ceramics.
 To determine the function of pots in relation to shape changes, we used geochemical methods to extract and identify lipids trapped in the fabric of potsherds to identify foods that may have contributed residues absorbed into the clay [cf.

### 2020-08-12 analysis/paper/paper.Rmd
finish discussion and conclusion editing

@@
 Firing resulted in orange and brownish color with a fully oxidized core, or a reduced core with oxidized fringes [@Chen2007].
-The vessels were believed to be made with pinching technique according to the hand-shaped traces on vessel interiors, such as finger impressions and seams.
+The vessels were believed to be made with pinching technique according to some hand-shaped traces on vessel interiors, such as finger impressions and seams.
 This kind of vessel has been widely found at archaeological sites during the late Iron Age and the historical period throughout the Yilan Plain [@NMTH2005].
@@
 
-However, intentional standardization could also contribute to shape standardization in our case.
-To ensure the standardization is caused by a small number of specialized potters, we investigate other factors that may contribute to shape standardization, including the function and decoration of ceramics.
-To determine the function of pots in relation to shape changes, we used geochemical methods to extract and identify lipids trapped in the fabric of potsherds to identify foods that may have contributed residues absorbed into the clay [cf.
+However, intentional standardization due to considerations of function or style could also contribute to the shape standardization in our case.
+To ensure shape standardization can reflect craft specialization, we investigated the function and surface decoration of the vessels.
+For function, we used geochemical methods to extract and identify lipids trapped in the fabric of potsherds to explore foods that may have contributed residues absorbed into the clay [cf.
 @Kwak2015].
@@
 909].
-Stylish attributes present a wide variation in geometric motifs from pot to pot across three phases.
-In general, the ceramics in the Chinese period have slightly fewer types of motifs according to the ratio of distinct motifs to the total number for each phase (Chinese = `r chinese_deco`, post-European = `r post_deco`, pre-European = `r pre_deco`) [@WangandMarwick2020].
-The evidence shows that we could not completely rule out the intentional standardization in our sample.
-
-The spatial pattern of ceramics could provide important information about the intensity and location of manufacturing or consumption.
-The spatial analysis (Figure \@ref(fig:kernel-plot)) reveals that pottery samples presents multiple clusters with high densities of pottery during European presence.
+For style, surface impressed decoration usually consists of multiple bands of geometric motifs that we define as one type of decoration.
+If two pots share the same set of motifs but different arrangements of single bands, they are considered as two types.
+In general, the ceramics in the Chinese period have slightly fewer variations in decoration according to the ratio of distinct types to the total number of pottery from each phase (Chinese = `r chinese_deco`, post-European = `r post_deco`, pre-European = `r pre_deco`) [@WangandMarwick2020].
+The current evidence is not sufficient for us to completely rule out the possibility of intentional standardization.
+
+The spatial pattern of ceramics could provide information about potential production units and production areas [@Costin2001].
+The spatial analysis (Figure \@ref(fig:kernel-plot)) reveals that pottery samples present a widespread distribution with high densities of pottery at some units during European presence.
 Hypothesis testing on spatial randomness indicates a non-randomly dispersed distribution before European contact and more extreme dispersed distribution after European presence.
 In contrast, the distribution of pottery is more similar to random distributions during the Chinese period.
-This contradicts our expectation that a clustered pattern will be observed with an increase in pottery standardization since the emergence of specialized groups.
+This contradicts our expectation that a clustered pattern will be observed with an increase in pottery standardization since the emergence of specialized groups [@Costin1995].
 The absence of clusters in the Chinese period is notable because this was a time of a historically-documented decline of Indigenous population [@Chen2007; @Hsieh2009].
@@
 This suggests that differences in population across our three occupation phases are probably not driving variation in shape standardization.
-Also, the pottery manufacture inclines to be household-produced since no workshops have been found [@Chen2007].
+Also, it is more likely that ceramics are household-produced based on the spatial pattern and no specific facilities of production have been found [@Chen2007].
 
@@
 
-Our results offer tentative support for the hypothesis that foreign presence at Kiwulan influenced the emergence of social inequality in the local Indigenous society.
+Our results offer tentative support for the hypothesis that foreign presence at Kiwulan influenced the shape of vessels in the local Indigenous society.
 If increased shape standardization is a reliable indicator of craft specialization, then we may be seeing evidence of a shift from corporate (group-based, distributed, collective, cooperative) to network (individual-based, competitive) organization [@Blanton1996; @Feinman1995; @Feinman2000; @Feinman2000political; @Feinman2010].
-However, strong claims for an emergence of social complexity resulting from foreign contact at Kiwulan will need support from multiple and diverse sources of evidence that are beyond the scope of this paper.
+However, strong claims for an emergence of social inequality resulting from foreign contact at Kiwulan will need support from multiple and diverse sources of evidence that are beyond the scope of this paper.
 We find that vessel shapes were more standardized during the Chinese period than the European period.
@@
 Our findings help to expand upon those factors that may lead to the standardization of ceramic production in a pericolonial interaction context.
-More homogeneous vessel shapes and sizes during the Chinese period, without any changes in clay paste composition, production technique, and spatial distribution, suggest that shape standardization may relate more to intentional behavior instead of the emergence of specialized groups.
+More homogeneous vessel shapes and sizes during the Chinese period, without any substantial changes in clay paste composition, production technique, and spatial distribution, suggest that shape standardization may relate more to intentional behavior instead of the emergence of specialized groups.
 Despite a small sample size in the Chinese period, the random distribution of ceramics does not support clustered patterns of manufacturing location, such as workshops.
-Instead, the production of ceramic is likely to be produced in households.
+Instead, the ceramic production is likely to be manufactured in households.
 The homogeneous appearance may suggest expressions of social identity or cultural boundaries in Indigenous societies through highly visible vessel qualities, such as shape, maybe heightened during periods of foreign contact in pericolonial contexts.
 Our analysis, with its openly available methods and data, is readily extensible to other pottery assemblages in the region to further explore related questions about craft specialization and standardization in Iron Age ceramic assemblages.
-This study also broadens the GMM field by focusing on ceramic technologies, which could motivate more ceramic studies and become a promising branch parallel to current applications to lithic topology and bone morphology.
+This study also broadens the GMM field by focusing on ceramic technologies, which could motivate more ceramic studies and become a promising branch parallel to current applications to lithic typology and bone morphology.
 

### 2020-08-13 OxCal/fs/fs_utils.js
add 14C data and modeling

@@ -1,0 +1,216 @@
+var fileDialogWindow;
+var filecontent="";
+var filename="";
+var nodeJS=false;
+
+function getArgs()
+{
+ var args = new Object();
+ var query = document.location.search.substring(1);
+ var pairs = query.split("\&");
+ for(var i=0;i < pairs.length; i++)
+ {
+  var pos=pairs[i].indexOf('=');
+  if(pos==-1){ continue;};
+  var argname = pairs[i].substring(0,pos);
+  var value = pairs[i].substring(pos+1);
+  args[argname]=decodeURIComponent(value);
+ };
+ return args;
+};
+function getCookie(name)
+{
+ var str;
+ var cookies
+ var i;
+ var cookie;
+ var re;
+ str=document.cookie+"";
+ if(str!="")
+ {
+  re=/\s*;\s*/;
+  cookies=str.split(re);
+  for(i=0;i<cookies.length;i++)
+  {
+   re=/\s*=\s*/;
+   cookie=cookies[i].split(re);
+   if(cookie.length==2)
+   {
+    if(name==cookie[0])
+    {
+     return decodeURIComponent(cookie[1]);
+    };
+   };
+  };
+  return "undefined";
+ };
+ return "undefined";
+};
+function delCookie(name)
+{
+ var expireDate = new Date();
+ expireDate.setFullYear(expireDate.getFullYear()-1);
+ document.cookie=name+"=;expires="+expireDate.toGMTString()+";path=/";
+};
+function setCookie(name,value)
+{
+ var expireDate = new Date();
+ expireDate.setFullYear(expireDate.getFullYear()+1);
+ document.cookie=name+"="+encodeURIComponent(value)+";expires="+expireDate.toGMTString()+";path=/";
+};
+function onAServer()
+{
+ var re=/^http/i;
+ if(nodeJS){return false;};
+ return re.test(document.location);
+};
+function fileDialog(action,ext)
+{
+ if(onAServer() || localFilePossible())
+ {
+  if(!action)
+  {
+   window.open("../fs/filedialog.html", "","width=650,height=700,toolbar=no,location=no,menubar=no,resizable=no,status=no,scrollbars=yes");
+  }
+  else
+  {
+   if(fileDialogWindow)
+   {
+    if(!fileDialogWindow.closed){fileDialogWindow.close();};
+   };
+   if(document.getElementById('fileFrame'))
+   {
[…]

### 2020-08-13 OxCal/jsobj/ObjectEditor.js

@@ -1,0 +1,4137 @@
+// (c) CB Ramsey 2007
+
+var actionSpec;
+var sortKey;
+var sortDirn;
+var exchangeSpec;
+var newExchange=false;
+var defaultBritish=false;
+var exchangeWindow;
+var timerNote;
+var editColorSpec;
+var appLink=false;
+var jsobjProps={"inherit":["name","prompt","required","type","className","expand","repeat"],
+ "inheritOption":["width","height","inline","tabbed","readonly","changed","hidden",
+ 	"popup","database","outline","transparent","hint","optionsIndex"],
+ "inheritArray":["options","optionsDisplay"],
+ "inheritFunc":["special","checker","changer","editor","action","appender","calculator"]};
+
+// code for backward compatability to deal with deprecated escape
+
+function encode(str)
+{
+ return encodeURIComponent(str).replace(/\!/g,'%21').replace(/\~/g,'%7E').replace(/\'/g,'%27').replace(/\(/g,'%28').replace(/\)/g,'%29');
+};
+
+function findJSONDate(str)
+{
+ var s,d,t,dt;
+ s=str.split("T");
+ if(s.length!=2){return str;};
+ d=s[0].split("-");
+ if(d.length!=3){return str;};
+ if(str.indexOf("Z")==-1){return str;};
+ t=s[1].replace("Z","").split(":");
+ if(t.length!=3){return str;};
+ dt=new Date(str);
+ return dt;
+};
+
+function findJSONDates(obj)
+{
+ switch(typeof(obj))
+ {
+ case "string":
+  return findJSONDate(obj); // catches Dates in JSON
+ case "object":
+  for(el in obj)
+  {
+   obj[el]=findJSONDates(obj[el]);
+  };
+  return obj;
+ };
+ return obj;
+};
+
+function myParseJSON(str)
+{
+ var obj=JSON.parse(str);
+ return findJSONDates(obj);
+};
+
+function decode(str)
+{
+ try
+ { 
+  return decodeURIComponent(str);
+ }
+ catch(err)
+ {
+  return unescape(str);
+ };
+};
+
+function decodeObject(obj)
+{
+ switch(typeof(obj))
+ {
+ case "string":
+  return findJSONDate(decode(obj)); // catches Dates in JSON
[…]

### 2020-08-13 OxCal/jsobj/ObjectWriter.js

@@ -1,0 +1,179 @@
+// (c) CB Ramsey 2005
+ function isSimple(it)
+ {
+  switch(typeof(it))
+  {
+  case "number":
+  case "string":
+  case "boolean":
+   return true;
+  case "object":
+   if(it==null){return true;};
+   if(it.constructor==Date)
+   {
+    return true;
+   };
+  };
+  return false;
+ };
+ function displaySimple(it)
+ {
+  switch(typeof(it))
+  {
+  case "number":
+   return it.toString();
+  case "string":
+   return '"'+it.toString().replace(/\"/g,'\\"').replace(/\n/g,'\\n')+'"';
+  case "boolean":
+   return it.toString();
+  case "object":
+   if(it==null){return "null";};
+   if(it.constructor==Date)
+   {
+    return "new Date("+it.getTime()+")";
+   };
+  };
+  return false;
+ };
+ function checkForArrayConstructorForObject(it)
+ {
+  if(typeof(it)=="object")
+  {
+   if(it.constructor==Array)
+   {
+    if(it.length<1)
+    {
+     for(el in it)
+     {
+      return true;
+     };
+    };
+   };
+  };
+  return false;
+ };
+ function JSONdisplay(it,name)
+ {
+  return name+"="+JSON.stringify(it)+";";
+ };
+ function displayItem(it,name,noheader)
+ {
+  var disp="";
+  var el;
+  var i;
+  if(typeof(it)=="undefined"){return "";};
+  if(!(noheader || checkForArrayConstructorForObject(it)))
+  {
+   return JSONdisplay(it,name);
+  };
+  if(isSimple(it))
+  {
+   disp += name + "=" + displaySimple(it) + ";\n";
+  }
+  else
+  {
+   if(typeof(it)!="object"){return "";};
+// check if it is an array
+   if((it.constructor==Array)&&(!checkForArrayConstructorForObject(it)))
+   {
+    if((it.length>1) && isSimple(it[it.length-1]))
+    {
+     disp += name + "= new Array(" + displaySimple(it[0]);
+     for(i=1;i<it.length;i++)
+     {
+      disp+= ", " + displaySimple(it[i]);
+     };
+     disp += ");\n";
+    }
+    else
+    {
+     if(!noheader)
+     {
+      disp += name + "= new Array();\n";
+     };
+     for(el in it)
+     {
+      if(isNaN(el))
+      {
[…]

### 2020-08-13 OxCal/oxcal/header.js

@@ -1,0 +1,1 @@
+header={"version":125,"by":"Christopher Ramsey","title":"OxCal 4.4","updated":"2020-08-12T11:38:48.896Z","issued":"2020-08-12T11:38:52.293Z","issuedBy":"Christopher Ramsey","document":true,"training":false,"internet":true,"comments":[{"title":"Changes","comment":"Updated to v4.4 with IntCal 20 options","by":"Christopher Ramsey","on":"2020-08-12T11:38:52.293Z"},{"title":"Reason","comment":"New version","by":"Christopher Ramsey","on":"2020-08-12T11:38:47.435Z"}],"symlink":"oxcal","loc":"16/2/OxCalWeb"};

### 2020-08-13 OxCal/oxcal/oxcal_plot.js

@@ -1,0 +1,1172 @@
+ var ocd;
+ var nan=null;
+ var inf=null;
+ var calib;
+ var tableWindow;
+ var textWindow;
+ var showControls=true;
+ var selectNo=0;
+ var source="ocp_null.js";
+ var args;
+ var wasWorking=false;
+ var optionCommand="";
+ var curveCommand="";
+ var is_focussed=false;
+ var ocdIndex;
+ var tableDrawn=false;
+ var leftLoaded=false;
+ var loaderLoaded=false;
+ var framesLoaded=false;
+ var showingSequence=false;
+ var svgContent="";
+ var page;
+ var headerSet=true;
+ var multiPlots={plots:[],showKey:false,keyColumn:0,keyRow:0,columns:3};
+ var canvas,plotInfo,plotData,ocPlotWindow;
+ var plotColor={min_col:"#0000ff",max_col:"#ff0000"};
+ var playerMin,playerMax,ocMapWindow;
+ var playerDisable=false;
+ 
+ function makePlotOptionsString()
+ {
+  var str="";
+  function addItem(ind)
+  {
+   str+="plotOptions."+ind+"="+plotOptions[ind]+";\n";
+  };
+  str+="plotOptions.showRange[1]="+plotOptions.showRange[1]+";\n";
+  str+="plotOptions.showRange[2]="+plotOptions.showRange[2]+";\n";
+  str+="plotOptions.showRange[3]="+plotOptions.showRange[3]+";\n";
+  addItem("reportingStyle");
+  addItem("showReversed");
+  addItem("showWhole");
+  addItem("plotsPerPage");
+  addItem("scale");
+  addItem("scaleFont");
+  addItem("scaleLine");
+  addItem("showIndices");
+  addItem("showPropPlace");
+  addItem("showADFirst");
+  addItem("showNormalised");
+  addItem("showAgreement");
+  addItem("showConvergence");
+  addItem("showOutliers");
+  addItem("colorOutliers");
+  addItem("showParameters");
+  addItem("showVerbs");
+  addItem("showItalics");
+  addItem("showGrid");
+  addItem("showText");
+  addItem("showBandW");
+  addItem("showBrackets");
+  return str;
+ };
+ 
+ function readPlotOptionsString(str)
+ {
+  eval(str);
+ };
+ 
+ function saveOptions()
+ {
+  setCookie("OxCalPlotOptions",makePlotOptionsString());
+ };
+ 
[…]

### 2020-08-13 OxCal/oxcalhelp/header.js

@@ -1,0 +1,1 @@
+header={"version":34,"by":"Christopher Ramsey","title":"OxCal Help","updated":"2020-08-12T11:15:43.116Z","issued":"2020-08-12T11:17:21.987Z","issuedBy":"Christopher Ramsey","document":true,"training":false,"internet":true,"symlink":"oxcalhelp","comments":[{"title":"Changes","comment":"Updated to v4.4","by":"Christopher Ramsey","on":"2020-08-12T11:17:21.987Z"},{"title":"Reason","comment":"Update with IntCal20 link","by":"Christopher Ramsey","on":"2020-08-12T11:15:35.343Z"}],"loc":"16/2/oxcalhelp"};

### 2020-08-13 OxCal/oxcalhelp/oc_commhelp.js

@@ -1,0 +1,297 @@
+if(typeof(parseCommands)!='undefined')
+{
+
+parseCommands();
+
+commandsRef["After"].help	="group function that returns a PDF for being after all of the grouped elements; in MCMC analysis defines terminus post quem (TPQ) events";
+commandsRef["Age"].help		="type conversion function that returns an age or PDF for an age from an expression; uses the Year parameter";
+commandsRef["Axis"].help		="plotting command that specifies the default range of the x-axis for dates";
+commandsRef["Before"].help	="group function that returns a PDF for being before all of the grouped elements; in MCMC analysis defines terminus ante quem (TAQ) events";
+commandsRef["Boundary"].help	="used with Sequence, P_Sequence, U_Sequence and V_Sequence; when paired with another Boundary defines a uniformly distributed group";
+commandsRef["C_Date"].help	="simple function that returns a PDF for calendar date with Normally distributed errors; definition of date depends on options (BP or BC/AD): use -ve numbers for BC";
+commandsRef["C_Combine"].help	="used to combine normally distributed errors defined by C_Date; the optional systematic uncertainty parameter can be used to define irreducible uncertainties";
+commandsRef["C_Simulate"].help	="used for simulating Normally distributed dating methods";
+commandsRef["Combine"].help	="group function that combines any number of PDFs which all give independent information on a parameter";
+commandsRef["Color"].help="allows the user to define the color used for a distribution or a curve in plots; any of the css style color definitions can be used such as Red or #ff0000";
+commandsRef["Comment"].help	="comment (ignored by the program)";
+commandsRef["Correl_Matrix"].help ="generates a matrix of Pearson product-moment correlation coefficients; if no parameters are given the query is applied to the parent group";
+commandsRef["Correlation"].help	="generates a correlation plot showing the relationship bewtween two parameters";
+commandsRef["Covar_Matrix"].help ="generates a covariance matrix for the parameters referred to; if no parameters are given the query is applied to the parent group";
+commandsRef["Curve"].help	="define the radiocarbon calibration curve to be used for R_Date and R_Combine";
+commandsRef["Date"].help		="type conversion function that returns a date or PDF for a date from an expression";
+commandsRef["Delta_R"].help	="defines the shift that is to be applied to dates before calibration; applies to the current Curve; if required a non-Normal prior can be applied to this parameter using the expression but this will only have an effect on MCMC models, for straight calibrations the mean and standard deviation of the distribution will be used";
[…]

### 2020-08-13 OxCal/oxplot/header.js

@@ -1,0 +1,1 @@
+header={"version":31,"by":"Christopher Ramsey","title":"WWW OxPlot files","loc":"16/2/oxplot","updated":"2020-04-22T21:00:13.649Z","issued":"2020-04-22T21:00:17.386Z","issuedBy":"Christopher Ramsey","document":true,"training":false,"internet":true,"symlink":"oxplot","comments":[{"title":"Changes","comment":"plot_link.js updated","by":"Christopher Ramsey","on":"2020-04-22T21:00:17.386Z"},{"title":"Reason","comment":"Update plotlink with default for showKey","by":"Christopher Ramsey","on":"2020-04-22T21:00:00.118Z"}]};

### 2020-08-13 OxCal/oxplot/oxplot.js

@@ -1,0 +1,2748 @@
+var canvas,plotInfo,plotOptions,plotData,editOptions,multiPlots,memInfo,tables,fileOptions;
+var plotCalc,plotMultiCalc,plotColor,memCalc;
+var editSpec,viewSpec,formatSpec,canvasSpec,multiSpec,dataSpec,fileSpec,colorSpec;
+var autofill=false;
+var plotPg="";
+var dataCount=0;
+var multiPos=0,xorg,yorg,worg,horg,yoffs;
+var ticker=false;
+var fileinclude=false;
+var markers=new Array("","circle","square","diamond","triangle","cross","x","label","histogram","rectangle","image");
+var colors=new Array("rgb(255,0,0)","rgb(0,255,0)","rgb(0,0,255)","rgb(204,204,0)", "rgb(204,0,204)","rgb(0,204,204)","rgb(127,127,127)");
+   canvas=new Object;
+   canvas.scale=1.0; // zoom scaling
+   canvas.scaleFont=0.8; // font scaling
+   canvas.scaleLine=1.3; // line scaling
+   canvas.pxPerCm=35; // px per cm
+   canvas.frameWidth=18.0; // cm
+   canvas.frameHeight=13.5; // cm
+   
+   fileOptions=new Object;
+   fileOptions.canvas=true; // covers canvas
+   fileOptions.view=true; //covers plotInfo and multiPlots
+   fileOptions.format=true; // covers plotOptions
+   fileOptions.data=true;
+   
+   plotInfo=new Object;
+   plotInfo.minx=-180;
+   plotInfo.maxx=180;
+   plotInfo.autox=0;
+   plotInfo.miny=-180;
+   plotInfo.nonliny="mercator";
+   plotInfo.mapPlot=true;
+   plotInfo.maxy=180;
+   plotInfo.autoy=0;
+   plotInfo.x_calc="longitude";
+   plotInfo.y_calc="latitude";
+   plotInfo.dx_calc="";
+   plotInfo.dy_calc="";
+   plotInfo.xlabel="Longitude";
+   plotInfo.ylabel="Latitude";
+   plotInfo.title="";
+   plotInfo.category="";
+   plotInfo.table="";
+   plotInfo.keyTitle="Key";
+   plotInfo.pcaVariables="";
+   
+   plotColor=new Object;
+   plotColor.showKey=false;
+   plotColor.z_calc="";
+   plotColor.dz_calc="";
[…]

### 2020-08-13 OxCal/oxplot/xy_plot.js

@@ -1,0 +1,1466 @@
+   var firstTime=true;
+   var plotNo;
+   var mover={active:false,frm_clientX:0,frm_clientY:0,
+ 	  to_clientX:0,to_clientY:0,tmr:false,relax:false,xmoved:false,ymoved:false,graph:-1};
+
+var   canvas=new Object;
+   canvas.scale=1.0; // zoom scaling
+   canvas.scaleFont=0.8; // font scaling
+   canvas.scaleLine=1.3; // line scaling
+   canvas.pxPerCm=35; // px per cm
+   canvas.frameWidth=18.0; // cm
+   canvas.frameHeight=13.5; // cm
+
+var   plotOptions=new Object;
+   plotOptions.multiPlot=0;
+   plotOptions.mapPlot="svg";
+   plotOptions.background="rgb(255,255,255)";
+   plotOptions.plotBackground="rgba(255,255,255,0)";
+   plotOptions.plotPosX=3.0; // cm
+   plotOptions.plotPosY=1.5;  // cm
+   plotOptions.plotPosYOffs=0;  // cm
+   plotOptions.plotWidth=11.5; // cm
+   plotOptions.plotHeight=11.5; // cm
+   plotOptions.showKey=0;
+   plotOptions.keyTitle="Key";
+   plotOptions.BandW=false;
+
+var   plotInfo=new Object;
+   plotInfo.minx=0;
+   plotInfo.maxx=1;
+   plotInfo.miny=-6;
+   plotInfo.maxy=6;
+   plotInfo.xlabel="Proportion";
+   plotInfo.ylabel="Standard Deviations";
+   plotInfo.title="Known Age";
+   plotInfo.mapPlot=false;
+   plotInfo.hideRectangle=false;
+   plotInfo.xtop=false;
+   plotInfo.yright=false;
+//   plotInfo.key=0;
+//   plotInfo.showKey=0;
+      
+var   plotColor=new Object;
+   plotColor.showKey=false;
+   plotColor.z_calc="";
+   plotColor.zlabel="";
+   plotColor.minz=0;
+   plotColor.maxz=100
+   plotColor.autoz=0;
+   plotColor.min_col="rgb(0,0,255)";
+   plotColor.max_col="rgb(255,0,0)";
+
+var   plotData=new Array;
[…]

### 2020-08-13 OxCal/utils/Menu.js

@@ -1,0 +1,458 @@
+// (c) CB Ramsey 2006
+ var menuControl=new Object;
+ menuControl.helpActive=false;
+ menuControl.lastId="";
+ menuControl.lastI=0;
+ menuControl.lastTab=false;
+ menuControl.menus=new Object;
+ var menuBars=new Object;
+ var clickMenus=new Object;
+
+ function toggleHelp()
+ {
+  menuControl.helpActive=!menuControl.helpActive;
+  makeActive(menuControl.lastId,menuControl.lastI,menuControl.lastTab);
+ };
+ function overHelp()
+ {
+  if(!menuControl.helpActive)
+  {
+   document.getElementById("helpButton").src="../img/HelpOver.gif"
+  };
+ };
+ function outHelp()
+ {
+  if(!menuControl.helpActive)
+  {
+   document.getElementById("helpButton").src="../img/Help.gif"
+  };
+ };
+/* function outButton(b)
+ {
+  b.firstChild.src="../img/Button.gif";
+ };
+ function overButton(b)
+ {
+  b.firstChild.src="../img/ButtonMid.gif";
+ };
+ function downButton(b)
+ {
+  b.firstChild.src="../img/ButtonDeep.gif";
+ };
+ function upButton(b)
+ {
+  b.firstChild.src="../img/Button.gif";
+ };*/
+ function outTButton(b)
+ {
+  b.className="button";
+ };
+ function downTButton(b)
+ {
+  b.className="buttonDown";
+ };
+ function upTButton(b)
+ {
+  b.className="button";
+ };
+ function Button(action,image)
+ {
+  document.writeln("<div class='button' onclick='"+action+"'>"+
+   "<img style='position:relative;top:-1px;left:-1px' src='"+image+"'\/><\/div>");
+/*  document.writeln("<div class='button' onmouseover='overButton(this)' "+
+   "onmouseout='outButton(this)' onmousedown='downButton(this)' onmouseup='upButton(this);"+action+"'>"+
+   "<img src='../img/Button.gif' \/><img style='position:relative;top:-23px' src='"+image+"'\/><\/div>");*/
+/*  document.writeln("<button style='height:25px;width:32px;background-color:rgba(255,255,255,0.5);border-width:1px;-moz-border-radius: 4px;-webkit-border-radius: 4px' onclick='"+action+"'>"+
[…]

### 2020-08-13 OxCal/utils/ViewPort.js

@@ -1,0 +1,860 @@
+// ViewPort.js
+// General Plotting routines
+// =======================
+  var xlinkNS = "http://www.w3.org/1999/xlink";
+  
+  function ViewPortDoc()
+  {
+   return document;
+  };
+  function roundFactor(num,fact)
+  {
+   return Math.round(num*fact)/fact;
+  };
+  function svgColor(col,typ,opac)
+  {
+   var arr,str="";
+   if(typeof(col)!="string"){col="none";};
+   if(typ){str=typ+":";};
+   if((col.indexOf("(")!=-1)&&(col.indexOf(")")!=-1))
+   {
+    arr=col.slice(col.indexOf("(")+1,col.indexOf(")")).split(",");
+    if(opac){arr[3]=opac;};
+    if(arr.length>3)
+    {
+     str+="rgb("+arr[0]+","+arr[1]+","+arr[2]+")";
+     switch(typ)
+     {
+     case "stroke":case "fill":
+      str+=";"+typ+"-opacity:"+arr[3];
+      break;
+     };
+     return str;
+    };
+   }
+   else
+   {
+    if(opac)
+    {
+     str+=col;
+     switch(typ)
+     {
+     case "stroke":case "fill":
+      str+=";"+typ+"-opacity:"+opac;
+      break;
+     };
+     return str;
+    };
+   };
+   str+=col;
+   return str;
+  };
+  function mercator(lat)
+  {
+   lat=Math.PI*lat/180;
+   return 180*Math.log(1/Math.cos(lat)+Math.tan(lat))/Math.PI;
+  };
+  function inv_mercator(lat)
+  {
+   return 360*Math.atan(Math.exp(Math.PI*lat/180))/Math.PI-90;
+  };
+  var scale,font_scale,line_scale;
+  function attachSVGDocument(w,h,s,font_stretch,line_stretch,background)
+  {
+   var mainFrame;
+   var el;
+   var pt_scale;
+   if(!font_stretch){font_stretch=0.8;};
+   if(!line_stretch){line_stretch=1.0;};
+   if(!background){background="white";};
+   scale=s;
+   font_scale=scale*font_stretch;
+   line_scale=line_stretch*font_scale;
+   pt_scale=s*28.3;
+   el=ViewPortDoc().getElementById("top");
+   el.setAttributeNS(null,"width",(w*pt_scale).toFixed(3)+"pt");
[…]

### 2020-08-13 OxCal/utils/header.js

@@ -1,0 +1,1 @@
+header={"version":29,"by":"Christopher Ramsey","title":"WWW Utils","updated":"2020-02-21T13:49:49.347Z","issued":"2020-02-21T13:49:55.867Z","issuedBy":"Christopher Ramsey","document":true,"training":false,"internet":true,"symlink":"utils","comments":[{"title":"Changes","comment":"Menu.js updated","by":"Christopher Ramsey","on":"2020-02-21T13:49:55.867Z"},{"title":"Reason","comment":"Update rocker to make it clearer this is a section - and to change cursor used over it","by":"Christopher Ramsey","on":"2020-02-21T13:49:24.173Z"}],"loc":"16/2/utils"};

### 2020-08-13 OxCal/utils/kde.js

@@ -1,0 +1,431 @@
+function kde(rows,columns,content,weights)
+{
+ var i,j;
+ this.data=new matrix(rows,columns,content,weights);
+ if(rows)
+ {
+  this.delta=this.data.subtract_mean();
+  this.cov=this.delta.covariance();
+ };
+ this.norm=1/Math.sqrt(2*Math.PI);
+ this.max=0;
+};
+new kde(0,0);
+kde.prototype.cvexp=function(k1,k2,x1,x2)
+{
+ return this.norm*Math.exp(-0.5*(x1*x2)/this.cov.m[k1][k2]);
+};
+kde.prototype.kde_1d_setup=function(k1)
+{
+ var d;
+ this.h=Math.sqrt(this.cov.m[k1][k1]);
+ d=1;
+ this.h=this.h*Math.exp((1/(d+4))*Math.log(4/((d+2)*this.data.rows)));
+ // using Scott 1992 as reported in Zhang et al 2004
+};
+kde.prototype.kde_1d=function(k1,x1)
+{
+ var i,p,pp;
+ p=0;
+ for(i=0;i<n;i++)
+ {
+  pp=this.norm*Math.exp(-0.5*((x1-this.data.m[i][k1])*(x1-this.data.m[i][k1]))/this.h);
+  p+=pp/n;
+ };
+ return p;
+};
+kde.prototype.angle_2d=function(k1,k2)
+{
+ var f;
+ if(this.cov.m[k1][k2]==0)
+ {
+  if(this.cov.m[k2][k2]>this.cov.m[k1][k1])
+  {
+   this.q=Math.PI/2;
+  }
+  else
+  {
+   this.q=0;
+  };
+ }
+ else
+ {
+  if(this.cov.m[k1][k1]==this.cov.m[k2][k2])
+  {
+   if(this.cov.m[k2][k2]>0)
+   {
+    this.q=Math.PI/4;
+   }
+   else
+   {
+    this.q=-Math.PI/4;
+   };
+  }
+  else
+  {
+   this.q=0.5*Math.atan(2*this.cov.m[k1][k2]/(this.cov.m[k1][k1]-this.cov.m[k2][k2]));
+   if(this.cov.m[k1][k2]>0)
+   {
+    if(this.cov.m[k2][k2]>this.cov.m[k1][k1])
+    {
+     this.q=Math.PI/2+this.q;
+    };
+   }
+   else
+   {
[…]

### 2020-08-13 analysis/code/KWL-dates-model.R

@@ -1,0 +1,72 @@
+kw_dates <- readxl::read_excel(here::here("analysis", "data", "raw_data", "KWL_C14_dates.xlsx"))
+library(tidyverse)
+# subset dates with error value
+kw_dates_with_error <-
+  kw_dates  %>%
+  filter(!is.na(se))
+
+# convert rce to numeric
+kw_dates_with_error$rce <- as.numeric(kw_dates_with_error$rce)
+
+# make some kind of ID
+kw_dates_with_error$id <- with(kw_dates_with_error,
+                               paste0(Square,"-",Layer))
+
+# Use oxcAAR to clibrate
+library(oxcAAR)
+quickSetupOxcal()
+date_kwl <- oxcalCalibrate(kw_dates_with_error$rce,
+                           kw_dates_with_error$se,
+                           kw_dates_with_error$`Lab code`)
+date_kwl
+plot(date_kwl)
+
+# for modeling
+kw_dates_with_error_three_phases <-
+  kw_dates_with_error %>%
+  filter(rce < 600) %>%
+  mutate(phase = rce) %>%
+  mutate(phase = ifelse(rce > 324, "Pre European", phase),
+         phase = ifelse(rce < 324, "Post European", phase)) %>%
+  select(`Lab code`, "rce", "se", "phase")
+
+
+three_phases <- 'Plot()
+{
+ Phase()
+ {
+    Sequence("Pre European")
+    {
+      Boundary("Start of Pre-European");
+      Phase()
+      {
+        R_Date("NTU-3791", 340, 30);
+        R_Date("NTU-4292", 510, 75);
+        R_Date("NTU-4310", 360, 100);
+        R_Date("NTU-4320", 340, 100);
+        Interval( "Duration of Pre-European");
+      };
+        Boundary("End of Pre-European");
+    };
+  Sequence()
+    {
+   Boundary( "Start of Post-European");
+   Phase( )
+        {
+        R_Date("NTU-3993", 250, 40);
+        R_Date("NTU-4016", 270, 45);
+        R_Date("NTU-4311", 310, 100);
+        R_Date("NTU-4419", 280, 70);
+        Interval("Duration of Post-European");
+        };
+      Boundary("End of Post-European");
+    };
+  };
+};
+'
+three_phases_results <- executeOxcalScript(three_phases)
+three_phases_text <- readOxcalOutput(three_phases_results)
[…]

### 2020-08-13 analysis/code/KWL-dates-model.R
added code to download package that fixed the issue

@@ -13,6 +13,8 @@ kw_dates_with_error$id <- with(kw_dates_with_error,
                                paste0(Square,"-",Layer))
 
 # Use oxcAAR to clibrate
+library(devtools)
+install_github('ISAAKiel/oxcAAR')
 library(oxcAAR)
 quickSetupOxcal()
 date_kwl <- oxcalCalibrate(kw_dates_with_error$rce,

### 2020-08-13 analysis/code/KWL-dates-model.R
modeling contuguous phases

@@ -35,9 +35,9 @@ kw_dates_with_error_three_phases <-
 
 three_phases <- 'Plot()
 {
- Phase()
+ Phase("Upper Layer")
  {
-    Sequence("Pre European")
+    Sequence()
     {
       Boundary("Start of Pre-European");
       Phase()
@@ -72,3 +72,37 @@ three_phases_result_data <- parseOxcalOutput(three_phases_text, only.R_Date = F)
 str(three_phases_result_data)
 print(three_phases_result_data)
 plot(three_phases_result_data)
+
+three_phases_con <- 'Plot()
+{
+    Sequence("Upper Layer")
+    {
+      Boundary("Start of Pre-European");
+      Phase()
+      {
+        R_Date("NTU-3791", 340, 30);
+        R_Date("NTU-4292", 510, 75);
+        R_Date("NTU-4310", 360, 100);
+        R_Date("NTU-4320", 340, 100);
+        Interval( "Duration of Pre-European");
+      };
+        Boundary("Pre-European - Post-European");
+      Phase()
+      {
+        R_Date("NTU-3993", 250, 40);
+        R_Date("NTU-4016", 270, 45);
+        R_Date("NTU-4311", 310, 100);
+        R_Date("NTU-4419", 280, 70);
+        Interval("Duration of Post-European");
+      };
+      Boundary("End of Post-European");
+    };
+};
+'
+three_phases_con_results <- executeOxcalScript(three_phases_con)
+three_phases_con_text <- readOxcalOutput(three_phases_con_results)
+three_phases_result_con_data <- parseOxcalOutput(three_phases_con_text, only.R_Date = F)
+str(three_phases_result_con_data)
+print(three_phases_result_con_data)
+plot(three_phases_result_con_data)
+

### 2020-08-13 analysis/paper/paper.Rmd
update chronology section to include modeling result

@@
 These phases were identified according to chronologically diagnostic artifacts based on their fined-grained time scale.
-Despite the presence of 17 radiocarbon ages [@Chen2007], they tend to have long time ranges that can only serve as a cross-validation if applicable.
+Despite the availability of 17 radiocarbon ages [@Chen2007], the chronological model demonstrates that time range are too coarse to provide sufficient information for the separation of three phases [@WangandMarwick2020].
+Thus, radiocarbon ages only serve as a cross-validation for the diagnostic artifacts if applicable.
 The diagnostic artifacts include blue and white porcelains, light grey glazed jars, and large dark brown glazed stoneware jars commonly used in the 17th century, and bricks and tiles employed by the Chinese in the 19th century [@Chen2007; @Hsieh2009; @Wang2011].

### 2020-08-14 analysis/paper/paper.Rmd
update discussion section and add references

@@
 
-The shape variation reported here is subtle, and invites consideration of another possible scenario that the absence of major changes in vessel shape at Kiwulan may have been an act of resistance to foreign influence.
+The shape variation reported here is subtle and invites consideration of another possible scenario that the absence of major changes in vessel shape at Kiwulan may have been an act to show ethnic identity when experiencing foreign influences [cf.
+@Torrence2000].
+We should note that the morphology of ceramics could be treated as a signal for message communication between potters that repeating the same shape or modifying may have occurred in a non-deterministic way [@Kubler1962].
+However, as the only type of locally made pottery throughout 600 years, the homogeneous and even more standardized shape in the Chinese period might convey some meaningful information.
 We recognize the decline in population may lead to a standardized shape, however, the evidence of ceramic spatial distribution shows a dispersed or random pattern across the whole area throughout three phases.

### 2020-08-19 analysis/code/KWL-dates-model.R
removed an old doc file

@@ -43,9 +43,9 @@ three_phases <- 'Plot()
       Phase()
       {
         R_Date("NTU-3791", 340, 30);
-        R_Date("NTU-4292", 510, 75);
-        R_Date("NTU-4310", 360, 100);
         R_Date("NTU-4320", 340, 100);
+        R_Date("NTU-4310", 360, 100);
+        R_Date("NTU-4292", 510, 75);
         Interval( "Duration of Pre-European");
       };
         Boundary("End of Pre-European");
@@ -57,8 +57,8 @@ three_phases <- 'Plot()
         {
         R_Date("NTU-3993", 250, 40);
         R_Date("NTU-4016", 270, 45);
-        R_Date("NTU-4311", 310, 100);
         R_Date("NTU-4419", 280, 70);
+        R_Date("NTU-4311", 310, 100);
         Interval("Duration of Post-European");
         };
       Boundary("End of Post-European");

### 2020-08-19 analysis/code/KWL-dates-model.R
fixed typo and added rmd for chronological model

@@ -32,47 +32,7 @@ kw_dates_with_error_three_phases <-
          phase = ifelse(rce < 324, "Post European", phase)) %>%
   select(`Lab code`, "rce", "se", "phase")
 
-
-three_phases <- 'Plot()
-{
- Phase("Upper Layer")
- {
-    Sequence()
-    {
-      Boundary("Start of Pre-European");
-      Phase()
-      {
-        R_Date("NTU-3791", 340, 30);
-        R_Date("NTU-4320", 340, 100);
-        R_Date("NTU-4310", 360, 100);
-        R_Date("NTU-4292", 510, 75);
-        Interval( "Duration of Pre-European");
-      };
-        Boundary("End of Pre-European");
-    };
-  Sequence()
-    {
-   Boundary( "Start of Post-European");
-   Phase( )
-        {
-        R_Date("NTU-3993", 250, 40);
-        R_Date("NTU-4016", 270, 45);
-        R_Date("NTU-4419", 280, 70);
-        R_Date("NTU-4311", 310, 100);
-        Interval("Duration of Post-European");
-        };
-      Boundary("End of Post-European");
-    };
-  };
-};
-'
-three_phases_results <- executeOxcalScript(three_phases)
-three_phases_text <- readOxcalOutput(three_phases_results)
-three_phases_result_data <- parseOxcalOutput(three_phases_text, only.R_Date = F)
-str(three_phases_result_data)
-print(three_phases_result_data)
-plot(three_phases_result_data)
-
+# contiguous phases, not overlapping
 three_phases_con <- 'Plot()
 {
     Sequence("Upper Layer")
@@ -105,4 +65,3 @@ three_phases_result_con_data <- parseOxcalOutput(three_phases_con_text, only.R_D
 str(three_phases_result_con_data)
 print(three_phases_result_con_data)
 plot(three_phases_result_con_data)
-

### 2020-08-20 analysis/code/KWL-dates-model.R
stash

@@ -12,7 +12,7 @@ kw_dates_with_error$rce <- as.numeric(kw_dates_with_error$rce)
 kw_dates_with_error$id <- with(kw_dates_with_error,
                                paste0(Square,"-",Layer))
 
-# Use oxcAAR to clibrate
+# Use oxcAAR to calibrate
 library(devtools)
 install_github('ISAAKiel/oxcAAR')
 library(oxcAAR)
@@ -32,7 +32,8 @@ kw_dates_with_error_three_phases <-
          phase = ifelse(rce < 324, "Post European", phase)) %>%
   select(`Lab code`, "rce", "se", "phase")
 
-
+# use continguous example from
+# https://c14.arch.ox.ac.uk/oxcalhelp/hlp_analysis_eg.html#phases
 three_phases <- 'Plot()
 {
  Phase("Upper Layer")
@@ -42,17 +43,13 @@ three_phases <- 'Plot()
       Boundary("Start of Pre-European");
       Phase()
       {
-        R_Date("NTU-3791", 340, 30);
         R_Date("NTU-4292", 510, 75);
         R_Date("NTU-4310", 360, 100);
         R_Date("NTU-4320", 340, 100);
+        R_Date("NTU-3791", 340, 30);
         Interval( "Duration of Pre-European");
       };
-        Boundary("End of Pre-European");
-    };
-  Sequence()
-    {
-   Boundary( "Start of Post-European");
+   Boundary("Pre-European - Post-European");
    Phase( )
         {
         R_Date("NTU-3993", 250, 40);
@@ -73,36 +70,5 @@ str(three_phases_result_data)
 print(three_phases_result_data)
 plot(three_phases_result_data)
 
-three_phases_con <- 'Plot()
-{
-    Sequence("Upper Layer")
-    {
-      Boundary("Start of Pre-European");
-      Phase()
-      {
-        R_Date("NTU-3791", 340, 30);
-        R_Date("NTU-4292", 510, 75);
-        R_Date("NTU-4310", 360, 100);
-        R_Date("NTU-4320", 340, 100);
-        Interval( "Duration of Pre-European");
-      };
-        Boundary("Pre-European - Post-European");
-      Phase()
-      {
-        R_Date("NTU-3993", 250, 40);
-        R_Date("NTU-4016", 270, 45);
-        R_Date("NTU-4311", 310, 100);
-        R_Date("NTU-4419", 280, 70);
-        Interval("Duration of Post-European");
-      };
[…]

### 2020-08-20 analysis/paper/KWL-chronology-modeling.Rmd

---
title: "Supplementary Data: 14C Modeling for Kiwualn Chronology"
author: "Liying Wang and Ben Marwick"
date: "8/18/2020"
output: html_document
bibliography: references.bib
---

# Introduction

This document provides data and code for modeling the chronology of the upper component of Kiwulan. The full discussion of chronology is in the paper: The paper Standardization of ceramic shape: A case study from the Iron Age pottery from northeastern Taiwan. In this document, we calibrated all C14 dates [@Chen2007] first and then further modeled different phases related to foreign influences, pre- and post- European periods. Our purpose is to answer whether the C14 dates support a time range that corresponds to historical events of the presence of Europeans in Taiwan. The methods and results are included below.

```{r load-data}
kw_dates <- readxl::read_excel(here::here("analysis", "data", "raw_data", "KWL_C14_dates.xlsx"))
```

```{r tidy-data}
library(tidyverse)

# subset dates with error value
kw_dates_with_error <-
  kw_dates  %>%
  filter(!is.na(se))

# convert rce to numeric
kw_dates_with_error$rce <- as.numeric(kw_dates_with_error$rce)

# make some kind of ID
kw_dates_with_error$id <- with(kw_dates_with_error,
                               paste0(Square,"-",Layer))
```

# Calibration

We calibrated 23 dates from different contexts, including normal deposit, burials, and middens, at Kiwulan using oxcAAR package, which executes OxCal v4.4.1 with radiocarbon calibration curve IntCal20 [@Reimer2020].

```{r calibrate-C14data}
# Use oxcAAR to calibrate, run the following two lines to download ocxAAR 
# library(devtools)
# install_github('ISAAKiel/oxcAAR')
library(oxcAAR)
quickSetupOxcal()
date_kwl <- oxcalCalibrate(kw_dates_with_error$rce,
                           kw_dates_with_error$se,
                           kw_dates_with_error$`Lab code`)
date_kwl
plot(date_kwl)
calcurve_plot(date_kwl)
```

```{r data-for-modeling}
# upper layer with estimation of phases
kw_dates_with_error_three_phases <-
  kw_dates_with_error %>%
  filter(rce < 600) %>%
  mutate(phase = rce) %>%
  mutate(phase = ifelse(rce > 324, "Pre European", phase),
         phase = ifelse(rce < 324, "Post European", phase)) %>%
  select(`Lab code`, "rce", "se", "phase")
```

# Chronological model

We then constructed a Bayesian radiocarbon calibration model to obtain the range of pre-European and post-European we examined. We assume the phases are in an order that represents before and after the European arrival in 1624 [@Andrade2007]. The end of post-European is identified by the start of the Chinese migration to Yilan in 1780s [@Chen2007]. The results suggest that pre-European started between 1331–1635 cal AD at a 95.4% probability range and ended between 1458–1670 cal AD at a 99.7% probability range. The end date of pre-European also represents the start date of post-European, while the end date of post-European is 1517-1753 cal AD at a 95.4% probability range. The overall model is consistent with the historical data.

```{r modeling-upper-layer}
# contiguous phases, not overlapping
three_phases_con <- 'Plot()
{
    Sequence("Upper Layer")
    {
      Boundary("Start of Pre-European");
      Phase()
      {
        R_Date("NTU-3791", 340, 30);
        R_Date("NTU-4292", 510, 75);
        R_Date("NTU-4310", 360, 100);
        R_Date("NTU-4320", 340, 100);
        Interval( "Duration of Pre-European");
      };
        Boundary("Pre-European - Post-European");
      Phase()
      {
        R_Date("NTU-3993", 250, 40);
        R_Date("NTU-4016", 270, 45);
        R_Date("NTU-4311", 310, 100);
        R_Date("NTU-4419", 280, 70);
        Interval("Duration of Post-European");
      };
      Boundary("End of Post-European");
    };
};
'
```

```{r plot}
three_phases_con_results <- executeOxcalScript(three_phases_con)
three_phases_con_text <- readOxcalOutput(three_phases_con_results)
three_phases_result_con_data <- parseOxcalOutput(three_phases_con_text, only.R_Date = F)
str(three_phases_result_con_data)
print(three_phases_result_con_data)
plot(three_phases_result_con_data)
```

### 2020-08-20 analysis/code/KWL-dates-model.R
resolve merge conflict

@@ -14,14 +14,14 @@ kw_dates_with_error$id <- with(kw_dates_with_error,
 
 # Use oxcAAR to calibrate
 library(devtools)
-install_github('ISAAKiel/oxcAAR')
+# install_github('ISAAKiel/oxcAAR')
 library(oxcAAR)
 quickSetupOxcal()
 date_kwl <- oxcalCalibrate(kw_dates_with_error$rce,
                            kw_dates_with_error$se,
                            kw_dates_with_error$`Lab code`)
-date_kwl
-plot(date_kwl)
+# date_kwl
+# plot(date_kwl)
 
 # for modeling
 kw_dates_with_error_three_phases <-
@@ -32,7 +32,6 @@ kw_dates_with_error_three_phases <-
          phase = ifelse(rce < 324, "Post European", phase)) %>%
   select(`Lab code`, "rce", "se", "phase")
 
-<<<<<<< HEAD
 # use continguous example from
 # https://c14.arch.ox.ac.uk/oxcalhelp/hlp_analysis_eg.html#phases
 three_phases <- 'Plot()
@@ -40,17 +39,10 @@ three_phases <- 'Plot()
  Phase("Upper Layer")
  {
     Sequence()
-=======
-# contiguous phases, not overlapping
-three_phases_con <- 'Plot()
-{
-    Sequence("Upper Layer")
->>>>>>> 03f3a0a616fb71ee4b7128279b677a8cf681b075
     {
       Boundary("Start of Pre-European");
       Phase()
       {
-<<<<<<< HEAD
         R_Date("NTU-4292", 510, 75);
         R_Date("NTU-4310", 360, 100);
         R_Date("NTU-4320", 340, 100);
@@ -60,23 +52,11 @@ three_phases_con <- 'Plot()
    Boundary("Pre-European - Post-European");
    Phase( )
         {
-=======
-        R_Date("NTU-3791", 340, 30);
-        R_Date("NTU-4292", 510, 75);
-        R_Date("NTU-4310", 360, 100);
-        R_Date("NTU-4320", 340, 100);
-        Interval( "Duration of Pre-European");
-      };
-        Boundary("Pre-European - Post-European");
-      Phase()
-      {
->>>>>>> 03f3a0a616fb71ee4b7128279b677a8cf681b075
         R_Date("NTU-3993", 250, 40);
         R_Date("NTU-4016", 270, 45);
         R_Date("NTU-4311", 310, 100);
         R_Date("NTU-4419", 280, 70);
[…]

### 2020-08-20 analysis/paper/paper.Rmd
minor edit to intro

@@
 ---
-title: "Standardization of ceramic shape: A case studay from the Iron Age pottery from northeastern Taiwan"
+title: "Standardization of ceramic shape: A case study from the Iron Age pottery from northeastern Taiwan"
 
@@
 Using standardization in ceramic shapes as a proxy for craft specialization, we ask: Did colonial trade impact the shape of locally-produced Indigenous pottery vessels?
-Did pottery shape become more statistically homogeneous after foreign contacts with European colonizers or Chinese immigrants?
+Did pottery shape become more homogeneous after foreign contacts with European colonizers or Chinese immigrants?
 

### 2020-08-21 analysis/paper/paper.Rmd
numerous edits to the text after revisions in response to peer reviews

@@
 
-This research investigates if there was increasing ceramic specialization resulting from Indigenous interaction with Europeans in the 17th century or Chinese in the 19th century, two major foreign influences in early historical Taiwan, which may indicate social changes in Indigenous societies.
-We predict that competition within the Indigenous community at Kiwulan for foreign resources and trade partnerships with European or Chinese colonizers may have led to the emergence of craft specialization caused by greater economic and social control of ceramic production by a small group of individuals.
+This research investigates if there was increasing ceramic specialization resulting from Indigenous interaction with Europeans in the 17th century, or Chinese in the 19th century.
+These were the two major foreign influences in early historical Taiwan that may relate to social changes in Indigenous societies.
+We predict that competition within the Indigenous community at Kiwulan for foreign resources and trade partnerships with European or Chinese colonizers may have led to the emergence of craft specialization, caused by greater economic and social control of ceramic production by a small group of individuals.
 Using standardization in ceramic shapes as a proxy for craft specialization, we ask: Did colonial trade impact the shape of locally-produced Indigenous pottery vessels?
@@
 Landmarks, curves or outlines of objects can be represented by coordinates in terms of their unique point locations with respect to numerical values on coordinate axes.
-There are two common morphometric methods: landmark and outline approaches [@Adams2004].
+There are two common morphometric approaches: landmark and outline methods [@Adams2004].
 Landmark GMM approaches assign a set of landmarks and/or semilandmarks onto objects as reference points.
 Generalized Procrustes analysis (GPA) is used to superimpose landmark data on a common coordinate system by translating, rotating, and scaling [@Bookstein1991].
-After the procedure of GPA, superimposed landmark coordinates become shape variables that allow further statistical analyses [@Slice2007].
+After the GPA procedure, superimposed landmark coordinates become shape variables that allow further statistical analyses [@Slice2007].
 A common procedure is using dimensional reduction techniques, such as Principal Components Analysis or Canonical Variate Analysis, to capture the key features that represent the overall shape.
@@
 Landmark-based morphometrics have been widely applied to archaeological objects with obvious morphological features that provide unambiguous reference points for landmark placement, such as tips and edges of stone or metal tools [@Birch2019; @Lycett2013], visually distinctive bone features [@Haruda2019; @Meloro2015], or ceramic assemblages with distinct components [@Selden2019; @Topi2017].
-GMM is often used to answer research questions related to lithic typological and technological change [@Doyon2019; @Eren2015; @Selden2018lithic; @Perez2007; @Presnyakova2018], animal domestication or mobility [@Haruda2019; @Owen2014], or hominid activities through cutmarks and taphonomic traces [@Aramendi2019; @Courtenay2019].
+This approach is often used to answer research questions related to lithic typological and technological change [@Doyon2019; @Eren2015; @Selden2018lithic; @Perez2007; @Presnyakova2018], animal domestication or mobility [@Haruda2019; @Owen2014], or hominid activities through cutmarks and taphonomic traces [@Aramendi2019; @Courtenay2019].
 
@@
 For example, @Archer2015's case study of stone points in Southern Africa suggests an increase in shape standardization over time that may relate to increased maintenance of finished points.
-@Buchanan2018 analyzed lithic morphology with metric data and identified a more uniform base-shape of Folsom points compared to Clovis points across the western US [@Buchanan2019].
-With similar methods, @Smith2016 found standardized bases of fluted point in Alaska and northern Yukon that might indicate a risk management strategy to ensure the ease of replacement during long-distance travel.
+@Buchanan2018 analyzed lithic morphology with metric data and identified a more uniform base-shape of Folsom points compared to Clovis points across the western US.
+With similar methods, @Smith2016 found standardized bases of fluted points in Alaska and northern Yukon that might indicate a risk management strategy to ensure the ease of replacement during long-distance travel.
 Other factors, such as low levels of cultural innovation in a small group, could also lead to an increase in standardization of point shapes [@Okumura2014].
-To test the effectiveness of measuring standardization, @Birch2019 compared landmark-based GMM to traditional metric analysis with CV using European iron weapons as an example.
+To test the effectiveness of measuring standardization, @Birch2019 compared landmark-based GMM to traditional metric analysis with CVs using European iron weapons as an example.
 They demonstrated that landmark-based GMM can capture more variation in not only overall shape, but also bilateral symmetry.
 
-For ceramics, @Topi2017 identified that two types of the Casas Grandes vessels in northwest Mexico tend to have standardized shapes using coefficients of variation for the positions of semi-landmarks across shape groups.
-They suggested standardization might hint the presence of specialized producers, reflecting social complexity.
-Another way to explore standardization is pairwise testing of variations in morphological disparity between shape groups by calculating their distances in morphospace, an n-dimensional space where shape groups occupy [@Wills2001].
-In this manner, @Selden2018ceramic2 examined Caddo ceramics in northeast Texas using semi-landmark approaches and found am increase in shape standardization over time, providing a basis for further discussion of craft specialization or group identity [@Selden2019].
-Similarly, a lithic case study suggests that the Gahagan bifaces from the central Texas exhibit less size standardization than those from the southern Caddo area, indicating different uses or tool types [@Selden2020].
-Other applications, such as the study of cranial deformation, demonstrate that landmark approaches with multivariate analyses of shape variances are useful to evaluate shape standardization [@Kuzminsky2016; @Natahi2019; @Perez2007].
-
-A key limitation of landmark approaches in archaeology is that landmarks may be difficult to reproducibly locate for structures that are mostly or entirely curves if not mathematically-defined.
-In those cases, outline approaches, such as Elliptic Fourier Analysis (EFA), is more effective for assessing morphological variations in the whole structure of two-dimensional closed shapes [@Cardillo2010].
+For ceramics, @Topi2017 identified that two types of the Casas Grandes vessels in northwest Mexico tend to have standardized shapes, using coefficients of variation for the positions of semi-landmarks across shape groups.
+They suggested standardization might hint at the presence of specialized producers, reflecting social complexity.
+Another way to explore standardization is pairwise testing of variations in morphological disparity between shape groups by calculating their distances in morphospace, an n-dimensional space that shape groups occupy [@Wills2001].
+In this manner, Seldon [-@Selden2018ceramic2; -@Selden2019] examined Caddo ceramics in northeast Texas using semi-landmark approaches and found an increase in shape standardization over time, providing a basis for further discussion of craft specialization or group identity.
+Similarly, the Gahagan bifaces from the central Texas exhibit less size standardization than those from the southern Caddo area, indicating different uses or tool types [@Selden2020].
+Other applications, such as studies of cranial deformation, demonstrate that landmark approaches with multivariate analyses of shape variances are useful to evaluate shape standardization [@Kuzminsky2016; @Natahi2019; @Perez2007].
+
+A key limitation of landmark approaches in archaeology is that landmarks may be difficult to reproducibly locate for structures that are mostly or entirely curves, if not mathematically-defined.
+In those cases, outline approaches, such as Elliptic Fourier Analysis (EFA), are more effective for assessing morphological variations in the whole structure of two-dimensional closed shapes [@Cardillo2010].
 EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions [@Adams2004; @Bonhomme2014; @Claude2008].
@@
 Later, @Saragusti2005 introduced more functions allowing the calculation of the specific shape attributes, such as symmetry, roughness, and deformation.
-This presents the potential of EFA for the analysis of curves in detail.
+This demonstrated the potential of EFA for the analysis of curves in detail.
 @Iovictua2009 demonstrated a protocol, including outline digitization, EFA procedure, and multivariate linear regression, to compare resharpening trajectories of European Middle Paleolithic stone tools.
-They found that resharpening can be independent of morphology, suggesting functional attributes should be studied separately.
-Recent case studies further support that EFA is applicable and effective to examine lithic assemblages, such as typological classification of Late Woodland points [@Fox2015], the function of flaked obsidian tools in Easter Island [@Lipo2016], the shape and symmetry standardization of the British Acheulean [@Hoggard2019], and the cultural taxonomies for the European Late Palaeolithic [@Ivanovaite2020].
-
-Despite rare ceramic studies using EFA, this approach is promising for analyzing ceramic taxonomy and standardization.
+He found that resharpening can be independent of morphology, suggesting that functional attributes should be studied separately.
+Recent case studies further support the effectiveness of EFA for examining lithic assemblages, e.g.
+for typological classification of Late Woodland points [@Fox2015], analysis of the function of flaked obsidian tools in Easter Island [@Lipo2016], study of the shape and symmetry standardization of the British Acheulean [@Hoggard2019], and investigating cultural taxonomies of the European Late Palaeolithic [@Ivanovaite2020].
+
+Despite few ceramic studies using EFA to date, this approach is promising for analyzing ceramic taxonomy and standardization.
 For example, @Wilczek2014 evaluated the concordance between EFA and Discrete Cosine Transform (DCT), and a traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France.
-The results present that the variation indicated by EFA and DCT matches the traditional ceramic typology, which supports the claim that outline-based approaches can be efficiently used for studying variations in ceramic shapes.
-Furthermore, @Wilczek2014 pointed out the potential of EFA for detecting the level of ceramic standardization.
-In this paper we use EFA to evaluate the level of standardization of ceramics data from Kiwulan, northeastern Taiwan, in relation to the foreign colonial presences to gain insights into the emergence of ceramic specialization.
-Since our ceramic sample lacks sufficient landmarks that could be identified straightforwardly due to their globular body, EFA is a suitable method by focusing the overall shape of an artifact.
-In addition, we use a significance test for the equality of coefficients of variation of shape variables to statistically compare the vessel standardization from different periods.
+They found that the variation indicated by EFA and DCT matches the traditional ceramic typology, which supports the claim that outline-based approaches can be efficiently used for studying variations in ceramic shapes.
+@Wilczek2014's findings demonstrate the potential of EFA for detecting variation in ceramic standardization.
+In this paper we use EFA to evaluate the level of standardization of ceramics data from Kiwulan, northeastern Taiwan around the time of foreign colonial presence to gain insight into the emergence of ceramic specialization.
+The globular shape of the vessels in our sample means that our specimens lacks visually distinctive landmarks, so EFA is an ideal method because of its focus on the overall shape of an artifact.
+In addition, we use a novel significance test for the equality of coefficients of variation of shape variables to statistically compare vessel standardization from different periods.
 
@@
 The chronology of the archaeological deposits consists of two cultural components, the upper and the lower, with a sterile layer in between [@Chen2007].
-We focus on the upper component, dated from AD 1350 to 1850, spans the late Iron Age and the historical period.
+We focus on the upper component, dated from AD 1350 to 1850, because it spans the late Iron Age and the historical period.
 The historical period in Taiwan started with the presence of the Europeans in the early 17th century.
 The Dutch first occupied southern Taiwan in 1624, followed by the Spanish in northern Taiwan in 1626 [@Andrade2007].
-In 1642, the Spanish was expelled by the Dutch, who then occupied the previous Spanish forts at Helping Dau in Keelung, and in Tamsui.
+In 1642, the Spanish were expelled by the Dutch, who then took over the Spanish forts at Helping Dau in Keelung, and in Tamsui.
 Western Taiwan remained under Dutch colonial rule until 1662 when the Kingdom of Tungning in Taiwan was founded by Koxinga, a loyalist of the Ming dynasty of China [@Andrade2007].
@@
 To compare different foreign influences, we classified the upper component into three chronological phases: pre-European, European, and Chinese.
-These phases were identified according to chronologically diagnostic artifacts based on their fined-grained time scale.
-Despite the availability of 17 radiocarbon ages [@Chen2007], the chronological model demonstrates that time range are too coarse to provide sufficient information for the separation of three phases [@WangandMarwick2020].
-Thus, radiocarbon ages only serve as a cross-validation for the diagnostic artifacts if applicable.
+These phases were identified according to chronologically diagnostic artifacts.
+@Chen2007 reports on 17 radiocarbon ages for Kiwulan, and our Bayesian modeling of these ages shows a consistent result with our artefact-based chronology.
+However, because the three phases are relatively brief and the number of ages is small, radiocarbon modeling is of limited value to chronology building in this case [more details in @WangandMarwick2020].
 The diagnostic artifacts include blue and white porcelains, light grey glazed jars, and large dark brown glazed stoneware jars commonly used in the 17th century, and bricks and tiles employed by the Chinese in the 19th century [@Chen2007; @Hsieh2009; @Wang2011].
-We also examined excavation depth measurement and stratigraphic details reported by the excavators (color, texture, disturbance, etc.) to reliably separate three phases.
+We also examined excavation depth measurements and stratigraphic details reported by the excavators (color, texture, disturbance, etc.) to reliably separate the three phases.
 The deposit exhibited signs of continuous human occupation in each of the three phases with no apparent breaks.
-More details for the assignment of different phases are in the Online Supplementary Materials.
+More details for the assignment of different phases are in the Online Supplementary Materials [@WangandMarwick2020].
 
 The most abundant artifacts in the upper component are locally manufactured ceramics, which are distributed throughout the temporal sequence, and across the study area.
-More than 550,000 sherds were revealed, and around 1,200 vessels could be completely or partially reconstructed (i.e.
+More than 550,000 sherds were recovered, and around 1,200 vessels could be completely or partially reconstructed (i.e.
 complete rim or base).
@@
 The exterior surface below the neck is decorated with a variety of impressed geometric motifs.
-These vessels were potentially used for cooking indicated by evidence of the frequent presence of charred residues and carbon deposits on vessel interiors, and soot on vessel exteriors.
+These vessels were likely used for cooking, as indicated by the frequent presence of charred residues and carbon deposits on vessel interiors, and soot on vessel exteriors.
 Firing resulted in orange and brownish color with a fully oxidized core, or a reduced core with oxidized fringes [@Chen2007].
@@
 
-The sample consists of `r total_no` reconstructed vessels with rim, body and base parts, and we securely provenanced to pre-contact (n = `r pre_no`), post-European (n = `r Euro_no`), and Chinese contact contexts (n = `r Chi_no`).
+The sample consists of `r total_no` reconstructed vessels with rim, body and base parts that were securely provenanced to pre-contact (n = `r pre_no`), post-European (n = `r Euro_no`), and Chinese contact contexts (n = `r Chi_no`).
 
@@
 
-We used scans of pottery drawings at 300 dpi acquired from the Bureau of Cultural Affairs in Yilan (Figure \@ref(fig:KWL-pots)).
+We used 300 dpi scans of pottery drawings acquired from the Bureau of Cultural Affairs in Yilan (Figure \@ref(fig:KWL-pots)).
 All drawings provide a two-dimensional view of vessel cross-sections based on metric measurements.
 The scanned drawings were imported into Inkscape (http://inkscape.org) for digitization where outlines were manually traced.
-In those instances where only one side of the cross-section image was available, or some sections were missing, we interpolated the curves and then mirrored and joined to create a closed outline for each vessel.
-Analyses were conducted in R software (R Core Team, 2019) using functions included in the Momocs package for quantifying and analyzing shapes [@Bonhomme2014].
-Outlines were converted into a list of successive x, y pixel coordinates for EFA.
+In those instances where only one side of the cross-section image was available, or small sections were missing, we interpolated the curves and then mirrored and joined to create a closed outline for each vessel.
+Analyses were conducted using R software (R Core Team, 2019) with functions from the Momocs package for quantifying and analyzing shapes [@Bonhomme2014].
+Outlines were converted into a list of successive x and y pixel coordinates for EFA.
 We analyzed harmonic coefficients by principal component analysis (PCA) for dimensionality reduction to illustrate the diversity of the shape data and identify major patterns of variation.
@@
 Following @Eerkens2001 and @Roux2003, we take this as our measurement of standardization in vessel shape variables: lower CV values reflect higher standardization, and thus increased craft specialization in the community.
-Given that CV is robust for positive values due to the representation by a ratio scale, we normalized PC scores to a range between 1 and 10 for the computation of CV.
-
-To answer the question of whether CV values of vessel samples across our three occupational phases are significantly different or not, we used the modified signed-likelihood ratio (MSLR) test for equality of CVs [@Krishnamoorthy2014].
+Given that CVs are most informative when computed on either all positive values or all negative values, we normalized PC scores to a range between 1 and 10 for the computation of CV.
+
+To answer the question of whether CV values across our three occupational phases are significantly different or not, we used the modified signed-likelihood ratio (MSLR) test for equality of CVs [@Krishnamoorthy2014].
 While previous work has used the @Feltz1996's asymptotic test for the equality of coefficients of variation from k populations [@Eerkens2000; @Eerkens2001; @Hoggard2017; @Lycett2008; @Okumura2014], we prefer the MSLR test for shape variables as a more recent development with lower rates of type I error, better performance with uneven sample numbers, and more power across a range of conditions [@Krishnamoorthy2014].
@@
 To complement our investigation of craft specialization through shape standardization, we investigated spatial patterns of ceramic vessels at Kiwulan.
-As craft specialization increases, pottery distribution we expect a shift from a pattern of vessels dispersed across the site to a pattern of clusters that reflects the loci of production [@Costin2001].
+As craft specialization increases, we expect a shift from a pattern of vessels dispersed across the site to a pattern of clusters that reflects the loci of production [@Costin2001].
 We used a Monte Carlo test for randomness in spatial locations of ceramics to robustly test whether their distribution is significantly clustered or dispersed.
@@
 
-Thirteen harmonics captured 99% of the total harmonic power in the elliptic Fourier coefficients of `r total_no` vessels from three discrete phases.
-Figure \@ref(fig:first-two-PCs-byplot-and-tps-plot) illustrates differences in vessel shapes described using thin-plate spline warping for paired periods, pre- and post-European periods, and post-European and Chinese periods, with the greatest differences evidenced between pre-European and Chinese periods.
+Thirteen harmonics captured 99% of the total harmonic power in the elliptic Fourier coefficients of `r total_no` vessels from three phases.
+Figure \@ref(fig:first-two-PCs-byplot-and-tps-plot) illustrates differences in vessel shapes using thin-plate spline warping for paired periods, pre- and post-European periods, and post-European and Chinese periods, with the greatest differences evidenced between pre-European and Chinese periods.
 
@@
 
-```{r, first-two-PCs-byplot-and-tps-plot, echo=FALSE, fig.cap='Left: Significant differences in average vessel shapes between the Chinese and the post-European period using thin plate splines (TPS) with outline deformations required to pass from an extreme of one morphospace to another.
+```{r, first-two-PCs-byplot-and-tps-plot, echo=FALSE, fig.cap='Left: Significant differences in average vessel shapes between the Chinese and the post-European period are visible using thin plate splines (TPS), with outline deformations required to pass from an extreme of one morphospace to another.
 Right: Pottery shape distribution by each occupation phase according to the first two PCs.'}
@@
 The CVs calculated of the three PC support a general trend toward a more standardized shape over time, especially the shape identified by PC1 that represents vessel height and roundness.
-The PC1 shows a higher variation in the pre-European period and post-European period compared to the Chinese period.
+PC1 shows a higher variation in the pre-European period and post-European period compared to the Chinese period.
 That is, a more standardized shape found in the Chinese period.
-However, the PC2 presents a similar diversity in ceramics assemblages across three phases, while the PC3 demonstrates a slightly standardized shape in the Chinese period.
+However, PC2 presents a similar diversity in ceramics assemblages across three phases, while PC3 demonstrates a slightly standardized shape in the Chinese period.
 
@@
 P-values for PC1 show significant differences in shape standardization across periods, between Chinese contact with either pre-European or post-European (Table \@ref(tab:CV-test-PCs-table)).
-This result statistically supports the observation of a higher standardized shape in the Chinese period.
-
-```{r PCs-distribution, fig.cap= "The distribution of normalized PC scores by phases with CV values (%)"}
+This result supports the observation of a more highly standardized shape in the Chinese period.
+
+```{r PCs-distribution, fig.cap= "The distribution of normalized PC scores by phases. CV values (%) are shown in the upper right of each plot."}
 
@@
   mutate(PC = as.character(PC)) %>% 
-  mutate_at(c("MSLRT", "p_value"), round, 4)
+  mutate_at(c("MSLRT", "p_value"), round, 4) %>% 
+  rename(`p-value` = p_value )
 
@@
 
-Vessel size is another important variable for detecting standardization, we used the body diameter of vessels as a proxy of size to examine their variations and relationships with vessel shape.
+Vessel size is another important variable for detecting standardization.
+We used the body diameter of vessels as a proxy of size to examine their variation and relationships with vessel shape.
 We measured body diameter directly from each physical vessel in the collection, and we focus on this metric because it is available for more vessels than any other metric.
-The body diameter of vessels from the Chinese period is larger than those from the two earlier periods, and the vessels before European contact have the smallest body diameter on average (Figure \@ref(fig:size-shape-regression-plots): A).
-To investigate vessel form standardization, represented by shape and size, we compared CV for PC1 (shape) in relation to CV for body diameter (size).
-The result (Figure \@ref(fig:size-shape-regression-plots): B) presents a higher standardization in vessel form in the Chinese period according to smaller CV values compared to those from the other two phases.
+The body diameter of vessels from the Chinese period is larger than those from the two earlier periods, and vessels from before European contact have the smallest body diameter on average (Figure \@ref(fig:size-shape-regression-plots): A).
+To investigate vessel form standardization, represented by shape and size, we compared CV for PC1 (as a shape variable) in relation to CV for body diameter (as a size variable).
+The result (Figure \@ref(fig:size-shape-regression-plots): B) shows a higher standardization in vessel form in the Chinese period, with smaller CV values compared to those from the other two phases.
 However, there are no obvious differences in form standardization before and after the European presence.
-To understand the relationship between shape and size, we modeled linear regression for PCs and the measurement of body diameter, with an examination of Pearson's correlation coefficient and p-value (Figure \@ref(fig:size-shape-regression-plots): C).
-The results demonstrate that shape changes with size in all phases indicated by moderate positive relationships (0.3 ≤ r ≤ 0.7) and small p-values (≤ 0.05), except for PC1 in the Chinese period and PC2 in the pre-European period.
+To understand the relationship between shape and size, we computed linear regression models for PCs and body diameter (Figure \@ref(fig:size-shape-regression-plots): C).
+The results demonstrate that shape and size are positively correlated in all phases, as indicated by moderate positive relationships (0.3 ≤ r ≤ 0.7) and small p-values (≤ 0.05), except for PC1 in the Chinese period and PC2 in the pre-European period.
 In general, the shorter vessels are larger in body diameter according to the significant positive correlation.
@@
 ```{r, size-shape-regression-plots, echo=FALSE, fig.cap= "A: Distribution of the body diameter of vessels with coefficients of variation by phases.
-B: Coefficients of variation of vessel shape represented by PC1 in relation to vessel size represetnted by body diameter, showing a more standardization in vessel form in the Chinese period.
-C: Correlation between shape variables (PC1 and PC2) and body diameter of vessles with Person's r and P-values by phases"}
+B: Coefficients of variation of vessel shape represented by PC1 in relation to vessel size represented by body diameter, showing more standardization (lower CV values) in vessel form and size in the Chinese period.
+C: Correlation between shape variables (PC1 and PC2) and body diameter of vessels with Pearson's r and p-values by phases"}
 knitr::include_graphics(here("analysis/figures/size-boxplot-regression.jpg"))
@@
 This hinted at the emergence of social inequality within the Indigenous community.
-To understand the relationship between social inequality and foreign presences, we examined shape standardization of ceramics to measure craft specialization as a proxy for social change [@Costin2001; @Junker1999].
+To investigate this possible relationship between social inequality and foreign presence, we examined shape standardization of ceramics to measure craft specialization as a proxy for social change [@Costin2001; @Junker1999].
 The result of our MANOVA demonstrates significant differences in shapes between the pre-European and Post-European periods, and pre-European and Chinese periods.
@@
 For the degree of shape standardization, our CV tests on PC1 indicate a significant difference between the Chinese period and either pre-European or post-European periods.
-A test of morphological disparity also demonstrates similar results between Chinese contact and either pre-European (p = $`r pre_chi`$) or post-European periods (p= $`r post_chi`$) [@WangandMarwick2020], suggesting a more standardized shape after contact with the Chinese.
-In addition, we found a more homogeneous shape accompanied by a more standardized size, represented by the body diameter of vessels, in the Chinese period.
-Humans tend to make mistakes in hand-crafting when the size of an object increases, leading to higher variations in larger artifacts [@Eerkens2001].
-However, it is opposite for the ceramics in the Chinese period when we use body diameter as our proxy variable of size.
-This might hint an intentional behavior to achieve a homogeneous form of vessels, but we still need more metric variables to assess the variation in size.
+Analysis of morphological disparity, which measures the positioning of specimens relative to one another in the morphospace [@hopkins2017morphological], supports our finds of shape differences between Chinese contact and either pre-European (p = $`r pre_chi`$) or post-European periods (p= $`r post_chi`$) [@WangandMarwick2020], suggesting a more standardized shape after contact with the Chinese.
+In addition, we found a more homogeneous shape accompanied by a more standardized but also larger size in the Chinese period.
+Generally, people tend to make mistakes in hand-crafting as the size of an object increases, leading to higher variations in larger artifacts [@Eerkens2001].
+However, we found the opposite for ceramics in the Chinese period when using body diameter as our proxy variable of size.
+This might hint an intentional behavior by Kiwulan potters to achieve a homogeneous form for the larger vessels.
 Mineral composition shows that the clay pastes are similar throughout three phases, regardless of the increasing standardization of the pottery shape, reflecting continuity in the raw material sources.
@@
 We note that a small sample size in the Chinese period may lead to a more standardized shape.
-However, this effect can be reduced using CV that scales variation to magnitude, allowing reliable comparison across uneven sample numbers, even for small sample size [@Eerkens2001].
+However, this effect can be reduced using CV statistics that scales variation to magnitude, allowing reliable comparison across uneven sample numbers, even for small sample size [@Eerkens2001].
 Moreover, the MSLR test for equality of CVs enables a robust test between different sample numbers [@Krishnamoorthy2014].
@@
 
-Whether the shape standardization we found suggests craft specialization depends on the number of producers, which distinguishes mechanical standardization from intentional standardization defined by stylish and functional attributes [@Costin2001; @Costin1995].
-Mechanical standardization is used to access the appearance of specialized production based on the assumption that increased skills, routinization, and lower diversity of producers will lead to morphological uniformity [@Arnold2000].
-The population at Kiwulan could be an indicator to estimate the potential number of producers.
+Whether the shape standardization we found results from craft specialization depends partly on the number of producers, which distinguishes mechanical standardization from intentional standardization defined by stylistic and functional attributes [@Costin2001; @Costin1995].
+Mechanical standardization is related to the appearance of specialized production based on the assumption that increased skills, routinization, and lower diversity of producers will lead to morphological uniformity [@Arnold2000].
+in our case, relative changes in the potential number of producers may be inferred from changes in population size at Kiwulan.
 According to the Dutch census in 1648 [@Nakamura1938, pp.
 12], the population at Kiwulan was large but declined in the Chinese period due to the movement of Indigenous people to the south [@Chen2007].
-The change in population size corresponds with the observation of ceramic abundance at Kiwulan.
+This change in population corresponds to a decline in ceramic abundance at Kiwulan.
 Thus, we model CV values as a function of the mass of ceramics using Poisson GLM with a link function.
 The model suggests that ceramic abundance strongly predicts the CV values (β = $`r beta_print`$, p = $`r pval_print`$).
-This indicates that a more standardized shape in the Chinese period may be influenced by a small population if ceramic abundance can reflect the population size.
+This indicates that the more standardized vessel shape of the Chinese period may be influenced by a small population, and thus smaller number of pottery producers, if ceramic abundance can be taken to reflect the population size.
 
 However, intentional standardization due to considerations of function or style could also contribute to the shape standardization in our case.
-To ensure shape standardization can reflect craft specialization, we investigated the function and surface decoration of the vessels.
-For function, we used geochemical methods to extract and identify lipids trapped in the fabric of potsherds to explore foods that may have contributed residues absorbed into the clay [cf.
+To explore this aspect of the relationship between shape standardization and craft specialization, we investigated the function and surface decoration of the vessels.
+We used geochemical methods to extract and identify lipids trapped in the fabric of potsherds to identify foods that may have contributed residues absorbed into the clay [cf.
 @Kwak2015].
-Unfortunately, we did not obtain useful results due to extremely low lipid yields, which is probably due to the very thin, dense, and low porosity fabric of Kiwulan pottery.
+Unfortunately, we did not obtain useful results due to extremely low lipid yields, which were probably due to the very thin, dense, and low porosity fabric of Kiwulan pottery.
 These physical characteristics of the clay offer limited spaces to trap and protect organic molecules from microbiological degradation [cf.
@@
 909].
-For style, surface impressed decoration usually consists of multiple bands of geometric motifs that we define as one type of decoration.
-If two pots share the same set of motifs but different arrangements of single bands, they are considered as two types.
+To analyse style, we defined surface impressed decorations, usually consisting of multiple bands of geometric motifs, as types of decorations.
+If two pots shared the same set of motifs but different arrangements of single bands, we considered them two different types.
 In general, the ceramics in the Chinese period have slightly fewer variations in decoration according to the ratio of distinct types to the total number of pottery from each phase (Chinese = `r chinese_deco`, post-European = `r post_deco`, pre-European = `r pre_deco`) [@WangandMarwick2020].
-The current evidence is not sufficient for us to completely rule out the possibility of intentional standardization.
-
-The spatial pattern of ceramics could provide information about potential production units and production areas [@Costin2001].
-The spatial analysis (Figure \@ref(fig:kernel-plot)) reveals that pottery samples present a widespread distribution with high densities of pottery at some units during European presence.
-Hypothesis testing on spatial randomness indicates a non-randomly dispersed distribution before European contact and more extreme dispersed distribution after European presence.
+The limited evidence about function, and slight differences in style suggest that intentional standardization may have played a minor role at Kiwulan, but further evidence is required to completely rule out this factor.
+
+Additional insight into craft specialization at Kiwulan comes from the spatial pattern of ceramics, which provides information about potential production units and production areas [@Costin2001].
+Figure \@ref(fig:kernel-plot) shows that the pottery samples have a widespread distribution with high densities of pottery at some units during the European presence.
+Hypothesis testing on spatial randomness indicates a non-randomly dispersed distribution before European contact and a more extreme dispersed distribution after European presence.
 In contrast, the distribution of pottery is more similar to random distributions during the Chinese period.
-This contradicts our expectation that a clustered pattern will be observed with an increase in pottery standardization since the emergence of specialized groups [@Costin1995].
+This contradicts our expectation that a clustered pattern will be observed with an increase in pottery standardization caused by the emergence of specialized groups [@Costin1995].
 The absence of clusters in the Chinese period is notable because this was a time of a historically-documented decline of Indigenous population [@Chen2007; @Hsieh2009].
@@
 This suggests that differences in population across our three occupation phases are probably not driving variation in shape standardization.
-Also, it is more likely that ceramics are household-produced based on the spatial pattern and no specific facilities of production have been found [@Chen2007].
+Also, it is more likely that ceramics are household-produced, based on the spatial pattern, and no specific facilities of production are evident [@Chen2007].
 
@@
 The quantity is indicated by the color scale.
-B: Kernel density map visulizes the probability of the density of pottery across space.
+B: Kernel density maps visulize the probability of the density of pottery across space.
 The maps show a major core area during the pre-European period, multiple core areas during the European period, and a single core during the Chinese period.
@@
 
-Our results offer tentative support for the hypothesis that foreign presence at Kiwulan influenced the shape of vessels in the local Indigenous society.
+Our results offer tentative support for the hypothesis that foreign presence at Kiwulan influenced the shape of vessels made by the local Indigenous society.
+We find that vessel shapes were more standardized during the Chinese period than the European period.
 If increased shape standardization is a reliable indicator of craft specialization, then we may be seeing evidence of a shift from corporate (group-based, distributed, collective, cooperative) to network (individual-based, competitive) organization [@Blanton1996; @Feinman1995; @Feinman2000; @Feinman2000political; @Feinman2010].
 However, strong claims for an emergence of social inequality resulting from foreign contact at Kiwulan will need support from multiple and diverse sources of evidence that are beyond the scope of this paper.
-We find that vessel shapes were more standardized during the Chinese period than the European period.
 
@@
 @Acabado2017].
-In contrast to the Indigenous-European interactions at Kiwulan, the interaction between Indigenous people and Chinese immigrants in the 19th century appears to have been more intense and direct.
+In contrast to the Indigenous-European interactions at Kiwulan, interaction between Indigenous people and Chinese immigrants in the 19th century appears to have been more intense and direct.
 Historical records indicate that Chinese groups settled in Yilan and lived closely with Kiwulan Indigenous societies [@Chen1963; @Ke1993].
-This direct influence is reflected by the archaeological evidence of large amounts of Chinese porcelains and distinctive architectural bricks and tiles used by Chinese [@Hsieh2009].
+This direct influence is reflected by the archaeological evidence of large amounts of Chinese porcelain and distinctive Chinese architectural bricks and tiles [@Hsieh2009].
 Similarly, burials at Kiwulan in this later phase show the adoption of coffins in mortuary practices, which @Chiu2004 interprets as the adoption of a symbol of ethnic Chinese.
 
-The shape variation reported here is subtle and invites consideration of another possible scenario that the absence of major changes in vessel shape at Kiwulan may have been an act to show ethnic identity when experiencing foreign influences [cf.
+The shape variation reported here is subtle and invites consideration of another possible scenario, namely that the absence of major changes in vessel shape at Kiwulan may have been an act to show ethnic identity when experiencing foreign influences [cf.
 @Torrence2000].
-We should note that the morphology of ceramics could be treated as a signal for message communication between potters that repeating the same shape or modifying may have occurred in a non-deterministic way [@Kubler1962].
-However, as the only type of locally made pottery throughout 600 years, the homogeneous and even more standardized shape in the Chinese period might convey some meaningful information.
+Ceramic morphology could be a signal for communication between potters, such that repeating the same shape or slight modifications may have occurred in a non-deterministic way [@Kubler1962].
+As the only type of locally made pottery throughout 600 years, the homogeneous and even more standardized shape in the Chinese period might convey some meaningful information about community identity.
 We recognize the decline in population may lead to a standardized shape, however, the evidence of ceramic spatial distribution shows a dispersed or random pattern across the whole area throughout three phases.
-This indicates manufacturing activities may be limited to household scale and reflect a common practice in the community.
+This indicates manufacturing activities may be limited to the household scale and reflect a common, shared practice in the community.
 The standardization in vessel shape over time draws our attention to the endurance of traditional pottery production practices amid intrusions from Europeans and Chinese.
-In a culture contact situation, social identity may be expressed through material practices as a means of expressing cultural homogeneity and distinction from other groups [@Voss2005].
+In a culture contact situation, we speculate that social identity might have be expressed through material practices as a means of expressing cultural homogeneity and distinction from other groups [@Voss2005].
 
 It is also important to recognize that social identity might be more complicated in a colonial context, and maybe representative of more than a colonized–colonizer or local/foreign dichotomy [@Voss2005; @Voss2008].
-Shamaoshan cemetery dating from BC 3 to AD 4 in Southwest China suggests that the process of the incorporation of Southwest China into the Han Empire involved a century of conflicts, resistance, and acceptance among social groups with different identities, especially in the historical context of Han immigrants [@Wu2019].
+For example, the Shamaoshan cemetery (BC 250 to 55 AD) in Southwest China suggests that the process of the incorporation of Southwest China into the Han Empire involved a century of conflicts, resistance, and acceptance among social groups with different identities, especially in the historical context of Han immigrants [@Wu2019].
 A similar dynamic may have occurred at Kiwulan, with vessel shape indicating both acceptance of foreign influence through increased shape standardization, and resistance through the overall continuity in vessel shape.
-Vessel shape can be viewed as a symbolic expression of the Indigenous identity and social boundaries because shape is a highly visible trait compared with other features of pottery [cf.
+Vessel shape may be viewed as a symbolic expression of Indigenous identity and social boundaries because shape is a highly visible trait compared with other features of pottery [cf.
 @Roux2015].
@@
 @Acabado2017].
-However, we still need more lines of evidence to confirm this speculation of an act of resistance.
+However, additional lines of evidence are necessary from Kiwulan to confirm this speculation.
 
@@
 This study demonstrated the first use of EFA on ceramic shapes to explore the emergence of ceramic specialization as indicative of foreign influences.
-Here, EFA is combined with a significance test for the equality of CVs of shape variables to provide a robust method for assessing differences in shape standardization.
+Here, EFA is combined with significance tests for the equality of CVs of shape variables to provide a robust method for assessing differences in shape standardization.
 The direct relationship between foreign influences and standardization of ceramic shape was tested on ceramics from Kiwulan, a large Iron Age Indigenous settlement in northeastern Taiwan.
@@
 Our findings help to expand upon those factors that may lead to the standardization of ceramic production in a pericolonial interaction context.
-More homogeneous vessel shapes and sizes during the Chinese period, without any substantial changes in clay paste composition, production technique, and spatial distribution, suggest that shape standardization may relate more to intentional behavior instead of the emergence of specialized groups.
-Despite a small sample size in the Chinese period, the random distribution of ceramics does not support clustered patterns of manufacturing location, such as workshops.
-Instead, the ceramic production is likely to be manufactured in households.
-The homogeneous appearance may suggest expressions of social identity or cultural boundaries in Indigenous societies through highly visible vessel qualities, such as shape, maybe heightened during periods of foreign contact in pericolonial contexts.
-Our analysis, with its openly available methods and data, is readily extensible to other pottery assemblages in the region to further explore related questions about craft specialization and standardization in Iron Age ceramic assemblages.
-This study also broadens the GMM field by focusing on ceramic technologies, which could motivate more ceramic studies and become a promising branch parallel to current applications to lithic typology and bone morphology.
+More homogeneous vessel shapes and sizes during the Chinese period, without any substantial changes in clay paste composition, production technique, and spatial distribution, suggest that shape standardization may relate more to intentional behaviors rather than the emergence of discrete groups of producers.
+Despite a smaller sample size in the Chinese period, the random distribution of ceramics does not support clustered patterns of manufacturing location, such as workshops.
+Instead, the ceramic production is likely to have occurred in households.
+We speculate that the relatively homogeneous appearance of the vessels may suggest an expression of social identity or cultural boundaries in Indigenous societies through highly visible vessel qualities, such as shape, which may be heightened during periods of foreign contact in pericolonial contexts.
+Our analysis, with its openly available methods and data, is readily extensible to other pottery assemblages in the region to further explore related questions about craft specialization and standardization in ceramic assemblages.
 

### 2020-08-21 analysis/code/check-diffs.R
add code to view rich diff of Rmd

@@ -1,0 +1,80 @@
+
+# How to see rich diffs of two commits of a single R Markdown document in a local Git repo
+
+# To browse the commits locally
+
+# https://github.com/lorenzwalthert/gitsum
+# devtools::install_github("lorenzwalthert/gitsum")
+library("gitsum")
+library("tidyverse")
+
+# Set the path within the project to the Rmd file that we want to compare versions of
+file_path <- "analysis/paper/paper.Rmd"
+
+# We make a data frame that store the git commit history
+init_gitsum()
+tbl_out <- parse_log_detailed() %>%
+  select(date, author_name, message, hash)
+
+# View this dataframe to pick the specific commits to compare
+# View(tbl_out)
+
+# choose two commits to compare, here I've got the most recent
+# one (the last row of the table), and the one 2-3 before that
+i <- nrow(tbl_out)
+j <- nrow(tbl_out) - 3
+
+# construct the terminal commands using the SHA of the commits we want to
+# compare, and the file path
+git_show_str_1 <- str_glue('git show ', tbl_out[i, ]$hash, ":", file_path)
+git_show_str_2 <- str_glue('git show ', tbl_out[j, ]$hash, ":", file_path)
+
+# get the contents of the file at each commit and
+# store in a vector
+earlier_commit <- system(git_show_str_2, intern = TRUE)
+later_commit <- system(git_show_str_1, intern = TRUE)
+
+# compute differences in visualise
+diffobj::diffChr(earlier_commit,
+                  later_commit,
+                  mode = "sidebyside")
+
+
+
+### --------------------------------------------------------------------
+# How to see rich diffs of two commits of a single R Markdown document on a GitHub repo
+
+# Edit this to use the repo that has the file you want to inspect
+gh_account_and_repo <- "LiYingWang/kwl.pottery"
+
+# Edit this to use the single file that you want to inspect
+# start the path here from the top level of the github repo
+gh_file_path <- "analysis/paper/paper.Rmd"
+
+# open this web page to see the commit history so we can
+# copy the SHA ids to use in the next step, look for the
+# little clipboard icon to click on to copy the full SHA,
+# and past them in the step below
+browseURL(file.path(
+  "https://github.com",
+  gh_account_and_repo,
+  "commits/master",
+  gh_file_path
+))
+
+# Edit these to get the full SHA for the commits you want to compare, get
[…]

### 2020-08-21 analysis/paper/KWL-chronology-modeling.Rmd
minor edits when talking about 14C ages

@@
 We calibrated 23 dates from different contexts, including normal deposit, burials, and middens, at Kiwulan using oxcAAR package, which executes OxCal v4.4.1 with radiocarbon calibration curve IntCal20 [@Reimer2020].
+<!-- if this is not the full set of ages in Chen, then please explain how you choose them -->
 
@@
 The end of post-European is identified by the start of the Chinese migration to Yilan in 1780s [@Chen2007].
-The results suggest that pre-European started between 1331–1635 cal AD at a 95.4% probability range and ended between 1458–1670 cal AD at a 99.7% probability range.
+<!-- please explain how you choose the subset of ages to go into this model -->
+
+The results suggest that the re-European period started between 1331–1635 cal AD at a 95.4% probability range and ended between 1458–1670 cal AD at a 99.7% probability range.
 The end date of pre-European also represents the start date of post-European, while the end date of post-European is 1517-1753 cal AD at a 95.4% probability range.
@@
 three_phases_result_con_data <- parseOxcalOutput(three_phases_con_text, only.R_Date = F)
-str(three_phases_result_con_data)
-print(three_phases_result_con_data)
+# str(three_phases_result_con_data)
+# print(three_phases_result_con_data)
 plot(three_phases_result_con_data)

### 2020-08-21 analysis/code/KWL-dates-model.R

@@ -1,75 +1,0 @@
-kw_dates <- readxl::read_excel(here::here("analysis", "data", "raw_data", "KWL_C14_dates.xlsx"))
-library(tidyverse)
-# subset dates with error value
-kw_dates_with_error <-
-  kw_dates  %>%
-  filter(!is.na(se))
-
-# convert rce to numeric
-kw_dates_with_error$rce <- as.numeric(kw_dates_with_error$rce)
-
-# make some kind of ID
-kw_dates_with_error$id <- with(kw_dates_with_error,
-                               paste0(Square,"-",Layer))
-
-# Use oxcAAR to calibrate
-library(devtools)
-# install_github('ISAAKiel/oxcAAR')
-library(oxcAAR)
-quickSetupOxcal()
-date_kwl <- oxcalCalibrate(kw_dates_with_error$rce,
-                           kw_dates_with_error$se,
-                           kw_dates_with_error$`Lab code`)
-# date_kwl
-# plot(date_kwl)
-
-# for modeling
-kw_dates_with_error_three_phases <-
-  kw_dates_with_error %>%
-  filter(rce < 600) %>%
-  mutate(phase = rce) %>%
-  mutate(phase = ifelse(rce > 324, "Pre European", phase),
-         phase = ifelse(rce < 324, "Post European", phase)) %>%
-  select(`Lab code`, "rce", "se", "phase")
-
-# use continguous example from
-# https://c14.arch.ox.ac.uk/oxcalhelp/hlp_analysis_eg.html#phases
-three_phases <- 'Plot()
-{
- Phase("Upper Layer")
- {
-    Sequence()
-    {
-      Boundary("Start of Pre-European");
-      Phase()
-      {
-        R_Date("NTU-4292", 510, 75);
-        R_Date("NTU-4310", 360, 100);
-        R_Date("NTU-4320", 340, 100);
-        R_Date("NTU-3791", 340, 30);
-        Interval( "Duration of Pre-European");
-      };
-   Boundary("Pre-European - Post-European");
-   Phase( )
-        {
-        R_Date("NTU-3993", 250, 40);
-        R_Date("NTU-4016", 270, 45);
-        R_Date("NTU-4311", 310, 100);
-        R_Date("NTU-4419", 280, 70);
-        Interval("Duration of Post-European");
-        };
-      Boundary("End of Post-European");
-    };
-  };
-};
-'
-three_phases_results <- executeOxcalScript(three_phases)
[…]

### 2020-08-21 analysis/paper/paper.Rmd
minor edits when talking about 14C ages
minor edits about intentional vs mechanical

@@
 These phases were identified according to chronologically diagnostic artifacts.
-@Chen2007 reports on 17 radiocarbon ages for Kiwulan, and our Bayesian modeling of these ages shows a consistent result with our artefact-based chronology.
+Our Bayesian modeling of eight ages from @Chen2007 shows a consistent result with our artefact-based chronology.
 However, because the three phases are relatively brief and the number of ages is small, radiocarbon modeling is of limited value to chronology building in this case [more details in @WangandMarwick2020].
@@
 Mechanical standardization is related to the appearance of specialized production based on the assumption that increased skills, routinization, and lower diversity of producers will lead to morphological uniformity [@Arnold2000].
-in our case, relative changes in the potential number of producers may be inferred from changes in population size at Kiwulan.
+In our case, relative changes in the potential number of producers may be inferred from changes in population size at Kiwulan.
 According to the Dutch census in 1648 [@Nakamura1938, pp.
@@
 In general, the ceramics in the Chinese period have slightly fewer variations in decoration according to the ratio of distinct types to the total number of pottery from each phase (Chinese = `r chinese_deco`, post-European = `r post_deco`, pre-European = `r pre_deco`) [@WangandMarwick2020].
-The limited evidence about function, and slight differences in style suggest that intentional standardization may have played a minor role at Kiwulan, but further evidence is required to completely rule out this factor.
+The limited evidence about function, and slight differences in style suggest that intentional standardization may have played only a minor role at Kiwulan, and further evidence is required to completely rule out this factor.
 
@@
 In contrast, the distribution of pottery is more similar to random distributions during the Chinese period.
-This contradicts our expectation that a clustered pattern will be observed with an increase in pottery standardization caused by the emergence of specialized groups [@Costin1995].
-The absence of clusters in the Chinese period is notable because this was a time of a historically-documented decline of Indigenous population [@Chen2007; @Hsieh2009].
+This is interesting because it contradicts our expectation that a clustered pattern will be observed with an increase in pottery standardization caused by the emergence of specialized groups [@Costin1995].
+The absence of clusters in the Chinese period is notable because this was a time of a historically-documented decline of the Indigenous population [@Chen2007; @Hsieh2009].
 We might expect reduced numbers of potters to result in pottery production shrinking to a few locations in the settlement during this time.
-Despite the small number of vessels during the Chinese contact period, Figure \@ref(fig:kernel-plot) shows that pottery is distributed randomly across the sampling area without any distinctive clusters during this time.
-This suggests that differences in population across our three occupation phases are probably not driving variation in shape standardization.
-Also, it is more likely that ceramics are household-produced, based on the spatial pattern, and no specific facilities of production are evident [@Chen2007].
+However, despite the small number of vessels during the Chinese contact period, Figure \@ref(fig:kernel-plot) shows that pottery is distributed randomly across the sampling area without any distinctive clusters during this time.
+As population across our three occupation phases declines, we see less clustered distributions of pottery, supporting an interpretation of intentional standardization rather than mechanical standardization.
+The spatial pattern shows that ceramics were mostly household-produced, and no specific facilities of production are evident [@Chen2007].
 
@@
 Our findings help to expand upon those factors that may lead to the standardization of ceramic production in a pericolonial interaction context.
-More homogeneous vessel shapes and sizes during the Chinese period, without any substantial changes in clay paste composition, production technique, and spatial distribution, suggest that shape standardization may relate more to intentional behaviors rather than the emergence of discrete groups of producers.
-Despite a smaller sample size in the Chinese period, the random distribution of ceramics does not support clustered patterns of manufacturing location, such as workshops.
-Instead, the ceramic production is likely to have occurred in households.
-We speculate that the relatively homogeneous appearance of the vessels may suggest an expression of social identity or cultural boundaries in Indigenous societies through highly visible vessel qualities, such as shape, which may be heightened during periods of foreign contact in pericolonial contexts.
+More homogeneous vessel shapes and sizes during the Chinese period, without any substantial changes in clay paste composition, production technique, and spatial distribution, suggest that shape standardization emerges from a combination of mechanical and intentional factors.
+Discrete groups of producers are not evident, favoring the role of intentional factors in this case.
+The distribution of ceramics in the Chinese period does not support clustered patterns of manufacturing locations, such as workshops.
+Instead, ceramic production is likely to have occurred in households.
+We speculate that the relatively homogeneous appearance of the vessels may suggest an expression of social identity or cultural boundaries in Indigenous societies through highly visible vessel qualities, such as shape.
+The symbolic value of these shapes may be heightened during periods of foreign contact in pericolonial contexts.
 Our analysis, with its openly available methods and data, is readily extensible to other pottery assemblages in the region to further explore related questions about craft specialization and standardization in ceramic assemblages.

### 2020-08-22 analysis/paper/KWL-chronology-modeling.Rmd
update file, code, and text for the C14 modeling file

@@
 ---
-title: "Supplementary Data: 14C Modeling for Kiwualn Chronology"
-author: "Liying Wang and Ben Marwick"
+title: "Supplementary Data: 14C Modeling for Kiwulan Chronology"
+author: "Li-Ying Wang and Ben Marwick"
 date: "8/18/2020"
@@
 
-This document provides data and code for modeling the chronology of the upper component of Kiwulan.
-The full discussion of chronology is in the paper: The paper Standardization of ceramic shape: A case study from the Iron Age pottery from northeastern Taiwan.
-In this document, we calibrated all C14 dates [@Chen2007] first and then further modeled different phases related to foreign influences, pre- and post- European periods.
-Our purpose is to answer whether the C14 dates support a time range that corresponds to historical events of the presence of Europeans in Taiwan.
-The methods and results are included below.
+This document provides data and code for calibrating C14 dates and modeling the chronology of the upper component of Kiwulan.
+The full discussion of chronology is in the paper: Standardization of ceramic shape: A case study from the Iron Age pottery from northeastern Taiwan.
+There are 17 C14 dates from the upper component according to the excavation report [@Chen2007].
+We calibrated 11 C14 dates after removing 6 dates that are not informative due to a long time range without standard deviation (< 200 yr).
+We further modeled two phases related to the European influence, pre- and post- European periods.
+Our purpose is to answer whether the C14 dates support a time range that corresponds to the historical events of the presence of the Europeans in Taiwan.
 
@@
                                paste0(Square,"-",Layer))
+
+# upper layer with estimation of phases
+kw_dates_with_error_three_phases <-
+  kw_dates_with_error %>%
+  filter(Component == "upper") %>%
+  mutate(phase = rce) %>%
+  mutate(phase = ifelse(rce > 324, "Pre European", phase),
+         phase = ifelse(rce < 324, "Post European", phase)) %>%
+  select(`Lab code`, "rce", "se", "phase", Context)
 ```
@@
 
-We calibrated 23 dates from different contexts, including normal deposit, burials, and middens, at Kiwulan using oxcAAR package, which executes OxCal v4.4.1 with radiocarbon calibration curve IntCal20 [@Reimer2020].
-<!-- if this is not the full set of ages in Chen, then please explain how you choose them -->
+We calibrated 11 dates from different contexts, including artifact-bearing deposit, burials, and middens, at Kiwulan using oxcAAR package, which executes OxCal v4.4.1 with radiocarbon calibration curve IntCal20 [@Reimer2020].
 
@@
 
-```{r data-for-modeling}
-# upper layer with estimation of phases
-kw_dates_with_error_three_phases <-
-  kw_dates_with_error %>%
-  filter(rce < 600) %>%
-  mutate(phase = rce) %>%
-  mutate(phase = ifelse(rce > 324, "Pre European", phase),
-         phase = ifelse(rce < 324, "Post European", phase)) %>%
-  select(`Lab code`, "rce", "se", "phase")
-```
-
 # Chronological model
 
-We then constructed a Bayesian radiocarbon calibration model to obtain the range of pre-European and post-European we examined.
-We assume the phases are in an order that represents before and after the European arrival in 1624 [@Andrade2007].
-The end of post-European is identified by the start of the Chinese migration to Yilan in 1780s [@Chen2007].
-<!-- please explain how you choose the subset of ages to go into this model -->
+We then constructed a Bayesian radiocarbon calibration model for the 11 dates to obtain the range of the pre-European and the post-European periods we examined.
+We assume that the two phases are in a time sequence with the arrival of the Europeans in 1624 as a break [@Andrade2007].
+The end of the post-European period is identified by the start of the Chinese migration to Yilan in 1780s [@Chen2007].
 
-The results suggest that the re-European period started between 1331–1635 cal AD at a 95.4% probability range and ended between 1458–1670 cal AD at a 99.7% probability range.
-The end date of pre-European also represents the start date of post-European, while the end date of post-European is 1517-1753 cal AD at a 95.4% probability range.
-The overall model is consistent with the historical data.
+The results suggest that the pre-European period started between 998–1386 cal AD at a 95.4% probability range and ended between 1496–1656 cal AD at a 95.4% probability range.
+The end date of pre-European is the same as the start date of post-European (1496–1656 cal AD).
+The end date of post-European is 1524-1770 cal AD at a 93.4% probability range.
+In general, the overall model is consistent with the historical data that we expect.
 
@@
         R_Date("NTU-4320", 340, 100);
+        R_Date("NTU-4304", 600, 75);
+        R_Date("NTU-4423", 610, 90);
+        R_Date("NTU-4319", 920, 105);
         Interval( "Duration of Pre-European");

### 2020-08-22 analysis/code/check-diffs.R

@@ -21,8 +21,8 @@ tbl_out <- parse_log_detailed() %>%
 
 # choose two commits to compare, here I've got the most recent
 # one (the last row of the table), and the one 2-3 before that
-i <- nrow(tbl_out)
-j <- nrow(tbl_out) - 3
+i <- nrow(tbl_out) -1
+j <- nrow(tbl_out) -4
 
 # construct the terminal commands using the SHA of the commits we want to
 # compare, and the file path
@@ -38,8 +38,6 @@ later_commit <- system(git_show_str_1, intern = TRUE)
 diffobj::diffChr(earlier_commit,
                   later_commit,
                   mode = "sidebyside")
-
-
 
 ### --------------------------------------------------------------------
 # How to see rich diffs of two commits of a single R Markdown document on a GitHub repo

### 2020-08-22 analysis/paper/paper.Rmd
update file, code, and text for the C14 modeling file
added a reference and text for it

@@
   1. Iron Age pottery shapes from Kiwulan, Taiwan, were analyzed using elliptical Fourier analysis
-  2. Statistical test for the equality of coefficients of variation of shape coefficients was applied
+  2. Statistical tests for the equality of coefficients of variation of shape coefficients were applied
   3. Significant differences in pottery shape were identified between culture contact phases
@@
 Key questions in archaeological shape analysis normally involve measuring shape standardization over time, or between geographical areas.
-Standardization is often investigated using multivariate analysis of shape variables computed from landmark data, coupled with coefficients of variation on associated metric data, especially for lithic assemblages.
+Standardization is often investigated using multivariate analysis of shape variables computed from landmark data, along with coefficients of variation on associated metric data, especially for lithic assemblages.
 For example, @Archer2015's case study of stone points in Southern Africa suggests an increase in shape standardization over time that may relate to increased maintenance of finished points.
@@
 These phases were identified according to chronologically diagnostic artifacts.
-Our Bayesian modeling of eight ages from @Chen2007 shows a consistent result with our artefact-based chronology.
+Our Bayesian modeling of 11 ages related to the upper component from @Chen2007 shows a consistent result with our artifact-based chronology.
 However, because the three phases are relatively brief and the number of ages is small, radiocarbon modeling is of limited value to chronology building in this case [more details in @WangandMarwick2020].
@@
 For example, the Shamaoshan cemetery (BC 250 to 55 AD) in Southwest China suggests that the process of the incorporation of Southwest China into the Han Empire involved a century of conflicts, resistance, and acceptance among social groups with different identities, especially in the historical context of Han immigrants [@Wu2019].
+Another example demonstrates Indigenous endurance and resilience through the long-term  persistence of mound use in Mississippian culture after contact with European colonizers [@Holland].
 A similar dynamic may have occurred at Kiwulan, with vessel shape indicating both acceptance of foreign influence through increased shape standardization, and resistance through the overall continuity in vessel shape.
@@
 Our analysis, with its openly available methods and data, is readily extensible to other pottery assemblages in the region to further explore related questions about craft specialization and standardization in ceramic assemblages.
+This study also broadens the GMM field by focusing on ceramic technologies, which may motivate more ceramic studies and become a promising branch parallel to current applications to lithic typology and bone morphology.
 

### 2020-08-22 analysis/paper/paper.Rmd
edit discussion section

@@
 In a culture contact situation, we speculate that social identity might have be expressed through material practices as a means of expressing cultural homogeneity and distinction from other groups [@Voss2005].
-
 It is also important to recognize that social identity might be more complicated in a colonial context, and maybe representative of more than a colonized–colonizer or local/foreign dichotomy [@Voss2005; @Voss2008].
 For example, the Shamaoshan cemetery (BC 250 to 55 AD) in Southwest China suggests that the process of the incorporation of Southwest China into the Han Empire involved a century of conflicts, resistance, and acceptance among social groups with different identities, especially in the historical context of Han immigrants [@Wu2019].
-Another example demonstrates Indigenous endurance and resilience through the long-term  persistence of mound use in Mississippian culture after contact with European colonizers [@Holland].
+At the Oconee Valley (AD 1540 to 1670) in the Southeastern United States, Indigenous endurance and resilience is indicated by the long-term persistence of mound use lasting for 130 years after contact with European colonizers [@Holland].
 A similar dynamic may have occurred at Kiwulan, with vessel shape indicating both acceptance of foreign influence through increased shape standardization, and resistance through the overall continuity in vessel shape.

### 2020-08-22 analysis/paper/KWL-chronology-modeling.Rmd
updated chronology modeling file

@@
 ---
-title: "Supplementary Data: 14C Modeling for Kiwulan Chronology"
+title: "Supplementary Data: Radiocarbon Age Modeling for Kiwulan Chronology"
 author: "Li-Ying Wang and Ben Marwick"
@@
 
-This document provides data and code for calibrating C14 dates and modeling the chronology of the upper component of Kiwulan.
+This document provides data and code for calibrating radiocarbon dates and modeling the chronology of the upper component of Kiwulan.
 The full discussion of chronology is in the paper: Standardization of ceramic shape: A case study from the Iron Age pottery from northeastern Taiwan.
-There are 17 C14 dates from the upper component according to the excavation report [@Chen2007].
-We calibrated 11 C14 dates after removing 6 dates that are not informative due to a long time range without standard deviation (< 200 yr).
-We further modeled two phases related to the European influence, pre- and post- European periods.
-Our purpose is to answer whether the C14 dates support a time range that corresponds to the historical events of the presence of the Europeans in Taiwan.
+There are 17 radiocarbon dates from the upper component according to the excavation report [@Chen2007].
+We calibrated 11 radiocarbon dates after removing 6 dates that are not informative due to their long time ranges without standard deviation (< 200 yr).
+We then modeled two phases related to the European influence, pre- and post- European periods.
+Our purpose is to answer whether the radiocarbon dates support a time range that corresponds to our artifact-based chronology in terms of the presence of Europeans in Taiwan.
 
@@
 
-We then constructed a Bayesian radiocarbon calibration model for the 11 dates to obtain the range of the pre-European and the post-European periods we examined.
-We assume that the two phases are in a time sequence with the arrival of the Europeans in 1624 as a break [@Andrade2007].
+We constructed a Bayesian radiocarbon calibration model for the 11 dates to obtain the range of the pre-European and the post-European periods we examined.
+We assume that those two phases are in a time sequence with the arrival of the Europeans in 1624 as a break [@Andrade2007].
 The end of the post-European period is identified by the start of the Chinese migration to Yilan in 1780s [@Chen2007].
@@
 The results suggest that the pre-European period started between 998–1386 cal AD at a 95.4% probability range and ended between 1496–1656 cal AD at a 95.4% probability range.
-The end date of pre-European is the same as the start date of post-European (1496–1656 cal AD).
-The end date of post-European is 1524-1770 cal AD at a 93.4% probability range.
-In general, the overall model is consistent with the historical data that we expect.
+The start date of the post-European period is the same as the end date of the pre-European period (1496–1656 cal AD).
+The end date of the post-European period is 1524-1770 cal AD at a 93.4% probability range.
+In general, the overall model for radiocarbon age relative to the European contact is consistent with our artifact-based chronology.
 

### 2020-08-23 analysis/paper/paper.Rmd
knit and added online supplementary materials to data folder

--- 2020-08-22 analysis/paper/paper.Rmd
@@
 In a culture contact situation, we speculate that social identity might have be expressed through material practices as a means of expressing cultural homogeneity and distinction from other groups [@Voss2005].
-
 It is also important to recognize that social identity might be more complicated in a colonial context, and maybe representative of more than a colonized–colonizer or local/foreign dichotomy [@Voss2005; @Voss2008].
 For example, the Shamaoshan cemetery (BC 250 to 55 AD) in Southwest China suggests that the process of the incorporation of Southwest China into the Han Empire involved a century of conflicts, resistance, and acceptance among social groups with different identities, especially in the historical context of Han immigrants [@Wu2019].
-Another example demonstrates Indigenous endurance and resilience through the long-term  persistence of mound use in Mississippian culture after contact with European colonizers [@Holland].
+At the Oconee Valley (AD 1540 to 1670) in the Southeastern United States, Indigenous endurance and resilience are indicated by the long-term persistence of mound use lasting for 130 years after the initial contact with European colonizers [@Holland].
 A similar dynamic may have occurred at Kiwulan, with vessel shape indicating both acceptance of foreign influence through increased shape standardization, and resistance through the overall continuity in vessel shape.
@@
 @Roux2015].
-Although there is an increase in number of imported ceramics through time at Kiwulan, production of local ceramics was continuous, and increasingly standardized.
+Although there is an increase in number of imported ceramics through time at Kiwulan, production of the same type of local ceramics was continuous, and increasingly standardized.
 This might imply not only the utilitarian function, but an intentional and increased emphasis on the local ceramic tradition, their cultural custom, as a response to intensified foreign contact [cf.

### 2020-08-29 analysis/paper/paper.Rmd
update acknowledgments

@@
                   phase = rep(op.p$fac$phase, 3),
-                  pcn = str_glue('PC{sort(rep(1:3, len))}'))
+                  pcn = as.character(str_glue('PC{sort(rep(1:3, len))}')))
 
@@
 This research used statistical consulting resources provided by the Center for Statistics and the Social Sciences, University of Washington.
+We also thank anonymous reviewers for their insightful comments that have greatly improved this manuscript.
 

### 2020-09-08 analysis/code/thin_section.R
get stat values for petrography

@@ -1,4 +1,5 @@
 # read data
+library(tidyverse)
 library(readxl)
 thin_section <- read_excel(here::here("analysis", "data", "raw_data", "kwl_petrography.xlsx"), 2)
 sherd <- read_excel(here::here("analysis", "data", "raw_data", "kwl_petrography.xlsx"), 1)
@@ -91,7 +92,7 @@ ggplot(data = scores,
   ggtitle("PCA plot of thin sections - Mineral")
 
 # MANOVA
-man_pca <- manova(pca1$x ~pca1$phase)
+man_pca <- manova(pca1$x ~ pca1$phase)
 sum_man_all <- summary(man_pca, test="Pillai")
 summary.aov(man_pca)
 
@@ -114,4 +115,7 @@ sum_man_all_df <-
          `degrees of freedom` = `stats.den.Df`,
          `Pr(>F)` = `stats.Pr..F.`)
 
+#  Pillai’s Trace = 0.465, F = 0.775, df = 46,
+
 pval_all <- signif(unname(sum_man_all_df[,'Pr(>F)'][1]), 4)
+

### 2020-09-08 analysis/paper/paper.Rmd
proof editing

@@
 ---
-title: "Standardization of ceramic shape: A case study from the Iron Age pottery from northeastern Taiwan"
+title: "Standardization of ceramic shape: A case study of Iron Age pottery from northeastern Taiwan"
 
@@
 abstract: |
-  The emergence of ceramic specialization in prehistoric societies is often linked to shifts in the complexity of social structures, because standardized ceramic production can reflect craft specialization and the presence of elite control.
+  The emergence of ceramic specialization in past societies is often linked to shifts in the complexity of social structures, because standardized ceramic production can reflect craft specialization and the presence of elite control.
 Previous work on identifying specialization relies on typological or linear metric analysis.
@@
 keywords: |
-  Geometric Morphometrics; craft specialization; Iron age; culture contact; ceramic standardization
+  Geometric Morphometrics; craft specialization; Iron age; culture contact; ceramic standardization; Taiwan
 highlights: |
@@
 However, this effect can be reduced using CV statistics that scales variation to magnitude, allowing reliable comparison across uneven sample numbers, even for small sample size [@Eerkens2001].
-Moreover, the MSLR test for equality of CVs enables a robust test between different sample numbers [@Krishnamoorthy2014].
+Moreover, the MSLR test for equality of CVs enables a robust test between different sample sizes [@Krishnamoorthy2014].
 
@@
 Figure \@ref(fig:kernel-plot) shows that the pottery samples have a widespread distribution with high densities of pottery at some units during the European presence.
-Hypothesis testing on spatial randomness indicates a non-randomly dispersed distribution before European contact and a more extreme dispersed distribution after European presence.
+Hypothesis testing on spatial randomness (Figure \@ref(fig:Monte-Carlo-spatial-pattern)) indicates a non-randomly dispersed distribution before European contact and a more extreme dispersed distribution after European presence.
 In contrast, the distribution of pottery is more similar to random distributions during the Chinese period.
@@
 We might expect reduced numbers of potters to result in pottery production shrinking to a few locations in the settlement during this time.
-However, despite the small number of vessels during the Chinese contact period, Figure \@ref(fig:kernel-plot) shows that pottery is distributed randomly across the sampling area without any distinctive clusters during this time.
+However, despite the small number of vessels during the Chinese contact period, Figure \@ref(fig:Monte-Carlo-spatial-pattern) shows that pottery is distributed randomly across the sampling area without any distinctive clusters during this time.
 As population across our three occupation phases declines, we see less clustered distributions of pottery, supporting an interpretation of intentional standardization rather than mechanical standardization.
-The spatial pattern shows that ceramics were mostly household-produced, and no specific facilities of production are evident [@Chen2007].
+The spatial pattern shows that ceramics were mostly household-produced, and no specific facilities of production are evident [@Chen2007] (Figure \@ref(fig:kernel-plot)).
 
@@
 We would like to thank the Yilan County Cultural Affairs Bureau in Taiwan for permitting access to the pottery used in this study and providing the shape images.
-We thank Dr.
-Wen-Shan Chen in the Department of Geosciences, National Taiwan University for his invaluable guidance of petrographic analysis at his lab.
+We thank Wen-Shan Chen in the Department of Geosciences, National Taiwan University for his invaluable guidance of petrographic analysis at his lab.
 We thank the Quaternary Research Center funding for supporting the organic residue analysis in this project.
-We thank Dr.
-Julian Sachs in the Department of Oceanography, University of Washington for his supports and suggestions for conducting organic geochemistry analysis of potsherds.
-We thank Dr.
-Matthew Wolhowe for his help in developing protocols for lipid extraction and his assistance with the GC-MS, GC-FID, and GC-C-IRMS analyses.
+We thank Julian Sachs in the Department of Oceanography, University of Washington for his supports and suggestions for conducting organic geochemistry analysis of potsherds.
+We thank Matthew Wolhowe for his help in developing protocols for lipid extraction and his assistance with the GC-MS, GC-FID, and GC-C-IRMS analyses.
 This research used statistical consulting resources provided by the Center for Statistics and the Social Sciences, University of Washington.
-We also thank anonymous reviewers for their insightful comments that have greatly improved this manuscript.
+We also thank Ben Fitzhugh, Peter Lape, Robert Z.
+Selden Jr.
+and the anonymous reviewers for their insightful comments that have greatly improved this manuscript.
 

### 2020-09-13 analysis/paper/paper.Rmd
update inline code for stats

@@
 
-pval_all <- signif(unname(sum_man_all_df[,'Pr(>F)'][1]), 4)
+pillai <- signif(unname(sum_man_all_df[,"Pillai's trace"][1]), 4)
+F_value <- signif(unname(sum_man_all_df[,"Approximate F value"][1]), 4)
+df <- signif(unname(sum_man_all_df[,"degrees of freedom"][1]), 3)
+pval_all <- signif(unname(sum_man_all_df[,'Pr(>F)'][1]), 3)
 ```
@@
 This composition is consistent with the mineralogical composition of local raw materials found in the Yilan Plain [@Chen2016].
-There are no significant changes in the inclusions over time, indicating continuity in pottery fabric composition across the three periods (p = $`r pval_all`$) [@WangandMarwick2020].
+There are no significant changes in the inclusions over time, indicating continuity in pottery fabric composition across the three periods (Pillai's Trace = $`r pillai`$, F = $`r F_value`$, df = $`r df`$, p = $`r pval_all`$) [@WangandMarwick2020].
 

### 2020-09-18 analysis/paper/paper.Rmd
removed some repetitive figures

@@
 C: Correlation between shape variables (PC1 and PC2) and body diameter of vessels with Pearson's r and p-values by phases"}
-knitr::include_graphics(here("analysis/figures/size-boxplot-regression.jpg"))
+knitr::include_graphics(here("analysis/figures/plot-size-boxplot-regression.jpg"))
 ```

### 2021-01-14 analysis/paper/paper.Rmd
edit code chunk labels

@@
 
-```{r, KWL-map, fig.cap = "Map illustrating the location of Kiwulan, and other locations in northern Taiwan that are named in the text.
+```{r KWL-map, fig.cap = "Map illustrating the location of Kiwulan, and other locations in northern Taiwan that are named in the text.
 Map data is from naturalearthdata.com", fig.width=image_width_2_col_inch}
@@
 
-```{r, KWL-sam-area, fig.cap = "Map showing the largest section of excavation areas at Kiwulan, and the distribution of forty squares sampled in this paper presented in red with square ID number.
+```{r KWL-sam-area, fig.cap = "Map showing the largest section of excavation areas at Kiwulan, and the distribution of forty squares sampled in this paper presented in red with square ID number.
 Small dots represent the location of post-holes.
@@
 
-```{r}
+```{r thin-section-stats}
 library(readxl)
@@
 
-```{r, KWL-pots, fig.cap = "A typical pot from Kiwulan (left) and an example of a pottery drawing used for outline analysis (right)", fig.width=image_width_2_col_inch}
+```{r KWL-pots-image, fig.cap = "A typical pot from Kiwulan (left) and an example of a pottery drawing used for outline analysis (right)", fig.width=image_width_2_col_inch}
 knitr::include_graphics(here("analysis", "figures", "pots.png"))
@@
 
-```{r}
+```{r PCA-byplot}
 # PCA byplot, we save it to disk to use in the next chunk
@@
 
-```{r}
+```{r combine-multiple-plots}
 library(magick)
@@
 
-```{r, first-two-PCs-byplot-and-tps-plot, echo=FALSE, fig.cap='Left: Significant differences in average vessel shapes between the Chinese and the post-European period are visible using thin plate splines (TPS), with outline deformations required to pass from an extreme of one morphospace to another.
+```{r first-two-PCs-byplot-and-tps-plot, echo=FALSE, fig.cap='Left: Significant differences in average vessel shapes between the Chinese and the post-European period are visible using thin plate splines (TPS), with outline deformations required to pass from an extreme of one morphospace to another.
 Right: Pottery shape distribution by each occupation phase according to the first two PCs.'}
@@
 
-```{r, size-shape-regression-plots, echo=FALSE, fig.cap= "A: Distribution of the body diameter of vessels with coefficients of variation by phases.
+```{r size-shape-regression-plots, echo=FALSE, fig.cap= "A: Distribution of the body diameter of vessels with coefficients of variation by phases.
 B: Coefficients of variation of vessel shape represented by PC1 in relation to vessel size represented by body diameter, showing more standardization (lower CV values) in vessel form and size in the Chinese period.
@@
 
-```{r}
+```{r git-commit-file}
 # what commit is this file at? 

### 2021-05-30 analysis/paper/KWL-chronology-modeling.Rmd
update the figure of chronological model

@@
   select(`Lab code`, "rce", "se", "phase", Context)
+```
+
+```{r edit-calibrate-function, include = FALSE}
+print.oxcAARCalibratedDatesList <- function(x, ...){
+  n <- length(x)
+  if (n == 1) print(x[[1]],...)
+  else{
+    cat(sprintf("List of %d calibrated dates:\n",n))
+    for (i in 1:n) {
+      print(x[[i]])
+    }
+  }
+}
+
+#' @export
+plot.oxcAARCalibratedDatesList <- function(x, use_ggplot = T, ...){
+  if (length(x) == 1) plot(x[[1]], use_ggplot, ...)
+  else{
+    if (requireNamespace("ggplot2", quietly = TRUE) & use_ggplot) {
+      plotoxcAARCalibratedDatesListGGPlot2(x, ...)
+    } else {
+      plotoxcAARCalibratedDatesListSystemGraphics(x, ...)
+    }
+  }
+}
+
+plotoxcAARCalibratedDatesListGGPlot2<-function(x, ...){
+  .data <- NULL
+  to_plot<-lapply(x,function(y){
+    alpha <- ifelse((!(is.null(y$posterior_probabilities))& !(is.na(y$posterior_probabilities))),
+                    c(0.25,0.75),
+                      c(0.75,0)
+                      )
+    outdata <- data.frame(
+      dates=y$raw_probabilities$dates,
+      probability=y$raw_probabilities$probabilities,
+      name=y$name,
+      class = "unmodelled",
+      alpha=alpha[1])
+    if(!(is.null(y$posterior_probabilities)) & !(is.na(y$posterior_probabilities))) {
+      outdata <- rbind(outdata,
+                   data.frame(
+                     dates=y$posterior_probabilities$dates,
+                     probability=y$posterior_probabilities$probabilities,
+                     name=y$name,
+                     class = "modelled",
+                     alpha=alpha[2])
+      )
+    }
+    return(outdata)
+
+  })
+  to_plot <- do.call(rbind,to_plot)
+
+  m <- ggplot2::ggplot(to_plot)
+
+  graph <- m + ggridges::geom_ridgeline(ggplot2::aes(x = .data$dates,
+                                  y = .data$name,
+                                  height = .data$probability,
+                                  alpha = .data$alpha,
+                                  group = interaction(.data$class,.data$name)),
+                              scale=100,
+                              fill = "#fc8d62",
+                              color="#00000077") +
+    ggplot2::theme_light() + ggplot2::labs(y="Dates", x = "AD")  +
+    ggplot2::scale_alpha_continuous(guide = FALSE)
+
+  methods::show(graph)
+}
+
+plotoxcAARCalibratedDatesListSystemGraphics <- function(x, ...){
+  op <- graphics::par(no.readonly = TRUE)
+  years <- years_post <- NA
+
+  indices <- 1:length(x)
+
+  min_year <- min(
+    sapply(
+      indices,
+      function(i) {
+        this_year_range <- get_years_range(x[[i]])
+        if (!all(is.na(this_year_range))) {
+          min(this_year_range, na.rm=T)
+        } else {
+          NA
+        }
+      }
+    ), na.rm=T
+  )
+
+  max_year <- max(
+    sapply(
+      indices,
+      function(i) {
+        this_year_range <- get_years_range(x[[i]])
+        if (!all(is.na(this_year_range))) {
+          max(this_year_range, na.rm=T)
+        } else {
+          NA
+        }
+      }
+    ), na.rm=T
+  )
+
+  graphics::par(mfrow=c(length(x)+1,1))
+
+  graphics::par(oma = c(3,1,2,2) + 0.1,
+                mar = c(0,1,0,1) + 0.1)
+
+  for (i in indices) {
+    max_prob <- 0
+    years <- probability <- NA
+    post_present <- prob_present <- FALSE
+
+    if(class(x[[i]]$raw_probabilities)=="data.frame") {
+      prob_present <- TRUE
+      years <- x[[i]]$raw_probabilities$dates
+      probability <- x[[i]]$raw_probabilities$probabilities
+      max_prob <- max(probability)
+    }
+
+    unmodelled_color <- "lightgrey"
+
+    years_post <- probability_post <- NA
+
+    if(class(x[[i]]$posterior_probabilities)=="data.frame") {
+      post_present <- TRUE
+      years_post <- x[[i]]$posterior_probabilities$dates
+      probability_post <- x[[i]]$posterior_probabilities$probabilities
+      unmodelled_color <- "#eeeeeeee"
+      max_prob <- max(max_prob, probability_post)
+    }
+
+    if(!prob_present & !post_present)
+    {
+      year_range <-c(0,1)
+    } else {
+      year_range <- get_years_range(x[[i]])
+    }
+
+    graphics::plot(
+      years,
+      probability,
+      type = "n",
+      ylim=c(max_prob / 7 * -1,max_prob),
+      xlim = c(min_year,max_year),
+      axes = FALSE
+    )
+    graphics::axis(side=4)
+    graphics::polygon(years, probability, border = "black", col = unmodelled_color)
+    if (unmodelled_color!="lightgrey"){
+      graphics::polygon(years_post, probability_post, border = "black", col = "#aaaaaaaa")
+    }
+    graphics::text(x=min_year, y=max_prob, labels=print_label(x[[i]]),las=2,cex=0.6, adj=0)
+    graphics::grid()
+  }
+  plot(c(min_year,max_year),c(0,0),
+       axes = FALSE,
+       type="n")
+  graphics::axis(side=1, pretty(c(min_year, max_year)))
+  graphics::par(op)
+}
 ```
@@
 three_phases_con_text <- readOxcalOutput(three_phases_con_results)
-three_phases_result_con_data <- parseOxcalOutput(three_phases_con_text, only.R_Date = F)
+three_phases_result_con_data <- parseOxcalOutput(three_phases_con_text, only.R_Date = T) 
 # str(three_phases_result_con_data)

### 2023-09-04 analysis/paper/paper.Rmd
checking text labels - they are correct

@@
   geom_text(data = cv_pcs_all,
-            mapping = aes(x = c(2.1, 1.1, 3.1, 2.1, 1.1, 3.1, 2.1, 1.1, 3.1), 
+            mapping = aes(x = c(2.1,   # upper middle 
+                                1.1,   # upper left
+                                3.1,   # upper right
+                                2.1,   # middle middle
+                                1.1,   # middle left
+                                3.1,   # middle right
+                                2.1,   # lower middle
+                                1.1,   # lower left
+                                3.1),  # lower right
                           y = 8.5, 
